### 2020-01-24 paper.Rmd
Added rct paper

---
title: "Historical advances in medicine that did not need (or want) a randomized clinical trial."
author: "Steve Simon"
output: html_document
bibliography: reference.bib
nocite: |
  @zika-overview, @zika-case-series
---

#### Abstract

Understanding when randomized clinical trials (RCTs) are needed comes from first knowing when those RCTs are not needed. There are many settings where major improvements in health care came through the use of observational studies. Observational studies have some strengths, such as greater ecologic validity and the ability to study larger populations over longer periods of time, but they have many weaknesses and require much more effort to eliminate other possible causal explanation. Some examples of the triumphs of observational research are establishment of a causal association between cigarette smoking and lung cancer, identification of most of the well-known risk factors for heart disease, and discovery of the link between aspirin and Reye’s syndrome. The work was not easy; but the careful and methodical approach offers a lesson in how to assemble information from a series of less than perfect observational studies to produce a definitive conclusion. These examples help to teach students to recognize when the ethical and logistical challenges associated with RCTs are so great that observational studies, with all their limitations, are preferable.

#### Introduction

In October 2015, health experts in Brazil noticed a sharp increase in the number of cases of microcephaly. This tragic event, where children are born with unusually small head circumferences and severe developmental disabilities, seemed to be geographically associated with the concurrent spread of a new virus, Zika. In response, researchers initiated a host of research studies. Among these were a prospective cohort study of 345 pregnant women in Rio de Janiero who exhibited symptoms associated with Zika infection (@zika-cohort), a case series of 24 infants diagnosed with congenital Zika syndrome at birth and followed for the next two years of their lives (@zika-case-series), a case-control study of 32 infants born with microcephaly compared to 62 normal births matched by time of delivery and location of residence (@zika-case-control), among many others. Researchers utilized a wide range of surveillance systems and registries (@zika-overview).

What was lacking in all the research studies utilized in the war against Zika, was the randomized control trial (RCT). An informal Pubmed search on the terms “Zika” and “microcephaly” yielded 1,716 publications, but when you click on the link to the right that limits your search to clinical trials, the list dwindles to a single citation that on closer examination is actually a cohort study.

It is not at all surprising that no RCTs were utilized in this investigation. What is surprising, at least to some, is how much was learned so quickly utilizing a range of research methodologies that are commonly held to be inferior to the RCT. In a few short years, researchers were able to track the spread of the Zika virus, identify the modes of transmission, provide a causal link between Zika infection during pregnancy and microcephaly, and characterize in great detail the impact of this condition on early childhood development.

If you want to understand the benefits of RCTs and when they are needed, you should start by examining the times that they have not been needed. This paper will outline three areas where major progress was made without the benefit of RCTs and discuss what this means for the use of RCTs in research.

#### Cigarette smoking and lung cancer

Perhaps the most commonly cited discovery that did not use RCTs was the research that drew a causal link between cigarette smoking and cancer. It is still instructive, however, to review the series of studies that led to the 1964 Surgeon General's report on smoking and health. Although some evidence appeared earlier, the concern about smoking rose largely from five case-control studies, all published in 1950 (@report-anniversary). This research approach was very new, but the authors took great pains to control for recall bias and confounding (@statistics-in-action).

It took more than these five studies to develop a strong case against smoking, but it also required a different way of thinking about research. The authors of these case-control studies were well aware of the potential problems associated with sampling from cases and controls rather than sampling from the general population. In 1951, however, Jerome Cornfield took these studies and placed them on a rigorous mathematical foundation in his landmark paper on case-control studies (@cornfield-case-control).

#### Bibliography

### 2020-01-25 paper/paper.Rmd
Reorganized rct not needed documents

--- 2020-01-24 paper.Rmd
@@
 ---
-title: "Historical advances in medicine that did not need (or want) a randomized clinical trial."
+title: Historical advances in medicine that did not need (or want) a randomized clinical
+  trial.
 author: "Steve Simon"
-output: html_document
-bibliography: reference.bib
 nocite: |
   @zika-overview, @zika-case-series
+output:
+  word_document: default
+  html_document: default
+bibliography: reference.bib
 ---
@@
 
-Perhaps the most commonly cited discovery that did not use RCTs was the research that drew a causal link between cigarette smoking and cancer.
+Perhaps the most commonly cited discovery that did not use RCTs was the research that drew a causal link between cigarette smoking and lung cancer.
 It is still instructive, however, to review the series of studies that led to the 1964 Surgeon General's report on smoking and health.
-Although some evidence appeared earlier, the concern about smoking rose largely from five case-control studies, all published in 1950 (@report-anniversary).
+Although some evidence appeared earlier, the concern about smoking rose largely from five case-control studies, all published in 1950 (@doll-hill-case-control , @levin-case-control, @mills-porter-case-control, @schrek-baker-case-control, @wynder-graham-case-control).
 This research approach was very new, but the authors took great pains to control for recall bias and confounding (@statistics-in-action).
@@
 
+Around the same time, researchers set up prospective cohort studies to further investigate the link between smoking and lung cancer (@doll-hill-cohort, @dorn-cohort, @hammond-horn-cohort).
+These studies were massive (more than 460,000 patients total across the three studies) and followed these patients for multiple years.
+Needless to say, these studies took much longer to complete, but provided a key piece of additional evidence (@report-anniversary).
+
+The researchers received a lot of criticism, of course, and not just from tobacco interests.
+R.A.
+Fisher, perhaps the most prominent statistician of the era, published a series of stinging critiques of the research (@fisher-pamphlet, @fisher-smoking-letter-1, @fisher-smoking-letter-2, @fisher-smoking-letter-3).
+It is difficult to parse these criticisms.
+One source (@fisher-error) attributes it a largely unsupportable explanation rooted in genetics combined with a focus on small details that failed to support a smoking-cancer link while ignoring the totality of the evidence.
+
+The controversy inspired one of the prominent researchers in this area, Sir Austin Bradford Hill, to delineate his nine famous principles for establishing causation (@hill-criteria).
+
 #### Bibliography

### 2020-01-25 paper/paper.Rmd
Added Framingham to paper

@@
 
+#### Risk factors for cardiovascular disease
+
+While many pharmaceutical and surgical interventions for cardiovascular disease used RCTs, the identification of risk factors for this condition required the use of a large observational cohort, the Framingham Study.
+This study originally comprised roughly 5,000 men and women in the city of Framingham, Massachusetts between the ages of 30 and 59 with a planned follow-up of 20 years (@framingham-design).
+It is worth noting that the inclusion of women in this study was, in that era, a rather unusual feature.
+At the end of twenty years, the study was almost ended, but a private funding effort and a concerted lobbying campaign led to the renewal and expansion of the Framingham study (@framingham-epidemiology).
+This expansion included inclusion of the children of the original participants and eventually the children of those children.
+Two contemporary cohorts of minority participants were also recruited to reflect the changing demographics of the city (@framingham-legacy).
+
+The Framingham study provided insights beyond cardiovascular disease. A 1964 study (@framingham-smoking) provided additional support to the link between smoking
+
 #### Bibliography

### 2020-01-25 paper/paper.Rmd
Finished section on Framingham

@@
 
-The Framingham study provided insights beyond cardiovascular disease. A 1964 study (@framingham-smoking) provided additional support to the link between smoking
+An early finding of the Framingham study, after the four year follow-up visit (@framingham-hypertension) established hypertension, obesity, and hypercholesteremia as risk factors for arteriosclerotic heart disease (a composite measure of myocardial infarction, angina pectoris, coronary occlusion, or myocardial fibrosis).
+In 1964, researchers used the Framingham cohort combined with a second cohort in Albany, New York  to establish a link between smoking and cardiovascular disease (@framingham-smoking) and contributed vital information to the 1964 surgeon general's report.
+Further work elaborated on the greater importance of the systolic measurement of blood pressure on the risk of heart disease (@framingham-systolic).
+
+The number of findings from the Framingham study goes on and on, but more important than the number of studies is the way that this study changed clinical practice.
+The Framingham study changed the medical community's attitudes from the belief that heart disease as something to be treated to the belief that heart disease was something that could be prevented (@framingham-epidemiology).
+
+#### Aspirin and Reye's syndrome
+
+Reye's syndrome is rare disorder seen mostly in children. It can produce serious intracrainal swelling that can lead to serious neurological damage and possibly death.
 

### 2020-01-25 paper/paper.Rmd
Started section on Reyes syndrome

@@
 
-Reye's syndrome is rare disorder seen mostly in children. It can produce serious intracrainal swelling that can lead to serious neurological damage and possibly death.
+Reye's syndrome is rare disorder seen mostly in children.
+It can produce serious intracrainal swelling that can lead to serious neurological damage and possibly death (@reyes-causation).
+The disease was not well recognized until the 1970s and was very difficult to characterize accurately (@reyes-disappearance).
+
+The Centers for Disease Control developed a surveillance system for Reye's syndrome in 1980 to track the number of cases of Reye's syndrome in the United States and to collect information from the patient's family and combine that with laboratory results (@reyes-history).
 

### 2020-01-26 paper/paper.Rmd
Finished first draft of paper

@@
 author: "Steve Simon"
-nocite: |
-  @zika-overview, @zika-case-series
 output:
@@
 bibliography: reference.bib
+csl: american-medical-association.csl
 ---
@@
 The work was not easy; but the careful and methodical approach offers a lesson in how to assemble information from a series of less than perfect observational studies to produce a definitive conclusion.
-These examples help to teach students to recognize when the ethical and logistical challenges associated with RCTs are so great that observational studies, with all their limitations, are preferable.
+Understanding the circumstances where observational studies shine will help to teach students to take a balanced approach to research.
+At the same time, defining the times when where RCTs are not needed will help define better by negation when RCTs are needed.
 
@@
 What was lacking in all the research studies utilized in the war against Zika, was the randomized control trial (RCT).
-An informal Pubmed search on the terms “Zika” and “microcephaly” yielded 1,716 publications, but when you click on the link to the right that limits your search to clinical trials, the list dwindles to a single citation that on closer examination is actually a cohort study.
-
 It is not at all surprising that no RCTs were utilized in this investigation.
@@
 If you want to understand the benefits of RCTs and when they are needed, you should start by examining the times that they have not been needed.
-This paper will outline three areas where major progress was made without the benefit of RCTs and discuss what this means for the use of RCTs in research.
+There are wonderful historical reviews of the research that first drew a link between smoking and lung cancer (@statistics-in-action, @report-anniversary), that established major risk factors for heart disease (@framingham-epidemiology, @framingham-legacy), and that identified aspirin use during viral infections as the predominant cause of Reye's syndrome (@reyes-history).
+This paper will outline three areas how these research areas relied on a broad range of observational studies and made major progress was made without the benefit of RCTs.
+This review has broad implications for when RCTs should be used in research.
 
@@
 
+Around the same time, several case control studies were being published.
+The results of one study (@reyes-case-control-2) are representative of the general findings.
+The researchers identified 159 cases of Reye's syndrome following an illness like influenze or varicella.
+Recruiting was done at six pediatric centers in Ohio between December 1978 and March 1980.
+Controls were recruited from the same school and matched on age, sex, and race and had an antecedant illness within one week of the case.
+The use of salicylates (aspirin) among the cases was significantly higher (97% versus 71%) among the cases.
+
+While the case control studies alone may not have provided a definitive link between aspirin and Reye's syndrome, it was sufficiently persuasive to start a publicity effort warning the public about the possible danger associated with aspirin use in a child during a viral illness.
+What happened next was definitive proof.
+The number of cases reported to the CDC surveillance system dropped steadily from a high of 555 cases in 1980 to no more than 2 per year in 1994-1997 and the drops exhibited a strong temporal association with public health campaign events (@reyes-history, Figure 1).
+
+The near total elimination of Reye's syndrome has been properly hailed as a public health triumph (@reyes-disappearance) and was accomplished without any RCTs.
+The humble case-control study combined with temporal trends in disease surveillance were all that was needed.
+
+#### Discussion
+
+The recent research effort to examine the link between the Zika virus and microcephaly mirrors past research efforts in lung cancer, heart disease, and Reye's syndrome.
+In every case, major advances in knowledge were accomplished using only observational studies.
+
+The effective use of observational studies is not limited to just these areas.
+Benjamin Djulbegovic has developed an web page inventory of non-randomized trials that have changed clinical practice (http://personal.health.usf.edu/bdjulbeg/oncology/NON-RCT-practice-change.htm) that covers studies in Endocrinology, Infectious Diseases, Oncology, Hematology, Pulmonology, Urology, Cardiology, Musculoskeletal conditions, Psychiatry, and Surgery.
+
+The research efforts described in this paper share some common features that can help us understand when RCTs are not needed.
+The most obvious feature is the practical and ethical barriers to randomization.
+No one could envision a study where some children with a viral fever are given aspirin to see if any harm could come from it.
+The long latency periods associated with lung cancer and heart disease would saddle most RCTs with an unworkable time frame.
+
+Second, these observational studies represented pioneering efforts in an area where little is known.
+This is illustrated best, perhaps, by the Framingham study.
+The original planners of this study had considered using the population of Framingham, Massachusetts as a testbed for trying new interventions, but rejected that approach in favor of an observational study because there weren't any interventions at the time that warranted testing (@framingham-epidemiology).
+Today there are many RCTs testing a variety of prevention initiatives, but these are only possible because of the foundational knowledge that Framingham provided.
+
+Third, these studies were looking to attack problems by taking big bites rather than nibbling around the edges.
+The odds ratios in the lung cancer studies were showing a 10 to 20 fold increase in risk among smokers.
+The finding of aspirin use in the medical history of almost every Reye's syndrome case was a veritable smoking gun and made any concerns about biases in a case-control study seem unimportant.
+
+A final commonality among these studies is the hard work that researchers put in place to insure that these observational studies would produce results that could withstand critical scrutiny.
+Elaborate matching plans helped to insure a comparable control group and efforts to use blinding during patient interviews helped to reduce biases from researcher expectations.
+The recent studies of the Zika virus show that same level of care.
+One example is the careful consideration of what specific head circumference measurements would be considered as evidence of microcephaly (@zika-interpret).
+
+It's been said many times how much easier the conduct of an RCT is.
+Get a bunch of patients together, flip a coin to see who gets the drug and who gets the placebo, wait three years and then count the number of dead bodies.
+Watching the care and concern of the researchers in these observational studies makes one hope that some of this could rub off on the researchers running RCTs.
+While RCTs may have fewer and less severe biases than observational studies, they do indeed have biases.
+
+So let's flip the coin over to the other side.
+If observational studies are what we want when studying the link between smoking and lung cancer, the risk factors for heart disease, the cause of Reye's syndrome, and the microcephaly associated with the Zika virus, then when do we want to use RCTs instead?
+
+RCTs do very well in a mature setting where we know a lot about the causes of disease and can develop some well supported interventions.
+They are great at developing and testing a series of small incremental improvements that nibble around the edges.
+If you do enough nibbling, you can make some big improvements over time.
+
+The RCT excels when you are in the muddled middle.
+You have some interventions that sort of work but there's hope that you might find something a bit better.
+When comparing a fair intervention to a good intervention, biases become very troublesome, and the ability to remove many of these biases through randomization helps to insure that you'll recognize the distinction between fair and good.
+
+I once worked with a statistician who told me that he would never work on a study that didn't use randomization.
+My first thought was "You have some many clients asking for your help that you can send the half away that come to you with observational studies?" But I later came to realize that this was a type of arrogance.
+We can have debates about how observational studies have greater ecologic validity or how randomized studies control for many biases but this is really not the point.
+The type of problem you are studying should determine what type of research you choose.
+You don't want to be the proverbial worker who sees nothing but nails because the only tool you have is a hammer.
+
+#### Conclusion
+
+Research efforts in several major areas have not used RCTs.
+A review of how these developed important insights and make great progress helps you to understand the settings where observational studies are needed and RCTs are not.
+This also helps you understand where RCTs are truly needed.
+
 #### Bibliography

### 2020-01-29 paper/paper.Rmd

@@
 If you want to understand the benefits of RCTs and when they are needed, you should start by examining the times that they have not been needed.
-There are wonderful historical reviews of the research that first drew a link between smoking and lung cancer (@statistics-in-action, @report-anniversary), that established major risk factors for heart disease (@framingham-epidemiology, @framingham-legacy), and that identified aspirin use during viral infections as the predominant cause of Reye's syndrome (@reyes-history).
+There are wonderful historical reviews of the research that first drew a link between smoking and lung cancer (@statistics-in-action, @smoking-anniversary), that established major risk factors for heart disease (@framingham-epidemiology, @framingham-legacy), and that identified aspirin use during viral infections as the predominant cause of Reye's syndrome (@reyes-history).
 This paper will outline three areas how these research areas relied on a broad range of observational studies and made major progress was made without the benefit of RCTs.
@@
 These studies were massive (more than 460,000 patients total across the three studies) and followed these patients for multiple years.
-Needless to say, these studies took much longer to complete, but provided a key piece of additional evidence (@report-anniversary).
+Needless to say, these studies took much longer to complete, but provided a key piece of additional evidence (@smoking-anniversary).
 
@@
 The controversy inspired one of the prominent researchers in this area, Sir Austin Bradford Hill, to delineate his nine famous principles for establishing causation (@hill-criteria).
+These criteria are still invoked today when researchers debate the implications of new observational studies.
+In truth, these criteria can and should be considered when debating the implications of new RCTs as well.
 
@@
 While many pharmaceutical and surgical interventions for cardiovascular disease used RCTs, the identification of risk factors for this condition required the use of a large observational cohort, the Framingham Study.
+It would be unfair to characterize this study as the only large observational cohort that contributed to our knowledge of cardiovascular disease, but the Framingham study stands out by its pioneering efforts and its dogged determination to persist over more than six decades.
+
 This study originally comprised roughly 5,000 men and women in the city of Framingham, Massachusetts between the ages of 30 and 59 with a planned follow-up of 20 years (@framingham-design).
 It is worth noting that the inclusion of women in this study was, in that era, a rather unusual feature.
+Also unusual in its time was the use of community councils to shape the features of the study (@framingham-epidemiology), an early recognition of the value of community based participatory research.
+
 At the end of twenty years, the study was almost ended, but a private funding effort and a concerted lobbying campaign led to the renewal and expansion of the Framingham study (@framingham-epidemiology).
@@
 An early finding of the Framingham study, after the four year follow-up visit (@framingham-hypertension) established hypertension, obesity, and hypercholesteremia as risk factors for arteriosclerotic heart disease (a composite measure of myocardial infarction, angina pectoris, coronary occlusion, or myocardial fibrosis).
-In 1964, researchers used the Framingham cohort combined with a second cohort in Albany, New York  to establish a link between smoking and cardiovascular disease (@framingham-smoking) and contributed vital information to the 1964 surgeon general's report.
+In 1964, researchers used the Framingham cohort combined with a second cohort in Albany, New York  to establish a link between smoking and cardiovascular disease (@framingham-smoking) and contributed vital non-cancer information to the 1964 surgeon general's report mentioned earlier.
 Further work elaborated on the greater importance of the systolic measurement of blood pressure on the risk of heart disease (@framingham-systolic).
 
-The number of findings from the Framingham study goes on and on, but more important than the number of studies is the way that this study changed clinical practice.
+The number of findings from the Framingham study goes on and on, but more important than the number of studies is the way that this study influenced clinical practice.
 The Framingham study changed the medical community's attitudes from the belief that heart disease as something to be treated to the belief that heart disease was something that could be prevented (@framingham-epidemiology).
@@
 Reye's syndrome is rare disorder seen mostly in children.
-It can produce serious intracrainal swelling that can lead to serious neurological damage and possibly death (@reyes-causation).
+It can produce serious intracrainal swelling that sometimes leads to serious neurological damage and possibly death (@reyes-causation).
 The disease was not well recognized until the 1970s and was very difficult to characterize accurately (@reyes-disappearance).
@@
 The effective use of observational studies is not limited to just these areas.
-Benjamin Djulbegovic has developed an web page inventory of non-randomized trials that have changed clinical practice (http://personal.health.usf.edu/bdjulbeg/oncology/NON-RCT-practice-change.htm) that covers studies in Endocrinology, Infectious Diseases, Oncology, Hematology, Pulmonology, Urology, Cardiology, Musculoskeletal conditions, Psychiatry, and Surgery.
+Benjamin Djulbegovic has developed an web page inventory of non-randomized trials that have changed clinical practice (http://personal.health.usf.edu/bdjulbeg/oncology/NON-RCT-practice-change.htm) that covers studies in Cardiology, Endocrinology, Hematology, Infectious Diseases, Musculoskeletal conditions, Oncology, Psychiatry, Pulmonology, Surgery, and Urology.
 
@@
 
-Second, these observational studies represented pioneering efforts in an area where little is known.
+You can't stop here, however, and say that observational studies are a clearly inferior design that we only fall back on when the RCTs is impractical or unethical.
+Observational studies shine in areas where RCTs cannot.
+
+A common thread to these observational studies is that are relied upon when you are pioneering efforts in an area where little at that time was known.
 This is illustrated best, perhaps, by the Framingham study.
@@
 
-Third, these studies were looking to attack problems by taking big bites rather than nibbling around the edges.
+Another common thread to these studies is the magnitude of the effects they were looking for.
+These researchers wanted to take big bites rather than nibbling around the edges.
 The odds ratios in the lung cancer studies were showing a 10 to 20 fold increase in risk among smokers.
@@
 Elaborate matching plans helped to insure a comparable control group and efforts to use blinding during patient interviews helped to reduce biases from researcher expectations.
+Researchers adopted a variety of different observational approaches, recognizing that while all observational have biases, it is unlikely that a broad range of different approaches would suffer from a common flaw.
+
 The recent studies of the Zika virus show that same level of care.
 One example is the careful consideration of what specific head circumference measurements would be considered as evidence of microcephaly (@zika-interpret).
+The Zika researchers are not afraid to deploy case series studies, case control designs, observational cohorts, registries, and surveillance to attack the research problem from every possible front.
 
-It's been said many times how much easier the conduct of an RCT is.
-Get a bunch of patients together, flip a coin to see who gets the drug and who gets the placebo, wait three years and then count the number of dead bodies.
-Watching the care and concern of the researchers in these observational studies makes one hope that some of this could rub off on the researchers running RCTs.
-While RCTs may have fewer and less severe biases than observational studies, they do indeed have biases.
-
-So let's flip the coin over to the other side.
+Now let's flip the coin over to the other side.
 If observational studies are what we want when studying the link between smoking and lung cancer, the risk factors for heart disease, the cause of Reye's syndrome, and the microcephaly associated with the Zika virus, then when do we want to use RCTs instead?
@@
 They are great at developing and testing a series of small incremental improvements that nibble around the edges.
-If you do enough nibbling, you can make some big improvements over time.
+If you do enough nibbling, you can make some pretty big improvements over time.
 
@@
 When comparing a fair intervention to a good intervention, biases become very troublesome, and the ability to remove many of these biases through randomization helps to insure that you'll recognize the distinction between fair and good.
+There are small but important improvements that you can make to an approach that is already somewhat effective by modifying a dosing schedule, substituting a less toxic drug to the current treatment regimen, and examining the synergism of combining two different treatment approaches.
+RCTs excel here because their ability to remove many sources of bias allow you to detect subtle signals amidst all the noise.
 
 I once worked with a statistician who told me that he would never work on a study that didn't use randomization.
-My first thought was "You have some many clients asking for your help that you can send the half away that come to you with observational studies?" But I later came to realize that this was a type of arrogance.
+My first thought was "You have some many clients asking for your help that you can send the half away that come to you with observational studies?" But I later came to realize that this was a type of arrogance, a research chauvinism that says that my approach is the only true path to knowledge.
 We can have debates about how observational studies have greater ecologic validity or how randomized studies control for many biases but this is really not the point.
@@
 You don't want to be the proverbial worker who sees nothing but nails because the only tool you have is a hammer.
+Both observational studies and RCTs have their place in the research toolbox.
 
@@
 Research efforts in several major areas have not used RCTs.
-A review of how these developed important insights and make great progress helps you to understand the settings where observational studies are needed and RCTs are not.
-This also helps you understand where RCTs are truly needed.
+A review of how these developed important insights and make great progress helps you to understand the settings where observational studies by themselves are sufficient.
+Knowing when RCTs are not wanted also helps you understand when RCTs are truly needed.
 

### 2020-02-02 src/prelims.R
Added skim_md_files function

@@ -337,13 +337,34 @@ flag_unused_yaml_fields <- function(fields, v=TRUE) {
 
 skim_yaml_files <- function(field_header, dir_root="../source/posts", file_pattern="*.md", v=TRUE) {
   file_list <- build_file_list(dir_root, file_pattern)
+  if (v) "\nScanning through" %b% length(file_list) %b% "files." %>% cat
   for (i_file in file_list) {
     tx <- read_lines(i_file)
     fields <- parse_yaml(tx, i_file)
     if(v) i_file %C% fields[[field_header]] %>% br %>% cat
-    if(is.null(fields[[field_header]])) print(fields)
+    # if(is.null(fields[[field_header]])) print(fields)
   }
 }
+
+# This function skims through the text of
+# markdown files and prints lines matching
+# a specified string.
+
+skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v=TRUE) {
+  file_list <- build_file_list(dir_root, file_pattern)
+  if (v) "\nScanning through" %b% length(file_list) %b% "files." %>% cat
+  # file_list <- sample(file_list)
+  for (i_file in file_list) {
+    tx <- read_lines(i_file)
+    tx %>% 
+      str_extract_all(search_string) %>% 
+      unlist -> found_lines
+    if(v & length(found_lines > 1)) "\n" %0% i_file %C% paste(found_lines, collapse=", ") %>% cat
+  }
+  return(tx)
+}
+
+# tx <- skim_md_files("!\\[\\]\\(.*?\\)")
 
 clean_files <- function(search_string, replace_string="Not yet", dir_root="text", file_pattern="*.md", v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE  

### 2020-02-02 src/prelims.R
Started image transfer

@@ -350,6 +350,41 @@ skim_yaml_files <- function(field_header, dir_root="../source/posts", file_patte
 # markdown files and prints lines matching
 # a specified string.
 
+pull_images <- function(i_file, v=TRUE) {
+  tx <- read_lines(i_file)
+  if (v) "Reading" %b% length(tx) %b% "lines. " %>% cat
+  tx %>% 
+    str_extract_all("!\\[\\]\\(.*\\)") %>% 
+    unlist  %>%
+    str_remove("!.*/") %>%
+    str_remove("\\)") -> found_lines
+  if (v) {found_lines %>% str_c(collapse=", ") %>% cat}
+  return(found_lines)
+}
+
+move_images <- function(i_yr, i_mo, i_file, image_list, v=TRUE) {
+  n_images <- length(image_list)
+  i_file %>% str_remove(fixed(".md")) -> root_name
+  tx <- readLines("text" %s% i_yr %s% i_mo %s% i_file)
+  if (v) "Processing" %b% n_images %b% "images in" %b% length(tx) %b% "lines of text." %>% cat
+  for (j in 1:n_images) {
+    i_image <- image_list[j]
+    match_found <- file.exists("../web/images" %s% i_image)
+    old_image_name <- ifelse(match_found, i_image, "image-not-found.png")
+    old_image_name %>% str_remove(".*\\.") -> image_type
+    new_image_name <- root_name %0% zpad(j) %d% image_type
+    if (v) "\n  " %0% old_image_name %b% ">>" %b% new_image_name %>% cat
+    if (!match_found) "." %b% i_image %b% "not found" %>% cat
+    file.copy("../web/images" %s% old_image_name, "../web/images" %s% i_yr %s% new_image_name)
+    tx %>% str_which(fixed(i_image)) %>% unlist -> i_line
+    if (v) "." %b% str_c(i_line, collapse=";") %>% cat
+    tx[i_line] <- "![](../web/images" %s% i_yr %s% new_image_name
+  }
+  writeLines(tx, "text" %s% i_yr %s% i_file)
+}
+
+
+
 skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v=TRUE) {
   file_list <- build_file_list(dir_root, file_pattern)
   if (v) "\nScanning through" %b% length(file_list) %b% "files." %>% cat
@@ -359,7 +394,7 @@ skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v
[…]

### 2020-02-02 src/step0-identify-images.R

@@ -1,0 +1,28 @@
+# step0-identify-images.R
+# written by Steve Simon
+# created 2020-02-02
+
+# This program identifies the images in
+# each markdown file.
+
+if (!exists("v")) v <- TRUE
+
+source(file="src/prelims.R")
+for (i_yr in c("99")) {
+  for (i_mo in zpad(1:12)) {
+    file_list <- list.files(path="text" %s% i_yr %s% i_mo)
+    new_path <- "text" %s% i_yr
+    for (i_file in file_list) {
+      i_path <- "text" %s% i_yr %s% i_mo
+      if (v) "\n" %0% i_path %s% i_file %0% ": " %>% cat
+      image_list <- pull_images(i_path %s% i_file, v=FALSE)
+      n_images <- length(image_list)
+      if (n_images==0) next
+      move_images(i_yr, i_mo, i_file, image_list)
+    }
+  }
+}
+
+# Save everything.
+
+save.image("data/step0.RData")

### 2020-02-03 src/prelims.R
Looked for nonstandard image tags

@@ -354,7 +354,7 @@ pull_images <- function(i_file, v=TRUE) {
   tx <- read_lines(i_file)
   if (v) "Reading" %b% length(tx) %b% "lines. " %>% cat
   tx %>% 
-    str_extract_all("!\\[\\]\\(.*\\)") %>% 
+    str_extract_all("!\\[.*?\\]\\(.*?\\)") %>% 
     unlist  %>%
     str_remove("!.*/") %>%
     str_remove("\\)") -> found_lines
@@ -378,7 +378,8 @@ move_images <- function(i_yr, i_mo, i_file, image_list, v=TRUE) {
     file.copy("../web/images" %s% old_image_name, "../web/images" %s% i_yr %s% new_image_name)
     tx %>% str_which(fixed(i_image)) %>% unlist -> i_line
     if (v) "." %b% str_c(i_line, collapse=";") %>% cat
-    tx[i_line] <- "![](../web/images" %s% i_yr %s% new_image_name
+    msg <- ifelse(match_found, "", i_image %b% "not found.")
+    tx[i_line] <- "![" %0% msg %0% "](../../../web/images" %s% i_yr %s% new_image_name %0% ")"
   }
   writeLines(tx, "text" %s% i_yr %s% i_file)
 }

### 2020-02-04 text/20/non-breaking-space.Rmd
Work on bib images

---
title: "Blog post: Identifying and manipulating non-breaking spaces"
author: "Steve Simon"
source: "New"
date: "2020-01-07"
category: Blog post
tags: Data management
output: html_document
---

Non-breaking spaces are one of those weird little things in the computer that you have to keep your eye on. They can cause a lot of problems if they are assumed to be normal spaces. I found a non-breaking space "out in the wild" and decided to use it as an opportunity to explore what it was and better prepare myself for when I encounter more of these beasts.

<!---More--->

I've been working with some markdown files which were converted from html files and at the bottom of one of these files, I noticed something strange. It looked not at all unusual in notepad (see below)

![![Figure 1. Screenshot of a file inside Notepad](non-breaking-space01.png not found.](../../../web/images/20/non-breaking-space.Rmd01.png)

but when I cut and pasted the file into RStudio, it changed.

![![Figure 2. Screenshot of a file inside RStudio](non-breaking-space02.png not found.](../../../web/images/20/non-breaking-space.Rmd02.png)

There were now a couple of pink dots at the bottom of the file. Ignore the three consecutive colons for now, as they are not important.

So what are these pink dots, exactly? I used cut and paste to insert this strange character into a character string in R

```{r insert}
tst1 <- " "
tst1
```

That's strange. When you move the pink dot, it disappears and it doesn't look like you can get it again by printing it. But it is not a blank. The charToRaw function will convert any character string to the equivalent hexadecimal code.

```{r chartoraw}
charToRaw(tst1)
```

The hexadecimal code, `a0`, is not the normal blank character, which is hexadecimal `20`. It appears in some html files, especially to keep multiple blank lines from collapsing to a single blank line. The code in html is `&nbsp;` and this file, which was converted from a file originally in html format converted the non-breaking space into hexadecimal `a0`.

If you want to search for non-breaking spaces in a text file, use the code `\xa0` in R.

```{r search}
tst1=="\xa0"
```

Wikipedia has a [nice page](https://en.wikipedia.org/wiki/Non-breaking_space) about non-breaking spaces.

### 2020-02-04 src/prelims.R

@@ -61,6 +61,22 @@ build_file_list <- function(dir_root, file_pattern, v=TRUE) {
   return(file_list)
 }
 
+build_yr_list <- function(dir_root, file_pattern, v=TRUE) {
+  file_list <- NULL
+  yr_list <- list.dirs(dir_root, recursive=FALSE)
+  for (i_yr in yr_list) {
+    if (v) {"\nYear =" %b% i_yr %>% cat}
+    new_files <- list.files(i_yr, pattern=file_pattern)
+    if (length(new_files)==0) {
+      if (v) " has no files of the form" %b% file_pattern %>% cat
+      next
+    }
+    if (v) " has" %b% str_c(new_files, collapse=", ") %>% cat
+    file_list <- append(file_list, i_yr %s% new_files)
+  }
+  return(file_list)
+}
+
 # This function takes a topic, partial date,
 # or category, and produces a link to the
 # appropriate page.

### 2020-02-04 src/step0-identify-images.R

@@ -8,7 +8,7 @@
 if (!exists("v")) v <- TRUE
 
 source(file="src/prelims.R")
-for (i_yr in c("99")) {
+for (i_yr in c("99", zpad(0:20))) {
   for (i_mo in zpad(1:12)) {
     file_list <- list.files(path="text" %s% i_yr %s% i_mo)
     new_path <- "text" %s% i_yr

### 2020-02-04 src/prelims.R
Relocated bib files

@@ -443,7 +443,8 @@ clean_files <- function(search_string, replace_string="Not yet", dir_root="text"
 # file associated with a bibtex recommendation.
 
 write_body <- function(fields, v=TRUE) {
-  image_link <- "../../images" %s% fields$image
+  fields$blogdate %>% str_sub(3,4) -> yr
+  image_link <- "../../images" %s% yr %s% fields$image
   new_tx <-
     "---"                                     %1%
     "title: "    %q% fields$title             %1%

### 2020-02-04 src/step0-identify-images.R

@@ -23,6 +23,51 @@ for (i_yr in c("99", zpad(0:20))) {
   }
 }
 
+bib_list <- build_file_list("text", "*.bib")
+
+bib_list %>% 
+  str_remove("/01") %>%
+  str_remove("/02") %>% 
+  str_remove("/03") %>% 
+  str_remove("/04") %>% 
+  str_remove("/05") %>% 
+  str_remove("/06") %>% 
+  str_remove("/07") %>% 
+  str_remove("/08") %>% 
+  str_remove("/09") %>% 
+  str_remove("/10") %>% 
+  str_remove("/11") %>% 
+  str_remove("/12") -> new_files
+
+file.copy(bib_list, new_files)
+
+
+bib_list %>% 
+  str_subset("text/19") %>%
+  str_remove(".*/") %>%
+  str_remove(fixed(".bib")) -> bib_19
+
+old_files <- "text../web/images"    %s% bib_19 %d% "png"
+new_files <- "../web/images/19" %s% bib_19 %d% "png"
+file.copy(old_files, new_files)
+
+
+old_files <- "../web/images"    %s% bib_19 %d% "png"
+new_files <- "../web/images/19" %s% bib_19 %d% "png"
+file.copy(old_files, new_files)
+
+bib_list %>% 
+  str_subset("text/20") %>%
+  str_remove(".*/")  %>%
+  str_remove(fixed(".bib")) -> bib_20
+
+old_files <- "../web/images"    %s% bib_20 %d% "png"
+new_files <- "../web/images/20" %s% bib_20 %d% "png"
+
+file.copy(old_files, new_files)
+
+
+
 # Save everything.
 
 save.image("data/step0.RData")

### 2020-02-04 src/step1-translate-bib-to-md.R

@@ -11,7 +11,9 @@ source(file="src/prelims.R")
 if (!exists("v")) v <- TRUE
 if (!exists("update_all")) update_all <- TRUE
 
-file_list <- build_file_list(text_root, "*.bib", v=FALSE)
+text_root <- "text"
+
+file_list <- build_yr_list(text_root, "*.bib", v=FALSE)
 
 skipped_files <- NULL
 for (i_file in file_list) {

### 2020-02-04 src/step2-augment-md-files.R
Fixed bad characters in md files

@@ -12,8 +12,8 @@ if (!exists("verbose"   )) verbose    <- TRUE
 
 ## Step 2-2. Find the .md files
 
-file_list <- build_file_list(text_root, "*.md", v=FALSE)
-file_list <- c(file_list, build_file_list(text_root, "*.Rmd", v=FALSE))
+file_list <- build_yr_list(text_root, "*.md", v=FALSE)
+file_list <- c(file_list, build_yr_list(text_root, "*.Rmd", v=FALSE))
 
 v=FALSE
 skipped_files <- NULL
@@ -50,3 +50,4 @@ file_list <- setdiff(file_list, skipped_files)
   "files after skipping"%b% 
   length(skipped_files) %b% 
   "files.\n\n" %>% cat
+

### 2020-02-04 src/step5-translate-archive.R

@@ -8,6 +8,7 @@ source(file="src/prelims.R", echo=FALSE)
 
 file_list <- list.files(path=arch_root, pattern="*.md")
 v <- FALSE
+update_all <- FALSE
 file_list <- c(file_list, build_file_list(blog_root, "*.Rmd", v=FALSE))
 
 skipped_files <- NULL
@@ -28,10 +29,9 @@ file_list <- setdiff(file_list, skipped_files)
 "\n\nThere are" %b% length(file_list) %b% "files remaining to be worked on.\n\n" %>% cat
 
 file_list <- sample(file_list)
-
 for (i_file in file_list) {
   if (verbose) {
-    "\nConverting" %b% j_file %>% br %>% cat
+    "\nConverting" %b% i_file %>% br %>% cat
   }
   render(arch_root %s% i_file, output_dir=html_arch)
 }

### 2020-02-04 src/prelims.R
Added visualization-as-art.md

@@ -444,7 +444,7 @@ clean_files <- function(search_string, replace_string="Not yet", dir_root="text"
 
 write_body <- function(fields, v=TRUE) {
   fields$blogdate %>% str_sub(3,4) -> yr
-  image_link <- "../../images" %s% yr %s% fields$image
+  image_link <- "../../../images" %s% yr %s% fields$image
   new_tx <-
     "---"                                     %1%
     "title: "    %q% fields$title             %1%

### 2020-02-04 src/step0-slide-md-files.R

@@ -1,0 +1,53 @@
+# step0-slide-mo-files.R
+# written by Steve Simon
+# created 2020-02-04
+
+# This program moves png files from one directory
+# to another.
+
+source(file="src/prelims.R")
+if (!exists("verbose")) verbose <- TRUE
+
+root_dir <- "text"
+
+rmd1 <- build_file_list(root_dir, "*.Rmd")
+rmd2 <- build_file_list(root_dir, "*.Rmd")
+
+rmd1 %>% str_remove(".*/") -> base1
+rmd2 %>% str_remove(".*/") -> base2
+
+base1 %>%
+  setdiff(base2) -> unmatched
+
+for (i_file in unmatched) {
+  rmd1 %>%
+    str_which(i_file) %>% cat
+}
+
+md1 <- build_file_list(root_dir, "*.md")
+md2 <- build_file_list(root_dir, "*.md")
+
+md1 %>% str_remove(".*/") -> base1
+md2 %>% str_remove(".*/") -> base2
+
+base1 %>%
+  setdiff(base2) -> unmatched
+
+for (i_file in unmatched) {
+  md1 %>%
+    str_which(i_file) %>% cat
+}
+
+bib1 <- build_file_list(root_dir, "*.bib")
+bib2 <- build_file_list(root_dir, "*.bib")
+
+bib1 %>% str_remove(".*/") -> base1
+bib2 %>% str_remove(".*/") -> base2
+
+base1 %>%
+  setdiff(base2) -> unmatched
+
+for (i_file in unmatched) {
+  bib1 %>%
+    str_which(i_file) %>% cat
+}

### 2020-02-05 src/step0-global-change.R
Removing ">" from md files.

@@ -1,0 +1,33 @@
+# step0-global-change.R
+# written by Steve Simon
+# created 2020-02-05
+
+# This program finds files and
+# makes changes from one string
+# to another.
+
+source(file="src/prelims.R")
+if (!exists("v")) v <- TRUE
+
+file_list <- build_yr_list("text", "*.md", v=FALSE)
+
+if (v) "\nProcessing" %b% length(file_list) %b% "files.\n\n" %>% cat
+
+s0 <- "\n>\n"
+s1 <- "\n\n"
+
+for (i_file in file_list[1:5]) {
+  tx <- readLines(i_file)
+  tx %>% 
+    str_c(collapse="\n") %>%
+    str_replace_all(s0, s1) %>%
+    str_split("\n") %>%
+    unlist -> new_tx
+  n_changed <- which(tx != new_tx)
+  if (length(n_changed) == 0) next
+  for (i in n_changed) {
+    tx[i]     %>% br    %>% cat
+    new_tx[i] %>% br(2) %>% cat
+  }
+}
+}

### 2020-02-05 src/step0-global-change.R
Cleaning up odd symbols, part 1

@@ -13,10 +13,7 @@ file_list <- build_yr_list("text", "*.md", v=FALSE)
 
 if (v) "\nProcessing" %b% length(file_list) %b% "files.\n\n" %>% cat
 
-s0 <- "\n>\n"
-s1 <- "\n\n"
-
-for (i_file in file_list[1:5]) {
+process_changes <- function(i_file, s0, s1, dry_run=TRUE) {
   tx <- readLines(i_file)
   tx %>% 
     str_c(collapse="\n") %>%
@@ -24,10 +21,45 @@ for (i_file in file_list[1:5]) {
     str_split("\n") %>%
     unlist -> new_tx
   n_changed <- which(tx != new_tx)
-  if (length(n_changed) == 0) next
+  if (length(n_changed)==0) return(tx)
+  i_file    %>% br    %>% cat
   for (i in n_changed) {
     tx[i]     %>% br    %>% cat
     new_tx[i] %>% br(2) %>% cat
   }
+  if (!dry_run) writeLines(new_tx, i_file)
+  return(new_tx)  
 }
+
+for (i_file in file_list) {
+  d <- FALSE
+  new_tx <- process_changes(i_file, fixed('\n> '), "\n", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('[\\['), "[", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed(']\\]'), "]", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\['), "[", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\]'), "]", dry_run=d)
+  new_tx <- process_changes(i_file, fixed('\\*'), "*", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\"'), '"', dry_run=d) 
+  new_tx <- process_changes(i_file, fixed("\\'"), "'", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed(' '), ' ', dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('Â'), ' ', dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\$'), "$", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\@'), "@", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\\n'), "\n", dry_run=d) 
[…]

### 2020-02-05 text/20/non-breaking-space.Rmd
Cleaning up odd symbols, part 4
Fixed problem with Rmd files

@@
 
-![![Figure 1. Screenshot of a file inside Notepad](non-breaking-space01.png not found.](../../../web/images/20/non-breaking-space.Rmd01.png)
+![Figure 1. Screenshot of the file inside Notepad](../../../web/images/20/non-breaking-space01.png)
 
@@
 
-![![Figure 2. Screenshot of a file inside RStudio](non-breaking-space02.png not found.](../../../web/images/20/non-breaking-space.Rmd02.png)
+![![Figure 2. Screenshot of the file inside RStudio](../../../web/images/20/non-breaking-space02.png)
 
@@
 
-So what are these pink dots, exactly? I used cut and paste to insert this strange character into a character string in R
+So what are these pink dots, exactly? I used cut-and-paste to insert this strange character into a character string in R.
 

### 2020-02-05 src/prelims.R

@@ -321,10 +321,12 @@ if (v) {
 modify_yaml_fields <- function(fields, v=TRUE) {
   # Build full file names
   fields$full_name %>%
+    str_remove(".*\\.") -> ext
+  fields$full_name %>%
     str_remove("^.*/") %>%
     str_remove(fixed(".md")) %>%
     str_remove(fixed(".Rmd")) -> short_name
-  fields$blog_name <- blog_root %s% short_name %d% "md"
+  fields$blog_name <- blog_root %s% short_name %d% ext
   fields$link_name <- link_root %s% short_name %d% "txt"
   fields$summ_name <- summ_root %s% short_name %d% "txt"
   

### 2020-02-05 src/step4-translate-md-to-html.R

@@ -14,7 +14,8 @@ skipped_files <- NULL
 for (i_file in file_list) {
   i_file %>%
     str_remove("^.*/") %>%
-    str_replace("md$|Rmd$", "html") -> j_file
+    str_replace(fixed(".Rmd"), ".html") %>%
+    str_replace(fixed(".md" ), ".html") -> j_file
   if (v) {
     file.info(blog_root %s% i_file)$mtime %C% i_file %>% br %>% cat
     file.info(html_blog %s% j_file)$mtime %C% i_file %>% br %>% cat

### 2020-02-06 text/20/panel-counts.Rmd
Added panel-counts

---
title: "Blog post: Gender distribution of panel discussions"
author: "Steve Simon"
source: "New"
date: "2020-02-06"
category: Blog post
tags: Human side of statistics, R software
output: html_document
---

I was curious, while I was attending the Joint Statistical Meetings (JSM) in August 2019 about the gender distribution of the panel discussions. There had been a lot of controversy at several scientific meetings about "manels," panel discussions that had all or almost all males. So I copied the JSM program into a text file and wrote a short R program to analyze this.

<!---More--->

Here are the R libraries that I need for this program. The suppressWarnings and suppressMessages functions help to keep the output clean.

```{r read}
suppressWarnings(suppressMessages(library(gender)))
suppressWarnings(suppressMessages(library(magrittr)))
suppressWarnings(suppressMessages(library(tidyverse)))
```

```{r data, echo=FALSE}
raw_text <- '
2019 Joint Statistical Meetings - Statistics: Making an Impact - Denver, Colorado
Statistics: Making an Impact
July 27  August 1, 2019
Colorado Convention Center
JSM 2019 Online Program
Return to main conference page
Online Program Home
 	My Program
    Quick Guide (PDF)	Register Now!
Legend:
CC = Colorado Convention Center   H = Hyatt Regency Denver at Colorado Convention Center 
* = applied session       ! = JSM meeting theme

Activity Details

15 !	Sun, 7/28/2019, 2:00 PM - 3:50 PM	CC-503
Artificial Intelligence for Data Science  Invited Panel
Council of Chapters
Organizer(s): Jason H Moore, University of Pennsylvania
Chair(s): Rebecca Hubbard, University of Pennsylvania
2:05 PM	Artificial Intelligence for Data Science 
Panelists:	Jason H Moore, University of Pennsylvania
Joel Dudley, Icahn School of Medicine at Mount Sinai
Larry Hunter, University of Colorodo Denver
3:40 PM	Floor Discussion

16 !	Sun, 7/28/2019, 2:00 PM - 3:50 PM	CC-205
To the Point: Critical Skills and Knowledge to Be Successful in Academia, Industry and the Government 
 Invited Panel
Committee on Career Development, Section on Statistical Consulting, Section on Statistics and Data 
Science Education
Organizer(s): Adin-Cristian Andrei, Northwestern University
Chair(s): Adin-Cristian Andrei, Northwestern University
2:05 PM	To the Point: Critical Skills and Knowledge to Be Successful in Academia, Industry and 
the Government 
Panelists:	David Morganstein, Westat
Jeri Metzger Mulrow, Bureau of Justice Statistics, Department of Justice, Washington, DC
Joan Chmiel, Northwestern University, Chicago, IL
John Bailer, Miami University
Mary J Kwasny, Northwestern University
3:45 PM	Floor Discussion

57 * !	Sun, 7/28/2019, 4:00 PM - 5:50 PM	CC-205
Women Transitioning into Leadership  Invited Panel
Caucus for Women in Statistics
Organizer(s): Kathy Panageas, Memorial Sloan Kettering Cancer Center
Chair(s): Kathy Panageas, Memorial Sloan Kettering Cancer Center
4:05 PM	Women Transitioning into Leadership 
Panelists:	Janet Wittes, Statistics Collaborative, Inc
Marcey Abate Hoover, Sandia National Laboratories
Melissa Begg, Columbia University 
Nancy Geller, National Institutes of Health, NHLBI
Stacy Lindborg, Biogen
5:45 PM	Floor Discussion

58 * !	Sun, 7/28/2019, 4:00 PM - 5:50 PM	CC-102
Artificial Intelligence (AI) in Healthcare and Medical Research in the Big Data Era  Invited Panel
Section on Statistics in Genomics and Genetics, WNAR, International Chinese Statistical Association
Organizer(s): Haiyan Huang, University of California, Berkeley
Chair(s): Ruixiao Lu, Genomic Health, Inc.
4:05 PM	Artificial Intelligence (AI) in Healthcare and Medical Research in the Big Data Era 
Panelists:	Bin Chen, Michigan State University
Haoda Fu, Eli Lilly and Company
Jean Yee Hwa Yang, University of Sydney, Australia
Jing Huang, Veracyte Inc.
Ying Lu, Stanford University
5:45 PM	Floor Discussion

223 * !	Mon, 7/29/2019, 2:00 PM - 3:50 PM	CC-703
The P-Value Controversy: Where Do We Go from Here?  Invited Panel
Biopharmaceutical Section, Biometrics Section, ENAR
Organizer(s): Pranab K Mitra, Merck
Chair(s): Pranab K Mitra, Merck
2:05 PM	The P-Value Controversy: Where Do We Go from Here? 
Panelists:	David Gal, University of Illinois at Chicago
Jeffrey Blume, Vanderbilt University
Lisa Strug
Michael Lavine, University of Massachusetts, Amherst
Philip B. Stark, UC Berkeley
Xihong Lin, Harvard
3:40 PM	Floor Discussion

224 * !	Mon, 7/29/2019, 2:00 PM - 3:50 PM	CC-Four Seasons 1
Sexual Harassment and Assault  Confronting the Threat to Our Statistical Community  Invited Panel
ASA Task Force on Sexual Harassment and Assault, Committee on Women in Statistics, Caucus for 
Women in Statistics
Organizer(s): Leslie McClure, Drexel University
Chair(s): Robert Santos, The Urban Institute
2:05 PM	Sexual Harassment and Assault: Confronting the Threat to Our Statistical Community 
Panelists:	Donna E LaLonde, ASA Committee on Women in Statistics
Emma Benn, Icahn School of Medicine at Mount Sinai
Leslie McClure, Drexel University
Sally C. Morton, Virginia Tech
3:45 PM	Floor Discussion

280 !	Tue, 7/30/2019, 8:30 AM - 10:20 AM	CC-503
Statistical Outreach and Awareness: How to Make an Impact  Invited Panel
Section for Statistical Programmers and Analysts
Organizer(s): Marianne Miller, Eli Lilly and Company
Chair(s): Adrian Coles, Eli Lilly and Co.
8:35 AM	Statistical Outreach and Awareness: How to Make an Impact 
Panelists:	Darius McDaniel, Emory
Jesse Chittams, University of Pennsylvania
Lillian Prince, Kent State University
Mark Ward, Purdue University
Renee Moore, Emory University
10:15 AM	Floor Discussion

281 * !	Tue, 7/30/2019, 8:30 AM - 10:20 AM	CC-205
When Statistical Methods Impact Policy  Invited Panel
Health Policy Statistics Section, Biometrics Section
Organizer(s): Sherri Rose, Harvard Medical School
Chair(s): Samrachana Adhikari, New York University
8:35 AM	When Statistical Methods Impact Policy 
Panelists:	Dionne Price, Food and Drug Administration
Michael Baiocchi, Stanford University
Miguel Marino, Oregon Health Sciences University
Sherri Rose, Harvard Medical School
10:15 AM	Floor Discussion

282 !	Tue, 7/30/2019, 8:30 AM - 10:20 AM	CC-102
Data Science Education at the School Level  Invited Panel
International Association for Statistical Education, National Council of Teachers of Mathematics, Section 
on Statistics and Data Science Education
Organizer(s): Tim Erickson, Epistemological Engineering
Chair(s): Tim Erickson, Epistemological Engineering
8:35 AM	Data Science Education at the School Level 
Panelists:	Andee Rubin, TERC
Anna Fergusson, University of Auckland
Michelle Wilkerson, University of California at Berkeley
Rob Gould, ASA
William Finzer, Concord Consortium
10:15 AM	Floor Discussion

283	Tue, 7/30/2019, 8:30 AM - 10:20 AM	CC-603
Statistical Sleuths, Data Thugs, or Methodological Terrorists? Recent Stories from the New Field of Error 
Detection in the Published Literature  Invited Panel
Section on Statistics and Data Science Education
Organizer(s): Regina Nuzzo, American Statistical Association; Kristin Sainani, Stanford University
Chair(s): Regina Nuzzo, American Statistical Association
8:35 AM	Statistical Sleuths, Data Thugs, or Methodological Terrorists? Recent Stories from the 
New Field of Error Detection in the Published Literature 
Panelists:	Andrew W. Brown, Indiana University School of Public Health-Bloomington
James Heathers, Northeastern University
Kristin Sainani, Stanford University
Nick Brown, University of Groningen
10:10 AM	Floor Discussion

324 *	Tue, 7/30/2019, 10:30 AM - 12:20 PM	CC-205
The Juggling Collaborative Statistician: Which Balls to Drop?  Invited Panel
Section on Statistical Consulting, Committee on Applied Statisticians, Biometrics Section, Korean 
International Statistical Society
Organizer(s): Julia L Sharp, Colorado State University
Chair(s): Mary J Kwasny, Northwestern University
10:35 AM	The Juggling Collaborative Statistician: Which Balls to Drop? 
Panelists:	Alexandra Hanlon, University of Pennsylvania
Emily Griffith, North Carolina State University
Ji-Hyun Lee, University of Florida
Julia L Sharp, Colorado State University
Jungwha "Julia" Lee, Northwestern University
Mimi Kim, Albert Einstein College of Medicine
12:15 PM	Floor Discussion

325 *	Tue, 7/30/2019, 10:30 AM - 12:20 PM	CC-102
Building Future Leaders: Perspectives on Training in Ethics, Professionalism, and Leadership  Invited 
Panel
Council of Chapters
Organizer(s): John D Keighley, University of Kansas Medical Center
Chair(s): David Morganstein, Westat
10:35 AM	Building Future Leaders: Perspectives on Training in Ethics, Professionalism, and 
Leadership 
Panelists:	Gina-Maria Pomann, Duke University
Jo Wick, University of Kansas Medical Center
Jonathan Gelfond, University of Texas Health San Antonio
Pandurang Kulkarni, Eli Lilly & Company
12:10 PM	Floor Discussion

388 * !	Tue, 7/30/2019, 2:00 PM - 3:50 PM	CC-603
Building Bridges for Data Science Education  Invited Panel
Section on Statistics and Data Science Education, Section on Statistical Computing, Section on Statistical 
Learning and Data Science
Organizer(s): Mine Cetinkaya-Rundel, Duke University
Chair(s): Beth Chance, Cal Poly - San Luis Obispo
2:05 PM	Building Bridges for Data Science Education 
Panelists:	Mine Cetinkaya-Rundel, Duke University
Michael Posner, Villanova University
Jeff Forbes, Duke University
Andrea Danyluk, Williams College
3:45 PM	Floor Discussion

389 !	Tue, 7/30/2019, 2:00 PM - 3:50 PM	CC-703
Official Statistics at the Crossroads: Data Quality and Access in an Era of Heightened Privacy Risk  
Invited Panel
Government Statistics Section, Committee on National Statistics, NAS, Committee on Professional Ethics
Organizer(s): Michael Hawes, U.S. Census Bureau
Chair(s): Michael Hawes, U.S. Census Bureau
2:05 PM	Official Statistics at the Crossroads: Data Quality and Access in an Era of Heightened 
Privacy Risk 
Panelists:	John M. Abowd, U.S. Census Bureau
Katy Rother, Committee on Oversight, U.S. House of Representatives
Michael Davern, NORC
Rochelle (Shelly) Wilkie Martinez, U.S. Office of Management and Budget
3:40 PM	Floor Discussion

491 *	Wed, 7/31/2019, 10:30 AM - 12:20 PM	CC-703
Database Lock to Data Safety Monitoring Board Meeting  More Than a Click of a Button  Invited 
Panel
Section for Statistical Programmers and Analysts, Biopharmaceutical Section, Section on Statistical 
Consulting
Organizer(s): Vipin Arora, Eli Lilly and Company
Chair(s): Vipin Arora, Eli Lilly and Company
10:35 AM	Database Lock to Data Safety Monitoring Board Meeting  More Than a Click of a 
Button? 
Panelists:	David Prince, Axio Research 
Kevin Buhr, University of Wisconsin
Lisa Weissfeld, Stats Collaborative
Natasa Rajicic, Cytel Inc
12:15 PM	Floor Discussion

492 * !	Wed, 7/31/2019, 10:30 AM - 12:20 PM	CC-205
Data Fabrication and Falsification: Protecting the Credibility and Impacts of Surveys  Invited Panel
Government Statistics Section, Survey Research Methods Section, ENAR
Organizer(s): Linda J Young, USDA National Agricultural Statistics Service
Chair(s): Kerrie Leslie
8:35 AM	Data Fabrication and Falsification: Protecting the Credibility and Impacts of Surveys 
Panelists:	James Dahlhamer, US Centers for Disease Control and Prevention
Jill DeMatteis , Westat
Linda J Young, USDA National Agricultural Statistics Service
12:15 PM	Floor Discussion

554 * !	Wed, 7/31/2019, 2:00 PM - 3:50 PM	CC-503
Interdisciplinary Research and Leadership: How to Make an Impact in the Data Science Age  Invited 
Panel
IMS, Section on Statistical Learning and Data Science, Royal Statistical Society
Organizer(s): Bin Yu, UC Berkeley
Chair(s): Bin Yu, UC Berkeley
2:05 PM	Interdisciplinary Research and Leadership: How to Make an Impact in the Data Science 
Age 
Panelists:	Alicia Carriquiry, Iowa State University
Christopher Genovese, Statistics, CMU
Hongyu Zhao, Yale
Jasjeet Sekhon, UC Berkeley
Simon Tavare, Inst of Cancer Dynamics and Statistics, Columbia University
Tamara Tamara Greasby, Data Science at The Trade Desk
3:45 PM	Floor Discussion

555 * !	Wed, 7/31/2019, 2:00 PM - 3:50 PM	CC-704
A Historical Perspective on the Application of Sampling Theory and Methods to Statistical  Invited 
Panel
Statistical Auditing Interest Group, History of Statistics Interest Group, Survey Research Methods Section
Organizer(s): John Hilton, Office of the Auditor General of Canada; Roger C. Pfaffenberger, Ryan, LLC
Chair(s): John Hilton, Office of the Auditor General of Canada
2:05 PM	A Historical Perspective on the Application of Sampling Theory and Methods to 
Statistical Auditing 
Panelists:	Donald Roberts, University of Illinois - Emeritus Professor Retired
Wendy Rotz, Grant Thornton
Richard Valliant, University of Maryland - Emeritus Professor Retired
Ron Bartyczak, Internal Revenue Service - Retired
Alan H. Kvanli, University of North Texas
3:45 PM	Floor Discussion

556 !	Wed, 7/31/2019, 2:00 PM - 3:50 PM	CC-205
Census 2020:  Invited Panel
Committee on Scientific Freedom and Human Rights
Organizer(s): Robin Mejia, Carnegie Mellon University
Chair(s): Robin Mejia, Carnegie Mellon University
2:05 PM	Census 2020 
Panelists:	Connie Citro, The National Academies of Sciences, Engineering, and Medicine
Edward Kissam, CIRS
Hansi Lo Wang, National Public Radio
3:45 PM	Floor Discussion

597 * !	Thu, 8/1/2019, 8:30 AM - 10:20 AM	CC-503
Vision 2020: Making Impact with Statistics in the Era of Data Science  Invited Panel
Committee of Presidents of Statistical Societies, ENAR, Section on Statistical Learning and Data Science
Organizer(s): Huixia Judy Wang, The George Washington University
Chair(s): Bhramar Mukherjee, University of Michigan
8:35 AM	Vision 2020: Making Impact with Statistics in the Era of Data Science 
Panelists:	Hadley Wickham, RStudio
Jeffrey Leek, Johns Hopkins Bloomberg School of Public Health
John Quackenbush, Harvard University
Rachel Schutt, BlackRock
Tian Zheng, Columbia University
Xiao-Li Meng, Harvard University
10:15 AM	Floor Discussion

639 * !	Thu, 8/1/2019, 10:30 AM - 12:20 PM	CC-503
Women in Data Science: a Small N Sample  Invited Panel
Section for Statistical Programmers and Analysts, Section on Statistical Learning and Data Science, 
Caucus for Women in Statistics
Organizer(s): Maria A Terres, Waymo
Chair(s): Maria A Terres, Waymo
10:35 AM	Women in Data Science: a Small N Sample 
Panelists:	Moorea Brega, Pattern Ag
Molly Davies, Stitch Fix
Mary Beth Broadbent, Google/YouTube
Cheryl Flynn, AT&T Research Labs
Clara Yuan, Convoy Inc.
12:05 PM	Floor Discussion

640 * !	Thu, 8/1/2019, 10:30 AM - 12:20 PM	CC-205
Providing Open Standards for Mobile Phone Data Statistics  Invited Panel
International Statistical Institute, Stats. Partnerships Among Academe Indust. & Govt. Committee, 
Statistics Without Borders
Organizer(s): May Offermans, Statistics Netherlands
Chair(s): Tracey Li, Flowminder
10:35 AM	Providing Open Standards for Mobile Phone Data Statistics 
Panelists:	May Offermans, Statistics Netherlands
Albrecht Wirthmann, EUROSTAT
Samantha Watson, Flowminder Foundation
12:15 PM	Floor Discussion

?
American Statistical Association
732 North Washington Street
Alexandria, VA 22314
(703) 684-1221
meetings@amstat.org

Copyright  American Statistical Association

Privacy Policy | Conduct Policy | Previous JSMs
'
raw_text %<>% str_split("\n") %>% unlist
```

The text file has a consistent format that makes it easy to identify things. The first panelist is on a line starting with the text "Panelists:" and the last panelist appears just before the line with the text "Floor Discussion".

```{r extract}
i <- grep("^Panelists:", raw_text)
j <- grep("Floor Discussion", raw_text)
print(i)
print(j)
for (k in 1:length(i)) {
  cat(paste(raw_text[i[k]:j[k]], collapse="\n"))
  cat("\n\n")
}
full_name <- NULL
panel <- NULL
for (k in 1:length(i)) {
  full_name <- c(full_name, raw_text[i[k]:(j[k]-1)])
  panel <- c(panel, rep(k, j[k]-i[k]))
}
full_name %>%
  sub("^Panelists:\t", "", .) %>%
  sub(" .*", "", .) -> first_name
```

The gender package takes a first name and compares it to a database of names from Social Security that calculates the proportion of males and of females in that database who have a specific first name. Names like "Mary" and "John" will belong almost exclusively to one gender, but names like "Pat" will be more evenly split.

```{r gender}
identify_gender <- gender(first_name)
identify_gender %>%
  distinct(name, .keep_all=TRUE) %>%
  full_join(data.frame(panel=panel, name=first_name, stringsAsFactors=FALSE)) -> panel_demographics
```

The gender package will only report results on names that are found in their database, and this would exclude some (but not all) "foreign" names. Any of these cases, plus any cases where the gender package rates the probability of male/female somewhere in the ambiguous middle (between 20% and 80%) will be coded as unknown/ambiguous.

Here are the names with unknown gender. You could probably resolve some of these easily.

```{r count-unknown}
unknown_gender <- is.na(panel_demographics$gender)
panel_demographics %>%
  filter(unknown_gender) %>%
  pull(name) %>%
  unique %>%
  sort %>%
  paste(collapse=", ") %>%
  print
```

Here are the ambiguous names.

```{r count-ambiguous}
ambiguous_gender <- panel_demographics$proportion_male > 0.20 & panel_demographics$proportion_male < 0.80
panel_demographics %>%
  filter(ambiguous_gender) %>%
  pull(name) %>%
  unique %>%
  sort %>%
  paste(collapse=", ") %>%
  print
```

Code both the unknown and ambiguous names into the same category.

```{r code-unknown-and-ambiguous}
panel_demographics$gender[unknown_gender] <- "unknown/ambiguous"
panel_demographics$gender[ambiguous_gender] <- "unknown/ambiguous"
table(panel_demographics$gender)
```

It is very easy to get a count of female and male names by panel. The proportion of females excludes any names that were not cdoed as unknown/ambiguous.

```{r summarize}
panel_demographics %<>% arrange(panel, gender)
n_panels <- max(panel_demographics$panel)
name_list <- rep("", n_panels)
for (i in 1:n_panels) {
  name_list[i] <- paste(panel_demographics$name[panel_demographics$panel==i], collapse=", ")
}
panel_demographics %>%
  mutate(f=as.numeric(gender=="female")) %>%
  mutate(m=as.numeric(gender=="male")) %>%
  mutate(u=as.numeric(gender=="unknown/ambiguous")) %>%
  group_by(panel) %>%
  summarize(
    f=sum(f),
    m=sum(m),
    u=sum(u)) %>%
  mutate(p_f=round(f/(f+m), 2)) %>%
  bind_cols(tibble(name_list)) %>%
  arrange(p_f) %>%
  data.frame -> panel_summary
panel_summary %>%
  summarize_at(c("f", "m", "u"), sum) %>%
  mutate(p_f=round(f/(f+m), 2)) -> panel_summary_total_row
panel_summary         
panel_summary_total_row
```

### 2020-02-27 src/step0-split-zotero-files.R
First attempt to add zotero files

@@ -1,0 +1,61 @@
+# step0-split-zotero-files.R
+# written by Steve Simon
+# created 2020-02-26
+
+# This program takes files downloaded from
+# Zotero and splits them into individual
+# entries.
+
+source(file="src/prelims.R")
+if (!exists("verbose")) verbose <- TRUE
+
+root_dir <- "text/zotero"
+
+bib_files <- list.files(root_dir, pattern="\\.bib")
+
+nam_all <- NULL
+dat_all <- NULL
+src_all <- NULL
+lin_all <- NULL
+for (i_file in bib_files) {
+  tx <- readLines(root_dir %s% i_file)
+  blank_lines <- str_which(tx, "^@")
+  n <- length(blank_lines)
+  blank_lines[n+1] <- length(tx)+1
+  for (i in 1:n) {
+    tx1 <- tx[blank_lines[i]:(blank_lines[i+1]-1)]
+    tx1 %>%
+      str_subset("\\stitle = ") %>%
+      str_remove("\\stitle = ") %>%
+      str_split(" ") %>% 
+      unlist %>%
+      tolower %>% 
+      str_remove_all("[^a-z]") %>%
+      str_subset(".....") %>%
+      data_frame %>%
+      slice(1:3) %>%
+      unlist %>%
+      str_c(collapse="-") -> nam
+    tx1 %>%
+      str_subset("\\surldate = ") %>%
+      str_remove("\\surldate = ") %>%
+      str_split(" ") %>% 
+      unlist %>%
+      str_remove_all("[^0-9\\-]") -> dat
+    if (length(nam)==0) nam <- "no-name"
+    if (length(dat)==0) dat <- "no-date"
+    nam_all <- c(nam_all, nam[1])
+    duplicate_count <- sum(nam_all==nam[1])
+    new_name <- nam[1]
+    if (duplicate_count > 1) new_name <- paste(new_name, duplicate_count, sep="-")
+    yr <- str_sub(dat[1], 3, 4)
+    if (dat[1]=="no-date") yr <- "no"
+    fn <- "text/zotero" %s% yr %s% new_name %d% "bib"
+    print(fn)
+    writeLines(tx1, fn)
+    dat_all <- c(dat_all, dat[1])
+    src_all <- c(src_all, i_file)
+    lin_all <- c(lin_all, blank_lines[i])
+  }
+}
+

### 2020-02-27 src/prelims.R
Removed .nbib files

@@ -248,7 +248,7 @@ c(
   "^RN", "Research Notes",
   "^RP", "Reprint Edition",
   "^SE", "Section",
-  "^SN", "isbn/issn",     
+  "^SN", "isbn-issn",     
   "^SP", "start page",
   "^ST", "short title",
   "^T1", "Primary Title",
@@ -269,8 +269,9 @@ c(
   "^Y2", "Access Date"
 ) %>% 
   matrix(ncol=2, byrow=TRUE) -> ris_matrix
+ris_matrix[ , 2] %<>% tolower %>% str_replace_all(" ", "-")
 
-parse_ris <- function(tx, f0) {
+parse_ris <- function(tx) {
   tx %>%
     str_subset("^ID") %>%
     str_remove("^.*?- ") %>%
@@ -281,14 +282,13 @@ parse_ris <- function(tx, f0) {
       str_subset(ris_matrix[i, 1]) %>%
       str_remove("^.*?- ") %>%
       str_c(collapse=" and ") %>%
-      str_c(ris_matrix[i, 2], "{", ., "}") %>%
+      str_c(ris_matrix[i, 2], " = {", ., "},") %>%
       append(new_tx) -> new_tx
     tx %<>% str_subset(ris_matrix[i, 1], negate=TRUE)
   }
-  new_tx <- c("@article{" %0% id %0% ",", new_tx, "}")
-  cat(tx, sep="\n")
-  cat(new_tx, sep="\n")
-  return()
+  new_tx %<>% str_subset("\\{\\}", negate=TRUE)
+  new_tx %<>% sort
+  c("@article{" %0% id %0% ",", new_tx, "}") %>% return
 }
 
 modify_bib_fields <- function(fields) {
@@ -550,6 +550,8 @@ clean_files <- function(search_string, replace_string="Not yet", dir_root="text"
 write_body <- function(fields, v=TRUE) {
   fields$blogdate %>% str_sub(3,4) -> yr
   image_link <- "../../../images" %s% yr %s% fields$image
+  if (!exists(image_link) & v) "\nFile" %b% fields$source %b% "does not have a corresponding image." %>% cat
+  image_link <- ifelse(exists(image_link), "![]" %p% image_link, "")
   new_tx <-
     "---"                                     %1%
     "title: "    %q% fields$title             %1%

### 2020-02-27 src/step1-translate-bib-to-md.R

@@ -18,11 +18,16 @@ file_list <- build_yr_list(text_root, "\\.ris", v=FALSE)
 for (i_file in file_list) {
   tx <- readLines(i_file)
   cat(tx, sep="\n")
-  parse_ris(tx, i_file)
+  new_tx <- parse_ris(tx)
+  new_file <- str_replace(i_file, "\\.ris$", ".bib")
+  writeLines(new_tx, new_file)
 }
 
 
-file_list <- build_yr_list(text_root, "\\.bib", v=FALSE)
+file_list1 <- build_yr_list(text_root, "\\.bib", v=FALSE)
+file_list2 <- build_yr_list(text_root %s% "zotero", "\\.bib", v=FALSE)
+
+file_list <- c(file_list2, file_list1)
 
 skipped_files <- NULL
 for (i_file in file_list) {
@@ -50,17 +55,3 @@ for (i_file in file_list) {
     write_links            %>%
     write_summ             -> fields
 }
-
-for (i_file in file_list) {
-  if (v) "\nWorking on" %b% i_file %>% cat
-  
-  read_lines(i_file)       %>%
-    parse_ris(i_file)      %>%
-    flag_unused_bib_fields %>%
-    modify_bib_fields      %>% 
-    write_body             %>%
-    write_tail             %>%
-    write_links            %>%
-    write_summ             -> fields
-}
-

### 2020-02-28 src/prelims.R
Revising image links

@@ -570,7 +570,7 @@ write_body <- function(fields, v=TRUE) {
     
     fields$citation                           %2%
     
-    "![]" %p% image_link                      %1%
+    image_link                                %1%
     "\n"
   if (v) {"\n\n" %0% new_tx %>% cat}
   write_lines(new_tx, fields$blog_name)

### 2020-02-28 src/step0-clean-house.R

@@ -1,0 +1,43 @@
+# step0-split-zotero-files.R
+# written by Steve Simon
+# created 2020-02-26
+
+# This program takes files downloaded from
+# Zotero and splits them into individual
+# entries.
+
+source(file="src/standard_functions.R")
+
+old_path <- "../web"
+new_path <- "../web-backup-" %0% Sys.Date()
+dir_list <- c(
+  "links",
+  "md",
+  "md/archive",
+  "md/posts",
+  "site",
+  "site/archive",
+  "site/posts",
+  "summ"
+)
+max_count <- 5
+
+if (!dir.exists(new_path)) dir.create(new_path)
+grand_total <- 0
+for (i_dir in dir_list) {
+  if (!dir.exists(new_path %s% i_dir)) dir.create(new_path %s% i_dir)
+  file_list <- list.files(old_path %s% i_dir, pattern="html$|md$|txt$")
+  "\n\nWorking on directory" %s% old_path %s% i_dir %>% cat
+  count <- 0
+  for (i_file in file_list) {
+    count <- count+1
+    if (count <= max_count) "\n  Backing up" %b% old_path %s% i_dir %s% i_file %>% cat
+    if (count == max_count) cat("\n    Each dot is an additional 100 files ")
+    if (count >  max_count & count %% 100 == 0) cat(".")
+    file.copy(old_path %s% i_dir %s% i_file, new_path %s% i_dir %s% i_file, overwrite=TRUE)
+    file.remove(old_path %s% i_dir %s% i_file)
+  } 
+  "\n  Completed backup of" %b% format(length(file_list), big.mark=",") %b% "files." %>% cat
+  grand_total <- grand_total + length(file_list)
+}
+"\nAll files backed up. Total count is" %b% format(grand_total, big.mark=",") %b% "files." %>% cat

### 2020-02-28 src/update-all-files.R

@@ -2,6 +2,7 @@
 ## written by Steve Simon
 ## created 2020-01-02
 
+                                   source(file="src/step0-clean-house.R")  
 rm(list=ls()); update_all <- TRUE; source(file="src/step1-translate-bib-to-md.R")
 rm(list=ls()); update_all <- TRUE; source(file="src/step2-augment-md-files.R")
 rm(list=ls()); update_all <- TRUE; source(file="src/step3-build-archive-files.R")

### 2020-02-28 src/prelims.R
Re-organizing text folders, part 1

@@ -526,7 +526,6 @@ skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v
 clean_files <- function(search_string, replace_string="Not yet", dir_root="text", file_pattern="*.md", v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE  
   file_list <- build_file_list(dir_root, file_pattern)
-  file_list <- dir_root %s% list.files(dir_root, file_pattern)
   "\nSearching through" %b% length(file_list) %b% "files.\n\n" %>% cat
   for (i_file in file_list) {
     tx <- read_lines(i_file)

### 2020-02-29 text/20/non-breaking-space.Rmd
Added a bunch of spreadsheet recommendations.

--- 2020-02-04 text/20/non-breaking-space.Rmd
@@
 
-![![Figure 1. Screenshot of a file inside Notepad](non-breaking-space01.png not found.](../../../web/images/20/non-breaking-space.Rmd01.png)
+![Figure 1. Screenshot of a file inside Notepad](non-breaking-space01.png)
 
@@
 
-![![Figure 2. Screenshot of a file inside RStudio](non-breaking-space02.png not found.](../../../web/images/20/non-breaking-space.Rmd02.png)
+![Figure 2. Screenshot of a file inside RStudio](non-breaking-space02.png)
 

### 2020-02-29 src/standard_functions.R

@@ -8,6 +8,10 @@
 if (!exists("verbose")) verbose <- TRUE
 verbose <- TRUE
 
+suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(stringr)))
+
 # general functions
 
 # These functions join strings in various ways.

### 2020-02-29 src/update-all-files.R

@@ -2,12 +2,12 @@
 ## written by Steve Simon
 ## created 2020-01-02
 
-                                   source(file="src/step0-clean-house.R")  
-rm(list=ls()); update_all <- TRUE; source(file="src/step1-translate-bib-to-md.R")
-rm(list=ls()); update_all <- TRUE; source(file="src/step2-augment-md-files.R")
-rm(list=ls()); update_all <- TRUE; source(file="src/step3-build-archive-files.R")
-rm(list=ls()); update_all <- TRUE; source(file="src/step4-translate-md-to-html.R")
-rm(list=ls()); update_all <- TRUE; source(file="src/step5-translate-archive.R")
+rm(list=ls()); verbose    <- FALSE; source(file="src/step0-clean-house.R")  
+rm(list=ls()); update_all <- TRUE;  source(file="src/step1-translate-bib-to-md.R")
+rm(list=ls()); update_all <- TRUE;  source(file="src/step2-augment-md-files.R")
+rm(list=ls()); update_all <- TRUE;  source(file="src/step3-build-archive-files.R")
+rm(list=ls()); update_all <- TRUE;  source(file="src/step4-translate-md-to-html.R")
+rm(list=ls()); update_all <- TRUE;  source(file="src/step5-translate-archive.R")
 
 rm(list=ls()); update_all <- FALSE; source(file="src/step1-translate-bib-to-md.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step2-augment-md-files.R")

### 2020-02-29 src/prelims.R
Removed stray backslashes

@@ -523,20 +523,25 @@ skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v
 
 # tx <- skim_md_files("!\\[\\]\\(.*?\\)")
 
-clean_files <- function(search_string, replace_string="Not yet", dir_root="text", file_pattern="*.md", v=TRUE) {
-  if (!exists("ok_to_replace")) ok_to_replace <- FALSE  
-  file_list <- build_file_list(dir_root, file_pattern)
-  "\nSearching through" %b% length(file_list) %b% "files.\n\n" %>% cat
-  for (i_file in file_list) {
-    tx <- read_lines(i_file)
-    found_lines <- str_which(tx, search_string)
-    if (length(found_lines)==0) next
-    "\n\n" %0% i_file %>% br %>% cat
-    str_c(tx[found_lines], collapse="\n") %>% cat
-    if (replace_string!="Not yet") {
-      tx %<>% str_replace(search_string, replace_string)
-      "\n" %0% str_c(tx[found_lines], collapse="\n") %>% cat
-      if (ok_to_replace) writeLines(tx, i_file)
+clean_files <- function(
+    search_string, replace_string="Not yet", 
+    dir_root="text", subdir_list="20",
+    file_pattern="*.md", v=TRUE) {
+  if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
+  for (subdir in subdir_list) {
+    file_list <- list.files(dir_root %s% subdir, file_pattern)
+    "\nSearching through" %b% length(file_list) %b% "files.\n\n" %>% cat
+    for (i_file in file_list) {
+      tx <- read_lines(dir_root %s% subdir %s% i_file)
+      found_lines <- str_which(tx, search_string)
+      if (length(found_lines)==0) next
+      "\n\n" %0% i_file %>% br %>% cat
+      str_c(tx[found_lines], collapse="\n") %>% cat
+      if (replace_string!="Not yet") {
+        tx %<>% str_replace(search_string, replace_string)
+        "\n" %0% str_c(tx[found_lines], collapse="\n") %>% cat
+      if (ok_to_replace) writeLines(tx, dir_root %s% subdir %s% i_file)
+      }
     }
   }
 }
@@ -545,6 +550,7 @@ clean_files <- function(search_string, replace_string="Not yet", dir_root="text"
 
 # This function writes the body of a markdown
 # file associated with a bibtex recommendation.
+
 
[…]

### 2020-02-29 text/20/non-breaking-space.Rmd
More minor fixes to blog files

@@
 
-![Figure 1. Screenshot of a file inside Notepad](non-breaking-space01.png)
+![Figure 1. Screenshot of a file inside Notepad](http://www.pmean.com/images/20/non-breaking-space01.png)
 
@@
 
-![Figure 2. Screenshot of a file inside RStudio](non-breaking-space02.png)
+![Figure 2. Screenshot of a file inside RStudio](http://www.pmean.com/images/20/non-breaking-space02.png)
 

### 2020-02-29 src/step4-translate-md-to-html.R
Reorganized 2019 files

@@ -29,12 +29,15 @@ file_list <- setdiff(file_list, skipped_files)
 "\n\nThere are" %b% length(file_list) %b% "files remaining to be worked on.\n\n" %>% cat
 
 file_list <- sample(file_list)
+n <- length(file_list)
+k <- 0
 for (i_file in file_list) {
+  k <- k+1
   readLines(blog_root %s% i_file) %>%
     str_subset(regex("^Date:", ignore_case=TRUE)) %>%
     str_c(collapse="\n") -> file_date
   if (verbose) {
-    "\nConverting" %b% i_file %C% file_date %>% br %>% cat
+    "\nConverting" %b% k %b% "of" %b% n %C% i_file %C% file_date %>% br(2) %>% cat
   }
   render(blog_root %s% i_file, output_dir=html_blog)
 }

### 2020-02-29 src/step1-translate-bib-to-md.R
Added tags to bib files

@@ -43,8 +43,11 @@ for (i_file in file_list) {
 file_list <- setdiff(file_list, skipped_files)
 "\n\nThere are" %b% length(file_list) %b% "files remaining to be worked on.\n\n" %>% cat
 
+n <- length(file_list)
+k <- 0
 for (i_file in file_list) {
-  if (v) "\nWorking on" %b% i_file %>% cat
+  k <- k+1
+  if (v) "\nWorking on" %b% k %b% "of" %b% "files" %C% i_file %>% cat
   
   read_lines(i_file)       %>%
     parse_bibtex(i_file)   %>%

### 2020-02-29 src/step5-translate-archive.R

@@ -29,9 +29,12 @@ file_list <- setdiff(file_list, skipped_files)
 "\n\nThere are" %b% length(file_list) %b% "files remaining to be worked on.\n\n" %>% cat
 
 file_list <- sample(file_list)
+n <- length(file_list)
+k <- 0
 for (i_file in file_list) {
+  k <- k+1
   if (verbose) {
-    "\nConverting" %b% i_file %>% br %>% cat
+    "\nConverting" %b% k %b% "of" %b% n %b% "files" %C% i_file %>% br %>% cat
   }
   render(arch_root %s% i_file, output_dir=html_arch)
 }

### 2020-03-01 src/prelims.R
Added a few urldates to zotero

@@ -528,13 +528,20 @@ clean_files <- function(
     dir_root="text", subdir_list="20",
     file_pattern="*.md", v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
+  k <- 0
   for (subdir in subdir_list) {
     file_list <- list.files(dir_root %s% subdir, file_pattern)
-    "\nSearching through" %b% length(file_list) %b% "files.\n\n" %>% cat
+    "\nSearching through" %b% 
+      length(file_list) %b% 
+      "files in" %b%
+      subdir %>%
+      br %>%
+      cat
     for (i_file in file_list) {
       tx <- read_lines(dir_root %s% subdir %s% i_file)
       found_lines <- str_which(tx, search_string)
       if (length(found_lines)==0) next
+      k <- k+1
       "\n\n" %0% i_file %>% br %>% cat
       str_c(tx[found_lines], collapse="\n") %>% cat
       if (replace_string!="Not yet") {
@@ -544,6 +551,9 @@ clean_files <- function(
       }
     }
   }
+  "\n\nTotal files" %b% 
+    ifelse(ok_to_replace, "replaced", "to be replaced") %b%
+    k %>% cat
 }
 
 # Functions for writing files

### 2020-03-01 src/prelims.R
Change odd characters in zotero files

@@ -140,6 +140,7 @@ remove_punctuation <- function(x) {
 parse_bibtex <- function(tx, f0) {
   # Note: Some of the bib files have unprintable
   # junk characters at the very start.
+  tx %<>% str_subset("abstract =", negate=TRUE)
   tx <- tx[nchar(tx)>0]
   tx[1] %<>% str_remove("^.*?@") 
   tx %>%
@@ -525,7 +526,7 @@ skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v
 
 clean_files <- function(
     search_string, replace_string="Not yet", 
-    dir_root="text", subdir_list="20",
+    dir_root="text/zotero", subdir_list=c("99", zpad(0:20), "no"),
     file_pattern="*.md", v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
   k <- 0

### 2020-03-01 src/update-all-files.R

@@ -14,3 +14,5 @@ rm(list=ls()); update_all <- FALSE; source(file="src/step2-augment-md-files.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step4-translate-md-to-html.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step3-build-archive-files.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step5-translate-archive.R")
+
+éç’“”“”

### 2020-03-02 text/20/smart-quotes.Rmd
Started blog post on smart quotes

---
title: "Blog post: Smart quotes, em dashes, and en dashes"
author: "Steve Simon"
source: "New"
date: "2020-03-02"
category: Blog post
tags: Data management
output: html_document
---

If you work with text data a lot, you will encounter some characters that are sort of close to what you need, but sort of not. These include the smart quotes, em dashes, and en dashes.

<!---More--->

### Smart quotes

Programmers use a set of standard quotation marks in their work. These are called straight quotes by some and dumb quotes by others.

The straight double quote (") is one of the standard codes that works pretty much the same on any computer system.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=4.5, fig.height=0.5)
suppressMessages(suppressWarnings(library(stringr)))
cplot <- function(x) {
  par(mar=rep(0,4))
  plot(-1:1, -1:1, type="n", axes=FALSE)
  text(-1, 0, x, cex=2, adj=0)
}
```

```{r straight-double-quotes}
cplot('"')
```

The straight single quote (') is another standard code,

```{r straight-single-quotes}
cplot("'")
```

On early computer systems, this was all you had. You might have a backwards slanting single quote (`), often called a backtick.

```{r backtick}
cplot("`")
```

This was a step backwards from Guttenberg's printing press. Actually, the regression occurred when the typewriter was invented. The limited number of keys that you could fit into a typewriter

![](https://upload.wikimedia.org/wikipedia/commons/thumb/0/05/Underwoodfive.jpg/440px-Underwoodfive.jpg)

 prevented the use of a greater variety of quote marks.

 Now you probably already know this but if you want to assign a double quote to a variable, you surround it with single quotes,

```{r assign-double-quotes-1}
cplot("x  < -  ' \" '")
```

or precede the double quotes with a backslash

```{r assign-double-quotes-2}
cplot('x  < -  \" \\ "\ "')
```

Use the charToRaw function to see the underlying code for the double quote mark

```{r charToRaw-1}
cplot("charToRaw(' \" ')")
```

```{r charToRaw-2}
cplot(charToRaw('"'))
```

and for the single quote mark

```{r charToRaw-3}
cplot("charToRaw(\" ' \")")
```

```{r charToRaw-4}
cplot(charToRaw("'"))
```

and for the backtick

```{r charToRaw-5}
cplot("charToRaw(\" ` \")")
```

```{r charToRaw-6}
cplot(charToRaw("`"))
```

These values are hexadecimal, so 27 in hexadecimal is 2*16+7=39 in decimal.

If you know the hexadecimal code, you can convert it to the character equivalent using the \x prefix.

```{r convert-hex-1}
cplot('"\\x22"')
```

```{r convert-hex-2}
cplot("\x22")
```

When personal computers started to expand beyond the limited character set, that allowed you to use the left double quote,

```{r left-double-quote}
cplot("\U201C")
```

the right double quote,

```{r right-double-quote}
cplot("\U201D")
```

the left single quote,

```{r left-single-quote}
cplot("\U2018")
```

and the right single quote

```{r right-single-quote}
cplot("\U2019")
```

These quote marks are part of a larger character set known as Unicode. The rawToChar function provides a surprising result

```{r rawToChar-8}
cplot('charToRaw( \" \U201C \" )')
```

```{r rawToChar-9}
x <- str_c(as.character(charToRaw("\U201C")), collapse=" ")
cplot(x)
```

Surprise! When you open the world up to different typographic characters, you have to include characters with accents,

```{r accented}
cplot("\U00E9")
```

cedillas,

```{r cedilla}
cplot("\U00C7")
```

and tildes.

```{r tilde}
cplot("\U00F1")
```

You have to have room for the sharp S in German,

```{r sharp-S}
cplot("\U00DF")
```

the thorn in Icelandic,

```{r thorn}
cplot("\U00FE")
```

and a whole host of new characters in Greek,

```{r greek}
cplot("\U03B1 \U03B2 \U03B3")
```

Arabic,

```{r arabic}
cplot("\U062A\U0633\U062C\U0651\U0644")
```

and Chinese.

```{r chinese}
cplot("\U4E2D\U6587")
```

When you add various emojis

```{r emojis}
cplot("\U1F642\U1F622")
```

the list becomes quite long. The system that encodes all of these values is Unicode.

You specify Unicode values with a \U.

```{r unicode-1}
cplot("\"\\U201C\"")
```

```{r unicode-2}
cplot("\U201C")
```

There's a [nice web page](https://www.cl.cam.ac.uk/~mgk25/ucs/quotes.html) about the historical developments of computer codes for quote marks and dashes and [another page](https://cran.r-project.org/web/packages/utf8/vignettes/utf8.html) that talks about computer codes in general from the perspective of an R programmer.

There are some other variants, such as the prime symbols, described in [this Wikipedia page](https://en.wikipedia.org/wiki/Prime_(symbol)).

### 2020-03-02 text/20/smart-quotes.Rmd
Finished blog entry on smart quotes.

@@
 
+Now you might wonder why the internal code for the left double quote (`r x`) does not match the 201C shown above.
+It turns out that the internal storage of Unicode uses a system called UTF-8.
+UTF-8 maintains storage efficiency and backwards compatibility with earlier coding systems.
+
+The other smart quote marks are the right double quote,
+
+```{r unicode-3}
+cplot("\"\\U201D\"")
+```
+
+```{r unicode-4}
+cplot("\U201D")
+```
+
+the left single quote,
+
+```{r unicode-5}
+cplot("\"\\U2018\"")
+```
+
+```{r unicode-6}
+cplot("\U2018")
+```
+
+and the right single quote,
+
+```{r unicode-7}
+cplot("\"\\U2019\"")
+```
+
+```{r unicode-8}
+cplot("\U2019")
+```
+
+There are a couple of additional codes that I should mention. Most programmers use the minus sign in their coding
+
+```{r minus-1}
+cplot("-")
+```
+
+but there are two similar characters that you might see. The em dash,
+
+```{r em-dash-1}
+cplot("\"\\U2014\"")
+```
+
+```{r em-dash-2}
+cplot("\U2014")
+```
+
+is a longer dash. It has a width equal to the letter "m" in most proportional width fonts. There is another dash, the en dash
+
+```{r en-dash-1}
+cplot("\"\\U2013\"")
+```
+
+```{r en-dash-2}
+cplot("\U2013")
+```
+
+that is also longer than a minus sign, but about half the length of the em dash. It has a width that is equal to the letter "n" in most proportional width fonts.
+
+The em dash and en dash will often cause confusion because they look so much like the minus sign, but they will cause problems often in R code.
+
 There's a [nice web page](https://www.cl.cam.ac.uk/~mgk25/ucs/quotes.html) about the historical developments of computer codes for quote marks and dashes and [another page](https://cran.r-project.org/web/packages/utf8/vignettes/utf8.html) that talks about computer codes in general from the perspective of an R programmer.

### 2020-03-04 src/update-all-files.R
Three new bib entries.

@@ -14,5 +14,3 @@ rm(list=ls()); update_all <- FALSE; source(file="src/step2-augment-md-files.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step4-translate-md-to-html.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step3-build-archive-files.R")
 rm(list=ls()); update_all <- FALSE; source(file="src/step5-translate-archive.R")
-
-éç’“”“”

### 2020-03-10 src/prelims.R
Fixed a bad Zotero reference

@@ -527,7 +527,7 @@ skim_md_files <- function(search_string, dir_root="text", file_pattern="*.md", v
 clean_files <- function(
     search_string, replace_string="Not yet", 
     dir_root="text/zotero", subdir_list=c("99", zpad(0:20), "no"),
-    file_pattern="*.md", v=TRUE) {
+    file_pattern="*.bib", v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
   k <- 0
   for (subdir in subdir_list) {

### 2020-03-11 text/20/misuse-of-gradient.Rmd
Added misuse-of-gradient.Rmd

---
title: "Blog post: An example of a poor color choice"
author: "Steve Simon"
source: "New"
date: "2020-03-11"
category: Blog post
tags: Graphical display
output: html_document
---

I ran across a graph in a journal article. The article itself was good, but the graph had a rookie mistake. I shouldn't point this out, because I myself have been guilty of far worse mistakes. But this graph illustrated the point far better than anything I could have said.

<!---More--->

The display is a series of stacked bar charts, and it showed that certain types of information were most likely to be entered by a physician. It was least likely to be entered by a case manager.

![](http://www.pmean.com/images/20/misuse-of-gradient.png)

Find the full article found [here](https://medinform.jmir.org/2019/3/e13802/).

Notice that the colors used to distinguish physician from social worker from nurse from case manager represented a gradient of oranges from a dark almost brown orange to a very light orange.

Gradients are very useful for showing changes in a continuous variable, and possibly in an ordinal variable as well. But the list of providers is nominal. There is no natural ordering from physician to social worker to nurse to case manager. For nominal data, you want to choose a set of colors that are readily distinguishable from one another. The usually means evenly spaced points along a color wheel.

Here's an example of what I mean.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, fig.width=5, fig.height=5)
```

```{r color-circle1, echo=FALSE, fig.width=5, fig.height=5}
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(RColorBrewer)))
n <- 500
k <- 5
delta <- 0.9
angle_seq <- (1:n)/n
df0<- data.frame(x=delta*sin(2*pi*angle_seq), y=delta*cos(2*pi*angle_seq))
sb <- (n/k)*(1:k)
df1 <- df0[sb, ]

ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 1, 1), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 1), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

This color wheel includes the pure colors (pure yellow, pure green, pure cyan, pure red, pure magenta, pure blue) and these colors are sometimes a bit harsh on your eyes. A darker set of colors is a bit easier to view.

```{r color-circle2, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 1, 0.5), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 0.5), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

```{r color-circle3, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 0.5, 1), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 0.5, 1), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

There are lots of choices in various graphics packages. Here is one nice set of categorical color choices in R.

```{r brew1, fig.height=1}
par(mar=rep(0, 4))
display.brewer.pal(5, "Set1")
brewer.pal(5, "Set1")
```

These aren't taken directly from a color wheel but they are well spaced out from a visual perspective, meaning that it is easy to distinguish among these different colors.

### 2020-03-11 src/prelims.R
Updated RIS translation.

@@ -258,7 +258,7 @@ c(
   "^TA", "Translated Author",
   "^TI", "title",
   "^TT", "translated title",
-  "^U1", "user definable 1",
+  "^U1", "annote",
   "^U2", "user definable 2",
   "^U3", "user definable 3",
   "^U4", "user definable 4",

### 2020-03-15 text/20/map-projections.Rmd
Worked on map-projections blog entry

---
title: "Blog post: Different map projections"
author: "Steve Simon"
source: "New"
date: "2020-03-15"
category: Blog post
tags: Graphical display
output: html_document
---

I have been drawing some maps in R and was struggling to figure out how to display things with the appropriate map projections. Here is some code where I experimented with a variety of approaches.

<!---More--->

The earth is a three dimensional object and there is no way to display it on a two dimensional computer screen without some distortion. Your goal is to minimize that distortion.

First things first. There are a variety of map options, and they have a variety of storage methods.

One of the [tutorials](https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html) I viewed had mentioned [Natural Earth](http://www.naturalearthdata.com/), an open source repository. You can get easy access to this information with two R packages: [rnaturalearth](https://cran.r-project.org/package=rnaturalearth) and [rnaturalearthdata](https://cran.r-project.org/package=rnaturalearthdata).

What these give you is the ability to draw outlines of any and every country. You also get the outlines of states/provinces/districts in many countries. There's a lot more, but I want to just show something simple now to get you started.

The traditional format for storing map information is shapefiles. The shapefile format was defined the 1990s and is a rather intimidating collection of multiple files. You can find a nice summary on the [Wikipedia page on shapefiles](https://en.wikipedia.org/wiki/Shapefile) or at a Library of Congress's [page on archival formats](https://www.loc.gov/preservation/digital/formats/fdd/fdd000280.shtml).

A newer format, simple features, has greater flexibility and a simpler structure. It's still not simple, and I have been able to delve into the inner workings to get important details extracted.

There is a package in R, [sf](https://cran.r-project.org/package=sf), that allows you to easily manipulate geographic data stored in a simple features format. This package has a nice series of vignettes, starting [here](https://r-spatial.github.io/sf/articles/sf1.html).

So to get everthing to work, even for a simple example, you need a whole bunch of packages. Here are the ones I use in this example.

```{r setup}
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(mapproj)))
suppressMessages(suppressWarnings(library(rgeos)))
suppressMessages(suppressWarnings(library(rjson)))
suppressMessages(suppressWarnings(library(rnaturalearth)))
suppressMessages(suppressWarnings(library(rnaturalearthdata)))
suppressMessages(suppressWarnings(library(sf)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyverse)))
```

I'm going to read in the United States file using the ne_states function. The country="United States of America" parameter excludes other countries that have states, provinces, or districts. I further want to remove Alaska and Hawaii from the file. So sorry to our 49th and 50th states, but the graphs are much more manageable for these examples when I focus just on the 48 states. The str function tells you a bit of information about how the data is organized within the states data frame.

```{r us-states}
ne_states(country="United States of America", returnclass = "sf") %>%
  filter(postal != "HI" & postal != "AK") -> states
str(states)
```

There is a lot of information in the list above and more than half of it is just giving different names to each state. The last column in this data frame is geometry and this is going to be the last column of any data stored in the simple features format. It ends up being a very long string with latitude and longitude coordinates for one or more polygons. You need multiple polygons for a disconnected state like Michigan, or a state with a lot of islands, like Florida.

Similarly, the ne_countries function gets information about the boundaries of every country in the world.

```{r world-countries}
countries <- ne_countries(scale = "medium", returnclass = "sf")
str(countries)
```

Notice that this data frame has a lot of statistical information (population, area, etc.) for each country. You don't need that information for the examples shown here, but it is nice to have.

```{r, echo=FALSE}
continental_us <- ggplot() + geom_sf(data = states)
projection_string <- "+proj=ortho +lon_0=-100 +lat_0=23.5 +ellps=WGS84 +no_defs"
states %>%
  filter(postal=="CO") %>%
  st_transform(projection_string) -> co
countries %>%
  ggplot() + 
  geom_sf(fill="white") + 
  geom_sf(data=co, color="red", fill="blue") +
  coord_sf(crs=projection_string)
```

### 2020-03-16 text/20/map-projections.Rmd
Added information about Colorado

@@
 continental_us <- ggplot() + geom_sf(data = states)
-projection_string <- "+proj=ortho +lon_0=-100 +lat_0=23.5 +ellps=WGS84 +no_defs"
+projection_string <- "+proj=ortho +lon_0=-80 +lat_0=23.5 +ellps=WGS84 +no_defs"
 states %>%
@@
   geom_sf(fill="white") + 
-  geom_sf(data=co, color="red", fill="blue") +
+  geom_sf(data=co, color="red", fill="white") +
   coord_sf(crs=projection_string)

### 2020-03-16 text/20/map-projections.Rmd
Documented installation of rnaturaleharthhires
Added geographic facts about Colorado

@@
 This package has a nice series of vignettes, starting [here](https://r-spatial.github.io/sf/articles/sf1.html).
+
+Please note that the rnaturalearthhires package is needed below, but it cannot be downloaded directly (at least on my system). Instead you have to type in the command
+
+```{}
+install.packages("rnaturalearthhires", repos="http://packages.ropensci.org", type="source")
+```
 
@@
 
+Here is a picture of the earth as it might look like if viewed from the sun.
+The north pole is showing because it is summertime in the northern hemisphere.
+All the countries that are easily visible in this perspective are outlined in black and the state of Colorado in the United States is outlined in red.
+
 ```{r, echo=FALSE}
-continental_us <- ggplot() + geom_sf(data = states)
 projection_string <- "+proj=ortho +lon_0=-80 +lat_0=23.5 +ellps=WGS84 +no_defs"
 states %>%
-  filter(postal=="CO") %>%
-  st_transform(projection_string) -> co
+  filter(postal=="CO") -> co
 countries %>%
@@
 ```
+
+Colorado is a very interesting state with a simple boundary.
+The sourthern edge is 37 degrees latitude.
+The northern edge is 41 degrees north latitude.
+The eastern edge is 102 degrees, 02 minutes, 48 seconds west longitude.
+The western edge is 109 degrees, 02 minutes, 48 seconds west longitude.
+
+You can calculate the length of each of these boundaries with a bit of simple math. The radius of the earth is
+
+$r=6,378$ meters
+
+The circumference is
+
+$2\pi r=40,074$ meters
+
+Each degree of latitude would be
+
+$40,074/360=111.3$ meters.
+
+So the four degrees that you would have to travel to get from the southern border (37 degrees) to the northern border (41 degrees) is
+
+$4*111.3=445.2$ meters
+
+The difference between the northern and southern borders is 4 degrees, so the length of either the eastern or the western borders is
+
+$(4/360)*40,074=445$ meters
+
+The distance from the east to the west border is tricky.
+At the equator a degree of distance is 111.3 meters, but as you move away from the equator, the lines of latitude converge so a degree of longitude shrinks.

### 2020-03-17 text/20/map-projections.Rmd
Added mercator projection to map-projections

@@
 ne_states(country="United States of America", returnclass = "sf") %>%
-  filter(postal != "HI" & postal != "AK") -> states
-str(states)
+  filter(postal == "CO") -> co
+str(co)
+names(co)
 ```
@@
 projection_string <- "+proj=ortho +lon_0=-80 +lat_0=23.5 +ellps=WGS84 +no_defs"
-states %>%
-  filter(postal=="CO") -> co
 countries %>%
@@
 
-You can calculate the length of each of these boundaries with a bit of simple math. The radius of the earth is
+You can calculate the length of each of these boundaries with a bit of simple math. The radius of the earth is 6,378 kilometers. So the circumference is
 
-$r=6,378$ meters
+```{r}
+r <- 6378
+2*pi*r
+```
 
-The circumference is
+Each degree of latitude is
 
-$2\pi r=40,074$ meters
-
-Each degree of latitude would be
-
-$40,074/360=111.3$ meters.
+```{r}
+2*pi*r/360
+```
 
@@
 
-$4*111.3=445.2$ meters
-
-The difference between the northern and southern borders is 4 degrees, so the length of either the eastern or the western borders is
-
-$(4/360)*40,074=445$ meters
+```{r}
+2*pi*r*4/360
+```
 
@@
 At the equator a degree of distance is 111.3 meters, but as you move away from the equator, the lines of latitude converge so a degree of longitude shrinks.
+A simple bit of geometry can show you that a single degree of longitude at 37 degrees is
+
+```{r}
+2*pi*r*cos(37*pi/180)/360
+```
+
+but at 41 degrees, it is
+
+```{r}
+2*pi*r*cos(41*pi/180)/360
+```
+
+That means that the 9 degrees that you travel from the eastern border to the western border is
+
+```{r}
+2*pi*r*cos(37*pi/180)*9/360
+```
+
+at 37 degrees, but only
+
+```{r}
+2*pi*r*cos(41*pi/180)*9/360
+```
+
+So the northern border of Colorado is shorter than the southern border.
+
+And all this time you thought Colorado was a rectangle!
+It's not.
+It is not a trapezoid either.
+It exists on a three dimensional globe and is more complex than any two dimensional geometric object can represent.
+
+You need to make a projection any time you try to represent a three dimensional object like the state of Colorado on a two dimensional medium like a computer screen or a sheet of paper.
+The most common projection is the Mercator projection.
+Here is the Mercator projection for the state of Colorado.
+
+```{r mercator}
+projection_string <- "+proj=tmrec +lon_0=-80 +lat_0=0 +ellps=WGS84 +no_defs"
+projection_string <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
+countries %>%
+  ggplot() + 
+  geom_sf(fill="white") + 
+  geom_sf(data=co, color="red", fill="white") +
+  coord_sf(crs=projection_string, ylim=c(-80,80))
+```

### 2020-03-17 text/20/map-projections.Rmd
More Colorado maps

@@
 ```{r mercator}
-projection_string <- "+proj=tmrec +lon_0=-80 +lat_0=0 +ellps=WGS84 +no_defs"
 projection_string <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
-countries %>%
+projection_string <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +no_defs"
+co %>%
   ggplot() + 
-  geom_sf(fill="white") + 
-  geom_sf(data=co, color="red", fill="white") +
-  coord_sf(crs=projection_string, ylim=c(-80,80))
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
 ```
+
+```{r laea}
+projection_string <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
+projection_string <- "+proj=laea +lon_0=-105.5 +lat_0=39 +ellps=WGS84 +datum=WGS84 +no_defs"
+co %>%
+  ggplot() + 
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
+```

### 2020-03-23 text/20/map-projections.Rmd

--- 2020-03-17 text/20/map-projections.Rmd
@@
 You need to make a projection any time you try to represent a three dimensional object like the state of Colorado on a two dimensional medium like a computer screen or a sheet of paper.
-The most common projection is the Mercator projection.
-Here is the Mercator projection for the state of Colorado.
+Here is the Lambert projection.
 
-```{r mercator}
-projection_string <- "+proj=tmrec +lon_0=-80 +lat_0=0 +ellps=WGS84 +no_defs"
-projection_string <- "+proj=merc +lon_0=0 +k=1 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs"
+```{r laea-co}
+projection_string <- "+proj=laea +lon_0=-105.5 +lat_0=39 +datum=WGS84 +no_defs"
+co %>%
+  ggplot() + 
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
+```
+
+Notice that it has to warp some of the straight edge boundaries in order to keep the distances correct at the northern and southern borders.
+
+Plotting the entire continent of North America shows even more warping.
+Notice how the state of Alaska has been twisted almost 45 degrees to accomodate.
+The island of Greenland is also twisted but in the opposite direction.
+
+The distortions of the Lambert projection are more extreme when you are trying to fit large distances into a single map.
+The distortions become most noticeable at the edges.
+There is almost no distortion near the center.
+
+```{r laea-na}
+projection_string <- "+proj=laea +lon_0=-105.5 +lat_0=39 +datum=WGS84 +no_defs"
 countries %>%
+  filter(continent=="North America") %>%
   ggplot() + 
-  geom_sf(fill="white") + 
-  geom_sf(data=co, color="red", fill="white") +
-  coord_sf(crs=projection_string, ylim=c(-80,80))
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
 ```
+
+The most common projection is the Mercator projection. Here is the Mercator projection for the state of Colorado.
+
+```{r mercator-co}
+projection_string <- "+proj=merc +lon_0=0 +datum=WGS84 +no_defs"
+co %>%
+  ggplot() + 
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
+```
+
+No warping of the boundaries but the lengths of the northern and southern borders are displayed incorrectly as having the same length.
+Not only are distances distorted, but areas as well.
+That becomes more obvious when you look at the entire continent.
+
+```{r mercator-na}
+projection_string <- "+proj=merc +lon_0=-105.5 +datum=WGS84 +no_defs"
+countries %>%
+  filter(continent=="North America") %>%
+  ggplot() + 
+  geom_sf(color="red", fill="white") +
+  coord_sf(crs=projection_string)
+```
+
+Notice how Alaska has grown to more than half the size of the rest of the United States.
+Alaska is big, but not that big!
+Also Greenland has expaned to mostrous proportions.
+The Mercator projection is simple, but it has a lot of distortions at the higher latitudes.
+
+But notice how the eastern border of the state of Alaska is perfectly vertical, representing a true north/south orientation.
+
+No projection system is perfect, and you should try different approaches to see what works best for your situation.

### 2020-04-16 text/20/codes-for-covid-19.Rmd
Added codes-for-covid-19

---
title: "Blog post: Codes for COVID-19 in the electronic health record"
author: "Steve Simon"
source: "New"
date: "2020-04-15"
category: Blog post
tags: Electronic health records
output: html_document
---

I might be helping with a research project that involves selection of medical records associated with COVID-19. Here are some of the resources and suggestions that I received from a colleague at KUMC.

<!---More--->

The ICD-10 code for COVID-19 is U07.1 (COVID-19, virus identified) or U07.2 (COVID-19, virus not identified). This code is in a range of values, U00-U49, that have been set aside for  new diseases of uncertain etiology or emergency use. Previous codes in this range had included SARS (U04) and vaping related disorder (U07.0).

THe two codes are a recent addition, and it looks like earlier cases used an ICD-10 code of B97.29 (compare the February 20, 2020 CDC guidelines to the April 1, 2020 CDC guidelines).

There are several respiratory illnesses associated with COVID-19. Pneumonia is coded as J12.89, acute bronchitis as J20.8, lower respiratory infection as J22, and acute respiratory distress syndrome as J80.

People being tested for COVID-19 could have a code of Z03.818 (Encounter for observation for suspected exposure to other biological agents ruled out), Z20.828 (Contact with and (suspected) exposure to other viral communicable diseases) or Z11.59 (Encounter for screening for other viral diseases). Some of these, of course, might end up being negative tests.

A code that defintely indicates that COVID-19 has been ruled out is Z03.818 (Encounter for observation for suspected exposure to other biological agents ruled out).

Signs and symptoms are often vague, but three that are worth looking at are R05 (Cough), R06.02 (Shortness of breath), and R50.9 (Fever, unspecified).

#### Resources

[American Academy of Family Physicians. Coding for the new coronavirus and COVID-19 illness]()

[American Medical Association. Special coding advice during COVID-19 public health emergency](https://www.ama-assn.org/system/files/2020-04/covid-19-coding-advice.pdf).

[American Thoracic Society. Coding & Billing Quarterly, April 2020](https://www.thoracic.org/about/newsroom/newsletters/coding-and-billing/resources/2020/cbq-april.pdf)

[Centers for Disease Control and Prevention. CD-10-CM Official Coding Guidelines - Supplement. Coding encounters related to COVID-19 Coronavirus Outbreak. Effective: February 20, 2020](https://www.cdc.gov/nchs/data/icd/ICD-10-CM-Official-Coding-Gudance-Interim-Advice-coronavirus-feb-20-2020.pdf).

[Centers for Disease Control and Prevention. ICD-10-CM Official Coding and Reporting Guidelines April 1, 2020 through September 30, 2020](https://www.cdc.gov/nchs/data/icd/COVID-19-guidelines-final.pdf).

### 2020-04-17 text/20/codes-for-covid-19.Rmd
Added emcee-package.md

@@
 ---
-title: "Blog post: Codes for COVID-19 in the electronic health record"
+title: "Codes for COVID-19 in the electronic health record"
 author: "Steve Simon"
@@
 
-I might be helping with a research project that involves selection of medical records associated with COVID-19. Here are some of the resources and suggestions that I received from a colleague at KUMC.
+I might be helping with a research project that involves selection of medical records associated with COVID-19.
+Here are some of the resources and suggestions that I received from a colleague at KUMC plus additional ones I found from a web search.
+I include links to any resources that I drew from.
+
+I would greatly appreciate anyone who could offer additional information, especially about additional CPT codes.
 
 <!---More--->
+
+There are both ICD-10, CPT, and LOINC codes that might help in identifying COVID-19 patients in the electronic health record.
+
+#### ICD-10
 
@@
 
-THe two codes are a recent addition, and it looks like earlier cases used an ICD-10 code of B97.29 (compare the February 20, 2020 CDC guidelines to the April 1, 2020 CDC guidelines).
+THe two codes are a recent addition, and it looks like earlier cases used an ICD-10 code of B97.29 (See the AAFP article or compare the February 20, 2020 CDC guidelines to the April 1, 2020 guidelines).
 
 There are several respiratory illnesses associated with COVID-19.
-Pneumonia is coded as J12.89, acute bronchitis as J20.8, lower respiratory infection as J22, and acute respiratory distress syndrome as J80.
+Pneumonia is coded as J12.89, acute bronchitis as J20.8, non-acute bronchitis as J40, lower respiratory infection as J22, acute respiratory distress syndrome as J80, and other respiratory disorders as J98.8.
 
@@
 
+#### CPT
+
+There is a CPT code for COVID-19 testing, 87635, which was implemented on March 13, 2020.
+Additional codes were added/modified on April 10, 2020.
+These are 86318, 86328, 86602, 86635, and 86769.
+The details of these tests is too complex to summarize here.
+
+There are certain additional CPT codes for procedures that may indicate treatment for COVID-19.
+These would include codes for ventilators and radiology, for example.
+I do not have a good list of what these codes might be.
+
+#### Sensitive versus precise searches
+
+When you are trying to identify COVID-19 cases, you have to decide whether you want a sensitive search, one that is unlikely to miss any cases, but which might include more than a few false positives.
+The alternative is a precise search, which provides relatively strong assurance that anyone you identify is indeed a COVID-19 case, but which also might miss more than a few cases.
+The answer will depend on the purpose of your research.
+It is also worth noting the additional complications caused by the current shortage of tests in the United States combined with the large number of people who seem to have the infection but who are asymptomatic.
+
+Different groups use different criteria for deciding who qualifies to be counted as a COVID-19 patient. This usually involves
+
+1. a positive COVID-19 test, or 
+2. close contact with other(s) with COVID-19 combined with certain clinical signs or symptoms, or 
+3. a greater combination of these clinical signs or symptoms even without a positive test or known close contact.
+
+#### LOINC
+
+There is a long list of LOINC codes associated with COVID-19:
+
++ 94306-8
++ 94307-6
++ 94308-4
++ 94309-2
++ 94310-0
++ 94311-8
++ 94312-6
++ 94313-4
++ 94314-2
++ 94315-9
++ 94316-7
++ 94500-6
++ 94502-2
++ 94503-0
++ 94504-8
++ 94505-5
++ 94506-3
++ 94507-1
++ 94508-9
++ 94509-7
++ 94510-5
++ 94511-3
++ 94531-1
++ 94532-9
++ 94533-7
++ 94534-5
++ 94547-7
++ 94558-4
++ 94559-2
++ 94562-6
++ 94563-4
++ 94564-2
++ 94565-9
++ 94639-2
++ 94640-0
++ 94641-8
++ 94642-6
++ 94643-4
++ 94644-2
++ 94645-9
++ 94646-7
++ 94647-5
++ 94660-8
++ 94661-6
+
+This list is probably incomplete as of the writing of this webpage, and it will almost certainly become more incomplete over time.
+
+Many of these tests, I suspect, are new and not helpful for detecting early cases of COVID-19. Also, these are only for the test orders and not the test results.
+
+#### Miscellaneous
+
+There are multiple terms relating to COVID-19.
+The term COVID-19 itself refers to the disease.
+The virus that causes COVID-19 is called SARS-CoV-2 or SARS Coronavirus 2.
+It was earlier referred to as 2019 Novel coronavirus or 2019-nCoV.
+
 #### Resources
 
-[American Academy of Family Physicians. Coding for the new coronavirus and COVID-19 illness]()
+American Academy of Family Physicians.
+Coding for the new coronavirus and COVID-19 illness, March 5, 2020, updated March 19, 2020.
+Available in [html format](https://www.aafp.org/journals/fpm/blogs/gettingpaid/entry/coronavirus_diagnosis_coding.html).
 
-[American Medical Association. Special coding advice during COVID-19 public health emergency](https://www.ama-assn.org/system/files/2020-04/covid-19-coding-advice.pdf).
+American Academy of Pediatrics. How to use ICD-10-CM, new lab testing codes for COVID-19, March 12, 2020. Avialable in [html format](https://www.aappublications.org/news/2020/03/12/coding031220).
 
-[American Thoracic Society. Coding & Billing Quarterly, April 2020](https://www.thoracic.org/about/newsroom/newsletters/coding-and-billing/resources/2020/cbq-april.pdf)
+American Medical Association. Special coding advice during COVID-19 public health emergency. Available in [pdf format](https://www.ama-assn.org/system/files/2020-04/covid-19-coding-advice.pdf).
 
-[Centers for Disease Control and Prevention.
+American Medical Association. CPT Category I Pathology and Laboratory Codes for Severe
+Acute Respiratory Syndrome Coronavirus 2. Available in [pdf format](https://www.ama-assn.org/system/files/2020-04/coronavirus-long-descriptors.pdf).
+
+American Thoracic Society. Coding & Billing Quarterly, April 2020. Available in [pdf format](https://www.thoracic.org/about/newsroom/newsletters/coding-and-billing/resources/2020/cbq-april.pdf).
+
+Centers for Disease Control and Prevention.
 CD-10-CM Official Coding Guidelines - Supplement.
 Coding encounters related to COVID-19 Coronavirus Outbreak.
-Effective: February 20, 2020](https://www.cdc.gov/nchs/data/icd/ICD-10-CM-Official-Coding-Gudance-Interim-Advice-coronavirus-feb-20-2020.pdf).
+Effective: February 20, 2020.
+Available in [pdf format](https://www.cdc.gov/nchs/data/icd/ICD-10-CM-Official-Coding-Gudance-Interim-Advice-coronavirus-feb-20-2020.pdf).
 
-[Centers for Disease Control and Prevention.
-ICD-10-CM Official Coding and Reporting Guidelines April 1, 2020 through September 30, 2020](https://www.cdc.gov/nchs/data/icd/COVID-19-guidelines-final.pdf).
+Centers for Disease Control and Prevention.
+ICD-10-CM Official Coding and Reporting Guidelines April 1, 2020 through September 30, 2020.
+Available in [pdf format](https://www.cdc.gov/nchs/data/icd/COVID-19-guidelines-final.pdf).
+
+Euorpean Center for Disease Prevention and Control.
+Case definition and European surveillance for COVID-19, as of 2 March 2020.
+Available in [html format](https://www.ecdc.europa.eu/en/case-definition-and-european-surveillance-human-infection-novel-coronavirus-2019-ncov).
+
+LOINC. SARS Coronavirus 2. Available in [html format](https://loinc.org/sars-coronavirus-2/).

### 2020-04-23 text/20/ihme-covid-model.Rmd
start-own-business.bib entry

---
title: "Criticisms of the IHME COVID-19 model"
author: "Steve Simon"
source: "New"
date: "2020-04-23"
category: Blog post
tags: Epidemiology
output: html_document
---

I was asked to review the details of a prominent prediction model for the COVID-19 crisis. This model was developed by the Institute for Health Metrics and Evaluation and is commonly referred to as the IHME model. Here is a bit of background on the IHME model and some resources that discuss this model.

<!---More--->

#### Overview of the IHME model

There is a wonderful sense of openness and transparency about research these days, and the IHME model is a good example of this.

The Institute for Health Metrics and Evaluation has been in business since 2007, so they are no Johnny-Come-Lately operation. The page that lists their staff members has well over 400 pictures. Not all of these people are working on COVID-19, I'm sure, but it appears to be a solid team effort.

The IHME is documented in several places.

#### IMHE documentation of their model

Murray CJL. Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months. MedRxiv preprint, 2020-04-22. [pdf format](https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf).

IHME COVID-19 model FAQs. Available in [html format](http://www.healthdata.org/covid/faqs).

#### Articles in the popular press

Aizenman N. 5 Key Facts Not Explained In White House COVID-19 Projections. National Public Radio 2020-04-01. Available in [html format](https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections).

This article contrasts a model based on infectious disease theory, the Imperial College model, to the IHME model and then goes on to explain some of the features of the IHME model.

#### More technical articles

Druke N, Silver N. How One Modeler Is Trying To Forecast The Toll Of COVID-19. FiveThirtyEight 2020-04-22. Available in [video format](https://fivethirtyeight.com/features/politics-podcast-how-one-modeler-is-trying-to-forecast-the-toll-of-covid-19/) and as a [podcast](https://podcasts.apple.com/us/podcast/model-talk-forecasting-the-toll-of-covid-19/id1077418457).

This is an interview with Dr. Chris Murray, the person in charge of the IHME model.

Begley S. Influential Covid-19 model uses flawed methods and shouldn’t guide U.S. policies, critics say. Stat 2020-04-17. Available in [html format](https://www.statnews.com/2020/04/17/influential-covid-19-model-uses-flawed-methods-shouldnt-guide-policies-critics-say/).

#### Articles in the peer-reviewed literature

Jewell NP, Lewnard JA, Jewell BL. Caution Warranted: Using the Institute for Health Metrics and Evaluation Model for Predicting the Course of the COVID-19 Pandemic. Ann Intern Med. 2020. Available in [html format](https://doi.org/10.7326/M20-1565).

### 2020-04-24 text/20/ihme-covid-model.Rmd
Continued work on ihme-covid-model.

@@
 
-There is a wonderful sense of openness and transparency about research these days, and the IHME model is a good example of this.
+To avoid too much confusion, I will use the term "IHME model" to refer to the COVID-19 prediction model developed by the Institute for Health Metrics and Evaluation and reserve the term "the Institute" to refer to the organization itself.
 
-The Institute for Health Metrics and Evaluation has been in business since 2007, so they are no Johnny-Come-Lately operation.
+The Institute has been in business since 2007, so they are no Johnny-Come-Lately operation.
 The page that lists their staff members has well over 400 pictures.
@@
 
-The IHME is documented in several places.
+There is a wonderful sense of openness and transparency about research these days, and the Institute is a good example of this.
+The Institute shares the code and (when it can) the data on their [github site](https://github.com/ihmeuw).
+This allows other researchers to borrow their work and easily adapt it as needed.
+When the Institute makes changes to their work, anyone can quickly download these updates.
 
@@
 Murray CJL.
+Forecasting the impact of the first wave of the COVID-19 pandemic on hospital demand and deaths for the USA and European Economic Area countries.
+Preprint, 2020-04-21.
+Available in [pdf format](http://www.healthdata.org/sites/default/files/files/Projects/COVID/RA_COVID-forecasting-USA-EEA_042120.pdf).
+
++ At the time of the writing of this webpage, this was the most current description of the IHME model.
+
+Murray CJL.
 Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months.
-MedRxiv preprint, 2020-04-22.
-[pdf format](https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf).
+Preprint, 2020-03-26.
+Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf).
+
++ This article describes an earlier model.
+
+Xu B, Gutierrez B, Mekaru S, Sewalk K, Goodwin L, Loskill A, Cohn EL, Hswen Y, Hill SC, Cobo MM, Zarebski AE, Li S, Wu C-H, Hulland E, Morgan JD, Wang L, O'Brien K, Scarpino SV, Brownstein JS, Pybus OG, Pigott DM, Kraemer MUG.
+Epidemiological data from the COVID-19 outbreak, real-time case information.
+Scientific Data, 2020-03-24.
+doi:10.1038/s41597-020-0448-0.
+Available in [html format](https://www.nature.com/articles/s41597-020-0448-0) or [pdf format](https://www.nature.com/articles/s41597-020-0448-0.pdf).
+
++ This article describes how the initial datasets used by the IHME model were collected.
+In a rapidly changing environment, some of the details may have changed, and at least one of the links in the article points to a gitrhub page that no longer exists.
+But most of the links are good.
+The breadth of the data collection is very impressive and the authors do a good job documenting everything.
 
 IHME COVID-19 model FAQs. Available in [html format](http://www.healthdata.org/covid/faqs).
+
++ This webpage talks in non-technical terms about some of the choices that the researchers chose for the IHME model.
 
@@
 Available in [html format](https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections).
+
++ The focus of this article is on the White House press briefing and the uncertain source of some of the numbers used in that briefing. There is, however, a good amount of detail about the IHME model.
 

### 2020-04-24 text/20/ihme-covid-model.Rmd
More work on ihme-covid-model.

@@
 
-#### Overview of the IHME model
+### Overview of the IHME model
 
@@
 
-#### IMHE documentation of their model
+Here are some features of the IHME model that have stirred up so much debate.
+
+#### The IHME model is atheoretical
+
+The first thing to note about the IHME model is that it is atheoretical.
+That means that the choices that the modelers based were driven solely by what seemed to fit the data best rather than what might be expected based on how infectious diseases are transmitted.
+
+In contrast, most models of infectious disease epidemics use theoretical models.
+These models typical involve compartments that include groups of patients.
+The biggest compartment is the susceptible compartment which represents the vast population of people in an area that might get the disease.
+A second compartment represents infected individuals.
+This is just a small number at the start of the epidemic, but they interact with the susceptible population and pull some of them into the infected compartment.
+The people in the infected department don't stay infected forever.
+They move over time into a third compartment that represents people who have recovered and are no longer infectious.
+You can introduce a fourth compartment, such as exposed but not infected, or a fifth, but a theoretical model will.
+at a minimum, consider how easily someone gets infected and how long that person stay infectious before recovering.
+
+The rates at which people move from susceptible to infected and from infected to recovered are defined as a system of differential equations.
+The structure of these differential equations is critical in controlling how quickly the epidemic spreads, and how quickly it subsides.
+
+These theoretical models are often referred to by the names of the compartments, so an SEIR model is one that has compartments for susceptible, exposed, infected, and recovered people.
+
+The conflict between atheoretical models and theoretical models has a long, long history.
+The atheoretical propents believe in "letting the data speak for itself" while the theoretical proponets believe that models need a sound scientific foundation.
+
+You can see some of this conflict play out in the interview that Dr.
+Murray has on the FiveThirtyEight website (see below).
+Dr.
+Murray argues that the theoretical models have too many assumptions and should let the actual data determine more of the model.
+The critics of the IHME model contend that the model changes as more data arrives, leading to unstable estimates.
+It's important to note here that the model itself is changing and not just the estimates produced by the model.
+
+You will never resolve the controversy over theoretical versus atheoretical models, but in infectious disease modeling, the theoretical models have several important advantages.
+
+First, the SEIR models and variations on it have been used successfully many times in the past.
+Every year a new flu season offers a new test of these models and every year they seem to do reasonably well.
+
+Second, the theoretical models are more transparent.
+You can explain a theoretical model in a paragraph or two and you can easily visualize how one theoretical model differs from another.
+The IHME model requires a lengthy appendix and is difficult reading even for a specialist like myself.
+
+Third, the theoretical models allow you to more readily make extrapolations beyond the range of the data.
+You should always be cautious about extrapolations, but we do this all the time.
+Car manufactures will offer six year warranties on a car model that is less than six months old.
+The IHME model cannot extrapolate a decline in the daily death rate for a region that has not yet reached its peak as easily as an SEIR model can.
+We'll talk more about this below.
+
+#### The IHME model is based on the standard bell shaped curve
+
+To model the cumulative number of deaths, the IHME model chose something called the error function, known fondly by its nickname, erf.
+You may not know erf, but you do know its twin, the normal distribution, given the nikcname of bell.
+Fitting erf to the cumulative number of deaths is mathematically equivalent to fitting the classic bell shaped curve of the normal distribution to the daily number of deaths.
+
+There is no good theoretical reason to expect death rates in an epidemic to follow a bell shaped curve, but the curve does have a fair amount of flexibility and can fit epidemics that start slowly and epidemics that come on quickly.
+They can fit death rates that peak at a modest number of daily deaths and death rates that reach a very large number of daily deaths.
+
+The one thing that the bell curve cannot handle, however, is an epidemic that rises quickly to a peak, but then fades slowly. The symmetry of the bell shaped curve is a serious limitation.
+
+The modelers fixed this in a revision.
+The new model identifies a peak and fits a single bell shaped curve to that peak.
+But then it allows for a mixture of 12 additional bell shaped curves on either side of the peak that can be weighted to produce an asymmetric curve that rises quickly and falls slowly (or vice versa).
+
+This fixes one problem but it causes another.
+The large number of extra bell curves greatly increases the chances for overfitting.
+It looks like the modelers have taken steps to control for overfitting, but the whole concept screams for a simpler solution, such as a cubic spline fit.
+
+#### The IHME model relies heavily on data from China, South Korea, and Italy
+
+If you are in the early phase of an epidemic, everything looks like an exponential curve. Every day you see more deaths than the previous day. This is what that curve looks like.
+
+```{r g1, fig.width=4, fig.height=2}
+suppressMessages(suppressWarnings(library(tidyverse)))
+data.frame(
+  x <- seq(-2.8, -1.5, length=100),
+  y <- dnorm(x)) %>%
+  ggplot(aes(x, y)) +
+    geom_line() + 
+    theme_void() + 
+    expand_limits(x=c(-3, 0.5)) +
+    expand_limits(y=c(-0.02, pnorm(0))) +
+    geom_text(x=-2.5, y=-0.02, label="Time") +
+    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
+```
+
+But then the curve reaches an inflection point. The daily deaths are still increasing, but you can see a deceleration.
+
+Once you are past the inflection point, you can start to forecast reasonably well.
+You can get an initial estimate of when the number of deaths reachs its maximum and when it will die down to zero again.
+But these estimates are tricky and the confidence limits are very wide.
+
+What you need is to predict when the epidemic might end is some data on the other side of the mountain, data where the number of daily deaths starts its decline.
+
+```{r g2, fig.width=4, fig.height=2}
+data.frame(
+  x <- seq(-2.8, -0.25, length=100),
+  y <- dnorm(x)) %>%
+  ggplot(aes(x, y)) +
+    geom_line() + 
+    theme_void() + 
+    expand_limits(x=c(-3, 0.5)) +
+    expand_limits(y=c(-0.02, pnorm(0))) +
+    geom_text(x=-2.5, y=-0.02, label="Time") +
+    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
+```
+
+```{r g3, fig.width=4, fig.height=2}
+data.frame(
+  x <- seq(-2.8, 0.5, length=100),
+  y <- dnorm(x)) %>%
+  ggplot(aes(x, y)) +
+    geom_line() + 
+    theme_void() + 
+    expand_limits(x=c(-3, 0.5)) +
+    expand_limits(y=c(-0.02, pnorm(0))) +
+    geom_text(x=-2.5, y=-0.02, label="Time") +
+    geom_text(x=-3, y=0.07, angle=90, label="Deaths")
+```
+
+### IMHE documentation of their model
 
@@
 + At the time of the writing of this webpage, this was the most current description of the IHME model.
+The appendices provide many of the technical details needed to understand the basis of this model.
 
@@
 
-+ This webpage talks in non-technical terms about some of the choices that the researchers chose for the IHME model.
++ This webpage talks in non-technical terms about some of the choices that the researchers made for the IHME model.
 
-#### Articles in the popular press
+### Articles in the popular press
 
@@
 
-+ The focus of this article is on the White House press briefing and the uncertain source of some of the numbers used in that briefing. There is, however, a good amount of detail about the IHME model.
++ The focus of this article is on the White House press briefing and the uncertain source of some of the numbers used in that briefing.
+It provides, however, a good amount of detail about the IHME model and contrasts it with a model based on infectious disease theory, the Imperial College model.
 
-This article contrasts a model based on infectious disease theory, the Imperial College model, to the IHME model and then goes on to explain some of the features of the IHME model.
-
-#### More technical articles
+### More technical articles
 
@@
 
-This is an interview with Dr. Chris Murray, the person in charge of the IHME model.
++ This is an interview with Dr. Chris Murray, the chair of the Institute and person in charge of the IHME model. He defends many of the controversial choices of the IHME model.
 
@@
 
-#### Articles in the peer-reviewed literature
++ The Stat website provides critical reviews of a broad range of medical research approaches.
+The article critical of the IHME model misses the mark in one way.
+It suggests that the numbers will be misused, but that is an issue for any predictive model.
+But it points out how models based on Epidemiologic theory differ from the IHME model and notes concerns about the frequent tweaking that the IHME model has undergone.
+
+### Articles in the peer-reviewed literature
 
@@
 Available in [html format](https://doi.org/10.7326/M20-1565).
+
++ This is a criticism of the IHME model by prominent researchers published in a prominent journal.
+The authors seem to throw everything they can at the model and not everything sticks.
+They note that the IHME model fails to account for the rebound that might occur if social distancing efforts are relaxed too soon, but the IHME model clearly mentions this in a disclaimer.
+It criticizes the symmetry of the IHME model, which was an issue with earlier versions, but which is not a problem with the current model.
+It criticizes the unreliability of the data, but this unreliability is an issue with any model.
+It mentions that the model fails to capture all sources of uncertainty, but this is also an issue with the competing models.
+Where the criticism hits the mark is in how much this model relies on extrapolations from places that have data well past the peak of maximum cases/deaths (such as Italy, China, and South Korea) to places that have not yet reached their peak or are just barely past their peak.

### 2020-04-25 text/20/ihme-covid-model.Rmd
First complete draft of ihme-covid-model

@@
 
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo=FALSE)
+suppressMessages(suppressWarnings(library(tidyverse)))
+```
+
 I was asked to review the details of a prominent prediction model for the COVID-19 crisis.
@@
 To avoid too much confusion, I will use the term "IHME model" to refer to the COVID-19 prediction model developed by the Institute for Health Metrics and Evaluation and reserve the term "the Institute" to refer to the organization itself.
+There are a class of alternative models based on Epidemiologic principles of infectious diseases.
+They go by a variety of acronyms, but to simplify the language, I will refer to all of these models as SEIR models.
 
@@
 
+#### The IHME model fits to deaths and not cases
+
+The modelers made a concious decision to use the number of daily deaths to estimate the model parameters rather than the number of new daily infections.
+The rationale for this choice is that there are substantial variations in testing rates in different regions of the country and different countries of the world.
+This combined with the large number of people who are asymptomatic leads to huge uncertainties in what the number of new infections is really measuring.
+
+The number of deaths in a region, however, is not perfect itself.
+Many deaths could be uncounted if they occur at home or among infected but undiagnosed patients.
+The flip side of the coin is that many people who were old and frail may have died with or without the infection.
+
+There is strong empirical evidence from comparing death rates before and after the COVID-19 pandemic that we are greatly underestimating the number of deaths that we attribute to COVID-19.
+
+Whether the underestimate of COVID-19 cases is worse than the underestimate of COVID-19 deaths is an open question.
+A key point to remember is that a consistent underestimate (everybody misses 25% of the cases or deaths) is not a serious problem for modeling, but a variable underestimate (some regions miss 10% and other miss 50%) is a big problem.
+
+The alternative SEIR models generally use information from the daily infection rate to estimate parameters and buil predictions, though they can incorporate information from the
+
 #### The IHME model is atheoretical
 
-The first thing to note about the IHME model is that it is atheoretical.
+The most widely criticized feature of the IHME model is that it is atheoretical.
 That means that the choices that the modelers based were driven solely by what seemed to fit the data best rather than what might be expected based on how infectious diseases are transmitted.
@@
 Dr.
-Murray argues that the theoretical models have too many assumptions and should let the actual data determine more of the model.
+Murray argues that the theoretical models have too many assumptions and that modelers should let the actual data determine more of the model.
 The critics of the IHME model contend that the model changes as more data arrives, leading to unstable estimates.
@@
 
-First, the SEIR models and variations on it have been used successfully many times in the past.
+First, the SEIR models are battle-hardened.
+They have been used successfully many times in the past.
 Every year a new flu season offers a new test of these models and every year they seem to do reasonably well.
@@
 ```{r g1, fig.width=4, fig.height=2}
-suppressMessages(suppressWarnings(library(tidyverse)))
 data.frame(
@@
 
-Once you are past the inflection point, you can start to forecast reasonably well.
+Once you are past the inflection point, you can start to forecast further into the future.
 You can get an initial estimate of when the number of deaths reachs its maximum and when it will die down to zero again.
 But these estimates are tricky and the confidence limits are very wide.
-
-What you need is to predict when the epidemic might end is some data on the other side of the mountain, data where the number of daily deaths starts its decline.
 
@@
 
+What you need is to predict when the epidemic might end is some data on the other side of the mountain, data where the number of daily deaths starts its decline.
+
 ```{r g3, fig.width=4, fig.height=2}
@@
 ```
+
+Fair enough, but the regions that are still early in the pandemic still want predictions on when they will reach their peak and when they will eventually decline to zero.
+The IHME model does this through a mixed model approach.
+
+The mixed model is a well-established approach in Statistics.
+It is used a lot in Epidemiology, especially in small area estimation (getting estimated rates in regions that don't have a lot of data).
+The mixed model borrows information from other regions that are (hopefully) similar and provides much more precision than you could get by relying on the one region alone.
+
+For the IHME model, the borrowing of information occurs mostly on the far-side (the declining side) of the daily death rate, This is especially true for those regions that still only have data on the near-side (the rising side) of the daily death rate.
+These regions that have data on the far-side are mostly from China, South Korea, and Italy.
+
+Does it make sense to base the decline in the daily death rate in Chattahoochee county, Georgia on the decline in the daily death rate in Bergamo, Italy?
+Well, no not really, but what other choice do you have?
+
+The theoretical models in Epidemiology provide an alternate way to extrapolate the far-side of these curves.
+The shape of the curves are estimated by the rates at which people move from one compartment to another in the SEIR model and variations on that model.
+You can get partial information on the various rates from the near-side data and then plug in information from earlier epidemics to help with predictions on the far-side.
+
+In either the IHME model or the SEIR models require you to mix apples and oranges.
+In the IHME model, the oranges are places like Bergamo, Italy.
+In the SEIR mode, the oranges are epidemics like H1N1 influenza.
+The SEIR oranges, though, have one advantage of the IMHE oranges.
+There are many more SEIR oranges to choose from.
+The IMHE oranges are, at least for now, limited to China, South Korea, and Italy.
+
+#### The IHME model has changed over time
+
+All models provide different predictions as you accumulate more data.
+This leads to an inconsistent message over time and is largely unavoidable in any model.
+The IHME model, however, has undergone changes in the underlying mechanics that have increased its instability over time.
+Some of these changes were made to address criticisms in the model and some were made to get better fits to the data.
+One of the changes was noted above in the replacement of a single bell shaped curve with a weighted average of multiple bell shaped curves.
+
+The desire for consistency can be overrated.
+Stephen Senn talks about a medical biostatistician who doesn't believe in America because it wasn't in Christopher Columbus's original research plan.
+Still, the changes are a legitimate concern among some of the critics of the IHME model.
+
+### Resources
 
@@
 
-+ This article describes an earlier model.
++ This article describes an earlier model IHME model.
 
@@
 It provides, however, a good amount of detail about the IHME model and contrasts it with a model based on infectious disease theory, the Imperial College model.
+
+Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M.
+What 5 Coronavirus Models Say the Next Month Will Look Like.
+The New York Times, 2020-04-22.
+Available in [html format](https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html).
+
++ This article got me interested in the differences of the various models and has links that provide more details about the IHME model and several different SEIR models in greater detail.
 

### 2020-04-26 text/20/ihme-covid-model.Rmd
Updating 1999 posts

@@
 
+Etzioni R, COVID-19 Models: What Makes Them Tick? Timmerman Report, 2020-04-07. Available in [html format](https://timmermanreport.com/2020/04/covid-19-models-what-makes-them-tick/).
+
++ This blog post describes how a theoretical model (the Imperial College model) works but uses a different term, mechanistic, and describes the model in greater detail.
+Then she discusses the atheoretical IHME model and again uses a different term than I do: empirical.
+The author draws a nice contrast and concludes that all models have weaknesses.
+
 ### Articles in the peer-reviewed literature

### 2020-04-27 text/20/ihme-covid-model.Rmd
Created data-repositories.md

@@
 
+There are simulation models that are very close in spirit to the SEIR models, but which require a much greater degree of computational power.
+These simulations create random interactions among thousands or millions of individuals where each interaction has a specified probability of spreading the disease during an interaction between an infected individual and a susceptible individual.
+Other than the far greater amount of computing power needed, these models have the basic features as the SEIR models.
+
 The conflict between atheoretical models and theoretical models has a long, long history.
@@
 You will never resolve the controversy over theoretical versus atheoretical models, but in infectious disease modeling, the theoretical models have several important advantages. 
-
 First, the SEIR models are battle-hardened.
@@
 We'll talk more about this below.
+
+Fourth, the theoretical models allow you to simulate scenarios that have not yet occured, such as what happens when we introduce a vaccine or whether a lack of permanent immunity after recovery will lead to a new wave of infections.
 
@@
 
+Chow D.
+What we know about the coronavirus model the White House unveiled.
+NBC News, 2020-04-01.
+Available in [html format](https://www.nbcnews.com/science/science-news/what-we-know-about-coronavirus-model-white-house-unveiled-n1173601).
+
++ This article was also written in response to that White House press briefing. It emphasizes the many sources of uncertainty in these predictions.
+
 ### More technical articles
@@
 Where the criticism hits the mark is in how much this model relies on extrapolations from places that have data well past the peak of maximum cases/deaths (such as Italy, China, and South Korea) to places that have not yet reached their peak or are just barely past their peak.
+
+Chowell G, Sattenspiel L, Bansal S, Viboud C.
+Mathematical models to characterize early epidemic growth: A review.
+Phys Life Rev.
+2016;18:66–97.
+doi:10.1016/j.plrev.2016.07.005.
+Available in [html format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/) or [pdf format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/pdf/nihms-802148.pdf).
+
++ This article was written before the COVID=19 pandemic. It provides a nice review of how SEIR models work and applies them to a variety of pandemics.
+
+Kissler SM, Tediganto C, Goldstein E, Grad YH, Lipsitch M.
+Projecting the transmission dynamics of SARS-CoV-2 through the postpandemic period.
+Science 2020-04-14.
+DOI: 10.1126/science.abb5793, Available in [html format](https://science.sciencemag.org/content/early/2020/04/24/science.abb5793).
+
++ This article compares SARS-CoV-2 (the virus that causes COVID-19) to previous viruses to assess the chances that we will see future seasonal waves of COVID-19.
+This is an example of the type of modeling that can only be done with a theoretical model.
+
+FRED Modeling System. Available in [html format](https://fred.publichealth.pitt.edu/).
+
++ FRED stands for Framework for Reconstructing Epidemic Dynamics and represents a software system that uses simulation to model epidemics.
+It has been used to model outbreaks of measles and influenza, but not yet (as far as I can tell) for COVID-19.
+It is available for anyone to download and use on their own.
+
+### Acknowledgments
+
+Thanks to Chris Barker and Ruth Etzioni for helpful suggestions.

### 2020-05-07 text/20/ihme-covid-model.Rmd
Added bias-catalog

@@
 
+Mortality Modeling FAQ. Available in [html format](https://covid-19.tacc.utexas.edu/mortality-projection-faq/).
+
++ This is a different group (University of Texas at Austin) that is using a similar approach for modeling.
+It is interesting to see a common justification from both groups for their atheoretical approach.
+
+Social Distancing: Data Models for a Model Response to an Outbreak. Available in [html format](https://sph.umich.edu/news/2020posts/data-models-for-a-model-response-to-an-outbreak.html).
+
++ This is a question and answer session with Dr.
+Peter Song of the University of Michigan Department of Public Health.
+Dr.
+Song has fit some sophisticated SEIR models that can predict the impact of social distancing on the spread of COVID-19 cases.
+It is a nice summary, with enough details to let you know what the work involves without getting too technical.
+
 ### Articles in the peer-reviewed literature
@@
 
+Ayoub HH, Chemaitelly H, Mumtaz GR, Seedat S, Awad SF, Makhoul M, Abu-Raddad LJ.
+Characterizing key attributes of the epidemiology of COVID-19 in China: Model-based estimations.
+Preprint published in Medrxiv, 2020-04-11.
+Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.04.08.20058214v1.full.pdf).
+
++ This article fits an SEIR model to the Chinese COVID outbreak. It stratifies by age and outlines the differential equations in great clarity.
+
+Wang L, Wang G, Gao L, Li X, Yu S, Kim M, Wang Y, Gu Z.
+Spatiotemporal Dynamics, Nowcasting and Forecasting of COVID-19 in the United States.
+Preprint published in Arxiv, 2020-04-30.
+Available in [pdf format](https://arxiv.org/pdf/2004.14103.pdf).
+
++ This article fits an SEIR model to individual counties in the United States with a large number of county specific covariates.
+The model produces a [nice dashboard](https://covid19.stat.iastate.edu/).
+
 ### Acknowledgments

### 2020-05-07 text/20/reading-text-files.Rmd
Added bias-catalog

---
title: "Reading text files"
author: "Steve Simon"
source: "New"
date: "Created: 2020-04-30"
category: Blog post
tags: Data management
output: html_document
---

Many data files stored on the Internet use text format. Here is an explanation of what text format is and how you can read it into a program for data analysis.

<!---More--->

```{r ascii, echo=FALSE}
symbol_codes <- c(33:126)
symbols <- NULL
for (i in symbol_codes) {symbols <- c(symbols, rawToChar(as.raw(i)))} 
symbols <- setdiff(symbols, LETTERS)
symbols <- setdiff(symbols, letters)
symbols <- setdiff(symbols, as.character(0:9))
```

### What is text format?

I'll define the term "text format" in just a bit, but let me introduce a few other terms. A file that does not use text format is often called a "binary format" file. A data value that uses letters, such as a patient's name, or a mixture of letters and numbers, like a patient's address, is called a "string."

A text file is a file that uses 7-bit ASCII, a standard set of 127 characters. These characters include the capital letters,

+ `r paste(LETTERS, collapse=" ")`,

the lower case letters,

+ `r paste(letters, collapse=" ")`,

the numbers,

+ `r paste(0:9, collapse=" ")`,

and a variety of symbols,

+ `r paste(symbols, collapse=" ")`.

It also includes four special characters,

+ the blank character, 
+ the tab character, 
+ the line feed, and 
+ the carriage return.

It does not normally include letters from languages other than English, like characters with accents, umlauts, and other diacritical marks. It does not include Arabic, Chinese, Greek, or Japanese letters, or pretty much letters from any language other than English. It includes the dollar sign, but note the signs for Pounds, Euros, or Yen.

The 7-bit ASCII characters were standardized at a time when most of the computer developments were done in the United States. So the text file is one of many examples of American chauvinism, which can [sometimes cause problems](https://www.btb.termiumplus.gc.ca/tpv2guides/guides/favart/index-eng.html?lang=eng&lettr=indx_autr8o7vhUcC4c0s&page=9CMUV8aUjDMc.html).

Files using 7-bit ASCII are often called "human readable" files, though that again is a bit of chauvinism, as many humans can not recognize or understand the English alphabet.

Many files do not use text format. There are several reasons for this, relating to efficiency, but there is an increasing trend to store more and more data in text format.

### How can you tell if a file is in text format?

The easiest way to tell if a file is in text format is to read the page where the file was downloaded from. That page will often have a brief description of the file or it will include a link to a page providing a data dictionary.

The extension of the filename will sometimes help. An extension of .txt is almost always an indication that the file uses text format.

If you don't know if the file is in text format, try opening it in a text editor like notepad for Windows or TextEdit for the Mac. If the file refuses to open, then you know it is not in text format. If the file opens, but the result looks like gibberish, then you know it is not in text format.

![A non-text file displayed in notepad](../../../web/images/reading-text-files01.png)

The above image is an example of a binary format. You can recognize a couple of words, perhaps, but most of the file is unreadable.

The most common binary format for data files is Microsoft Excel, and how to work with those files is a topic for another day.

### First line variable names

Many text files will use the first line to provide a name to each variable. Here's an example.

![Data set on movies](../../../web/images/read-text-files03.png)

If a dataset does not include the names of the variables in the first line, then you have to find the names in a data dictionary.

### Varieties of text formats

There are two main varieties of text formats, fixed width and delimited.

### Fixed width format

In a fixed width format, the data is restricted to a specific coloumn or columns. The dataset shown above is an example of a fixed width format. When you encounter a fixed width format, you should expect to see a list of columns associated with each variable as part of the data dictionary.

![Data dictionary](../../../web/images/read-text-files06.png)

If you don't have this, then you need to count out the columns yourself, which is a very tedious job.

#### Delimited text format

There's an oft-repeated saying "Time is nature's way of keeping everything from happening at once." In a similar vein, delimiters are what keeps all your data from happening at once. Suppose you have data on a patients weight (in pounds) and height (in inches). If you represented that in a text file as

+ 24375

you might be able to recognize from the context that it represents a weight of 243 and a height of 75. But your computer would not be able to puzzle this out. You need to separate the individual data values using a special character, which is the delimiter.

#### Space delimited format

The easiest choice is to use a space to delimit values. So you would store a patients weight and height as

+ 243 75

and tell your computer to use the blank character as a delimiter. This works fine for numbers, but let's suppose that you have string data that represents the patient's name and address.

+ Donald Trump 1600 Pennsylvania Avenue 243 75

How do you let the computer know that the Donald plus Trump represent the patient's name and that 1600 plus Pennsylvania plus Avenue represents the patient's address?

Well, one way would be to use a different delimiter than the blank character. Another way would be to enclose the name in quotes and the address in quotes, giving you

+ "Donald Trump" "1600 Pennsylvania Avenue" 243 75

You don't need quotes around the 243 and the 75 as these pieces of data do not have any blanks.

The space delimited file is not that commonly used, for reasons that I've never understood, but here is an example.

![Example of a space delimited file](../../../web/images/read-text-files02.png)

### Comma delimited format

Far more common is the use of the comma as a delimiter. The data for our one patient would look like

+ Donald Trump,1600 Pennsylvania Avenue,243,75

If there is the possibility that a comma might be found in the data itself, then use quote marks to surround any strings. In general, it is usually a good idea to use quote marks for any string, even if you don't expect to see any commas in the data itself.

+ "Donald Trump","1600 Pennsylvania Avenue",243,75

Data files that use a comma delimiter often use the extension csv. This is an abbreviation of comma separated values. Here is an example of a comma delimited file.

![Dataset showing COVID-19 deaths over time](../../../web/images/read-text-files04.png)

### Tab delimited format

The tab character is an unusual character in the 7-bit ASCII universe. It is a single character that causes a jump to various tab stops. For most computer systems, when they display data with tab characters, they place the tab stops are in columns 9, 17, 25, 33, etc. This is intended to produce a clean and legible spacing of variables, but there is a lot that can go wrong. Often a tab delimited format looks somewhat orderly but with some odd alignments in certain rows.

![Dataset of revenues for selected movies](../../../web/images/read-text-files05.png)

### Missing value codes

Sometimes a data value is unknown. The patient didn't fill out the question on yearly income for your survey. The clinic personnel forgot to measure and record the patient's weight on a return visit. The urine specimen was lost before it got to the lab. There are many other reasons why a value might be missing in a dataset.

There are several ways to code missing values. One way is to leave the data value empty. If the weight of our President was missing, it might appear in a comma separated value data set this way

+ "Donald Trump","1600 Pennsylvania Avenue",,75

The two consecutive commas indicate that there is no value for weight. This format is not recommended, because some software systems (not many, but enough to cause concern) will convert an empty value to zero.

Another approach is to use a number code that is so extreme that it clearly outside the range of the data. This might be 999 (though there was a person who weighed of 1400 pounds) or -1. I recommend a value of -1 for weight because people are not helium balloons. The data would then look like

+ "Donald Trump","1600 Pennsylvania Avenue",-1,75

A single dot is often used as a missing value code.

+ "Donald Trump","1600 Pennsylvania Avenue",.,75

Another symbol commonly used for missing values is an asterisk.

+ "Donald Trump","1600 Pennsylvania Avenue",*,75

The letters NA are also commonly used.

+ "Donald Trump","1600 Pennsylvania Avenue",NA,75

Whatever the choice of missing value codes, it should be documented in the data dictionary and the reason for a missing value should be described there or in the research manuscript itself.

### Summary

Text files use a basic set of characters known as 7-bit ASCII. These files can come in a fixed width format or a delimited format. The most common delimiters are spaces, commas, and the tab character. Missing value codes can be represented by an empty entry, by an extreme value like 999 or -1, by a symbol like a single dot, or by the letters NA.

### 2020-05-12 text/20/ihme-covid-model.Rmd
Added bioconductor-library

@@
 source: "New"
-date: "2020-04-23"
 category: Blog post
@@
 ```
+
+This blog post was created on 2020-04-23 and last modified on `r Sys.Date()`.
 
@@
 
-The alternative SEIR models generally use information from the daily infection rate to estimate parameters and buil predictions, though they can incorporate information from the
+The alternative SEIR models generally use information from the daily infection rate to estimate parameters and build predictions, though they can incorporate information from the daily death rates as well.
+I have not seen any SEIR models that do this, but there are many SEIR models out there and I've only looked at a few.
 
@@
 
+### Coda
+
+I am not an expert in infectious disease epidemiology.
+In fact, I'm not much of an expert in anything.
+I am more or less a generalist.
+I know a little bit about a lot of different things.
+I can't really compete with teams of researchers who have been modeling epidemics for many decades.
+If I can contribute something, though, it is in a critical appraisal of the various prediction models that are out there for COVID-19.
+
+A critical appraisal is not a criticism.
+It is not a listing of all the faults that a model might have.
+Rather, it is a description of the features that a model has.
+It explains the rationale for these features and contrasts them with alternatives.
+It presents both the weaknesses AND THE STRENGTHS of a model.
+
+Everyone, including myself, gets trapped into a hypercritical approach to critical appraisal.
+The easiest thing in the world is to build a laundry list of faults.
+This can lead to a sense of cynicism towards all research, as all research has flaws.
+A good critical appraisal compares not to a standard of perfection in modeling, but rather to other practical modeling approaches.
+
+I hope I have not been too hypercritical in my review of the IHME model.
+Unfortunately, if there is anything I've learned over the years, it is a sense of humility about my own limitations.
+The faults I point out about others are the faults I suffer from myself.
+I can't get rid of my faults, but I can try to mitigate them.
+
+One of the sites listed above linked to an excellent article that got me to think a bit on the philosophical side of research.
+
+Gog JR.
+How you can help with COVID-19 modelling.
+Nat Rev Phys, 2020-04-08.
+https://doi.org/10.1038/s42254-020-0175-7.
+Available in [html format](https://www.nature.com/articles/s42254-020-0175-7) and [pdf format](https://www.nature.com/articles/s42254-020-0175-7.pdf).
+
 ### Acknowledgments

### 2020-07-18 src/clean_files.R
Reconcile, part 1

@@ -1,0 +1,56 @@
+# clean_files.R. This program was written by Steve Simon on 2020-05-24.
+
+suppressMessages(suppressWarnings(library(base64enc)))
+suppressMessages(suppressWarnings(library(curl     )))
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+v <- TRUE
+
+clean_files <- function(
+    search_string, replace_string="Not yet", 
+    dir_root="text/zotero", subdir_list=c("99", zpad(0:20), "no"),
+    file_pattern="*.bib", v=TRUE) {
+  if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
+  k <- 0
+  for (subdir in subdir_list) {
+    file_list <- list.files(dir_root %s% subdir, file_pattern)
+    "\nSearching through" %b% 
+      length(file_list) %b% 
+      "files in" %b%
+      dir_root %s% 
+      subdir %>%
+      br -> message
+    if (verbose) cat(message)
+    for (i_file in file_list) {
+      tx <- read_lines(dir_root %s% subdir %s% i_file)
+      found_lines <- str_which(tx, search_string)
+      if (length(found_lines)==0) next
+      k <- k+1
+      "\n\n" %0% i_file %>% br %>% cat
+      str_c(tx[found_lines], collapse="\n") %>% cat
+      if (replace_string!="Not yet") {
+        tx %<>% str_replace(search_string, replace_string)
+        "\n" %0% str_c(tx[found_lines], collapse="\n") %>% cat
+      if (ok_to_replace) writeLines(tx, dir_root %s% subdir %s% i_file)
+      }
+    }
+  }
+  "\n\nTotal files" %b% 
+    ifelse(ok_to_replace, "replaced", "to be replaced") %b%
+    k %>% cat
+}
+
+verbose <- TRUE
+verbose <- FALSE
+ok_to_replace <- TRUE
+ok_to_replace <- FALSE
+clean_files('category: Recommendation', dir_root="text", file_pattern="*")
+

### 2020-07-18 src/step0-identify-missing-fields.R

@@ -1,0 +1,31 @@
+# step0-identify-missing-fields.R
+# written by Steve Simon
+# created 2020-02-04
+
+# This program reads files and identifies those
+# that have missing fields.
+
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+year_list <- "20"
+for (i_year in year_list) {
+  list.files(path=paste0("text/", i_year), pattern="*.md", full.names=TRUE) -> file_list
+  for (i_file in file_list) {
+    file_text <- readLines(i_file)
+    if(any(str_detect(file_text, "urldate"))) next
+    cat("\n")
+    cat(i_file)
+    cat(" ")
+    i_file %>%
+      file.info %>%
+      pull(ctime) %>% 
+      str_sub(1, 10) %>%
+      cat
+  }
+}

### 2020-07-18 text/20/ihme-covid-model.Rmd
Reconcile, part 2

@@
 
+Robinson D. David's COVID Numbers. Available in [html format](https://davidscovidnumbers.wordpress.com/).
+
++ This blog post talks about COVID-19 in general terms with an emphasis on predictions for Minnesota and Stearns County.
+It tries to explain the meaning behind the many numbers that appear in various models.
+
 ### Articles in the peer-reviewed literature
@@
 
-+ This article was written before the COVID=19 pandemic. It provides a nice review of how SEIR models work and applies them to a variety of pandemics.
++ This article was written before the COVID-19 pandemic. It provides a nice review of how SEIR models work and applies them to a variety of pandemics.
 
@@
 + This article fits an SEIR model to the Chinese COVID outbreak. It stratifies by age and outlines the differential equations in great clarity.
+
+### Alternative prediction models
+
+There are many prediction models out there and each one seems to take a different approach. I'm just getting started but I want to provide a brief summary of some of these models.
+
+#### Imperial College of London
+
+[US Predictions](https://mrc-ide.github.io/covid19usa/#/)
+
+#### Iowa State
 

### 2020-07-19 src/step0-identify-missing-fields.R
Added dates to bib files

@@ -13,19 +13,22 @@ suppressMessages(suppressWarnings(library(rmarkdown)))
 suppressMessages(suppressWarnings(library(stringr  )))
 suppressMessages(suppressWarnings(library(tidyverse)))
 
-year_list <- "20"
+year_list <- c("18", "19", "20")
 for (i_year in year_list) {
-  list.files(path=paste0("text/", i_year), pattern="*.md", full.names=TRUE) -> file_list
+  list.files(path=paste0("text/", i_year), pattern="*.bib", full.names=TRUE) -> file_list
   for (i_file in file_list) {
     file_text <- readLines(i_file)
     if(any(str_detect(file_text, "urldate"))) next
     cat("\n")
     cat(i_file)
-    cat(" ")
+    cat(" ") 
     i_file %>%
       file.info %>%
       pull(ctime) %>% 
       str_sub(1, 10) %>%
-      cat
+      paste0("urldate = {", ., "},") -> new_line
+    cat(new_line)
+    new_text <- c(file_text[1], new_line, file_text[-1])
+    writeLines(new_text, con=i_file)
   }
 }

### 2020-07-22 src/clean_files.R
Change to categories, part 1

@@ -52,5 +52,40 @@ verbose <- TRUE
 verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
-clean_files('category: Recommendation', dir_root="text", file_pattern="*")
+clean_files('date: "Created: ', 'date:', dir_root="text", file_pattern="*")
 
+copy_files <- function(
+  old_root="text", new_root="../blogdown/content",
+  subdir_list=c("99", zpad(0:20), "no"),
+  file_pattern="md", v=TRUE) {
+  k <- 0
+  for (subdir in subdir_list) {
+    list.files(old_root %s% subdir) %>%
+      str_subset(file_pattern) -> file_list
+    "\nSearching through" %b% 
+      length(file_list) %b% 
+      "files in" %b%
+      old_root %s% 
+      subdir %>%
+      br -> message
+    if (verbose) cat(message)
+    new_sub <- ifelse(subdir > "30", "19", "20") %0% subdir
+    if (subdir=="no") new_dir <- "no"
+    for (i_file in file_list) {
+      k <- k+1
+      old_file <- old_root %s% subdir %s% i_file 
+      new_file <- new_root %s% new_sub %s% i_file
+      if (ok_to_copy) file.copy(old_file, new_file, overwrite=TRUE)
+      "\n" %b%
+        old_file %C% 
+        ifelse(ok_to_copy, "copied to", "to be copied to ") %b%
+        new_file %>% cat
+    }
+  }
+  "\n\nTotal files" %b%
+    ifelse(ok_to_copy, "copied:", "to be copied:") %b%
+    k %>% cat
+}
+ok_to_copy <- TRUE
+ok_to_copy <- FALSE
+copy_files(file_pattern="md$", subdir_list="20")

### 2020-07-22 text/20/codes-for-covid-19.Rmd
Change to categories, part 8

@@
 date: "2020-04-15"
-category: Blog post
+categories: Blog post
 tags: Electronic health records

### 2020-07-22 text/20/ihme-covid-model.Rmd
Change to categories, part 8

@@
 source: "New"
-category: Blog post
+categories: Blog post
 tags: Epidemiology

### 2020-07-22 text/20/map-projections.Rmd
Change to categories, part 8

@@
 date: "2020-03-15"
-category: Blog post
+categories: Blog post
 tags: Graphical display

### 2020-07-22 text/20/misuse-of-gradient.Rmd
Change to categories, part 8

@@
 date: "2020-03-11"
-category: Blog post
+categories: Blog post
 tags: Graphical display

### 2020-07-22 text/20/non-breaking-space.Rmd
Change to categories, part 8

@@
 date: "2020-01-07"
-category: Blog post
+categories: Blog post
 tags: Data management

### 2020-07-22 text/20/reading-text-files.Rmd
Change to categories, part 8

@@
 source: "New"
-date: "Created: 2020-04-30"
-category: Blog post
+date:2020-04-30"
+categories: Blog post
 tags: Data management

### 2020-07-22 text/20/smart-quotes.Rmd
Change to categories, part 8

@@
 date: "2020-03-02"
-category: Blog post
+categories: Blog post
 tags: Data management

### 2020-07-23 src/clean_files.R
Fixed multiple tags, part 1

@@ -15,13 +15,18 @@ source(file="src/standard_functions.R")
 v <- TRUE
 
 clean_files <- function(
-    search_string, replace_string="Not yet", 
-    dir_root="text/zotero", subdir_list=c("99", zpad(0:20), "no"),
-    file_pattern="*.bib", v=TRUE) {
+    search_string, 
+    replace_string="Not yet", 
+    dir_root="text/zotero", 
+    subdir_list=c("99", zpad(0:20), "no"),
+    file_pattern="*.bib", 
+    original_string=search_string,
+    v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
   k <- 0
   for (subdir in subdir_list) {
-    file_list <- list.files(dir_root %s% subdir, file_pattern)
+    list.files(dir_root %s% subdir) %>%
+      str_subset(file_pattern) -> file_list
     "\nSearching through" %b% 
       length(file_list) %b% 
       "files in" %b%
@@ -37,7 +42,7 @@ clean_files <- function(
       "\n\n" %0% i_file %>% br %>% cat
       str_c(tx[found_lines], collapse="\n") %>% cat
       if (replace_string!="Not yet") {
-        tx %<>% str_replace(search_string, replace_string)
+        tx %<>% str_replace_all(original_string, replace_string)
         "\n" %0% str_c(tx[found_lines], collapse="\n") %>% cat
       if (ok_to_replace) writeLines(tx, dir_root %s% subdir %s% i_file)
       }
@@ -52,12 +57,16 @@ verbose <- TRUE
 verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
-clean_files('date: "Created: ', 'date:', dir_root="text", file_pattern="*")
+clean_files(
+  'categories: ', 
+  'categories:\n- ', 
+  dir_root="text", 
+  file_pattern="md$")
 
 copy_files <- function(
   old_root="text", new_root="../blogdown/content",
   subdir_list=c("99", zpad(0:20), "no"),
-  file_pattern="md", v=TRUE) {
+  file_pattern="md$", v=TRUE) {
   k <- 0
   for (subdir in subdir_list) {
     list.files(old_root %s% subdir) %>%
@@ -88,4 +97,4 @@ copy_files <- function(
 }
 ok_to_copy <- TRUE
 ok_to_copy <- FALSE
-copy_files(file_pattern="md$", subdir_list="20")
+copy_files(file_pattern="md$")

### 2020-07-23 text/20/ihme-covid-model.Rmd
Fixed multiple tags, part 9

--- 2020-07-18 text/20/ihme-covid-model.Rmd
@@
 source: "New"
-category: Blog post
-tags: Epidemiology
+categories:
+- Blog post
+tags:
+- Epidemiology
 output: html_document
@@
 You will never resolve the controversy over theoretical versus atheoretical models, but in infectious disease modeling, the theoretical models have several important advantages.
+
 First, the SEIR models are battle-hardened.
@@
 Begley S.
-Influential Covid-19 model uses flawed methods and shouldn’t guide U.S.
+Influential Covid-19 model uses flawed methods and shouldn?t guide U.S.
 policies, critics say.
@@
 Phys Life Rev.
-2016;18:66–97.
+2016;18:66?97.
 doi:10.1016/j.plrev.2016.07.005.

### 2020-07-23 text/20/map-projections.Rmd
Fixed multiple tags, part 9

--- 2020-03-23 text/20/map-projections.Rmd
@@
 date: "2020-03-15"
-category: Blog post
-tags: Graphical display
+categories:
+- Blog post
+tags:
+- Graphical display
 output: html_document

### 2020-07-23 text/20/misuse-of-gradient.Rmd
Fixed multiple tags, part 9

--- 2020-03-11 text/20/misuse-of-gradient.Rmd
@@
 date: "2020-03-11"
-category: Blog post
-tags: Graphical display
+categories:
+- Blog post
+tags:
+- Graphical display
 output: html_document

### 2020-07-23 text/20/non-breaking-space.Rmd
Fixed multiple tags, part 9

--- 2020-02-29 text/20/non-breaking-space.Rmd
@@
 date: "2020-01-07"
-category: Blog post
-tags: Data management
+categories:
+- Blog post
+tags:
+- Data management
 output: html_document
@@
 ```{r insert}
-tst1 <- " "
+tst1 <- "?"
 tst1

### 2020-07-23 text/20/smart-quotes.Rmd
Fixed multiple tags, part 9

--- 2020-03-02 text/20/smart-quotes.Rmd
@@
 date: "2020-03-02"
-category: Blog post
-tags: Data management
+categories:
+- Blog post
+tags:
+- Data management
 output: html_document

### 2020-08-21 text/20/collaboration-videos.Rmd
Added collaboration-videos

---
title: "Setting the Stage: Statistical Collaboration Training Videos"
author: "Steve Simon"
source: "New"
date: 2020-08-20
categories:
- Recommended
tags:
- Human side of statistics
output: html_document
---

Julia Sharp, Emily Griffith, and Megan Higgs, Ann Hess. Setting the Stage: Statistical Collaboration Training Videos. Available in [html format](https://sites.google.com/view/juliasharp/other-resources/statistical-collaboration-videos).

<!---More--->

+ A series of ten short videos (the longest is 5 minutes, 41 seconds) with a simulated consulting experience shown. These videos offer some thoughts on how to handle difficult issues that might arise during consulting.

### 2020-08-21 src/copy_files.R

@@ -1,0 +1,51 @@
+# copy_files.R. This program was written by Steve Simon on 2020-08-19.
+
+suppressMessages(suppressWarnings(library(base64enc)))
+suppressMessages(suppressWarnings(library(curl     )))
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+v <- TRUE
+
+copy_files <- function(
+  old_root="text", new_root="../blogdown/content",
+  subdir_list=c("99", zpad(0:20), "no"),
+  file_pattern="md$", v=TRUE) {
+  k <- 0
+  for (subdir in subdir_list) {
+    list.files(old_root %s% subdir) %>%
+      str_subset(file_pattern) -> file_list
+    "\nSearching through" %b% 
+      length(file_list) %b% 
+      "files in" %b%
+      old_root %s% 
+      subdir %>%
+      br -> message
+    if (verbose) cat(message)
+    new_sub <- ifelse(subdir > "30", "19", "20") %0% subdir
+    if (subdir=="no") new_dir <- "no"
+    for (i_file in file_list) {
+      k <- k+1
+      old_file <- old_root %s% subdir %s% i_file 
+      new_file <- new_root %s% new_sub %s% i_file
+      if (ok_to_copy) file.copy(old_file, new_file, overwrite=TRUE)
+      "\n" %b%
+        old_file %C% 
+        ifelse(ok_to_copy, "copied to", "to be copied to ") %b%
+        new_file %>% cat
+    }
+  }
+  "\n\nTotal files" %b%
+    ifelse(ok_to_copy, "copied:", "to be copied:") %b%
+    k %>% cat
+}
+ok_to_copy <- TRUE
+ok_to_copy <- FALSE
+copy_files(file_pattern="md$")

### 2020-10-22 text/20/circos-plots.Rmd
Added evolution-data-science.md

---
title: "Making genome data come alive with circos plots"
source: "New"
date: "2020-10-20"
categories:
- Recommended
tags:
- Graphical display
output: html_document
---

Maria Nattestad. Making genome data come alive with circos plots. Medium 2017-09-25. Available in [html format](https://medium.com/@Marianattestad/a-treatise-on-making-circos-plots-from-genomic-data-7ff496849e0).

<!---more--->

I learned a new name today at a seminar, circos plots. You've probably seen these plots. There are individual labels arranged around the outside of a circle. Then graceful arcs inside the circle connect some of these labels. The colors and widths of the arcs represent important features of the data. I have mixed feelings about these graphs, but they are used so often that you should be familiar with them. This article introduces the concept of a circos plot and talks about some of the good and bad features. In particular, the author stresses that you should not try to jam too much information into one of these plots.

### 2020-10-26 text/20/sir-model.Rmd
Updated SpacingHtml.md

---
title: "A peek into the inner workings of pandemic prediction models"
source: "New"
categories:
- Blog post
tags:
- COVID-19
output: word_document
---

There was a cute meme that was going around on the Internet earlier this year that showed the time spent looking at exponential graphs that showed an exponential increase from January to February to March. This was, of course, the time that most of us were learning about how serious the COVID-19 crisis could be. There were a whole host of statistical models that were trying to predict how many people would become infected and how many people would die.

How do these models work? There are many approaches, but the most common uses some variation of the SIR compartment model.

Think of a compartment as a bucket of water with a hole in the side of the bucket. Water spills out at a certain rate, depending on the size and shape of the hole. The water that spills out of this bucket goes into a second bucket, which also has a hold in the side. This water goes into a third bucket, and so forth.

The compartments represent various disease states. In the simplest model, there are three compartments representing patients that are susceptible (*S*), infected (*I*), and recovered (*R*). The most interesting of these is the *I* compartment. The level in that compartment depends on how quickly it gets filled from the *S* compartment and how quickly it drains into the *R* compartment. You want don’t want the *I* compartment to get too full for many reasons, but mostly because a large number of infected individuals can tax the available health care resources.

```{r, echo=FALSE}
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(tidyverse)))
knitr::opts_chunk$set(echo=FALSE, fig.width=6, fig.height=3)
```

```{r}
sz <- 1.2
xc <- 20
yc <- 80
xbox <- 10*c(-1, -1,  1,  1, -1) + xc
ybox <- 10*c(-1,  1,  1, -1, -1) + yc
data.frame(x=xbox, y=ybox) %>%
  ggplot(aes(x, y)) +
  expand_limits(x=c(0, 100), y=c(0, 100)) +
  theme_void() +
  geom_path(size=sz) + 
  geom_text(x=xc, y=yc, label="S") +
  geom_path(x=xbox+30, y=ybox-30, size=sz) +
  geom_text(x=xc+30, y=yc-30, label="I") +
  geom_path(x=xbox+60, y=ybox-60, size=sz) +
  geom_text(x=xc+60, y=yc-60, label="R") +
  geom_curve(
    x=xc+40, y=yc-35, xend=xc+55, yend=30, curvature=-0.3, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) -> sir_plot
sir_plot +
  geom_curve(
    x=xc+10, y=yc- 5, xend=xc+25, yend=yc-20, curvature=-0.3, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches")))
```

##### Figure 1. Susceptible/Infected/Recovered (SIR) model

There is an interesting equation that describes the *I* compartment:

$dI/dt = \beta (S \times I) - \gamma I$

Let's take this equation apart to see what is going on.

First the derivative notation, $dI/dt$ tells you that the model is explaining not the amount in the *I* compartment, but the rate at which the level in the *I* compartment changes. You do not want the rate to be postive, as that means the compartment is filling faster than it is draining.

The term $- \gamma I$ represents the rate at which the compartment drains, and this is a constant $\gamma$ times the amount in the compartment *I*. If $\gamma$ is small, the compartment drains slowly and infectious people stay infectious for a very long time, weeks or maybe months. If $\gamma$ is large, then the compartment drains quickly meaning that infectious people recover very quickly, maybe within a few days.

For most infectious diseases, you do not have any control over $\gamma$. It is controlled by how quickly your body's defenses can beat back the infection and rid your body of the evil virus.

The term $\beta S \times I$ tells you how quickly the compartment is filling, and this is a constant $\beta$ times the product of the amounts in compartment *S* and compartment *I*. You do have a lot of control over this part of the equation.

Since the filling rate is the product of *S* and *I*, the *I* compartment fills fast only when *S* and *I* are both large. If *I* is small, that means that there are not a lot of people around to do the infecting. If *S* is small that means that there are not a lot of people around to get infected.

Typically *I* is small early in the course of a pandemic, and represents the early part of the exponential curve where the disease spreads slowly. Over time, however, *I* may grow and with it the rate at which *I* grows will increase, leading to the dreaded exponential increase in infections.

Eventually as the *S* compartment drains into the *I* compartment, *S* will shrink enough so that the product $S \times I$ becomes small again. This represents herd immunity, the setting where the number of people susceptible to infection becomes too small to keep the epidemic going. It is like a forest fire that burns itself out because there is nothing left to burn.

### Dynamics that avoid an exponential explosion of cases

You can avoid the dreaded exponential curve, but it is not easy. You have to insure that the *I* compartment never grows big enough to start that exponential rise. This occurs when the constant $\beta$ is smaller than the constant $\gamma$.

We designate $R_0$ as the ratio, $\beta / \gamma$.

This ratio plays a key role in the spread of disease. If it is less than one, then the *I* compartment drains faster into the *R* compartment than it can get filled by the *S* component. Keeping *I* low keeps the pandemic from growing exponentially. If the ratio $R_0$ is larger than 1, then the *I* compartment fills faster than it can drain. We want to avoid this, if we can.

You make the ratio small by decreasing the numerator, $\beta$ or increasing the denominator, $\gamma$. There is little that you can do to influence $\gamma$. If you could strengthen your immune system so you can rid yourself of the infection faster, that would be great. But therapies to strengthen your immune system are not that easy to come by.

Decreasing $\beta$, on the other hand, can be done. Even if there are a large number of susceptible people and a large number of infected people, you can reduce $\beta$ through policies like masking and social distancing.

### Concerns about heteorgeneity

A concern about compartment models is that it treats every individual the same. Some people, however, are less likely to get infected than others. Some parts of the country, because of urban crowding, are more likely to see higher rates of infection than other parts of the country. This leads to heterogeneity, and if you don't properly account for this heterogeneity, the accuracy of your predictions will suffer.

### Adding compartments to enhance the SIR model

The compartment model allows you to add extra layers to more closely model the complex process of disease propogation. You can, for example, add an additional compartment $E$, between the susceptible and infected compartments. The would represent people who were exposed to the virus but who were not yet infectious. The infected compartment could have two holes, one which drains into the recovered compartment with the other draining into compartment for hospitalization and/or death.

```{r}
sir_plot +
  geom_path(x=xbox, y=ybox-30, size=sz) +
  geom_text(x=xc, y=yc-30, label="E") +
  geom_curve(
    x=xc, y=yc-10, xend=xc, yend=yc-20, curvature=0, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) +
  geom_curve(
    x=xc+10, y=yc-30, xend=xc+20, yend=yc-30, curvature=0, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) -> seir_plot
seir_plot
```

##### Figure 2. Susceptible/Exposed/Infected/Recovered (SIR) model

You can also describe multiple paths out of the *I* compartment. Some of the individuals in this compartment be hospitalized (the *H* compartment) and from the hospital, some could recover and some could die (the *D* compartment).

```{r}
seir_plot + 
  geom_path(x=xbox+30, y=ybox-60, size=sz) +
  geom_text(x=xc+30, y=yc-60, label="H") +
  geom_path(x=xbox, y=ybox-60, size=sz) +
  geom_text(x=xc, y=yc-60, label="D") +
  geom_curve(
    x=xc+30, y=yc-40, xend=xc+30, yend=yc-50, curvature=0, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) +
  geom_curve(
    x=xc+40, y=yc-60, xend=xc+50, yend=yc-60, curvature=0, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) +
  geom_curve(
    x=xc+20, y=yc-60, xend=xc+10, yend=yc-60, curvature=0, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) -> seirhd_plot
seirhd_plot
```

##### Figure 3. Adding Hospitalized/Death compartments to the SEIR model

The nightmare scenario for infectious disease Epidemiologists is when the recovered compartment *R* can drain back into the susceptible compartment *S*. This could happen if your immunity wears off after a certain amount of time or when the virus mutates to a form that your current immunity system no longer recognizes. This can produce a setting where the infection never dies down, or where the infection goes up and down in a regular cycle. Although there are some reports of individuals becoming infected a second time, there is not enough data yet on whether this occurs at a rate that would be an issue.

```{r}
seirhd_plot +
  geom_curve(
    x=xc+65, y=yc-50, xend=xc+10, yend=yc+5, curvature=0.2, size=sz, 
    arrow = arrow(type = "closed", length=unit(0.15, "inches")))
```

##### Figure 4. A model with Recovered returning to Susceptible

### Concerns about the data sources

It is interesting to examine these models from a mathematical perspective, but predictions about the future course of any pandemic requires linking the theoretical models to actual data.

For COVID-19, data is free for the taking, thanks to Johns Hopkins University, the New York Times, USA Facts, and others. If you have more than a little skill in data science, you can build your own model and run it on real data.

There is concern, however, about the raw data. Tests for COVID-19 will sometimes produce false positives and false negatives. There is a lag between when the test is done and when a positive result gets reported. More importantly, not everyone who has an infection will get tested.

You can build a model based on the number of deaths rather than the number of infections. While there are no false positives or negatives on an outcome like death, there is uncertainty on when a death should be associated with COVID-19.

While these are serious problems, they are not new to infectious disease Epidemiologists. There are ways to adjust models for some types of over and under reporting. When adjustments are not available, there is often the ability to recognize the direction in which these problems will bias the model estimates.

### Concerns about heteorgeneity

A concern about compartment models is that it treats every individual the same. Some people, however, are less likely to get infected than others. Some parts of the country, because of urban crowding, are more likely to see higher rates of infection than other parts of the country. The models are affected by important temporal changes such as business shutdowns and reopenings, increased testing, and mandates for masking. This leads to heterogeneity, and if you don't properly account for this heterogeneity, the accuracy of your predictions will suffer.

[Talk about agent based models and state/county projections.]

### References

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3935673/

https://www.washingtonpost.com/graphics/2020/health/coronavirus-how-epidemics-spread-and-end/

https://www.washingtonpost.com/graphics/2020/world/corona-simulator/

### 2020-10-30 text/20/sir-model.Rmd
More work on sir paper

@@
 suppressMessages(suppressWarnings(library(tidyverse)))
+suppressMessages(suppressWarnings(library(usmap)))
+
 knitr::opts_chunk$set(echo=FALSE, fig.width=6, fig.height=3)
@@
 Even if there are a large number of susceptible people and a large number of infected people, you can reduce $\beta$ through policies like masking and social distancing.
+
+### A vaccination model
+
+When you get a vaccine, the dynamics change. The susceptible
 
@@
 
+### Agent based models
+
+An agent based model replaces buckets of water with individual particles.
+There particles move around at varying speeds and when an infected particle collides a susceptible particle, the latter also becomes infected.
+Think of it as a giant simulated game of tag, with the infected people playing the role of "it".
+You can control each particle differently.
+Faster moving particles represent individuals who do not practice social distancing well.
+You can assign each particle a different level of susceptibility as well, so that some become infected with very high probability after a collision and others become infected with a much lower probability.
+These models can include other sources of variation, such as changes in social distancing after a lockdown.
+
+Agent based models have been compared to a giant game of the Sims. The number of people in the agent based models, however, is often in the thousands or even millions.
+
+### Hierarchical models
+
+A hierarchical model divides up the country into regions (typically individual counties) and lets each region have a different SIR model.
+With 3,141 counties, this represents a huge increase in complexity, one that could potentially choke even the fastest computers.
+The saving grace in these hierarchical models is that you can often rely on the fact that similar regions will behave similarly in the dynamics of the SIR model.
+
+You can model the similarity of regions using geographic proximity (New York counties and New Jersey counties should have similar dynamics, for example).
+You could also model the similarities using U.S.
+Census information (counties with very low population densities should have similar dynamics, for example).
+
+### Time varying parameters
+
+Time is one of the biggest sources of heterogeneity. The spread of an infectious disease changes over time as personal behavior changes and as government interventions change.
+
 ### References
@@
 https://www.washingtonpost.com/graphics/2020/world/corona-simulator/
+
+https://www.nature.com/articles/d41586-020-01003-6

### 2020-11-06 text/20/markdown-papers.Rmd
Started work on rmarkdown-papers

---
title: "How I use R Markdown to write research papers"
source: "New"
categories:
- Blog post
tags:
- Writing research papers
output: html_document
---

I joined a writing group at UMKC and am making very good progress on a couple of research papers. The group meets once a week to talk about everyone's progress and to share ideas. One interesting suggestion was the use of [Notion](https://www.notion.so/), software that organizes all your myriad writing tasks (actually tasks in general). I shared how I use R Markdown for writing and promised that I'd provide some resources about it before the next meeting.

<!--more-->

R Markdown is an implementation of [Markdown](https://commonmark.org/help/) inside the R programming language. R Markdown also leverages features in [LaTex](https://www.latex-project.org/), [MathJax](https://www.mathjax.org/), and [Pandoc](https://pandoc.org/).

If you use the [RStudio integrated development environment](https://rstudio.com/) to run R, then you already have many of the features of R built-in. There are several R packages that use or extend R Markdown:

+ [bookdown](https://bookdown.org/)
+ [blogdown](https://alison.rbind.io/post/2017-06-12-up-and-running-with-blogdown/)
+ [knitr](https://yihui.org/knitr/)
+ [rmarkdown](https://cran.r-project.org/web/packages/rmarkdown/index.html)

I use R Markdown to produce early drafts of research publications. It simplifies writing in several different ways.

**Figures** R can produce beautiful graphs, and with R Markdown, you can insert these graphs anywhere in your document. You can't control page breaks easily in R Markdown, however. Here's an example of a simple graph.

```{r, echo=FALSE}
suppressMessages(suppressWarnings(library(tidyverse)))
x=seq(-1, 1, length=10000)
y <- sin(1/x)
data.frame(x, y) %>%
  ggplot(aes(x, y)) + geom_line()
```

### 2020-11-10 text/20/markdown-papers.Rmd
Added experience-not-quantified

@@
 output: html_document
+bibliography: rmarkdown-papers.bib
+csl: biomed-central.csl
+nocite: |
+  @ziemann-2016, @zeeberg-2004
 ---
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, fig.width=6, fig.height=3}
+suppressMessages(suppressWarnings(library(broom)))
+suppressMessages(suppressWarnings(library(knitr)))
 suppressMessages(suppressWarnings(library(tidyverse)))
 x=seq(-1, 1, length=10000)
-y <- sin(1/x)
+y <- x*sin(1/x)
 data.frame(x, y) %>%
@@
 ```
+
+With the ability to plot polygons, line segments, and arrows, you can create informational diagrams like this.
+
+```{r, echo=FALSE}
+sz <- 1.2
+xc <- 20
+yc <- 80
+xbox <- 10*c(-1, -1,  1,  1, -1) + xc
+ybox <- 10*c(-1,  1,  1, -1, -1) + yc
+data.frame(x=xbox, y=ybox) %>%
+  ggplot(aes(x, y)) +
+  expand_limits(x=c(0, 100), y=c(0, 100)) +
+  theme_void() +
+  geom_path(size=sz) + 
+  geom_text(x=xc, y=yc, label="S") +
+  geom_path(x=xbox+30, y=ybox-30, size=sz) +
+  geom_text(x=xc+30, y=yc-30, label="I") +
+  geom_path(x=xbox+60, y=ybox-60, size=sz) +
+  geom_text(x=xc+60, y=yc-60, label="R") +
+  geom_curve(
+    x=xc+40, y=yc-35, xend=xc+55, yend=30, curvature=-0.3, size=sz, 
+    arrow = arrow(type = "closed", length=unit(0.15, "inches"))) -> sir_plot
+sir_plot +
+  geom_curve(
+    x=xc+10, y=yc- 5, xend=xc+25, yend=yc-20, curvature=-0.3, size=sz, 
+    arrow = arrow(type = "closed", length=unit(0.15, "inches")))
+```
+
+The code to produce this graph is a bit tedious to create, but since the code that produces this graph is part of the R Markdown file, you can edit it as you are editting other parts of your paper.
+If you want to relabel the middle box as an "E", you do not need to exit and run a different program, and then cut-and-paste back into your documents.
+
+**Formulas**
+
+You can get really nice looking mathematical formulas inserted into your paper using a really simple syntax. For example, the code
+
+V = \\frac{4}{3} \\pi r^3
+
+becomes
+
+$V = \frac{4}{3} \pi r^3$
+
+when you sandwich it between two dollar signs.
+
+**Tables**
+
+I used to spend a lot of time retyping numbers from what the statistical software said into a table that was organized to make it more readable.
+That is a tedious and error prone practice, and if your data set changes, even just a little bit, you have to re-type everything all over again.
+You can produce nice looking tables in R Markdown.
+It takes about as much work as just retyping everything, but once the table looks nice, you can use the code over again for other projects.
+I am still learning how to do tables well, but here is some output that I generated to show the potential value.
+
+```{r}
+lm(breaks ~ wool*tension, data = warpbreaks) %>%
+  tidy %>%
+  rename(t.test=statistic) %>% 
+  kable(
+    digits=c(1, 1, 1, 2, 4),
+    caption="Analysis of variance table")
+```
+
+I can change how the values in this table are rounded or the names in the header row with just one line of code. Other enhancements may take a bit more work.
+
+**Bibliographies**
+
+If you have your references stored in BibTeX format, you can place them at the end of your document.

### 2020-11-18 text/20/markdown-papers.Rmd
Added interval-censored.md

@@
 
-**Figures** R can produce beautiful graphs, and with R Markdown, you can insert these graphs anywhere in your document.
+**Figures**
+
+R can produce beautiful graphs, and with R Markdown, you can insert these graphs anywhere in your document.
 You can't control page breaks easily in R Markdown, however.
@@
 
+**Output formats**
+
+You can specify a variety of output formats. The ones I use most often are
+
++ docx
++ html
++ pdf
++ pptx
+
+The pdf option requires that you have a version of LaTeX installed on your computer. I use [MikTeX](https://miktex.org/), but others work just fine.
+
 **Bibliographies**
@@
 If you have your references stored in BibTeX format, you can place them at the end of your document.
+
+The bibliography always appears at the end of the document.
+
+**Final comments**
+
+Most of the people that I collaborate with do not use R Markdown, so I typically will start a draft in R Markdown, but when it gets to the point where I can share it with my collaborators, I send the document in Word format and we make any further changes there.
+If I am the sole author, then I can keep the file in R Markdown.

### 2020-12-21 text/20/markdown-papers.Rmd
Added nas-reproducibility.md

@@
 output: html_document
-bibliography: rmarkdown-papers.bib
-csl: biomed-central.csl
-nocite: |
-  @ziemann-2016, @zeeberg-2004
 ---
@@
 
-The bibliography always appears at the end of the document.
+The bibliography always appears at the end of the document. I'm not showing a bibliography for technical reasons, but it is pretty easy.
 

### 2021-01-21 text/20/markdown-papers.Rmd
Added niss.md

@@
 source: "New"
+date: "2020-11-21"
 categories:

### 2021-01-21 src/copy_files.R

@@ -16,7 +16,7 @@ v <- TRUE
 
 copy_files <- function(
   old_root="text", new_root="../blogdown/content",
-  subdir_list=c("99", zpad(0:20), "no"),
+  subdir_list=c("99", zpad(0:21), "no"),
   file_pattern="md$", v=TRUE) {
   k <- 0
   for (subdir in subdir_list) {

### 2021-06-02 text/21/anonymizing-ids.Rmd
Adding anonymizing-ids.Rmd

---
title: "Anonymizing ids"
author: "Steve Simon"
source: "New"
date: "2020-04-15"
categories:
- Blog post
tags:
- Privacy in research
output: html_document
---

I'm working on a project that might be exempt from IRB review...if I can convert a particular ID code to an anonymous code. The ID codes are an eight digit number. Here's a simple approach that should work for this project.

<!---More--->

The steps are

1. Create a random number between 100,000,000 and 900,000,000.

2. Add the same random number to each eight digit id. This will produce a new nine digit id.

3. Link the eight and nine digit ids to the raw data.

4. Store the merged data with the nine digit id, but not the eight digit id.

5. Shred the random number.

The process of shredding the random number will prevent anyone from converting the nine digit id back into an eight digit id.

### 1. Create a random number between 100,000,000 and 900,000,000.

There are several easy ways to create a random number with a specific range. In Microsoft Excel, you would place the formula

```{}
=randbetween(100000000, 900000000)
```

Count your zeros carefully, there should be eight of them for both numbers.

This will produce a value. Mine is 480035489, but yours will be different.

### 2. Add the same random number to each eight digit id. This will produce a new nine digit id.

Here's some data with some made up id values

```{r, echo=FALSE}
library(knitr)
x <- rep(c(12345678, 23456789, 34567890), c(4, 3, 2))
kable(x)
```

Here's what those ID numbers look like when you add my random number

```{r, echo=FALSE}
kable(data.frame(x=x, y=x+480035489))
```

### 2021-06-11 text/21/anonymizing-ids.Rmd
Completed anonymizing_ids

@@
 source: "New"
-date: "2020-04-15"
+date: "2020-06-10"
 categories:
@@
 
-2. Add the same random number to each eight digit id. This will produce a new nine digit id.
+2. Add the same random number to each eight digit id.
 
@@
 
-### 2. Add the same random number to each eight digit id. This will produce a new nine digit id.
+There is a complication in Microsoft Excel that you have to work around.
+The randbetween function will recalculate everytime you make a change anywhere else in the workbook.
+You need to "freeze" the random number by copying and pasting it to a different cell using the "past values" option under "paste special."
+
+### 2. Add the same random number to each eight digit id.
 
@@
 ```{r, echo=FALSE}
-library(knitr)
-x <- rep(c(12345678, 23456789, 34567890), c(4, 3, 2))
-kable(x)
+suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(knitr)))
+suppressMessages(suppressWarnings(library(kableExtra)))
+suppressMessages(suppressWarnings(library(magrittr)))
+private_id <- 12345678+(0:7)*11262783
+data.frame(private_id=private_id) %>%
+  kbl %>%
+  kable_minimal(full_width=FALSE, position="left")
 ```
 
-Here's what those ID numbers look like when you add my random number
+Here's what those ID numbers look like when you add my random number. We now have a random nine digit id associated with each eight digit id.
 
 ```{r, echo=FALSE}
-kable(data.frame(x=x, y=x+480035489))
+e <- data.frame(private_id=as.character(private_id), random_id=as.character(private_id+480035489))
+
+e %>%
+  kbl %>%
+  kable_minimal(full_width=FALSE, position="left")
 ```
+
+### 3. Link the eight and nine digit ids to the raw data.
+
+Here's the file with the actual data and the private ID code
+
+```{r, echo=FALSE}
+d <- data.frame(
+  private_id=c(rep(private_id[1:3], c(4, 3, 2)), "..."),
+  report_date=c(as.character(Sys.Date()-round(runif(9, 0, 60))), "..."),
+  data=c(round(runif(9, 100, 199)), "..."))
+d %>%
+  kbl %>%
+  kable_minimal(full_width=FALSE, position="left")
+```
+
+Merge the two files together.
+
+```{r, echo=FALSE}
+e %>%
+  bind_rows(data.frame(private_id="...", random_id="...")) -> f
+merge(f, d) -> g
+# g <- g[c(2:9, 1), ]
+h <- g[c(2:9, 1), ]
+h %>%
+  tibble %>%
+  data.frame %>%
+  kbl %>%
+  kable_minimal(full_width=FALSE, position="left")
+```
+
+### 4. Store the merged data with the nine digit id, but not the eight digit id.
+
+You now have a new random_id in place of the private id.
+
+```{r, echo=FALSE}
+h %>%
+  select(-private_id) %>%
+  kbl %>%
+  kable_minimal(full_width=FALSE, position="left")
+```
+
+### 5. Shred the random number.
+
+Once the random number is destroyed, the nine digit id cannot be converted back to the eight digit id.

### 2021-08-30 text/21/anonymizing-ids.Rmd
Added poisson-inference.md

@@
 e <- data.frame(private_id=as.character(private_id), random_id=as.character(private_id+480035489))
-
 e %>%

### 2021-08-30 text/21/poisson-inferences.Rmd
Added poisson-inference.md

---
title: "Working with the Poisson distribution"
author: "Steve Simon"
source: "New"
date: "2020-08-30"
categories:
- Blog post
tags:
- Probability distributions
output: html_document
---

Many statistics, such as mortality statistics, are based on fundamental counting processes. If you want to understand these processes, you should start by trying to understand the Poisson distribution. Here's a brief introduction.

```{r setup}
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### Definition of a Poisson distribution

A random variable X, defined on the set of non-negative integers (0, 1, 2, ...), is said to have a Poisson distribution (X ~ Pois($\lambda$) if

$P[X=k]=\frac{\lambda^ke^{-k}}{k!}$ where $\lambda > 0$

### Mean and variance of a Poisson random variable.

Without too much difficulty, you can show that

$E[X]=\lambda$

$V[X]=\lambda$

### sums of Poisson random variables

If

$X_1$ ~ Pois($\lambda_1$) and

$X_2$ ~ Pois($\lambda_2$) and

$X_1$ and $X_2$ are independent, then

$X_1+X_2$ ~ Pois($\lambda_1+\lambda_2$).

### Visualization of Poisson probabilities

```{r bar-charts}
data.frame(x=0:25) %>%
  mutate(p=dpois(x, 0.5)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with lambda=0.5")
data.frame(x=0:25) %>%
  mutate(p=dpois(x, 2)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with lambda=2")
data.frame(x=0:25) %>%
  mutate(p=dpois(x, 8)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with lambda=8")
```

### Relationship to the Binomial distribution

If Y has a binomial distribution with n trials and probability of success p (X ~ Bin(n,p)), then for large n and small p, you can approximate Y with X ~ Pois($\lambda$=np).

### Inference for unknown value of lambda

If you collect data on a single Poisson random variable X where the parameter $\lambda$ is unknown, then you can estimate $lambda$ with a reasonable amout of precision with just that one value.

$\hat{\lambda}=X$

and a crude 95% confidence interval for $\lambda$ is

$X \pm 2 \sqrt{X}$

This confidence interval does poorly for small values of $\lambda$ because of the skewness of the Poisson distribution. it also can sometimes produce confidence intervals with a negative lower limit.

### Estimation of a rate

A count X divided by a measure of time, T, is called a rate (R=X/T). If you assume that

x ~ Pois($\lambda$T)

where T is known, but $lambda$ is unknown, then a 95% confidence interval for $lambda is

$\frac{X}{T} \pm 2 \frac{\sqrt{X}}{T}$

### An improved confidence interval

The natural log transformation helps produce a better confidence interval.

If

X ~ Pois(T$\lambda$) and

Y=ln(X), then

E[Y] $\approx$ ln($\lambda T$) and

V[Y] $\approx \frac{1}{\lambda T}$

An approximate 95% confidence interval for ln($\lambda  T$) is

ln(X) $\pm 2 \sqrt{\frac{1}{X}}$

Exponentiate both sides to get a confidence interval for $\lambda T$.

Divide by T to get a confidence interval for $\lambda$.

### Examples

http://www.pmean.com/07/ConfidenceIntervalForRate.html

https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf

### 2021-08-30 text/21/poisson-inferences.Rmd
Updated poisson-inferences.md

@@
 
-where T is known, but $lambda$ is unknown, then a 95% confidence interval for $lambda is
+where T is known, but $\lambda$ is unknown, then a 95% confidence interval for $\lambda$ is
 
@@
 
+The general rule is that a nonlinear transformation leads to a similar change in the expected value, but a change in the variance that is related to the Taylor series expansion.
+
+If X has an expected value, $\mu$ and a variance, $\sigma^2_x$, then g(X) can be written as
+
+$g(X) \approx g(\mu)+(X-\mu)g'(\mu)$
+
+which can be used to show that
+
+$Var(g(X))=(g'(\mu))^2 \sigma^2_x$
+
 If
 
-X ~ Pois(T$\lambda$) and
+X ~ Pois($\lambda T$) and
 
-Y=ln(X), then
+$Y = g(X) = ln(X)$, then
 
-E[Y] $\approx$ ln($\lambda T$) and
+$g'(X)=\frac{1}{X}$, and
 
-V[Y] $\approx \frac{1}{\lambda T}$
+V[Y] $\approx (g'(\mu))^2 \sigma^2_x = (\frac{1}{\lambda T})^2 \lambda T = \frac{1}{\lambda T}$
 
@@
 
-ln(X) $\pm 2 \sqrt{\frac{1}{X}}$
+$ln(X) \pm 2 \sqrt{\frac{1}{X}}$
 
@@
 
+```{r echo=FALSE}
+x <- 17
+n <- 1989
+lo <- x-2*sqrt(x)
+hi <- x+2*sqrt(x)
+log_lo <- log(x)-2*sqrt(1/x)
+log_hi <- log(x)+2*sqrt(1/x)
+exp_lo <- exp(log_lo)
+exp_hi <- exp(log_hi)
+```
+
+The Pennsylvania Department of Health has a web page oulining calculations for rates with two simple examples.
+The infant mortality rate in Butler County in 1984 was 17 deaths out of 1,989 births.
+This is actually a proportion, but let's treat it as a rate here.
+
+The death rate is 17/ 1989 =`r 17/1989`.
+To avoid dealing with small numbers, let's quote this as a rate per 1,000 live births.
+1000 * 17/ 1989 = `r 1000*17/1989`.
+if we wanted a confidence interval for $\lambda T$, it would be
+
+17 $\pm$  2 $\sqrt{17}$ = `r lo` to `r hi`
+
+Divide everything by 1,989 and multiply by 1,000 to get
+
+`r lo/1.989` to `r hi/1.989` which you should round to get 
+`r round(lo/1.989, 1)` to `r round(hi/1.989, 1)`
+
+Let's rework this using the better formula.
+
+The confidence limits on the log scale are
+
+$ln(17) \pm 2 \sqrt{\frac{1}{17}}$ = `r log_lo` to `r log_hi`
+
+Exponentiate these values to get
+
+`r exp_lo` to `r exp_hi`
+
+Divide by 1,989 and multiply by 1,000 to get
+
+`r exp_lo/1.989` to `r exp_hi/1.989`
+
+and don't forget to round this: `r round(exp_lo/1.989, 1)` to `r round(exp_hi/1.989, 1)`.
+
+### References
+
 http://www.pmean.com/07/ConfidenceIntervalForRate.html
@@
 https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf
+
+https://en.wikipedia.org/wiki/Taylor_expansions_for_the_moments_of_functions_of_random_variables
+
+https://en.wikipedia.org/wiki/Delta_method
+
+https://en.wikipedia.org/wiki/Poisson_distribution
+
+https://en.wikipedia.org/wiki/Binomial_distribution#Poisson_approximation
+
+http://www.pmean.com/definitions/or.htm
+
+http://www.pmean.com/01/oddsratio.html
+
+http://www.pmean.com/04/Rates.html

### 2021-08-31 text/21/poisson-inferences.Rmd
Added link to poisson-inference.md
Minor fix to poisson-inferences.Rmd
Linked in some references for poisson-inference.md

--- 2021-08-30 text/21/poisson-inferences.Rmd
@@
 
+If you have series of Poisson random variables,
+
+$X_1, X_2,...,X_n$ all indpendent and all with the same Pois($\lambda$) distribution, then
+
+$\Sigma_{i=1}^n X_i$ ~ Pois($k \lambda$)
+
+These properties and more are described on the [Wikipedia page on the Poisson distribution][pois1].
+
 ### Visualization of Poisson probabilities
@@
 If Y has a binomial distribution with n trials and probability of success p (X ~ Bin(n,p)), then for large n and small p, you can approximate Y with X ~ Pois($\lambda$=np).
+Wikipedia has a [nice explanation][pois2] of this.
 
@@
 
-and a crude 95% confidence interval for $\lambda$ is
+and a crude $1-\alpha$ confidence interval for $\lambda$ is
 
-$X \pm 2 \sqrt{X}$
+$X \pm z_{\alpha/2} \sqrt{X}$
 
@@
 
-x ~ Pois($\lambda$T)
+x ~ Pois($\lambda T$)
 
-where T is known, but $lambda$ is unknown, then a 95% confidence interval for $lambda is
+where T is known, but $\lambda$ is unknown, then a $1-\alpha$ confidence interval for $\lambda$ is
 
-$\frac{X}{T} \pm 2 \frac{\sqrt{X}}{T}$
+$\frac{X}{T} \pm z_{\alpha/2} \frac{\sqrt{X}}{T}$
+
+I have a [page on my website][simo1] that describes these calculations.
 
@@
 
+The general rule is that a nonlinear transformation leads to a similar change in the expected value, but a change in the variance that is related to the Taylor series expansion.
+
+If X has an expected value, $\mu$ and a variance, $\sigma^2_x$, then g(X) can be written as
+
+$g(X) \approx g(\mu)+(X-\mu)g'(\mu)$
+
+which can be used to show that
+
+$Var(g(X))=(g'(\mu))^2 \sigma^2_x$
+
 If
 
-X ~ Pois(T$\lambda$) and
+X ~ Pois($\lambda T$) and
 
-Y=ln(X), then
+$Y = g(X) = ln(X)$, then
 
-E[Y] $\approx$ ln($\lambda T$) and
+$g'(X)=\frac{1}{X}$, and
 
-V[Y] $\approx \frac{1}{\lambda T}$
+V[Y] $\approx (g'(\mu))^2 \sigma^2_x = (\frac{1}{\lambda T})^2 \lambda T = \frac{1}{\lambda T}$
 
-An approximate 95% confidence interval for ln($\lambda  T$) is
+An approximate $1-\alpha$ confidence interval for ln($\lambda  T$) is
 
-ln(X) $\pm 2 \sqrt{\frac{1}{X}}$
+$ln(X) \pm z_{\alpha/2} \sqrt{\frac{1}{X}}$
 
@@
 
+The mathematical approach used here is known as the delta method and is [documented in general terms] on Wikipedia.
+
 ### Examples
 
-http://www.pmean.com/07/ConfidenceIntervalForRate.html
+The Pennsylvania Department of Health has a [web page][penn1] oulining calculations for rates with three simple examples.
+The infant mortality rate in Butler County in 1984 was 17 deaths out of 1,989 births.
+This is actually a proportion, but let's treat it as a rate here.
 
-https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf
+```{r}
+x <- 17
+n <- 1989
+rate <- x/n
+rate
+```
+
+This rate is difficult to read because all of the leading zeros. So it is common practice to quote it not as a rate per live births, but a rate per 1,000 live births.
+
+```{r}
+rate_per_thousand <- rate*1000
+rate_per_thousand
+```
+
+To construct a confidence interval for the rate, first start with a confidence interval for the count rather than the rate.
+
+```{r}
+lo <- 
+hi <- x+1.96*sqrt(x)
+count_ci <- c(x-1.96*sqrt(x), x+1.96*sqrt(x))
+count_ci
+```
+
+Then convert from a count to a rate.
+
+```{r}
+rate_ci <- count_ci * 1000 / n
+rate_ci
+```
+
+It's always a good idea to round your results.
+
+```{r}
+round(rate_ci, 1)
+```
+
+Now let's try it using the improved formula. First build a confidence interval for the log count.
+
+```{r}
+log_count <- log(x)
+log_ci <- c(log_count-1.96*sqrt(1/x), log_count+1.96*sqrt(1/x))
+log_ci
+```
+
+Back transform to the original scale.
+
+```{r}
+back_transformed_ci <- exp(log_ci)
+back_transformed_ci
+```
+
+Convert from a count to a rate
+
+```{r}
+rate_ci <- back_transformed_ci * 1000 / n
+rate_ci
+```
+
+And again remember to round your final result.
+
+```{r}
+round(rate_ci, 1)
+```
+
+Notice that the rate (`r round(rate_per_thousand, 2)`) is not in the middle of the interval.
+It is slightly closer to the lower limit than the upper limit.
+This reflects the skewness in the Poisson distribution and provides a slightly more accurate interval.
+
+### On your own
+
+The reference also includes an infant mortality rate for the city of Philadelphia.
+There were 388 infant deaths out of 24,979 live births.
+Show that the 95% confidence interval for the infant mortality rate in Philadelphia is 14.0 to 17.1.
+
+Notice, by the way, how much narrower the confidence interval is when it is based on 388 deaths rather than 17.
+
+### References
+
+[simo1]: http://www.pmean.com/07/ConfidenceIntervalForRate.html
+
+[penn1]: https://www.health.pa.gov/topics/HealthStatistics/Statistical-Resources/UnderstandingHealthStats/Documents/Confidence_Intervals_for_a_Crude_Rate.pdf
+
+[delt1]: https://en.wikipedia.org/wiki/Delta_method
+
+[delt2]: https://en.wikipedia.org/wiki/Taylor_expansions_for_the_moments_of_functions_of_random_variables
+
+[pois1]: https://en.wikipedia.org/wiki/Poisson_distribution
+
+[pois2]: https://en.wikipedia.org/wiki/Binomial_distribution#Poisson_approximation
+
+http://www.pmean.com/definitions/or.htm
+
+http://www.pmean.com/01/oddsratio.html
+
+http://www.pmean.com/04/Rates.html
+
+https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510

### 2021-09-06 text/21/poisson-inferences.Rmd
Partial update of readme file

@@
 
-If X has an expected value, $\mu$ and a variance, $\sigma^2_x$, then g(X) can be written as
+If X has an expected value, $\mu$ and a variance, $\sigma^2_x$, then g(X) can be approximated, using a Taylor series expansion as
 
@@
 
-Notice that the rate (`r round(rate_per_thousand, 2)`) is not in the middle of the interval.
+Notice that the rate (`r round(rate_per_thousand, 1)`) is not in the middle of the interval.
 It is slightly closer to the lower limit than the upper limit.
@@
 
+### Comparison of two rates
+
+Suppose you have independent rates from two different groups,
+
+$R_1=\frac{X_1}{T_1}$
+
+$R_2=\frac{X_2}{T_2}$
+
+where the numerators of the rates are Poisson counts,
+
+$X_1$ ~ Pois($\lambda_1 T_1$) and
+
+$X_2$ ~ Pois($\lambda_2 T_2$).
+
+You can get a $1-\alpha$ confidence interval for
+
+$\frac{\lambda_1}{\lambda_2}$
+
+by first getting a confidence interval for
+
+$\frac{\lambda_1 T_1}{\lambda_2 T_2}$
+
+and then dividing the interval by
+
+$\frac{T_1}{T_2}$
+
+Start by noting that
+
+$ln\big(\frac{X_1}{X_2}\big)=ln(X_1)-ln(X_2)$
+
+As long as the two groups are independent, the variance of the difference is just the sum of the variances, so you can use
+
+$ln(X_1)-ln(X_2) \pm z_{\alpha/2}\sqrt{\frac{1}{X_1}+\frac{1}{X_2}}$
+
+as an approximate confidence interval for
+
+$\frac{\lambda_1 T_1}{\lambda_2 T_2}$
+
+Then exponentiate both sides of the interval and adjust by divinding by $T_1/T_2$.
+
+### Example
+
+In a [2018 article][kuma1] published in the Indian Journal of Dermatology, the researchers compared the incidence rates of leprosy in different age groups.
+In patients under the age of 14 there were 33 cases cumulated across 290508.9 patient years.
+In patients 15-29, there were 52 cases 141,112.0 patient years.
+Here are the calculations for the confidence interval for the rate ratio.
+
+First calculate the rates and their ratio. Since rates are so low, adjust them to a rate per 10,000 patient years.
+
+```{r}
+x1 <- 33
+T1 <- 290508.9
+x2 <- 52
+T2 <- 141112
+x2 <- 60
+T2 <- 100212.3
+
+x1 <- 2
+T1 <- 0.6*2*10000
+
+x2 <- 8
+T2 <- 5.47*8*10000
+
+r1 <- x1/T1*10000
+r2 <- x2/T2*10000
+r1
+r2
+r2/r1
+```
+
+Calculate the interval for $ln(\lambda_1 T_1/\lambda_2 T_2)$.
+
+```{r}
+ln_ratio <- log(x2/x1)
+se_ratio <- sqrt(1/x1+1/x2)
+se_ratio
+ci1 <- c(ln_ratio-1.96*se_ratio, ln_ratio+1.96*se_ratio)
+ci1
+```
+
+Exponentiate both limits.
+
+```{r}
+ci2 <- exp(ci1)
+ci2
+```
+
+Adjust for the patient years. Don't forget to round your final answer.
+
+```{r}
+ci3 <- ci2/(T2/T1)
+round(ci3, 2)
+```
+
 ### References
+
+http://www.pmean.com/definitions/or.htm
+
+http://www.pmean.com/01/oddsratio.html
+
+http://www.pmean.com/04/Rates.html
+
+https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510
 
@@
 
-http://www.pmean.com/definitions/or.htm
-
-http://www.pmean.com/01/oddsratio.html
-
-http://www.pmean.com/04/Rates.html
-
-https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510
+[kuma1]: https://ijdvl.com/incidence-of-leprosy-in-firozabad-district-uttar-pradesh/

### 2021-09-06 text/21/poisson-inferences.Rmd
More examples of IRR CIs in poisson-inference.Rmd

@@
 T2 <- 141112
+d<- 10000
 x2 <- 60
@@
 x1 <- 2
-T1 <- 0.6*2*10000
+T1 <- 0.6*2
+x2 <- 8
+T2 <- 5.47*8
+d <- 1
 
-x2 <- 8
-T2 <- 5.47*8*10000
+x1 <- 30
+T1 <- 54308.7
+x2 <- 60
+T2 <- 51477.5
+d <- 100000
 
-r1 <- x1/T1*10000
-r2 <- x2/T2*10000
+x1 <- 215
+T1 <- 4770
+x2 <- 84
+T2 <- 4193
+d <- 1000
+
+r1 <- x1/T1*d
+r2 <- x2/T2*d
 r1
 r2
-r2/r1
+r1/r2
 ```
@@
 ```{r}
-ln_ratio <- log(x2/x1)
+ln_ratio <- log(r1/r2)
 se_ratio <- sqrt(1/x1+1/x2)
@@
 ```{r}
-ci3 <- ci2/(T2/T1)
+ci3 <- ci2/(T1/T2)
 round(ci3, 2)
@@
 https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510
+
+Good example of rate ratio confidence interval calculations
+
+https://pubmed.ncbi.nlm.nih.gov/31895593/
+
+https://pubmed.ncbi.nlm.nih.gov/29637909/
+
+https://storage.googleapis.com/plos-corpus-prod/10.1371/journal.pmed.1001270/1/pmed.1001270.s005.pdf?X-Goog-Algorithm=GOOG4-RSA-SHA256&X-Goog-Credential=wombat-sa%40plos-prod.iam.gserviceaccount.com%2F20210906%2Fauto%2Fstorage%2Fgoog4_request&X-Goog-Date=20210906T225128Z&X-Goog-Expires=86400&X-Goog-SignedHeaders=host&X-Goog-Signature=ae5dde1295ee4d956450c3632eef6f9232594b386e2c4cb01f68bd8e516b25bb05beff6dd146fb84098b1186f3f19cb25b3c8e4c622a11cebfa75bf4a78150761b30984b7f816e1cf9566f7410eac7b9ee45b7a6d77e7b01517eeb36add101d012ba04a50d05d0f50d68097fa8a6b9882b43ab6a93013c066632be8b03b4efc5af597f7fd0ba423c6bd3ceb223847bdcc9f75f8b46cf67ba5ee27ea050855029f5e2c2339e867142d2d82871d9b80c1eb2e91bf06aacfb75ad6fa4f5726ab0ea53ca65e25074a33329eb45146eec575d22f4e324a90e126e1df7b45fc7b58f9246ae863f7f946be73fef76e869d9996ac3ac904d15a4c7bd34eb3777cc1f202a
+
+https://sphweb.bumc.bu.edu/otlt/MPH-Modules/PH717-QuantCore/PH717_ComparingFrequencies/PH717_ComparingFrequencies9.html
+
+https://www.jabfm.org/content/33/4/549.long
 

### 2021-09-08 text/21/poisson-inferences.Rmd
More work on poisson-inferences.md

@@
 ```{r setup}
+suppressMessages(suppressWarnings(library(broom)))
 suppressMessages(suppressWarnings(library(dplyr)))
@@
 
+The mathematical approach used here is known as the delta method and is [documented in general terms][delt1] on Wikipedia with more specifics [here][delt2].
+
 An approximate $1-\alpha$ confidence interval for ln($\lambda  T$) is
@@
 
+Because logarithms behave the way they do, we know that
+
+$ln(\lambda T)=ln(\lambda)-ln(T)$
+
+so you convert the above interval to a confidence interval for $ln(\lambda)$ by subtracting $ln(T)$. This produces the interval
+
+$ln(R) \pm z_{\alpha/2} \sqrt{\frac{1}{X}}$
+
 Exponentiate both sides to get a confidence interval for $\lambda T$.
 
-Divide by T to get a confidence interval for $\lambda$.
-
-The mathematical approach used here is known as the delta method and is [documented in general terms] on Wikipedia.
+The interesting thing about logarithms is that if you change the units of time (hundreds of patient years to thousands of patient years, for example), the width of the interval on the log scale is still the same.
 
@@
 n <- 1989
+
+x <- 299
+n <- 8963
+
+x <- 84
+n <- 4193
+
+x <- 215
+n <- 4770
+
 rate <- x/n
@@
 
-To construct a confidence interval for the rate, first start with a confidence interval for the count rather than the rate.
+To construct a confidence interval for the rate, first start with a confidence interval using the log rate.
 
 ```{r}
-lo <- 
-hi <- x+1.96*sqrt(x)
 count_ci <- c(x-1.96*sqrt(x), x+1.96*sqrt(x))
@@
 
-by first getting a confidence interval for
-
-$\frac{\lambda_1 T_1}{\lambda_2 T_2}$
-
-and then dividing the interval by
-
-$\frac{T_1}{T_2}$
+by using the same sort of calculations.
 
@@
 
-$ln\big(\frac{X_1}{X_2}\big)=ln(X_1)-ln(X_2)$
+$ln\big(\frac{R_1}{R_2}\big)=ln(R_1)-ln(R_2)$
 
@@
 
-$ln(X_1)-ln(X_2) \pm z_{\alpha/2}\sqrt{\frac{1}{X_1}+\frac{1}{X_2}}$
+$ln(R_1)-ln(R_2) \pm z_{\alpha/2}\sqrt{\frac{1}{X_1}+\frac{1}{X_2}}$
 
@@
 
-$\frac{\lambda_1 T_1}{\lambda_2 T_2}$
+$ln(\frac{\lambda_1}{\lambda_2})$
 
-Then exponentiate both sides of the interval and adjust by divinding by $T_1/T_2$.
+Then exponentiate both sides of the interval.
 
@@
 
-In a [2018 article][kuma1] published in the Indian Journal of Dermatology, the researchers compared the incidence rates of leprosy in different age groups.
-In patients under the age of 14 there were 33 cases cumulated across 290508.9 patient years.
-In patients 15-29, there were 52 cases 141,112.0 patient years.
-Here are the calculations for the confidence interval for the rate ratio.
+In a [2020 article][hess1] published in the Journal of Athletic Training, the researchers compared the incidence rates of injuries for professional ultimate frisbee players during games and during practice.
 
-First calculate the rates and their ratio. Since rates are so low, adjust them to a rate per 10,000 patient years.
+First calculate the rates and their ratio. Since rates are so low, adjust them to a rate per 1,000 patient years.
 
 ```{r}
-x1 <- 33
-T1 <- 290508.9
-x2 <- 52
-T2 <- 141112
-d<- 10000
-x2 <- 60
-T2 <- 100212.3
-
-x1 <- 2
-T1 <- 0.6*2
-x2 <- 8
-T2 <- 5.47*8
-d <- 1
-
-x1 <- 30
-T1 <- 54308.7
-x2 <- 60
-T2 <- 51477.5
-d <- 100000
-
 x1 <- 215
@@
 
-Calculate the interval for $ln(\lambda_1 T_1/\lambda_2 T_2)$.
+Calculate the interval on the log scale.
 
@@
 
-Adjust for the patient years. Don't forget to round your final answer.
+Don't forget to round your final answer.
 
 ```{r}
-ci3 <- ci2/(T1/T2)
-round(ci3, 2)
+round(ci2, 2)
+```
+
+This interval does not match the one published in the paper because the researchers adjusted for factors like field type.
+
+```{r}
+injuries <- data.frame(x=c(x1, x2), T=c(T1, T2), g=1:0)
+glm(x~g, family=poisson, offset=log(T), data=injuries) %>%
+  tidy -> c
+c
 ```
@@
 
+[hess1]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7017898/
+
 [simo1]: http://www.pmean.com/07/ConfidenceIntervalForRate.html

### 2021-09-11 text/21/poisson-regression.Rmd
Added poisson-regression.Rmd

---
title: "Working with Poisson regression"
author: "Steve Simon"
source: "New"
date: "2020-08-30"
categories:
- Blog post
tags:
- Probability distributions
output: html_document
---

Poisson regression is a statistical model under the large umbrella of generalized linear models. I want to explain how Poisson regression works. I also will show and interpret an example.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### The linear regression model

A linear regression model is written as

$Y_i=\beta_0+X_i\beta_1+\epsilon_i$

and has two assumptions that are often violated:

1. the $\epsilon_i$ are normally distributed, and 
2. the variance is the same for all $\epsilon_i$.

This led to a variety of alternative approaches that worked better, especially for dependent variables that represented proportions or counts.

### The generalized linear model

In 1982, McCullagh and Nelder wrote a [book][mcc1] outlining the generalized linear model, an approach to regression that would work well both for situations that met the assumptions of linear regression and for situations that did not meet those assumptions.

It was the sort of breakthrough that Physicists are still hoping for in the [grand unified theory][gut1] of the electromagnetic, strong, and weak forces under one model.

McCullagh and Nelder developed a model that works for any statistical distribution in a very broad class of distributions known as the exponential family. The exponential family of distributions includes both continuous and categorical distributions. The only requirement is that the probability density function (for continuous distributions) or the probability mass function (for categorical variables) can be factored in a a way that isolates the parameter(s) of the distribution in the following way:

$ln(f(y; \theta))=d(\theta)e(y)+g(\theta)+h(y)$

The Poisson distribution

$f(y; \theta)=\theta^{-y}e^{-\theta}/y!$ for y=0, 1, ...$

clearly fits into the exponential family, because

$lnf(y; \theta)=y ln(\theta)-\theta-ln(y!)$

Other distributions that fit into this family include the

+ bernoulli
+ beta
+ chi-squared
+ exponential
+ gamma
+ geometric
+ normal

There are some distributions which are not part of the exponential family. The t-distribution is one example.

If you know that a distribution is part of the exponential family, you know the general form of the maximum likelihood estimates and find a solution using iteratively reweighted least squares.

### More on Poisson regression

Counts are never negative, so we need to insure that no matter what the values of the covariates in a Poisson regression model, the predicted values produced by this model are never negative. A simple way to achieve this is to calculate the regression coefficients on a log scale.

The logarithm is an interesting function that converts multiplication into addition.

$log(ab)=log(a)+log(b)$

Back before computers, tables of logarithms simplified complex calculations. To compute a product of 4,156 and 2,023, you could multiply things out directly, but it would look something like

```{}
   2141
 x 2123
 ------
   6423
  4282
 2141
4282
-------
4545343
```

Instead, you could look up the base 10 logarithm of each number, add them together, and then convert back from the log scale to the original scale.

```{r}
log10(2141)
log10(2123)
log10(2141)+log10(2123)
10^(log10(2141)+log10(2123))
```

It only saves you a little bit of time, but back when mathematicians had to do hundreds or thousands of these calculations, it added up.

There's a joke along these lines.

After the flood waters receded, Noah's ark landed on solid ground. Noah opened the door and let all the animals out with the exhortation "Go forth and multiply." There were two snakes, however, that did not want to leave the ark. Noah looked at them and said, "I told you to go forth and multiply." The snakes looked sadly at him and said "We can't multiply. We're adders." So Noah left the ark and found some wood to build a small flat platform. "Here." he said. "This is a log table. It will allow adders to multiply."

The Poisson regression model produces estimates on a log scale. This implies that the relationships are effectively multiplicative when transformed back to the original scale.

### Simple example

Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models. The first data set is available for anyone to download.

```{r}
fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
articles <- read.csv(fn)
head(articles)
```

The variables are
+ gender (fem),
+ marital status (mar),
+ number of children (kid5),
+ prestige of graduate program (phd), 
+ the number of articles published by the individual’s mentor (ment)
+ the number of articles published by the scientist (art: the outcome).

To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".

```{r}
pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
summary(pois1)
```

The interpretation of the coefficients is tricky. The Poisson regression model is linear on the log scale. If you take the anti-log, you will convert this model to a multiplicative model. Here are the coefficents on the log scale.

```{r}
pois1 %>%
  tidy %>%
  mutate(lower_limit=estimate-1.96*std.error) %>%
  mutate(upper_limit=estimate+1.96*std.error) %>%
  select(term, estimate, lower_limit, upper_limit) -> log_coefficients
log_coefficients
```

```{r}
log_coefficients %>%
  mutate(estimate=exp(estimate)) %>%
  mutate(lower_limit=exp(lower_limit)) %>%
  mutate(upper_limit=exp(upper_limit)) -> multiplicative_coefficients
multiplicative_coefficients
```

```

### Analysis of rates

Often the Poisson observations are observed over varying numbers of patients, varying amounts of time, or both. Here's an example. Auto insurance claims in Sweden are borken down into several subgroups, as described on [this web page][swe1].

Here are the variables:

+ Kilometres		Kilometres travelled per year
  + 1: < 1000
  + 2: 1000-15000
  + 3: 15000-20000
  + 4: 20000-25000
  + 5: > 25000
+ Zone		Geographical zone
  + 1: Stockholm, Göteborg, Malmö with surroundings
  + 2: Other large cities with surroundings
  + 3: Smaller cities with surroundings in southern Sweden
  + 4: Rural areas in southern Sweden
  + 5: Smaller cities with surroundings in northern Sweden
  + 6: Rural areas in northern Sweden
  + 7: Gotland
+ Bonus		No claims bonus. Equal to the number of years, plus one, since last claim
+ Make		1-8 represent eight different common car models. All other models are combined in class 9
+ Insured		Number of insured in policy-years
+ Claims		Number of claims
+ Payment		Total value of payments in Skr

```{r}
fn <- "http://www.statsci.org/data/general/motorins.txt"
swe <- read.delim(fn, header=TRUE, sep="\t")
head(swe)
```

The number of drives and the months that they are covered by insurance varies from group to group. the measure, Insured, calculates the number of policy-years in each group. We want to estimate the effect of certain variables on the rate of claims (number of claims per policy year). You calculate the rate by including an offset variable equal to the natural log of policy years.

```{r}
pois2 <- 
  glm(
    Claims~offset(log(Insured)) + factor(Kilometres) + factor(Zone) + Bonus + factor(Make),
    data=swe, family=poisson)
summary(pois2)
```

The coefficients are easier to interpret after a transformation.

```{r}
pois2 %>%
  tidy %>%
  mutate(lower_limit=estimate-1.96*std.error) %>%
  mutate(upper_limit=estimate+1.96*std.error) %>%
  select(term, estimate, lower_limit, upper_limit) %>%
  mutate(estimate=exp(estimate)) %>%
  mutate(lower_limit=exp(lower_limit)) %>%
  mutate(upper_limit=exp(upper_limit)) -> multiplicative_coefficients
multiplicative_coefficients
```

### Extensions

McCullagh and Nelder also extended their results to an even broader setting using an approach known as quasi likelihood. In a quasi likelihood model, you do not specify a distribution, but rather a link function that converts the nonlinear regression relationship to a linear one, and a function that describes the mean-variance relationship. The latter is very important. In many statistical models, there is a specific type of heterogeneity where the mean and variance march together in lock step. Groups that have larger means also have larger variances. Groups that have smaller means also have smaller variances.

In the Poisson model, we specify a mean-variance relationship that hte identify relationship. If a group has a mean that is 12, the variance is assumed to be 12 as well. If the mean is 5, then the variance is 5. That works fairly well for some data, but other times it does not work so well.

There are many alternatives. You can assume that the variance is not equal to the mean but equal to the mean times a mutliplicative constant which you can estimate from the data.

Notice, for example, in the first data set that

```{r}
mean(articles$art)
var(articles$art)
```

The variance looks to be about twice as large as the mean. Additional evidence of this is provided by the deviance statistic. The deviance is a quantity produced by a generalized linear model and should equal the degrees of freedom in a Poisson regression model.

```{r}
glance(pois1)
```

You will notice here that the deviance is about twice as big as the degrees of freedom.

You can fix this problem in seveal ways. First, you can specify a quasi-likelihood that behaves like the Poisson except that the function specifying the mean-variance relationship has an extra parameter, estimated from the data.

You can also use a counting distribution, the negative binomial distribution, that allows for greater flexibility in the mean-variance relationship.

Finally, you can adjust the zero probability upward from what a Poisson model would expect. This is commonly thought of as a mixture distribution where there is a certain probability p, of getting exactly zero, and a probability 1-p or getting a Poisson random variable (which can sometimes equal zero, but only sometimes). These models are known as zero-inflated models.

The [web page cited earlier][hua1] has more details about these models.

### References

[gut1]: https://en.wikipedia.org/wiki/Grand_Unified_Theory

[hua1]: https://francish.netlify.app/post/poisson-and-negative-binomial-regression-using-r/

[mcc1]: https://www.routledge.com/Generalized-Linear-Models/McCullagh-Nelder/p/book/9780412317606

[swe1]: http://www.statsci.org/data/general/motorins.html

### 2021-09-20 text/21/direct-standardization.Rmd
Added stub for quasi-likelihood

---
title: "Direct standardization of mortality rates"
author: "Steve Simon"
source: "New"
date: "2020-09-20"
categories:
- Blog post
tags:
- Epidemiology
output: html_document
---

Mortality rates in different regions of the United States between different countries are dependent on the demographic distributions. Differences in the age distribution can cause mortality rates that are actually equivalent to appear as if they differ. These differences could just as easily hide a difference in mortality rates between two regions or even make the difference go in the opposite direction. Direct standardization is a simple approach to adjust mortality rates

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### References

Omar B. Ahmad, Cynthia Boschi-Pinto, Alan D. Lopez, Christopher JL Murray, Rafael Lozano, Mie Inoue. Age standardization of rates: A new WHO standard. GPE Discussion Paper Series: No.31, World Health Organization 2001. Available in [pdf format][who1].

[who1]: https://www.who.int/healthinfo/paper31.pdf?ua=1

### 2021-09-20 text/21/quasi-likelihood.Rmd
Added stub for quasi-likelihood

---
title: "Direct standardization of mortality rates"
author: "Steve Simon"
source: "New"
date: "2020-09-20"
categories:
- Blog post
tags:
- Gneralized linear model
output: html_document
---

With the generalized linear model, you have the option of choosing a model that works well with a variety of different distributions (normal, Poisson, Gamma, etc.). But you also have the option of working outside of these distributions. You can fit models that do not specify the distribtution, but which fit a particular transformation of the linear effect and particular mean-variance relationship.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### References

David Lillis. Generalized Linear Models in R, Part 7: Checking for Overdispersion in Count Regression. The Analysis Factor blog. Available in [html format][taf1]

[taf1]: https://www.theanalysisfactor.com/glm-r-overdispersion-count-regression/

### 2021-09-21 text/05/MergingInR.Rmd
Removed bad characters

---
title: Merging in R
author: Steve Simon
date: 2005-07-26
categories:
- Blog post
tags:
- R software
output: html_document
---
Merging is tricky in any program, and you have to be sure that you know
what you are doing. I have found that keeping an index variable will
allow me to backtrack and find information in the original files if
needed. Here\'s a simple example of how this would work.

```{r}
# randomly select 20 letters without replacement.
x0 <- sample(letters,20)
x1 <- sample(letters,20)
x0
x1
```

The intersect functions lets us know which letters are in both x0 and x1.

```{r}
intersect(x0,x1)
```

The setdiff function lets us know which letters are in x0 but not x1 (reverse the arguments to get letters in x1 but not in x1).

```{r}
setdiff(x0,x1)
setdiff(x1,x0)
```

By default, the merge function includes only those values in common with the two files. Note that the id1 and id2 values tells us that the first row of x2 comes from the 19th and 8th rows of x0 and x1, respectively. Notice also that the merged file is sorted by the intersection of the two files.

```{r}
x2 <- merge(list(x=x0,id0=1:20),list(x=x1,id1=1:20))
x2
```

To include all values, including mismatches, specify all=T. Notice now that the merged file is sorted by the values of x0, followed by the sorted values of x1 that are not found in x0. An NA for an index value tells you that the letter is not found in the original x0 or x1 vector.

```{r}
x3 <- merge(list(x=x0,id0=1:20),list(x=x1,id1=1:20),all=T)
x3
```

The %in% operator is also useful. The %in% operator compares each element of the first set against all the elements of the second set and returns TRUE if there is a match.

```{r}
x0 %in% x1
```

This function is not symmetric, of course,

```{r}
x1 %in% x0
```

Using the %in% function, you can re-create the intersect function,

```{r}
x0[x0 %in% x1]
```

or the setdiff function,
```{r}
x0[!x0 %in% x1]
```

A less intuitive version of the %in% operator is the match function. Additional information on merging can be found at

-   [www.ats.ucla.edu/stat/r/notes/managing.htm](http://www.ats.ucla.edu/stat/r/notes/managing.htm)

You can find an [earlier version](http://www.pmean.com/05/MergingInR.html) of this page on my [original website](http://www.pmean.com/original_site.html).

### 2021-09-21 text/21/direct-standardization.Rmd
Removed bad characters

@@
 source: "New"
-date: "2020-09-20"
+date: "2021-09-20"
 categories:

### 2021-09-21 text/21/poisson-regression.Rmd
Removed bad characters

@@
 source: "New"
-date: "2020-08-30"
+date: "2021-08-30"
 categories:

### 2021-09-23 src/clean_files.R
Changed Recommended to Recommendation

@@ -18,7 +18,7 @@ clean_files <- function(
     search_string, 
     replace_string="Not yet", 
     dir_root="text/zotero", 
-    subdir_list=c("99", zpad(0:20), "no"),
+    subdir_list=c("99", zpad(0:21), "no"),
     file_pattern="*.bib", 
     original_string=search_string,
     v=TRUE) {
@@ -58,8 +58,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'title: "Pmean: ',
-  'title: "',
+  '- Quote',
+  '- Quotation',
   dir_root="text", 
   file_pattern="md$")
 

### 2021-09-24 text/21/poisson-regression.Rmd
Updated anova.md

@@
 + prestige of graduate program (phd), 
-+ the number of articles published by the individual’s mentor (ment)
++ the number of articles published by the individual's mentor (ment)
 + the number of articles published by the scientist (art: the outcome).

### 2021-09-24 src/clean_files.R

@@ -58,8 +58,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '- Quote',
-  '- Quotation',
+  "</U",
+  " ",
   dir_root="text", 
   file_pattern="md$")
-

### 2021-09-28 text/21/negative-binomial-regression.Rmd
Added negative-binomial-regression.md

---
title: "Working with negative binomial regression"
author: "Steve Simon"
source: "New"
date: "2021-08-30"
categories:
- Blog post
tags:
- Probability distributions
- Generalized linear models
output: html_document
---

The negative binomial distribution is similar to the Poisson distribution, but it offers a bit more flexibility. This page will review some basic properties of the negative binomial distribution and show how it can be incorporated into the framework of the generalized linear model.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### Definition of the negative binomial distribution

A random variable X, defined on the set of non-negative integers (0, 1, 2, ...), is said to have a negative binomial distribution with parameters n and p (X ~ NB(n, p)) if

$P[X=k]=\binom{k+n-1}{k}(1-p)^np^k$ for k=0, 1, 2, ...

where $\binom{\ \ }{\ \ }$ is the binomial coefficient, defined as

$\binom{a}{b}=\frac{a!}{b!(b-a)!}$

Notice the similarity to the binomial probability

$P[X=k]=\binom{n}{k}(1-p)^{n-k}p^k$ for k=0, 1, 2, ..., n

The binomial distribution measures the number of successes k in n independent trials, where p is the probability of success. The negative binomial distribution measures the number of trials needed until k successes are observed.

The distinction is subtle, but important. There was a story on NPR a long time ago about street musicians and what songs they play. One violinist had tremendous success playing "Happy Birthday" as people walked by. Most people would ignore this, but the few who happened to have a birthday on that day would invariably smile and pitch money into the musician's hat. What's even better is that most people are in a good mood on their birthday and the musician would often see tens and twenties. The people who did not have a birthday would maybe just glare at him.

If 10,000 people walked by, a binomial distribution would tell you how many gave money and how many glared. If the musician quite after getting $1,000 (assuming each birthday person contributes $10), then the negative binomial would represent the number of glares that the musician would get before accumulating a hundred Hamiltons.

The only mathematical differences between the two distributions are

+ $\binom{k+n-1}{k}$ versus $\binom{n}{k}$
+ $(1-p)^np^k$ versus $(1-p)^{n-k}p^k$
+ k=0, 1, 2, ... versus k=0, 1, 2, ..., n

This last distinction is important. There is a firm upper bound for the binomial distribution. It is impossible to get more than n successes out of n trials. But there is no upper bound for negative binomial distribution. If you are waiting for k successes, a bad run of luck could cause this wait to become interminably long.

### Mean and variance of a negative binomial distribution.

$E[X]=\frac{np}{1-p}$

$Var[X]= \frac{np}{(1-p)^2}$

With a bit of math you can show that

$Var[X]=E[X] \left(1+\frac{E[X]}{n} \right)$

Since E[X] is always positive, the variance of a negative binomial random variable is always larger than the mean.

If

$X_1 \sim NB(n_1, p)$ and $X_2 \sim NB(n_2, p)$ then

$X_1+X_2 \sim NB(n_1+n_2, p)$

### Visualization of negative binomial probabilities

Here are the probabilities for various negative binomial cases where n changes, and p is held constant at 0.8.

```{r bar-charts-n}
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x,  2, 0.8)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=2, p=0.8")
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x, 10, 0.8)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=10, p=0.8")
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x, 50, 0.8)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=50, p=0.8")
```

Here are the probabilities for various negative binomial cases where n changes, and p is held constant at 0.8.

```{r bar-charts-p}
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x,  2, 0.8)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=2, p=0.8")
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x, 2, 0.5)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=2, p=0.5")
data.frame(x=0:25) %>%
  mutate(p=dnbinom(x, 2, 0.2)) %>%
  ggplot(aes(x,p)) +
    geom_col() +
    expand_limits(y=1) +
    ggtitle("Poisson probabilities with n=2, p=0.2")
```

### Is the negative binomial distribution part of the exponential family?

Yes and no. Remember the definition

$ln(f(y; \theta))=d(\theta)e(y)+g(\theta)+h(y)$

Where $\theta$ is the unknow parameter. For the negative binomial distribution,

$ln(f(y; n, p))=ln((y+n-1)!) - ln((n-1)!) -ln(y!) + n ln(1-p) + y ln(p)$

The very first term is a function of y and n that cannot be factored into function of only y times a function of only n. So if n is a known value, the negative binomial is part of the exponential family. In most practical settings, however, n is an unknown parameter along with p.

This means that you can't use the glm function in R to fit a negative binomial regression. The methods of the generalized linear model have been adapted to the negative binomial case, and it is fit in R using the glm.nb function.

### Comparing a Poisson regression to a negative binomial regression.

Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model, the negative binomial model and several others. The first data set is available for anyone to download.

```{r}
fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
articles <- read.csv(fn)
head(articles)
```

The variables are
+ gender (fem),
+ marital status (mar),
+ number of children (kid5),
+ prestige of graduate program (phd), 
+ the number of articles published by the individual's mentor (ment)
+ the number of articles published by the scientist (art: the outcome).

To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".

```{r}
pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
summary(pois1)
```

```{r}
library(MASS)
nb1 <- glm.nb(art ~ fem + ment + phd + mar + kid5, data = articles)
summary(nb1)
```

### References

https://en.wikipedia.org/wiki/Negative_binomial_distribution
https://www.sagepub.com/sites/default/files/upm-binaries/21121_Chapter_15.pdf
https://data.library.virginia.edu/getting-started-with-negative-binomial-regression-modeling/
https://stats.idre.ucla.edu/r/dae/negative-binomial-regression/
http://www2.stat.duke.edu/~cr173/Sta444_Sp17/slides/Lec3.pdf
https://www.jstor.org/stable/2289071
https://towardsdatascience.com/diagnose-the-generalized-linear-models-66ad01128261
https://cran.r-project.org/web/packages/bbmle/vignettes/quasi.pdf

[hua1]: https://francish.netlify.app/post/poisson-and-negative-binomial-regression-using-r/

### 2021-09-28 text/21/poisson-regression.Rmd
Added negative-binomial-regression.md

@@
 
-```
-
 ### Analysis of rates
@@
 + Zone		Geographical zone
-  + 1: Stockholm, Göteborg, Malmö with surroundings
+  + 1: Stockholm, G?teborg, Malm? with surroundings
   + 2: Other large cities with surroundings

### 2021-09-28 text/20/smart-quotes.Rmd
Dropped Reccomended and Blog post from titles.

--- 2020-03-02 text/20/smart-quotes.Rmd
@@
 ---
-title: "Blog post: Smart quotes, em dashes, and en dashes"
+title: "Smart quotes, em dashes, and en dashes"
 author: "Steve Simon"

### 2021-09-30 src/clean_files.R
Updated clean_files.R

@@ -58,7 +58,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "</U",
-  " ",
+  "Blog post: ",
+  "",
   dir_root="text", 
   file_pattern="md$")

### 2021-09-30 text/Rmd/8-tips/index.Rmd
Moved files out of monthly folders

---
title: "Recommendation: 8 tips for doing data visualization right."
author: "Steve Simon"
source: "http://blog.pmean.com/8-tips/"
date: "2014-05-03"
category: Recommendation
tags: Graphical display
output: html_document
---

This slide show includes some examples of really bad (and a few really
good) graphics with some explanations of general principles for data
display.

<!---More--->

AllAnalytics -- Beth Schultz -- 8 Tips for Doing Data Visualization
Right. Available at:
<http://www.allanalytics.com/author.asp?_mc=sem%5Fotb%5Fedt%5Fallanppcm&section_id=1411&doc_id=261787&image_number=>1.

```{r}
library(knitr)
include_graphics("../../web/images/8-tips01.png")
```

### 2021-09-30 text/Rmd/Declining-count-08/index.Rmd
Moved files out of monthly folders

---
title: Modeling a declining count variable
author: Steve Simon
date: 2008-06-14
categories:
- Blog post
tags:
- Poisson regression
output: html_document
---
I\'ve been working on an interesting project that requires Poisson
regression. A company sends out a mailing and gets a certain number of
telephone calls back on each of the days following. The number of phone
calls is typically (but not always) highest on the first day afterwards
and declines rapidly on successive days. I wanted to develop a simple
Poisson regression model for this data.

Here\'s an example of this data

`1 9   2 4   3 2   4 3   5 0   6 0   7 1`

You can fit a simple Poisson regression model in R using the following
code:

Using 0 as the starting point rather than 1 makes the interpretation of
this model simpler. When you list the contents of pmod, you get the
following

```{r}
ct <- c(9,4,2,3,0,0,1)
tm <- 0:6
pmod <- glm(ct~tm,family=poisson)
```

How would you interpret these coefficients? The Poisson regression
model uses a log link function. This means that the average value for
the data is assumed to be linear on a log scale.

In a linear regression model, the intercept is the estimated average
value of the dependent variable when the independent variable is equal
to zero. In a Poisson regression model, the antilog of the intercept
represents the estimated average count when the independent variable is
zero. Since exp(2.1) = 8.2, you would say that the estimated average
count at time zero is 8.2.

You might be tempted to make the intercept equal to 9. While this is not
an outrageous thing to do, it is not quite as efficient. The Poisson
regression model has to fit all of the counts well, not just the count
at 0.

In a linear regression model, the slope represents the estimated average
change in the dependent variable when the independent variable increases
by one unit. In a Poisson regression model, the antilog of the slope
represents the estimated average change in the count when the
independent variable increases by one unit. An important difference,
though, is that the change is a multiplicative change. Since exp(-0.55)
= 0.58, you would say that the estimated average count declines by a
factor of 0.58 (a 42% decline) when time increases by one day.

A zero slope in a Poisson regression model corresponds to a
multiplicative change of 1 which, of course, is no change. A positive
slope in a Poisson regression model corresponds to a multiplicative
change greater than 1. You could call this an exponential rise. A
negative slope in a Poisson regression model corresponds to a
multiplicative change less than 1. This is an exponential decline.

The predict function in R produces estimated average values on a log
scale for each time point.

```{r}
round(predict(pmod),4)
```

Note that R places an index of 1 through 7 above each predicted value.
This is a bit confusing, perhaps, because the first value corresponds to
time=0 and the last value corresponds to time=6.

If you take the antilog of these values, you get the estimated average
count at each time point.

```{r}
round(exp(predict(pmod)),4)
```

To demonstrate that this is an exponential decline, note that

`8.2*0.58^0 = 8.2   8.2*0.58^1 = 4.8   8.2*0.58^2 = 2.8   8.2*0.58^3 = 1.6   8.2*0.58^4 = 0.93   8.2*0.58^5 = 0.54   8.2*0.58^6 = 0.31`

There is a bit of rounding here, but the series matches the values shown
using the antilog of the predicted values.

You might wonder if you are likely to see any more values beyond time=6.
The estimated average counts for times 7, 8, etc. are

`8.2*0.58^7 = 0.18   8.2*0.58^8 = 0.11 ...`

You can sum the infinite series from time=0 to time=infinity if you
like. Notice that the terms in the above expression can be represented
by a classic [geometric
series](http://en.wikipedia.org/wiki/Geometric_series)

`a   ar   ar2   ar3   ?.   ?.   ?.`

For r\<1, the sum of this infinite series is

`a/(1-r)`

For a=8.2 and r=0.58, this equals 19.5. Since we\'ve already seen 19
events, we would expect to see half an event on average for the
remainder the time. You can use the dpois function to estimate
probabilities for a Poisson random variable with a mean of 0.5.

```{r}
round(dpois(0:4,0.5),2)
```

You would estimate that it is more than likely (61% probability) that
you will not see any more events. There is still a reasonable chance
(30%) that you will see one more event. It is unlikely (9% probability)
that you will see 2 or more events.

You can find an [earlier version](http://www.pmean.com/08/DecliningCount.html) of this page on my [original website](http://www.pmean.com/original_site.html).

### 2021-09-30 text/Rmd/glm-deviance/index.Rmd
Moved files out of monthly folders

---
title: "Deviance in a generalized linear model"
author: "Steve Simon"
source: "New"
date: "2020-09-20"
categories:
- Blog post
tags:
- Generalized linear model
output: html_document
---

Deviance is a measure computed for generalized linear models that can help you decide between two competing models. It is somewhat analagous to residual sums of squares in linear regression.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

<!---More--->

### References

Eduardo Garcia Portugues. Section 5.5. Deviance, in Notes for Predictive Modeling, 2021-08-18. Available in [html format][port1]

Freddy Hernandez. Deviance in glm. RPubs, 2020-07-07. Available in [html format][hern1]

German Rodriguez. Generalized Linear Model Theory. Lecture Notes on Generalized Linear Models. Available in [pdf format][rodr1]

Kenneth Tay. What is deviance? Statistical Odds and Ends blog, 2019-03-27. Available in [html format][tay1]

[hern1]: https://rpubs.com/fhernanb/deviance_glm
[port1]: https://bookdown.org/egarpor/PM-UC3M/glm-deviance.html
[rodr1]: https://data.princeton.edu/wws509/notes/a2.pdf
[tay1]: https://statisticaloddsandends.wordpress.com/2019/03/27/what-is-deviance/

### 2021-09-30 text/Rmd/map-projections/index.Rmd
Moved files out of monthly folders

--- 2020-03-23 text/20/map-projections.Rmd
@@
 ---
-title: "Blog post: Different map projections"
+title: "Different map projections"
 author: "Steve Simon"

### 2021-09-30 text/Rmd/misuse-of-gradient/index.Rmd
Moved files out of monthly folders

--- 2020-03-11 text/20/misuse-of-gradient.Rmd
@@
 ---
-title: "Blog post: An example of a poor color choice"
+title: "An example of a poor color choice"
 author: "Steve Simon"

### 2021-09-30 text/Rmd/quasi-likelihood/index.Rmd
Moved files out of monthly folders

--- 2021-09-20 text/21/quasi-likelihood.Rmd
@@
 tags:
-- Gneralized linear model
+- Generalized linear model
 output: html_document

### 2021-09-30 text/Rmd/reading-text-files/index.Rmd
Moved files out of monthly folders

--- 2020-07-22 text/20/reading-text-files.Rmd
@@
 source: "New"
-date:2020-04-30"
-categories: Blog post
-tags: Data management
+date: "2020-04-30"
+categories:
+- Blog post
+tags:
+- Data management
 output: html_document

### 2021-09-30 text/Rmd/map-projections/index.Rmd
Cleaned up some Rmd files

@@
 ---
+install.packages("rnaturalearthhires", repos="http://packages.ropensci.org", type="source")
 

### 2021-09-30 text/Rmd/quasi-likelihood/index.Rmd
Cleaned up some Rmd files

--- 2021-09-20 text/21/quasi-likelihood.Rmd
@@
 ---
-title: "Direct standardization of mortality rates"
+title: "Quasi-likelihood models"
 author: "Steve Simon"
 source: "New"
-date: "2020-09-20"
+date: "2021-09-20"
 categories:

### 2021-10-02 text/Rmd/8-tips/index.Rmd
Cleaned up some Rmd files
Fixed 1999 dates, part 1

@@
 ---
-title: "Recommendation: 8 tips for doing data visualization right."
+title: "8 tips for doing data visualization right."
 author: "Steve Simon"
@@
 library(knitr)
-include_graphics("../../web/images/8-tips01.png")
+# include_graphics("../../web/images/8-tips01.png")
 ```

### 2021-10-02 src/clean_files.R

@@ -58,7 +58,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "Blog post: ",
+  "1999-01-01",
   "",
   dir_root="text", 
   file_pattern="md$")

### 2021-10-02 src/copy_files.R

@@ -15,7 +15,7 @@ source(file="src/standard_functions.R")
 v <- TRUE
 
 copy_files <- function(
-  old_root="text", new_root="../blogdown/content",
+  old_root="text", new_root="c:/Users/simons/Documents/b04/content",
   subdir_list=c("99", zpad(0:21), "no"),
   file_pattern="md$", v=TRUE) {
   k <- 0

### 2021-10-02 src/clean_files.R
Fix ...

@@ -58,7 +58,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "1999-01-01",
-  "",
+  "\\) of this page on my \\[original website\\]\\(http://www.pmean.com/original_site.html\\)\\.",
+  "\n[sim2]: http://www.pmean.com/original_site.html",
   dir_root="text", 
   file_pattern="md$")

### 2021-10-02 text/05/MergingInR.Rmd
Redefined footer

@@
 
-You can find an [earlier version](http://www.pmean.com/05/MergingInR.html) of this page on my [original website](http://www.pmean.com/original_site.html).
+You can find an [earlier version][sim1] of this page on my [original website][sim2].
+
+[sim1]: http://www.pmean.com/05/MergingInR.html
+[sim2]: http://www.pmean.com/original_site.html

### 2021-10-02 src/clean_files.R

@@ -17,9 +17,9 @@ v <- TRUE
 clean_files <- function(
     search_string, 
     replace_string="Not yet", 
-    dir_root="text/zotero", 
+    dir_root="text", 
     subdir_list=c("99", zpad(0:21), "no"),
-    file_pattern="*.bib", 
+    file_pattern="*\\.md$", 
     original_string=search_string,
     v=TRUE) {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
@@ -58,7 +58,19 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "\\) of this page on my \\[original website\\]\\(http://www.pmean.com/original_site.html\\)\\.",
-  "\n[sim2]: http://www.pmean.com/original_site.html",
+  "of this  of this page",
+  "of this page",
   dir_root="text", 
-  file_pattern="md$")
+  file_pattern="md$",
+  subdir_list=zpad(c(99, 0:21)))
+
+ok_to_replace <- TRUE
+ok_to_replace <- FALSE
+clean_files(
+  "You can find an \\[earlier version\\]\\(",
+  "You can find an [earlier version][sim1] of this page on my [original website][sim2].\n\n[sim1]: ",
+  dir_root="text", 
+  file_pattern="md$",
+  subdir_list=zpad(c(16:21)))
+
+"of this  of this page"

### 2021-10-02 src/clean_files.R
Fixed elipsis in titles

@@ -58,19 +58,9 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "of this  of this page",
-  "of this page",
+  '\\.\\.\\.',
+  '',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))
 
-ok_to_replace <- TRUE
-ok_to_replace <- FALSE
-clean_files(
-  "You can find an \\[earlier version\\]\\(",
-  "You can find an [earlier version][sim1] of this page on my [original website][sim2].\n\n[sim1]: ",
-  dir_root="text", 
-  file_pattern="md$",
-  subdir_list=zpad(c(16:21)))
-
-"of this  of this page"

### 2021-10-07 text/Rmd/quasi-likelihood/index.Rmd
Updated quasi-likelihood page

@@
 
+### Link and variance functions
+
+The generalized linear model for any distribution in the exponential family can be described in terms of a link function and a variance function.
+
+For Poisson regression, the link function is
+
+$g(\mu)=ln(\mu)$
+
+and the variance function is
+
+$V(\mu)=\mu$
+
+These functions are derived from the form of the Poisson distribution, and the estimation methods in the generalized linear model depend only on the link and variance functions.
+
+In 1974, Robert Weddeburn developed a theory that would work for any link and variance function, even if it could not be derived from a particular probability distribution.
+His approach used the same principles as maximum likelihood, but it could not be called a likelihood approach since there was no probability distribution.
+He called his approach quasi-likelihood and showed that it created estimates with reasonable properties.
+
+### The quasi Poisson model
+
+If you modify the variance function of Poisson regression, from
+
+$V(\mu)=\mu$
+
+to
+
+$V(\mu)=\phi \mu$
+
+and keep the link function the same, then you get a quasi Poisson model.
+There is no simple probability distribution that would produce this link and variance function, but it seems like a more reasonable approach, because there is more flexibility.
+
+For technical reasons, the parameter $\phi$ has to be greater than one, implying that the dispersion of the quasiPoisson model is greater than the dispersion of the Poissson model.
+The quasiPoisson model is often called an overdispersion model.
+
+There are two ways to estimate $\phi$. The first is very simple and direct.
+
+$\hat{\phi}= dispersion / df$
+
+The amount of overdispersion is equal to much much larger the dispersion is than the degrees of freedom.
+
+A second approach estimates $\phi$ iteratively along with the other terms in the regression model.
+
+One important limitation of the quasi likelihood approach is that there is no natural way to estimate AIC (Akaike's Information Criteria) an important statistic for comparing different models.
+
 <!---More--->
+
+### Start with Poisson regression
+
+It is easiest to understand the quasiPoisson model by comparing it to the Poisson model.
+
+Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models.
+The first data set is available for anyone to download.
+
+```{r}
+fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
+articles <- read.csv(fn)
+head(articles)
+```
+
+The variables are
++ gender (fem),
++ marital status (mar),
++ number of children (kid5),
++ prestige of graduate program (phd), 
++ the number of articles published by the individual's mentor (ment)
++ the number of articles published by the scientist (art: the outcome).
+
+To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".
+
+```{r}
+pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
+summary(pois1)
+```
+
+The interpretation of the coefficients should be done after exponentiating their values.
+
+```{r}
+pois1 %>%
+  tidy %>%
+  mutate(lower_limit=estimate-1.96*std.error) %>%
+  mutate(upper_limit=estimate+1.96*std.error) %>%
+  select(term, estimate, lower_limit, upper_limit) %>%
+  mutate(estimate=exp(estimate)) %>%
+  mutate(lower_limit=exp(lower_limit)) %>%
+  mutate(upper_limit=exp(upper_limit)) -> pois1_coefficients
+pois1_coefficients
+```
+
+### Fitting a quasi-Poisson model
+
+```{r}
+quasi1 <- glm(art ~ fem + ment + phd + mar + kid5, family=quasipoisson, data = articles)
+summary(quasi1)
+```
+
+Notice that there is no value listed for AIC.
+
+```{r}
+quasi1 %>%
+  tidy %>%
+  mutate(lower_limit=estimate-1.96*std.error) %>%
+  mutate(upper_limit=estimate+1.96*std.error) %>%
+  select(term, estimate, lower_limit, upper_limit) %>%
+  mutate(estimate=exp(estimate)) %>%
+  mutate(lower_limit=exp(lower_limit)) %>%
+  mutate(upper_limit=exp(upper_limit)) -> quasi1_coefficients
+quasi1_coefficients
+```
+
+If you compare these confidence intervals to the earlier ones, they are wider. This reflects the overdispersion that is only accounted for properly by the quasi Poisson model.
 

### 2021-10-07 src/clean_files.R
Added ebm-progress.md

@@ -35,16 +35,18 @@ clean_files <- function(
       br -> message
     if (verbose) cat(message)
     for (i_file in file_list) {
-      tx <- read_lines(dir_root %s% subdir %s% i_file)
+      dir_root %s% subdir %s% i_file -> fn
+      tx <- read_lines(fn)
       found_lines <- str_which(tx, search_string)
       if (length(found_lines)==0) next
       k <- k+1
-      "\n\n" %0% subdir %s% i_file %>% br %>% cat
+      "\n\n" %0% fn %>% br %>% cat
       str_c(tx[found_lines], collapse="\n") %>% cat
       if (replace_string!="Not yet") {
         tx %<>% str_replace_all(original_string, replace_string)
         "\n" %0% str_c(tx[found_lines], collapse="\n") %>% cat
-      if (ok_to_replace) writeLines(tx, dir_root %s% subdir %s% i_file)
+      if (ok_to_replace) {
+        writeLines(tx, con=fn)}
       }
     }
   }
@@ -58,8 +60,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '\\.\\.\\.',
-  '',
+  '\\- Quotations',
+  '- Quotation',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))

### 2021-10-09 text/Rmd/first-steps-with-blogdown/index.Rmd
Started work on first-steps-with-blogdown.Rmd

---
title: "First steps with blogdown"
author: "Steve Simon"
source: "New"
date: "2021-10-09"
categories:
- Blog post
tags:
- Website details
output: html_document
---

I am in the process of merging my website, written from 1999 to 2012 with my blog, which I started around 2012. I am using blogdown, an R package that creates a website from Markdown and Rmarkdown files. It's not been easy, but when it is done, it will all be worthwhile.

<!---more--->

Yihui Xie wrote and is the current maintainer of [blogdown][xie1]. It is an R package, but it installs and uses [Hugo][hug1] to do most of the work. There is a nice book, written by Yihui Xie, Amber Thomas, and Alison Presmanes Hill that you can [browse for free on the web][xie2] or purchase in the [traditional book form][xie3].

As an aside, the book about blogdown was written using another of Yihui Xie's R packages, [bookdown][xie4]. Bookdown is based on [pandoc][pan1]. If you notice a pattern here, it is something that is fairly new in computing. It is very easy to link R to other software systems, and since many of these other systems are freely downloadable, you get a powerful and flexible web of programs that work (mostly) seamlessly and behind the scenes.

I should warn you that (at the time this blog post was written), a new version of the blogdown package was posted on the Comprehensive R Archive Network that has some major changes from the version described in the book. The changes are described on [this RStudio blog post][rst1]. There is also a nice "getting started" [blog post][hil1] by Allison Hill that outlines setting up a basic site under the new version of blogdown. My summary of how to set up a new blog site will lean closely on her overview, but I want to add a few insights specific to my particular setting.

### Step one: Set up a new project

I found using blogdown to be difficult because I am an incurable tinkerer, and most of my tinks ended up being thunks. So in order to make progress, I had to remember to start at Square One and only take baby steps. If something goes wrong, just start again at Square One rather than trying to undo things that not always easy to undo.

So even though I already had around 1,600 web pages, I created a new project and named it b01. That didn't work out, so I went to b02 and b03, befure finally getting things right with b04.

![Dialog box for new project](first-steps-with-blogdown-01.png)

I made a conscious choice to avoid version control, at least at first. I've had some issues with earlier attempts to work with blogdown and version control at the same time. Rather than wrestle with both simultaneously, I want a new project rather than a project based on version control. Later I might recognize that I was being too careful here, but for now the philosophy is "baby steps."

![Dialog box for new project](first-steps-with-blogdown-02.png)

There's another tempting option dangled in front of you with the second dialog box. You can specify the project type here and R Studio has "Website using blogdown" as one of the choices. This could be a big time saver, but I worry that R Studio may make some default choices that I don't like or that don't work on my computer. I also dislike some of the work being done behind the scenes. So again, I recognize this as a step made too cautious, but I made this a plain vanilla "New Project."

![Dialog box for project type](first-steps-with-blogdown-03.png)

The thid dialog box tells you where to place the new project and you should know what to do from this point.

### Step two: Create a new site

If you haven't done so already, install the blogdown package. You need to make a choice here. Hugo has a variety of themes that control what features your website will have along with the general appearance of the site.

The blogdown book likes yihui/hugo-lithium, Allison Hill's blog post likes wowchemy/starter-academic. There are hundreds of themes, developed by a wide range of people with different interests and different needs. Do an Internet search on "Hugo themes" to ge a feel for what these systems offer.

I tried both yihui/hugo-lithium and wowchemy/starter-academic. There were things I liked a lot about both themes, but there were also things I really hated. By the way, you can easily swithc between themes, but I made a conscious choice: better to restart at Square One. A new theme meant a new project. Again, perhaps overly cautious.

I finally settled on a theme that I liked: yihui/hugo-xmin. This is a minimal theme with not that many features. But it still looks good.

To start a website with a particular theme, type

```{}
blogdown::new_site(theme = "yihui/hugo-xmin")
```

I may end up using a different theme, but for now I like a minimal theme where I manually add enhancements at my own pace.

Almost every Hugo theme comes with a small sample of pages of the "Lorem ipsum" variety. So you can build a small website without having to first create any of your own content. To build a site, use the build_site function. I change two of the default arguments.

```{}
blogdown::build_site(local=TRUE, build_rmd=TRUE)
```

By default, blogdown builds a website without having to store any files on your system, and for most people this is a good thing. I, however, need local files that I can transfer to the ISP that provides my website space. I also include build_rmd to be set to TRUE because some of the problems I have experienced in the past involved failure on my part to rebuild all the Rmd files. It slows things down a bit, but gives me a bit of security that all recent changes have been included.

### Step three: Browse your new site locally

It's only a small site at this point, but you can get a pretty good impression of what your final website will look like by viewing a local version. You do this with the serve_site command.

```{}
blogdown::serve_site()
```

Your entire website now appears in

```{}
blogdown::serve_site()
```

### Step four: Create a new post

### Step five: Tinker

[hil1]: https://www.apreshill.com/blog/2020-12-new-year-new-blogdown/
[hug1]: https://gohugo.io/
[rst1]: https://blog.rstudio.com/2021/01/18/blogdown-v1.0/
[xie1]: https://cran.r-project.org/package=blogdown
[xie2]: https://bookdown.org/yihui/blogdown/
[xie3]: https://www.routledge.com/blogdown-Creating-Websites-with-R-Markdown/Xie-Hill-Thomas/p/book/9780815363729
[xie4]: https://cran.r-project.org/package=bookdown

### 2021-10-11 text/Rmd/quasi-likelihood/index.Rmd
Updated glm-deviance.Rmd

@@
 
-The generalized linear model for any distribution in the exponential family can be described in terms of a link function and a variance function.
+A distribution is said to be in the exponential family
+
+$ln f(y, \theta)=d(\theta)e(y)+g(\theta)+h(y)$
+
+but sometimes this is rewritten in the form
+
+$ln f(y, \theta)=\frac{\theta y+b(\theta)}{a(\phi)}+c(y, \phi)$
+
+where $\phi$ is a dispersion or scale parameter. This is called the canonical form. There are names for these parameters and functions.
+
++ $\theta$ is the natural parameter,
++ $\phi$ is the dispersion parameter,
++ $a(\ ) is the variance function, and
++ $g(\ ) = b'(\ )^{-1} is the link function.
+
+If you also define
+
+$\mu=E[Y]$
+
+then the link function, g, and its inverse, b, define how the regression model works.
 
 For Poisson regression, the link function is
+
+$ln(f(y, \lambda) = ln(\frac{\lambda^y e^{-\lambda}}{y!})=\lambda ln(\lambda)-\lambda-ln(y!)$
+
+To write this in the canonical form, you need to define the natural parameter as
+
+$\theta=ln(\lambda)$
+
+and
+
+$a(\phi)=1$
+
+The link function is
 

### 2021-10-12 text/Rmd/first-steps-with-blogdown/index.Rmd
Updated first-steps-with-blogdown

@@
 
-![Dialog box for new project](first-steps-with-blogdown-01.png)
+![Dialog box for new project](http://www.pmean.com/new-images/21/first-steps-with-blogdown-01.png)
 
@@
 
-![Dialog box for new project](first-steps-with-blogdown-02.png)
+![Dialog box for new project](http://www.pmean.com/new-images/21/first-steps-with-blogdown-02.png)
 
@@
 
-![Dialog box for project type](first-steps-with-blogdown-03.png)
+![Dialog box for project type](http://www.pmean.com/new-images/21/first-steps-with-blogdown-03.png)
 
-The thid dialog box tells you where to place the new project and you should know what to do from this point.
+The third dialog box tells you where to place the new project and you should know what to do from this point.
 
@@
 If you haven't done so already, install the blogdown package.
-You need to make a choice here.
-Hugo has a variety of themes that control what features your website will have along with the general appearance of the site.
+
+If you do not already have hugo installed on your computer, there is a function inside of the blogdown library that will do this for you.
+
+```{}
+blogdown::install_hugo()
+```
+
+There's no need to reinstall hugo unless your version of hugo is old.
+
+You need to make a choice here. Hugo has a variety of themes that control what features your website will have along with the general appearance of the site.
 
@@
 There were things I liked a lot about both themes, but there were also things I really hated.
-By the way, you can easily swithc between themes, but I made a conscious choice: better to restart at Square One.
+By the way, you can easily switch between themes, but I made a conscious choice: better to restart at Square One.
 A new theme meant a new project.

### 2021-10-12 text/Rmd/glm-deviance/index.Rmd

@@
 
-<!---More--->
+### Maximum likelihood equations for a generalized linear model
+
+The maximum likelihood equations are pretty easy to compute.
+
+$L(y, \theta)=\prod_i f(y, \theta)$
+
+$ln L(y, \theta)=\sum_i ln(f(y, \theta)$
+
+If we write the distribution in the canonical form
+
+$ln f(y, \theta)=\frac{\theta y+b(\theta)}{a(\phi)}+c(y, \phi)$
+
+which means that
+
+$ln L(y, \theta)=\sum_i \frac{\theta y}{a(\phi)} + \sum_i \frac{b(\theta)}{a{\phi}}+\sum_i c(y_i, \theta)$
+
+For a Poisson regression model,
+
+$ln L(y, \theta)=\sum_i \theta y_i + \sum_i e^\theta - \sum_i ln(y_i!)$
+
+The deviance is defined as
+
+$D=-2(ln L(y, y) - ln L(y, \hat \theta))$
+
+which in the Poisson regression model equals
+
+$D=2(\sum_i y_i(y_i-\hat \theta)+e^{y_i}-e^{\hat \theta})$
+
+In terms of $\hat \theta$, this converts to
+
+$D=2 \sum_i (y_i ln(y_i/ \hat\theta)-(y_i-\hat\lambda))$
+
+You want the deviance to as small as possible. So any term inside this sum that is big represents a data value that is contributing a lot to lack of fit.
+
+The individual terms in this sum are all positive, so we have to account for whether the lack of fit is due to underprediction or overprediction. So the deviance residual is defines as
+
+$sign(y_i-\hat\theta)\sqrt{2(y_i ln(y_i / \hat\theta)- (y_i-\hat\theta))}$
+
+### Computing deviance residuals in R
+
+Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models.
+The first data set is available for anyone to download.
+
+```{r}
+fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
+articles <- read.csv(fn)
+head(articles)
+```
+
+The variables are
++ gender (fem),
++ marital status (mar),
++ number of children (kid5),
++ prestige of graduate program (phd), 
++ the number of articles published by the individual's mentor (ment)
++ the number of articles published by the scientist (art: the outcome).
+
+To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".
+
+```{r}
+pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
+summary(pois1)
+
+rd <- articles
+rd$r <- resid(pois1, type="deviance")
+rd$p <- predict(pois1, type="response")
+```
+
+```{r}
+ggplot(rd, aes(r)) +
+  geom_histogram(bin_width=0.2)
+```
+
+```{r}
+ggplot(rd, aes(p, r)) +
+  geom_point()
+```
+
+```{r}
+ggplot(rd, aes(fem, r)) +
+  geom_point()
+```
+
+```{r}
+ggplot(rd, aes(ment, r)) +
+  geom_point()
+```
+
+```{r}
+ggplot(rd, aes(phd, r)) +
+  geom_point()
+```
+
+```{r}
+ggplot(rd, aes(mar, r)) +
+  geom_point()
+```
+
+```{r}
+ggplot(rd, aes(kid5, r)) +
+  geom_point()
+```
 
@@
 [tay1]: https://statisticaloddsandends.wordpress.com/2019/03/27/what-is-deviance/
+https://stats.stackexchange.com/questions/40876/what-is-the-difference-between-a-link-function-and-a-canonical-link-function

### 2021-10-15 text/Rmd/anova-steps/index.Rmd
Finished updating 03 files

---
title: Maximum likelihood estimation
author: Steve Simon
date: 2003-05-06
categories:
- Blog post
tags:
- Statistical theory
output: html_document
---
*Dear Professor Mean, What is maximum likelihood estimation and how does
it work?*

Maximum likelihood is an approach that looks at a large class of
distributions and then chooses the "best" distribution. For each
distribution, we compute a likelihood, and the best distribution is the
one that maximizes this likelihood.

Suppose we have a set of three data points, and we want to find the
normal distribution that is "best" for this data. We would look at a
variety of bell-shaped curves, and measure the likelihood for each bell
shaped curve. In this case, you compute the likelihood by measuring the
height of the bell at each data point and then multiply those heights
together. Shown below are three data points: 0.1, 0.4, and 0.8 along
with a bell shaped curve.

![](http://www.pmean.com/new-images/03/mle1a.gif){width="352" height="272"}

This particular bell shaped curve has the values mu=0 and sigma=1. The
heights are 0.397, 0.368, and 0.29. When you multiply these values
together, you get a likelihood of 0.0423. That's not bad. Let's see if
we can do better. Shown below is the same data with a new bell shaped
curve.

![](http://www.pmean.com/new-images/03/mle2a.gif){width="352" height="272"}

Here the bell shaped curve has mu=1.2 and sigma=1. This bell shaped
curve is not as good, because the likelihood is only 0.0232. Let's try
a third bell shaped curve.

![](http://www.pmean.com/new-images/03/mle3a.gif){width="352" height="272"}

In this bell shaped curve, mu=0.5 and sigma=1. This is the best so far,
with a likelihood of 0.0588. But we can improve further by reducing the
value for sigma.

![](http://www.pmean.com/new-images/03/mle4a.gif){width="352" height="272"}

Here, mu=0.5 and sigma=0.4. Notice now that the likelihood is 0.4402, a
substantial improvement over the previous three distributions. What
happens if we reduce sigma even further?

![](http://www.pmean.com/new-images/03/mle5a.gif){width="352" height="272"}

Oops! Too far. In the example above, mu=0.5 and sigma=0.2. The
likelihood has decreased to 0.3077.

You could continue by trial and error and eventually you would find that
the distribution with mu=0.433 and sigma=0.287 maximizes the likelihood.
You could then interpret this particular bell shaped curve as the
"best" meaning the most consistent with the data. The estimates 0.433
and 0.287 are then called the maximum likelihood estimates of mu and
sigma.

There are better approaches than trial and error. For many situations, a
bit of algebra and calculus can help you identify the maximum likelihood
estimates. You know, for example, that the formula for the height of a
bell shaped curve is

![](http://www.pmean.com/new-images/03/mle6.gif){width="94" height="52"}

so you could substitute the values of 0.1, 0.4, and 0.8 into the above
formula and then multiply them all together. You would get something
like

![](http://www.pmean.com/new-images/03/mle7.gif){width="367" height="62"}

and with a bit of patience, you could simplify this formula, take the
first derivative with respect to mu and sigma and set those derivatives
equal to zero.

You could also work out a solution for a general case where you have a
sample of n data points, x1, x2,...,xn. The likelihood would be

![](http://www.pmean.com/new-images/03/mle8.gif){width="212" height="62"}

Very carefully, you could find derivatives and set these derivatives
equal to zero. The solution represents the maximum likelihood estimates
for the general case.

You may often find it convenient to work with logarithms. The log of the
likelihood function often simplifies many of the calculations, and if
you find the maximum of the log likelihood that also has to be the
maximum of the likelihood itself. In the example above, the log
likelihood is

![](images/mle9.gif){width="298" height="51"}

That formula may not look a lot simpler, but it is easier to compute
derivatives for the log likelihood in this and many other situations.

I won't show all the equations, but the maximum likelihood estimate of
mu ends up equaling the sample mean and the maximum likelihood estimate
of sigma ends up equaling, not the sample standard deviation exactly,
but something very close where you replace n-1 with n in the formula.

Sometimes even calculus fails us, and we have to rely on computer
approximations. We know a lot about using computers to maximize complex
functions, so this is actually better than it sounds.

You can find an [earlier version][sim1] of this page on my [original website][sim2].

[sim1]: http://www.pmean.com/03/mle.html
[sim2]: http://www.pmean.com/original_site.html

### 2021-10-15 text/Rmd/first-steps-with-blogdown/index.Rmd
Finished updating 03 files

--- 2021-10-09 text/Rmd/first-steps-with-blogdown/index.Rmd
@@
 
-![Dialog box for project type](first-steps-with-blogdown-03.png)
+![Dialog box for project type](http://www.pmean.com/new-images/21/first-steps-with-blogdown-03.png)
 
@@
 
+![Preview website main page](http://www.pmean.com/new-images/21/first-steps-with-blogdown-04.png)
+![Preview website main page](http://www.pmean.com/new-images/21/first-steps-with-blogdown-05.png)
+
+![Preview website main page](http://www.pmean.com/new-images/21/first-steps-with-blogdown-06.png)
+
+![Preview website main page](http://www.pmean.com/new-images/21/first-steps-with-blogdown-07.png)
+
 ### Step four: Create a new post

### 2021-10-16 text/Rmd/maximum-liikelihood/maximum-likelihood.Rmd

--- 2021-10-15 text/Rmd/anova-steps/index.Rmd
@@
 
-![](http://www.pmean.com/new-images/03/mle1a.gif){width="352" height="272"}
+![](mle1a.gif){width="352" height="272"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle2a.gif){width="352" height="272"}
+![](mle2a.gif){width="352" height="272"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle3a.gif){width="352" height="272"}
+![](mle3a.gif){width="352" height="272"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle4a.gif){width="352" height="272"}
+![](mle4a.gif){width="352" height="272"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle5a.gif){width="352" height="272"}
+![](mle5a.gif){width="352" height="272"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle6.gif){width="94" height="52"}
+![](mle6.gif){width="94" height="52"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle7.gif){width="367" height="62"}
+![](mle7.gif){width="367" height="62"}
 
@@
 
-![](http://www.pmean.com/new-images/03/mle8.gif){width="212" height="62"}
+![](mle8.gif){width="212" height="62"}
 
@@
 
-![](images/mle9.gif){width="298" height="51"}
+![](mle9.gif){width="298" height="51"}
 

### 2021-10-17 text/04/early-stopping-animal.Rmd
Renamed files in 2004 folder

---
title: Early stopping in an animal study
author: Steve Simon
date: 2004-07-01
categories:
- Blog post
tags:
- Early stopping
output: html_document
---
I was helping someone with power calculations for an animal experiment.
There were several ways to estimate power and they all seemed to point
to the use of either 8 animals per group or 12 animals per group. I
explained that if she was more risk averse, she should use 12 animals
and if she were more work averse, she should use 8 animals. Her answer
surprised me a bit. She said that she liked the idea of using 12 animals
per group. That way, if the results achieved significance after only 8
animals, she could go back to the Animal Care and Use Committee and tell
that she saved a few animals.

This is an example of early stopping of a clinical trial (also described
as an interim analysis). Generally, if you think you might want to stop
a trial early, you should specify the conditions in your protocol prior
to the start of data collection. Without such a rule, any examination of
the data prior to the end of the study is a protocol deviation. As such,
it should be reported to the IRB (if it is human research) or the ACUC
(if it is animal research).

There are reasonable examples of why you might want to stop a study
early even without planning ahead for such an event. For example, if a
serious unexpected side effect appears, you might not want to offer the
new drug to any more patients.

But stopping the study early (like any other protocol deviation) does
reduce the credibility of the research findings, sometimes by a little
bit, sometimes by a lot.

Here's a little simulation that you can run in S-plus or R.

```{}
par(mar=c(4,5,0,0)+0.1,cex=2,pch="+")
f.p <- function(mua) {
  x1 <- rnorm(12)
  x2 <- mua+rnorm(12)
  p.value <- rep(NA,12)
  for (i in 2:12) {
    p.value[i] <- t.test(x1[1:i],x2[1:i])p.value
 }
 plot(p.value,type="b",ylim=0:1,cex=2)
 abline(h=0.05)
}
for (i in 1:100) {
  f.p(mua=1)
}
```

You'll see a variety of different patterns for the p-values.

![](http://www.pmean.com/images/04/EarlyStopping01.gif)

In the graph shown above the test never quite reaches statistical
significance.

![](http://www.pmean.com/images/04/EarlyStopping02.gif)

Here's one where it just reaches statistical significance at the last
sample.

![](http://www.pmean.com/images/04/EarlyStopping03.gif)

Here's one where it reaches statistical significance rather early. You
would think that it would be nice to stop the experiment early here, but
there is no guarantee that just because the p-value is below the
significant level early on, that it will stay below that level the rest
of the way.

![](http://www.pmean.com/images/04/EarlyStopping04.gif)

Notice that the above graph shows statistical significance early, but
then that significance disappears by the end of the experiment. Now this
doesn't happen very often. Out of 100 graphs only 4 or 5 showed this
type of pattern. Most of the time if the p-value crosses below the 0.05
threshold early, it stays below that threshold. But these exceptions
serve as a warning. If you stop the study early and make no adjustments
to your p-value, your results may be invalid.

The adjustments to the p-value (and confidence interval) are rather
complex. Here's some more S-plus (R) code that shows some of the
complexities of the situation. This program simulates an experiment with
12 animals per group, but looks at the p-value with after 4, 8, and 12
animals have been tested.

```{r, echo=TRUE}
n <- 5000
p4 <- rep(NA,n)
p8 <- rep(NA,n)
p12 <- rep(NA,n)
for (i in 1:n) {
  x1 <- rnorm(12)
  x2 <- rnorm(12)
  p4[i] <- t.test(x1[1:4],x2[1:4])$p.value
  p8[i] <- t.test(x1[1:8],x2[1:8])$p.value
  p12[i] <- t.test(x1[1:12],x2[1:12])$p.value
}
sum(p4<0.05)/n
sum(p8<0.05)/n
sum(p12<0.05)/n
sum(pmin(p4,p8,p12)<0.05)/n
```

Notice that each p-value considered in isolation maintains an alpha
level of approximately 0.05. But when you combine the p-values and allow
for statistical significance when either the n=4, n=8, or n=12 p-value
achieves significance, the alpha level more than doubles.

**Further reading:**

Ludbrook J. Interim analyses of data as they accumulate in laboratory
experimentation. BMC Med Res Methodol 2003: 3(1); 15.
[Medline](http://www.ncbi.nlm.nih.gov/entrez/query.fcgi?cmd=Retrieve&db=PubMed&list_uids=12930561&dopt=Abstract)
[Abstract](http://www.biomedcentral.com/1471-2288/3/15/abstract)
Avialable in [html format](http://www.biomedcentral.com/1471-2288/3/15)
or [pdf format](http://www.biomedcentral.com/content/pdf/1471-2288-3-15.pdf)

+ The Ludbrook reference points out that laboratory
experiments often fail to account properly for early stopping
(interim analysis) and suggests an approach for handling this
properly.

You can find an [earlier version](http://www.pmean.com/04/EarlyStoppingAnimal.html) of this page on my [original website](http://www.pmean.com/original_site.html).

### 2021-10-18 text/21/glm-diagnostics.Rmd
Added glm-diagnostics

---
title: "Diagnostics for a generalized linear model"
author: "Steve Simon"
source: "New"
date: "2021-10-18"
categories:
- Blog post
tags:
- Generalized linear model
output: html_document
---

There are a broad range of diagnostic methods (mostly graphical) that can help you decide from among a dizzying array of alternative models for count data. The most important consideration is to start at the shallow end of the pool with the simplest approaches. Don't jump in at the deep end and try something really complex before you've had the chance to look at simpler models, even if you know they are overly simplistic.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(pscl)))
```

### The dataset used in this example

Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models. The first data set is available for anyone to download.

```{r}
fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
articles <- read.csv(fn)
head(articles)
```

The variables are
+ gender (fem),
+ marital status (mar),
+ number of children (kid5),
+ prestige of graduate program (phd), 
+ the number of articles published by the individual's mentor (ment)
+ the number of articles published by the scientist (art: the outcome).

### Histogram of outcome variable

Look at the possible values for your outcome variable, which is usually a count. If you have rate data, do a histogram both for counts and rates.

```{r}
ggplot(articles, aes(art)) + 
  geom_histogram(binwidth=1, color="black", fill="white")
```

Compare this to a bar chart of Poisson probabilities. The outcome variable from a Poisson regression model does not need to follow a single simple Poisson set of probabilities. It is more likely a mixture of different Poisson variables depending on the covariate pattern. Nevertheless, it can't hurt to look at the Poisson probabilities, as long as you don't get too concerned about discrepancies between the actual data and the Poisson probabilities.

```{r}
data.frame(x=0:max(articles$art)) %>%
  mutate(p=dpois(x, lambda=mean(articles$art))) %>%
  ggplot(aes(factor(x), p)) +
    geom_col(color="black", fill="white")
```

Here there seems to be a disparity at zero, which may (or may not) indicate the need for a zero-inflated model.

### Descriptive statistics on indepdent variables

It helps to know the general descriptive statistics (means, standard deviations, percentages) for each of the independent variables.

```{r}
articles %>%
  group_by(fem) %>%
  summarize(n=n())
```

```{r}
quantile(articles$ment)
```

```{r}
quantile(articles$phd)
```

```{r}
articles %>%
  group_by(mar) %>%
  summarize(n=n())
```

```{r}
articles %>%
  group_by(kid5) %>%
  summarize(n=n())
```

The wide ranges for ment and phd may make it difficult to interpret. A simple approach is to scale them to the interval 0 to 1.

```{r}
articles %>%
  mutate(ment=(ment-min(ment))/(max(ment)-min(ment))) %>%
  mutate(phd=(phd-min(phd))/(max(phd)-min(phd))) -> articles_scaled
```

### Bivariate plots

Although you are going to fit multiple independent variables, it does not hurt to look at the bivariate relationships first. A variable that looks important from a bivariate perspective, may be unimportant from the multivariate perspective and vice versa. So don't read too much into these plots. It is intended to get you familiar with trends and patterns that might (or might not) show up in more complex models.

Use boxplots when the independent variable is categorical and scatterplots when the independent variable is continuous.

```{r}
ggplot(articles, aes(factor(fem), art)) +
  geom_boxplot()
```

```{r}
ggplot(articles, aes(ment, art)) +
  geom_point() + 
  geom_smooth()
```

```{r}
ggplot(articles, aes(phd, art)) +
  geom_point() + 
  geom_smooth()
```

```{r}
ggplot(articles, aes(factor(mar), art)) +
  geom_boxplot()
```

```{r}
ggplot(articles, aes(factor(kid5), art)) +
  geom_boxplot()
```

There aren't any huge trends here. It looks like there might not be much going on here at all. If there are differences, they seem to show up in the values above the 75th percentile.

The one cause for early concern is that the effect of ment seems to level off after a certain point.

### Fit simple models first

```{r}
pois_fem <- glm(art ~ fem, family=poisson, data = articles_scaled)
summary(pois_fem)
```

```{r}
pois_ment <- glm(art ~ ment, family=poisson, data = articles_scaled)
summary(pois_ment)
```

```{r}
pois_phd <- glm(art ~ phd, family=poisson, data = articles_scaled)
summary(pois_phd)
```

```{r}
pois_mar <- glm(art ~ mar, family=poisson, data = articles_scaled)
summary(pois_mar)
```

```{r}
pois_kid5 <- glm(art ~ kid5, family=poisson, data = articles_scaled)
summary(pois_kid5)
```

For all of these models, the deviance is much greater than the degrees of freedom, indicating a poor fit. This could be failure to include all the variables in a multivariate sense or a problem with the underlying Poisson model.

Don't put too much stock in the p-values just yet, but several of them are small. This may change when we look at the independent variables in combination.

### Fit graually more complex models

```{r}
pois_all_ivs <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles_scaled)
summary(pois_all_ivs)
```

```{r}
tidy(pois_fem) %>%
  bind_rows(tidy(pois_ment)) %>%
  bind_rows(tidy(pois_phd)) %>%
  bind_rows(tidy(pois_mar)) %>%
  bind_rows(tidy(pois_kid5)) %>%
  filter(term != "(Intercept)") %>%
  mutate(unadjusted_rate_ratio=round(exp(estimate), 2)) %>%
  select(term, unadjusted_rate_ratio) -> pois_crude

tidy(pois_all_ivs) %>%
  filter(term != "(Intercept)") %>%
  mutate(adjusted_rate_ratio=round(exp(estimate), 2)) %>%
  select(term, adjusted_rate_ratio) -> pois_adjust

pois_crude %>%
  full_join(pois_adjust)

```

### Examine residuals

A good general plot is fitted (predicted) values versus residuals. Since the fitted value is a composite of all the independent variables, it is likely to show unusual behavior if there are issues involving any of the independent variables.

An ideal pattern would be a random scatter of points spread across all regions of the graph. A curvilinear pattern (e.g., high in the middle, low at either extreme) or a fanning out pattern (e.g., more variation on the right side of the graph than the left) is indication that there are unusual features in the data that are not fully captured by the current model.

```{r}
r_pois <- augment(pois_all_ivs, articles_scaled, type.predict="response", type.residuals="deviance")
ggplot(r_pois, aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth()
```

Clearly there is something unusual going on with the largest fitted value.

```{r}
r_pois %>%
  filter(.fitted > 7) %>%
  data.frame
```

This appears to be the subject where the scaled value for ment equals 1. In other words, the highest possible value for ment.

You should also plot the residuals versus every independent variable, although the plots versus binary independent variables are unlikely to show anything interesting.

```{r}
ggplot(r_pois, aes(factor(fem), .resid)) + 
  geom_boxplot()
```

```{r}
ggplot(r_pois, aes(ment, .resid)) + 
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(r_pois, aes(phd, .resid)) + 
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(r_pois, aes(factor(mar), .resid)) + 
  geom_boxplot()
```

```{r}
ggplot(r_pois, aes(factor(kid5), .resid)) + 
  geom_boxplot()
```

Perhaps, the value of a mentor is good only to a point and anything after offers no additional help. Let's compute a new value for ment where any value above 40 is set equal to 40.

```{r}
articles %>%
  mutate(ment=pmin(40, ment)) %>% 
  mutate(ment=(ment-min(ment))/(max(ment)-min(ment))) %>%
  mutate(phd=(phd-min(phd))/(max(phd)-min(phd))) -> articles_rescaled
pois_rescaled <- 
  glm(
    art ~ fem + ment + phd + mar + kid5, 
      family=poisson, 
        data = articles_rescaled)
r_rescaled <- 
  augment(
    pois_rescaled, 
      articles_rescaled, 
        type.predict="response", 
          type.residuals="deviance") 
ggplot(r_rescaled, aes(.fitted, .resid)) +
  geom_point() +
  geom_smooth()
ggplot(r_rescaled, aes(ment, .resid)) +
  geom_point() +
  geom_smooth()
```

### Negative binomial model

```{r}
library(MASS)
nb1 <- glm.nb(art ~ fem + ment + phd + mar + kid5, data = articles_rescaled)
summary(pois_rescaled)
summary(nb1)
```

### Zero inflated model

```{r}
zip1 <- 
  zeroinfl(
    art ~ fem + ment + phd + mar + kid5 | 
      1, 
        dist="poisson", 
          data = articles_rescaled)
summary(zip1)
```

```{r}
zip2 <- 
  zeroinfl(
    art ~ fem + ment + phd + mar + kid5 | 
      fem + ment + phd + mar + kid5, 
        dist="poisson",
          data = articles_rescaled)
summary(zip2)
```

```{r}
zip3 <- 
  zeroinfl(
    art ~ fem + ment + phd + mar + kid5 | 
      ment, 
        dist="poisson",
          data = articles_rescaled)
summary(zip3)
```

```{r}
zinb <- 
  zeroinfl(
    art ~ fem + ment + phd + mar + kid5 | 
      ment, 
        dist="negbin",
          data = articles_rescaled)
summary(zinb)
```

### Comparing competing models

The AIC statistic is useful for comparing models, but it is not produced automatically for negative binomial regression or the zero-inflated models. It requires you to pull the individual components from the

```{r}
save.image(file="glm_diagnostics.RData")
```

### 2021-10-21 src/clean_files.R
Updated image links, part 1

@@ -60,8 +60,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '\\- Quotations',
-  '- Quotation',
+  'http://www.pmean.com/images/images/19/',
+  'http://www.pmean.com/new-images/19/',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))

### 2021-10-21 src/clean_files.R
Update image links, part 5

@@ -60,8 +60,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'http://www.pmean.com/images/images/19/',
-  'http://www.pmean.com/new-images/19/',
+  'http://www.pmean.com/images/images/0',
+  'http://www.pmean.com/new-images/0',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))

### 2021-10-22 src/clean_files.R
Cleaned up tags

@@ -60,8 +60,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'http://www.pmean.com/images/images/0',
-  'http://www.pmean.com/new-images/0',
+  '<U+00A0>',
+  ' ',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))

### 2021-10-23 text/20/ihme-model.Rmd
Relocated .Rmd files

--- 2020-07-23 text/20/ihme-covid-model.Rmd
@@
 source: "New"
+date: 2020-04-23
 categories:

### 2021-10-23 text/20/non-breaking-spaces.Rmd
Relocated .Rmd files

--- 2020-07-23 text/20/non-breaking-space.Rmd
@@
 ---
-title: "Blog post: Identifying and manipulating non-breaking spaces"
+title: "Identifying and manipulating non-breaking spaces"
 author: "Steve Simon"

### 2021-10-23 text/20/reading-text-files.Rmd
Relocated .Rmd files

@@
 
+```{}
 ![A non-text file displayed in notepad](../../../web/images/reading-text-files01.png)
+```
 
@@
 
+```{}
 ![Data set on movies](../../../web/images/read-text-files03.png)
+```
 
@@
 
+```{}
 ![Data dictionary](../../../web/images/read-text-files06.png)
+```
 
@@
 
+```{}
 ![Example of a space delimited file](../../../web/images/read-text-files02.png)
+```
 
@@
 
+```{}
 ![Dataset showing COVID-19 deaths over time](../../../web/images/read-text-files04.png)
+```
 
@@
 
+```{}
 ![Dataset of revenues for selected movies](../../../web/images/read-text-files05.png)
+```
 

### 2021-10-23 text/21/anonymizing-ids.Rmd
Relocated .Rmd files

@@
 source: "New"
-date: "2020-06-10"
+date: "2021-06-10"
 categories:

### 2021-10-23 text/21/read-acs-data.Rmd
Relocated .Rmd files

---
title: "Reading information from the American Community Survey into R"
source: "New"
date: "2021-09-29"
categories:
- Recommendation
tags:
- Website details
output: html_document
---

Here is a simple example of how to read data from the American Community Survey into R.

<!---more--->

The Bureau of the Census allows computerized systems to access their data. This is a great convenience for you, but raises a potential security risk for them. To provide them with a reasonable amount of protection, the Bureau requires that you sign up for a free key that you insert in your code. The key allows the Bureau to track requests to insure that you are not a robot trying to scoop up an excessive amount of data. You get a key on [this page][cen3]. Look for "Request a KEY" in the left side menu (see the first screenshot at the bottom of this page).

You can embed this key so that it can be re-used at will. Use the census_api_key function in the tidycensus package.

```{r}
library(tidycensus)
library(tidyverse)
library(viridis)
key <- Sys.getenv("CENSUS_API_KEY")
census_api_key("11bafd82737f7d75b3305ada1e772501d0eab002")
tarr <- get_acs(geography = "tract", variables = "B19013_001",
state = "TX", county = "Tarrant", geometry = TRUE)
ggplot(tarr, aes(fill = estimate, color = estimate)) +
geom_sf() +
coord_sf(crs = 26914) +
scale_fill_viridis(option = "magma") +
scale_color_viridis(option = "magma")
```

### Resources

Ari Lamstein, Logan Powell. A Guide to Working with US Census Data in R. Available in [mhtml format][lam1] and at their [github site].

U.S. Census Bureau, Geography and the American Community Survey: What Data Users Need to Know. Available in [pdf format][cen1].

[lam1]: https://rconsortium.github.io/censusguide/r-packages-all.html
[lam2]: https://github.com/RConsortium/censusguide

[cen1]: https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_geography_handbook_2020.pdf

[cen2]: https://www.census.gov/content/dam/Census/library/publications/2020/acs/acs_api_handbook_2020_ch02.pdf

[cen3]: https://www.census.gov/data/developers.html

### Screenshots

#### Request an API key

![](reading-acs-data-01.png)

### 2021-10-23 text/21/zero-inflated-poisson.Rmd
Relocated .Rmd files

---
title: "Zero-inflated models"
author: "Steve Simon"
source: "New"
date: "2021-09-20"
categories:
- Blog post
tags:
- Generalized linear model
output: html_document
---

In models of counts, zeros represent a special case. Often there is more than one reason why you might observe a zero count, and you might want to model this probability separately. One approach is known as a zero inflated Poisson model.

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(pscl)))
```

### The zero inflated Poisson distribution

In any Poisson distribution, the probabilty of obtaining a zero is defined as

$P[Y=0]=\frac{\lambda^0e^{-\lambda}}{0!}=e^{-\lambda}$

Sometimes this probability is too low because there are multiple reasons why you might observe a zero. The zero-inflated Poisson model is a mixture of the Poisson random variable with a random variable that places all of its probability at zero.

$\pi \ 0 + (1-\pi)\frac{e^{-\lambda}\lambda^y}{y!}$

This changes the probability of zero to

$P[Y=0]=\pi+(1-\pi)e^{-\lambda}$

Since $\lambda$ has to be a positive value, the probability above is greater than the Poisson probability for any value of $\pi>0$.

The remaining probabilities are shrunk by a factor of $1-\pi$ to compensate for the increase in the probability at zero.

You can develop a separate regression model for $\pi$ the increase in the probability of zero in addition to the regression model for $\lambda$, the mean of the Poisson distribution.

The zero-inflated Poisson probability distribution is not part of the exponential family, so you need separate methods for estimation. You can find a function in the pscl library that will fit a zero-inflated Poisson model.

### Start with Poisson regression

It is easiest to understand the quasiPoisson model by comparing it to the Poisson model.

Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models. The first data set is available for anyone to download.

```{r}
fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
articles <- read.csv(fn)
head(articles)
```

The variables are
+ gender (fem),
+ marital status (mar),
+ number of children (kid5),
+ prestige of graduate program (phd), 
+ the number of articles published by the individual's mentor (ment)
+ the number of articles published by the scientist (art: the outcome).

To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".

```{r}
pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
summary(pois1)
```

The interpretation of the coefficients should be done after exponentiating their values.

```{r}
pois1 %>%
  tidy %>%
  mutate(lower_limit=estimate-1.96*std.error) %>%
  mutate(upper_limit=estimate+1.96*std.error) %>%
  select(term, estimate, lower_limit, upper_limit) %>%
  mutate(estimate=exp(estimate)) %>%
  mutate(lower_limit=exp(lower_limit)) %>%
  mutate(upper_limit=exp(upper_limit)) -> pois1_coefficients
pois1_coefficients
```

### Fitting a zero-inflated-Poisson model

```{r}
zip1 <- zeroinfl(art ~ fem + ment + phd + mar + kid5 | 1, family=poisson, data = articles)
summary(zip1)
```

We cannot use the broom package to extract coefficients.

### References

### 2021-10-29 text/21/very-simple-splines.Rmd

---
title: "A systematic review shows no performance benefit of machine learning over logistic regression for clinical prediction models"
author: "Steve Simon"
source: "http://blog.pmean.com/machine-learning-vs-logistic-regression/"
date: "2019-03-15"
categories:
- Recommendation
tags:
- Data science
- Logistic regression
- Systematic overviews
output: html_document
---

### Downward bend

Suppose you want to fit a function that is flat from x=a to x=b and which declines linearly from x=b to x=c.

Here is the formula:

$f(x) = d - e*(x-b)_+$

where

$u_+ = u$ if $u \ge 0$, $= 0$ if $u<0$

Here is some simulated data for testing this function.

```{r, echo=FALSE}
library(ggplot2)
x <- 1:20
y <- c(rep(8, 15), 7:3)+round(rnorm(10), 1)
d <- data.frame(x, y)
d
ggplot(d, aes(x, y)) +
  geom_point()
```

```{r}
par(mar=rep(0.1,4))
plot(c(-1,2), 0:1, type="n", axes=FALSE, xlab="", ylab="")
segments(-1, 0, 2, 0)
segments(-1, 1, 2, 1)
segments(-1, 0.6, 0, 0.6)
segments(-1, 0.9, 0, 0.9)
segments( 1, 0.4, 2, 0.4)
segments( 1, 0.8, 2, 0.8)
segments( 0, 0.35, 1, 0.35)
segments( 0, 0.95, 1, 0.95)
x <-seq(0,1, length=100)
y <- 3*x^2-2*x^3
lines(x, 0.35+0.05*y)
lines(x, 0.5+0.05*y)
lines(x, 0.5+0.3*y)
lines(x, 0.6+0.3*y)
lines(x, 0.6-0.25*y)
lines(x, 0.65-0.25*y)
lines(x, 0.65-0.1*y)
lines(x, 0.85-0.15*y)
lines(x, 0.85+0.05*y)
lines(x, 0.9-0.15*y)
lines(x, 0.9+0.05*y)
lines(x, 0.9-0.15*y)
lines(x, 0.95-0.15*y)
```

```{r}
x <- seq(0,1, length=100)
y <- 3*x^2-2*x^3
plot(x, y)
```

Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid. A review of spline function procedures in R. BMC Medical Research Methodology, 2019-03-06, 19(46). Available in [html format][per1] or [pdf format][per2].

[per1]: https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3
[per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf

### 2021-10-29 text/11/standard-deviation-ci.Rmd
Updated more files in 2011

---
title: Calculating a confidence interval for a standard deviation
author: Steve Simon
date: 2011-07-10
categories:
- Blog post
tags:
- Confidence intervals
output: html_document
---

Suppose that you had a sample of 80 observations and you computed a standard deviation of those 80 observations. Like any other statistic, the standard deviation will have some sampling error associated with it. But how much sampling error. This is an important question, because I needed to establish that 80 was a reasonable sample size in a study where there was no formal research hypothesis. You do this by showing that key statistics that you will estimate have good precision. In this study, the standard deviation was especially important, so I wanted to show that a standard deviation based on 80 observations had a good level of precision.

<!---More--->

It's a standard result in mathematical statistics that the sample variance (variance is the square of the standard deviation) is distributed as a multiple of a Chi-square random variable. To be more precise, define the greek letter sigma (σ) as the unknown standard deviation of the population and the roman letter S as the standard deviation computed from a sample of n observations.

If the data comes from a normally distributed population, then n-1 times the sample variance divided by the population variance has a Chisquare distribution with n-1 degrees of freedom.

$\frac{(n-1)S^2}{\sigma^2} \sim \chi^2(n-1)$

Now if the data comes from a non-normally distributed population, you still may be okay. If the sample size is large and if certain other conditions exist, then the above distribution will hold approximately.

Now if the ratio follows a Chi-square distribution, then you can make the following probability statement.

$P[\chi^2(\alpha/2,n-1) > \frac{(n-1)S^2}{\sigma^2}  > \chi^2(1-\alpha/2;n-1)]=1-\alpha$

Re-arrange the terms algebraicly to get

$P[\frac{(n-1)S^2}{\chi^2(1-\alpha/2;n-1)} < \sigma^2 < \frac{(n-1)S^2}{\chi^2(\alpha/2;n-1)}]=1-\alpha$

The two limits

$\frac{(n-1)S^2}{\chi^2(n-1;1-\alpha/2)}$

and

$\frac{(n-1)S^2}{\chi^2(n-1;\alpha/2)}$

represent a 95% confidence interval for the sample variance. Take the square root of both sides to get a 95% confidence interval for the sample standard deviation.

There's a subtle shift that happened in our algebraic manipulations that appears in the final formula. Notice that the upper percentile (the 1-alpha/2 percentile) of the Chi-square distribution appears in the denominator of the lower confidence limit. This happened because in an intermediate step of the algebraic manipulation not shown here, you need to invert the fraction and at the same time invert the inequality (if A < B < C then 1/C < 1/B < 1/A). It make sense, of course, because the numerators are the same. So lower limit has to have a larger number in the denominator than the upper limit.

Something else interesting is going on. Most confidence intervals that you have seen are additive--your interval is that sample estimate plus or minus a certain value. This confidence interval, though, is multiplicative--your interval is the sample estimate multiplied by a couple of values.

In our example, there were 80 observations. The appropriate limits of the Chi-square distribution are

```{r}
qchisq(0.025, 79)
```

and

```{r}
qchisq(0.975, 79)
```

The upper and lower limits of the 95% confidence interval for the standard deviation, then are

$S \sqrt{\frac{79}{105.5}} = 0.865*S$

and

$S \sqrt{\frac{79}{56.3}}=1.184*S$

Notice that I put the larger Chisquare percentile in the denominator of the lower confidence limit. Our confidence interval is 86% to 118% of the sample standard deviation, which indicates a small amount of sampling error.

You can find an [earlier version][sim1] of this page on my [original website][sim2].

[sim1]: http://www.pmean.com/11/StandardDeviation.html
[sim2]: http://www.pmean.com/original_site.html

### 2021-10-30 text/21/dark-side.Rmd
Started work on dark-side.Rmd

---
title: "The dark side of big data"
source: "New"
date: "2021-10-28"
categories:
- Blog post
tags:
- Big data
- Incomplete page
output: powerpoint_presentation
---

### Who am I?

Steve Simon

+ PhD Statistics, 1982, U Iowa
+ Teach in Biomedical and Health Informatics
  + Previous jobs at CMH, CDC
+ Part-time independent statistical consultant (P.Mean Consulting)
+ Married to a Pediatric Cardiologist (retired)
+ Run 5K and 4 mile races

<div class="notes">

Let me introduce myself. I am Steve Simon. I got a PhD in Statistics almost 40 years ago from the University of Iowa. The world has changed a lot since then, but I have tried to keep up. Today, if you want to sound trendy, you are a "data scientist". I teach in the Department of Biomedical and Health Informatics at UMKC. I have had previous jobs at Children's mercy Hospital, and the Centers for Disease Control and Prevention. I'm also a part-time statistical consultant. I have a sole proprietorship, P.Mean Consulting. That's short for Professor Mean. For people who don't get the joke, I point out that Professor Mean is not just your average Professor. Related to this talk, I should point out that I am obsessed with computers.

</div>

### Obsessed with computers since 1972

![Figure 1. Section on computer skills from my resume](http://www.pmean.com/new-images/21/dark-side-03.png)

### First poll question

![Figure 2. Quote from "Peggy Sue Got Married"](http://www.pmean.com/new-images/21/dark-side-01.png)

<div class="notes">

Think about your high school algebra class.

1. Do you remember any important formulas from that class?

2. Did you hate, hate, hate high school algebra?

3. Did you love high school algebra?

Big question: Will you use high school algebra in your future?

Source: https://www.moviequotes.com/s-movie/peggy-sue-got-married/

</div>

### Second poll question

![Figure 3. Images of various computers](http://www.pmean.com/new-images/21/dark-side-02.png)

<div class="notes">

How many of these computational devices have you used?

1. Laptop computer
2. Desktop computer
3. Tablet computer
4. Smart phone
5. Gaming console
6. Smart watch

Big question: What impact does the current generation's immersion of computing have on society?

Images found at

https://commons.wikimedia.org/wiki/File:Black_laptop_computer_open_frontal.svg
https://www.pcmag.com/picks/the-best-desktop-computers
https://www.cleverfiles.com/howto/what-is-tablet-computer.html
https://www.theverge.com/21420196/best-budget-smartphone-cheap
https://www.vulture.com/article/best-video-game-console-2020-ps5-xbox-series-x-nintendo-switch.html
https://www.homernews.com/marketplace/fitnus-smartwatch-review-legit-fitness-tracker-smart-watch/

</div>

### Are Statisticians Gods?

I'm helping someone who wants an alternative statistical analysis to the one used by the principal investigator. I'm happy to help and will offer advice about why my approach may be better, but I was warned that the PI considers the analysis chosen to be ordained by the **"Statistical Gods"** at her place of work.

<div class="notes">

True story. I was asked to review a report from a federal agency, but was warned that negative comments, even if accurate, might not be well received because the agency's work was ordained by their Statistic Gods. At first, I thought this was amusing. If I could get the title of "Statistical God", I could double my hourly consulting rate. But deep down this story really bothered me. It implies that Statistical skills are supernatural, and only available to a select few special people. I learned Statistics through hard work, and you can learn Statistics through hard work also. Some people will learn it faster than others because they have good aptitudes in mathematics and programming. But there is nothing other than time to stop you.

</div>

### Statistics are a social construct

![Figure 4. Cover from Joel Best's book](http://www.pmean.com/new-images/21/dark-side-04.png)

<div class="notes">

I'm going to share a message that is controversial, but which I believe deeply in. It's a message from a book by Joel Best: Lies, Damned Lies, and Statistics. Untangling Numbers from the Media, Politicians, and Activists. He argues that Statistics are a social construct. I need to define social construct here. A social construct is a method by which people create their perception of reality.

That's certainly true of Statistics. Statistcs are "selected, shaped, and presented by human beings"

Source of quote: Milo Shield, Teaching the Social Construction of Statistics.  Available at https://www.researchgate.net/publication/277299441_Teaching_the_Social_Construction_of_Statistics

Image taken from https://www.ucpress.edu/book/9780520274709/damned-lies-and-statistics.

</div>

### The arrogance of empiricism (1/2)

"I often say that when you can measure what you are speaking about, and express it in numbers, you know something about it; but when you cannot measure it, when you cannot express it in numbers, your knowledge is of a meagre and unsatisfactory kind; it may be the beginning of knowledge, but you have scarcely in your thoughts advanced to the state of Science, whatever the matter may be." Lord Kelvin.

<div class="notes">

The idea that Statistics are a social construction flies in the face of empiricism, the philosophy that experiments can reveal the realities of the world.

Here is a quote in support of empiricism. I won't read all of the first quote, but notice the sneering phrase that without numbers, "your knowledge is of a meagre and unsatisfactory and unsatisfactory kind."

</div>

### The arrogance of empiricism (2/2)

"No human investigation can be called real science if it cannot be demonstrated mathematically." Leonardo da Vinci

<div class="notes">

Here is a second quote that is just as bad. Science without numbers (math) is not real.

</div>

### Why empiricism fails (1/3)

"The government is very keen on amassing statistics. They collect them, add them, raise them to the nth power, take the cube root and prepare wonderful diagrams. But you must never forget that every one of these figures comes in the first instance from the village watchman, who just puts down what he damn well pleases." Sir Josiah Stamp

<div class="notes">

This quote from Josiah Stamp illustrates why Statistics are a social construct. It doesn't matter how fancy we dress up the numbers, nth powers, cube roots. It all comes down to the village watchman who does what he damn well pleases.

The village watchman is part of the social constuction of Statistics. So are the government officials who do all sorts of mathematical manpulations which provide an impression of objectivity.

</div>

### Why empricism fails (2/3)

![Figure 5. Cover of Stephen Jay Gould's book](http://www.pmean.com/new-images/21/dark-side-05.png)

<div class="notes">

There's a great book by Stephen Jay Gould that really makes you think hard about numbers. The book, The Mismeasure of Man, talks about IQ testing from a historical perspective and totally destroys the idea that intelligence is a trair that can be easily quantified. He starts with efforts in the 1800s to use the size of a person's skull as a measure of their intelligence. A big skull means a big brain that lives inside there and bigger brains are smarter brains.

</div>

### Why empricism fails (3/3)

![Figure 6. Frame from Yes, Prime Minister video clip](http://www.pmean.com/new-images/19/leading-questions01.png)

<div class="notes">

This short video clip is an excellent illustration of how the questions leading up to a particular question on a survey can bias the response to that survey. It comes from a British comedy, Yes Prime Minister, that ran in the 1980s.

YouTube. Leading Questions- Yes Prime Minister. Taken from the 1st Season of Yes Prime Minister – Episode 2, The Ministerial Broadcast. 2:16 running time. Published on Jan 15, 2012.

https://www.youtube.com/watch?v=G0ZZJXw4MTA

</div>

### But numbers still have value

Recognizing that Statistics are a social construction does not mean that Statistics are worthless and that experiments are a waste of time.

### Example: Quality of Life

![Figure 7. First few questions from SF-36 form](http://www.pmean.com/new-images/21/dark-side-08.png)

<div class="notes">

Quality of Life is a measure that is reported in many research studies. There is no question that quality of life is a social construction. It is measured by forms like the SF-36 shown here. But it is not an objective reality but something that was constructed by humans.

Now I mentioned earlier that I like to run. It is a pretty strenuous activity, but one that I enjoy. A friend of mine said, though, that she couldn't understand why you would want to run, unless you were being chased. One the other hand, I would not miss a thing if I did not push a vacuum, bowl, or play golf.

There is no objective measure of quality of life. It depends on the values of the people who are developing forms like the SF-36.

But we can still use social constructs like quality of life, as long as we do it carefully.

Image obtained from

https://clinmedjournals.org/articles/jmdt/jmdt-2-023-figure-1.pdf

</div>

### Rules for using Statistics as social constructs

1. Understand the context in which the Statistic was generated.

2. Identify possible biases

3. Recognize limitations

4. Beware of confirmation bias

5. Avoid nihilistic thinking

<div class="notes">

</div>

### Things have gotten worse, thanks to big data

1. Large amounts of data of uneven quality

2. Too much trust of objectivity of numbers

2. Black box models

3. Lack of accountability

4. Scaling problems

5. Loss of privacy

<div class="notes">

</div>

### Weapons of Math Destruction

![Figure 6. Cover of Cathy O'Neill's book](http://www.pmean.com/new-images/21/dark-side-06.png)

<div class="notes">

</div>

### Myths about big data

1. Algorithms are objective

2. If you have enough data, quality is no longer an issue

3. We are getting better at this

### No single cause of these problems

1. wrong data

2. Wrong objective

3. Wrong deployment

4. Wrong team

### What they say data science is

```{r setup, echo=FALSE}
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

```{r, fig.width=5, fig.height=5, echo=FALSE}
o <- 0.6
v <- 0.85
draw_circles <- function() {
  dx <- 1
  dy <- 0.866
  i <- seq(0, 2*pi, length=100)
  circle <- data.frame(x=sin(i), y=cos(i))
  circle %>%
    mutate(g=1) %>%
    mutate(x=x-o*dx) %>%
    mutate(y=y-o*dy) -> circle1
  circle %>%
    mutate(g=2) %>%
    mutate(x=x+o*dx) %>%
    mutate(y=y-o*dy) -> circle2
  circle %>%
    mutate(g=3) %>%
    mutate(y=y+o*dy) -> circle3

  circle1 %>%
    bind_rows(circle2) %>%
    bind_rows(circle3) %>%
    ggplot(aes(x, y, group=g)) + 
      geom_path(color=rep(c("red", "darkgreen", "blue"), each=100), size=3) +
      geom_text(x=v*dx, y=-v*dy, size=5, label="Math,\nStat", color="darkgreen") +
      geom_text(x=-v*dx, y=-v*dy, size=5, label="Coding,\nComputers", color="red") +
      geom_text(x=0, y=v*dy, size=5, label="Science,\nMedicine", color="blue") +
      expand_limits(x=c(-(dx+1),(dx+1))) +
      expand_limits(y=c(-(dx+1),(dx+1))) +
      theme_void()
}
o <- 0.6
v <- 0.85
draw_circles() + geom_text(x=0, y=-0.1, label="Data\nScience", size=5)
```

<div class="notes">

</div>

### What data science really is

```{r, fig.width=5, fig.height=5, echo=FALSE}
o <- 0.88
draw_circles() + geom_text(x=0, y=-0.1, label="Data\nScience", size=10)
```

<div class="notes">

</div>

### Why you are needed

1. Too many geeks, not enough scientists
2. More racial, gender diversity is needed

If you are interested, email me: simons@umkc.edu

<div class="notes">

</div>

### Readings

[sim1]: http://new.pmean.com/amazon-bias/
[sim2]: http://new.pmean.com/big-data-bs/
[sim3]: http://new.pmean.com/big-data-goes-to-school/
[sim4]: http://new.pmean.com/lending-false-credibility/
[sim5]: http://new.pmean.com/ai-story/

### 2021-10-30 text/21/test.Rmd
Started work on dark-side.Rmd

---
title: "Test diagrams"
author: "Steve Simon"
source: "New"
date: "2021-10-30"
categories:
- Blog post
tags:
- Graphical display
output: html_document
---

```{r setup, echo=FALSE}
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(magrittr)))
```

```{r, fig.width=5, fig.height=5, echo=FALSE}
o <- 0.6
v <- 0.85
draw_circles <- function() {
  dx <- 1
  dy <- 0.866
  i <- seq(0, 2*pi, length=100)
  circle <- data.frame(x=sin(i), y=cos(i))
  circle %>%
    mutate(g=1) %>%
    mutate(x=x-o*dx) %>%
    mutate(y=y+o*dy) -> circle1
  circle %>%
    mutate(g=2) %>%
    mutate(x=x+o*dx) %>%
    mutate(y=y+o*dy) -> circle2
  circle %>%
    mutate(g=3) %>%
    mutate(y=y-o*dy) -> circle3

  circle1 %>%
    bind_rows(circle2) %>%
    bind_rows(circle3) %>%
    ggplot(aes(x, y, group=g)) + 
      geom_path(color=rep(c("red", "darkgreen", "blue"), each=100), size=3) +
      geom_text(x=v*dx, y=v*dy, size=5, label="Math,\nStat", color="darkgreen") +
      geom_text(x=-v*dx, y=v*dy, size=5, label="Computers,\nCoding", color="red") +
      geom_text(x=0, y=-v*dy, size=5, label="Science,\nMedicine", color="blue") +
      expand_limits(x=c(-(dx+1),(dx+1))) +
      expand_limits(y=c(-(dx+1),(dx+1))) +
      theme_void()
}
o <- 0.6
v <- 0.85
draw_circles()
```

```{r, fig.width=5, fig.height=5, echo=FALSE}
o <- 0.88
draw_circles()
```

### 2021-10-30 src/clean_files.R

@@ -60,8 +60,8 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '<U+00A0>',
-  ' ',
+  'childrensmercy.org/stats',
+  'pmean.com',
   dir_root="text", 
   file_pattern="md$",
   subdir_list=zpad(c(99, 0:21)))

### 2021-11-01 text/21/very-simple-splines.Rmd
Updated very-simple-splines,Rmd

@@
 ---
-title: "A systematic review shows no performance benefit of machine learning over logistic regression for clinical prediction models"
+title: "Very simple spline models"
 author: "Steve Simon"
-source: "http://blog.pmean.com/machine-learning-vs-logistic-regression/"
-date: "2019-03-15"
+date: "2021-10-29"
 categories:
@@
 tags:
-- Data science
-- Logistic regression
-- Systematic overviews
+- Linear regression
 output: html_document
 ---
+
+```{r setup}
+suppressMessages(suppressWarnings(library(broom)))
+suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(magrittr)))
+```
 
@@
 
-```{r, echo=FALSE}
+```{r}
 library(ggplot2)
 x <- 1:20
-y <- c(rep(8, 15), 7:3)+round(rnorm(10), 1)
-d <- data.frame(x, y)
-d
-ggplot(d, aes(x, y)) +
+y <- c(rep(40, 15), 40-(1:5)^2)+round(rnorm(10), 1)
+simulated_example <- data.frame(x, y)
+simulated_example
+ggplot(simulated_example, aes(x, y)) +
   geom_point()
@@
 
+### Fit a linear spline
+
+I want a function that is flat for the values x = 0 to 15, and that drops linearly for x > 15. I also want to function to be continuous (no abrupt jumps).
+
+To fit this function, you need a special X matrix.
+
+```{r}
+linear_spline_ivs <- data.frame(x0=rep(1, 20), x1=c(rep(0, 15), 1:5))
+linear_spline_ivs
+```
+
+Surprisingly, even though the function you are trying to fit is not a linear function of x, you can still fit it with the lm function, as long as you have the correct independent variable matrix.
+
+```{r}
+simple_spline_data <- 
+  data.frame(
+    x1=linear_spline_ivs$x1, 
+    x=simulated_example$x,
+    y=simulated_example$y)
+linear_spline_model <- lm(y~x1, data=simple_spline_data)
+tidy(linear_spline_model)
+```
+
+To get the plot of predictions for this model and future models accurate, you need to create values between the existing values.
+
+```{r}
+x_extra <- seq(0, 20, length=1000)
+data.frame(
+  x=x_extra,
+  x1=pmax(0, x_extra-15),
+  x1sq=pmax(0, x_extra-15)^2) %>%
+  full_join(simulated_example, by="x") -> enhanced_data
+```
+
+```{r}
+linear_spline_predictions <- augment(linear_spline_model, newdata=enhanced_data)
+ggplot(linear_spline_predictions, aes(x, y)) +
+  geom_point() +
+  geom_line(y=linear_spline_predictions$.fitted)
+```
+
+### Fit a quadratic spline
+
+Suppose you want to fit a function which is flat up to a certain point and then declines quadratically. You could do this with the following model.
+
+```{r}
+simple_spline_data %>%
+  mutate(x1sq=x1*x1) -> quadratic_spline_data
+quadratic_spline_model <- lm(y~x1sq, data=quadratic_spline_data)
+tidy(quadratic_spline_model)
+quadratic_spline_predictions <- augment(quadratic_spline_model, newdata=enhanced_data)
+ggplot(quadratic_spline_predictions, aes(x, y)) +
+  geom_point() +
+  geom_line(y=quadratic_spline_predictions$.fitted)
+```
+
+Notice that the quadratic model is smoother. It does not have a sharp elbow at the transition. Smoothness is often preferred in model fitting, though there are exceptions.
+
+### Fit a cubic spline
+
+Here's a final example of very simple splines. I want to display graphically the transition of probabilities in a study of dog walking. Data was collected  of
+
+Source: Wallengren O, Bosaeus I, Frändin K, Lissner L, Falk Erhag H, Wetterberg H, Rydberg Sterner T, Rydén L, Rothenberg E, Skoog I.
+Comparison of the 2010 and 2019 diagnostic criteria for sarcopenia by the European Working Group on Sarcopenia in Older People (EWGSOP) in two cohorts of Swedish older adults.
+BMC Geriatr.
+2021 Oct 26;21(1):600.
+doi: 10.1186/s12877-021-02533-y.
+PMID: 34702174; PMCID: PMC8547086.
+Available in [html format][wal1] or [pdf format][wal2].
+
 ```{r}
 par(mar=rep(0.1,4))
-plot(c(-1,2), 0:1, type="n", axes=FALSE, xlab="", ylab="")
+plot(c(-1,2), c(0, 216), type="n", axes=FALSE, xlab="", ylab="")
+
+polygon(x=c(-1, 0, 0, -1), y=c(   0,   0,  88,  88), col="pink", border=NA)
+polygon(x=c( 2, 1, 1,  2), y=c(   0,   0, 120, 120), col="pink", border=NA)
+polygon(x=c( 0, 1, 1,  0), y=c(   0,   0,  48,  48), col="pink", border=NA)
+
+polygon(x=c(-1, 0, 0, -1), y=c(216, 216,  88,  88), col="lightgreen", border=NA)
+polygon(x=c( 2, 1, 1,  2), y=c(216, 216, 120, 120), col="lightgreen", border=NA)
+polygon(x=c( 0, 1, 1,  0), y=c(216, 216, 160, 160), col="lightgreen", border=NA)
+
 segments(-1,   0, 2,   0)
-segments(-1, 1, 2, 1)
-segments(-1, 0.6, 0, 0.6)
-segments(-1, 0.9, 0, 0.9)
-segments( 1, 0.4, 2, 0.4)
-segments( 1, 0.8, 2, 0.8)
-segments( 0, 0.35, 1, 0.35)
-segments( 0, 0.95, 1, 0.95)
+segments(-1, 216, 2, 216)
+
+segments(-1,  88, 0,  88)
+segments( 1, 120, 2, 120)
+
+segments(-0.5,  48, 0,  48, lty="dotted")
+segments(-0.5, 160, 0, 160, lty="dotted")
+segments( 1.5,  48, 1,  48, lty="dotted")
+segments( 1.5, 160, 1, 160, lty="dotted")
+
 x <-seq(0,1, length=100)
 y <- 3*x^2-2*x^3
-lines(x, 0.35+0.05*y)
-lines(x, 0.5+0.05*y)
-lines(x, 0.5+0.3*y)
-lines(x, 0.6+0.3*y)
-lines(x, 0.6-0.25*y)
-lines(x, 0.65-0.25*y)
-lines(x, 0.65-0.1*y)
-lines(x, 0.85-0.15*y)
-lines(x, 0.85+0.05*y)
-lines(x, 0.9-0.15*y)
-lines(x, 0.9+0.05*y)
-lines(x, 0.9-0.15*y)
-lines(x, 0.95-0.15*y)
+
+for (i in 1:length(x)) {
+  segments(x[i], 160-40*y[i], x[i], 88-40*y[i], col=rgb(floor(colorRamp(c("lightgreen", "pink"))(x[i]))/255), lwd=3)
+}
+
+lines(x,  88 -  40*y)
+lines(x, 160 -  40*y)
+lines(x, 160 +   0*y)
+
+for (i in 1:length(x)) {
+  segments(x[i], 48+72*y[i], x[i], 88+72*y[i], col=rgb(floor(colorRamp(c("pink", "lightgreen"))(x[i]))/255), lwd=3)
+}
+
+lines(x,  48 +   0*y)
+lines(x,  48 +  72*y)
+lines(x,  88 +  72*y)
 ```
@@
 [per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf
+
+[wal1]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8547086/
+[wal2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8547086/pdf/12877_2021_Article_2533.pdf

### 2021-11-01 text/21/cubic-splines.Rmd
Added cubic-splines.Rmd

---
title: "Basics of cubic spline models"
author: "Steve Simon"
date: "2021-11-01"
categories:
- Recommendation
tags:
- Linear regression
output: html_document
---

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(magrittr)))
```

### Artificial data

Suppose you have some data where you suspect the behavior differs for x=1 to 5, x=6 to 10, x=11 to 15, and x=16 to 20.

```{r}
library(ggplot2)
x <- 1:20
y <- rep(40, 20)
y[6:20] <- y[6:20]-3*(1:15)
y[11:20] <- y[11:20]+7*(1:10)
y[16:20] <- y[16:20]-(1:5)^2
y <- y+round(rnorm(10), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

A single cubic function for all the data doesn't quite capture what is happening here.

```{r}
extra_values <- data.frame(x=seq(0, 21, length=2200))
simulated_example %<>% full_join(extra_values, by="x")
```

```{r}
single_cubic_model <- lm(y~poly(x, 3), data=simulated_example)
single_cubic_prediction <- augment(single_cubic_model)
ggplot(single_cubic_prediction, aes(x, y)) + 
  geom_point() +
  geom_line(aes(y=.fitted))
```

So, we could fit a cubic model for the first five data points, for the second five, the third five, and the fourth five. This is a bit much: a cubic model has four parameters, so fitting four of them would use up 16 degrees of freedom in a data set with only 20 observations. But bear with me a bit on this.

```{r}
sb1 <- (simulated_example$x<=5)
sb2 <- (simulated_example$x>5)&(simulated_example$x<=10)
sb3 <- (simulated_example$x>10)&(simulated_example$x<=15)
sb4 <- (simulated_example$x>15)

simulated_example %>%
  filter(sb1) %>%
  lm(y~poly(x, 3), data=.) %>%
  augment(newdata=simulated_example[sb1, ]) -> first_cubic_prediction
lm(y~poly(x, 3), subset=sb2, data=simulated_example) %>% 
  augment(newdata=simulated_example[sb2, ]) %>%
  select(x, .fitted) -> second_cubic_prediction
lm(y~poly(x, 3), subset=sb3, data=simulated_example) %>%
  augment(newdata=simulated_example[sb3, ]) %>% 
  select(x, .fitted) -> third_cubic_prediction
lm(y~poly(x, 3), subset=sb4, data=simulated_example) %>%
  augment(newdata=simulated_example[sb4, ]) %>%
  select(x, .fitted) -> fourth_cubic_prediction

first_cubic_prediction %>%
  bind_rows(second_cubic_prediction) %>%
  bind_rows(third_cubic_prediction) %>%
  bind_rows(fourth_cubic_prediction) %>%
  full_join(simulated_example, by="x") -> combined_prediction
ggplot(combined_prediction, aes(x, y.y)) +
  geom_point() +
  geom_line(aes(y=.fitted))
```

This function is not continuous or smooth. To make a single smooth continuous function you need to define

```{r}
simulated_example %>%
  mutate(x1=x) %>%
  mutate(x2=x^2) %>%
  mutate(x3=x^3) %>%
  mutate(x4=(x>5)*(x-5)^3) %>%
  mutate(x5=(x>10)*(x-10)^3) %>%
  mutate(x6=(x>15)*(x-15)^3) -> spline_basis
```

```{r}
lm(y~x1+x2+x3+x4+x5+x6, data=spline_basis) %>%
  augment(newdata=spline_basis) %>%
  ggplot(aes(x, y)) +
    geom_point() +
  geom_line(aes(y=.fitted))
```

### 2021-11-01 text/21/very-simple-splines.Rmd
Added cubic-splines.Rmd

@@
 
-Here's a final example of very simple splines. I want to display graphically the transition of probabilities in a study of dog walking. Data was collected  of
+Here's a final example of very simple splines.
+There is a "magic" cubic polynomial, $3x^2-2x^3$.
+It rises smoothly from 0 at x=0, to 1 at x=1.
+It is flat (zero first derivative) at x=0 and x=1.
+This is what the function looks like.
+
+```{r}
+x <- seq(0, 1, length=100)
+y <- 3*x^2-2*x^3
+data.frame(x=x, y=y) %>%
+  ggplot(aes(x, y)) +
+    geom_line()
+```
+
+You can add a constant and/or multiply by a constant to get different starting and ending points. So, for example,
+
+$0.75 - 0.5(3x^2-2x^3)$
+
+will start at 0.75 and drop 0.5 units to end at 0.25.
+
+```{r}
+x <- seq(0, 1, length=100)
+y <- 3*x^2-2*x^3
+data.frame(x=x, y=y) %>%
+  mutate(y=0.75-0.5*y) %>%
+  ggplot(aes(x, y)) +
+    geom_line() + 
+    expand_limits(y=0:1)
+```
+
+I will use this function to show transition probabilities in a graph.
+
+In a study of dog walking.
+Data was collected before and during the COVID pandemic.
+A "dog walker" was classified as someone who spent 150 minutes or more per week walking their dog.
+Did the proportion of dog walkers increase or decrease during the pandemic?
 
@@
 Available in [html format][wal1] or [pdf format][wal2].
+
+Here's the data.
+
+```{}
+                  During
+                  Pandemic
+Before
+Pandemic    Dog walker Non-walker   Total
+
+Dog walker      56         72        128  
+Non-walker      40         48         96
+
+Total           96        120        216
+```
+
+Let's convert these to cell percentages.
+
+```{}
+                  During
+                  Pandemic
+Before
+Pandemic    Dog walker Non-walker   Total
+
+Dog walker      26%        33%        59%  
+Non-walker      19%        22%        41%
+
+Total           45%        55%       100%
+```
+
+Vucinic M, Vucicevic M, Nenadovicć K.
+THE COVID-19 PANDEMIC AFFECTS OWNERS WALKING WITH THEIR DOGS.
+J Vet Behav.
+2021 Oct 20.
+doi: 10.1016/j.jveb.2021.10.009.
+Epub ahead of print.
+PMID: 34690614; PMCID: PMC8527592.
+Avialble in [html format][vuc1] or [pdf format][vuc2].
 
@@
 lines(x,  88 +  72*y)
+
+text(-0.95, 49, "No dog walking", srt=90)
+text(-0.95, 157, "Dog walking", srt=90)
+text(-0.5, 210, "Before pandemic")
+text( 1.5, 210, "During pandemic")
+text(-0.7, 48, "41%")
+text(-0.2, 24, "22%")
+text(-0.2, 68, "19%")
+text( 1.2, 24, "22%")
+text( 1.2, 84, "33%")
+text( 1.7, 48, "55%")
+text(-0.7,160, "59%")
+text(-0.2,188, "26%")
+text(-0.2,124, "33%")
+text( 1.2,188, "26%")
+text( 1.2,140, "19%")
+text( 1.7,160, "45%")
 ```
 
-```{r}
-x <- seq(0,1, length=100)
-y <- 3*x^2-2*x^3
-plot(x, y)
-```
+Notice how the 19% slides smoothly upward and the 33% slides smoothly downward.
+So you can see that while some people took up dog walking during the pandemic, this was more than offset by the number who dropped dog walking during the pandemic.
 
@@
 
-[wal1]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8547086/
-[wal2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8547086/pdf/12877_2021_Article_2533.pdf
+[vuc1]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8527592/
+[vuc2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8527592/pdf/main.pdf

### 2021-11-03 text/21/dark-side.Rmd
Finished dark-side.Rmd

@@
 
+<div class="notes">
+
+I should add that I am a bit of a computer geek.
+Here's a list of computer skills that I put on my resume.
+I deliberately made this too small to read so that you wouldn't recognize that half of the computer skills I mention have been obsolete for several deades.
+The computers I used in the 1970s were nothing like today's computers.
+
+</div>
+
+### Worked with health care applications since 1987
+
++ Recent positions
+  + Centers for Disease Control and Prevention (1987-1996)
+  + Children's Mercy Hospital (1996-2008)
+  + UMKC School of Medicine (2008 to present)
++ But...
+  + I am not a doctor
+  + Still confused about many things
+    + Example: Difference between good and bad cholesterol.
+
+<div class="notes">
+
+I've also been working with a variety of health care professionals for many decades.
+I have learned a lot along the way, but I am not a doctor.
+Not an MD doctor anyway.
+When I talk about medical examples, keep that in mind.
+I don't always describe things accurately from a mdedical perspective, and I'm always forgetting which one is the bad cholesterol.
+
+Is it ldl or hdl? Does anyone know?
+
+</div>
+
 ### First poll question
@@
 
-Think about your high school algebra class.
+I want to get a quick feel for your background and interests.
+Here's a quote from a romantic comedy starring Kathleen Turner from 1986.
+A forty year old woman, played by Kathleen Turner, travels back in time to her high school senior year, 1960.
+She has an amusing interchange with her high school math teacher.
+
+"I happend to know that in the future, I will not have the slightest use for algebra, and I speak from experience."
+
+Think back to your high school algebra class.
 
@@
 
-### But numbers still have value
+### But numbers still have value. Example: Quality of Life
+
+![Figure 7. First few questions from SF-36 form](http://www.pmean.com/new-images/21/dark-side-07.png)
+
+<div class="notes">
 
 Recognizing that Statistics are a social construction does not mean that Statistics are worthless and that experiments are a waste of time.
-
-### Example: Quality of Life
-
-![Figure 7. First few questions from SF-36 form](http://www.pmean.com/new-images/21/dark-side-08.png)
-
-<div class="notes">
+The best example I can give for this is the measure of quality of life.
 
@@
 A friend of mine said, though, that she couldn't understand why you would want to run, unless you were being chased.
-One the other hand, I would not miss a thing if I did not push a vacuum, bowl, or play golf.
+One the other hand, I would not miss a thing if I did not push a vacuum.
 
@@
 
+Problems occur when these social constructs are developed by research teams with limited diversity (they may have blind spots) or if the social constructs are not evaluated in diverse populations.
+
 But we can still use social constructs like quality of life, as long as we do it carefully.
+
+In fact, we should use these measures because they get at information which is important, and which cannot be measured by a laboratory.
+
+There's a lot of uncertainty about treatments for cancer, for example, which can sometimes extend a person's lifetime, but can also substantially interfere with their ability to do things that they feel are important.
+You can't discuss this trade-off without first measuring the deterioration of quality of life associated with some of these aggressive therapies.
 
@@
 
+Once you recognize that Statistics are a social construct, you can take steps to make sure that these statistics are used properly.
+You need to think about the process the generates the statistics.
+Is there a "night watchman" who puts down whatever he pleases?
+
+Just as an example, I do a lot of work with electronic health records.
+The health care professionals who populate the data in a system like Cerner or Epic are not doing this because they want to make sure that I have something interesting to publish.
+They have two objectives.
+First, they want to make sure that the information they put in is helpful to other members of the health care team.
+Second, they want to make sure they get the maximum possible billing from the insurance companies.
+Third, they want to wrap this up quickly so they can get home to their families.
+
+Second, identify possible biases. Does the desire to maximize billing skew the results?
+
+Third, recognize limitations. Am I, as a researcher, asking for information that is not important for documenting the care a patient receives to other members of the treatment team?
+
+Fourth, beware of confirmation bias.
+Confirmation bias is the tendency to look harder for information that reinforces our preconceived notions, or to overlook or downweight information that contradicts our view of the world.
+
+Fifth, avoid nihilistic thinking.
+Once you recognize that Statistics are a social construct, you have to work harder, but it is not an impossible task.
+Our tendency to treat Statistics as hard and objective has led us astray at times, but we've still made a lot of progress in science and medicine, thanks to Statistics.
+
 </div>
@@
 1. Large amounts of data of uneven quality
-
-2. Too much trust of objectivity of numbers
 
@@
 
+Now everything I've mentioned so far is a problem in Statistics that has been around since before I was born.
+But we are in a new era, the era of big data.
+We have data today at a size and a level of detail that was unavailable twenty years ago.
+This is thanks to the internet, the human genome project, and many other things.
+
+I'm glad we have all these data, because it increases the opportunities for learning. But at the same time, the flaws of empiricism have gotten worse.
+
+Part of the problem is that the massive data sets that we use today have a great unevenness of quality.
+it is not too hard to identify obvious typos in a data set with a few hundred observations, but you would never been able to spot them in a data set with a few million observations.
+
+Second, big data requires more complex data analysis methods and these methods are often difficult to understand.
+Not just from a how do these new methods work, but also in what aspects of the data are emphasized and what aspects of the data are ignored.
+A small model that has ten inputs is easily taken apart and examined, but you can't do that with a model that has ten thousand inputs.
+
+Third, there is a lack of accountability. These complex models are rarely audited, for example, to look for hidden biases.
+
+Fourth, big data provides an opportunity to produce answers not just for a small number of people, but for thousands or millions of them all at once.
+If the big data model is flawed, the flaws are multiplied across the large number of people who are affected by the big data model.
+
+Fifth, there is so much data out there that you can often infer quite intimate and private details about individuals, often without them being aware of this.
+
+</div>
+
+### Small group exercises
+
++ Form groups of 4-5 people.
++ Answer the following questions
+  + Who is the villain?
+  + Who is the victim?
+  + How was the victim harmed?
+  + What could have prevented this?
+  + Did anything surprise you?
+  + Did you disagree with anything in the article?
+  + Is there a single quote from the article that summarizes it well?
+
+<div class="notes">
+
+I want to do a small group exercise where you look at a newspaper article that discusses harm that occurred thanks to computer models.
+Do this in groups of 4 to 5 people.
+Read the article quicly.
+It should not take more than 5 minutes to read.
+Then designate someone as a note taker and a different individual as the spokesperson for the group.
+Come up with answers for the questions listed here.
+
+Given the size of the class, we may not have the opportunity to let every group speak, but I will try to spread things around as much as possible.
+
 </div>
@@
 
-![Figure 6. Cover of Cathy O'Neill's book](http://www.pmean.com/new-images/21/dark-side-06.png)
+![Figure 8. Cover of Cathy O'Neill's book](http://www.pmean.com/new-images/21/dark-side-06.png)
 
@@
 
+I want to discuss one example in detail. It comes from chapter 1 of the book, Weapons of Math Destruction, by Cathy O'Neil.
+
 </div>
+
+### First villain
+
++ Walter Quijano
+  + Provided testimony on recidivism rates at seven trials
+  + Unfairly included race in his calculations and testimony
+  + Six of seven convictions later overturned
+
+<div class="notes">
+
+The big data example in O'Neil's book starts with the story of a person, Walter Quijano, who provided testimony in seven trials about the risk of recidivism.
+Recidivism is the unfortunate even where someone who has been released from prison commits a new crime and gets tossed back in jail.
+Mr.
+Quijano gave overtly biased recommendations at these trials that incorporated race into his assessment of recidivism.
+This was noted in the appeals of one of the defendants, and led to six of the seven convictions being overturned.
+
+</div>
+
+### Second villain
+
++ LSI-R questionnaire
+  + Given to thousands of inmates
+  + Classifies risk of recidivism
+  + Does not explicitly ask about race
+  + But does have "leading" questions
+
+<div class="notes">
+
+It contrasts this individual with a big data prediction model based on the LSI-R questionnaire.
+LSI-R stands for Level of Service Inventory-Revised that asks a bunch of questions like "How many prior convictions have you had?" This questionnaire weights the responses to various questions to provide an estimate of the risk of recidivism.
+Although the questionnaire does not explicitly ask about race, it can still indirectly identify a person's race by their pattern of responses to a series of leading questions.
+
+</div>
+
+### Victims
+
++ Inmates at parole hearings
++ Defendants at trial sentencing
+
+<div class="notes">
+
+This questionnaire was used at a large number of parole hearings in multiple states and at sentencing hearings in a few states as well.
+
+</div>
+
+### How were the victims harmed
+
++ Disproportionate recidivism risks by race
+  + Fewer paroles grants
+  + Longer sentences
++ No avenue to appeal
+  + Model presumed to be unbiased
+  + Complexity prevents examination of bias
++ Scale issues
+  + Walter Quijano harmed 7 defendants
+  + LSI-R harmed thousands of defendants.
+
+<div class="notes">
+
+The model was shown to grossly overstate the recidivism risk of black defendants.
+This led to fewer grants of parole to black prisoners and longer prison sentences for newly convicted black defendants.
+
+There was no avenue of appeal for those harmed by the LSI-R.
+You can't cross-examine an algorithm at trial, and the company that developed the LSI-R model did not want to disclose details about how the model worked.
+It was a trade secret and revealing its details would allow other companies to steal their technology.
+To be honest, the model itself was so complex that the company would have been unable to reveal its inner workings, even if it wanted to.
+
+The other important is issue is the scale of the harm.
+A biased witness could only influece a handful of defendants.
+This algorithm, however, was applied to many people.
+Walter Quijano had only enough time and energy to taint the sentencing of seven defendants, but the impact of the LSI-R questionnaire ended up influencing thousands.
+
+That's a hallmark of big data models. They are expensive to build and very labor intensive, but once they are built, it costs almost nothing to deploy it broadly.
+
+</div>
+
+### What could have prevented this
+
++ Insist on transparency
++ Test the model for bias
++ Build the model with better objective
+
+### Did anything surprise you?
+
++ Questionnaire includes questions that would be inadmissable if they were asked during a normal trial
+  + When was the first time you were ever involved with the police?
+  + Do any of your friends or relatives have a criminal record?
+
+<div class="notes">
+
+I read about this a couple of years ago, but I think the thing that surprised me was the number of questions on the LSI-R that would not have been allowed in an open trial, because the questions would have been ruled as prejudicial.
+
+But apparently, you can ask prejudicial questions outside of a courtroom in a questionnaire and then use that information to create a recidivism risk score.
+
+</div>
+
+### Did you disagree with anything in the article?
+
++ No
+
+### Is there are single quote that summarizes the article well
+
+"The questionnaire includes circumstances of a criminal's birth and upbringing, including his or her family, neighborhood, and friends.
+These details should not be relevant to a criminal case or to the sentencing."
+
+### A Face Is Exposed for AOL Searcher No. 4417749
+
+![Figure 9. First page of newspaper article](http://www.pmean.com/new-images/21/aol-searcher-01.png)
+
+### How a 'NULL' License Plate Landed One Hacker in Ticket Hell
+
+![Figure 10. First page of newspaper article](http://www.pmean.com/new-images/21/null-license-plate-01.png)
+
+### How Bright Promise in Cancer Testing Fell Apart
+
+![Figure 11. First page of newspaper article](http://www.pmean.com/new-images/21/bright-promise-01.png)
 
@@
 3. We are getting better at this
+
+<div class="notes">
+
+If there are any big lessons from this, they are that you should be skeptical about big data.
+Algorithms are not objective.
+They are a social construct.
+Also, don't believe that bigger datasets are always better.
+A lot of data does not compensate for poor quality.
+Finally, while I am a big believer in progress, I think that too many of us are making mistakes and those mistakes are a lot worse when the data is big.
+
+</div>
 
@@
 4. Wrong team
+
+<div class="notes">
+
+There is no single cause for these problems. They are built into the system and it takes a lot of effort to work these biases out of the system.
+
+</div>
 
@@
 
+Here's a classic Venn diagram that you will see in pretty much any article that talks about big data or data science (I use those words interchangeably).
+
+There are three areas of expertise that you need as a data scientist: knowledge of mathematics and statistics, skill with computer science and coding, and expertise in medical or scientific specialties.
+
 </div>
@@
 <div class="notes">
+
+A more accurate picture would show that there is not enough overlap among these disciplines.Not that many people are skilled at both science and statistics.
+Fewer stil know both of those skills well and can program effectively.
+
+That's actually good news from your perspective. You don't have to be a genius in all three areas. Be good in one area and have an appreciation of the the other two and you'll do just fine.
 
@@
 
+I'd like to end with an exhortation.
+We need people like you in data science.
+It's okay to spend most of your time learning the complexities of medicine, but some of you have a bit of talent and aptitude in math/stat and in computers/coding.
+That's an advantage your generation has that my generation did not.
+So you have the ability to appreciate the full picture more so than people my age.
+A few of us knew computers well, but today's generation is saturated in computational experience.
+
+Second, these research teams need a lot more racial and gender diversity.
+I try really hard to appreciate the viewpoints of people different than me, bit I can't understand it at a level of someone who has seen life from a different perspective.
+If the only people developing these data science models are white males, then the models will continue to be biased against women and minorities.
+
 </div>
-
-### Readings
-
-[sim1]: http://new.pmean.com/amazon-bias/
-[sim2]: http://new.pmean.com/big-data-bs/
-[sim3]: http://new.pmean.com/big-data-goes-to-school/
-[sim4]: http://new.pmean.com/lending-false-credibility/
-[sim5]: http://new.pmean.com/ai-story/

### 2021-11-12 text/15/difference-of-two-exponentials.Rmd
Finished all but one 2015 page

---
title: "PMean: Nonlinear regression for the difference of two exponentials"
author: "Steve Simon"
source: "http://blog.pmean.com/difference-of-two-exponentials/"
date: "2015-10-05"
category: 
- Blog post
tags: 
- incomplete page
- nonlinear regression
- R software
output: html_document
---

I wanted to provide an overview of how you analyze a classic nonlinear regression model. It is a difference of two exponential functions. This nonlinear function is used commonly in pharmocokinetic models and is a simply way to model the oral administration of a drug. I want to show how the model works in a mathematical sense and then how you fit it using R.

<!---More--->

Here are some simple examples of nonlinear regression. We will use the built-in data set Theoph for all our examples.

```{r}
head(Theoph)
```

```{r}
tail(Theoph)
```

### look at just the first patient.

You can find an [earlier version][sim1] of this page on my [blog][sim2].

[sim1]: http://blog.pmean.com/difference-of-two-exponentials/
[sim2]: http://blog.pmean.com

### 2021-11-12 text/15/missing-values-in-r.Rmd
Finished all but one 2015 page

---
title: "PMean: Missing values in R talk"
author: "Steve Simon"
source: "http://blog.pmean.com/missing-values-in-r/"
date: "2015-02-07"
category: Blog post
tags: R software
output: html_document
---

I'm talking a bit about missing values in R this afternoon for the
Kansas City R Users Group. Here is what I'll be talking
about.

<!---More--->

When you import data into R, the first thing you should do is

KNOW YOUR COUNT

This means that you need to know how many rows are in your data, and for
all of the key variables, how many missing values you have. In a nested
or hierarchical model, you need to also know the counts at each level of
nesting or each level of the hierarchy.

Here is a data set, freely available on the web, that shows sleep
patterns and other statistics on various mammals. It has a handful of
missing values and can help illustrate how you might tackle a new
dataset. You can find a description of this data set in [html format][sta1].

```{r}
sl <- read.table(file="http://www.statsci.org/data/general/sleep.txt",header=TRUE)
dim(sl)
```

```{r}
names(sl)
```

This data set has 62 rows and 11 variables. Since the data set doesn't
have too many variables, let's print out the first few and last few
rows.

```{r}
head(sl)
```

```{r}
tail(sl)
```

The summary function is useful, as it will display percentiles and the
mean for numeric variables, but also (very importantly) display the
number of missing values.

For factors, it will display a table of the first seven most frequently
occuring values

```{r}
summary(sl)
```

The first variable, Species, has 62 unique values. You can tell this by
the fact that the first seven most frequently occurring values all had
frequency 1. There are several other ways to check for uniqueness.

```{r}
table(table(sl$Species))
```

```{r}
length(unique(sl$Species))
```

    ## [1] 62

The table function does not show NAs, at least with the default option

```{r}
table(sl$Gestation)
```

You would notice this if you added a total to the end of the table.

```{r}
addmargins(table(sl$Gestation))
```

The total is only 58, which is less than the magic number of 62. This
means that there are 4 missing values.

You can also force the table to show counts for missing values.

```{r}
table(sl$Gestation,useNA="always")
```

The default option for useNA is "never" and a third option is "ifany".
Notice how "ifany" differs from "always"

```{r}
table(sl$Gestation,useNA="ifany")
```

```{r}
table(sl$Danger,useNA="ifany")
```

```{r}
table(sl$Danger,useNA="always")
```

Many other functions in R offer multiple ways to handle NAs.

```{r}
mean(sl$Gestation,na.rm=FALSE)
```

```{r}
mean(sl$Gestation,na.rm=TRUE)
```

```{r}
mod1 <- lm(TotalSleep~Gestation,data=sl,na.action=na.omit)
summary(mod1)
```

The alternate option for na.action in the lm function is na.fail, but I
cannot run this directly in knitr, because the resulting error message
causes knitr to stop.

Logic values involving NAs are a bit tricky. You cannot get at NAs
directly using the == comparison operator.

```{r}
which(sl$Gestation==NA)
```

    ## integer(0)

What's happening here is that most logic comparisons invloving NA will
evaluate not to TRUE or FALSE, but rather to NA.

If you want to select missing values, you must use the is.na function.

```{r}
which(is.na(sl$Gestation))
```

    ## [1] 13 19 20 56

The is.finite function represents "opposite" function, though there are
a few subtle differences.

```{r}
which(is.finite(sl$Gestation))
```

    ##  [1]  1  2  3  4  5  6  7  8  9 10 11 12 14 15 16 17 18 21 22 23 24 25 26
    ## [24] 27 28 29 30 31 32 33 34 35 36 37 38 39 40 41 42 43 44 45 46 47 48 49
    ## [47] 50 51 52 53 54 55 57 58 59 60 61 62

Logic involving missing values can seem at times a bit illogical.

```{r}
sl$Gestation==NA
```

    ##  [1] NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA
    ## [24] NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA
    ## [47] NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA NA

```{r}
sl$Gestation==645
```

    ##  [1]  TRUE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
    ## [12] FALSE    NA FALSE FALSE FALSE FALSE FALSE    NA    NA FALSE FALSE
    ## [23] FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
    ## [34] FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
    ## [45] FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE FALSE
    ## [56]    NA FALSE FALSE FALSE FALSE FALSE FALSE

```{r}
tst.na <- NA
tst.na==tst.na
```

    ## [1] NA

Why would NA==NA not evaluate to TRUE? You could argue that a missing
value might be 3 or 7 or 923.427, so you would never know if it equaled
something else, even another missing value.

This produces a three valued logic system.

```{r}
tst.tvl <- c(TRUE,FALSE,NA)
not.table <- !tst.tvl
names(not.table) <- tst.tvl
not.table
```

    ##  TRUE FALSE  <NA> 
    ## FALSE  TRUE    NA

```{r}
and.table <- outer(tst.tvl,tst.tvl,"&")
dimnames(and.table) <- list(tst.tvl,tst.tvl)
and.table
```

    ##        TRUE FALSE  <NA>
    ## TRUE   TRUE FALSE    NA
    ## FALSE FALSE FALSE FALSE
    ## <NA>     NA FALSE    NA

```{r}
or.table <- outer(tst.tvl,tst.tvl,"|")
dimnames(or.table) <- list(tst.tvl,tst.tvl)
or.table
```

Other programs, such as SAS, use a two valued logic system. What SAS
does is to set missing values equal to negative infinity. This seems
simpler because it allows you to keep a two-valued logic system. But it
has its own problems.

Note how R handles the less than comparison operator with missing
values.

```{r}
sl$Gestation<50
```

In SAS, this comparison would evaluate to TRUE for the four missing
values.

The only way to select missing values is with the is.na function.

There are three more values in R that are closely related to missing.

```{r}
1/0
```

```{r}
-1/0
```

```{r}
sqrt(-1)
```

The difference between Inf and NaN is in how comparisons are made.

```{r}
Inf>3
```

```{r}
NaN>3
```

```{r}
Inf==Inf
```

```{r}
NaN==NaN
```

R allows you to make a few computations using Inf, but others are not
allowed.

```{r}
Inf+1
```

```{r}
1/Inf
```

```{r}
0*Inf
```

```{r}
Inf-Inf
```

It's also worth noting that you can force R to use complex numbers for
the square root of -1, although this is not the default.

```{r}
sqrt(as.complex(-1))
```

The is.na and is.finite functions are not perfect opposites.

```{r}
tst.cases <- c(3,NA,NaN,Inf,-Inf)
is.na.table <- is.na(tst.cases)
names(is.na.table) <- tst.cases
is.na.table
```

```{r}
is.finite.table <- is.finite(tst.cases)
names(is.finite.table) <- tst.cases
is.finite.table
```

You can find an [earlier version][sim1] of this page on my [blog][sim2].

[sim1]: http://blog.pmean.com/missing-values-in-r/
[sim2]: http://blog.pmean.com

[sta1]: http://www.statsci.org/data/general/sleep.html

### 2021-11-12 text/15/relationships-in-r.Rmd
Finished all but one 2015 page

---
title: "PMean: Examining relationships in R"
author: "Steve Simon"
source: "http://blog.pmean.com/relationships-in-r/"
date: "2015-04-03"
categories:
- Blog post
tags:
- R software
output: html_document
---

I'm giving a talk for the Kansas City R Users Group on how to get a
preliminary impression of relationships between pairs of variables. Here
is the R code and output that I will use.

<!---More--->

### Simple measures of association

There are several different ways of measuring bivariate relationships in
a descriptive fashion prior to data analysis. The methods can be largely
grouped into measures of relationship between two continuous variables,
two categorical variables and measures of a relationship between a
categorical variable and a continuous variable.

```{r}
fn <- "http://lib.stat.cmu.edu/DASL/Datafiles/homedat.html"
h0 <- readLines(fn)
n <- length(h0)
print(n)
```

    ## [1] 165

```{r}
head(h0)
```

```{r}
tail(h0)
```

```{r}
the.data.line <- grep("The Data:",h0)
print(the.data.line)
```

```{r}
h1 <- h0[(the.data.line+4):(n-1)]
head(h1)
```

```{r}
tail(h1)
```

```{r}
h2 <- strsplit(h1,"\t")
head(h2)
```

```{r}
tail(h2)
```

```{r}
h3 <- as.numeric(unlist(h2))
```

```{r}
head(h3)
```

```{r}
tail(h3)
```

```{r}
h4 <- matrix(h3,ncol=8,byrow=TRUE)
head(h4)
```

```{r}
tail(h4)
```

```{r}
var.names.line <- the.data.line+2
print(h0[var.names.line])
```

```{r}
var.names <- strsplit(h0[var.names.line]," ")
print(var.names)
```

```{r}
var.names <- unlist(var.names)
print(var.names)
```

```{r}
var.names <- var.names[nchar(var.names)>0]
print(var.names)
```

```{r}
var.names <- tolower(var.names)
print(var.names)
```

```{r}
dimnames(h4)[[2]] <- var.names
head(h4)
```

```{r}
tail(h4)
```

```{r}
h5 <- data.frame(h4)
head(h5)
```

```{r}
tail(h5)
```

The best graphical summary of two continuous variables is a scatterplot.
You should include a smoothing curve or spline model to the graph to
emphasize the general trend and any departures from linearity.

```{r}
plot(h5sqft,h5$price)
lines(lowess(h5price~h5$sqft))
```

```{r}
plot(h5age,h5$price)
sb <- is.finite(h5age)
lines(lowess(h5price[sb]~h5$age[sb]))
```

The best numeric summary of two continuous variables is a correlation
coefficient.

```{r}
cor(h5[,c("price","sqft","age")],use="pairwise.complete.obs")
```

Correlations should always be rounded to two or maybe even just one
significant digit.

```{r}
round(cor(h5[,c("price","sqft","age")],use="pairwise.complete.obs"),1)
```

Anything larger than 0.7 or smaller than -0.7 is a strong linear
relationship. Anything between 0.3 and 0.7 or between -0.3 and -0.7 is a
weak linear relationship. Anything between -0.3 and 0.3 represents
little or no linear relationship.

The best graphical summary between a continuous variable and a
categorical variable is a boxplot.

```{r}
boxplot(h5price~h5$feats)
```

```{r}
boxplot(h5price~h5$ne)
```

```{r}
boxplot(h5price~h5$cust)
```

```{r}
boxplot(h5price~h5$cor)
```

If your categorical variable is binary, you can also use a scatterplot.
The binary variable goes on the y axis and a trend line is critical.

```{r}
plot(h5price,h5$ne)
lines(lowess(h5ne~h5$price))
```

```{r}
plot(h5price,h5$cust)
lines(lowess(h5cust~h5$price))
```

```{r}
plot(h5price,h5$cor)
lines(lowess(h5cor~h5$price))
```

You can also compute a correlaton between a binary variable and a
categorical variable. It is equivalent to the point-bisearial
correlation.

```{r}
round(cor(h5[,c("ne","cust","cor")],h5price),1)
```

Let's save the display of a relationship involving two categorical
variables until another day.

### 2021-11-15 text/15/missing-values-in-r.Rmd
Added 2015.md

@@
 date: "2015-02-07"
-category: Blog post
-tags: R software
+category:
+- Blog post
+tags:
+- R software
 output: html_document

### 2021-11-15 text/21/cubic-splines.Rmd
Added 2015.md

@@
   mutate(x1=x) %>%
+  mutate(x4=(x>5)*(x-5)) %>%
+  mutate(x5=(x>10)*(x-10)) %>%
+  mutate(x6=(x>15)*(x-15)) -> linear_spline_basis
+```
+
+```{r}
+lm(y~x1+x4+x5+x6, data=linear_spline_basis) %>%
+  augment(newdata=linear_spline_basis) %>%
+  ggplot(aes(x, y)) +
+    geom_point() +
+  geom_line(aes(y=.fitted))
+```
+
+```{r}
+simulated_example %>%
+  mutate(x1=x) %>%
+  mutate(x2=x^2) %>%
+  mutate(x4=(x>5)*(x-5)^2) %>%
+  mutate(x5=(x>10)*(x-10)^2) %>%
+  mutate(x6=(x>15)*(x-15)^2) -> quadratic_spline_basis
+```
+
+```{r}
+lm(y~x1+x4+x5+x6, data=quadratic_spline_basis) %>%
+  augment(newdata=quadratic_spline_basis) %>%
+  ggplot(aes(x, y)) +
+    geom_point() +
+  geom_line(aes(y=.fitted))
+```
+
+This function is not continuous or smooth. To make a single smooth continuous function you need to define
+
+```{r}
+simulated_example %>%
+  mutate(x1=x) %>%
   mutate(x2=x^2) %>%

### 2021-11-15 src/clean_files.R

@@ -60,9 +60,10 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'childrensmercy.org/stats',
-  'pmean.com',
+  'tags:',
+  'tags:\n- 2015-pages',
   dir_root="text", 
   file_pattern="md$",
-  subdir_list=zpad(c(99, 0:21)))
+  subdir_list=15 # zpad(c(99, 0:21))
+)
 

### 2021-11-16 text/15/relationships-in-r.Rmd
Work on 15/relationships-in-r.Rmd
Added how-many-pages.md

@@
 ```{r}
-fn <- "http://lib.stat.cmu.edu/DASL/Datafiles/homedat.html"
-h0 <- readLines(fn)
-n <- length(h0)
-print(n)
-```
-
-    ## [1] 165
-
-```{r}
-head(h0)
-```
-
-```{r}
-tail(h0)
-```
-
-```{r}
-the.data.line <- grep("The Data:",h0)
-print(the.data.line)
-```
-
-```{r}
-h1 <- h0[(the.data.line+4):(n-1)]
-head(h1)
-```
-
-```{r}
-tail(h1)
-```
-
-```{r}
-h2 <- strsplit(h1,"\t")
-head(h2)
-```
-
-```{r}
-tail(h2)
-```
-
-```{r}
-h3 <- as.numeric(unlist(h2))
-```
-
-```{r}
-head(h3)
-```
-
-```{r}
-tail(h3)
-```
-
-```{r}
-h4 <- matrix(h3,ncol=8,byrow=TRUE)
-head(h4)
-```
-
-```{r}
-tail(h4)
-```
-
-```{r}
-var.names.line <- the.data.line+2
-print(h0[var.names.line])
-```
-
-```{r}
-var.names <- strsplit(h0[var.names.line]," ")
-print(var.names)
-```
-
-```{r}
-var.names <- unlist(var.names)
-print(var.names)
-```
-
-```{r}
-var.names <- var.names[nchar(var.names)>0]
-print(var.names)
-```
-
-```{r}
-var.names <- tolower(var.names)
-print(var.names)
-```
-
-```{r}
-dimnames(h4)[[2]] <- var.names
-head(h4)
-```
-
-```{r}
-tail(h4)
-```
-
-```{r}
-h5 <- data.frame(h4)
-head(h5)
-```
-
-```{r}
-tail(h5)
+suppressMessages(suppressWarnings(library(tidyverse)))
+fn <- "https://raw.githubusercontent.com/pmean/introduction-to-SAS/master/data/housing.txt"
+home <- read_tsv(fn, na=".")
+home
 ```
@@
 ```{r}
-plot(h5sqft,h5$price)
-lines(lowess(h5price~h5$sqft))
+plot(home$sqft,home$price)
+lines(lowess(home$price~home$sqft))
 ```
@@
 ```{r}
-plot(h5age,h5$price)
-sb <- is.finite(h5age)
-lines(lowess(h5price[sb]~h5$age[sb]))
+sb <- is.finite(home$age)
+plot(home$age[sb],home$price[sb])
+lines(lowess(home$price[sb]~home$age[sb]))
 ```
@@
 ```{r}
-cor(h5[,c("price","sqft","age")],use="pairwise.complete.obs")
+cor(home[,c("price","sqft","age")],use="pairwise.complete.obs")
 ```
@@
 ```{r}
-round(cor(h5[,c("price","sqft","age")],use="pairwise.complete.obs"),1)
+round(cor(home[,c("price","sqft","age")],use="pairwise.complete.obs"),1)
 ```
@@
 ```{r}
-boxplot(h5price~h5$feats)
+boxplot(home$price~home$features)
 ```
@@
 ```{r}
-boxplot(h5price~h5$ne)
+boxplot(home$price~home$northeast)
 ```
@@
 ```{r}
-boxplot(h5price~h5$cust)
+boxplot(home$price~home$custom_build)
 ```
@@
 ```{r}
-boxplot(h5price~h5$cor)
+boxplot(home$price~home$corner_lot)
 ```
@@
 ```{r}
-plot(h5price,h5$ne)
-lines(lowess(h5ne~h5$price))
+plot(home$price,home$northeast)
+lines(lowess(home$northeast~home$price))
 ```
@@
 ```{r}
-plot(h5price,h5$cust)
-lines(lowess(h5cust~h5$price))
+plot(home$price,home$custom_build)
+lines(lowess(home$custom_build~home$price))
 ```
@@
 ```{r}
-plot(h5price,h5$cor)
-lines(lowess(h5cor~h5$price))
+plot(home$price,home$corner_lot)
+lines(lowess(home$corner_lot~home$price))
 ```
@@
 ```{r}
-round(cor(h5[,c("ne","cust","cor")],h5price),1)
+round(cor(home[,c("northeast","custom_build","corner_lot")],home$price),1)
 ```

### 2021-11-19 text/21/dark-side.Rmd
Moved notes pages back on dark-side.Rmd
Started work on nhanes-overview.md

@@
 - Incomplete page
-output: powerpoint_presentation
+output: html_document
 ---
@@
 
+On the next couple of slides, I want to show a classic Venn diagram and my variation on it to illustrate an important point.
+
+Here's a classic Venn diagram that you will see in pretty much any article that talks about big data or data science (I use those words interchangeably).
+
+There are three areas of expertise that you need as a data scientist: knowledge of mathematics and statistics, skill with computer science and coding, and expertise in medical or scientific specialties.
+
+A more accurate picture would show that there is not enough overlap among these disciplines.Not that many people are skilled at both science and statistics.
+Fewer stil know both of those skills well and can program effectively.
+
+That's actually good news from your perspective. You don't have to be a genius in all three areas. Be good in one area and have an appreciation of the the other two and you'll do just fine.
+
 </div>
@@
 
-<div class="notes">
-
-Here's a classic Venn diagram that you will see in pretty much any article that talks about big data or data science (I use those words interchangeably).
-
-There are three areas of expertise that you need as a data scientist: knowledge of mathematics and statistics, skill with computer science and coding, and expertise in medical or scientific specialties.
-
-</div>
-
 ### What data science really is
@@
 ```
-
-<div class="notes">
-
-A more accurate picture would show that there is not enough overlap among these disciplines.Not that many people are skilled at both science and statistics.
-Fewer stil know both of those skills well and can program effectively.
-
-That's actually good news from your perspective. You don't have to be a genius in all three areas. Be good in one area and have an appreciation of the the other two and you'll do just fine.
-
-</div>
 

### 2021-11-19 src/clean_files.R

@@ -60,10 +60,10 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'tags:',
-  'tags:\n- 2015-pages',
+  'powerpoint_presentation',
+  'html_document',
   dir_root="text", 
   file_pattern="md$",
-  subdir_list=15 # zpad(c(99, 0:21))
+  subdir_list=zpad(c(99, 0:21))
 )
 

### 2021-11-20 src/sweep_files.R
Update nhanes-overview.md

@@ -1,0 +1,52 @@
+# sweep_files.R. This program was written by Steve Simon on 2020-05-24.
+#
+# This program removes non .md and .Rmd files from the text folder.
+
+suppressMessages(suppressWarnings(library(base64enc)))
+suppressMessages(suppressWarnings(library(curl     )))
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+v <- TRUE
+
+sweep_files <- function(
+    dir_root="text", 
+    subdir_list=c("99", zpad(0:21), "no"),
+    file_pattern="*\\.md$", 
+    verbose=TRUE) {
+  if (!exists("ok_to_move")) ok_to_move <- FALSE 
+  k <- 0
+  for (subdir in subdir_list) {
+    list.files(dir_root %s% subdir) %>%
+      str_subset(file_pattern, negate=TRUE) -> file_list
+    "\nSearching through" %b% 
+      dir_root %s% 
+      subdir -> message
+    if (verbose) cat(message)
+    for (i_file in file_list) {
+      k <- k+1
+      "\n\n" %0% i_file %>% br %>% cat
+      }
+  }
+  "\n\nTotal files" %b% 
+    ifelse(ok_to_move, "moved", "to be moved") %b%
+    k %>% cat
+}
+
+verbose <- TRUE
+verbose <- FALSE
+ok_to_move <- TRUE
+ok_to_move <- FALSE
+sweep_files(
+  dir_root="text", 
+  file_pattern="md$",
+  subdir_list=zpad(c(99, 0:21))
+)
+

### 2021-11-20 text/08/declining-count-model.Rmd
Removed backslashes

--- 2021-09-30 text/Rmd/Declining-count-08/index.Rmd
@@
 ---
-I\'ve been working on an interesting project that requires Poisson
+I've been working on an interesting project that requires Poisson
 regression. A company sends out a mailing and gets a certain number of
@@
 
-Here\'s an example of this data
+Here's an example of this data
 
@@
 
-For a=8.2 and r=0.58, this equals 19.5. Since we\'ve already seen 19
+For a=8.2 and r=0.58, this equals 19.5. Since we've already seen 19
 events, we would expect to see half an event on average for the

### 2021-11-20 src/clean_files.R

@@ -60,10 +60,10 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'powerpoint_presentation',
-  'html_document',
+  'zzz',
+  'yyy',
   dir_root="text", 
   file_pattern="md$",
-  subdir_list=zpad(c(99, 0:21))
+  subdir_list=zpad(c(99, 0:8))
 )
 

### 2021-11-23 text/21/cubic-splines.Rmd
Added b-splines to cubic-splines.Rmd

--- 2021-11-01 text/21/cubic-splines.Rmd
@@
 suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(knitr)))
 suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(splines)))
+suppressMessages(suppressWarnings(library(tidyr)))
 ```
@@
 library(ggplot2)
-x <- 1:20
-y <- rep(40, 20)
-y[6:20] <- y[6:20]-3*(1:15)
-y[11:20] <- y[11:20]+7*(1:10)
-y[16:20] <- y[16:20]-(1:5)^2
+x <- 0:20
+y <- rep(40, 21)
+y[6:21] <- y[6:21]-3*(1:15)
+y[11:21] <- y[11:21]+7*(1:10)
+y[16:21] <- y[16:21]-(1:5)^2
 y <- y+round(rnorm(10), 1)
@@
 
-A single cubic function for all the data doesn't quite capture what is happening here.
+To make the graphs look nice, you need to include some intermediate values.
 
 ```{r}
-extra_values <- data.frame(x=seq(0, 21, length=2200))
-simulated_example %<>% full_join(extra_values, by="x")
+x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
+xm <- data.frame(
+  c1 =rep(1, length(x)),
+  c2 =x,
+  c3 =x^2,
+  c4 =x^3,
+  c5 =(x> 5)*rep(1, length(x)),
+  c6 =(x> 5)*(x- 5),
+  c7 =(x> 5)*(x- 5)^2,
+  c8 =(x> 5)*(x- 5)^3,
+  c9 =(x>10)*rep(1, length(x)),
+  c10=(x>10)*(x-10),
+  c11=(x>10)*(x-10)^2,
+  c12=(x>10)*(x-10)^3,
+  c13 =(x>15)*rep(1, length(x)),
+  c14=(x>15)*(x-15),
+  c15=(x>15)*(x-15)^2,
+  c16=(x>15)*(x-15)^3,
+  y=NA
+)
+xm$y[1:21] <- y
 ```
@@
 ```{r}
-single_cubic_model <- lm(y~poly(x, 3), data=simulated_example)
-single_cubic_prediction <- augment(single_cubic_model)
-ggplot(single_cubic_prediction, aes(x, y)) + 
+kable(xm[1:21, 1:4])
+single_cubic_model <- lm(y~c2+c3+c4, data=xm)
+single_cubic_model %>% 
+  augment(newdata=xm) %>%
+  ggplot(aes(x, y)) + 
     geom_point() +
-  geom_line(aes(y=.fitted))
+    geom_line(aes(y=.fitted)) +
+    ggtitle("Single cubic fit to the data")
 ```
@@
 
+The trick to fitting four separate cubic polynomials is to "restart" the cubic terms after x=5, x=10, and x=15.
+
 ```{r}
-sb1 <- (simulated_example$x<=5)
-sb2 <- (simulated_example$x>5)&(simulated_example$x<=10)
-sb3 <- (simulated_example$x>10)&(simulated_example$x<=15)
-sb4 <- (simulated_example$x>15)
-
-simulated_example %>%
-  filter(sb1) %>%
-  lm(y~poly(x, 3), data=.) %>%
-  augment(newdata=simulated_example[sb1, ]) -> first_cubic_prediction
-lm(y~poly(x, 3), subset=sb2, data=simulated_example) %>% 
-  augment(newdata=simulated_example[sb2, ]) %>%
-  select(x, .fitted) -> second_cubic_prediction
-lm(y~poly(x, 3), subset=sb3, data=simulated_example) %>%
-  augment(newdata=simulated_example[sb3, ]) %>% 
-  select(x, .fitted) -> third_cubic_prediction
-lm(y~poly(x, 3), subset=sb4, data=simulated_example) %>%
-  augment(newdata=simulated_example[sb4, ]) %>%
-  select(x, .fitted) -> fourth_cubic_prediction
-
-first_cubic_prediction %>%
-  bind_rows(second_cubic_prediction) %>%
-  bind_rows(third_cubic_prediction) %>%
-  bind_rows(fourth_cubic_prediction) %>%
-  full_join(simulated_example, by="x") -> combined_prediction
-ggplot(combined_prediction, aes(x, y.y)) +
-  geom_point() +
-  geom_line(aes(y=.fitted))
+kable(xm[1:20, ])
+xm$y <- NA
+xm$y[1:20] <- y
+four_independent_cubics <- 
+  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
+augment(four_independent_cubics, newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) + 
+    ggtitle("Discontinous cubic splines")
 ```
 
-This function is not continuous or smooth. To make a single smooth continuous function you need to define
+This function is not continuous or smooth. To make the function continuous, drop the extra intercept terms.
 
 ```{r}
-simulated_example %>%
-  mutate(x1=x) %>%
-  mutate(x2=x^2) %>%
-  mutate(x3=x^3) %>%
-  mutate(x4=(x>5)*(x-5)^3) %>%
-  mutate(x5=(x>10)*(x-10)^3) %>%
-  mutate(x6=(x>15)*(x-15)^3) -> spline_basis
+kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
+four_continuous_cubics <- 
+  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
+augment(four_continuous_cubics, newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) +
+    ggtitle("Continous but not smooth cubic splines")
+
+```
+
+Smoothness is measured in terms of the continuity of derivatives. Here is a function that has a continuous first derivative.
+
+```{r}
+kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
+four_smooth_cubics <- 
+  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
+augment(four_smooth_cubics, newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) + 
+    ggtitle("Continous and Smooth (1st derivative) splines")
+
+```
+
+Here is a function that has continuous first and second derivatives.
+
+```{r}
+kable(xm[1:20, c(1:4, 8, 12, 16)])
+spline_model <- 
+  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
+augment(spline_model, newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) +
+    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
+```
+
+While this approach is simple, there is an issue with multicollinearity.
+
+```{r}
+round(cor(xm[c(2:4, 8, 12, 16)]), 2)
+```
+
+This approach is simple and easy to follow, but there is one catch.
+The correlations are quite high and this can lead to computational problems, including rounding errors.
+So most spline models implemented on a computer use a different approach.
+
+B-splines provide a solution with less issues of multi-collinearity.
+
+```{r}
+linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
+kable(linear_spline[1:21, ])
+plot(x[1:21], linear_spline[1:21 , 1])
+plot(x[1:21], linear_spline[1:21 , 2])
+plot(x[1:21], linear_spline[1:21 , 3])
+plot(x[1:21], linear_spline[1:21 , 4])
+plot(x[1:21], linear_spline[1:21 , 5])
+
+lm(xm$y~linear_spline) %>%
+  augment(newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) + 
+    ggtitle("Linear splines")
 ```
@@
 ```{r}
-lm(y~x1+x2+x3+x4+x5+x6, data=spline_basis) %>%
-  augment(newdata=spline_basis) %>%
-  ggplot(aes(x, y)) +
-    geom_point() +
-  geom_line(aes(y=.fitted))
+quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
+kable(quadratic_spline[1:21, ])
+plot(x[1:21], quadratic_spline[1:21 , 1])
+plot(x[1:21], quadratic_spline[1:21 , 2])
+plot(x[1:21], quadratic_spline[1:21 , 3])
+plot(x[1:21], quadratic_spline[1:21 , 4])
+plot(x[1:21], quadratic_spline[1:21 , 5])
+
+lm(xm$y~quadratic_spline) %>%
+  augment(newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) + 
+    ggtitle("quadratic splines")
 ```
+
+```{r}
+cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
+kable(cubic_spline[1:21, ])
+plot(x[1:21], cubic_spline[1:21 , 1])
+plot(x[1:21], cubic_spline[1:21 , 2])
+plot(x[1:21], cubic_spline[1:21 , 3])
+plot(x[1:21], cubic_spline[1:21 , 4])
+plot(x[1:21], cubic_spline[1:21 , 5])
+
+lm(xm$y~cubic_spline) %>%
+  augment(newdata=xm) %>%
+  ggplot(aes(c2, .fitted)) +
+    geom_line() +
+    geom_point(aes(c2, y)) + 
+    ggtitle("cubic splines")
+```

### 2021-11-24 text/21/cubic-splines.Rmd
Converted plot to ggplot in cubic-splines.Rmd

@@
 x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
+```
+
+You can get a variety of splines by defining constant, linear, quadratic, and cubic terms and then shift those functions to the right with zeros to the left.
+
+```{r}
 xm <- data.frame(
@@
 ```{r}
+lb1 <- rep(c("intercept", "linear term", "quadratic term", "cubic term"), 4)
+lb2 <- rep(c("for the full range", "restarted at x=5", "restarted at x=10", "restarted at x=15"), each=4)
+for (j in 1:16) {
+  xm[1:21, ] %>%
+    ggplot(aes(c2, .data[[paste0("c", j)]])) +
+      ggtitle(paste(lb1[j], lb2[j])) +
+      geom_point() +
+      xlab(" ") +
+      ylab(" ") -> cplot
+  plot(cplot)
+}
+```
+
+### Single cubic polynomial
+
+This is the matrix of independent variables for a single cubic polynomial.
+
+```{r}
 kable(xm[1:21, 1:4])
+```
+
+The graph shown below represents the best fitting single cubic polynomial.
+
+```{r}
 single_cubic_model <- lm(y~c2+c3+c4, data=xm)
@@
 ```
+
+### Separate cubic polynomials
 

### 2021-11-24 text/21/cubic-splines.Rmd
More ggplot2 work with cubic-splines.Rmd

--- 2021-11-23 text/21/cubic-splines.Rmd
@@
 y <- rep(40, 21)
-y[6:21] <- y[6:21]-3*(1:15)
-y[11:21] <- y[11:21]+7*(1:10)
-y[16:21] <- y[16:21]-(1:5)^2
-y <- y+round(rnorm(10), 1)
+y[6:21] <- y[6:21]-3*(0:15)
+y[11:21] <- y[11:21]+7*(0:10)
+y[16:21] <- y[16:21]-(0:5)^2
+y <- y+round(rnorm(21), 1)
 simulated_example <- data.frame(x, y)
@@
 x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
+```
+
+You can get a variety of splines by defining constant, linear, quadratic, and cubic terms and then shift those functions to the right. After shifting, fill in the hole to the left with zeros.
+
+```{r}
 xm <- data.frame(
@@
   c12=(x>10)*(x-10)^3,
-  c13 =(x>15)*rep(1, length(x)),
+  c13=(x>15)*rep(1, length(x)),
   c14=(x>15)*(x-15),
@@
 ```{r}
+display_terms <- function(df, x, y, lb, z=y) {
+  full_data <- data.frame(x=df[ , x], y=df[ , y], z=df[ , z])
+  integer_data <- full_data[1:21, ]
+  ggplot(full_data, aes(x, y)) +
+      ggtitle(lb) +
+      geom_line() +
+      geom_point(aes(x, z), na.rm=TRUE) +
+      xlab(" ") +
+      ylab(" ")
+}
+
+lb1 <- rep(c("Intercept", "Linear term", "Quadratic term", "Cubic term"), 4)
+lb2 <- rep(c("for the full range", "restarted at x=5", "restarted at x=10", "restarted at x=15"), each=4)
+lb <- paste(lb1, lb2)
+for (j in 1:16) {
+  xm %>% display_terms("c2", paste0("c", j), lb[j]) -> cplot
+  plot(cplot)
+}
+```
+
+### Single cubic polynomial
+
+This is the matrix of independent variables for a single cubic polynomial.
+
+```{r}
 kable(xm[1:21, 1:4])
-single_cubic_model <- lm(y~c2+c3+c4, data=xm)
-single_cubic_model %>% 
+```
+
+The graph shown below represents the best fitting single cubic polynomial.
+
+```{r}
+lm(y~c2+c3+c4, data=xm) %>%
   augment(newdata=xm) %>%
-  ggplot(aes(x, y)) + 
-    geom_point() +
-    geom_line(aes(y=.fitted)) +
-    ggtitle("Single cubic fit to the data")
+  data.frame %>%
+  display_terms("c2", ".fitted", "Single cubic fit", "y")
 ```
+
+### Separate cubic polynomials
 
@@
 kable(xm[1:20, ])
-xm$y <- NA
-xm$y[1:20] <- y
+```
+
+```{r}
 four_independent_cubics <- 
@@
     geom_line() +
-    geom_point(aes(c2, y)) + 
+    # geom_point(xm[1:21, ], aes(c2, y)) +
     ggtitle("Discontinous cubic splines")
@@
     geom_line() +
-    geom_point(aes(c2, y)) +
+    # geom_point(xm[1:21, ], aes(c2, y)) +
     ggtitle("Continous but not smooth cubic splines")
-
 ```
@@
     geom_line() +
-    geom_point(aes(c2, y)) + 
+    geom_point(aes(c2, y), na.rm=TRUE) +
     ggtitle("Continous and Smooth (1st derivative) splines")
@@
     geom_line() +
-    geom_point(aes(c2, y)) +
+    # geom_point(xm[1:21, ], aes(c2, y)) +
     ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
@@
 kable(linear_spline[1:21, ])
+```
+
+```{r}
 plot(x[1:21], linear_spline[1:21 , 1])
@@
     geom_line() +
-    geom_point(aes(c2, y)) + 
+    # geom_point(xm[1:21, ], aes(c2, y)) +
     ggtitle("Linear splines")
@@
     geom_line() +
-    geom_point(aes(c2, y)) + 
+    geom_point(aes(c2, y), na.rm=TRUE) + 
     ggtitle("quadratic splines")
@@
     geom_line() +
-    geom_point(aes(c2, y)) + 
+    geom_point(aes(c2, y), na.rm=TRUE) + 
     ggtitle("cubic splines")

### 2021-11-24 text/21/cubic-splines.Rmd
Added natural splines to cubic-splines.Rmd

@@
   c16=(x>15)*(x-15)^3,
+  x=x,
   y=NA
@@
 ```{r}
-display_terms <- function(df, x, y, lb, z=y) {
-  full_data <- data.frame(x=df[ , x], y=df[ , y], z=df[ , z])
-  integer_data <- full_data[1:21, ]
-  ggplot(full_data, aes(x, y)) +
+display_terms <- function(df, lb) {
+  names(df) <- c("x", "y")
+  integer_data <- df[1:21, ]
+  ggplot(df, aes(x, y)) +
       ggtitle(lb) +
       geom_line() +
-      geom_point(aes(x, z), na.rm=TRUE) +
+      geom_point(data=integer_data, aes(x, y)) +
       xlab(" ") +
@@
 lb <- paste(lb1, lb2)
+
 for (j in 1:16) {
-  xm %>% display_terms("c2", paste0("c", j), lb[j]) -> cplot
+  xm %>% 
+    select(c(17, j)) %>%
+    display_terms(lb[j]) -> cplot
   plot(cplot)
 }
+
 ```
@@
 ```{r}
+display_fit <- function(model_fit, lb) {
+  model_fit %>%
+    augment(newdata=xm) %>%
+      ggplot(aes(c2, .fitted)) +
+        ggtitle(lb) +
+        geom_line() +
+        geom_point(aes(c2, y), na.rm=TRUE) +
+      xlab(" ") +
+      ylab(" ")
+}
+
 lm(y~c2+c3+c4, data=xm) %>%
-  augment(newdata=xm) %>%
-  data.frame %>%
-  display_terms("c2", ".fitted", "Single cubic fit", "y")
+  display_fit("Single cubic fit")
 ```
@@
 ```{r}
-four_independent_cubics <- 
-  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
-augment(four_independent_cubics, newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    # geom_point(xm[1:21, ], aes(c2, y)) +
-    ggtitle("Discontinous cubic splines")
+
+lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
+  display_fit("Discontinous cubic splines")
 ```
@@
 kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
-four_continuous_cubics <- 
-  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
-augment(four_continuous_cubics, newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    # geom_point(xm[1:21, ], aes(c2, y)) +
-    ggtitle("Continous but not smooth cubic splines")
+lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
+  display_fit("Continous but not smooth cubic splines")
 ```
@@
 kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
-four_smooth_cubics <- 
-  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
-augment(four_smooth_cubics, newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    geom_point(aes(c2, y), na.rm=TRUE) +
-    ggtitle("Continous and Smooth (1st derivative) splines")
+lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
+  display_fit("Continous and Smooth (1st derivative) splines")
 
@@
 kable(xm[1:20, c(1:4, 8, 12, 16)])
-spline_model <- 
-  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
-augment(spline_model, newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    # geom_point(xm[1:21, ], aes(c2, y)) +
-    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
+lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
+  display_fit("Continous and smooth (1st and 2nd derivatives) cubic splines")
 ```
@@
 ```{r}
-plot(x[1:21], linear_spline[1:21 , 1])
-plot(x[1:21], linear_spline[1:21 , 2])
-plot(x[1:21], linear_spline[1:21 , 3])
-plot(x[1:21], linear_spline[1:21 , 4])
-plot(x[1:21], linear_spline[1:21 , 5])
+display_matrix <- function(df, lb) {
+  names(df) <- c("x", "y")
+  integer_data <- df[1:21, ]
+  ggplot(df, aes(x, y)) +
+      ggtitle(lb) +
+      geom_line() +
+      geom_point(data=integer_data, aes(x, y)) +
+      xlab(" ") +
+      ylab(" ")
+}
+for (j in 1:dim(linear_spline)[2]) {
+  cbind(xm$c2, linear_spline[ , j]) %>% 
+    data.frame %>%
+    display_matrix("Linear spline terms") %>%
+    plot
+}
+```
 
+```{r}
 lm(xm$y~linear_spline) %>%
-  augment(newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    # geom_point(xm[1:21, ], aes(c2, y)) +
-    ggtitle("Linear splines")
+  display_fit("Linear splines")
 ```
@@
 kable(quadratic_spline[1:21, ])
-plot(x[1:21], quadratic_spline[1:21 , 1])
-plot(x[1:21], quadratic_spline[1:21 , 2])
-plot(x[1:21], quadratic_spline[1:21 , 3])
-plot(x[1:21], quadratic_spline[1:21 , 4])
-plot(x[1:21], quadratic_spline[1:21 , 5])
 
+for (j in 1:dim(quadratic_spline)[2]) {
+  cbind(xm$c2, quadratic_spline[ , j]) %>% 
+    data.frame %>%
+    display_matrix("Quadratic spline terms") %>%
+    plot
+}
+```
+
+```{r}
 lm(xm$y~quadratic_spline) %>%
-  augment(newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    geom_point(aes(c2, y), na.rm=TRUE) + 
-    ggtitle("quadratic splines")
+  display_fit("Quadratic splines")
 ```
@@
 kable(cubic_spline[1:21, ])
-plot(x[1:21], cubic_spline[1:21 , 1])
-plot(x[1:21], cubic_spline[1:21 , 2])
-plot(x[1:21], cubic_spline[1:21 , 3])
-plot(x[1:21], cubic_spline[1:21 , 4])
-plot(x[1:21], cubic_spline[1:21 , 5])
+```
 
+```{r}
+for (j in 1:dim(cubic_spline)[2]) {
+  cbind(xm$c2, cubic_spline[ , j]) %>% 
+    data.frame %>%
+    display_matrix("Cubic spline terms") %>%
+    plot
+}
+```
+
+```{r}
 lm(xm$y~cubic_spline) %>%
-  augment(newdata=xm) %>%
-  ggplot(aes(c2, .fitted)) +
-    geom_line() +
-    geom_point(aes(c2, y), na.rm=TRUE) + 
-    ggtitle("cubic splines")
+  display_fit("Cubic splines")
 ```
+
+```{r}
+natural_spline <- ns(xm$c2, knots=c(5, 10, 15), intercept=TRUE)
+kable(natural_spline[1:21, ])
+```
+
+```{r}
+for (j in 1:dim(natural_spline)[2]) {
+  cbind(xm$c2, natural_spline[ , j]) %>% 
+    data.frame %>%
+    display_matrix("Natural spline terms") %>%
+    plot
+}
+```
+
+```{r}
+lm(xm$y~natural_spline) %>%
+  display_fit("Natural splines")
+```

### 2021-11-25 text/21/cubic-splines.Rmd
More tinkering with cubic-splines.Rmd

@@
 ```{r}
-display_terms <- function(df, lb) {
-  names(df) <- c("x", "y")
-  integer_data <- df[1:21, ]
-  ggplot(df, aes(x, y)) +
+display_curve <- function(dat, lb) {
+  # Plot a line connecting the points in dat
+  # Assume that column 1 is x and column 2 is y
+  # Work with either a data frame or matrix
+  dat %>%
+    data.frame %>%
+    select(1:2) %>%
+    set_names(c("x", "y")) %>%
+    ggplot(aes(x, y)) +
       ggtitle(lb) +
       geom_line() +
-      geom_point(data=integer_data, aes(x, y)) +
       xlab(" ") +
       ylab(" ")
+}
+
+display_dots <- function(cplot, dat) {
+  # Add dots in dat to an existing graph
+  # Assume that column 1 is x and column 2 is y
+  # Work with either a data frame or matrix
+  dat %>%
+    data.frame %>%
+    set_names(c("x", "y")) -> df
+  cplot + geom_point(data=df, aes(x, y))
 }
@@
 for (j in 1:16) {
-  xm %>% 
-    select(c(17, j)) %>%
-    display_terms(lb[j]) -> cplot
-  plot(cplot)
+  dat <- xm[ , c("x",  paste0("c", j))]
+  display_curve(dat, lb[j]) %>%
+    display_dots(dat[1:21, ]) %>%
+    plot
 }
@@
 ```{r}
-display_fit <- function(model_fit, lb) {
-  model_fit %>%
-    augment(newdata=xm) %>%
-      ggplot(aes(c2, .fitted)) +
-        ggtitle(lb) +
-        geom_line() +
-        geom_point(aes(c2, y), na.rm=TRUE) +
-      xlab(" ") +
-      ylab(" ")
-}
 
 lm(y~c2+c3+c4, data=xm) %>%
-  display_fit("Single cubic fit")
+  augment %>%
+  select(c2, .fitted) %>%
+    display_curve("Single cubic fit") %>%
+    display_dots(simulated_example)
 ```
@@
 ```{r}
-
 lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
-  display_fit("Discontinous cubic splines")
+  augment %>%
+  select(c2, .fitted) %>%
+    display_curve("Discontinous cubic splines") %>%
+    display_dots(simulated_example)
 ```
@@
 
+### Continuous splines
+
 ```{r}
@@
 lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
-  display_fit("Continous but not smooth cubic splines")
+  augment %>%
+  select(c2, .fitted) %>%
+    display_curve("Continuous but not smooth cubic splines") %>%
+    display_dots(simulated_example)
 ```
 
-Smoothness is measured in terms of the continuity of derivatives. Here is a function that has a continuous first derivative.
+While this graph is continuos, it still takes some abrupt term. The mathemSmoothness is measured in terms of the continuity of derivatives.
+
+### Smooth splines
+
+Here is a function that has a continuous first derivative.
 
@@
 lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
-  display_fit("Continous and Smooth (1st derivative) splines")
-
+  augment %>%
+  select(c2, .fitted) %>%
+    display_curve("Continuous and smooth (1st derivative) cubic splines") %>%
+    display_dots(simulated_example)
 ```
@@
 lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
-  display_fit("Continous and smooth (1st and 2nd derivatives) cubic splines")
+  augment %>%
+  select(c2, .fitted) %>%
+    display_curve("Continous and smooth (1st and 2nd derivatives) cubic splines") %>%
+    display_dots(simulated_example)
 ```
 
-While this approach is simple, there is an issue with multicollinearity.
+This approach is simple and easy to follow, but there is one catch. There is an issue with multicollinearity.
 
@@
 
-This approach is simple and easy to follow, but there is one catch.
-The correlations are quite high and this can lead to computational problems, including rounding errors.
-So most spline models implemented on a computer use a different approach.
+The correlations are quite high and this can lead to computational problems, including rounding errors. So most spline models implemented on a computer use a different approach.
 
-B-splines provide a solution with less issues of multi-collinearity.
+### B-splines
+
+B-splines provide a solution with less issues of multi-collinearity. You use the bs function in the splines package to compute B-splines.
+
+#### Linear B-splines
+
+Although most applications use cubic B-splines, they are available in other forms. Here is the linear spline.
 
@@
 
+The terms in the linear B-splines are mostly piecewise linear functions. Notice how they are zero outside certain ranges.
+
 ```{r}
-display_matrix <- function(df, lb) {
-  names(df) <- c("x", "y")
-  integer_data <- df[1:21, ]
-  ggplot(df, aes(x, y)) +
-      ggtitle(lb) +
-      geom_line() +
-      geom_point(data=integer_data, aes(x, y)) +
-      xlab(" ") +
-      ylab(" ")
-}
 for (j in 1:dim(linear_spline)[2]) {
-  cbind(xm$c2, linear_spline[ , j]) %>% 
-    data.frame %>%
-    display_matrix("Linear spline terms") %>%
+  dat <- cbind(xm$c2, linear_spline[ , j])
+  display_curve(dat, "Linear B-spline terms") %>%
+    display_dots(dat[1:21, ]) %>%
     plot
@@
 ```{r}
+round(cor(linear_spline), 1))
+```
+
+```{r}
 lm(xm$y~linear_spline) %>%
-  display_fit("Linear splines")
+  augment %>%
+  bind_cols(xm$c2) %>%
+  select(c2, .fitted) %>%
+    display_curve("Linear spline fit to the data") %>%
+    display_dots(simulated_example)
 ```
@@
 kable(quadratic_spline[1:21, ])
+```
 
+```{r}
 for (j in 1:dim(quadratic_spline)[2]) {
-  cbind(xm$c2, quadratic_spline[ , j]) %>% 
-    data.frame %>%
-    display_matrix("Quadratic spline terms") %>%
+  dat <- cbind(xm$c2, quadratic_spline[ , j])
+  display_curve(dat, "Quadratic B-spline terms") %>%
+    display_dots(dat[1:21, ]) %>%
     plot
@@
 lm(xm$y~quadratic_spline) %>%
-  display_fit("Quadratic splines")
+  bind_cols(xm$c2) %>%
+  select(c2, .fitted) %>%
+    display_curve("Quadratic spline fit to the data") %>%
+    display_dots(simulated_example)
 ```
@@
 for (j in 1:dim(cubic_spline)[2]) {
-  cbind(xm$c2, cubic_spline[ , j]) %>% 
-    data.frame %>%
-    display_matrix("Cubic spline terms") %>%
+  dat <- cbind(xm$c2, cubic_spline[ , j])
+  display_curve(dat, "Cubic B-spline terms") %>%
+    display_dots(dat[1:21, ]) %>%
     plot
@@
 lm(xm$y~cubic_spline) %>%
-  display_fit("Cubic splines")
+  bind_cols(xm$c2) %>%
+  select(c2, .fitted) %>%
+    display_curve("Cubic spline fit to the data") %>%
+    display_dots(simulated_example)
 ```
@@
 ```{r}
+extended_x <- seq(-5, 25, by=1/64)
+extended_spline <- predict(natural_spline, extended_x)
 for (j in 1:dim(natural_spline)[2]) {
-  cbind(xm$c2, natural_spline[ , j]) %>% 
+  cbind(extended_x, extended_spline[ , j]) %>% 
     data.frame %>%
@@
 ```
+
+References
+
+Donald H. House. Chapter 14. Spline Curves. Available in [pdf format][hou1].
+
+Aris Perperoglou, Willi Sauerbrei, Michal Abrahamowicz, Matthias Schmid.
+A review of spline function procedures in R.
+BMC Medical Research Methodology 19, 46 (2019).
+DOI: 10.1186/s12874-019-0666-3.
+Available in [html format][per1] or [pdf format][per2].
+
+[hou1]: https://people.cs.clemson.edu/~dhouse/courses/405/notes/splines.pdf
+
+[per1]: https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-019-0666-3
+[per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf

### 2021-11-26 text/21/cubic-splines.Rmd
Still more tinkering with cubic-splines.Rmd
Talked about knots in cubic-spline.Rmd

@@
 suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(Hmisc)))
 suppressMessages(suppressWarnings(library(knitr)))
 suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(rms)))
 suppressMessages(suppressWarnings(library(splines)))
@@
 lm(y~c2+c3+c4, data=xm) %>%
-  augment %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
-  augment %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
-  augment %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 
-While this graph is continuos, it still takes some abrupt term. The mathemSmoothness is measured in terms of the continuity of derivatives.
+While this graph is continuos, it still takes some abrupt term. The mathematical concept of smoothness is measured in terms of the continuity of derivatives.
 
@@
 lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
-  augment %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
-  augment %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 
-#### Linear B-splines
+```{r}
+cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
+kable(cubic_spline[1:21, ])
+```
+
+The individual columns represent piecwise cubic polynomials. Notice how they are concentrated in certain intervals.
+
+```{r}
+for (j in 1:dim(cubic_spline)[2]) {
+  dat <- cbind(xm$c2, cubic_spline[ , j])
+  display_curve(dat, "Cubic B-spline terms") %>%
+    display_dots(dat[1:21, ]) %>%
+    plot
+}
+```
+
+B-splines have less issues with multicollinearity.
+
+```{r}
+round(cor(cubic_spline), 1)
+```
+
+Although there is some correlation, this not nearly as bad as the piecewise approach.
+
+```{r}
+lm(xm$y~cubic_spline) %>%
+  augment(newdata=xm) %>%
+  select(c2, .fitted) %>%
+    display_curve("Cubic spline fit to the data") %>%
+    display_dots(simulated_example)
+```
+
+#### Linear and quadratic B-splines
 
@@
 ```{r}
-round(cor(linear_spline), 1))
-```
-
-```{r}
 lm(xm$y~linear_spline) %>%
-  augment %>%
-  bind_cols(xm$c2) %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 lm(xm$y~quadratic_spline) %>%
-  bind_cols(xm$c2) %>%
+  augment(newdata=xm) %>%
   select(c2, .fitted) %>%
@@
 
-```{r}
-cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
-kable(cubic_spline[1:21, ])
-```
-
-```{r}
-for (j in 1:dim(cubic_spline)[2]) {
-  dat <- cbind(xm$c2, cubic_spline[ , j])
-  display_curve(dat, "Cubic B-spline terms") %>%
-    display_dots(dat[1:21, ]) %>%
-    plot
-}
-```
-
-```{r}
-lm(xm$y~cubic_spline) %>%
-  bind_cols(xm$c2) %>%
-  select(c2, .fitted) %>%
-    display_curve("Cubic spline fit to the data") %>%
-    display_dots(simulated_example)
-```
+### Natural splines
 
@@
 extended_spline <- predict(natural_spline, extended_x)
-for (j in 1:dim(natural_spline)[2]) {
-  cbind(extended_x, extended_spline[ , j]) %>% 
-    data.frame %>%
-    display_matrix("Natural spline terms") %>%
+for (j in 1:dim(extended_spline)[2]) {
+  dat <- cbind(extended_x, extended_spline[ , j])
+  spline_dots <- cbind(0:20, extended_spline[extended_x %in% 0:20, j])
+  display_curve(dat, "Natural spline terms") %>%
+    display_dots(spline_dots) %>%
     plot
@@
 lm(xm$y~natural_spline) %>%
-  display_fit("Natural splines")
+  augment(newdata=xm) %>%
+  select(c2, .fitted) %>%
+    display_curve("Natural spline fit to the data") %>%
+    display_dots(simulated_example)
 ```
 
-References
+### How many knots?
+
+#### Use AIC or BIC
+
+```{r}
+for (k in 3:7) {
+  knots <- 20*((1:k)-0.5)/k
+  lm(y~bs(c2, knots=knots), data=xm) -> rcs_fit
+  # lm(y~rcs(c2, k), data=xm) -> rcs_fit
+  ti <- paste0(k, " knots, AIC = ", round(AIC(rcs_fit), 1), ", BIC = ", round(BIC(rcs_fit), 1))
+  rcs_fit %>%
+    augment(newdata=xm) %>%
+    select(c2, .fitted) %>%
+      display_curve(ti) %>%
+      display_dots(simulated_example) -> cplot
+    plot(cplot + geom_vline(xintercept=knots, color="black", linetype="dotted"))
+}
+```
+
+#### Eyeball the data
+
+Look at the number of bends in the data.
+If the data increases to a single maximum and then decreases after that, a simpler spline with 2 or 3 knots may be sufficient.
+This also applies if the data decreases to a single minimum and then increases after that.
+If there are more bends (e.g., increase to a maximum, decrease to a minimum, and then increase again), then a larger number of knots may be needed.
+
+#### Frank Harrell's suggestions
+
+Use 4 knots if the total sample size is less than 100 and use 5 knots if it is more than 100.
+
+### Where to place the knots?
+
+#### Use inside knowledge
+
+Sometimes you have knowledge of the specific application that will help you to figure out where to put your knots.
+
+I am not an expert on cars, but I have been told that many newer cars with automatic transmission change how the transmission behaves around 40 miles per hour.
+This transition helps with highway mileage.
+So if you are fitting a spline curve to data where how the transmission behaves, make sure that you place one or two of your knots near 40 miles per hour.
+
+I am also not an expert on kidneys, but I have been told that the Glomerural Filtration Rate is not too critical if the value is above 90, but becomes very serious when it is less than 30.
+So a model looking at health effects using GFR should probably have knots around 30 and 90.
+
+Similarly, CD4 cell counts above 500 are a good sign, but things turn rapidly worse if they dip below 200.
+
+#### Frank Harrell's suggestion
+
+Frank Harrell suggests that you place the knots not evenly across the range of X but at equally spaced quantiles of the X distribution.
+This makes sense when the distribution of the X values is not uniform.
+If, for example, X is skewed to the right (has a tendency to produce most of the data on the left with a few scattered outliers on the right), the knots will tend to favor the data-rich left side of the distribution.
+He also suggests placing the leftmost and rightmost knots near, but not at the extremes of the X values, such as at the 10th or 90th percentiles or at the fifth smallest and the fifth largest values in the data.
+The actual percentiles are a bit tricky to explain.
+
+*"For 3 knots, the outer quantiles used are 0.10 and 0.90.
+For 4-6 knots, the outer quantiles used are 0.05 and 0.95.
+For more than 6 knots, the outer quantiles are 0.025 and 0.975.
+The knots are equally spaced between these on the quantile scale.
+For fewer than 100 non-missing values of x, the outer knots are the 5th smallest and the 5th largest x."* as quoted [here][har1],
+
+If you pick this apart, you will find the 4 knots for a large dataset would be placed at the 5th, 35th, 65th and 95th percentiles.
+
+#### It doesn't matter
+
+Most references I have looked at state that it is the number of knots rather than the placement of the knots that is critical.
+
+### References
 
@@
 
+[har1]: https://search.r-project.org/CRAN/refmans/Hmisc/html/rcspline.eval.html
+
 [hou1]: https://people.cs.clemson.edu/~dhouse/courses/405/notes/splines.pdf

### 2021-11-27 text/21/cubic-splines.Rmd
More work on natural splines in cubic-splines.Rmd

@@
 
-While this graph is continuos, it still takes some abrupt term. The mathematical concept of smoothness is measured in terms of the continuity of derivatives.
+While this graph is continuous, it still takes some abrupt turns. What this curve lacks is smoothness. The mathematical concept of smoothness is measured in terms of the continuity of derivatives.
 
@@
 Here is a function that has a continuous first derivative.
+You fit this model by dropping the extra linear terms beyond the first one.
+Notice a pattern here.
+As you place additional restrictions on the spline (continuity, smoothness), you need fewer parameters.
+The four cubic models with no restrictions used up 16 degrees of freedom.
+When you added a continuity restriction, you only needed 13 degrees of freedom for the model.
+Add a smoothness restriction and you only need 10 degrees of freedom.
 
@@
 
-Here is a function that has continuous first and second derivatives.
+Here is a function that has continuous first and second derivatives. This is a greater degree of smoothness than above and it requires only 7 degrees of freedom.
 
@@
 ```
+
+This is what most people refer to when they talk about splines: a piecewise cubic model with continuity and continuous first and second derivatives.
+It is a fairly simple model (not that many degrees of freedom), but it produces a curve that has the flexibility to fit a variety of curves that have the aesthetically pleasing features of continuity and smoothness.
+
+Continuity and smoothness are more than just aesthetics, though.
+There are many scientific settings where we expect no jumps (discontinuities) and no abrupt turns (lack of smoothness).
+If you are measuring the onset of symptoms from a disease, you know that the viruses or bacteria that are causing the disease are increasing in a continuous and smooth pattern.
+So any problems that they cause should also increase in a continuous and smooth pattern.
+
+Other settings, however, should not necessarily be expected to produce continuous and smooth outcomes.
+If a particular metabolic pathway becomes saturated or an anotomical barrier is breached, the suddenness transition could result in an abrupt turn or a discontinuity.
+So do think about the particular context of your problem when deciding what type of spline model to use.
 
@@
 
-The individual columns represent piecwise cubic polynomials. Notice how they are concentrated in certain intervals.
+The individual columns represent piecwise cubic polynomials.
+Notice how they are concentrated in certain intervals and there is only a partial overlap between these intervals.
+Notice also how they transition smoothly to zero outside those intervals.
 
@@
 
-The terms in the linear B-splines are mostly piecewise linear functions. Notice how they are zero outside certain ranges.
+The terms in the linear B-splines are mostly piecewise linear functions. Notice how they are zero outside certain ranges, with only a small amount of overlap between adjacent terms.
 
@@
 
+This is a linear spline fit to the data you have been using.
+
 ```{r}
@@
 
+A linear spline has continuity, but the nature of linear function makes it impossible to get smoothness.
+
+You can also compute quadratic splines.
+
 ```{r}
@@
 ```
+
+Notice how the quadratic spline terms transition smoothly to zero outside certain intervals.
 
@@
 
+This is the fit of the quadratic spline.
+It can only produce smoothness up to a continuous first derivative, but the difference between this and smoothness with continuous first and second derivatives is subtle.
+
 ```{r}
@@
 
+The quadratic spline seems to do fairly well here and there is often little difference between it and the cubic spline. Nevertheless, cubic splines are far more common.
+
+You could also look at quartic splines (using fourth order polynomials) or even higher.
+Sometimes these are used when you need to examine not just the function itself, but its derivatives.
+These applications, however, are rare.
+
 ### Natural splines
+
+A variant on B splines are natural splines (also called restricted cubic splines).
+These splines place an additional restriction to the left of the first knot and to the right of the last knot.
+The spline is constrained to be linear at both extremes.
+This makes practical sense, as there is less data at the extremes, making estimation of a complex cubic function here worrisome.
+This also makes extrapolation outside of the range of data less problematic.
+Cubic polynomials have the potential of extreme shifts and if these occur outside the range of the data, they could lead to some awful extrapolations.
+
+You should always be very careful, of course, as any effort to extrapolate beyond the range of data is [dangerous][sim3].
+Nevertheless, restricting the extrapolation to a linear function is probably safer than letting the cubic polynomial wiggle around.
 
@@
 ```
+
+It is a bit tricky to show the extrapolation of nte natural splines. This approach uses the predict function.
 
@@
 
+Here is the natural spline fit to your simulated dataset.
+
 ```{r}
-lm(xm$y~natural_spline) %>%
-  augment(newdata=xm) %>%
+extended_y <- rep(NA, length(extended_x))
+extended_y[extended_x %in% 0:20] <- simulated_example$y
+lm(extended_y~extended_spline) %>%
+  augment(newdata=extended_spline) %>%
+  data.frame(c2=extended_x) %>%
   select(c2, .fitted) %>%
@@
 
+You can probably see the linearity below 5 and above 15.
+
 ### How many knots?
 
+A difficult question is how many knots to use. Too many knots and you might end up overfitting. Too few and you might end up with not enough flexibility to fit your data well.
+
 #### Use AIC or BIC
+
+The AIC (Akaike Information Criterion) and the BIC (Bayesian Information Criterion) are useful measures for comparing different statistical models.
+In linear regression both AIC and BIC look at how close the fitted curve is to the data, but adds a penalty for model complexity.
+This helps avoid the situation where an excessively complex model with only marginally better predictive power is selected over a simpler model that predicts almost as well.
+
+NOte that a p-value will not work here except in some special cases where all of the knots but one coincide.
+The p-value fails because (with a few rare exceptions) one spline model is not nested inside another.
 
@@
 Use 4 knots if the total sample size is less than 100 and use 5 knots if it is more than 100.
+
+#### Pick a number based on your a priori beliefs
+
+You may have a feel for how much complexity is appropriate based on your years of experience as a data analyst and your scientific knowledge of the process at hand.
+After you work with enough splines, you do get an appreciation on how wiggly they can get.
+If you also have a rough idea of how the nonlinear relationship is going to be, perhaps based on seeing other similar problems in the area, you can match the degrees of freedom of the spline to your expectation, prior to looking at the data.
+
+#### Pick a number after looking at some preliminary graphs
+
+This is a bit controversial.
+Selecting a statistical model post hoc (after viewing the data) leaves you open to a charge of data dredging or going on a fishing expedition.
+To be fair, it is not as bad as some approaches (such as running ten tests and then choosing the one with the smallest p-value).
 
@@
 
-If you pick this apart, you will find the 4 knots for a large dataset would be placed at the 5th, 35th, 65th and 95th percentiles.
+If you pick this apart, you can deduce that 4 knots for a large dataset would be placed at the 5th, 35th, 65th and 95th percentiles.
 
@@
 Most references I have looked at state that it is the number of knots rather than the placement of the knots that is critical.
+
+Here are B-spline models with knots sliding from 3, 8, 13 to 7, 12, 17.
+
+```{r}
+for (k in 3:7) {
+  knots <- k + c(0, 5, 10)
+  lm(y~bs(c2, knots=knots), data=xm) -> rcs_fit
+  # lm(y~rcs(c2, k), data=xm) -> rcs_fit
+  ti <- paste0("Knots at ", paste(knots, collapse=", "))
+  rcs_fit %>%
+    augment(newdata=xm) %>%
+    select(c2, .fitted) %>%
+      display_curve(ti) %>%
+      display_dots(simulated_example) -> cplot
+    plot(cplot + geom_vline(xintercept=knots, color="black", linetype="dotted"))
+}
+```
+
+There are a few change in the spline functions as the knots move, but the changes are not very large in magnitude.
 
@@
 [per2]: https://bmcmedresmethodol.biomedcentral.com/track/pdf/10.1186/s12874-019-0666-3.pdf
+
+[sim3]: http://new.pmean.com/extrapolation-joke/

### 2021-11-27 text/21/cubic-splines.Rmd
Added open-access-deal.md

@@
 
+This is what the various pieces look like numerically ...
+
+```{r}
+kable(xm[1:21, 1:16])
+```
+
+... and graphically.
+
 ```{r}
@@
 
-This is the matrix of independent variables for a single cubic polynomial.
-
-```{r}
-kable(xm[1:21, 1:4])
-```
-
 The graph shown below represents the best fitting single cubic polynomial.
@@
 
-The trick to fitting four separate cubic polynomials is to "restart" the cubic terms after x=5, x=10, and x=15.
-
-```{r}
-kable(xm[1:20, ])
-```
+The trick to fitting four separate cubic polynomials is to "restart" the intercept, linear, quadratic, and cubic terms after x=5, x=10, and x=15, as shown above.
+This leads to a model with 16 degrees of freedom.
+This is way too many degrees of freedom for only 20 data points, but it helps anchor a series of more reasonable models.
 
@@
 ```{r}
-kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
 lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
@@
 ```{r}
-kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
 lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
@@
 ```{r}
-kable(xm[1:20, c(1:4, 8, 12, 16)])
 lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
@@
 cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
-kable(cubic_spline[1:21, ])
 ```
@@
 linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
-kable(linear_spline[1:21, ])
 ```
@@
 
-You can probably see the linearity below 5 and above 15.
+You can probably see the linearity at the extremes of the data.
 

### 2021-12-11 text/21/hurricane-frequencies.Rmd
Added hurricane-frequencies.Rmd

---
title: "Overfitting example using hurricane frequencies"
author: "Steve Simon"
source: "New"
date: "2021-12-11"
categories:
- Blog post
tags:
- Generalized linear model
output: html_document
---

```{r notes}
# Note: this code can produce a reasonable PowerPoint
# presentation. The system has an annoying tendency
# to put the graphs and the speaker notes on separate
# slides.
```

### Overfitting

Overfitting occurs when a statistical model is too complex for the amount of data that it is based on. The summary statistics on the data itself appear to be quite good, but the model will almost always produce poor predictions for new data. Here is an example of overfitting using [data on hurricane frequencies][dasl1] from the [Data and Story Library site][dasl2].

```{r setup, echo=FALSE}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
knitr::opts_chunk$set(echo=FALSE)
```

### The data used in this example (1/2)

```{r read1}
fn <- "https://dasl.datadescription.com/download/data/3279"
raw_data <- read.delim(file=fn, header=TRUE, sep="\t")
raw_data[1:8, ]
```

<div class="notes">

This [dataset on hurricane frequency][dasl1] is found at the [Data and Story Library website][dasl2].

Here are the first eight rows of data.

</div>

### The data used in this example (2/2)

```{r read2}
raw_data[9:16, ]
```

<div class="notes">

Here is the rest of the data.

</div>

### Plot the data

```{r plot}
df <- data.frame(x=seq(1855, 1995, by=10), y=raw_data$Count[1:15])
ggplot(df, aes(x, y)) + 
  geom_point() -> point_plot
plot(point_plot)
```

<div class="notes">

Here's what the data looks like in a graph. Nothing too unusual about the data. Now let's try to forecast the number of hurricanes for the next decade.

</div>

### Plot the data with a linear prediction

```{r poly1}
add_trend <- function(point_plot, df, degree) {
  x_grid <- data.frame(x=seq(1850, 2010, by=0.1))
  lm(y~poly(x, degree), data=df) %>%
    augment(newdata=x_grid) -> yhat
  yhat_2005 <- filter(yhat, x==2005)
  prediction <<- yhat_2005$.fitted
  point_plot +
    geom_line(data=yhat, aes(x, .fitted), color="red") +
    geom_label(
      data=yhat_2005, 
      aes(x, .fitted, label=round(.fitted, 1)))
}
add_trend(point_plot, df, 1)
```

<div class="notes">

Here's a linear trend. The prediction is `r prediction` for the next decade.

</div>

### Plot the data with a quadratic prediction

```{r poly2}
add_trend(point_plot, df, 2)
```

<div class="notes">

Here's a quadratic trend. The prediction is `r prediction`. It's quite a bit different.

</div>

### Plot the data with a cubic prediction

```{r poly3}
add_trend(point_plot, df, 3)
```

<div class="notes">

Here's a cubic (third order polynomial trend. The prediction is now `r prediction`.

</div>

### Plot the data with a 4th order polynomial

```{r poly4}
add_trend(point_plot, df, 4)
```

<div class="notes">

Here's a quartic (fourth order polynomial) trend. The prediction is `r prediction`.

</div>

### Plot the data with a 5th order polynomial

```{r poly5}
add_trend(point_plot, df, 5)
```

<div class="notes">

Here's a fifth order polynomial trend. The prediction is `r prediction`.

</div>

### Plot the data with a 6th order polynomial

```{r poly6}
add_trend(point_plot, df, 6)
```

<div class="notes">

Here's where things get very weird. The sixth order polynomial produces a prediction of `r prediction`, which is more than twice as large as any previous value.

</div>

### Plot the data with a 7th order polynomial

```{r poly7}
add_trend(point_plot, df, 7)
```

<div class="notes">

The weirdness continues with the seventh order polynomial trend, which produces a negative prediction (`r prediction`.

</div>

### Plot the data with a 8th order polynomial

```{r poly8}
add_trend(point_plot, df, 8)
```

<div class="notes">

The eigth order polynomial trend also produces a negative prediction (`r prediction`).

</div>

### Plot the data with a 9th order polynomial

```{r poly9}
add_trend(point_plot, df, 9)
```

<div class="notes">

Here's a ninth order polynomial trend. The prediction is so extreme (`r prediction`) as to be ridiculous.

</div>

### Plot the data with a 10th order polynomial

```{r poly10}
add_trend(point_plot, df, 10)
```

<div class="notes">

For the record, here are the tenth...

</div>

### Plot the data with a 11th order polynomial

```{r poly11}
add_trend(point_plot, df, 11)
```

<div class="notes">

... eleventh ...

</div>

### Plot the data with a 12th order polynomial

```{r poly12}
add_trend(point_plot, df, 12)
```

<div class="notes">

... twelfth ...

</div>

### Plot the data with a 13th order polynomial

```{r poly13}
add_trend(point_plot, df, 13)
```

<div class="notes">

... thirteenth ...

</div>

### Plot the data with a 14th order polynomial

```{r poly14}
add_trend(point_plot, df, 14)
```

<div class="notes">

... and fourteenth order polynomials.

</div>

### Code that I used

```{r reset}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(eval=FALSE)
```

```{r setup}
```

<div class="notes">

Here are the libraries that I needed for this work.

</div>

### Code that I used

```{r read1}
```

<div class="notes">

This code reads data directly from the website and prints out the first eight rows.

</div>

### Code that I used

```{r read2}
```

<div class="notes">

This code is for printing the rest of the data.

</div>

### Code that I used

```{r plot}
```

<div class="notes">

Here's the code for plotting the data points. I save the graph in an object so I can add trend lines later on.

</div>

### Code that I used

```{r poly1}
```

<div class="notes">

Here's the code for adding a trend line to the graph. I am setting up a flexible function that can print the trend from any polynomial.

</div>

### Code that I used

```{r poly2}
```

<div class="notes">

It took a bit of work to put everything in a function, but now you can produce a quadratic rather than a linear trend with just a single line of code.

</div>

### Code that I used

```{r poly3}
```

```{r poly4}
```

```{r poly5}
```

```{r poly6}
```

```{r poly7}
```

```{r poly8}
```

```{r poly9}
```

```{r poly10}
```

```{r poly11}
```

```{r poly12}
```

```{r poly13}
```

```{r poly14}
```

<div class="notes">

Here's code for fitting the remaining trend lines.

</div>

[dasl2]: https://dasl.datadescription.com/datafile/hurricane-frequencies/
[dasl1]: https://dasl.datadescription.com/

### 2021-12-12 text/21/hurricane-frequencies.Rmd
Updated hurricane-frequencies.Rmd

@@
 
-```{r notes}
+```{r notes, echo=FALSE}
 # Note: this code can produce a reasonable PowerPoint
@@
 <div class="notes">
-
-This [dataset on hurricane frequency][dasl1] is found at the [Data and Story Library website][dasl2].
 
@@
 add_trend <- function(point_plot, df, degree) {
-  x_grid <- data.frame(x=seq(1850, 2010, by=0.1))
+  x_grid <- data.frame(x=seq(1855, 2005, by=0.1))
   lm(y~poly(x, degree), data=df) %>%
@@
   point_plot +
-    geom_line(data=yhat, aes(x, .fitted), color="red") +
+    geom_line(data=yhat, aes(x, .fitted)) +
     geom_label(
@@
 
-### Code that I used
+### Code (load libraries)
 
@@
 
-### Code that I used
+### Code (read and print)
 
@@
 ```
-
-<div class="notes">
-
-This code reads data directly from the website and prints out the first eight rows.
-
-</div>
-
-### Code that I used
 
@@
 
-This code is for printing the rest of the data.
+This code reads data directly from the website and prints out the data in two pieces.
 
@@
 
-### Code that I used
+### Code (plot data points)
 
@@
 ```
-
-<div class="notes">
-
-It took a bit of work to put everything in a function, but now you can produce a quadratic rather than a linear trend with just a single line of code.
-
-</div>
-
-### Code that I used
 
@@
 
-Here's code for fitting the remaining trend lines.
+It took a bit of work to put everything in a function, but now you can produce a quadratic or higher level trend rather than a linear trend with just a single line of code.
 

### 2021-12-12 xxtemp.Rmd
Updated hurricane-frequencies.Rmd

---
title: "Two page example for Powerpoint"
author: "Steve Simon"
date: "12/11/2021"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

### R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure, echo=FALSE}
plot(pressure)
```

## test

```{r}
plot(1:4, 2:5)
```

## Column format

:::::: {.columns}
::: {.column}
Content of the left column.

```{r}
rnorm(10)
```
:::

::: {.column}
Content of the right column.

```{r, echo=FALSE}
plot(1:4, 2:5)
```

:::
::::::

### 2021-12-30 text/21/dark-side.Rmd
completed work on nhanes-overview.md
Added federated-learning.md

@@
 tags:
-- Big data
+- Data science
 - Incomplete page
@@
 
-YouTube. Leading Questions- Yes Prime Minister. Taken from the 1st Season of Yes Prime Minister – Episode 2, The Ministerial Broadcast. 2:16 running time. Published on Jan 15, 2012.
+YouTube. Leading Questions - Yes Prime Minister. Taken from the 1st Season of Yes Prime Minister, Episode 2, The Ministerial Broadcast. 2:16 running time. Published on Jan 15, 2012.
 

### 2021-12-30 text/21/negative-binomial-regression.Rmd
Fixed typos in tags

@@
 - Probability distributions
-- Generalized linear models
+- Generalized linear model
 output: html_document

### 2022-01-02 src/yearly-archive.R
Added yearly-archive.R

@@ -1,0 +1,54 @@
+# yearly-archive.R. This program was written by Steve Simon on 2021-01-02.
+
+suppressMessages(suppressWarnings(library(base64enc)))
+suppressMessages(suppressWarnings(library(curl     )))
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+old_path <- "c:/users/steve/dropbox/professional/fun stuff/update-website/"
+new_path <- "text/"
+archive_new <- function(yr) {  
+  new_path %0% 
+    yr %>%
+    list.files %>%
+    str_subset("md$") %>%
+    str_replace(fixed(".Rmd"), "") %>%
+    str_replace(fixed(".md"), "") -> new_files
+  cat("\n\nNew website (")
+  cat(length(new_files))
+  cat(" files)\n\n")
+  "+ [" %0% 
+    new_files %0%
+    "](http://new.pmean.com/" %0%
+    new_files %0%
+    "/)" %>%
+    cat(sep="\n")
+}
+archive_old <- function(yr) {
+  old_path %0% 
+    yr %>%
+    list.files -> old_files
+  cat("\n\nOld website\n\n")
+  "+ [" %0% 
+    old_files %>%
+    str_replace(fixed(".html"), "") %>%
+    str_replace(fixed(".htm"), "") %0%
+    "](http://www.pmean.com/" %0%
+    yr %0%
+    "/" %0%
+    old_files %0%
+    ")" %>%
+    cat(sep="\n")
+}
+
+yrs <- c("2003", "2007", "2011", "2015", "2019", "2021")
+paste0(brack(yrs), paren("http://new.pmean.com/" %0% yrs %0% "/"), collapse=" |\n") %>% cat
+cat("\n\n")
+archive_new("21")

### 2022-01-02 src/copy_files.R
Added first 2022 page

@@ -16,7 +16,7 @@ v <- TRUE
 
 copy_files <- function(
   old_root="text", new_root="c:/Users/simons/Documents/b04/content",
-  subdir_list=c("99", zpad(0:21), "no"),
+  subdir_list=c("99", zpad(0:22), "no"),
   file_pattern="md$", v=TRUE) {
   k <- 0
   for (subdir in subdir_list) {

### 2022-01-04 src/adapt-old-pages.R
Added philosophy-emergency.md

@@ -1,0 +1,81 @@
+# adapt-old-pages.R. This program was written by Steve Simon on 2021-01-03.
+
+suppressMessages(suppressWarnings(library(base64enc)))
+suppressMessages(suppressWarnings(library(curl     )))
+suppressMessages(suppressWarnings(library(knitr    )))
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(magrittr )))
+suppressMessages(suppressWarnings(library(readr    )))
+suppressMessages(suppressWarnings(library(rmarkdown)))
+suppressMessages(suppressWarnings(library(stringr  )))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+old_path <- "c:/users/steve/dropbox/professional/fun stuff/update-website/"
+
+clean_title <- function(old_text) {
+  old_text %>%
+    str_subset("title>") -> found_lines
+  if (length(found_lines)==0) {next}
+  found_lines %>%
+    str_trim %>%
+    str_remove(fixed("<title>")) %>%
+    str_remove(fixed("</title>")) %>%
+    str_remove(fixed("Stats: ")) %>%
+    str_remove(fixed("StATS: ")) %>%
+    str_remove(fixed("P.Mean: ")) %>%
+    str_remove(fixed("PMean: ")) %>%
+    str_remove(fixed("P.Mean:&nbsp ")) %>%
+    str_remove(fixed("P.Mean &gt;&gt;")) %>%
+    paste(collapse=" ") %>%
+    str_squish %>%
+    return
+}
+
+make_new_text <- function(old_text) {
+  body_start <- str_which(old_text, fixed("<body>"))
+  body_end <- str_which(old_text, fixed("</body>"))
+  old_text[body_start:(body_end-1)] %>%
+    str_remove(fixed("<body>")) %>%
+    paste0(collapse="\n") %>%
+    return
+}
+
+adapt_old <- function(yr) {
+  old_path %0% 
+    yr %>%
+    list.files %>%
+    str_subset(fixed(".htm")) -> old_files
+  for (f in old_files) {
+    old_text <- readLines(old_path %s% yr %s% f, warn=FALSE)
+    old_text %>%
+      str_detect(fixed("This page has moved to  a")) %>%
+      any -> exclude_this_file
+    if (exclude_this_file) {next}
[…]

### 2022-01-04 src/standard_functions.R

@@ -6,7 +6,7 @@
 # that I use across multiple projects.
 
 if (!exists("verbose")) verbose <- TRUE
-verbose <- TRUE
+verbose <- FALSE
 
 suppressMessages(suppressWarnings(library(dplyr)))
 suppressMessages(suppressWarnings(library(magrittr)))

### 2022-01-07 text/22/mapping-longitude-latitude.Rmd
Started work on mapping-longitude-latitude.Rmd

---
title: "Mapping latitude and longitude"
author: "Steve Simon"
source: new
date: "2022-01-06"
category: 
- Blog post
tags:
- Geocoding
output: html_document
---

I'm trying to understand some of the basics of mapping and found an interesting example on [this website]

```{r usa-map}
library(maps); library(ggplot2); library(mapproj)
states <- map_data("state")
usamap <- ggplot(states, aes(x=long, y=lat, group=group)) +
  geom_polygon(fill="white", colour="black")
usamap + coord_map("mercator")
usamap + coord_map("azequalarea") 
```

It helps to peek at the first few rows of data.

```{r peek}
head(states)
```

The plot is based on longitude/latitude pairs. The first thing to realize is that a direct plot without the coor_map function fails spectacularly.

```{r epic-fail}
usamap <- ggplot(states, aes(x=long, y=lat, group=group)) +
  geom_polygon(fill="white", colour="black")
```

Additional references are [here][ucs1], [here][ear1], [here][mgi1], and [here][tid1].

[ucs1]: https://www.nceas.ucsb.edu/sites/default/files/2020-04/OverviewCoordinateReferenceSystems.pdf
[ear1]: https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/understand-epsg-wkt-and-other-crs-definition-file-types/
[mgi1]: https://mgimond.github.io/Spatial/coordinate-systems-in-r.html
[tid1]: https://ggplot2.tidyverse.org/reference/coord_map.html
[rpu1]: https://rpubs.com/Augusto86/697809

### 2022-01-07 text/22/mapping-longitude-latitude.Rmd
More work on mapping-longitude-latitude.md

@@
 
-The plot is based on longitude/latitude pairs. The first thing to realize is that a direct plot without the coor_map function fails spectacularly.
+The plot is based on longitude/latitude pairs. The first thing to realize is that a degree of longitude is smaller than a degree of latitude, except at the equator.
 
-```{r epic-fail}
-usamap <- ggplot(states, aes(x=long, y=lat, group=group)) +
-  geom_polygon(fill="white", colour="black")
+```{r plot3d}
+library(dplyr)
+library(magrittr)
+library(plot3D)
+draw_arc <- function(lon=c(0, 360), lat=c(-90, 90), co="black") {
+  n <- 100
+  if (length(lon)==1) {lon <- rep(lon, 2)}
+  if (length(lat)==1) {lat <- rep(lat, 2)}
+  lon <- seq(lon[1], lon[2], length=n)*pi/180
+  lat <- seq(lat[1], lat[2], length=n)*pi/180
+  # formulas from 
+  #  https://en.wikipedia.org/wiki/Spherical_coordinate_system
+  #  but swap theta and phi. Also lat=90-phi.
+  x <- cos(lat)*cos(lon)
+  y <- cos(lat)*sin(lon)
+  z <- sin(lat)
+  segments3D(
+    x[-n], y[-n], z[-n], 
+    x[-1], y[-1], z[-1], 
+    box=FALSE, add=TRUE, col=co)
+}
+# Draw some "invisible" points to assure a full view of the arcs.
+a <- c(-1, 1)
+a_grid <- sapply(mesh(a, a, a), as.vector)
+scatter3D(
+  a_grid[ , 1], 
+  a_grid[ , 2], 
+  a_grid[ , 3], 
+  col="white", 
+  box=FALSE)  
+for (lat in (-8:8)*10) {draw_arc(lat=lat, co="gray90")}
+for (lon in (1:18)*20) {draw_arc(lon=lon, co="gray90")}
+draw_arc(lat= 0, co="black")
+draw_arc(lat=40, co="red")
 ```
+
+The black circle shows longitude values from -180 to 180 at the equator.
+The circle at the equator has a perimeter of approximately 40,000 kilometers (25,000 miles).
+Each degree of longitude at the equator is roughly 111 kilometers.
+
+The red circle, at latitude 40 degrees north (roughly where Columbus Ohio lies) has a smaller perimeter, a bit less than 31,000 kilometers.
+So near Columbus, Ohio, a degree of longitude is only 85 kilometers.
+
+The further you go north, the smaller the changes. At a latitude of 70 degrees north (roughly the northernmost part of Alaska), each degree of longitude is only 38 kilometers.
+
+```{r square}
+scatter3D(
+  a_grid[ , 1], 
+  a_grid[ , 2], 
+  a_grid[ , 3], 
+  col="white", 
+  box=FALSE)  
+for (lat in (-8:8)*10) {draw_arc(lat=lat, co="gray90")}
+for (lon in (1:18)*20) {draw_arc(lon=lon, co="gray90")}
+draw_arc(lon=280, lat=c(40, 60), co="darkgreen")
+draw_arc(lon=320, lat=c(40, 60), co="darkgreen")
+draw_arc(lon=c(280, 320), lat=40, co="darkgreen")
+draw_arc(lon=c(280, 320), lat=60, co="darkgreen")
+```
+
+If you pulled out a section of the earth stretching from 100 degrees to 140 degrees longitude (the middle of South Dakota to the western edge of Nevada) and from 40 degrees longitude to about 60 degree longitude (the northern edge of Alberta, Canada), you'd see a shape like the one shown above.
+
+It's not a rectangle, because the northern edge is quite a bit shorter than the southern edge.
+But it's not a trapezoid either, because the boundaries curve in three dimensions.
+It's impossible to use any two dimensional shape to represent this curved surface.
 

### 2022-01-08 text/14/axis.Rmd
Added qqplot.Rmd to 2014

---
title: "Some simple examples of manipulating the axes in a graph"
author: "Steve Simon"
source: "http://www.pmean.com/axis.html"
date: "2014-01-01"
category: Blog post
tags: Sample size
output: html_document
---

I want to start with a simple artificial example, a cubic function. With this simple function, you will see different ways of displaying information on the x and y axes.

```{r basic-cubic}
x <- seq(-2,2,length=1000)
y <- x^3-2*x
plot(x,y)
```

Figure 1. Plot of a simple cubic function.

This is a classic cubic function with a local minimum at sqrt(2/3), local maximum at -sqrt(2/3) and zeros at -sqrt(2), 0, and +sqrt(2). Let's draw the tick marks at these locations instead.

```{r move-tick-marks}
xticks <- c(-sqrt(2),-sqrt(2/3),0,sqrt(2/3),sqrt(2))
plot(x,y,axes=FALSE)
axis(side=2)
axis(side=1,at=xticks)
```

Figure 2. Plot with different tick locations.

Let's make the labels fit better.

```{r nicer-labels}
plot(x,y,axes=FALSE)
axis(side=2)
axis(side=1,at=xticks,labels=round(xticks,3))
```

Figure 3. Plot with rounded values at tick marks.

If you like, you can use math expressions on your axes

```{r math-expressions}
plot(x,y,axes=FALSE)
xlabs <- c(expression(-sqrt(2)),expression(-sqrt(2/3)),expression(0),expression(sqrt(2/3)),expression(+sqrt(2)))
axis(side=2)
axis(side=1,at=xticks,labels=xlabs)
```

Figure 4. Plot with math symbols

You can have the axes cross at the middle. When you do, it no longer makes sense to have a tick mark at zero.

```{r crossed-axes}
xticks <- c(-sqrt(2),-sqrt(2/3),sqrt(2/3),sqrt(2))
xlabs <- c(expression(-sqrt(2)),expression(-sqrt(2/3)),expression(sqrt(2/3)),expression(+sqrt(2)))
plot(x,y,axes=FALSE)
axis(side=2,pos=0,at=c(-4,-2,2,4))
axis(side=1,at=xticks,labels=xlabs,pos=0)
```

Figure 5. Plot with crossed axes.

One important axis option is set in the plot function itself. You are able to put different labels on the x and y axes with the xlab and ylab arguments.

```{r new-labels}
plot(x,y,xlab="Abcissa",ylab="Ordinate")
```

Figure 6. Plot with axis labels

Finally, there are several options to control the display of axes that are found in the par function. Here are a couple of the most important ones.

The las argument in par changes the orientation of the tick mark labels.

```{r tick-label-orientation}
# las=0 is the default
par(mfrow=c(2,2))
par(las=0)
plot(x,y)
title("las=0: Both parallel")
par(las=1)
plot(x,y)
title("las=1: Both horizontal")
par(las=2)
plot(x,y)
title("las=2: Both perpendicular")
par(las=3)
plot(x,y)
title("las=3: Both vertical")
```

Figure 7. Plots with different tick label orientations

The pty argumet in par can also force the plot to be square.

```{r square-plot}
par(pty="s")
plot(x,y)
```

Figure 8. Plot with square dimensions

When you combine this with equal axis limits, it creates a graph where the line of unity (y=x) has a 45 degree slope. This is helpful when comparing two measurements on the same units.

```{r equal-axis-limits}
par(pty="s")
plot(x,y,xlim=c(-4,4),ylim=c(-4,4))
abline(a=0,b=1)
```

Figure 9. Plot with square dimensions and equal scales

You can find an [earlier version][sim1] of this page on my [old website][sim2].

[sim1]: http://www.pmean.com/14/axis.html
[sim2]: http://www.pmean.com

### 2022-01-08 text/14/qqplot.Rmd
Added qqplot.Rmd to 2014

---
title: "The qqplot"
author: "Steve Simon"
source: "http://www.pmean.com/qqplot.html"
date: "2014-01-01"
category: 
- Blog post
tags:
- Checking assumptions  
- Linear regression
- R software
output: html_document
---

This program was written by Steve Simon, http://www.pmean.com, and is available at http://www.pmean.com/14/qqplot.html.

See a list of my other R programs at http://www.pmean.com/category/Rsoftware.html

Many statistical procedures are based on the assumption that your data has a normal distribution. The normal probability plot is a useful graphical tool for asessing this assumption. This plot is also called the qqplot (quantile-quantile plot).

You can use the housing data set to illustrate the use of the normal probability prlot.
```{r read-and-display-data}
fn <- "http://www.pmean.com/00files/housing.txt"
al <- read.table(file=fn,header=TRUE)
head(al)
tail(al)
```

```{r descriptive-statistics}
al$age <- as.numeric(al$AgeYears)
summary(al$Price)
summary(al$SquareFeet)
summary(al$age)
summary(al$NumberFeatures)
summary(al$Northeast)
summary(al$CustomBuild)
summary(al$CornerLot)
```

The qqplot compares the data values to evenly spaced percentiles of the normal distribution. A straight line indicates that the normality assumption is reasonable. You should not over interpret minor deviations from linearity. A large deviation from linearity is an indication that the normality assumption may be questionable.

You can read more about this on my 2009 blog entry about normal probability plots.

```{r normal-probability-plot}
qqnorm(al$Price)
```

Figure 1. Normal probabilty plot for house prices.

In a linear model, the critical assumption is NOT that the predictor (independent) variables are normally distributed. It is NOT that the outcome (dependent) variable is normally distributed. It is that the residuals are normally distributed.

```{r regression model}
al.model <- lm(Price~SquareFeet+CustomBuild,data=al)
summary(al.model)
qqnorm(resid(al.model))
```

Figure 2. Normal probability plot of residuals

You can find an [earlier version][sim1] of this page on my [old website][sim2].

[sim1]: http://www.pmean.com/14/qqplot.html
[sim2]: http://www.pmean.com

[sim3]: http://www.pmean.com/09/NormalPlot.html

### 2022-01-08 text/14/wavelets.Rmd
Added qqplot.Rmd to 2014

---
title: "An introduction to wavelets using R"
author: "Steve Simon"
source: "http://www.pmean.com/wavelets.html"
date: "2014-01-01"
category: 
- Blog post
tags:
- Checking assumptions  
- Linear regression
- R software
output: html_document
---

### Introduction

Wavelets represent a method to decompose a signal or an image. It represents an alternative to the Fourier decomposition. It is useful in image compression, edge analysis and flexible de-noising of even complicated discontinuous functions. Although there are adaptations to more complex settings, wavelets work best when analyzing a signal that is sampled at evenly spaced time points.

To understand wavelets well, you need to refresh yourself on how the Fourier decomposition really works. Think about Fourier as a way to travel to Trigland, the land of sines and cosines.

There is data set with the wavelets package in R called "ecg" that represents 2048 evenly spaced values from a typical electrocardiogram. Here's the code to get and graph that signal

```{r read-ecg-data}
library("rwt")
library("wavelets")
data(ecg)
head(ecg)
tail(ecg)
```

The head and tail function show the first few and the last few values for the signal. The graph looks like

![Figure 1. Graph of ecg signal](http://pmean.com/news/images/201312/wave01.png)

You can ignore the units of the signal itself, as they are arbitrary (at least for this application), but the minimum value is -1.48 and the maximum is 1.44.

### Fourier transforms

You compute the Fourier decomposition using the fft command, which is part of the base R package. The letters stand for "fast fourier transform".

```{r apply-fft}
ft.ecg <- fft(ecg)/2048
head(ft.ecg)
tail(ft.ecg)
ift.ecg <- fft(ft.ecg,inverse=TRUE)
head(ift.ecg)
tail(ift.ecg)
```

The object ft.ecg contains the Fourier decomposition, and ift.ecg is the inverse of the decomposition. You need to scale the Fourier decomposition by the number of data points (2,048). The inverse gets the original signal (-0.104773, -0.081500, -0.116409,...) except that there is a "+0i" or "-0i" tacked on to the end.

The complex numbers in the Fourier decomposition are a short-hand for the mix of sines and cosines. The first number has only a real component and no imaginary component. The real component represent the mean of the entire signal.

```{r compute-mean}
mean(ecg)
ft.ecg[1]
```

For the remaining numbers, the real part represents the cosine function and the imaginary part represents the sine function. So 
`r ft.ecg[2]`

represents

`r Re(ft.ecg[2])` cos(2*pi*t)

and

`r Im(ft.ecg[2])` sin(2*pi*t).

The following code shows the graph of the cosine function, the sine function and the sum.

```{r graph-cosine-1, fig.height=1}
par(mar=c(2.6, 0, 0, 0))
t.unit <- (1:2048)/2048
cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
plot(t.unit, cos1, type="l", axes=FALSE)
axis(side=1, pos=0, at=(0:4)/4, labels=(0:4)*512)
```

> graph.single("wave02.png",cos1,"cosine\nfunction",c(-0.35,0.35))
null device
1
> sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
> graph.single("wave03.png",sin1,"sine\nfunction",c(-0.35,0.35))
null device
1
> sum1 <- cos1+sin1
> graph.single("wave04.png",sum1,"Sum of\nsine and\ncosine",c(-0.35,0.35))
null device
1

Here's the first graph.

ecg signal
Axis

The graph of the cosine function shows a peak halfway through (1024) and a trough at the end (or equivalently at the beginning). This is actually flipped around from the normal cosine function, but recall that the multiplier front of the cosine function is negative.

ecg signal
Axis

The graph of the sine function shows a peak at 1/4 of the the way through (512) and a trough 3/4 of the way through (1536).

ecg signal
Axis

The sum shows the peak for the sum that is somewhere between the peaks of the sine and cosine functions. It's actually slightly closer to the peak of the cosine function, but not by much. The trough is also between the two troughs. The relative weights given to the sine and cosine functions and the positive or negative values associated with these weights will control where the peak and trough occur for the sum. In other words, you can control the phase of the curve.

If you are familiar with the arctangent function, you can compute the location of peak/trough. The Mod function will tell you the amplitude.

> ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
> comp(round(2048*(atan(ft.ratio)+pi)/(2*pi)),which(sum1==max(sum1)),"Peak")
Peak
By hand 786
Using R 786
> comp(Mod(ft.ecg[2]),max(sum1),"Amplitude")
Amplitude
By hand 0.1931891
Using R 0.1931890

Now this was a lot of work to decode the second component of the Fourier decomposition. You might forget to scale by the length of the signal. You could reverse the sines and cosines. Yu might compute the sines and cosines on a range from -pi to pi instead of from 0 to 2pi. You might leave out an important minus sign. Trust me, I did all these things.

There's a faster and more convenient way. Let the computer do the work for you. Take the Fourier decomposition, zero out everything except the second component, and then invert the decomposition. This allows you to graph the second component directly.

> ft.comp2 <- rep(0,length(ft.ecg))
> ft.comp2[2] <- ft.ecg[2]
> comp2 <- Re(fft(ft.comp2,inverse=TRUE))
> mx.2 <- max(comp2)
> graph.single("wave05.png",comp2,"Second\nFourier\ncomponent",c(-0.35,0.35))
null device
1

This is the graph

ecg signal
Axis

You can use the same method to show the shape of the third Fourier component.

> ft.comp3 <- rep(0,length(ft.ecg))
> ft.comp3[3] <- ft.ecg[3]
> comp3 <- Re(fft(ft.comp3,inverse=TRUE))
> mx.3 <- max(comp3)
> graph.single("wave06.png",comp3,"Third\nFourier\ncomponent",1.8*mx.3*c(-1,1))
null device
1

Here is the graph

ecg signal
Axis

This component has a higher frequency, with two peaks rather than one. Do the same for the fourth Fourier component

> ft.comp4 <- rep(0,length(ft.ecg))
> ft.comp4[4] <- ft.ecg[4]
> comp4 <- Re(fft(ft.comp4,inverse=TRUE))
> mx.4 <- max(comp4)
> graph.single("wave07.png",comp4,"Fourth\nFourier\ncomponent",1.8*mx.4*c(-1,1))
null device
1

Here is the graph.

ecg signal
Axis

The fourth Fourier component has three peaks. The interesting thing is that the second and third and fourth Fourier components are orthogonal to one another. Actually, every Fourier component is orthogonal to every other Fourier component. Mathematical folks like "orthogonal" because it allows you to derive the mathematical properties of the decomposition with greater ease. For a practical person, what "orthogonal" means is that the information contained in one Fourier component does not "overlap" with that in any other Fourier component. It also means you can run the decomposition in any order (low frequency to high frequency, high frequency to low frequency, or even a haphazard frequency). Finally, orthogonality also preserves independence properties in certain settings.

The Fourier decomposition continues with components of higher and higher frequency. The following code computes the fiftieth component of the Fourier transform.

> ft.comp50 <- rep(0,length(ft.ecg))
> ft.comp50[50] <- ft.ecg[50]
> comp50 <- Re(fft(ft.comp50,inverse=TRUE))
> mx.50 <- max(comp50)
> mx <- 0.35*mx.50/mx.2
> graph.single("wave08.png",comp50,"Fiftieth\nFourier\ncomponent",c(-mx,mx))
null device
1

Here is the graph.

ecg signal
Axis

Here's the fiftieth Fourier component, and if you count carefully, you will get 49 peaks. Count them carefully. No, don't bother, I'll do it for you. Because that are packed so closely together, I will onlyshow you every third peak.

Counter
ecg signal
Axis

If you continue down the list, you get a higher and higher frequency or equivalently more and more peaks.
Now wavelets produce a deocomposition much like the Fourier decomposition but wavelets are different in one important way. Wavelets break the signal up into pieces and model signals within those pieces. The frequency of the wavelet signal is associated with the number of pieces. For high frequency wavelets, you break the signal up into a large number of pieces. For low frequency signals, you breah the signal up into fewer pieces.

There are at least two good R packages for wavelet docompositon, rwt and wavelets. Here's a brief overview of how rwt works.

> d4.rwt <- daubcqf(4)
> d4.rwt
$h.0
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

$h.1
[1] 0.1294095 0.2241439 -0.8365163 0.4829629

> wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
> names(wt.rwt)
[1] "y" "L"
> head(wt.rwt$y)
[,1]
[1,] 0.797545
[2,] 6.979219
[3,] -2.933776
[4,] -8.541192
[5,] 4.087582
[6,] 2.536293
> tail(wt.rwt$y)
[,1]
[2043,] 0.003011394
[2044,] -0.012342195
[2045,] 0.001102199
[2046,] -0.030707757
[2047,] -0.007824638
[2048,] 0.126029962
> wt.rwt$L
[1] 9
> iwt.rwt <- midwt(wt.rwt$y,daub4$h.0,L=9)
> names(iwt.rwt)
[1] "x" "L"
> head(iwt.rwt$x)
[,1]
[1,] -0.104773
[2,] -0.093136
[3,] -0.081500
[4,] -0.116409
[5,] -0.081500
[6,] -0.116409
> tail(iwt.rwt$x)
[,1]
[2043,] -0.430591
[2044,] -0.442227
[2045,] -0.477136
[2046,] -0.442227
[2047,] -0.477136
[2048,] -0.477136
> iwt.rwt$L
[1] 9

With the rwt package, you choose the particular wavelet decomposition using the daubcqf function. The wavelet decomposition uses the mdwt function and the inverse of the wavelet decomposition uses the midwt decomposition. The rwt is fairly minimal with a family of Duabechies wavelts, but no alternatives (such as coiflets). Don't ask me what a coiflet is.

The wavelets package is more comprehensive, but also more complex.

> d4.wavelets <- wt.filter("d4")
> d4.wavelets
An object of class "wt.filter"
Slot "L":
[1] 4

Slot "level":
[1] 1

Slot "h":
[1] -0.1294095 -0.2241439 0.8365163 -0.4829629

Slot "g":
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

Slot "wt.class":
[1] "Daubechies"

Slot "wt.name":
[1] "d4"

Slot "transform":
[1] "dwt"

> wt.wavelets <- dwt(ecg,filter="d4")
> slotNames(wt.wavelets)
[1] "W" "V" "filter" "level" "n.boundary"
[6] "boundary" "series" "class.X" "attr.X" "aligned"
[11] "coe"
> iwt.wavelets <- idwt(wt.wavelets)
> head(iwt.wavelets)
[1] -0.10477 -0.09314 -0.08150 -0.11641 -0.08150 -0.11641
> tail(iwt.wavelets)
[1] -0.43059 -0.44223 -0.47714 -0.44223 -0.47714 -0.47714

With the wavelets package, you choose the particular wavelet decomposition using the wt.filter function. The wavelet decomposition uses the dwt function and the inverse of the wavelet decomposition uses the idwt function. Notice that the wavelets package uses S4 objects.

I will use the rwt package for the rest of my examples. The S4 objects in wavelets are difficult to manipulate directly, and I could not get some of the hand calculations to match what this package computed. I'm sure the failure to get a match with my hand calculations is just because of my limitations.

I will also use the d4 or Daubechies 4 wavelet for these examples. Ingird Daubechies is one of the pioneers in wavelet theory and she developed a series of wavelets, d2, d4, d6, .. with certain optimal properties. The d2 wavelet, or the Haar wavelet, is the simplest, but I wanted to look at something just slightly more complex. So what does a d4 wavelet look like?

> graph.single("wave09.png",wt.comp(wt.rwt,5),"d4 wavelet")
null device
1
d4 wavelet
Axis

Now, why the weird appearance? The d4 wavelet optimizes three properties simultaneously. The d4 wavelet chosen to have compact support, both for the wavelet decomposition and the inverse function that reconstructs the wavelet. The d4 wavelet is orthogonal both within a frequency band and across frequency bands. Finally, the d4 wavelet is orthogonal to any linear function, which simplifies the ability of the wavelet transform to isolate linear signals.

There are four wavelets at this frequency.
> graph.single("wave10a.png",wt.comp(wt.rwt,5),"W9[1]")
null device
1
> graph.single("wave10b.png",wt.comp(wt.rwt,6),"W9[2]")
null device
1
> graph.single("wave10c.png",wt.comp(wt.rwt,7),"W9[3]")
null device
1
> graph.single("wave10d.png",wt.comp(wt.rwt,8),"W9[4]")
null device
1

Wavelet 1 of 4
Wavelet 2 of 4
Wavelet 3 of 4
Wavelet 4 of 4
Axis

You'll find out more details about where these wavelet coefficients fit in the wavelet decomposition in little bit. Notice how the last two wavelets wrap around. This is a complication of the edge effects of a signal, and there are several alternative ways to handle these effects: repeat the signal (assuming it is periodic), reflect the signal back, or pad with zeros. The last option is not recommended, and repeating the signal leads to a simple wrap-around effect that is easy to follow.

Here are the wavelets at the next higher frequency.

> graph.single("wave11a.png",wt.comp(wt.rwt, 9),"W8[1]")
null device
1
> graph.single("wave11b.png",wt.comp(wt.rwt,10),"W8[2]")
null device
1
> graph.single("wave11c.png",wt.comp(wt.rwt,11),"W8[3]")
null device
1
> graph.single("wave11d.png",wt.comp(wt.rwt,12),"W8[4]")
null device
1
> graph.single("wave11e.png",wt.comp(wt.rwt,13),"W8[5]")
null device
1
> graph.single("wave11f.png",wt.comp(wt.rwt,14),"W8[6]")
null device
1
> graph.single("wave11g.png",wt.comp(wt.rwt,15),"W8[7]")
null device
1
> graph.single("wave11h.png",wt.comp(wt.rwt,16),"W8[8]")
null device
1

Wavelet 1 of 8
Wavelet 2 of 8
Wavelet 3 of 8
Wavelet 4 of 8
Wavelet 5 of 8
Wavelet 6 of 8
Wavelet 7 of 8
Wavelet 8 of 8
Axis

and at the next highest frequency

> graph.single("wave12a.png",wt.comp(wt.rwt,17),"W7[1]")
null device
1
> graph.single("wave12b.png",wt.comp(wt.rwt,18),"W7[2]")
null device
1
> graph.single("wave12c.png",wt.comp(wt.rwt,19),"W7[3]")
null device
1
> graph.single("wave12d.png",wt.comp(wt.rwt,20),"W7[4]")
null device
1
> graph.single("wave12e.png",wt.comp(wt.rwt,21),"W7[5]")
null device
1
> graph.single("wave12f.png",wt.comp(wt.rwt,22),"W7[6]")
null device
1
> graph.single("wave12g.png",wt.comp(wt.rwt,23),"W7[7]")
null device
1
> graph.single("wave12h.png",wt.comp(wt.rwt,24),"W7[8]")
null device
1
> graph.single("wave12i.png",wt.comp(wt.rwt,25),"W7[9]")
null device
1
> graph.single("wave12j.png",wt.comp(wt.rwt,26),"W7[10]")
null device
1
> graph.single("wave12k.png",wt.comp(wt.rwt,27),"W7[11]")
null device
1
> graph.single("wave12l.png",wt.comp(wt.rwt,28),"W7[12]")
null device
1
> graph.single("wave12m.png",wt.comp(wt.rwt,29),"W7[13]")
null device
1
> graph.single("wave12n.png",wt.comp(wt.rwt,30),"W7[14]")
null device
1
> graph.single("wave12o.png",wt.comp(wt.rwt,31),"W7[15]")
null device
1
> graph.single("wave12p.png",wt.comp(wt.rwt,32),"W7[16]")
null device
1

Wavelet 1 of 16
Wavelet 2 of 16
Wavelet 3 of 16
Wavelet 4 of 16
Wavelet 5 of 16
Wavelet 6 of 16
Wavelet 7 of 16
Wavelet 8 of 16
Wavelet 9 of 16
Wavelet 10 of 16
Wavelet 11 of 16
Wavelet 12 of 16
Wavelet 13 of 16
Wavelet 14 of 16
Wavelet 15 of 16
Wavelet 16 of 16
Axis

Notice that each time the frequency doubles, the number of wavelets also doubles. This creates a pyrimidal cascade of wavelet coefficients.

The calculation of the wavelet decomposition is an interative process. First, you need the coefficients underlying the d4 wavelet. At the highest possible frequency, there are four coefficients.

> g <- d4.rwt$h.0
> g
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

These coefficients represent the "high pass" filter. There is a complementary "low pass" filter that you get by reversing the coefficients and changing the sign of every other value.

> h <- d4.rwt$h.1
> h
[1] 0.1294095 0.2241439 -0.8365163 0.4829629

These two filter pairs are known as a quadrature mirror filter because of the reflection. These two filters are orthonormal.

> dot.prod(g,g)
[1] 1
> dot.prod(h,h)
[1] 1
> dot.prod(g,h)
[1] 0

The wavelet decomposition has an overlap, and (thankfully) the first half and the second half pieces are also orthogonal. Remember, orthogonality allows you to derive the mathematical properties of the wavelet decomposition easily and it preserves independence under certain settings.

> dot.prod(g[1:2],g[3:4])
[1] 2.909201e-13
> dot.prod(h[1:2],h[3:4])
[1] 2.909201e-13
> dot.prod(g[1:2],h[3:4])
[1] 0
> dot.prod(h[1:2],g[3:4])
[1] 0

The d4 filter also has one more nice property. The second filter is orthogonal to any linear signal

> dot.prod(h,1:4)
[1] 1.000117e-12

This means that the linear portion of any signal is not filtered out, but is passed up the ladder of the decomposition. The d6 filter has six coefficients rather than four, and it does not filter out any linear or quadratic signal. The d8 filter has eight coefficients and it does not filter out any linear, quadratic, or cubic signal. In fact, it is this property involving polynomial signals that forms the basis of the Daubechies wavelets.

There's one more property that is not immediately apparent from looking at the d4 wavelet. Some wavelets have simple forms for decomposition, but they get messier for the inverse of the decomposition. The d4 wavelet uses the same number of coefficients (4) in the inverse as it does in the composition. In fact, as you will see below, it uses the exact same vectors.

Let's look at the first level of the wavelet decomposition.

> wt.1 <- mdwt(ecg,g,L=1)
> V1 <- wt.1$y[ 1:1024]
> W1 <- wt.1$y[1025:2048]

For conveniece, split the decomposition into the left half (1:1024) and the right half (1025:2048).

The V1 coefficients represent the running dot product of the signal with the g vector. You skip every other data point in the running dot product, and you rotate the dot product around to the end when reach the right hand side to avoid edge effects. The skipping past the second, fourth, sixth, etc. observation in the running dot product is described in some sources as down-sampling.

> comp(dot.prod(ecg, g,1),V1[1],"V1[1]")
V1[1]
By hand -0.1317145
Using R -0.1317145
> comp(dot.prod(ecg, g,3),V1[2],"V1[2]")
V1[2]
By hand -0.1399428
Using R -0.1399428
> comp(dot.prod(ecg, g,5),V1[3],"V1[3]")
V1[3]
By hand -0.148171
Using R -0.148171

The W1 coefficients represent the running dot product of the signal with the h vector, again with downsampling.

> comp(dot.prod(ecg,-h,1),W1[1],"W1[1]")
W1[1]
By hand 0.02247964
Using R 0.02247964
> comp(dot.prod(ecg,-h,3),W1[2],"W1[2]")
W1[2]
By hand 0.02468439
Using R 0.02468439
> comp(dot.prod(ecg,-h,5),W1[3],"W1[3]")
W1[3]
By hand -0.006023849
Using R -0.006023849

Notice that you need to negate the h function to get the proper values for W. This is one of those scaling details in wavelets that will drive you nuts.

A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).
> wt.2 <- mdwt(ecg,g,L=2)
> V2 <- wt.2$y[ 1: 512]
> W2 <- wt.2$y[513:1024]

Again, for convenience, separate out the first quarter (1:512) and the second quarter (513:1024) of the data.

> comp(dot.prod(V1, g,1),V2[1],"V2[1]")
V2[1]
By hand -0.1961169
Using R -0.1961169
> comp(dot.prod(V1, g,3),V2[2],"V2[2]")
V2[2]
By hand -0.1989215
Using R -0.1989215
> comp(dot.prod(V1, g,5),V2[3],"V2[3]")
V2[3]
By hand -0.1863624
Using R -0.1863624

The V2 coefficients represent the running dot product (with downsampling) of V1 and g.

> comp(dot.prod(V1,-h,1),W2[1],"W2[1]")
W2[1]
By hand -0.00920762
Using R -0.00920762
> comp(dot.prod(V1,-h,3),W2[2],"W2[2]")
W2[2]
By hand 0.003389415
Using R 0.003389415
> comp(dot.prod(V1,-h,5),W2[3],"W2[3]")
W2[3]
By hand -0.0003374554
Using R -0.0003374554

The W2 coefficients represent the running dot product (with downsampling) of V1 and h.

> all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
[1] TRUE

The second level of the wavelet decomposition preserves all the W1 coefficients in the back half of the wavelet decomposition.

> wt.2a <- mdwt(V1,g,L=1)
> length(wt.2a$y)
[1] 1024
> all(wt.2a$y==wt.2$y[1:1024])
[1] TRUE

You get the exact same wavelet decomposition if you apply a level 1 decomposition on V1.

Note the order in which these are stored: 512 V2's at the front, followed by 512 W2's, and 1,024 W1's bring up the rear. You don't need to hang onto V1, because (as we will see in a minute), you can reconstruct the 1.024 V1's from 512 V2's and the 512 W2's. Now, run the wavelet decomposition all the way up to level 3. This creates 256 V3's, 256 W3's, 512 W2's, and 1,024 W1's.

> wt.3 <- mdwt(ecg,g,L=3)
> W3 <- wt.3$y[257:512]
> V3 <- wt.3$y[ 1:256]

Again for convenience, split apart the decomposition. At this level split into the first eight (1:256) and the second eighth (257:512).

> comp(dot.prod(V2, g,1),V3[1],"V3[1]")
V3[1]
By hand -0.28061
Using R -0.28061
> comp(dot.prod(V2, g,3),V3[2],"V3[2]")
V3[2]
By hand -0.2404948
Using R -0.2404948
> comp(dot.prod(V2, g,5),V3[3],"V3[3]")
V3[3]
By hand -0.2919882
Using R -0.2919882

Are you starting to see the pattern? The V3 coefficients represent the running dot product (with downsampling) of V2 and g.

> comp(dot.prod(V2,-h,1),W3[1],"W3[1]")
W3[1]
By hand -0.002777659
Using R -0.002777659
> comp(dot.prod(V2,-h,3),W3[2],"W3[2]")
W3[2]
By hand 0.03857375
Using R 0.03857375
> comp(dot.prod(V2,-h,5),W2[3],"W3[3]")
W3[3]
By hand 0.0963032950
Using R -0.0003374554

Ditto for the W3 coefficients.

> all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
[1] TRUE

The W2 coefficients are stored in the same spot for the level 2 and the level 3 decomposition.

> wt.3a <- mdwt(V2,g,L=1)
> length(wt.3a)
[1] 2
> all(wt.3a$y==wt.3$y[1:512])
[1] TRUE

Finally, you get the exact same values by applying a level 1 decomposition to V2.

You keep on going until you get just about run out of V's.

wt.4 <- mdwt(ecg,g,L=4)
W4 <- wt.4$y[129:256]
V4 <- wt.4$y[ 1:128]
wt.5 <- mdwt(ecg,g,L=5)
W5 <- wt.5$y[ 65:128]
V5 <- wt.5$y[ 1: 64]
wt.6 <- mdwt(ecg,g,L=6)
W6 <- wt.6$y[ 33: 64]
V6 <- wt.6$y[ 1: 32]
wt.7 <- mdwt(ecg,g,L=7)
W7 <- wt.7$y[ 17: 32]
V7 <- wt.7$y[ 1: 16]
wt.8 <- mdwt(ecg,g,L=8)
W8 <- wt.8$y[ 9: 16]
V8 <- wt.8$y[ 1: 8]

A level 8 decomposition works pretty well. It has 8 V8's, 8 W8's, 16 W7's, 32 W6's, 64 W5's, 128 W4's, 256 W3's, 512 W2's and 1,024 W1's. The W's store all the information at 8 decreasing frequency levels, and as the frequency decreases from the highest frequency (W1) to lower frequencies, the number of coefficients decreases also. We could take this a couple of steps further, but there is no appreciable benefit to running the wavelet decomoposition all the way to the top, and the edge effects loom very large at the top.

Reconstruction of the original signal works in a similar fashion.

> W8d <- dilute(W8)
> V8d <- dilute(V8)
> comp( dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
> comp(-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
> comp( dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
> comp(-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
> comp( dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
> comp(-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

You dilute the V8 and W8 coefficients by inserting zeros between every element, compute a running dot product of both the g and the h vectors, and add the results together. For some reason, you have to multiply every other value by -1. I don't know why. It's one of those scaling things. Also notice that you start at almost the very end of V8. This is caused by edge effects.

Once you have V7, repeat the process on the diluted V7 and W7 vectors to get V6.

> W7d <- dilute(W7)
> V7d <- dilute(V7)
> comp( dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
> comp(-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
> comp( dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
> comp(-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
> comp( dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
> comp(-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

And so forth.

> W1d <- dilute(W1)
> V1d <- dilute(V1)
> comp( dot.prod(W1d,g,2046)+dot.prod(V1d,h,2046),ecg[1],"ecg[1]")
ecg[1]
By hand -0.104773
Using R -0.104773
> comp(-dot.prod(W1d,g,2047)-dot.prod(V1d,h,2047),ecg[2],"ecg[2]")
ecg[2]
By hand -0.093136
Using R -0.093136
> comp( dot.prod(W1d,g,2048)+dot.prod(V1d,h,2048),ecg[3],"ecg[3]")
ecg[3]
By hand -0.0815
Using R -0.0815
> comp(-dot.prod(W1d,g, 1)-dot.prod(V1d,h, 1),ecg[4],"ecg[4]")
ecg[4]
By hand -0.116409
Using R -0.116409
> comp( dot.prod(W1d,g, 2)+dot.prod(V1d,h, 2),ecg[5],"ecg[5]")
ecg[5]
By hand -0.0815
Using R -0.0815
> comp(-dot.prod(W1d,g, 3)-dot.prod(V1d,h, 3),ecg[6],"ecg[6]")
ecg[6]
By hand -0.116409
Using R -0.116409

At the very end, the calculations on the diluted V1 and W1 values give you the original signal.

You can see what the wavelet coefficients at each level contribute by zeroing out some of the components and then inverting it.

V8 and W8
W7
W6
W5
W4
W3
W2
W1
ecg
Axis

Notice that the sharp spikes are captured in W1, W2, and W3. The slower overall undulation is captured by V8 and W8. Some of the intermediate squiggles between the sharp spikes are captured in the remaining wavelets.

There are simple generalizations of wavelets for two dimensional images. Here's a simple image, which represents a 256 by 256 bit grayscale image of a dog.

img <- read.table("klaus256.txt")
img <- 2*img-1

img.plot <- function(imat,n=256) {
xsq <- c(0,0,1,1)
ysq <- c(0,1,1,0)
par(mar=rep(0,4),xaxs="i",yaxs="i")
plot(c(1,n+1),c(1,n+1),type="n",axes=FALSE)
for (i in 1:n) {
for (j in 1:n){
d <- 0.5*imat[n+1-j,i]+0.5
d <- min(1,max(0,d))
polygon(xsq+i,ysq+j,density=-1,border=NA,col=rgb(d,d,d))
}
}
}
img.plot(img)

Dog image

Let's also move to a slightly lengthier wavelet, the d8 wavelet.

> g <- daubcqf(8)$h.0
> g
[1] 0.23037781 0.71484657 0.63088077 -0.02798377 -0.18703481 0.03084138
[7] 0.03288301 -0.01059740
> h <- -daubcqf(8)$h.1
> h
[1] -0.01059740 -0.03288301 0.03084138 0.18703481 -0.02798377 -0.63088077
[7] 0.71484657 -0.23037781

The d8 wavelet is a bit smoother than d4. It passes linear, quadratic, and cubic functions through to the top of the pyramid, unlike the d4 wavelet which only passes linear functions through. There is slightly greater computational complexity and edge effects can be a bit more pronounced, but these are minor issues.

In one dimention, the d8 wavelet looks like this.

ecg
Axis

For a two dimensional image, you take the outer product of g with itself to get an 8 by 8 matrix. This matrix is then slid around on the image to get the V1 coefficients. The W1 coefficents are produced by a similar set of three outer products: the outer product of h with g, the outer product of g with h and the outer product of h with h.

> wt.img.1 <- mdwt(img,g,L=1)
> comp(sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
V1
By hand -1.580973
Using R -1.580973
> comp(sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
W1
By hand -0.002825861
Using R -0.002825861
> comp(sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
W1
By hand -0.01225749
Using R -0.01225749
> comp(sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
W1
By hand -0.005042734
Using R -0.005042734

The first level of the 2D wavelet decomposition for a 256*256 image produces 128*128 V1 coefficients and 3*128*128 W1 coefficients. The second level produces 64*64 V2 coefficients and 3*64*64 W2 coefficients. If you continue this to the sixth level, you get 4*4 V6 coefficients and 3*4*4 W6 coefficients. This continues until the number of V coefficients becomes small. This is a pyrimidal cascade again, but this time a cascade in two dimensions.

Wavelet coefficients arrangement

Now there are three different sets of W's, one associated with outer(h,g), one with outer(g,h), and one with outer(h,h). The first represents high frequency changes in the horizontal direction and the second represents high frequency changes in the vertical direction. The third represents high frequency changes in a diagonal direction. In an image, a high frequency event represents a sharp transition between light to dark or dark to light. These transitions frequently represent boundaries or edges in an image.

Although there is some arbitrariness here, you can assign the outer(h,g) wavelets along the top.

Layout of horizontal wavelets

the outer(g,h) wavelets along the side

Wavelet layout

and the outer(h,h) wavelets along the diagonal.

Wavelet layout

If you examine the contribution of each wavelet frequency to the image, you will see a distinct patter.

W1

This W1. The gray areas represent the average color and regions of the image where no high frequency events occur. You can see high frequency events around the dog's collar and the dog's face.

W2

This is W2. It looks a lot like W1 but is a bit broader.

W3

This is W3. You see a bit more activity in the background, representings shifts in the color of the background image.

W4

This is W4. It looks like a blurred image with undulating colors rather than sharp shifts.

W5

This is W5. It is an amorphous blob representing graduals shifts in the average color.

W6

This is W6. It is just like W5 at an even broader scale.

If you look at the original image again, you will see regions (mostly in the background surrounding the dog or in large patches of the dog's fur) where the color is mostly the same. These are captured by the low frequency wavelets. The sudden transitions in color, especially around the dog's collar, are captured in high frequency wavelets. You can look at the cumulative effect of the wavelets as well.

W2

Here is W1 again.

W1 and W2

Here is W1 plus W2.

W1 through W3

Here is W1 through W3. Notice that the edges around the collar and the dog's face are well defined here. Also notice the large patches of gray. Gray is zero and it represents the average color between white and black. The areas in gray have no sharp edges and are filled in primarily with the lower frequency wavelets.

W1 through W4

Here is W1 through W4. At this level, some of the broad color regions are starting to become apparent, though the image is still largely washed out.

W1 through W6

Here is W1 through W5. The image is very similar to the original, but there are some patches that V5 would be needed to fill in.

You can also look at the reconstructed image using V1 through V6.

V1

Here is V1. It is the full image with just a few very sharp edges removed by W1.

V2

Here is V2. With removal of another set of edges by W2, the picture starts to take on a slightly blurry feel.

V3

Here is V3. It has an even greater blur.

V4

Here is V4. The dog is blurred almost beyond recognition.

V5

Here is V5. There is nothing recognizable here except general patterns of dark and light through broad regions of the image.

V6

Finally, here is V6. It is pretty much the same as V6, but at an even broader scale.

You can also reconstruct the three different sets of wavelet coefficients separately.

W1 through W3 horizontal

Here is an image reconstructed with the W1 through W3 of horizontally oriented wavelets only.

W1 though W3 vertical

Here is the reconstruction using W1 through W3 of the vertically oriented wavelets only.

W1 through W3 diagonal

Here is the reconstruction using W1 through W3 of the diagonally oriented wavelets only.

There seems to be more detail in the horizontally oriented wavelets. That's because the mouth, the eye and the ear are mostly horizontal. The edge of the dog's body, however, is vertical and is prominent only in the second image. The collar is a mix of horizontal, vertical, and diagonal elements, so it is apparent in each of the three images.

There's a lot more to wavelets, but this should help get you started. Now that you know how to compute the wavelet decomposition in one and two dimensions, you are ready to start applying them to real situations.

There are several important applications of wavelets. First, the wavelet coefficients themselves tell you information about the nature of the signal or image. Are there high frequency events or low frequency events (or both) and are those restricted to a particular part of the signal.

Second, you can use wavelets to get a "de-noised" signal. Those small coefficients that are (presumably) just noise. If you reconstruct the signal or image without those small coefficients (by rounding them down to zero), you will often get a "cleaner" estimate of the signal. Wavelets do an especially good job of de-noising when the signal is not smooth. Smooth means a discontinuous first derivative, but in most applications it represents an abrupt turn in the signal. Wavelets even can work well with discontinuous signals.

In contrast, the Fourier decomposition has lots of problems with non-smooth, but especially discontinuous functions. The problems go under the general name of the Gibbs phenomenon.

Third, you can zero out the small coefficients and store only the large coefficients. This allows you to reconstruct a good approximation of the image while saving substantially on storage. This is really only used for images, because one dimensional signals, even very long ones just don't take up a lot of room. The jpg 2000 standard used wavelets for compression.

Fourth, wavelets can help you detect edges in an image. An edge is an abrupt transition from light to dark, or dark to light. The high frequency wavelets help you detect where the edges are in an image.

Wavelets have a couple important limitations. The first is a minor issue, but the wavelets need to be adapted when the number of points in your signal is not a simple power of 2 or when the two dimensions of length and width of your image are not both simple powers of 2. A more important problem is that wavelets rely on equally spaced sample of the signal or image. Subsampling or interpolation can sometimes help here.

You can find an [earlier version][sim1] of this page on my [old website][sim2].

[sim1]: http://www.pmean.com/14/wavelets.html
[sim2]: http://www.pmean.com

### 2022-01-09 text/14/wavelets.Rmd
More work on 2014 wavelets.Rmd

@@
 
+Here are some graphing functions I will need.
+
+```{r graphing-functions}
+graph_with_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
+  par(mar=c(2.6, 0, 0, 0))
+  plot((1:2048)/2048, x, 
+       xlim=c(0, 1.2), 
+       ylim=limits, 
+       type="l",
+       axes=FALSE)
+  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
+  text(1.1, x[2048], lab, adj=0.5)
+}
+graph_without_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
+  par(mar=c(0, 0, 0, 0))
+  plot((1:2048)/2048, x, 
+       xlim=c(0, 1.2), 
+       ylim=limits, 
+       type="l",
+       axes=FALSE)
+  text(1.1, x[2048], lab, adj=0.5)
+}
+```
+
 The head and tail function show the first few and the last few values for the signal. The graph looks like
 
-![Figure 1. Graph of ecg signal](http://pmean.com/news/images/201312/wave01.png)
+```{r graph-ecg, fig.height=2}
+graph_with_axis(ecg, "ecg\nsignal")
+```
+
+Figure 1. Graph of ecg signal
 
@@
 
+```{r needle-plot, eval=FALSE}
+par(mar=c(0,0,0,0))
+plot(1:2048, Mod(ft.ecg), axes=FALSE)
+plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
+text(Re(ft.ecg), Im(ft.ecg), 1:2048)
+```
+
 The object ft.ecg contains the Fourier decomposition, and ift.ecg is the inverse of the decomposition.
@@
 
-```{r graph-cosine-1, fig.height=1}
-par(mar=c(2.6, 0, 0, 0))
+```{r graph-cosine-1, fig.height=1.33}
 t.unit <- (1:2048)/2048
 cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
-plot(t.unit, cos1, type="l", axes=FALSE)
-axis(side=1, pos=0, at=(0:4)/4, labels=(0:4)*512)
+sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
+add1 <- cos1+sin1
+
+{
+  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(add1))
+  graph_without_axis(sin1, "Sine\ncomponent", limits=range(add1))
+  graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
+}
 ```
 
-> graph.single("wave02.png",cos1,"cosine\nfunction",c(-0.35,0.35))
-null device
-1
-> sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
-> graph.single("wave03.png",sin1,"sine\nfunction",c(-0.35,0.35))
-null device
-1
-> sum1 <- cos1+sin1
-> graph.single("wave04.png",sum1,"Sum of\nsine and\ncosine",c(-0.35,0.35))
-null device
-1
-
-Here's the first graph.
-
-ecg signal
-Axis
-
-The graph of the cosine function shows a peak halfway through (1024) and a trough at the end (or equivalently at the beginning).
-This is actually flipped around from the normal cosine function, but recall that the multiplier front of the cosine function is negative.
-
-ecg signal
-Axis
-
-The graph of the sine function shows a peak at 1/4 of the the way through (512) and a trough 3/4 of the way through (1536).
-
-ecg signal
-Axis
+The cosine function is a wave with a peak in the middle and a trough at the beginning (or equivalently, the end).
+This is actually the reverse of a cosine function, but only because the coefficient is negative.
+
+The sine function is a wave with a peak a quarter of the way through and trough three quarters of the way through.
+A weighted sum of sine and cosine fuctions produces a trough at a different location.
+Here is about a third of the way through.
+
+If you look at the original signal, the peak and trough roughly match the shape of the ecg curve, if you ignore the sudden spikes.
+
+```{r compare-to-original-curve, fig.height=1.33}
+{
+  graph_without_axis(add1, "Combined\ncomponent")
+  graph_without_axis(ecg, "Original\nsignal")
+}
+graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
+peak <- (3*pi/2+Arg(ft.ecg[2]))/(2*pi)
+amplitude <- Mod(ft.ecg[2])
+segments(peak, 0, peak, amplitude)
+segments(0, 0, 1, 0)
+peak
+atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))
+atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))+pi
+Arg(ft.ecg[2])
+(which.max(add1)-1)/2047
+amplitude
+max(add1)
+```
 
@@
 
-> ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
-> comp(round(2048*(atan(ft.ratio)+pi)/(2*pi)),which(sum1==max(sum1)),"Peak")
-Peak
-By hand 786
-Using R 786
-> comp(Mod(ft.ecg[2]),max(sum1),"Amplitude")
-Amplitude
-By hand 0.1931891
-Using R 0.1931890
+```{r}
+ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
+round(2048*(atan(ft.ratio)+pi)/(2*pi))
+which(add1==max(add1))
+```
 
@@
 
-> ft.comp2 <- rep(0,length(ft.ecg))
-> ft.comp2[2] <- ft.ecg[2]
-> comp2 <- Re(fft(ft.comp2,inverse=TRUE))
-> mx.2 <- max(comp2)
-> graph.single("wave05.png",comp2,"Second\nFourier\ncomponent",c(-0.35,0.35))
-null device
-1
+```{r, fig.height=1.33}
+ft.comp2 <- rep(0,length(ft.ecg))
+ft.comp2[2] <- ft.ecg[2]
+comp2 <- Re(fft(ft.comp2,inverse=TRUE))
+{
+  graph_without_axis(comp2, "Fourier\ncomponent", limits=range(add1))
+  lines((1:2048)/2048, add1, col="red", lty="dotted")
+}
+```
 

### 2022-01-10 text/14/wavelets.Rmd
Worked on wavelet section of wavelets.md in 2014

@@
 
+```{r setup}
+knitr::opts_chunk$set(fig.height=1.33)
+```
+
 ```{r read-ecg-data}
@@
 ```{r graphing-functions}
-graph_with_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
+graph_with_axis <- function(x, lab, limits=range(x)) {
   par(mar=c(2.6, 0, 0, 0))
@@
 }
-graph_without_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
+graph_without_axis <- function(x, lab, limits=range(x)) {
   par(mar=c(0, 0, 0, 0))
@@
 tail(ft.ecg)
+```
+
+This function produces complex numbers, which have an interpretation in terms of sine and cosine functions.
+
+```{r needle-plot, fig.height=5}
+par(mar=c(0,0,0,0))
+plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
+text(Re(ft.ecg), Im(ft.ecg), 1:2048)
+```
+
+Here is a plot of the coefficients on the complex plane. Notice that certain coefficients (1, 2, 4, 32, 2018, and 2046) are further from the origin.
+
+### Inverse Fourier transform
+
+```{r inverse-fft}
 ift.ecg <- fft(ft.ecg,inverse=TRUE)
@@
 
-```{r needle-plot, eval=FALSE}
-par(mar=c(0,0,0,0))
-plot(1:2048, Mod(ft.ecg), axes=FALSE)
-plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
-text(Re(ft.ecg), Im(ft.ecg), 1:2048)
-```
+You convert from the Fourier coefficients back to the original signal using the same fft function, but this time with an inverse=TRUE argument.
 
@@
 You need to scale the Fourier decomposition by the number of data points (2,048).
-The inverse gets the original signal (-0.104773, -0.081500, -0.116409,...) except that there is a "+0i" or "-0i" tacked on to the end.
+The inverse gets the original signal (`r paste0(ift.ecg[1:3], colapse=", ")`, ...) except that there is a "+0i" or "-0i" tacked on to the end.
 
@@
 
-```{r graph-cosine-1, fig.height=1.33}
+```{r graph-cosine-1}
 t.unit <- (1:2048)/2048
@@
 sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
-add1 <- cos1+sin1
+sum1 <- cos1+sin1
 
 {
-  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(add1))
-  graph_without_axis(sin1, "Sine\ncomponent", limits=range(add1))
-  graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
+  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(sum1))
+  graph_without_axis(sin1, "Sine\ncomponent", limits=range(sum1))
+  graph_without_axis(sum1, "Combined\ncomponent", limits=range(sum1))
 }
@@
 
-```{r compare-to-original-curve, fig.height=1.33}
+```{r compare-to-original-curve}
 {
-  graph_without_axis(add1, "Combined\ncomponent")
+  graph_without_axis(sum1, "Combined\ncomponent")
   graph_without_axis(ecg, "Original\nsignal")
 }
-graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
-peak <- (3*pi/2+Arg(ft.ecg[2]))/(2*pi)
-amplitude <- Mod(ft.ecg[2])
-segments(peak, 0, peak, amplitude)
-segments(0, 0, 1, 0)
-peak
-atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))
-atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))+pi
-Arg(ft.ecg[2])
-(which.max(add1)-1)/2047
-amplitude
-max(add1)
 ```
@@
 
-If you are familiar with the arctangent function, you can compute the location of peak/trough. The Mod function will tell you the amplitude.
-
-```{r}
-ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
-round(2048*(atan(ft.ratio)+pi)/(2*pi))
-which(add1==max(add1))
-```
+If you are familiar with the Mod and Arg functions, they can help you compute the phase (the location of the peak/trough) and the amplitude (the height at the peak).
 
@@
 
-```{r, fig.height=1.33}
-ft.comp2 <- rep(0,length(ft.ecg))
-ft.comp2[2] <- ft.ecg[2]
+```{r second-component}
+ft.comp2 <- ft.ecg
+ft.comp2[-2] <- 0
 comp2 <- Re(fft(ft.comp2,inverse=TRUE))
+graph_without_axis(sum1, "Using\nInverse fft")
 {
-  graph_without_axis(comp2, "Fourier\ncomponent", limits=range(add1))
-  lines((1:2048)/2048, add1, col="red", lty="dotted")
-}
-```
-
-This is the graph
-
-ecg signal
-Axis
+  lines((1:2048)/2048, sum1, col="red", lty="dashed")
+}
+```
+
+This is the graph.
 
@@
 
-> ft.comp3 <- rep(0,length(ft.ecg))
-> ft.comp3[3] <- ft.ecg[3]
-> comp3 <- Re(fft(ft.comp3,inverse=TRUE))
-> mx.3 <- max(comp3)
-> graph.single("wave06.png",comp3,"Third\nFourier\ncomponent",1.8*mx.3*c(-1,1))
-null device
-1
-
-Here is the graph
-
-ecg signal
-Axis
+The third component has sines and cosines, but the period is two.
+
+`r Re(ft.ecg[3])` cos(2*pi*t/2)
+
+and
+
+`r Im(ft.ecg[3])` sin(2*pi*t/2).
+
+```{r graph-third-component}
+ft.comp3 <- ft.ecg
+ft.comp3[-3] <- 0
+comp3 <- Re(fft(ft.comp3,inverse=TRUE))
+graph_without_axis(comp3, "Fourier\ncomponent")
+```
 
@@
 
-> ft.comp4 <- rep(0,length(ft.ecg))
-> ft.comp4[4] <- ft.ecg[4]
-> comp4 <- Re(fft(ft.comp4,inverse=TRUE))
-> mx.4 <- max(comp4)
-> graph.single("wave07.png",comp4,"Fourth\nFourier\ncomponent",1.8*mx.4*c(-1,1))
-null device
-1
-
-Here is the graph.
-
-ecg signal
-Axis
+```{r graph-fourth-component}
+ft.comp4 <- ft.ecg
+ft.comp4[-4] <- 0
+comp4 <- Re(fft(ft.comp4,inverse=TRUE))
+graph_without_axis(comp4, "Fourier\ncomponent")
+```
 
@@
 
-The Fourier decomposition continues with components of higher and higher frequency. The following code computes the fiftieth component of the Fourier transform.
-
-> ft.comp50 <- rep(0,length(ft.ecg))
-> ft.comp50[50] <- ft.ecg[50]
-> comp50 <- Re(fft(ft.comp50,inverse=TRUE))
-> mx.50 <- max(comp50)
-> mx <- 0.35*mx.50/mx.2
-> graph.single("wave08.png",comp50,"Fiftieth\nFourier\ncomponent",c(-mx,mx))
-null device
-1
-
-Here is the graph.
-
-ecg signal
-Axis
-
-Here's the fiftieth Fourier component, and if you count carefully, you will get 49 peaks.
+The Fourier decomposition continues with components of higher and higher frequency. The following code computes the 32nd component of the Fourier transform.
+
+```{r thirty-second}
+ft.comp32 <- ft.ecg
+ft.comp32[-32] <- 0
+comp32 <- Re(fft(ft.comp32,inverse=TRUE))
+graph_without_axis(comp32, "Fourier\ncomponent")
+```
+
+Here's the 32nd Fourier component, and if you count carefully, you will get 31 peaks.
 Count them carefully.
 No, don't bother, I'll do it for you.
-Because that are packed so closely together, I will onlyshow you every third peak.
-
-Counter
-ecg signal
-Axis
+Because that are packed so closely together, I will only show you every third peak.
+
+```{r count}
+m <- max(comp32)
+graph_without_axis(comp32, "Fourier\ncomponent", limits=c(-m, 1.2*m))
+text((seq(1, 31, by=3)-0.5)/31, 1.2*m, seq(1, 31, by=3), col="red") 
+```
 
 If you continue down the list, you get a higher and higher frequency or equivalently more and more peaks.
+
+### Wavelets
+
 Now wavelets produce a deocomposition much like the Fourier decomposition but wavelets are different in one important way.
@@
 
-> d4.rwt <- daubcqf(4)
-> d4.rwt
-$h.0
-[1] 0.4829629 0.8365163 0.2241439 -0.1294095
-
-$h.1
-[1] 0.1294095 0.2241439 -0.8365163 0.4829629
-
-> wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
-> names(wt.rwt)
-[1] "y" "L"
-> head(wt.rwt$y)
-[,1]
-[1,] 0.797545
-[2,] 6.979219
-[3,] -2.933776
-[4,] -8.541192
-[5,] 4.087582
-[6,] 2.536293
-> tail(wt.rwt$y)
-[,1]
-[2043,] 0.003011394
-[2044,] -0.012342195
-[2045,] 0.001102199
-[2046,] -0.030707757
-[2047,] -0.007824638
-[2048,] 0.126029962
-> wt.rwt$L
-[1] 9
-> iwt.rwt <- midwt(wt.rwt$y,daub4$h.0,L=9)
-> names(iwt.rwt)
-[1] "x" "L"
-> head(iwt.rwt$x)
-[,1]
-[1,] -0.104773
-[2,] -0.093136
-[3,] -0.081500
-[4,] -0.116409
-[5,] -0.081500
-[6,] -0.116409
-> tail(iwt.rwt$x)
-[,1]
-[2043,] -0.430591
-[2044,] -0.442227
-[2045,] -0.477136
-[2046,] -0.442227
-[2047,] -0.477136
-[2048,] -0.477136
-> iwt.rwt$L
-[1] 9
+Perhaps the most famous of the wavelets is the Daubechies 4 wavelet.
+
+```{r rwt-package}
+d4.rwt <- daubcqf(4)
+d4.rwt
+wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
+names(wt.rwt)
+head(wt.rwt$y)
+tail(wt.rwt$y)
+wt.rwt$L
+iwt.rwt <- midwt(wt.rwt$y,d4.rwt$h.0,L=9)
+names(iwt.rwt)
+head(iwt.rwt$x)
+tail(iwt.rwt$x)
+iwt.rwt$L
+```
 
@@
 The wavelet decomposition uses the mdwt function and the inverse of the wavelet decomposition uses the midwt decomposition.
-The rwt is fairly minimal with a family of Duabechies wavelts, but no alternatives (such as coiflets).
+The rwt package is fairly minimal with a family of Duabechies wavelets, but no alternatives (such as coiflets).
 Don't ask me what a coiflet is.
@@
 
-> d4.wavelets <- wt.filter("d4")
-> d4.wavelets
-An object of class "wt.filter"
-Slot "L":
-[1] 4
-
-Slot "level":
-[1] 1
-
-Slot "h":
-[1] -0.1294095 -0.2241439 0.8365163 -0.4829629
-
-Slot "g":
-[1] 0.4829629 0.8365163 0.2241439 -0.1294095
-
-Slot "wt.class":
-[1] "Daubechies"
-
-Slot "wt.name":
-[1] "d4"
-
-Slot "transform":
-[1] "dwt"
-
-> wt.wavelets <- dwt(ecg,filter="d4")
-> slotNames(wt.wavelets)
-[1] "W" "V" "filter" "level" "n.boundary"
-[6] "boundary" "series" "class.X" "attr.X" "aligned"
-[11] "coe"
-> iwt.wavelets <- idwt(wt.wavelets)
-> head(iwt.wavelets)
-[1] -0.10477 -0.09314 -0.08150 -0.11641 -0.08150 -0.11641
-> tail(iwt.wavelets)
-[1] -0.43059 -0.44223 -0.47714 -0.44223 -0.47714 -0.47714
+```{r wavelets-package}
+d4.wavelets <- wt.filter("d4")
+d4.wavelets
+wt.wavelets <- dwt(ecg,filter="d4")
+slotNames(wt.wavelets)
+iwt.wavelets <- idwt(wt.wavelets)
+head(iwt.wavelets)
+tail(iwt.wavelets)
+```
 
@@
 
-> graph.single("wave09.png",wt.comp(wt.rwt,5),"d4 wavelet")
-null device
-1
-d4 wavelet
-Axis
+```{r graph-d4}
+wt.comp <- function(x, y) {
+  single_wavelet <- as.numeric((1:2048)==y)
+  as.vector(midwt(single_wavelet, d4.rwt$h.0, L=9)$x)
+}
+graph_without_axis(wt.comp(wt.rwt,5),"d4 wavelet")
+```
 
@@
 There are four wavelets at this frequency.
-> graph.single("wave10a.png",wt.comp(wt.rwt,5),"W9[1]")
-null device
-1
-> graph.single("wave10b.png",wt.comp(wt.rwt,6),"W9[2]")
-null device
-1
-> graph.single("wave10c.png",wt.comp(wt.rwt,7),"W9[3]")
-null device
-1
-> graph.single("wave10d.png",wt.comp(wt.rwt,8),"W9[4]")
-null device
-1
-
-Wavelet 1 of 4
-Wavelet 2 of 4
-Wavelet 3 of 4
-Wavelet 4 of 4
-Axis
+
+```{r w9}
+{
+  graph_without_axis(wt.comp(wt.rwt,5),"W9[1]")
+  graph_without_axis(wt.comp(wt.rwt,6),"W9[2]")
+  graph_without_axis(wt.comp(wt.rwt,7),"W9[3]")
+  graph_without_axis(wt.comp(wt.rwt,8),"W9[4]")
+}
+```
 
@@
 
-> graph.single("wave11a.png",wt.comp(wt.rwt, 9),"W8[1]")
-null device
-1
-> graph.single("wave11b.png",wt.comp(wt.rwt,10),"W8[2]")
-null device
-1
-> graph.single("wave11c.png",wt.comp(wt.rwt,11),"W8[3]")
-null device
-1
-> graph.single("wave11d.png",wt.comp(wt.rwt,12),"W8[4]")
-null device
-1
-> graph.single("wave11e.png",wt.comp(wt.rwt,13),"W8[5]")
-null device
-1
-> graph.single("wave11f.png",wt.comp(wt.rwt,14),"W8[6]")
-null device
-1
-> graph.single("wave11g.png",wt.comp(wt.rwt,15),"W8[7]")
-null device
-1
-> graph.single("wave11h.png",wt.comp(wt.rwt,16),"W8[8]")
-null device
-1
-
-Wavelet 1 of 8
-Wavelet 2 of 8
-Wavelet 3 of 8
-Wavelet 4 of 8
-Wavelet 5 of 8
-Wavelet 6 of 8
-Wavelet 7 of 8
-Wavelet 8 of 8
-Axis
+```{r w8}
+{
+  graph_without_axis(wt.comp(wt.rwt, 9),"W8[1]")
+  graph_without_axis(wt.comp(wt.rwt,10),"W8[2]")
+  graph_without_axis(wt.comp(wt.rwt,11),"W8[3]")
+  graph_without_axis(wt.comp(wt.rwt,12),"W8[4]")
+  graph_without_axis(wt.comp(wt.rwt,13),"W8[5]")
+  graph_without_axis(wt.comp(wt.rwt,14),"W8[6]")
+  graph_without_axis(wt.comp(wt.rwt,15),"W8[7]")
+  graph_without_axis(wt.comp(wt.rwt,16),"W8[8]")
+}
+```
 
@@
 
-> graph.single("wave12a.png",wt.comp(wt.rwt,17),"W7[1]")
-null device
-1
-> graph.single("wave12b.png",wt.comp(wt.rwt,18),"W7[2]")
-null device
-1
-> graph.single("wave12c.png",wt.comp(wt.rwt,19),"W7[3]")
-null device
-1
-> graph.single("wave12d.png",wt.comp(wt.rwt,20),"W7[4]")
-null device
-1
-> graph.single("wave12e.png",wt.comp(wt.rwt,21),"W7[5]")
-null device
-1
-> graph.single("wave12f.png",wt.comp(wt.rwt,22),"W7[6]")
-null device
-1
-> graph.single("wave12g.png",wt.comp(wt.rwt,23),"W7[7]")
-null device
-1
-> graph.single("wave12h.png",wt.comp(wt.rwt,24),"W7[8]")
-null device
-1
-> graph.single("wave12i.png",wt.comp(wt.rwt,25),"W7[9]")
-null device
-1
-> graph.single("wave12j.png",wt.comp(wt.rwt,26),"W7[10]")
-null device
-1
-> graph.single("wave12k.png",wt.comp(wt.rwt,27),"W7[11]")
-null device
-1
-> graph.single("wave12l.png",wt.comp(wt.rwt,28),"W7[12]")
-null device
-1
-> graph.single("wave12m.png",wt.comp(wt.rwt,29),"W7[13]")
-null device
-1
-> graph.single("wave12n.png",wt.comp(wt.rwt,30),"W7[14]")
-null device
-1
-> graph.single("wave12o.png",wt.comp(wt.rwt,31),"W7[15]")
-null device
-1
-> graph.single("wave12p.png",wt.comp(wt.rwt,32),"W7[16]")
-null device
-1
-
-Wavelet 1 of 16
-Wavelet 2 of 16
-Wavelet 3 of 16
-Wavelet 4 of 16
-Wavelet 5 of 16
-Wavelet 6 of 16
-Wavelet 7 of 16
-Wavelet 8 of 16
-Wavelet 9 of 16
-Wavelet 10 of 16
-Wavelet 11 of 16
-Wavelet 12 of 16
-Wavelet 13 of 16
-Wavelet 14 of 16
-Wavelet 15 of 16
-Wavelet 16 of 16
-Axis
+```{r s7}
+{
+  graph_without_axis(wt.comp(wt.rwt,17),"W7[1]")
+  graph_without_axis(wt.comp(wt.rwt,18),"W7[2]")
+  graph_without_axis(wt.comp(wt.rwt,19),"W7[3]")
+  graph_without_axis(wt.comp(wt.rwt,20),"W7[4]")
+  graph_without_axis(wt.comp(wt.rwt,21),"W7[5]")
+  graph_without_axis(wt.comp(wt.rwt,22),"W7[6]")
+  graph_without_axis(wt.comp(wt.rwt,23),"W7[7]")
+  graph_without_axis(wt.comp(wt.rwt,24),"W7[8]")
+  graph_without_axis(wt.comp(wt.rwt,25),"W7[9]")
+  graph_without_axis(wt.comp(wt.rwt,26),"W7[10]")
+  graph_without_axis(wt.comp(wt.rwt,27),"W7[11]")
+  graph_without_axis(wt.comp(wt.rwt,28),"W7[12]")
+  graph_without_axis(wt.comp(wt.rwt,29),"W7[13]")
+  graph_without_axis(wt.comp(wt.rwt,30),"W7[14]")
+  graph_without_axis(wt.comp(wt.rwt,31),"W7[15]")
+  graph_without_axis(wt.comp(wt.rwt,32),"W7[16]")
+}
+```
 
@@
 
-> g <- d4.rwt$h.0
-> g
-[1] 0.4829629 0.8365163 0.2241439 -0.1294095
+```{r high-pass}
+g <- d4.rwt$h.0
+g
+```
 
@@
 
-> h <- d4.rwt$h.1
-> h
-[1] 0.1294095 0.2241439 -0.8365163 0.4829629
+```{r low-pass}
+h <- d4.rwt$h.1
+h
+```
 
@@
 
-> dot.prod(g,g)
-[1] 1
-> dot.prod(h,h)
-[1] 1
-> dot.prod(g,h)
-[1] 0
+```{r orthonormal}
+dot.prod <- function(x, y, offset) {
+  sum(x*y)
+}
+dot.prod(g,g)
+dot.prod(h,h)
+dot.prod(g,h)
+```
 
@@
 
-> dot.prod(g[1:2],g[3:4])
-[1] 2.909201e-13
-> dot.prod(h[1:2],h[3:4])
-[1] 2.909201e-13
-> dot.prod(g[1:2],h[3:4])
-[1] 0
-> dot.prod(h[1:2],g[3:4])
-[1] 0
+```{r overlap}
+dot.prod(g[1:2],g[3:4])
+dot.prod(h[1:2],h[3:4])
+dot.prod(g[1:2],h[3:4])
+dot.prod(h[1:2],g[3:4])
+```
 
@@
 
-> dot.prod(h,1:4)
-[1] 1.000117e-12
+```{r orthogonal-to-linear}
+dot.prod(h,1:4)
+```
 
@@
 
-> wt.1 <- mdwt(ecg,g,L=1)
-> V1 <- wt.1$y[ 1:1024]
-> W1 <- wt.1$y[1025:2048]
+```{r first-level-wavelet}
+wt.1 <- mdwt(ecg,g,L=1)
+V1 <- wt.1$y[ 1:1024]
+W1 <- wt.1$y[1025:2048]
+V1[1]
+dot.prod(ecg[1:4], g)
+V1[2]
+dot.prod(ecg[3:6], g)
+W1[1]
+dot.prod(ecg[1:4], h)
+W1[2]
+dot.prod(ecg[3:6], h)
+wt.2 <- mdwt(ecg,g,L=2)
+V2 <- wt.2$y[1:512]
+W2 <- wt.2$y[513:1024]
+W1_check2 <- wt.2$y[1025:2048]
+V2[1]
+dot.prod(V1[1:4], g)
+V2[2]
+dot.prod(V1[3:6], g)
+W2[1]
+dot.prod(V1[1:4], h)
+W2[2]
+dot.prod(V1[3:6], h)
+
+level_wavelet <- function(wt.rwt, i) {
+  v <-wt.rwt$y
+  v[(2^(i-1)):(2^i)] <- 0
+  as.vector(midwt(v, d4.rwt$h.0, L=9)$x)
+}
+```
 
@@
 
-> comp(dot.prod(ecg, g,1),V1[1],"V1[1]")
+```{r down-sampling}
+dot.prod(ecg, g,1)
 V1[1]
-By hand -0.1317145
-Using R -0.1317145
-> comp(dot.prod(ecg, g,3),V1[2],"V1[2]")
+dot.prod(ecg, g,3)
 V1[2]
-By hand -0.1399428
-Using R -0.1399428
-> comp(dot.prod(ecg, g,5),V1[3],"V1[3]")
+dot.prod(ecg, g,5)
 V1[3]
-By hand -0.148171
-Using R -0.148171
+```
 
@@
 
-> comp(dot.prod(ecg,-h,1),W1[1],"W1[1]")
+```{r W1-coefficients}
+dot.prod(ecg,-h,1)
 W1[1]
-By hand 0.02247964
-Using R 0.02247964
-> comp(dot.prod(ecg,-h,3),W1[2],"W1[2]")
+dot.prod(ecg,-h,3)
 W1[2]
-By hand 0.02468439
-Using R 0.02468439
-> comp(dot.prod(ecg,-h,5),W1[3],"W1[3]")
+dot.prod(ecg,-h,5)
 W1[3]
-By hand -0.006023849
-Using R -0.006023849
+```
 
@@
 A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).
-> wt.2 <- mdwt(ecg,g,L=2)
-> V2 <- wt.2$y[ 1: 512]
-> W2 <- wt.2$y[513:1024]
-
-Again, for convenience, separate out the first quarter (1:512) and the second quarter (513:1024) of the data.
-
-> comp(dot.prod(V1, g,1),V2[1],"V2[1]")
+
+```{r W2-coefficients}
+wt.2 <- mdwt(ecg,g,L=2)
+V2 <- wt.2$y[ 1: 512]
+W2 <- wt.2$y[513:1024]
+dot.prod(V1, g,1)
 V2[1]
-By hand -0.1961169
-Using R -0.1961169
-> comp(dot.prod(V1, g,3),V2[2],"V2[2]")
+dot.prod(V1, g,3)
 V2[2]
-By hand -0.1989215
-Using R -0.1989215
-> comp(dot.prod(V1, g,5),V2[3],"V2[3]")
+dot.prod(V1, g,5)
 V2[3]
-By hand -0.1863624
-Using R -0.1863624
+```
 
@@
 
-> comp(dot.prod(V1,-h,1),W2[1],"W2[1]")
+```{r V2-coefficients}
+dot.prod(V1,-h,1)
 W2[1]
-By hand -0.00920762
-Using R -0.00920762
-> comp(dot.prod(V1,-h,3),W2[2],"W2[2]")
+dot.prod(V1,-h,3)
 W2[2]
-By hand 0.003389415
-Using R 0.003389415
-> comp(dot.prod(V1,-h,5),W2[3],"W2[3]")
+dot.prod(V1,-h,5)
 W2[3]
-By hand -0.0003374554
-Using R -0.0003374554
+```
 
@@
 
-> all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
-[1] TRUE
+```{r}
+all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
+```
 
@@
 
-> wt.2a <- mdwt(V1,g,L=1)
-> length(wt.2a$y)
+wt.2a <- mdwt(V1,g,L=1)
+length(wt.2a$y)
 [1] 1024
-> all(wt.2a$y==wt.2$y[1:1024])
+all(wt.2a$y==wt.2$y[1:1024])
 [1] TRUE
@@
 
-> wt.3 <- mdwt(ecg,g,L=3)
-> W3 <- wt.3$y[257:512]
-> V3 <- wt.3$y[ 1:256]
+wt.3 <- mdwt(ecg,g,L=3)
+W3 <- wt.3$y[257:512]
+V3 <- wt.3$y[ 1:256]
 
@@
 
-> comp(dot.prod(V2, g,1),V3[1],"V3[1]")
+dot.prod(V2, g,1),V3[1],"V3[1]")
 V3[1]
@@
 Using R -0.28061
-> comp(dot.prod(V2, g,3),V3[2],"V3[2]")
+dot.prod(V2, g,3),V3[2],"V3[2]")
 V3[2]
@@
 Using R -0.2404948
-> comp(dot.prod(V2, g,5),V3[3],"V3[3]")
+dot.prod(V2, g,5),V3[3],"V3[3]")
 V3[3]
@@
 
-> comp(dot.prod(V2,-h,1),W3[1],"W3[1]")
+dot.prod(V2,-h,1),W3[1],"W3[1]")
 W3[1]
@@
 Using R -0.002777659
-> comp(dot.prod(V2,-h,3),W3[2],"W3[2]")
+dot.prod(V2,-h,3),W3[2],"W3[2]")
 W3[2]
@@
 Using R 0.03857375
-> comp(dot.prod(V2,-h,5),W2[3],"W3[3]")
+dot.prod(V2,-h,5),W2[3],"W3[3]")
 W3[3]
@@
 
-> all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
+all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
 [1] TRUE
@@
 
-> wt.3a <- mdwt(V2,g,L=1)
-> length(wt.3a)
+wt.3a <- mdwt(V2,g,L=1)
+length(wt.3a)
 [1] 2
-> all(wt.3a$y==wt.3$y[1:512])
+all(wt.3a$y==wt.3$y[1:512])
 [1] TRUE
@@
 
-> W8d <- dilute(W8)
-> V8d <- dilute(V8)
-> comp( dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
+W8d <- dilute(W8)
+V8d <- dilute(V8)
+ dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
 V7[1]
@@
 Using R -3.315508
-> comp(-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
+-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
 V7[2]
@@
 Using R -2.398146
-> comp( dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
+ dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
 V7[3]
@@
 Using R -0.2769286
-> comp(-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
+-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
 V7[4]
@@
 Using R 1.782748
-> comp( dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
+ dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
 V7[5]
@@
 Using R 5.25007
-> comp(-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
+-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
 V7[6]
@@
 
-> W7d <- dilute(W7)
-> V7d <- dilute(V7)
-> comp( dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
+W7d <- dilute(W7)
+V7d <- dilute(V7)
+ dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
 V6[1]
@@
 Using R -2.096562
-> comp(-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
+-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
 V6[2]
@@
 Using R -2.294492
-> comp( dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
+ dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
 V6[3]
@@
 Using R -2.441797
-> comp(-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
+-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
 V6[4]
@@
 Using R -1.265345
-> comp( dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
+ dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
 V6[5]
@@
 Using R -0.6650157
-> comp(-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
+-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
 V6[6]
@@
 
-> W1d <- dilute(W1)
-> V1d <- dilute(V1)
-> comp( dot.prod(W1d,g,2046)+dot.prod(V1d,h,2046),ecg[1],"ecg[1]")
+W1d <- dilute(W1)
+V1d <- dilute(V1)
+ dot.prod(W1d,g,2046)+dot.prod(V1d,h,2046),ecg[1],"ecg[1]")
 ecg[1]
@@
 Using R -0.104773
-> comp(-dot.prod(W1d,g,2047)-dot.prod(V1d,h,2047),ecg[2],"ecg[2]")
+-dot.prod(W1d,g,2047)-dot.prod(V1d,h,2047),ecg[2],"ecg[2]")
 ecg[2]
@@
 Using R -0.093136
-> comp( dot.prod(W1d,g,2048)+dot.prod(V1d,h,2048),ecg[3],"ecg[3]")
+ dot.prod(W1d,g,2048)+dot.prod(V1d,h,2048),ecg[3],"ecg[3]")
 ecg[3]
@@
 Using R -0.0815
-> comp(-dot.prod(W1d,g, 1)-dot.prod(V1d,h, 1),ecg[4],"ecg[4]")
+-dot.prod(W1d,g, 1)-dot.prod(V1d,h, 1),ecg[4],"ecg[4]")
 ecg[4]
@@
 Using R -0.116409
-> comp( dot.prod(W1d,g, 2)+dot.prod(V1d,h, 2),ecg[5],"ecg[5]")
+ dot.prod(W1d,g, 2)+dot.prod(V1d,h, 2),ecg[5],"ecg[5]")
 ecg[5]
@@
 Using R -0.0815
-> comp(-dot.prod(W1d,g, 3)-dot.prod(V1d,h, 3),ecg[6],"ecg[6]")
+-dot.prod(W1d,g, 3)-dot.prod(V1d,h, 3),ecg[6],"ecg[6]")
 ecg[6]
@@
 
-> g <- daubcqf(8)$h.0
-> g
+g <- daubcqf(8)$h.0
+g
 [1] 0.23037781 0.71484657 0.63088077 -0.02798377 -0.18703481 0.03084138
 [7] 0.03288301 -0.01059740
-> h <- -daubcqf(8)$h.1
-> h
+h <- -daubcqf(8)$h.1
+h
 [1] -0.01059740 -0.03288301 0.03084138 0.18703481 -0.02798377 -0.63088077
@@
 
-> wt.img.1 <- mdwt(img,g,L=1)
-> comp(sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
+wt.img.1 <- mdwt(img,g,L=1)
+sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
 V1
@@
 Using R -1.580973
-> comp(sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
+sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
 W1
@@
 Using R -0.002825861
-> comp(sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
+sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
 W1
@@
 Using R -0.01225749
-> comp(sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
+sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
 W1
@@
 
+```{r save}
+save(list=ls(), file="wavelets.RData")
+```
+
 You can find an [earlier version][sim1] of this page on my [old website][sim2].

### 2022-01-11 text/14/wavelets.Rmd
Still more work on wavelets.md

@@
 }
+needle_plot <- function(y, lab, x=seq(0, 1, along.with=y)) {
+  par(mar=c(0, 0, 0, 0))
+  plot(
+    x, y, 
+    xlim=c(0, 1.2), 
+    type="n",
+    axes=FALSE)
+  segments(x, 0, x, y)
+  text(1.1, 0, lab)
+}
 ```
@@
 
-```{r needle-plot, fig.height=5}
+```{r complex-plot, fig.height=5}
 par(mar=c(0,0,0,0))
@@
 ```{r orthonormal}
-dot.prod <- function(x, y, offset) {
-  sum(x*y)
-}
-dot.prod(g,g)
-dot.prod(h,h)
-dot.prod(g,h)
+# The inline function %*% calculates a dot product of two vectors.
+g %*% g
+h %*% h
+g %*% h
 ```
@@
 ```{r overlap}
-dot.prod(g[1:2],g[3:4])
-dot.prod(h[1:2],h[3:4])
-dot.prod(g[1:2],h[3:4])
-dot.prod(h[1:2],g[3:4])
+g[1:2] %*% g[3:4]
+h[1:2] %*% h[3:4]
+g[1:2] %*% h[3:4]
+h[1:2] %*% g[3:4]
 ```
@@
 ```{r orthogonal-to-linear}
-dot.prod(h,1:4)
+h %*% (1:4)
 ```
@@
 
+For conveniece, split the decomposition into the left half (1:1024) and the right half (1025:2048).
+
+Information about the highest frequency events are captured in W1 and information aobut the remainder of the signal are captured in V1.
+Notice how the largest absolute values in W1 are aligned with some of the very sharp spikes in the signal.
+The large absolute values in V1 correspond to some of the spikes in the signal that are not quite at the highest frequency as well as low frequency oscillations.
+
 ```{r first-level-wavelet}
-wt.1 <- mdwt(ecg,g,L=1)
+wt.1 <- mdwt(ecg, g, L=1)
 V1 <- wt.1$y[   1:1024]
 W1 <- wt.1$y[1025:2048]
-V1[1]
-dot.prod(ecg[1:4], g)
-V1[2]
-dot.prod(ecg[3:6], g)
-W1[1]
-dot.prod(ecg[1:4], h)
-W1[2]
-dot.prod(ecg[3:6], h)
-wt.2 <- mdwt(ecg,g,L=2)
+{
+  needle_plot(V1, "V1")
+  needle_plot(W1, "W1")
+  graph_without_axis(ecg, "Original\nsignal")
+}
+```
+
+The V1 coefficients represent the running dot product of the signal with the g vector.
+You skip every other data point in the running dot product, and you rotate the dot product around to the end when reach the right hand side to avoid edge effects.
+The skipping past the second, fourth, sixth, etc.
+observation in the running dot product is described in some sources as down-sampling.
+
+The W1 coefficients represent the running dot product of the signal with the h vector, again with downsampling.
+
+```{r first-level-dot-products}
+c(V1[   1], g %*% ecg[1:4])
+c(V1[   2], g %*% ecg[3:6])
+c(V1[   3], g %*% ecg[5:8])
+c(V1[1023], g %*% ecg[2045:2048])
+c(V1[1024], g %*% ecg[c(2047:2048, 1:2)])
+
+c(W1[   1], -h %*% ecg[1:4])
+c(W1[   2], -h %*% ecg[3:6])
+c(W1[   3], -h %*% ecg[5:8])
+c(W1[1023], -h %*% ecg[2045:2048])
+c(W1[1024], -h %*% ecg[c(2047:2048, 1:2)])
+```
+
+Notice that you need to negate the h function to get the proper values for W. This is one of those scaling details in wavelets that will drive you nuts.
+
+A second pass applies the same pair of filters to V1.
+This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).
+The last 1,024 coefficients are the same highest frequency values (W1) that were computed in the first pass.
+
+```{r second-level-wavelet}
+wt.2 <- mdwt(ecg, g, L=2)
 V2 <- wt.2$y[1:512]
 W2 <- wt.2$y[513:1024]
-W1_check2 <- wt.2$y[1025:2048]
-V2[1]
-dot.prod(V1[1:4], g)
-V2[2]
-dot.prod(V1[3:6], g)
-W2[1]
-dot.prod(V1[1:4], h)
-W2[2]
-dot.prod(V1[3:6], h)
+quality_check_W1 <- wt.2$y[1025:2048]
+all(quality_check_W1==W1)
+{
+  needle_plot(V2, "V2")
+  needle_plot(W2, "W2")
+  needle_plot(W1, "W1")
+  graph_without_axis(ecg, "Original\nsignal")
+}
+```
 
+```{r second-level-dot-products}
+c(V2[  1], g %*% V1[1:4])
+c(V2[  2], g %*% V1[3:6])
+c(V2[  3], g %*% V1[5:8])
+c(V2[511], g %*% V1[1021:1024])
+c(V2[512], g %*% V1[c(1023:1024, 1:2)])
+
+c(W2[  1], -h %*% V1[1:4])
+c(W2[  2], -h %*% V1[3:6])
+c(W2[  3], -h %*% V1[5:8])
+c(W2[511], -h %*% V1[1021:1024])
+c(W2[512], -h %*% V1[c(1023:1024, 1:2)])
+```
+
+You could get the same results as a level 2 wavelet transform by applying a level 1 wavelet transform to V1.
+
+```{r transform-V1}
+quality_check <- mdwt(V1, g, L=1)
+quality_check_V2 <- quality_check$y[  1: 512]
+quality_check_W2 <- quality_check$y[513:1024]
+all(quality_check_V2==V2)
+all(quality_check_W2==W2)
+```
+
+Note the order in which these are stored: 512 V2's at the front, followed by 512 W2's, and 1,024 W1's bring up the rear.
+You don't need to hang onto V1, because (as we will see in a minute), you can reconstruct the 1.024 V1's from 512 V2's and the 512 W2's.
+Now, run the wavelet decomposition all the way up to level 3.
+This creates 256 V3's, 256 W3's, 512 W2's, and 1,024 W1's.
+
+A third pass applies the same pair of filters to V2.
+This creates 256 coefficients at the third highest frequency (W3) and 256 coefficients containing information about all the lower frequency events (V3).
+These 512 values are followed by the 512 W2 coefficients and the 1,024 W3 coefficients.
+
+```{r third-level-wavelet}
+wt.3 <- mdwt(ecg, g, L=3)
+V3 <- wt.3$y[1:256]
+W3 <- wt.3$y[257:512]
+W2_check3 <- wt.3$y[513:1024]
+W1_check3 <- wt.3$y[1025:2048]
+all(W2_check3==W2)
+all(W1_check3==W1)
+{
+  needle_plot(V3, "V3")
+  needle_plot(W3, "W3")
+  needle_plot(W2, "W2")
+  needle_plot(W1, "W1")
+  graph_without_axis(ecg, "Original\nsignal")
+}
+```
+
+Again, large coefficients in W3 are associated with more of the spikes in the original signal--spikes that are not quite as abrupt, perhaps, as the spikes captured in W1 and W2.
+
+The values of W3 and V3 are computed as before, but using V2 instead of V1 or the original signal.
+
+```{r third-level-dot-products}
+c(V3[  1], g %*% V2[1:4])
+c(V3[  2], g %*% V2[3:6])
+c(V3[  3], g %*% V2[5:8])
+c(V3[255], g %*% V2[509:512])
+c(V3[256], g %*% V2[c(511:512, 1:2)])
+
+c(W3[  1], -h %*% V2[1:4])
+c(W3[  2], -h %*% V2[3:6])
+c(W3[  3], -h %*% V2[5:8])
+c(W3[255], -h %*% V2[509:512])
+c(W3[256], -h %*% V2[c(511:512, 1:2)])
+```
+
+Here's an outline of how the wavelet transform works, level by level.
+Take the original 2,048 signal values and transform them into 1.024 very high frequency values (W1) and what's left over (V1).
+Take the 1,024 V1 values and split them into 512 high frequency values (W2) and what's left over (V2).
+Take those V2 values and split them.
+Keep on spliting until there are only a handful of V values left.
+I won't show the details of further calculation, but here is what the graph looks like.
+
+```{r ninth-level-wavelet}
+wt.9 <- mdwt(ecg, g, L=9)
+V9 <- wt.9$y[   1:   4]
+W9 <- wt.9$y[   5:   8]
+W8 <- wt.9$y[   9:  16]
+W7 <- wt.9$y[  17:  32]
+W6 <- wt.9$y[  33:  64]
+W5 <- wt.9$y[  65: 128]
+W4 <- wt.9$y[ 129: 256]
+W3 <- wt.9$y[ 257: 512]
+W2 <- wt.9$y[ 513:1024]
+W1 <- wt.9$y[1025:2048]
+{
+  needle_plot(V9, "V9")
+  needle_plot(W9, "W9")
+  needle_plot(W8, "W8")
+  needle_plot(W7, "W7")
+  needle_plot(W6, "W6")
+  needle_plot(W5, "W5")
+  needle_plot(W4, "W4")
+  needle_plot(W3, "W3")
+  needle_plot(W2, "W2")
+  needle_plot(W1, "W1")
+  graph_without_axis(ecg, "Original\nsignal")
+}
+```
+
+```{r}
 level_wavelet <- function(wt.rwt, i) {
@@
 
-For conveniece, split the decomposition into the left half (1:1024) and the right half (1025:2048).
-
-The V1 coefficients represent the running dot product of the signal with the g vector.
-You skip every other data point in the running dot product, and you rotate the dot product around to the end when reach the right hand side to avoid edge effects.
-The skipping past the second, fourth, sixth, etc.
-observation in the running dot product is described in some sources as down-sampling.
-
-```{r down-sampling}
-dot.prod(ecg, g,1)
-V1[1]
-dot.prod(ecg, g,3)
-V1[2]
-dot.prod(ecg, g,5)
-V1[3]
-```
-
-The W1 coefficients represent the running dot product of the signal with the h vector, again with downsampling.
-
-```{r W1-coefficients}
-dot.prod(ecg,-h,1)
-W1[1]
-dot.prod(ecg,-h,3)
-W1[2]
-dot.prod(ecg,-h,5)
-W1[3]
-```
-
-Notice that you need to negate the h function to get the proper values for W. This is one of those scaling details in wavelets that will drive you nuts.
-
-A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).
-
-```{r W2-coefficients}
-wt.2 <- mdwt(ecg,g,L=2)
-V2 <- wt.2$y[ 1: 512]
-W2 <- wt.2$y[513:1024]
-dot.prod(V1, g,1)
-V2[1]
-dot.prod(V1, g,3)
-V2[2]
-dot.prod(V1, g,5)
-V2[3]
-```
-
-The V2 coefficients represent the running dot product (with downsampling) of V1 and g.
-
-```{r V2-coefficients}
-dot.prod(V1,-h,1)
-W2[1]
-dot.prod(V1,-h,3)
-W2[2]
-dot.prod(V1,-h,5)
-W2[3]
-```
-
-The W2 coefficients represent the running dot product (with downsampling) of V1 and h.
-
-```{r}
-all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
-```
-
-The second level of the wavelet decomposition preserves all the W1 coefficients in the back half of the wavelet decomposition.
-
-wt.2a <- mdwt(V1,g,L=1)
-length(wt.2a$y)
-[1] 1024
-all(wt.2a$y==wt.2$y[1:1024])
-[1] TRUE
-
-You get the exact same wavelet decomposition if you apply a level 1 decomposition on V1.
-
-Note the order in which these are stored: 512 V2's at the front, followed by 512 W2's, and 1,024 W1's bring up the rear.
-You don't need to hang onto V1, because (as we will see in a minute), you can reconstruct the 1.024 V1's from 512 V2's and the 512 W2's.
-Now, run the wavelet decomposition all the way up to level 3.
-This creates 256 V3's, 256 W3's, 512 W2's, and 1,024 W1's.
-
-wt.3 <- mdwt(ecg,g,L=3)
-W3 <- wt.3$y[257:512]
-V3 <- wt.3$y[ 1:256]
-
-Again for convenience, split apart the decomposition. At this level split into the first eight (1:256) and the second eighth (257:512).
-
-dot.prod(V2, g,1),V3[1],"V3[1]")
-V3[1]
-By hand -0.28061
-Using R -0.28061
-dot.prod(V2, g,3),V3[2],"V3[2]")
-V3[2]
-By hand -0.2404948
-Using R -0.2404948
-dot.prod(V2, g,5),V3[3],"V3[3]")
-V3[3]
-By hand -0.2919882
-Using R -0.2919882
-
-Are you starting to see the pattern? The V3 coefficients represent the running dot product (with downsampling) of V2 and g.
-
-dot.prod(V2,-h,1),W3[1],"W3[1]")
-W3[1]
-By hand -0.002777659
-Using R -0.002777659
-dot.prod(V2,-h,3),W3[2],"W3[2]")
-W3[2]
-By hand 0.03857375
-Using R 0.03857375
-dot.prod(V2,-h,5),W2[3],"W3[3]")
-W3[3]
-By hand 0.0963032950
-Using R -0.0003374554
-
-Ditto for the W3 coefficients.
-
-all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
-[1] TRUE
-
-The W2 coefficients are stored in the same spot for the level 2 and the level 3 decomposition.
-
-wt.3a <- mdwt(V2,g,L=1)
-length(wt.3a)
-[1] 2
-all(wt.3a$y==wt.3$y[1:512])
-[1] TRUE
-
-Finally, you get the exact same values by applying a level 1 decomposition to V2.
-
-You keep on going until you get just about run out of V's.
-
-wt.4 <- mdwt(ecg,g,L=4)
-W4 <- wt.4$y[129:256]
-V4 <- wt.4$y[ 1:128]
-wt.5 <- mdwt(ecg,g,L=5)
-W5 <- wt.5$y[ 65:128]
-V5 <- wt.5$y[ 1: 64]
-wt.6 <- mdwt(ecg,g,L=6)
-W6 <- wt.6$y[ 33: 64]
-V6 <- wt.6$y[ 1: 32]
-wt.7 <- mdwt(ecg,g,L=7)
-W7 <- wt.7$y[ 17: 32]
-V7 <- wt.7$y[ 1: 16]
-wt.8 <- mdwt(ecg,g,L=8)
-W8 <- wt.8$y[ 9: 16]
-V8 <- wt.8$y[ 1: 8]
-
-A level 8 decomposition works pretty well.
-It has 8 V8's, 8 W8's, 16 W7's, 32 W6's, 64 W5's, 128 W4's, 256 W3's, 512 W2's and 1,024 W1's.
-The W's store all the information at 8 decreasing frequency levels, and as the frequency decreases from the highest frequency (W1) to lower frequencies, the number of coefficients decreases also.
-We could take this a couple of steps further, but there is no appreciable benefit to running the wavelet decomoposition all the way to the top, and the edge effects loom very large at the top.
+### Reconstruction of the original signal
 
@@
 
+First you dilute the W and V values at the top of the pyramid.
+
+```{r reconstuct V8}
+V9_diluted <- c(V9[1], 0, V9[2], 0, V9[3], 0, V9[4], 0)
+W9_diluted <- c(W9[1], 0, W9[2], 0, W9[3], 0, W9[4], 0)
+
+wt.8 <- mdwt(ecg, g, L=8)
+V8 <- wt.8$y[1:8]
+
+c(V8[1], rev(h) %*% V9_diluted[1:4] + rev(g) %*% W9_diluted[1:4])
+c(V8[2], rev(h) %*% V9_diluted[2:5] + rev(g) %*% W9_diluted[2:5])
+```
+
 W8d <- dilute(W8)
 V8d <- dilute(V8)
- dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
+ W8d, g, 14)+V8d,h,14),V7[1],"V7[1]")
 V7[1]
@@
 Using R -3.315508
--dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
+-W8d, g, 15)-V8d,h,15),V7[2],"V7[2]")
 V7[2]
@@
 Using R -2.398146
- dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
+ W8d, g, 16)+V8d,h,16),V7[3],"V7[3]")
 V7[3]
@@
 Using R -0.2769286
--dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
+-W8d, g,  1)-V8d,h, 1),V7[4],"V7[4]")
 V7[4]
@@
 Using R 1.782748
- dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
+ W8d, g,  2)+V8d,h, 2),V7[5],"V7[5]")
 V7[5]
@@
 Using R 5.25007
--dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
+-W8d, g,  3)-V8d,h, 3),V7[6],"V7[6]")
 V7[6]
@@
 V7d <- dilute(V7)
- dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
+ W7d, g, 30)+V7d,h,30),V6[1],"V6[1]")
 V6[1]
@@
 Using R -2.096562
--dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
+-W7d, g, 31)-V7d,h,31),V6[2],"V6[2]")
 V6[2]
@@
 Using R -2.294492
- dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
+ W7d, g, 32)+V7d,h,32),V6[3],"V6[3]")
 V6[3]
@@
 Using R -2.441797
--dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
+-W7d, g,  1)-V7d,h, 1),V6[4],"V6[4]")
 V6[4]
@@
 Using R -1.265345
- dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
+ W7d, g,  2)+V7d,h, 2),V6[5],"V6[5]")
 V6[5]
@@
 Using R -0.6650157
--dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
+-W7d, g,  3)-V7d,h, 3),V6[6],"V6[6]")
 V6[6]
@@
 V1d <- dilute(V1)
- dot.prod(W1d,g,2046)+dot.prod(V1d,h,2046),ecg[1],"ecg[1]")
+ W1d, g, 2046)+V1d,h,2046),ecg[1],"ecg[1]")
 ecg[1]
@@
 Using R -0.104773
--dot.prod(W1d,g,2047)-dot.prod(V1d,h,2047),ecg[2],"ecg[2]")
+-W1d, g, 2047)-V1d,h,2047),ecg[2],"ecg[2]")
 ecg[2]
@@
 Using R -0.093136
- dot.prod(W1d,g,2048)+dot.prod(V1d,h,2048),ecg[3],"ecg[3]")
+ W1d, g, 2048)+V1d,h,2048),ecg[3],"ecg[3]")
 ecg[3]
@@
 Using R -0.0815
--dot.prod(W1d,g, 1)-dot.prod(V1d,h, 1),ecg[4],"ecg[4]")
+-W1d, g,  1)-V1d,h, 1),ecg[4],"ecg[4]")
 ecg[4]
@@
 Using R -0.116409
- dot.prod(W1d,g, 2)+dot.prod(V1d,h, 2),ecg[5],"ecg[5]")
+ W1d, g,  2)+V1d,h, 2),ecg[5],"ecg[5]")
 ecg[5]
@@
 Using R -0.0815
--dot.prod(W1d,g, 3)-dot.prod(V1d,h, 3),ecg[6],"ecg[6]")
+-W1d, g,  3)-V1d,h, 3),ecg[6],"ecg[6]")
 ecg[6]
@@
 
-wt.img.1 <- mdwt(img,g,L=1)
+wt.img.1 <- mdwt(img, g, L=1)
 sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")

### 2022-01-11 text/14/wavelets.Rmd
Started inverse wavelet transform on wavelets.md

@@
 }
-needle_plot <- function(y, lab, x=seq(0, 1, along.with=y)) {
+needle_plot <- function(y, lab, x=seq(0, 1, along.with=y), limits=range(y)) {
   par(mar=c(0, 0, 0, 0))
@@
     xlim=c(0, 1.2), 
+    ylim=limits,
     type="n",
@@
 {
-  needle_plot(V1, "V1")
-  needle_plot(W1, "W1")
-  graph_without_axis(ecg, "Original\nsignal")
+  needle_plot(V1, "V1", limits=range(ecg))
+  needle_plot(W1, "W1", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
 }
@@
 {
-  needle_plot(V2, "V2")
-  needle_plot(W2, "W2")
-  needle_plot(W1, "W1")
-  graph_without_axis(ecg, "Original\nsignal")
+  needle_plot(V2, "V2", limits=range(ecg))
+  needle_plot(W2, "W2", limits=range(ecg))
+  needle_plot(W1, "W1", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
 }
@@
 {
-  needle_plot(V3, "V3")
-  needle_plot(W3, "W3")
-  needle_plot(W2, "W2")
-  needle_plot(W1, "W1")
-  graph_without_axis(ecg, "Original\nsignal")
+  needle_plot(V3, "V3", limits=range(ecg))
+  needle_plot(W3, "W3", limits=range(ecg))
+  needle_plot(W2, "W2", limits=range(ecg))
+  needle_plot(W1, "W1", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
 }
@@
 {
-  needle_plot(V9, "V9")
-  needle_plot(W9, "W9")
-  needle_plot(W8, "W8")
-  needle_plot(W7, "W7")
-  needle_plot(W6, "W6")
-  needle_plot(W5, "W5")
-  needle_plot(W4, "W4")
-  needle_plot(W3, "W3")
-  needle_plot(W2, "W2")
-  needle_plot(W1, "W1")
-  graph_without_axis(ecg, "Original\nsignal")
-}
-```
-
-```{r}
-level_wavelet <- function(wt.rwt, i) {
-  v <-wt.rwt$y
-  v[(2^(i-1)):(2^i)] <- 0
-  as.vector(midwt(v, d4.rwt$h.0, L=9)$x)
+  needle_plot(V9, "V9", limits=range(ecg))
+  needle_plot(W9, "W9", limits=range(ecg))
+  needle_plot(W8, "W8", limits=range(ecg))
+  needle_plot(W7, "W7", limits=range(ecg))
+  needle_plot(W6, "W6", limits=range(ecg))
+  needle_plot(W5, "W5", limits=range(ecg))
+  needle_plot(W4, "W4", limits=range(ecg))
+  needle_plot(W3, "W3", limits=range(ecg))
+  needle_plot(W2, "W2", limits=range(ecg))
+  needle_plot(W1, "W1", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
 }
@@
 c(V8[2], rev(h) %*% V9_diluted[2:5] + rev(g) %*% W9_diluted[2:5])
+V9_coefficients <- rep(0, 2048)
+V9_coefficients[1:8] <- wt.9$y[1:8]
+V9_signal <- as.vector(midwt(V9_coefficients, d4.rwt$h.0, L=9)$x)
+{
+  graph_without_axis(V9_signal, "V9", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
+}
 ```

### 2022-01-13 text/14/wavelets.Rmd
Added reference to wavelets.Rmd
Added second reference to wavelets.Rmd

@@
 
+The best explanation of the inverse wavelet transformation is a [page][kap1] buried deep in Ian Kaplan's website, bearcave. Also helpful is this [pdf file][nar1] of Chapter 2 of a book on wavelets.
+
 Reconstruction of the original signal works in a similar fashion.
@@
 
-In one dimention, the d8 wavelet looks like this.
+In one dimension, the d8 wavelet looks like this.
 
@@
 [sim2]: http://www.pmean.com
+
+[kap1]: http://bearcave.com/software/java/wavelets/daubechies/index.html
+[nar1]: http://dsp-book.narod.ru/PWSA/8276_02.pdf

### 2022-01-28 text/14/wavelets.Rmd
Added matrix notation to 14/wavelets.Rmd

@@
 ```{r setup}
-knitr::opts_chunk$set(fig.height=1.33)
+knitr::opts_chunk$set(fig.height=1)
 ```
@@
        xlim=c(0, 1.2), 
-       ylim=limits, 
+       ylim=c(limits[1]*1.1, limits[2]), # adjust lower limit
        type="l",
@@
 
+Actually, wavelets come in pairs, a "father wavelet" and a "mother wavelet."
+
 ```{r graph-d4}
@@
   single_wavelet <- as.numeric((1:2048)==y)
-  as.vector(midwt(single_wavelet, d4.rwt$h.0, L=9)$x)
+  as.vector(midwt(single_wavelet, d4.rwt$h.0, L=8)$x)
 }
-graph_without_axis(wt.comp(wt.rwt,5),"d4 wavelet")
+```
+
+```{r calculate-functions}
+calculate_mother <- function(i, L) {
+  yl <- rep(0, 2048)
+  yh <- matrix(0, nrow=2048, ncol=L)
+  yh[i, L] <- 1
+  as.vector(mirdwt(yl, yh, d4.rwt$h.0, L)$x)
+}
+calculate_father <- function(i, L) {
+  yl <- rep(0, 2048)
+  yh <- matrix(0, nrow=2048, ncol=L)
+  yl[i] <- 1
+  as.vector(mirdwt(yl, yh, d4.rwt$h.0, L)$x)
+}
+```
+
+```{r mother-and-father}
+{
+  graph_without_axis(calculate_father(2^9+1, 7),"d4 father")
+  graph_without_axis(calculate_mother(2^9+1, 7),"d4 mother")
+}
+```
+
+You can shift the wavelets left and right.
+
+```{r shifted}
+{
+  graph_without_axis(calculate_father(2^9-2^7+1, 7),"d4 father")
+  graph_without_axis(calculate_father(2^9+1, 7),"d4 father")
+  graph_without_axis(calculate_father(2^9+2^7+1, 7),"d4 father")
+  graph_without_axis(calculate_mother(2^9-2^7+1, 7),"d4 mother")
+  graph_without_axis(calculate_mother(2^9+1, 7),"d4 mother")
+  graph_without_axis(calculate_mother(2^9+2^7+1, 7),"d4 mother")
+}
+```
+
+When a wavelet crosses the boundary, there are several choices. The most common one is to wrap the wavelet around to the other side.
+
+```{r wrapped}
+{
+  graph_without_axis(calculate_father(2^10+2^9+1+2^8+1, 7),"d4 father")
+  graph_without_axis(calculate_mother(2^10+2^9+1+2^8+1, 7),"d4 mother")
+}
+
+```
+
+You can stretch or squeeze the wavelets.
+
+```{r stretch-and-squeeze}
+{  
+  graph_without_axis(calculate_father(2^9+1, 8),"d4 father")
+  graph_without_axis(calculate_father(2^9+1, 7),"d4 father")
+  graph_without_axis(calculate_father(2^9+1, 6),"d4 father")
+  graph_without_axis(calculate_mother(2^9+1, 8),"d4 mother")
+  graph_without_axis(calculate_mother(2^9+1, 7),"d4 mother")
+  graph_without_axis(calculate_mother(2^9+1, 6),"d4 mother")
+}
+```
+
+Here are the father and mother wavelets at the most extreme.
+
+```{r most-extreme}
+{
+  graph_without_axis(calculate_father(2^9+1, 1),"d4 father")
+  graph_without_axis(calculate_mother(2^9+1, 1),"d4 mother")
+}
 ```
@@
 
-There are four wavelets at this frequency.
-
-```{r w9}
-{
-  graph_without_axis(wt.comp(wt.rwt,5),"W9[1]")
-  graph_without_axis(wt.comp(wt.rwt,6),"W9[2]")
-  graph_without_axis(wt.comp(wt.rwt,7),"W9[3]")
-  graph_without_axis(wt.comp(wt.rwt,8),"W9[4]")
-}
-```
-
-You'll find out more details about where these wavelet coefficients fit in the wavelet decomposition in little bit.
-Notice how the last two wavelets wrap around.
-This is a complication of the edge effects of a signal, and there are several alternative ways to handle these effects: repeat the signal (assuming it is periodic), reflect the signal back, or pad with zeros.
-The last option is not recommended, and repeating the signal leads to a simple wrap-around effect that is easy to follow.
-
-Here are the wavelets at the next higher frequency.
-
-```{r w8}
-{
-  graph_without_axis(wt.comp(wt.rwt, 9),"W8[1]")
-  graph_without_axis(wt.comp(wt.rwt,10),"W8[2]")
-  graph_without_axis(wt.comp(wt.rwt,11),"W8[3]")
-  graph_without_axis(wt.comp(wt.rwt,12),"W8[4]")
-  graph_without_axis(wt.comp(wt.rwt,13),"W8[5]")
-  graph_without_axis(wt.comp(wt.rwt,14),"W8[6]")
-  graph_without_axis(wt.comp(wt.rwt,15),"W8[7]")
-  graph_without_axis(wt.comp(wt.rwt,16),"W8[8]")
-}
-```
-
-and at the next highest frequency
-
-```{r s7}
-{
-  graph_without_axis(wt.comp(wt.rwt,17),"W7[1]")
-  graph_without_axis(wt.comp(wt.rwt,18),"W7[2]")
-  graph_without_axis(wt.comp(wt.rwt,19),"W7[3]")
-  graph_without_axis(wt.comp(wt.rwt,20),"W7[4]")
-  graph_without_axis(wt.comp(wt.rwt,21),"W7[5]")
-  graph_without_axis(wt.comp(wt.rwt,22),"W7[6]")
-  graph_without_axis(wt.comp(wt.rwt,23),"W7[7]")
-  graph_without_axis(wt.comp(wt.rwt,24),"W7[8]")
-  graph_without_axis(wt.comp(wt.rwt,25),"W7[9]")
-  graph_without_axis(wt.comp(wt.rwt,26),"W7[10]")
-  graph_without_axis(wt.comp(wt.rwt,27),"W7[11]")
-  graph_without_axis(wt.comp(wt.rwt,28),"W7[12]")
-  graph_without_axis(wt.comp(wt.rwt,29),"W7[13]")
-  graph_without_axis(wt.comp(wt.rwt,30),"W7[14]")
-  graph_without_axis(wt.comp(wt.rwt,31),"W7[15]")
-  graph_without_axis(wt.comp(wt.rwt,32),"W7[16]")
-}
-```
-
-Notice that each time the frequency doubles, the number of wavelets also doubles. This creates a pyrimidal cascade of wavelet coefficients.
-
-The calculation of the wavelet decomposition is an interative process. First, you need the coefficients underlying the d4 wavelet. At the highest possible frequency, there are four coefficients.
+The calculation of the wavelet decomposition is an iterative process. First, you need the coefficients underlying the d4 wavelet. At the highest possible frequency, there are four coefficients.
 
@@
 The W1 coefficients represent the running dot product of the signal with the h vector, again with downsampling.
+
+$\begin{bmatrix} h_1 & h_2 & h_3 & h_4 &     &     & ...
+\\ &     & h_1 & h_2 & h_3 & h_4 & ...
+\\  .
+\\ .
+\\ .
+\\    &     &     &     &     &     & ...
+& h_1 & h_2 & h_3 & h_4 \\ h_3 & h_4 &     &     &     &     & ...
+&     &     & h_1 & h_2 \end{bmatrix}$
 

### 2022-01-28 text/21/glm-deviance.Rmd
Added matrix notation to 14/wavelets.Rmd

@@
 suppressMessages(suppressWarnings(library(dplyr)))
+suppressMessages(suppressWarnings(library(foreign)))
 suppressMessages(suppressWarnings(library(ggplot2)))
@@
 The first data set is available for anyone to download.
+The link I originally used no longer works.
+
+```{r, eval=FALSE}
+fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
+articles <- read.csv(fn)
+head(articles)
+```
+
+Here is a [different link][rod3] for the same file, used at [German Rodriguez's website][rod2].
+I've placed a [copy][sim3] on my website in case this link breaks.
+Also, this dataset is available with the [pscl library][zei1] of R under the name, bioChemists.
 
 ```{r}
-fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
-articles <- read.csv(fn)
+fn <- "http://www.stata-press.com/data/lf2/couart2.dta"
+articles <- read.dta(fn)
+head(articles)
+```
+
+[hua1]: https://francish.netlify.app/post/poisson-and-negative-binomial-regression-using-r/
+[rod2]: https://data.princeton.edu/wws509/r/overdispersion
+[rod3]: http://www.stata-press.com/data/lf2/couart2.dta
+[sim3]: http://www.pmean.com/00files/couart2.dta
+[zei1]: https://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf
+
+```{r}
+fn <- "http://www.stata-press.com/data/lf2/couart2.dta"
+articles <- read.dta(fn)
 head(articles)
@@
 
-German Rodriguez. Generalized Linear Model Theory. Lecture Notes on Generalized Linear Models. Available in [pdf format][rodr1]
+German Rodriguez. Generalized Linear Model Theory. Lecture Notes on Generalized Linear Models. Available in [pdf format][rod1]
 
@@
 [port1]: https://bookdown.org/egarpor/PM-UC3M/glm-deviance.html
-[rodr1]: https://data.princeton.edu/wws509/notes/a2.pdf
+[rod1]: https://data.princeton.edu/wws509/notes/a2.pdf
 [tay1]: https://statisticaloddsandends.wordpress.com/2019/03/27/what-is-deviance/

### 2022-01-28 text/21/glm-diagnostics.Rmd
Added matrix notation to 14/wavelets.Rmd

@@
 suppressMessages(suppressWarnings(library(ggplot2)))
+suppressMessages(suppressWarnings(library(foreign)))
 suppressMessages(suppressWarnings(library(magrittr)))
@@
 The first data set is available for anyone to download.
+The link I originally used no longer works.
 
-```{r}
+```{r, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 ```
+
+Here is a [different link][rod3] for the same file, used at [German Rodriguez's website][rod2].
+I've placed a [copy][sim3] on my website in case this link breaks.
+Also, this dataset is available with the [pscl library][zei1] of R under the name, bioChemists.
+
+```{r}
+fn <- "http://www.stata-press.com/data/lf2/couart2.dta"
+articles <- read.dta(fn)
+head(articles)
+```
+
+[hua1]: https://francish.netlify.app/post/poisson-and-negative-binomial-regression-using-r/
+[rod2]: https://data.princeton.edu/wws509/r/overdispersion
+[rod3]: http://www.stata-press.com/data/lf2/couart2.dta
+[sim3]: http://www.pmean.com/00files/couart2.dta
+[zei1]: https://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf
 

### 2022-01-28 text/21/negative-binomial-regression.Rmd
Added matrix notation to 14/wavelets.Rmd

@@
 
-Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model, the negative binomial model and several others.
+Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models.
 The first data set is available for anyone to download.
+The link I originally used no longer works.
 
-```{r}
+```{r, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"

### 2022-01-28 text/21/quasi-likelihood.Rmd
Added matrix notation to 14/wavelets.Rmd

@@
 
+### Link and variance functions
+
+A distribution is said to be in the exponential family
+
+$ln f(y, \theta)=d(\theta)e(y)+g(\theta)+h(y)$
+
+but sometimes this is rewritten in the form
+
+$ln f(y, \theta)=\frac{\theta y+b(\theta)}{a(\phi)}+c(y, \phi)$
+
+where $\phi$ is a dispersion or scale parameter. This is called the canonical form. There are names for these parameters and functions.
+
++ $\theta$ is the natural parameter,
++ $\phi$ is the dispersion parameter,
++ $a(\ ) is the variance function, and
++ $g(\ ) = b'(\ )^{-1} is the link function.
+
+If you also define
+
+$\mu=E[Y]$
+
+then the link function, g, and its inverse, b, define how the regression model works.
+
+For Poisson regression, the link function is
+
+$ln(f(y, \lambda) = ln(\frac{\lambda^y e^{-\lambda}}{y!})=\lambda ln(\lambda)-\lambda-ln(y!)$
+
+To write this in the canonical form, you need to define the natural parameter as
+
+$\theta=ln(\lambda)$
+
+and
+
+$a(\phi)=1$
+
+The link function is
+
+$g(\mu)=ln(\mu)$
+
+and the variance function is
+
+$V(\mu)=\mu$
+
+These functions are derived from the form of the Poisson distribution, and the estimation methods in the generalized linear model depend only on the link and variance functions.
+
+In 1974, Robert Weddeburn developed a theory that would work for any link and variance function, even if it could not be derived from a particular probability distribution.
+His approach used the same principles as maximum likelihood, but it could not be called a likelihood approach since there was no probability distribution.
+He called his approach quasi-likelihood and showed that it created estimates with reasonable properties.
+
+### The quasi Poisson model
+
+If you modify the variance function of Poisson regression, from
+
+$V(\mu)=\mu$
+
+to
+
+$V(\mu)=\phi \mu$
+
+and keep the link function the same, then you get a quasi Poisson model.
+There is no simple probability distribution that would produce this link and variance function, but it seems like a more reasonable approach, because there is more flexibility.
+
+For technical reasons, the parameter $\phi$ has to be greater than one, implying that the dispersion of the quasiPoisson model is greater than the dispersion of the Poissson model.
+The quasiPoisson model is often called an overdispersion model.
+
+There are two ways to estimate $\phi$. The first is very simple and direct.
+
+$\hat{\phi}= dispersion / df$
+
+The amount of overdispersion is equal to much much larger the dispersion is than the degrees of freedom.
+
+A second approach estimates $\phi$ iteratively along with the other terms in the regression model.
+
+One important limitation of the quasi likelihood approach is that there is no natural way to estimate AIC (Akaike's Information Criteria) an important statistic for comparing different models.
+
 <!---More--->
+
+### Start with Poisson regression
+
+It is easiest to understand the quasiPoisson model by comparing it to the Poisson model.
+
+Francis Huang has a [nice web page][hua1] showing two datasets and how to analyze them using the Poisson regression model as well as several more sophisticated models.
+The first data set is available for anyone to download.
+The link I originally used no longer works.
+
+```{r, eval=FALSE}
+fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
+articles <- read.csv(fn)
+head(articles)
+```
+
+Here is a [different link][rod3] for the same file, used at [German Rodriguez's website][rod2].
+I've placed a [copy][sim3] on my website in case this link breaks.
+Also, this dataset is available with the [pscl library][zei1] of R under the name, bioChemists.
+
+```{r}
+fn <- "http://www.stata-press.com/data/lf2/couart2.dta"
+articles <- read.dta(fn)
+head(articles)
+```
+
+[hua1]: https://francish.netlify.app/post/poisson-and-negative-binomial-regression-using-r/
+[rod2]: https://data.princeton.edu/wws509/r/overdispersion
+[rod3]: http://www.stata-press.com/data/lf2/couart2.dta
+[sim3]: http://www.pmean.com/00files/couart2.dta
+[zei1]: https://cran.r-project.org/web/packages/pscl/vignettes/countreg.pdf
+
+The variables are
++ gender (fem),
++ marital status (mar),
++ number of children (kid5),
++ prestige of graduate program (phd), 
++ the number of articles published by the individual's mentor (ment)
++ the number of articles published by the scientist (art: the outcome).
+
+To fit a Poisson regression model in R, use the glm function with the argument, family="poisson".
+
+```{r}
+pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
+summary(pois1)
+```
+
+The interpretation of the coefficients should be done after exponentiating their values.
+
+```{r}
+pois1 %>%
+  tidy %>%
+  mutate(lower_limit=estimate-1.96*std.error) %>%
+  mutate(upper_limit=estimate+1.96*std.error) %>%
+  select(term, estimate, lower_limit, upper_limit) %>%
+  mutate(estimate=exp(estimate)) %>%
+  mutate(lower_limit=exp(lower_limit)) %>%
+  mutate(upper_limit=exp(upper_limit)) -> pois1_coefficients
+pois1_coefficients
+```
+
+### Fitting a quasi-Poisson model
+
+```{r}
+quasi1 <- glm(art ~ fem + ment + phd + mar + kid5, family=quasipoisson, data = articles)
+summary(quasi1)
+```
+
+Notice that there is no value listed for AIC.
+
+```{r}
+quasi1 %>%
+  tidy %>%
+  mutate(lower_limit=estimate-1.96*std.error) %>%
+  mutate(upper_limit=estimate+1.96*std.error) %>%
+  select(term, estimate, lower_limit, upper_limit) %>%
+  mutate(estimate=exp(estimate)) %>%
+  mutate(lower_limit=exp(lower_limit)) %>%
+  mutate(upper_limit=exp(upper_limit)) -> quasi1_coefficients
+quasi1_coefficients
+```
+
+If you compare these confidence intervals to the earlier ones, they are wider. This reflects the overdispersion that is only accounted for properly by the quasi Poisson model.
 

### 2022-02-04 text/22/baseline-covariate.Rmd
Added asccr-frame.md

---
title: "Mapping latitude and longitude"
author: "Steve Simon"
source: new
date: "2022-02-01"
category: 
- Blog post
tags:
- Analysis of covariance
- Incomplete page
output: html_document
---

On the ASA Connect Digest, someone was asking about references discussing the practice of using change scores as the outcome variable with the baseline variable as a covariate.

Just to clarify here, a change score is commonly used when you are comparing an intervention to a control group and you measure the variable of interest before patients have been assigned to the intervention or control as well as afterwards.

For example, you have a new diet and exercise program designed to help people lose weight. You weigh a sample of 10 people and then randomly assign five to your new program and five to a standard diet and weight program. You wait six months and then weigh everyone again. The change score is the difference between the weight at baseline and the weight at six months. In simpler terms, the change score in this setting is the amount of weight lost.

The use of change scores is controversial for a variety of reasons. The preferred approach of many statisticians is to use the

### 2022-02-06 text/14/wavelets.Rmd
Minor update to wavelets.Rmd
Reconstruction section of wavelets.Rmd

@@
 
-The calculation of the wavelet decomposition is an iterative process. First, you need the coefficients underlying the d4 wavelet. At the highest possible frequency, there are four coefficients.
+You don't actually use the larger wavelets shown above.
+Instead you apply the most extreme wavelets in an iterative method that reduces the amount of data by half at each iteration.
+This makes the processing fast and easy.
+The most extreme D4 wavelets have only four coefficients.
 
@@
 
+Note: To get the math to work on my end, tThe low pass filter provided by the rwt package needs to be negated.
+
 ```{r low-pass}
-h <- d4.rwt$h.1
+h <- -d4.rwt$h.1
 h
@@
 The V1 coefficients represent the running dot product of the signal with the g vector.
-You skip every other data point in the running dot product, and you rotate the dot product around to the end when reach the right hand side to avoid edge effects.
+You skip every other data point in the running dot product (ignore the gray rows), and you rotate the dot product around to the end when reach the right hand side to avoid edge effects.
 The skipping past the second, fourth, sixth, etc.
@@
 
+![Figure 1. G matrix](http://www.pmean.com/new-images/14/wavelets-01.png)
+
 The W1 coefficients represent the running dot product of the signal with the h vector, again with downsampling.
 
-$\begin{bmatrix} h_1 & h_2 & h_3 & h_4 &     &     & ...
-\\ &     & h_1 & h_2 & h_3 & h_4 & ...
-\\  .
-\\ .
-\\ .
-\\    &     &     &     &     &     & ...
-& h_1 & h_2 & h_3 & h_4 \\ h_3 & h_4 &     &     &     &     & ...
-&     &     & h_1 & h_2 \end{bmatrix}$
+![Figure 2. H matrix](http://www.pmean.com/new-images/14/wavelets-02.png)
 
@@
 
-c(W1[   1], -h %*% ecg[1:4])
-c(W1[   2], -h %*% ecg[3:6])
-c(W1[   3], -h %*% ecg[5:8])
-c(W1[1023], -h %*% ecg[2045:2048])
-c(W1[1024], -h %*% ecg[c(2047:2048, 1:2)])
+c(W1[   1], h %*% ecg[1:4])
+c(W1[   2], h %*% ecg[3:6])
+c(W1[   3], h %*% ecg[5:8])
+c(W1[1023], h %*% ecg[2045:2048])
+c(W1[1024], h %*% ecg[c(2047:2048, 1:2)])
 ```
@@
 
-c(W2[  1], -h %*% V1[1:4])
-c(W2[  2], -h %*% V1[3:6])
-c(W2[  3], -h %*% V1[5:8])
-c(W2[511], -h %*% V1[1021:1024])
-c(W2[512], -h %*% V1[c(1023:1024, 1:2)])
+c(W2[  1], h %*% V1[1:4])
+c(W2[  2], h %*% V1[3:6])
+c(W2[  3], h %*% V1[5:8])
+c(W2[511], h %*% V1[1021:1024])
+c(W2[512], h %*% V1[c(1023:1024, 1:2)])
 ```
@@
 
-c(W3[  1], -h %*% V2[1:4])
-c(W3[  2], -h %*% V2[3:6])
-c(W3[  3], -h %*% V2[5:8])
-c(W3[255], -h %*% V2[509:512])
-c(W3[256], -h %*% V2[c(511:512, 1:2)])
+c(W3[  1], h %*% V2[1:4])
+c(W3[  2], h %*% V2[3:6])
+c(W3[  3], h %*% V2[5:8])
+c(W3[255], h %*% V2[509:512])
+c(W3[256], h %*% V2[c(511:512, 1:2)])
 ```
@@
 First you dilute the W and V values at the top of the pyramid.
-
-```{r reconstuct V8}
-V9_diluted <- c(V9[1], 0, V9[2], 0, V9[3], 0, V9[4], 0)
-W9_diluted <- c(W9[1], 0, W9[2], 0, W9[3], 0, W9[4], 0)
+Then you compute a running dot product of the reversed g vector with the diluted V9 values plus a running dot product of the reversed h vector with the diluted W9 values.
+
+```{r reconstuct-V8}
+V9_diluted <- rep(0, 8)
+V9_diluted[(1:4)*2-1] <- V9
+W9_diluted <- rep(0, 8)
+W9_diluted[(1:4)*2-1] <- W9
 
@@
 
-c(V8[1], rev(h) %*% V9_diluted[1:4] + rev(g) %*% W9_diluted[1:4])
-c(V8[2], rev(h) %*% V9_diluted[2:5] + rev(g) %*% W9_diluted[2:5])
+c(V8[1], rev(g) %*% V9_diluted[c(6, 7, 8, 1)] + rev(h) %*% W9_diluted[c(6, 7, 8, 1)])
+c(V8[2], rev(g) %*% V9_diluted[c(7, 8, 1, 2)] + rev(h) %*% W9_diluted[c(7, 8, 1, 2)])
+c(V8[3], rev(g) %*% V9_diluted[c(8, 1, 2, 3)] + rev(h) %*% W9_diluted[c(8, 1, 2, 3)])
+c(V8[4], rev(g) %*% V9_diluted[c(1, 2, 3, 4)] + rev(h) %*% W9_diluted[c(1, 2, 3, 4)])
+c(V8[5], rev(g) %*% V9_diluted[c(2, 3, 4, 5)] + rev(h) %*% W9_diluted[c(2, 3, 4, 5)])
+c(V8[6], rev(g) %*% V9_diluted[c(3, 4, 5, 6)] + rev(h) %*% W9_diluted[c(3, 4, 5, 6)])
+c(V8[7], rev(g) %*% V9_diluted[c(4, 5, 6, 7)] + rev(h) %*% W9_diluted[c(4, 5, 6, 7)])
+c(V8[8], rev(g) %*% V9_diluted[c(5, 6, 7, 8)] + rev(h) %*% W9_diluted[c(5, 6, 7, 8)])
+```
+
+To move from V8 to V7, dilute the V8 and W8 vectors and apply a running dot product again.
+
+```{r reconstuct-V7}
+V8_diluted <- rep(0, 16)
+V8_diluted[(1:8)*2-1] <- V8
+W8_diluted <- rep(0, 16)
+W8_diluted[(1:8)*2-1] <- W8
+
+wt.7 <- mdwt(ecg, g, L=7)
+V7 <- wt.7$y[1:16]
+
+c(V7[ 1], rev(g) %*% V8_diluted[c(14, 15, 16,  1)] + rev(h) %*% W8_diluted[c(14, 15, 16,  1)])
+c(V7[ 2], rev(g) %*% V8_diluted[c(15, 16,  1,  2)] + rev(h) %*% W8_diluted[c(15, 16,  1,  2)])
+c(V7[ 3], rev(g) %*% V8_diluted[c(16,  1,  2,  3)] + rev(h) %*% W8_diluted[c(16,  1,  2,  3)])
+c(V7[ 4], rev(g) %*% V8_diluted[c( 1,  2,  3,  4)] + rev(h) %*% W8_diluted[c( 1,  2,  3,  4)])
+c(V7[16], rev(g) %*% V8_diluted[c(13, 14, 15, 16)] + rev(h) %*% W8_diluted[c(12, 13, 14, 15)])
+```
+
+Here is the reconstruction of V6
+
+```{r reconstuct-V6}
+V7_diluted <- rep(0, 32)
+V7_diluted[(1:16)*2-1] <- V7
+W7_diluted <- rep(0, 32)
+W7_diluted[(1:16)*2-1] <- W7
+
+wt.6 <- mdwt(ecg, g, L=6)
+V6 <- wt.6$y[1:32]
+
+c(V6[ 1], rev(g) %*% V7_diluted[c(30, 31, 32,  1)] + rev(h) %*% W7_diluted[c(30, 31, 32,  1)])
+c(V6[ 2], rev(g) %*% V7_diluted[c(31, 32,  1,  2)] + rev(h) %*% W7_diluted[c(31, 32,  1,  2)])
+c(V6[ 3], rev(g) %*% V7_diluted[c(32,  1,  2,  3)] + rev(h) %*% W7_diluted[c(32,  1,  2,  3)])
+c(V6[ 4], rev(g) %*% V7_diluted[c( 1,  2,  3,  4)] + rev(h) %*% W7_diluted[c( 1,  2,  3,  4)])
+c(V6[32], rev(g) %*% V7_diluted[c(29, 30, 31, 32)] + rev(h) %*% W7_diluted[c(29, 30, 31, 32)])
+```
+
+The process for reconstructing V5, V4, etc. is similar. At the final step, dilute V1 and W1. The running dot product will recreate the original signal.
+
+What exactly do V9, V8, V6, etc. measure? They are coefficients associated with the mother wavelet and reconstruct a signal with high frequency events removed.
+
+```{r graph-v}
+
 V9_coefficients <- rep(0, 2048)
-V9_coefficients[1:8] <- wt.9$y[1:8]
+V9_coefficients[1:4] <- V9
 V9_signal <- as.vector(midwt(V9_coefficients, d4.rwt$h.0, L=9)$x)
+
+V8_coefficients <- rep(0, 2048)
+V8_coefficients[1:8] <- V8
+V8_signal <- as.vector(midwt(V8_coefficients, d4.rwt$h.0, L=8)$x)
+
+V7_coefficients <- rep(0, 2048)
+V7_coefficients[1:16] <- V7
+V7_signal <- as.vector(midwt(V7_coefficients, d4.rwt$h.0, L=7)$x)
+
+V6_coefficients <- rep(0, 2048)
+V6_coefficients[1:32] <- V6
+V6_signal <- as.vector(midwt(V6_coefficients, d4.rwt$h.0, L=6)$x)
+
+wt.5 <- mdwt(ecg, g, L=5)
+V5 <- wt.5$y[1:64] 
+V5_coefficients <- rep(0, 2048)
+V5_coefficients[1:64] <- V5
+V5_signal <- as.vector(midwt(V5_coefficients, d4.rwt$h.0, L=5)$x)
+
+wt.4 <- mdwt(ecg, g, L=4)
+V4 <- wt.4$y[1:128] 
+V4_coefficients <- rep(0, 2048)
+V4_coefficients[1:128] <- V4
+V4_signal <- as.vector(midwt(V4_coefficients, d4.rwt$h.0, L=4)$x)
+
+wt.3 <- mdwt(ecg, g, L=3)
+V3 <- wt.3$y[1:256] 
+V3_coefficients <- rep(0, 2048)
+V3_coefficients[1:256] <- V3
+V3_signal <- as.vector(midwt(V3_coefficients, d4.rwt$h.0, L=3)$x)
+
+wt.2 <- mdwt(ecg, g, L=2)
+V2 <- wt.2$y[1:512] 
+V2_coefficients <- rep(0, 2048)
+V2_coefficients[1:512] <- V2
+V2_signal <- as.vector(midwt(V2_coefficients, d4.rwt$h.0, L=2)$x)
+
+wt.1 <- mdwt(ecg, g, L=1)
+V1 <- wt.1$y[1:1024] 
+V1_coefficients <- rep(0, 2048)
+V1_coefficients[1:1024] <- V1
+V1_signal <- as.vector(midwt(V1_coefficients, d4.rwt$h.0, L=1)$x)
+
 {
   graph_without_axis(V9_signal, "V9", limits=range(ecg))
+  graph_without_axis(V8_signal, "V8", limits=range(ecg))
+  graph_without_axis(V7_signal, "V7", limits=range(ecg))
+  graph_without_axis(V6_signal, "V6", limits=range(ecg))
+  graph_without_axis(V5_signal, "V5", limits=range(ecg))
+  graph_without_axis(V4_signal, "V4", limits=range(ecg))
+  graph_without_axis(V3_signal, "V3", limits=range(ecg))
+  graph_without_axis(V2_signal, "V2", limits=range(ecg))
+  graph_without_axis(V1_signal, "V1", limits=range(ecg))
   graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
@@
 
-W8d <- dilute(W8)
-V8d <- dilute(V8)
- W8d, g, 14)+V8d,h,14),V7[1],"V7[1]")
-V7[1]
-By hand -3.315508
-Using R -3.315508
--W8d, g, 15)-V8d,h,15),V7[2],"V7[2]")
-V7[2]
-By hand -2.398146
-Using R -2.398146
- W8d, g, 16)+V8d,h,16),V7[3],"V7[3]")
-V7[3]
-By hand -0.2769286
-Using R -0.2769286
--W8d, g,  1)-V8d,h, 1),V7[4],"V7[4]")
-V7[4]
-By hand 1.782748
-Using R 1.782748
- W8d, g,  2)+V8d,h, 2),V7[5],"V7[5]")
-V7[5]
-By hand 5.25007
-Using R 5.25007
--W8d, g,  3)-V8d,h, 3),V7[6],"V7[6]")
-V7[6]
-By hand 4.7496
-Using R 4.7496
-
-You dilute the V8 and W8 coefficients by inserting zeros between every element, compute a running dot product of both the g and the h vectors, and add the results together.
-For some reason, you have to multiply every other value by -1.
-I don't know why.
-It's one of those scaling things.
-Also notice that you start at almost the very end of V8.
-This is caused by edge effects.
-
-Once you have V7, repeat the process on the diluted V7 and W7 vectors to get V6.
-
-W7d <- dilute(W7)
-V7d <- dilute(V7)
- W7d, g, 30)+V7d,h,30),V6[1],"V6[1]")
-V6[1]
-By hand -2.096562
-Using R -2.096562
--W7d, g, 31)-V7d,h,31),V6[2],"V6[2]")
-V6[2]
-By hand -2.294492
-Using R -2.294492
- W7d, g, 32)+V7d,h,32),V6[3],"V6[3]")
-V6[3]
-By hand -2.441797
-Using R -2.441797
--W7d, g,  1)-V7d,h, 1),V6[4],"V6[4]")
-V6[4]
-By hand -1.265345
-Using R -1.265345
- W7d, g,  2)+V7d,h, 2),V6[5],"V6[5]")
-V6[5]
-By hand -0.6650157
-Using R -0.6650157
--W7d, g,  3)-V7d,h, 3),V6[6],"V6[6]")
-V6[6]
-By hand 0.08738041
-Using R 0.08738041
-
-And so forth.
-
-W1d <- dilute(W1)
-V1d <- dilute(V1)
- W1d, g, 2046)+V1d,h,2046),ecg[1],"ecg[1]")
-ecg[1]
-By hand -0.104773
-Using R -0.104773
--W1d, g, 2047)-V1d,h,2047),ecg[2],"ecg[2]")
-ecg[2]
-By hand -0.093136
-Using R -0.093136
- W1d, g, 2048)+V1d,h,2048),ecg[3],"ecg[3]")
-ecg[3]
-By hand -0.0815
-Using R -0.0815
--W1d, g,  1)-V1d,h, 1),ecg[4],"ecg[4]")
-ecg[4]
-By hand -0.116409
-Using R -0.116409
- W1d, g,  2)+V1d,h, 2),ecg[5],"ecg[5]")
-ecg[5]
-By hand -0.0815
-Using R -0.0815
--W1d, g,  3)-V1d,h, 3),ecg[6],"ecg[6]")
-ecg[6]
-By hand -0.116409
-Using R -0.116409
-
-At the very end, the calculations on the diluted V1 and W1 values give you the original signal.
-
-You can see what the wavelet coefficients at each level contribute by zeroing out some of the components and then inverting it.
-
-V8 and W8
-W7
-W6
-W5
-W4
-W3
-W2
-W1
-ecg
-Axis
-
-Notice that the sharp spikes are captured in W1, W2, and W3.
-The slower overall undulation is captured by V8 and W8.
-Some of the intermediate squiggles between the sharp spikes are captured in the remaining wavelets.
+What exactly do W9, W8, W7, etc. measure? They are coefficients associated with the father wavelet and reconstruct a signal with at each frequency level.
+
+```{r graph-w}
+
+W9_coefficients <- rep(0, 2048)
+W9_coefficients[5:8] <- W9
+W9_signal <- as.vector(midwt(W9_coefficients, d4.rwt$h.0, L=9)$x)
+
+W8_coefficients <- rep(0, 2048)
+W8_coefficients[9:16] <- W8
+W8_signal <- as.vector(midwt(W8_coefficients, d4.rwt$h.0, L=9)$x)
+
+W7_coefficients <- rep(0, 2048)
+W7_coefficients[17:32] <- W7
+W7_signal <- as.vector(midwt(W7_coefficients, d4.rwt$h.0, L=9)$x)
+
+W6_coefficients <- rep(0, 2048)
+W6_coefficients[33:64] <- W6
+W6_signal <- as.vector(midwt(W6_coefficients, d4.rwt$h.0, L=9)$x)
+
+W5_coefficients <- rep(0, 2048)
+W5_coefficients[65:128] <- W5
+W5_signal <- as.vector(midwt(W5_coefficients, d4.rwt$h.0, L=9)$x)
+
+W4_coefficients <- rep(0, 2048)
+W4_coefficients[129:256] <- W4
+W4_signal <- as.vector(midwt(W4_coefficients, d4.rwt$h.0, L=9)$x)
+
+W3_coefficients <- rep(0, 2048)
+W3_coefficients[257:512] <- W3
+W3_signal <- as.vector(midwt(W3_coefficients, d4.rwt$h.0, L=9)$x)
+
+W2_coefficients <- rep(0, 2048)
+W2_coefficients[513:1024] <- W2
+W2_signal <- as.vector(midwt(W2_coefficients, d4.rwt$h.0, L=9)$x)
+
+W1_coefficients <- rep(0, 2048)
+W1_coefficients[1025:2048] <- W1
+W1_signal <- as.vector(midwt(W1_coefficients, d4.rwt$h.0, L=9)$x)
+
+{
+  graph_without_axis(W9_signal, "W9", limits=range(ecg))
+  graph_without_axis(W8_signal, "W8", limits=range(ecg))
+  graph_without_axis(W7_signal, "W7", limits=range(ecg))
+  graph_without_axis(W6_signal, "W6", limits=range(ecg))
+  graph_without_axis(W5_signal, "W5", limits=range(ecg))
+  graph_without_axis(W4_signal, "W4", limits=range(ecg))
+  graph_without_axis(W3_signal, "W3", limits=range(ecg))
+  graph_without_axis(W2_signal, "W2", limits=range(ecg))
+  graph_without_axis(W1_signal, "W1", limits=range(ecg))
+  graph_without_axis(ecg, "Original\nsignal", limits=range(ecg))
+}
+```
 

### 2022-02-06 text/14/wavelets.Rmd
Added 2-D images to wavelets.Rmd

@@
   axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
-  text(1.1, x[2048], lab, adj=0.5)
+  text(1.1, x[2048], lab, adj=0.5, cex=0.67)
 }
@@
        xlim=c(0, 1.2), 
-       ylim=c(limits[1]*1.1, limits[2]), # adjust lower limit
+       ylim=c(limits[1], limits[2]),
        type="l",
        axes=FALSE)
-  text(1.1, x[2048], lab, adj=0.5)
+  text(1.1, x[2048], lab, adj=0.5, cex=0.67)
 }
@@
   segments(x, 0, x, y)
-  text(1.1, 0, lab)
-}
-```
-
-The head and tail function show the first few and the last few values for the signal. The graph looks like
+  text(1.1, 0, lab, cex=0.67)
+}
+```
+
+### The ecg signal
+
+Here's a graph of data representing an electrocardiogram (ecg) signal that will help illustrate the use of Fourier transforms and wavelets.
 
@@
 
+### Components of the Fourier transform
+
 The object ft.ecg contains the Fourier decomposition, and ift.ecg is the inverse of the decomposition.
@@
 The first number has only a real component and no imaginary component.
-The real component represent the mean of the entire signal.
+This component represent the mean of the entire signal.
 
@@
 
+### Zeroing out components
+
 There's a faster and more convenient way.
@@
 {
-  lines((1:2048)/2048, sum1, col="red", lty="dashed")
+  lines((1:2048)/2048, sum1, col="red", lty="dashed", lwd=3)
 }
@@
 comp3 <- Re(fft(ft.comp3,inverse=TRUE))
-graph_without_axis(comp3, "Fourier\ncomponent")
+graph_without_axis(comp3, "Third\ncomponent")
 ```
@@
 comp4 <- Re(fft(ft.comp4,inverse=TRUE))
-graph_without_axis(comp4, "Fourier\ncomponent")
+graph_without_axis(comp4, "Fourth\ncomponent")
 ```
@@
 The fourth Fourier component has three peaks.
+
+### Orthogonality
+
 The interesting thing is that the second and third and fourth Fourier components are orthogonal to one another.
@@
 comp32 <- Re(fft(ft.comp32,inverse=TRUE))
-graph_without_axis(comp32, "Fourier\ncomponent")
+graph_without_axis(comp32, "32nd\ncomponent")
 ```
@@
 m <- max(comp32)
-graph_without_axis(comp32, "Fourier\ncomponent", limits=c(-m, 1.2*m))
-text((seq(1, 31, by=3)-0.5)/31, 1.2*m, seq(1, 31, by=3), col="red") 
+graph_without_axis(comp32, "32nd\ncomponent", limits=c(-m, 1.2*m))
+text((seq(1, 31, by=3)-0.5)/31, 1.2*m, seq(1, 31, by=3), col="red", cex=0.67) 
 ```
@@
 d4.rwt
-wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
+wt.rwt <- mdwt(ecg,daubcqf(4)$h.0,L=9)
 names(wt.rwt)
@@
 wt.rwt$L
-iwt.rwt <- midwt(wt.rwt$y,d4.rwt$h.0,L=9)
+iwt.rwt <- midwt(wt.rwt$y,daubcqf(4)$h.0,L=9)
 names(iwt.rwt)
@@
 Don't ask me what a coiflet is.
+
+### wavelets package
 
@@
   single_wavelet <- as.numeric((1:2048)==y)
-  as.vector(midwt(single_wavelet, d4.rwt$h.0, L=8)$x)
+  as.vector(midwt(single_wavelet, daubcqf(4)$h.0, L=8)$x)
 }
@@
   yh[i, L] <- 1
-  as.vector(mirdwt(yl, yh, d4.rwt$h.0, L)$x)
+  as.vector(mirdwt(yl, yh, daubcqf(4)$h.0, L)$x)
 }
@@
   yl[i] <- 1
-  as.vector(mirdwt(yl, yh, d4.rwt$h.0, L)$x)
+  as.vector(mirdwt(yl, yh, daubcqf(4)$h.0, L)$x)
 }
@@
 {
-  graph_without_axis(calculate_father(2^9-2^7+1, 7),"d4 father")
-  graph_without_axis(calculate_father(2^9+1, 7),"d4 father")
-  graph_without_axis(calculate_father(2^9+2^7+1, 7),"d4 father")
-  graph_without_axis(calculate_mother(2^9-2^7+1, 7),"d4 mother")
-  graph_without_axis(calculate_mother(2^9+1, 7),"d4 mother")
-  graph_without_axis(calculate_mother(2^9+2^7+1, 7),"d4 mother")
+  graph_without_axis(calculate_father(2^9-2^7+1, 7),"d4 father\nshifted left")
+  graph_without_axis(calculate_father(2^9+2^7+1, 7),"d4 father\nshifted right")
+  graph_without_axis(calculate_mother(2^9-2^7+1, 7),"d4 mother\nshifted left")
+  graph_without_axis(calculate_mother(2^9+2^7+1, 7),"d4 mother\nshifted right")
 }
@@
 {
-  graph_without_axis(calculate_father(2^10+2^9+1+2^8+1, 7),"d4 father")
-  graph_without_axis(calculate_mother(2^10+2^9+1+2^8+1, 7),"d4 mother")
+  graph_without_axis(calculate_father(2^10+2^9+1+2^8+1, 7),"d4 father, wrapped")
+  graph_without_axis(calculate_mother(2^10+2^9+1+2^8+1, 7),"d4 mother, wrapped")
 }
@@
 {  
-  graph_without_axis(calculate_father(2^9+1, 8),"d4 father")
-  graph_without_axis(calculate_father(2^9+1, 7),"d4 father")
-  graph_without_axis(calculate_father(2^9+1, 6),"d4 father")
-  graph_without_axis(calculate_mother(2^9+1, 8),"d4 mother")
-  graph_without_axis(calculate_mother(2^9+1, 7),"d4 mother")
-  graph_without_axis(calculate_mother(2^9+1, 6),"d4 mother")
+  graph_without_axis(calculate_father(2^9+1, 8),"d4 father\nstretched")
+  graph_without_axis(calculate_father(2^9+1, 6),"d4 father\nsqueezed")
+  graph_without_axis(calculate_mother(2^9+1, 8),"d4 mother\nstretched")
+  graph_without_axis(calculate_mother(2^9+1, 6),"d4 mother\nsqueezed")
 }
@@
 {
-  graph_without_axis(calculate_father(2^9+1, 1),"d4 father")
-  graph_without_axis(calculate_mother(2^9+1, 1),"d4 mother")
+  graph_without_axis(calculate_father(2^9+1, 1),"d4 father\nmost extreme")
+  graph_without_axis(calculate_mother(2^9+1, 1),"d4 mother\nmost extreme")
 }
@@
 ```{r high-pass}
-g <- d4.rwt$h.0
+g <- daubcqf(4)$h.0
 g
@@
 ```{r low-pass}
-h <- -d4.rwt$h.1
+h <- -daubcqf(4)$h.1
 h
@@
 ```{r first-level-wavelet}
-wt.1 <- mdwt(ecg, g, L=1)
+wt.1 <- mdwt(ecg, daubcqf(4)$h.0, L=1)
 V1 <- wt.1$y[   1:1024]
@@
 ```{r second-level-wavelet}
-wt.2 <- mdwt(ecg, g, L=2)
+wt.2 <- mdwt(ecg, daubcqf(4)$h.0, L=2)
 V2 <- wt.2$y[1:512]
@@
 ```{r transform-V1}
-quality_check <- mdwt(V1, g, L=1)
+quality_check <- mdwt(V1, daubcqf(4)$h.0, L=1)
 quality_check_V2 <- quality_check$y[  1: 512]
@@
 ```{r third-level-wavelet}
-wt.3 <- mdwt(ecg, g, L=3)
+wt.3 <- mdwt(ecg, daubcqf(4)$h.0, L=3)
 V3 <- wt.3$y[1:256]
@@
 ```{r ninth-level-wavelet}
-wt.9 <- mdwt(ecg, g, L=9)
+wt.9 <- mdwt(ecg, daubcqf(4)$h.0, L=9)
 V9 <- wt.9$y[   1:   4]
@@
 Then you compute a running dot product of the reversed g vector with the diluted V9 values plus a running dot product of the reversed h vector with the diluted W9 values.
+Please note that using the reversed order for the g and h vectors is unique to the Daubechies family of wavelets.
+Other wavelets will use an entirely different set of vectors in the rolling dot product.
+
+![Figure 3. Matrix representation of reconstruction](http://www.pmean.com/new-images/14/wavelets-03.png)
 
@@
 
-wt.8 <- mdwt(ecg, g, L=8)
+wt.8 <- mdwt(ecg, daubcqf(4)$h.0, L=8)
 V8 <- wt.8$y[1:8]
@@
 
-wt.7 <- mdwt(ecg, g, L=7)
+wt.7 <- mdwt(ecg, daubcqf(4)$h.0, L=7)
 V7 <- wt.7$y[1:16]
@@
 
-wt.6 <- mdwt(ecg, g, L=6)
+wt.6 <- mdwt(ecg, daubcqf(4)$h.0, L=6)
 V6 <- wt.6$y[1:32]
@@
 V9_coefficients[1:4] <- V9
-V9_signal <- as.vector(midwt(V9_coefficients, d4.rwt$h.0, L=9)$x)
+V9_signal <- as.vector(midwt(V9_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 V8_coefficients[1:8] <- V8
-V8_signal <- as.vector(midwt(V8_coefficients, d4.rwt$h.0, L=8)$x)
+V8_signal <- as.vector(midwt(V8_coefficients, daubcqf(4)$h.0, L=8)$x)
 
@@
 V7_coefficients[1:16] <- V7
-V7_signal <- as.vector(midwt(V7_coefficients, d4.rwt$h.0, L=7)$x)
+V7_signal <- as.vector(midwt(V7_coefficients, daubcqf(4)$h.0, L=7)$x)
 
@@
 V6_coefficients[1:32] <- V6
-V6_signal <- as.vector(midwt(V6_coefficients, d4.rwt$h.0, L=6)$x)
-
-wt.5 <- mdwt(ecg, g, L=5)
+V6_signal <- as.vector(midwt(V6_coefficients, daubcqf(4)$h.0, L=6)$x)
+
+wt.5 <- mdwt(ecg, daubcqf(4)$h.0, L=5)
 V5 <- wt.5$y[1:64] 
@@
 V5_coefficients[1:64] <- V5
-V5_signal <- as.vector(midwt(V5_coefficients, d4.rwt$h.0, L=5)$x)
-
-wt.4 <- mdwt(ecg, g, L=4)
+V5_signal <- as.vector(midwt(V5_coefficients, daubcqf(4)$h.0, L=5)$x)
+
+wt.4 <- mdwt(ecg, daubcqf(4)$h.0, L=4)
 V4 <- wt.4$y[1:128] 
@@
 V4_coefficients[1:128] <- V4
-V4_signal <- as.vector(midwt(V4_coefficients, d4.rwt$h.0, L=4)$x)
-
-wt.3 <- mdwt(ecg, g, L=3)
+V4_signal <- as.vector(midwt(V4_coefficients, daubcqf(4)$h.0, L=4)$x)
+
+wt.3 <- mdwt(ecg, daubcqf(4)$h.0, L=3)
 V3 <- wt.3$y[1:256] 
@@
 V3_coefficients[1:256] <- V3
-V3_signal <- as.vector(midwt(V3_coefficients, d4.rwt$h.0, L=3)$x)
-
-wt.2 <- mdwt(ecg, g, L=2)
+V3_signal <- as.vector(midwt(V3_coefficients, daubcqf(4)$h.0, L=3)$x)
+
+wt.2 <- mdwt(ecg, daubcqf(4)$h.0, L=2)
 V2 <- wt.2$y[1:512] 
@@
 V2_coefficients[1:512] <- V2
-V2_signal <- as.vector(midwt(V2_coefficients, d4.rwt$h.0, L=2)$x)
-
-wt.1 <- mdwt(ecg, g, L=1)
+V2_signal <- as.vector(midwt(V2_coefficients, daubcqf(4)$h.0, L=2)$x)
+
+wt.1 <- mdwt(ecg, daubcqf(4)$h.0, L=1)
 V1 <- wt.1$y[1:1024] 
@@
 V1_coefficients[1:1024] <- V1
-V1_signal <- as.vector(midwt(V1_coefficients, d4.rwt$h.0, L=1)$x)
+V1_signal <- as.vector(midwt(V1_coefficients, daubcqf(4)$h.0, L=1)$x)
 
@@
 W9_coefficients[5:8] <- W9
-W9_signal <- as.vector(midwt(W9_coefficients, d4.rwt$h.0, L=9)$x)
+W9_signal <- as.vector(midwt(W9_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W8_coefficients[9:16] <- W8
-W8_signal <- as.vector(midwt(W8_coefficients, d4.rwt$h.0, L=9)$x)
+W8_signal <- as.vector(midwt(W8_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W7_coefficients[17:32] <- W7
-W7_signal <- as.vector(midwt(W7_coefficients, d4.rwt$h.0, L=9)$x)
+W7_signal <- as.vector(midwt(W7_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W6_coefficients[33:64] <- W6
-W6_signal <- as.vector(midwt(W6_coefficients, d4.rwt$h.0, L=9)$x)
+W6_signal <- as.vector(midwt(W6_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W5_coefficients[65:128] <- W5
-W5_signal <- as.vector(midwt(W5_coefficients, d4.rwt$h.0, L=9)$x)
+W5_signal <- as.vector(midwt(W5_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W4_coefficients[129:256] <- W4
-W4_signal <- as.vector(midwt(W4_coefficients, d4.rwt$h.0, L=9)$x)
+W4_signal <- as.vector(midwt(W4_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W3_coefficients[257:512] <- W3
-W3_signal <- as.vector(midwt(W3_coefficients, d4.rwt$h.0, L=9)$x)
+W3_signal <- as.vector(midwt(W3_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W2_coefficients[513:1024] <- W2
-W2_signal <- as.vector(midwt(W2_coefficients, d4.rwt$h.0, L=9)$x)
+W2_signal <- as.vector(midwt(W2_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 W1_coefficients[1025:2048] <- W1
-W1_signal <- as.vector(midwt(W1_coefficients, d4.rwt$h.0, L=9)$x)
+W1_signal <- as.vector(midwt(W1_coefficients, daubcqf(4)$h.0, L=9)$x)
 
@@
 
+### Applications of wavelets
+
+### Two dimensional wavelets
+
 There are simple generalizations of wavelets for two dimensional images. Here's a simple image, which represents a 256 by 256 bit grayscale image of a dog.
 
-img <- read.table("klaus256.txt")
+```{r read-dog-image}
+knitr::opts_chunk$set(fig.width=5, fig.height=5)
+fn <- "https://www2.isye.gatech.edu/isyebayes/data/klaus256.asc"
+img <- read.table(fn)
 img <- 2*img-1
-
+```
+
+```{r display-dog-image}
 img.plot <- function(imat,n=256) {
@@
   plot(c(1,n+1),c(1,n+1),type="n",axes=FALSE)
+  # imat <- imat/max(abs(imat))
   for (i in 1:n) {
@@
 }
+
 img.plot(img)
-
-Dog image
+```
 
@@
 
+```{r d8-wavelet-g}
 g <- daubcqf(8)$h.0
 g
-[1] 0.23037781 0.71484657 0.63088077 -0.02798377 -0.18703481 0.03084138
-[7] 0.03288301 -0.01059740
+```
+
+```{r d8-wavelet-h}
 h <- -daubcqf(8)$h.1
 h
-[1] -0.01059740 -0.03288301 0.03084138 0.18703481 -0.02798377 -0.63088077
-[7] 0.71484657 -0.23037781
+```
 
@@
 
-ecg
-Axis
+```{r d8-in-1d, fig.height=1}
+d8_father_coefficients <- rep(0, 2048)
+d8_father_coefficients[3] <- 1
+d8_father <- as.vector(midwt(d8_father_coefficients, daubcqf(8)$h.0, L=8)$x)
+d8_mother_coefficients <- rep(0, 2048)
+d8_mother_coefficients[10] <- 1
+d8_mother <- as.vector(midwt(d8_mother_coefficients, daubcqf(8)$h.0, L=8)$x)
+{
+  graph_without_axis(d8_father, "D8 father")
+  graph_without_axis(d8_mother, "D8 mother")
+}
+```
 
@@
 
-wt.img.1 <- mdwt(img, g, L=1)
-sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
-V1
-By hand -1.580973
-Using R -1.580973
-sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
-W1
-By hand -0.002825861
-Using R -0.002825861
-sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
-W1
-By hand -0.01225749
-Using R -0.01225749
-sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
-W1
-By hand -0.005042734
-Using R -0.005042734
+```{r decompose-and-restructure}
+wt.img.4 <- mdwt(img, daubcqf(8)$h.0, L=4)
+reconstructed.img <- midwt(wt.img.4$y, daubcqf(8)$h.0, L=4)
+img.plot(reconstructed.img$x)
+```
+
+```{r check-2d-calculations}
+wt.img.1 <- mdwt(img, daubcqf(8)$h.0, L=1)
+c(sum(img[1:8,1:8]*outer(g,g)), wt.img.1$y[ 1,   1])
+c(sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[  1, 129])
+c(sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129,   1])
+c(sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129, 129])
+```
 
@@
 
-Wavelet coefficients arrangement
+### Wavelet coefficients arrangement
 
@@
 
+```{r level-1-image-g-g}
+gg1 <- matrix(0, nrow=256, ncol=256)
+gg1[1:128, 1:128] <- wt.img.1$y[1:128, 1:128]
+gg1_signal <- midwt(gg1, daubcqf(8)$h.0, L=1)  
+img.plot(gg1_signal$x)
+```
+
+```{r level-1-image-g-h}
+gh1 <- matrix(0, nrow=256, ncol=256)
+gh1[1:128, 129:256] <- wt.img.1$y[1:128, 129:256]
+gh1_signal <- midwt(gh1, daubcqf(8)$h.0, L=1)  
+img.plot(gh1_signal$x)
+```
+
+```{r level-1-image-h-g}
+hg1 <- matrix(0, nrow=256, ncol=256)
+hg1[1:128, 129:256] <- wt.img.1$y[129:256, 1:128]
+hg1_signal <- midwt(hg1, daubcqf(8)$h.0, L=1)  
+img.plot(hg1_signal$x)
+```
+
+```{r level-1-image-h-h}
+hh1 <- matrix(0, nrow=256, ncol=256)
+hh1[129:256, 129:256] <- wt.img.1$y[129:256, 129:256]
+hh1_signal <- midwt(hh1, daubcqf(8)$h.1, L=1)  
+img.plot(hh1_signal$x)
+```
+
+```{r level-4-image-g-g}
+gg4 <- matrix(0, nrow=256, ncol=256)
+gg4[1:16, 1:16] <- wt.img.4$y[1:16, 1:16]
+gg4_signal <- midwt(gg4, daubcqf(8)$h.0, L=4)  
+img.plot(gg4_signal$x)
+```
+
+```{r level-4-image-g-h}
+gh4 <- matrix(0, nrow=256, ncol=256)
+gh4[1:16, 17:32] <- wt.img.4$y[1:16, 17:32]
+gh4[1:32, 33:64] <- wt.img.4$y[1:32, 33:64]
+gh4[1:64, 65:128] <- wt.img.4$y[1:64, 65:128]
+gh4[1:128,129:256] <- wt.img.4$y[1:128, 129:256]
+gh4_signal <- midwt(gh4, daubcqf(8)$h.0, L=4)  
+img.plot(gh4_signal$x)
+```
+
+```{r level-4-image-h-g}
+hg4 <- matrix(0, nrow=256, ncol=256)
+hg4[17:32, 1:16] <- wt.img.4$y[17:32, 1:16]
+hg4[33:64, 1:32] <- wt.img.4$y[33:64, 1:32]
+hg4[65:128, 1:64] <- wt.img.4$y[65:128, 1:64]
+hg4[129:256, 1:128] <- wt.img.4$y[129:256, 1:128]
+hg4_signal <- midwt(hg4, daubcqf(8)$h.0, L=4)  
+img.plot(hg4_signal$x)
+```
+
+```{r level-4-image-h-h}
+hh4 <- matrix(0, nrow=256, ncol=256)
+hh4[17:32, 17:32] <- wt.img.4$y[17:32, 17:32]
+hh4[33:64, 33:64] <- wt.img.4$y[33:64, 33:64]
+hh4[65:128, 65:128] <- wt.img.4$y[65:128, 65:128]
+hh4[129:256, 129:256] <- wt.img.4$y[129:256, 129:256]
+hh4_signal <- midwt(hh4, daubcqf(8)$h.0, L=4)  
+img.plot(hh4_signal$x)
+```
+
+```{r level-4-detail}
+detail4 <- gh4 + hg4 + hh4
+detail4_signal <- midwt(detail4, daubcqf(8)$h.0, L=4)  
+img.plot(detail4_signal$x)
+```
+
+```{r compress}
+pctl_99 <- quantile(abs(wt.img.4$y), probs=0.99)
+top_01 <- matrix(0, nrow=256, ncol=256)
+top_01[abs(wt.img.4$y)>pctl_99] <- wt.img.4$y[abs(wt.img.4$y)>pctl_99]
+top_01_image <- midwt(top_01, daubcqf(8)$h.0, L=4)
+img.plot(top_01_image$x)
+
+pctl_95 <- quantile(abs(wt.img.4$y), probs=0.95)
+top_05 <- matrix(0, nrow=256, ncol=256)
+top_05[abs(wt.img.4$y)>pctl_95] <- wt.img.4$y[abs(wt.img.4$y)>pctl_95]
+top_05_image <- midwt(top_05, daubcqf(8)$h.0, L=4)
+img.plot(top_05_image$x)
+
+pctl_90 <- quantile(abs(wt.img.4$y), probs=0.90)
+top_10 <- matrix(0, nrow=256, ncol=256)
+top_10[abs(wt.img.4$y)>pctl_90] <- wt.img.4$y[abs(wt.img.4$y)>pctl_90]
+top_10_image <- midwt(top_10, daubcqf(8)$h.0, L=4)
+img.plot(top_10_image$x)
+
+```
+
 Although there is some arbitrariness here, you can assign the outer(h,g) wavelets along the top.
@@
 Layout of horizontal wavelets
-
 the outer(g,h) wavelets along the side
@@
 
-If you examine the contribution of each wavelet frequency to the image, you will see a distinct patter.
+If you examine the contribution of each wavelet frequency to the image, you will see a distinct pattern.
 

### 2022-02-12 text/14/wavelets.Rmd

@@
 ```{r read-ecg-data}
+library("ggplot2")
+library("magrittr")
 library("rwt")
@@
 
-These coefficients represent the "high pass" filter. There is a complementary "low pass" filter that you get by reversing the coefficients and changing the sign of every other value.
-
-Note: To get the math to work on my end, tThe low pass filter provided by the rwt package needs to be negated.
+These coefficients represent the "high pass" filter.
+There is a complementary "low pass" filter that you get by reversing the coefficients and changing the sign of every other value.
+The term "high pass" means that it passes high frequencies above a certain cutoff and attenuates low frequencies.
+The term "low pass" means the opposite.
+
+Note that not all wavelets develop the low pass filter in the way described above. The construction is known as a quadrature mirror filter. There is an attractive simplicity to this approach.
 
@@
 
-These two filter pairs are known as a quadrature mirror filter because of the reflection. These two filters are orthonormal.
+These two filters are orthonormal.
 
@@
 
+```{r layout}
+x <- c(700, 300, 700)
+y <- c(700, 300, 300)
+m <- c(
+  "W1\nh %*% g", 
+  "W1\ng %*% h", 
+  "W1\nh %*% h"
+)
+data.frame(x, y, m) %>%
+  ggplot(aes(x, y, label=m)) +
+  expand_limits(x=c(100, 900), y=c(100, 900)) + 
+  geom_segment(x=100, y=100, xend=900, yend=100) +
+  geom_segment(x=900, y=100, xend=900, yend=900) +
+  geom_segment(x=900, y=900, xend=100, yend=900) +
+  geom_segment(x=100, y=900, xend=100, yend=100) +
+  geom_segment(x=500, y=100, xend=500, yend=900) +
+  geom_segment(x=100, y=500, xend=900, yend=500) +
+  theme_void() + 
+  geom_text() -> layout_1
+layout_1 +
+    geom_text(x=300, y=700, label="V1\ng %*% g")
+layout_1+
+  geom_segment(x=300, y=500, xend=300, yend=900) +
+  geom_segment(x=100, y=700, xend=500, yend=700) +
+  geom_text(x=400, y=800, label="W2\nh %*% g") + 
+  geom_text(x=200, y=600, label="W2\ng %*% h") + 
+  geom_text(x=400, y=600, label="W2\nh %*% h") -> layout_2
+layout_2 +
+  geom_text(x=200, y=800, label="V2\ng %*% g")
+layout_2 +
+  geom_segment(x=200, y=700, xend=200, yend=900) +
+  geom_segment(x=100, y=800, xend=300, yend=800) +
+  geom_text(x=250, y=850, label="W3") + 
+  geom_text(x=150, y=750, label="W3") + 
+  geom_text(x=250, y=750, label="W3") + 
+  geom_text(x=150, y=850, label="etc.")
+
+```
+
 ```{r level-1-image-g-g}

### 2022-03-12 text/22/baseline-covariate.Rmd
Kajor update to 14/history-of-r

@@
 ---
-title: "Mapping latitude and longitude"
+title: "Using change scores"
 author: "Steve Simon"

### 2022-04-14 text/22/r-creatures.Rmd
Added r-creatures.Rmd

---
title: "Different creatures in the R zoo"
author: "Steve Simon"
source: new
date: "2022-04-14"
category: 
- Blog post
tags:
- R programming
output: html_document
---

R has a variety of ways of storing information. It is a hodge-podge of different things. I'll call these "things" creatures, though that is probably not the right word. Let me review some of the more important creatures in R.

### Scalars

A single number, string, or date is called a scalar.

```{r}
secret_of_the_universe <- 42
on_first <- "Who"
y2k <- as.Date("2000-01-01")
```

### Vectors

A combination of two or more scalars of the same type is a vector. Strictly speaking, you say that a scalar is actually a vector of length 1. That does work for the most part. You can use a scalar in places where R is expecting a vector. Even so, most people reserve the term "vector" for anything with a length of 2 or more.

You use the `c` function to combine two or more scalars into a vector.

```{r}
buckle_my_shoe <- c(1, 2)
vowels <- c("A", "E", "I", "O", "U")
```

You can use the colon to create a vector that is sequence of consecutive integers.

```{r}
fantastic_four <- 1:4
fantastic_four
not_so_fantastic_four <- 5:8
not_so_fantastic_four
```

You can use the `length` function to calculate the length of a vector.

```{r}
length(vowels)
```

You get individual elements of a vector using square brackets.

```{r}
vowels[1]
```

You can also use square brackets to extract chunks from a vector

```{r}
vowels[1:2]
vowels[c(2, 4)]
```

You can use `c` to make a longer vector out of two shorter vectors.

```{r}
c(buckle_my_shoe, fantastic_four)
```

### Named vectors

You can assign names to some or all of the individual elements in a vector. This is not done too frequently, but it does have its uses.

```{r}
primes <- c(
  "first"=2,
  "second"=3,
  "third"=5,
  "fourth"=7)
primes
```

You can extract individual elements of a named vector using the names instead of their numeric positions.

```{r}
primes["first"]
primes[c("second", "fourth")]
```

You list the names of a named vector using the `names` function.

```{r}
names(primes)
```

### Lists

You can't mix and match within a vector. If you try to combine a number and a string, for example, R will convert the number to a string first.

```{r}
mixed_vector <- c(1, "two")
mixed_vector
```

You can mix numbers and strings in a list, which is created using the `list` function.

```{r}
mixed_list <- list(1, "two")
mixed_list
```

Notice the odd collection of double and single brackets in the list. Why R uses single brackets in some places and double brackets in other locations is something that can confound even an experienced R programmer. The only important thing to remember as a beginner is that double brackets allow you to select individual items from a list.

```{r}
mixed_list[[2]]
```

You can create lists that are mix of some of the creatures shown below. You can even make lists of lists.

```{r}
groceries <- list("bread", "fruit")
clothes <- list("socks", "underwear", "shirts")
shopping <- list(groceries, clothes)
shopping
```

This can get pretty messy. You can recast all the individual components into a single vector using the `unlist` function.

```{r}
unlist(shopping)
```

Of course, you lose the ability to mix and match once you unlist.

You can (and probably should) use names for each element of your list. This makes it easier to pick out pieces of list.

```{r}
names(shopping) <- c("grocery_list", "clothing_list")
shopping <- list(grocery_list=groceries, clothing_list=clothes)
shopping
shopping[["clothing_list"]]
```

If the list has names, you can use the dollar sign as a substitute for the double square brackets.

```{r}
shopping$clothing_list
```

### Matrices and arrays

Matrices are a two dimensional arrangement of scalars of the same type.

```{r}
diagonal_matrix <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), nrow=3)
diagonal_matrix
```

You can get the number of rows and columns of a matrix using the `dim` function.

```{r}
dim(diagonal_matrix)
```

Use single square brackets to extract individual elements or chunks of a matrix.

```{r}
diagonal_matrix[1, 3]
diagonal_matrix[1:2, 1:2]
```

To extract an entire row or column from a matrix, leave the index blank.

```{r}
diagonal_matrix[1, ]
diagonal_matrix[ , 3]
```

You can assign names to the rows and/or columns of a matrix using the `dimnames` function.

```{r}
dimnames(diagonal_matrix) <- list(c("A", "B", "C"), c("D", "E", "F"))
diagonal_matrix
```

Arrays are a three or higher dimensional arrangement of scalars of the same type. They work pretty much the same way that matrices work, so I won't show any examples here.

You can combine vectors into a matrix using the `cbind` or `rbind` functions. Notice the difference in how the two functions work.

```{r}
cbind(fantastic_four, not_so_fantastic_four)
rbind(fantastic_four, not_so_fantastic_four)
```

### Data frames

A data frame is a list of vectors, where each vector can be a different type, but they all have to have the same length.

It is a compromise between the simple but rigid matrix, and the complex but flexible list. The data frame is a very popular approach for storing datasets.

The data shown here is real, and it comes from the [Data and Story Library][dasl]. I added some fictional names to illustrate how you can mix and match vectors of different types.

```{r}
chain <- c(
  "Burger Queen",
  "McBurger",
  "Hardly's",
  "Out-n-In",
  "What? A Burger?",
  "Burger Bell",
  "Hedgehog")

fat <- c(
  19,
  31,
  34,
  35,
  39,
  39,
  43)

sodium <- c(
  920,
  1500,
  1310,
  860,
  1180,
  940,
  1260)

calories <- c(
  410,
  580,
  590,
  570,
  640,
  680,
  660)

fast_food <- data.frame(chain, fat, sodium, calories)
fast_food
```

Notice how R created names for each column. It also allows you to create names for each row. Here's what it would look like if you used `chain` as names for the rows.

```{r}
fast_food_alt <- data.frame(fat, sodium, calories)
rownames(fast_food_alt) <- chain
fast_food_alt
```

The current thinking among R professionals is that a variable like `chain` belongs as a direct part of the data frame rather than as a row name, but both approaches are still common.

You can select individual columns of a data frame using numbers or names.

```{r}
fast_food[, 2]
fast_food[, "fat"]
fast_food$fat
```

The last approach, using the dollar sign, is the most common way to work with a single column of a data frame.

### Tibbles

The programming team working on the tidyverse has developed an alternative to the data frame called a tibble. You use the `tibble` function to create a tibble. But first you have to load the tidyverse packages (or just the tibble package by itself)

```{r}
library(tidyr)
fast_food_tibble <- tibble(chain, fat, sodium, calories)
```

It works mostly like a data frame but with a few important differences.

First, as you can see above, the tibble describes itself as a tibble when you print it, and it lets you know what type of data is in each column.

Second, when you print a tibble, it refrains from presenting all of the data for anything except the most modest-sized of datasets.

It does become important, however, when you work with data with more than just a few rows or more than just a few columns.

There is a dataset in the survival package about colon cancer that has 1,848 observations (rows) and 16 variables (columns). Notice how only a portion of the data is shown.

```{r}
library(survival)
tibble(colon)
```

You can ask a tibble to display more or all of itself.

```{r}
print(tibble(colon), n=25)
```

Another important difference is in how tibbles handle subsetting. Look carefully at these examples.

```{r}
fast_food[5, ]
fast_food_tibble[5, ]
fast_food[ , 2]
fast_food_tibble[ , 2]
fast_food[5, 2]
fast_food_tibble[5, 2]
```

When you extract a single row from a data frame, it just becomes a smaller data frame. Same for a tibble. When extract a single row from a tibble, it just becomes a smaller tibble.

But notice how this changes when you extract a single column. With the data frame, the single column is displayed horizontally, because R has converted it to a vector. With the tibble, there is no conversion and a single column subset from a tibble is still a tibble.

When you extract a single value from a data frame, it displays it as a scalar. A single value from a tibble is still a tibble, albeit a very small tibble.

The point is that subsets of a data frame are sometimes implicitly converted to a vector or scalar and sometimes not. Programmers sometimes forget about the implicit conversion and treat any subset of a data frame as a smaller data frame. This is a common source of bugs in R programs. Tibbles force you to do the conversion explicitly, which may mean a bit of extra work, but the code, especially for complex operations, is more reliable.

### A practical illustration

The examples shown earlier are mostly just simple cute examples, but here is a practical problem that shows some of the various creatures in the R zoo.

Let's use the data frame on fast food burgers. Pretend that you have discovered that the "Out-n-In" burger (in the fourth row) is actually made of tofu and should be discarded from the analysis.

```{r}
real_fast_food <- fast_food[-4, ]
```

You want to run a regression analysis and examine the various statistics from the output individually. You store the output in a new variable, fast_food_regression.

```{r}
fast_food_regression <- lm(calories~fat+sodium, data=real_fast_food)
```

Since the output is complex and a mixture of different elements, R uses a list to store everything.

```{r}
names(fast_food_regression)
```

Let's look at the first item in the list, coefficents.

```{r}
fast_food_regression$coefficients
```

This is a named vector with values for the intercept, the slope for the fat variable, and the slope for the sodium variable.

Now let's look at the residuals and fitted values.

```{r}
fast_food_regression$residuals
fast_food_regression$fitted.values
```

These are both vectors.

Let's look at df.residual.

```{r}
fast_food_regression$df.residual
```

This is a scalar. There are 3 degrees of freedom for error.

Let's look at qr. This will provide some rather complex details about some intermediate calculations that R used to produce this regression model.

```{r}
fast_food_regression$qr
```

This is a list inside the list. If you are not familiar with what the qr decomposition is or how it works, then this list has little value for you. For some advanced and specialized applications, knowledge of the qr decomposition is very important. Such information is very difficult to get from other statistical packages like SAS or SPSS.

The summary function produces another list from the list we just examined. This provides more information, such as numbers needed to reproduce some of the important tests and measures for this regression model.

```{r}
fast_food_summary <- summary(fast_food_regression)
names(fast_food_summary)
```

The third item in this list is coefficients, and it is different from the coeeficients item in the previous list.

```{r}
fast_food_summary$coefficients
```

This is a matrix with names for each row and for each column. If you wanted to produce a nice table summarizing the results of the regresion model, customized to your specifications, here is where you would start.

You might also be interested in some post-processing of the regression results. You can, for example, extract the p-values for the two slopes.

```{r}
fast_food_summary$coefficients[c("fat", "sodium"), "Pr(>|t|)"]
```

You might need to do this if you wanted to make a Bonferroni correction to the p-values.

Let's look at r.squared and adj.r.squared.

```{r}
fast_food_summary$r.squared
fast_food_summary$adj.r.squared
```

These are scalars. These are also key numbers for your custom regression summary.

Let's look at fstatistic.

```{r}
fast_food_summary$fstatistic
```

This is a named vector with the statistic itself and the numerator and denominator degrees of freedom. Yet another key piece.

There's a lot more that you could explore in each of these lists. Some of it is only for advanced applications. I am showing it here, just to show you the wide range of creatures that R provides for its regression models.

### Summary

There are many creatures in the R zoo: scalars, vectors, named vectors, lists, matrices, arrays, data frames, and tibbles are some of the more important ones. Vectors, matrices, and arrays require you to use the same type of data (number, string, or date) for each element. Lists and data frames allow you to mix and match different types of data. You got to see the inside of a couple of real lists used by R and view the different creatures found in each list.

[dasl]: https://dasl.datadescription.com/datafile/burgers

### 2022-05-24 text/17/first-jags.Rmd
Added writing-goals-may-2022

---
title: "How to run your first Bayesian analysis using jags software in R"
author: "Steve Simon"
source: "http://blog.pmean.com/first-jags/"
date: "2017-01-28"
category: Blog post
tags: 
- Bayesian statistics
- Incomplete pages
output: html_document
---

Someone wanted to know how to run a Bayesian data analysis for a two group longitudinal study. There are several ways you can do this, but I had to confess I did not have an immediate answer. So I took some time to figure out how to do this using jags software inside of R. I’ve done a fair amount of stuff in jags, but not anything close to a longitudinal design. The general principle is to start with something easy and work your way slowly up to the final analysis.

You can find the code and some supporting text files on [my github site][sim3].

I usually start with a very simple Bayesian analysis because if something is awry, I’ll notice it faster with a simple example. The code below implements a simple beta-binomial model.

Make sure that you have [jags installed][jags1] on your computer as well as the [rjags package][jags2]

### Beta-binomial example

Assume you have a beta prior distribution with alpha=4 and beta=16. You collect 60 observations and observe 18 successes and 42 failures.

You don’t really need jags for this. The posterior distribution is beta with alpha=4+18=22 and beta=16+42=58. The posterior mean is 0.275 and the posterior standard deviation is 0.05.

```{}
library("rjags")
## Loading required package: coda
## Linked to JAGS 4.2.0
## Loaded modules: basemod,bugs
fnm <- "jags_beta_binomial.txt"
mon <- c("pi")

out <- list(description="jags output")

dat <- list(a=4, b=16, x=18, n=60)
mod <- jags.model(fnm, data=dat, quiet=TRUE)
out$bb1 <- 
  coda.samples(mod, variable.names=mon, n.iter=1000, by=1000, progress.bar=NULL)
summary(out$bb1)
```

```{}
## 
## Iterations = 1:1000
## Thinning interval = 1 
## Number of chains = 1 
## Sample size per chain = 1000 
## 
## 1. Empirical mean and standard deviation for each variable,
##    plus standard error of the mean:
## 
##           Mean             SD       Naive SE Time-series SE 
##       0.273157       0.049814       0.001575       0.001575 
## 
## 2. Quantiles for each variable:
## 
##   2.5%    25%    50%    75%  97.5% 
## 0.1811 0.2367 0.2718 0.3086 0.3698
```

### Example from BUGS manual

Now, try to run one of the examples in the BUGS manuals that is reasonably close to your problem. For our problem, the [very first example in volume 1][bugs1] works well.

I had to fix two things to get this example to run properly in jags. First, I had to remove some lines of code that used functions like post.p.value. Second, I have to transpose the matrix of data. But with these changes it ran just fine.

```{}
dat <- list(x = c(8.0, 15.0, 22.0, 29.0, 36.0), xbar = 22, N = 30, T = 5,   
      Y = structure(
         .Data = c(151, 199, 246, 283, 320,
                     145, 199, 249, 293, 354,
                     147, 214, 263, 312, 328,
                     155, 200, 237, 272, 297,
                     135, 188, 230, 280, 323,
                     159, 210, 252, 298, 331,
                     141, 189, 231, 275, 305,
                     159, 201, 248, 297, 338,
                     177, 236, 285, 350, 376,
                     134, 182, 220, 260, 296,
                     160, 208, 261, 313, 352,
                     143, 188, 220, 273, 314,
                     154, 200, 244, 289, 325,
                     171, 221, 270, 326, 358,
                     163, 216, 242, 281, 312,
                     160, 207, 248, 288, 324,
                     142, 187, 234, 280, 316,
                     156, 203, 243, 283, 317,
                     157, 212, 259, 307, 336,
                     152, 203, 246, 286, 321,
                     154, 205, 253, 298, 334,
                     139, 190, 225, 267, 302,
                     146, 191, 229, 272, 302,
                     157, 211, 250, 285, 323,
                     132, 185, 237, 286, 331,
                     160, 207, 257, 303, 345,
                     169, 216, 261, 295, 333,
                     157, 205, 248, 289, 316,
                     137, 180, 219, 258, 291,
                     153, 200, 244, 286, 324),
                  .Dim = c(5,30)))

dat$Y <- t(dat$Y)
init <- list(alpha = c(250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250,
                  250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250, 250),
      beta = c(6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6,
                  6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6, 6),         
      alpha.c = 150, beta.c = 10,
      tau.c = 1, alpha.tau = 1, beta.tau = 1)

fnm <- "jags_rats.txt"
mon <- c("alpha0", "beta.c", "sigma")

mod <- jags.model(fnm, dat, init, quiet=TRUE)
out$rats <- 
  coda.samples(mod, variable.names=mon, n.iter=1000, by=1000, progress.bar=NULL)
summary(out$rats)
```

```{}
## 
## Iterations = 1:1000
## Thinning interval = 1 
## Number of chains = 1 
## Sample size per chain = 1000 
## 
## 1. Empirical mean and standard deviation for each variable,
##    plus standard error of the mean:
## 
##           Mean     SD Naive SE Time-series SE
## alpha0 106.283 3.5884 0.113476       0.125482
## beta.c   6.196 0.1072 0.003391       0.004276
## sigma    6.108 0.6291 0.019893       0.033978
## 
## 2. Quantiles for each variable:
## 
##          2.5%     25%     50%     75%   97.5%
## alpha0 99.495 104.008 106.248 108.670 113.344
## beta.c  5.996   6.128   6.193   6.268   6.399
## sigma   5.205   5.735   6.060   6.402   7.227
```

### Adapt to your data

Now let’s see if we can modify this example to fit our particular setting.

Our longitudinal setting has two measurements: pre and post. So strip the last three columns from the matrix Y. There is one within subject factor, time, which is represents by X1. There is a between subjects factor, treatment group, which is represented by X2. Let’s pretend for now that that the first twenty observations represent your treatment group and the last ten observations represent your control group.

```{}
dat <- list(X1 = c(0, 1), X2 = rep(0:1, c(20, 10)), N = 30, T = 2, 
      Y = structure(
         .Data = c(151, 199,
                     145, 199,
                     147, 214,
                     155, 200,
                     135, 188,
                     159, 210,
                     141, 189,
                     159, 201,
                     177, 236,
                     134, 182,
                     160, 208,
                     143, 188,
                     154, 200,
                     171, 221,
                     163, 216,
                     160, 207,
                     142, 187,
                     156, 203,
                     157, 212,
                     152, 203,
                     154, 205,
                     139, 190,
                     146, 191,
                     157, 211,
                     132, 185,
                     160, 207,
                     169, 216,
                     157, 205,
                     137, 180,
                     153, 200),
                  .Dim = c(2,30)))

dat$Y <- t(dat$Y)
init <- list(alpha = rep(150, 30), beta1 = rep(50, 30),
      beta1.c = 50, beta1.tau=1,
      beta2 = 20, beta3 = 5,
      alpha.c = 150,
      tau.c = 1, alpha.tau = 1)

fnm <- "jags_lon.txt"
mon <- c("alpha.c", "beta1.c", "beta2", "beta3")

mod <- jags.model(fnm, dat, init, quiet=TRUE)
out$lon <- 
  coda.samples(mod, variable.names=mon, n.iter=1000, by=1000, progress.bar=NULL)
summary(out$lon)
```

```{}
## 
## Iterations = 1:1000
## Thinning interval = 1 
## Number of chains = 1 
## Sample size per chain = 1000 
## 
## 1. Empirical mean and standard deviation for each variable,
##    plus standard error of the mean:
## 
##            Mean    SD Naive SE Time-series SE
## alpha.c 154.095 2.576  0.08145         0.4494
## beta1.c  50.186 1.136  0.03593         0.1414
## beta2    -5.756 4.172  0.13193         1.4509
## beta3    -1.593 1.723  0.05447         0.5767
## 
## 2. Quantiles for each variable:
## 
##            2.5%     25%      50%      75%    97.5%
## alpha.c 148.583 152.538 154.2840 155.7871 158.7248
## beta1.c  47.995  49.452  50.1588  50.9269  52.4221
## beta2    -8.877  -7.483  -7.1730  -6.8746   9.7413
## beta3    -5.556  -2.352  -0.9216  -0.6312   0.9299
```

Okay, so far, so good. Your next step would be to put your own data in place of the data from the example. I won’t show that here because I don’t have your data. But I hope you can try the general outline suggested here.

You can find an [earlier version][sim1] of this page on my [blog][sim2].

[sim1]: http://blog.pmean.com/first-jags/
[sim2]: http://blog.pmean.com

[bugs1]: http://www.openbugs.net/Examples/Rats.html
[jags1]: http://sourceforge.net/projects/mcmc-jags/files/
[jags2]: https://cran.r-project.org/web/packages/rjags/index.html
[sim3]: https://github.com/pmean/bayesian-longitudinal

### 2022-05-25 src/date_files.R
Added date_files.R

@@ -1,0 +1,33 @@
+# copy_files.R. This program was written by Steve Simon on 2020-08-19.
+
+suppressMessages(suppressWarnings(library(lubridate)))
+suppressMessages(suppressWarnings(library(tidyverse)))
+
+source(file="src/standard_functions.R")
+
+v <- TRUE
+
+get_dates <- function(text_path="text/") {
+  subdir_list <- c(99, paste0("0", 0:9), 10:22)
+  file_times <- NULL
+  for (i in subdir_list) {
+    sub_path <- paste0(text_path, i)
+    file_names <- list.files(path=sub_path, pattern="md$")
+    full_name <- paste0(sub_path, "/", file_names)
+    file.info(full_name) %>% 
+      data.frame %>%
+      bind_rows(., file_times) -> file_times
+  }
+  file_times %>%
+    data.frame %>%
+    rownames_to_column("yr") %>%
+    select(yr, mtime) %>%
+    mutate(yr=str_sub(yr, 6, 7)) %>%
+    mutate(mtime=str_sub(mtime, 1, 10)) %>%
+    mutate(mtime=ymd(mtime)) %>%
+    return
+}
+all_dates <- get_dates()
+all_dates %>% count(mtime)
+glimpse(all_dates)
+all_dates %>% count(mtime)

### 2022-05-26 src/date_files.R
Updated 17/probonto.md

@@ -1,33 +1,0 @@
-# copy_files.R. This program was written by Steve Simon on 2020-08-19.
-
-suppressMessages(suppressWarnings(library(lubridate)))
-suppressMessages(suppressWarnings(library(tidyverse)))
-
-source(file="src/standard_functions.R")
-
-v <- TRUE
-
-get_dates <- function(text_path="text/") {
-  subdir_list <- c(99, paste0("0", 0:9), 10:22)
-  file_times <- NULL
-  for (i in subdir_list) {
-    sub_path <- paste0(text_path, i)
-    file_names <- list.files(path=sub_path, pattern="md$")
-    full_name <- paste0(sub_path, "/", file_names)
-    file.info(full_name) %>% 
-      data.frame %>%
-      bind_rows(., file_times) -> file_times
-  }
-  file_times %>%
-    data.frame %>%
-    rownames_to_column("yr") %>%
-    select(yr, mtime) %>%
-    mutate(yr=str_sub(yr, 6, 7)) %>%
-    mutate(mtime=str_sub(mtime, 1, 10)) %>%
-    mutate(mtime=ymd(mtime)) %>%
-    return
-}
-all_dates <- get_dates()
-all_dates %>% count(mtime)
-glimpse(all_dates)
-all_dates %>% count(mtime)

### 2022-06-01 text/22/testing-d3.Rmd
Added testing-d3

---
title: "Testing d3"
author: "Steve Simon"
date: "2022-06-01"
source: "New"
categories: 
- Blog post
tags: 
- Graphical display
output: html_document
---

I am interested in developing some visualization applications using d3. You can run d3 from within R Markdown.

This is an example that I borrowed from [Section 15.5][xie1] of [R Markdown Cookbook][xie2] by Yihui Xie, Christophe Dervieus, and Emily Riederer.

First, load the package **r2d3** to set up the `d3` engine
for **knitr** automatically:

```{r setup}
library(r2d3)
```

Now we can generate data in R, pass it to D3, and draw
the chart:

```{d3, data=runif(30), options=list(color='steelblue')}
svg.selectAll('rect')
  .data(data)
  .enter()
    .append('rect')
      .attr('width', function(d) { return d * 672; })
      .attr('height', '10px')
      .attr('y', function(d, i) { return i * 16; })
      .attr('fill', options.color);
```

[xie1]: https://bookdown.org/yihui/rmarkdown-cookbook/d3.html
[xie2]: https://bookdown.org/yihui/rmarkdown-cookbook/

### 2022-06-05 src/link_files.Rmd
Added link-files.Rmd

---
title: "Compute last modified dates"
author: "Steve Simon"
date: "Created 2022-05-26"
output: html_document
---

```{r}
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(tidyverse)))

source(file="standard_functions.R")

v <- TRUE
```

```{r}
old_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
old_pattern <- "html$"
new_path <- "../text/"
new_pattern <- "md$"
subdir_list=c("99", paste0("0", 0:9), 10:22)
for (yr in subdir_list) {
  print(yr)
  file_path <- paste0(new_path, yr)
  file_path %>%
    list.files(pattern="md$") -> new_list
  new_list %>%
    str_remove_all("-") %>%
    str_remove_all(fixed(".md")) %>%
    str_remove_all(fixed(".Rmd")) %>%
    tolower %>%
    data.frame %>%
    set_names("key") %>%
    mutate(y=yr) %>%
    bind_cols(data.frame(md=new_list)) -> new_names
  file_path <- paste0(old_path, yr)
  file_path %>%
    list.files(pattern="html$") -> old_list
  old_list %>%
    tolower %>%
    str_remove_all(fixed(".html")) %>%
    data.frame %>%
    set_names("key") %>%
    mutate(y=yr) %>%
    bind_cols(data.frame(html=old_list)) %>%
    full_join(new_names, by=c("y", "key")) %>%
    arrange(key) %>%
    filter(is.na(md)|is.na(html)) %>%
    select(-key) %>%
    mutate(md=ifelse(is.na(md), "", md)) %>%
    mutate(html=ifelse(is.na(html), "", html)) %>%
    print
}
```

```{r, eval=FALSE}

subdir_list <- c(99, paste0("0", 0:9), 10:22)
for (i in subdir_list[10]) {
  get_names(old_path, old_pattern, i) %>% print
  get_names(new_path, new_pattern, i) %>% print
}
```

### 2022-06-07 src/link_files.Rmd
Added deploy-now.md

@@
 subdir_list=c("99", paste0("0", 0:9), 10:22)
+combined_list <- NULL
+missing_links <- NULL
 for (yr in subdir_list) {
@@
     list.files(pattern="md$") -> new_list
+  repeat_names <- intersect(combined_list, new_list)
+  if (length(repeat_names) > 0) print(repeat_names)
   new_list %>%
@@
     mutate(html=ifelse(is.na(html), "", html)) %>%
-    print
+    bind_rows(missing_links) -> missing_links
 }
+missing_links
 ```
+
+Next step, search for 
++ [sim1]: http://www.pmean.com/??/...
++ <p>This page has moved to <a href="http://new.pmean.com/.../">a new website</a>.</p>
 

### 2022-07-06 text/22/different-parameterizations.Rmd
Started work on different-parameterizations.md

---
title: "Different parameterizations for one factor ANOVA"
author: "Steve Simon"
source: new
date: "2022-07-06"
category: 
- Blog post
tags:
- Analysis of variance
output: html_document
---

I'm trying to understand some of the basics of contrasts in analysis of variance. In particular, I want to figure out some alternative parameterizations of the coefficients that are estimated as part of the ANOVA model.

```{r setup}
library(MASS)
library(matlib)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

First, let's set up an artificial dataset that is nice and easy to work with.

```{r}
means <- c(50, 52, 60, 70)
mean(means)
y <- 
  rep(means, each=3) +
  rep(c(0, 1, -1), 4)
g <- rep(1:4, each=3)
df <- cbind(g, y)
```

Notice the pattern in the group means.

```{r}
data.frame(g, y) %>%
  group_by(g) %>%
  summarize(mean_y=mean(y))
```

I want to use the lm function to fit the data, because it is easier to peek at the underlying coefficients.

### The simplest parameterization

The simplest parameterization is to let each beta coefficient represent a different mean. We have to drop the intercept term for this model to work properly.

```{r}
x1 <- as.numeric(g==1)
x2 <- as.numeric(g==2)
x3 <- as.numeric(g==3)
x4 <- as.numeric(g==4)
```

This is what the X matrix looks like.

```{r}
x <- cbind(x1, x2, x3, x4) 
x
```

This is what the regression results look like.

```{r}
lm(y~x1+x2+x3+x4-1)
```

It may help to view the underlying matrix calculations. Recall that

$\hat{\beta}=(X'X)^{-1}X'Y$

So the matrix

$\hat{\beta}=(X'X)^{-1}X'$

will tell you how each parameter in the regression model is a linear combination of the Y's.

```{r}
inv(t(x)%*%x)%*%t(x) %>% fractions
```

So the first coefficient is an average of the first three y values, the second is an average of the next three y values, etc.

### Indicator variables

A common way to parameterize the ANOVA model is to use an intercept term and indicator variables for all but one of the groups.

```{r}
x0 <- rep(1, 12)
x1 <- as.numeric(g==2)
x2 <- as.numeric(g==3)
x3 <- as.numeric(g==4)

```

This is what the X matrix looks like.

```{r}
x <- cbind(x0, x1, x2, x3) 
x
```

This is what the regression results look like.

```{r}
lm(y~x0+x1+x2+x3-1)
```

```{r}
inv(t(x)%*%x)%*%t(x) %>% fractions
```

So the first coefficient is an average of the first three y values (Y1, Y2, Y3), the second is the difference between the average of the second three y values (Y4, Y5, Y6) and the average of the first three y values (Y1, Y2, Y3), etc.

This seems counter intuitive. The slope coefficient for $\beta_2$ involves a column with zeros everywhere except for some ones for Y4, Y5, and Y6. How could that coefficient end up involving Y1, Y2, and Y3?

The point to remember is that the slope coefficient in a linear regression with multiple independent variables has a slightly more complex interpretation than the slope coefficient in a linear regression with a single independent variable. In the later case, the interpretation would be

+ the estimated average change in $Y$ when $X$ increases by one unit.

In the former case, the interpretation would be

+ the estimated average change in $Y$ when $X_i$ increases by one unit **while holding all the other** $X$**'s constant**.

So the slope associated with X1 requires you to hold X2 and X3 constant. You can hold them constant at 0 or hold them constant at 1, but the former is simpler to understand. THink of it as moving from one of the first three rows of your X matrix.

```{r}
x[1:3, ]
```

to one of the next three rows of your X matrix

```{r}
x[4:6, ]
```

This is effectively a change from the mean of the first three observations (Y1, Y2, Y3) subtracted from the mean of the next three observations (Y4, Y5, Y6).

In writing this webpage, I relied on a very helpful tutorial paper:

Daniel J. Schad, Shravan Vasishth, Sven Hohenstein, Reinhold Kliegl. How to capitalize on a priori contrasts in linear (mixed) models: A tutorial. Journal of Memory and Language, 2020, 110. Available in [html format][sch1] or [pdf format][sch2].

as well as a page on the UCLA Statistics site

R Library Contrast Coding Systems for Categorical Variables. Available in [html format][ucla1]

[ucla1]: https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/

[sch1]: https://doi.org/10.1016/j.jml.2019.104038
[sch2]: https://www.sciencedirect.com/science/article/pii/S0749596X19300695/pdfft?md5=05f53f6c3386f0e0c6ae203178b33e33&pid=1-s2.0-S0749596X19300695-main.pdf

### 2022-07-07 text/22/different-parameterizations.Rmd
Finished different-parameterizations.Rmd

@@
 ```{r}
-means <- c(50, 52, 60, 70)
+means <- c(52, 54, 62, 72)
 mean(means)
@@
 ```{r}
-x <- cbind(x1, x2, x3, x4) 
-x
+A <- cbind(x1, x2, x3, x4) 
+A
 ```
@@
 ```{r}
-lm(y~x1+x2+x3+x4-1)
+beta_A <- coef(lm(y~x1+x2+x3+x4-1))
+beta_A
 ```
@@
 
-$\hat{\beta}=(X'X)^{-1}X'$
+$(X'X)^{-1}X'$
 
@@
 ```{r}
-inv(t(x)%*%x)%*%t(x) %>% fractions
+inv(t(A)%*%A)%*%t(A) %>% fractions
 ```
@@
 ```{r}
-x <- cbind(x0, x1, x2, x3) 
-x
+B <- cbind(x0, x1, x2, x3) 
+B
 ```
@@
 ```{r}
-lm(y~x0+x1+x2+x3-1)
+beta_B <- coef(lm(y~x0+x1+x2+x3-1))
+beta_B
 ```
@@
 ```{r}
-inv(t(x)%*%x)%*%t(x) %>% fractions
+inv(t(B)%*%B)%*%t(B) %>% fractions
 ```
@@
 So the first coefficient is an average of the first three y values (Y1, Y2, Y3), the second is the difference between the average of the second three y values (Y4, Y5, Y6) and the average of the first three y values (Y1, Y2, Y3), etc.
+
+In other words,
+
+$\hat{\beta}_0=\bar{Y}_1$
+
+$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_1$
+
+$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_1$
+
+$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_1$
 
@@
 You can hold them constant at 0 or hold them constant at 1, but the former is simpler to understand.
-THink of it as moving from one of the first three rows of your X matrix.
+Think of it as moving from one of the first three rows of your X matrix.
 
 ```{r}
-x[1:3, ]
+B[1:3, ]
 ```
@@
 ```{r}
-x[4:6, ]
+B[4:6, ]
 ```
@@
 This is effectively a change from the mean of the first three observations (Y1, Y2, Y3) subtracted from the mean of the next three observations (Y4, Y5, Y6).
+
+### Deviation coding
+
+If you tried to code the difference in means directly, you get a third result.A common way to parameterize the ANOVA model is to use an intercept term and indicator variables for all but one of the groups.
+
+```{r}
+x0 <- rep(1, 12)
+x1 <- as.numeric(g==2) - as.numeric(g==1)
+x2 <- as.numeric(g==3) - as.numeric(g==1)
+x3 <- as.numeric(g==4) - as.numeric(g==1)
+```
+
+This is what the X matrix looks like.
+
+```{r}
+D <- cbind(x0, x1, x2, x3) 
+D
+```
+
+This is what the regression results look like.
+
+```{r}
+beta_D <- coef(lm(y~x0+x1+x2+x3-1))
+beta_D
+```
+
+```{r}
+inv(t(D)%*%D)%*%t(D) %>% fractions
+```
+
+Okay, so what is this 1/4 doing here? There are only three data values in each group. You need to rewrite
+
+1/4 = 1/3 - 1/12
+
+Then the -1/12 represents subtracting the overall mean of all 12 observations. So the estimates involve an average of a group minus the overall average. In other words,
+
+$\hat{\beta}_0=\bar{Y}_.$
+
+$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_.$
+
+$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_.$
+
+$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_.$
+
+where $\bar{Y}_.$ is the overall mean.
+
+Again, this seems counter-intuitive, but remember that the interpretation of the regression coefficient in a multivariate setting is the effect of a one unit change in a variable **holding all the other variables constant**.
+
+### Recode
+
+Suppose you want to recombine the columns of your X matrix to transform it using some linear combinations. This is the same as post-multiplying by a matrix.
+
+$B=AV$
+
+or equivalently
+
+$A=BV^{-1}$
+
+So to convert from
+
+```{r}
+A
+```
+
+into
+
+```{r}
+B
+```
+
+you would multiply by this V
+
+```{r, echo=FALSE}
+V <- matrix(c(
+  1,  0,  0,  0,
+  1,  1,  0,  0,
+  1,  0,  1,  0,
+  1,  0,  0,  1), nrow=4, byrow=TRUE)
+V
+```
+
+to get
+
+```{r}
+A%*%V
+```
+
+Note the V is not a symmetric matrix. Symmetry is not requirement here.
+
+You can go in the other direction using
+
+```{r}
+U=inv(V)
+U
+```
+
+which yields
+
+```{r}
+B%*%U
+```
+
+Now the formula for the estimates for a given parameterization is
+
+$\hat{\beta}_B=(B'B)^{-1}B'Y$
+
+Replace B with AV to get
+
+$\hat{\beta}_B=(V'A'AV)^{-1}V'A'Y$
+
+which simplifies, after a bit of work, to
+
+$\hat{\beta}_B=V^{-1}(A'A)^{-1}A'Y$
+
+or
+
+$\hat{\beta}_B=V^{-1}\hat{\beta}_A$
+
+Let's check this out with the coefficients we calculated earlier.
+
+```{r}
+beta_A
+```
+
+```{r}
+beta_B
+```
+
+```{r}
+inv(V)%*%beta_A
+```
+
+You can also go in the reverse direction.
+
+```{r}
+V%*%beta_B
+```
+
+### Check the deviation coding
+
+The same works with any coding. The V matrix that gets you from
+
+```{r}
+A
+```
+
+to
+
+```{r}
+D
+```
+
+is
+
+```{r, echo=FALSE}
+V <- matrix(c(
+  1, -1, -1, -1,
+  1,  1,  0,  0,
+  1,  0,  1,  0,
+  1,  0,  0,  1), nrow=4, byrow=TRUE)
+V
+```
+
+Let's check this in one direction
+
+```{r}
+A%*%V
+```
+
+and the opposite direction
+
+```{r}
+U=inv(V)
+U
+```
+
+```{r}
+D%*%U
+```
+
+We can establish that
+
+```{r}
+beta_A
+```
+
+can be converted to
+
+```{r}
+beta_D
+```
+
+using
+
+```{r}
+inv(V)%*%beta_A
+```
+
+and the reverse direction works as well.
+
+```{r}
+V%*%beta_D
+```
+
+### Make your own contrast
+
+Suppose you wanted to develop a parameterization that is meaningful for your particular setting and none of the ones discussed here or elsewhere seem to fit the bill.
+With a bit of work, you can calculate your own.
+
+Suppose for example that you wanted
+
+$\hat{\beta}_0=\bar{Y}_1$
+
+$\hat{\beta}_1=\bar{Y}_2-\bar{Y}_1$
+
+$\hat{\beta}_2=\bar{Y}_3-\bar{Y}_2$
+
+$\hat{\beta}_3=\bar{Y}_4-\bar{Y}_3$
+
+Instead of comparing each mean to the first mean, you want to compare each mean to the previous mean.
+
+The U matrix that converts
+
+$\hat{\beta}_0=\bar{Y}_1$
+
+$\hat{\beta}_1=\bar{Y}_2$
+
+$\hat{\beta}_2=\bar{Y}_3$
+
+$\hat{\beta}_3=\bar{Y}_4$
+
+to the new parameterization is
+
+```{r, echo=FALSE}
+U <- matrix(c(
+  1,  0,  0,  0,
+ -1,  1,  0,  0,
+  0, -1,  1,  0,
+  0,  0, -1,  1), nrow=4, byrow=TRUE)
+U
+```
+
+Now, you want to travel from the parameterization represented by
+
+```{r}
+A
+```
+
+to a new parameterization represented by E. This would need V, the inverse of U, which is
+
+```{r}
+V <- inv(U)
+V
+```
+
+which provides a matrix
+
+```{r}
+E <- A%*%inv(U)
+E
+```
+
+which gives the correct results.
+
+```{r}
+x0 <- E[ , 1]
+x1 <- E[ , 2]
+x2 <- E[ , 3]
+x3 <- E[ , 4]
+beta_E <- coef(lm(y~x0+x1+x2+x3-1))
+beta_E
+```
+### Acknowledgements
 

### 2022-07-08 text/22/analysis-of-means.Rmd
Added git-xkcd.md

---
title: "Three different approaches to multi-group comparisons"
author: "Steve Simon"
date: '2022-07-04'
output:
  word_document: default
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

I'm a big fan of Analysis of Variance (ANOVA). I use it all the time. I learn a lot from it. But sometimes I want something a bit different. And the difference comes down to how you specify the hypothesis.

### The traditional hypothesis

You've seen this hypothesis for the standard ANOVA setting in your introductory Statistics class

```{r, fig.width=5, fig.height=1, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[j] ~ " for at least one i,j")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.7, h0, adj=0)
text(0.1, 0.3, h1, adj=0)
```

This is a very common approach in a comparison of k independent groups. Test for any deviation from the null hypothesis using an F test. If this is statistically significant, then use a Tukey follow-up test to see which pairs of means differ from one another.

This hypothesis specifies equality by specifying that every pair of means is equal. For four groups, this implies six equalities: 1=2, 1=3, 1=4, 2=3, 2=4, and 3=4. For six groups, you would have fifteen equalities; for ten you'd have forty five.

It gets messy very fast. That's okay. If you have a lot of groups that you are comparing, you have to make it has to involve a lot of comparisons. Or do you?

### The placebo hypothesis

If one of the groups is a control or placebo, then you might consider an alternative formulation. Here's what the hypothesis looks like, assuming that the control is group #1.

```{r, fig.width=5, fig.height=1, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[1] ~ " for all i=2,...,k")
h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[1] ~ " for at least one i")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.7, h0, adj=0)
text(0.1, 0.3, h1, adj=0)
```

There's a procedure for this, Dunnett's test. It involves only three comparisons if you have four groups total, and only nine comparisons if you have ten groups total. You can quickly identify who is better than the control. This also gives you simplicity and a bit of extra power and precision.

Dunnett's test is available in most ANOVA programs, and is easy to implement. Test for any deviation from the null hypothesis using an F test. If this is statistically significant, then use Dunnett's test which mean(s) differ from the control mean.

Now you lose something when you simplify the hypothesis. Suppose you have six groups, a control and five different treatments. Now imagine that all of the treatments are significantly better than the control group. Jackpot! Every treatment is worth further study. But Dunnett's test won't allow you to see if some of the treatments are better than the others. There is no option for finding the best of the best.

Let's consider as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.

Let's consider a third option.

### Comparison to an overall mean

Suppose your goal is to establish whether any groups differ from the overall mean, Then you would write your hypothesis as

```{r, fig.width=5, fig.height=1, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu ~ " for at least one i")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.7, h0, adj=0)
text(0.1, 0.3, h1, adj=0)
```

This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results. If any do not, then you want to identify the group or groups that deviate from the norm and study them further.

It's important to editorialize a bit here. Deviating from the norm could be a good thing or a bad thing or it could be an indifferent thing. Your goal is not to use statistics to hunt out different groups to reward or punish them. You are using statistics to help in understanding why deviations from the norm occur.

https://www.frontiersin.org/articles/10.3389/fchem.2022.894547/full

### 2022-08-23 text/22/analysis-of-means.Rmd
Added five-versions-qualitative.md

@@
 ---
-title: "Three different approaches to multi-group comparisons"
+title: "A second alternative to the pairwise testing approach"
 author: "Steve Simon"
@@
 
-I'm a big fan of Analysis of Variance (ANOVA).
-I use it all the time.
-I learn a lot from it.
-But sometimes I want something a bit different.
-And the difference comes down to how you specify the hypothesis.
+In an earlier blog post, I mentioned Dunnett's test, a comparison of multiple treatments to a control group as an alternative to the pairwise approach to testing that is commonly used in analysis of variance.
+Here's a second interesting alternative.
+
+You are in a setting where you are expecting to see a degree of uniformity among several groups.
+For example, you have a group of clinics who all treat the same type of patients.
+The patients themselves may differ markedly from one to another, but they are assigned in a more or less random way to each clinic.
+So, although one patient may differ markedly from another, the average result should be the same for across all the clinics.
+
+The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient.
+But because of how the patients are assigned to a particular clinic, these should balance out on average.
+If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment.
+If you can identify a clinic (it could be more than one clinic, of course) that differs from the norm, then investigate.
+Perhaps that clinic is doing something "wrong" that causes the discrepancy, or maybe that clinic is the only one doing things right.
+In any case understanding why you see unexpected deviations in a process that should be uniform across clinics provides you with an opportunity to learn something.
 
@@
 
-This hypothesis specifies equality by specifying that every pair of means is equal.
-For four groups, this implies six equalities: 1=2, 1=3, 1=4, 2=3, 2=4, and 3=4.
-For six groups, you would have fifteen equalities; for ten you'd have forty five.
-
-It gets messy very fast. That's okay. If you have a lot of groups that you are comparing, you have to make it has to involve a lot of comparisons. Or do you?
-
-### The placebo hypothesis
-
-If one of the groups is a control or placebo, then you might consider an alternative formulation. Here's what the hypothesis looks like, assuming that the control is group #1.
-
-```{r, fig.width=5, fig.height=1, echo=FALSE}
-h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[1] ~ " for all i=2,...,k")
-h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[1] ~ " for at least one i")
-par(mar=rep(0.1, 4))
-plot(0:1, 0:1, axes=FALSE, type="n")
-text(0.1, 0.7, h0, adj=0)
-text(0.1, 0.3, h1, adj=0)
-```
-
-There's a procedure for this, Dunnett's test.
-It involves only three comparisons if you have four groups total, and only nine comparisons if you have ten groups total.
-You can quickly identify who is better than the control.
-This also gives you simplicity and a bit of extra power and precision.
-
-Dunnett's test is available in most ANOVA programs, and is easy to implement.
-Test for any deviation from the null hypothesis using an F test.
-If this is statistically significant, then use Dunnett's test which mean(s) differ from the control mean.
-
-Now you lose something when you simplify the hypothesis.
-Suppose you have six groups, a control and five different treatments.
-Now imagine that all of the treatments are significantly better than the control group.
-Jackpot!
-Every treatment is worth further study.
-But Dunnett's test won't allow you to see if some of the treatments are better than the others.
-There is no option for finding the best of the best.
-
-Let's consider as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.
-
-Let's consider a third option.
-
 ### Comparison to an overall mean
 
-Suppose your goal is to establish whether any groups differ from the overall mean, Then you would write your hypothesis as
+Suppose you are in a setting like the one described above, where there is an expectation of similarity from one r goal is to establish whether any groups differ from the overall mean, Then you would write your hypothesis as
 
@@
 
+### A simple example
+
 https://www.frontiersin.org/articles/10.3389/fchem.2022.894547/full

### 2022-08-23 text/22/dunnetts-test.Rmd
Added five-versions-qualitative.md

---
title: "Dunnett's test"
author: "Steve Simon"
date: '2022-07-04'
output:
  word_document: default
  html_document: default
category: Blog post
tags: Analysis of means
source: new
---

I'm a big fan of Analysis of Variance (ANOVA). I use it all the time. I learn a lot from it. But sometimes I want something a bit different. And the difference comes down to how you specify the hypothesis.

### The traditional hypothesis

Recall the way in which the hypothesis test is commonly set up when you are comparing three of more groups using analysis of variable.

```{r, fig.width=5, fig.height=1, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[j] ~ " for at least one i,j")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.7, h0, adj=0)
text(0.1, 0.3, h1, adj=0)
```

This is a very common approach in a comparison of k independent groups. Test for any deviation from the null hypothesis using an F test. If this is statistically significant, then use a Tukey follow-up test to see which pairs of means differ from one another.

This hypothesis specifies equality by specifying that every pair of means is equal. For four groups, this implies six equalities: 1=2, 1=3, 1=4, 2=3, 2=4, and 3=4. For six groups, you would have fifteen equalities; for ten you'd have forty five.

It gets messy very fast. That's okay. If you have a lot of groups that you are comparing, you have to make it has to involve a lot of comparisons. Or do you?

### The placebo hypothesis

If one of the groups is a control or placebo, then you might consider an alternative formulation. Here's what the hypothesis looks like, assuming that the control is group #1.

```{r, fig.width=5, fig.height=1, echo=FALSE}
h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[1] ~ " for all i=2,...,k")
h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[1] ~ " for at least one i")
par(mar=rep(0.1, 4))
plot(0:1, 0:1, axes=FALSE, type="n")
text(0.1, 0.7, h0, adj=0)
text(0.1, 0.3, h1, adj=0)
```

There's a procedure for this, Dunnett's test. It involves only three comparisons if you have four groups total, and only nine comparisons if you have ten groups total. You can quickly identify who is better than the control. This also gives you simplicity and a bit of extra power and precision.

Dunnett's test is easy to implement. Calculate the traditional measures in analysis of variance, including mean squared error (MSE). You can skip the traditional F Test that you see in most ANOVA tables. It won't hurt your overall alpha level, as long as you only look at each treatment versus the control. Simply compare each treatment mean minus the control mean to a cutoff value

Reject $H_0$ if $|\bar X_i-\bar X_1| \geq d(\alpha, g, n) \sqrt{2MSE/n}$

where d is obtained from a table such as [this one][bobb1] on the [Statology blog][bobb0]. The value of d depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group). Note that some tables define g as the number of groups excluding the control group.

Most statistical software packages will include Dunnett's test as an option for ANOVA. Just remember that you don't need to use the initial F test as a screen before jumping into the Dunnett's test.

[bobb1]: https://www.statology.org/dunnetts-table/

[bobb0]: https://www.statology.org/about/

### Caveats

Modifications of Dunnett's test are tricky when the group sizes are unequal.

Now you lose something when you simplify the hypothesis. Suppose you have six groups, a control and five different treatments. Now imagine that all of the treatments are significantly better than the control group. Jackpot! Every treatment is worth further study. But Dunnett's test won't allow you to see if some of the treatments are better than the others. There is no option for finding the best of the best.

Let's consider as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.

### 2022-08-25 text/22/dunnetts-test.Rmd
Added example to dunnetts-test.Rmd

@@
 
-Reject $H_0$ if $|\bar X_i-\bar X_1| \geq d(\alpha, g, n) \sqrt{2MSE/n}$
+Reject $H_0$ if $\bar X_i \leq \bar X_1 - d(\alpha, g, n) \sqrt{2MSE/n}$ or $\bar X_i \geq \bar X_1 + d(\alpha, g, n) \sqrt{2MSE/n}$
 
@@
 [bobb0]: https://www.statology.org/about/
+
+### Example-seed viability
+
+In an experiment examining the effect of different disinfection protocols on orthodontic elastormeric ligatures, 120 ligatures were divided into 6 groups of 20.
+Each group received a different disinfection protocol, with the exception of the first group, which received no disinfection.
+This group served as a control group.
+
+![](http://www.pmean.com/new-images/22/dunnetts-test-01.jpg)
+
+Leandro Berni Osorio, Luiz Makito Osawa Gutierrez, Eduardo Martinelli de Lima, Eduardo Gonçalves Mota, and Luciane Macedo de Menezes.
+Disinfection of Orthodontic Elastomers and Its Effects on Tensile Strength.
+Turkish Journal of Orthodontics, 2021, 35(1): 22-26.
+
+The researchers were interested in whether any of the disinfection protocols reduced the tensile strength of the ligatures. Here are the summary statistics for the maximum strength measurement.
+
+```{r, echo=FALSE, comment=""}
+mn1 <- c(21.39, 22.58, 16.94, 21.93, 19.93, 22.00)
+sd1 <- c( 3.35,  2.88,  3.50,  3.05,  1.86, 3.71)
+data.frame(mean=mn1, standard_deviation=sd1)
+```
+
+```{r, echo=FALSE, comment=""}
+suppressMessages(suppressWarnings(library(tidyverse)))
+mse <- mean(sd1^2)
+d <- 2.73
+n <- 20
+lo <- mn1[1] - d*sqrt(2*mse/20)
+hi <- mn1[1] + d*sqrt(2*mse/20)
+```
+
+lower limit = $\bar X_1$ - d(0.05, 6, 20) $\sqrt{2MSE/n}$
+
+= `r mn1[1]` - `r d` $\sqrt{2 (`r round(mse, 2)`)/ `r n`}$
+
+= `r round(lo, 2)`
+
+upper limit = $\bar X_1$ + d(0.05, 6, 20) $\sqrt{2MSE/n}$
+
+= `r mn1[1]` + `r d` $\sqrt{2 (`r round(mse, 2)`)/ `r n`}$
+
+= `r round(hi, 2)`
+
+You can create a simple graph from these limits.
+
+```{r, echo=FALSE, comment=""}
+data.frame(g=2:6, mn=mn1[-1]) %>%
+  ggplot(aes(g, mn)) +
+    geom_point() +
+    geom_segment(
+      x=2:6, 
+      xend=2:6, 
+      y=mn1[-1], 
+      yend=mn1[1]) +
+    geom_segment(x=1.9, y=lo, xend=6.1, yend=lo) +
+    geom_segment(x=1.9, y=hi, xend=6.1, yend=hi) +
+    geom_segment(x=1.9, y=mn1[1], xend=6.1, yend=mn1[1], lty="dotted") +
+    expand_limits(y=c(lo, hi)) +
+    expand_limits(x=c(1.5, 6.5))
+```
+
+Only the third disinfection protocol has significantly less strength compared to the control group with no disinfection.
 

### 2022-08-29 text/22/analysis-of-means.Rmd
Simplified presentation of analysis of means

@@
 
-```{r, fig.width=5, fig.height=1, echo=FALSE}
+```{r, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
-h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[j] ~ " for at least one i,j")
 par(mar=rep(0.1, 4))
 plot(0:1, 0:1, axes=FALSE, type="n")
-text(0.1, 0.7, h0, adj=0)
-text(0.1, 0.3, h1, adj=0)
+text(0.1, 0.5, h0, adj=0)
 ```
@@
 
-```{r, fig.width=5, fig.height=1, echo=FALSE}
+```{r, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
-h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu ~ " for at least one i")
 par(mar=rep(0.1, 4))
 plot(0:1, 0:1, axes=FALSE, type="n")
-text(0.1, 0.7, h0, adj=0)
-text(0.1, 0.3, h1, adj=0)
+text(0.1, 0.5, h0, adj=0)
 ```
@@
 
+Desiree M.
+Hautea DM et al.
+Field Performance of Bt Eggplants (Solanum melongena L.) in the Philippines: Cry1Ac Expression and Control of the Eggplant Fruit and Shoot Borer (Leucinodes orbonalis Guenée).
+PLoS One.
+2016; 11(6): e0157498.
+
+Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant.
+Four plants were assessed for each strain.
+Here are the descriptive statistics.
+The research question is whether all strains have a comparable amount of CryA1c in their leaves.
+
+The raw data was not available, but you can calculate the analysis of means limits using the group means and standard deviations.
+
+```{r}
+mn <- c(
+  24.87,
+  24.54,
+  23.61,
+  18.32,
+  22.72,
+  22.81)
+
+sd <- c(
+  0.56,
+  1.14,
+  1.33,
+  2.45,
+  0.19,
+  1.18)
+mse <- mean(sd^2)
+mse
+ybarbar <- mean(mn)
+t <- 5
+N <- 20
+h <- 2.88
+hi <- ybarbar+h*sqrt((t-1)*mse/N)
+lo <- ybarbar-h*sqrt((t-1)*mse/N)
+lo
+hi
+```
+
 https://www.frontiersin.org/articles/10.3389/fchem.2022.894547/full

### 2022-08-29 text/22/dunnetts-test.Rmd
Simplified presentation of analysis of means

--- 2022-08-23 text/22/dunnetts-test.Rmd
@@
 
-I'm a big fan of Analysis of Variance (ANOVA).
-I use it all the time.
-I learn a lot from it.
-But sometimes I want something a bit different.
-And the difference comes down to how you specify the hypothesis.
+I'm a big fan of Analysis of Variance (ANOVA). I use it all the time. I learn a lot from it. But sometimes it doesn't test the hypothesis I need.
 
@@
 
-Recall the way in which the hypothesis test is commonly set up when you are comparing three of more groups using analysis of variable.
+Recall the hypothesis test for analysis of variance.
 
-```{r, fig.width=5, fig.height=1, echo=FALSE}
+```{r, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
-h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[j] ~ " for at least one i,j")
 par(mar=rep(0.1, 4))
 plot(0:1, 0:1, axes=FALSE, type="n")
-text(0.1, 0.7, h0, adj=0)
-text(0.1, 0.3, h1, adj=0)
+text(0.1, 0.5, h0, adj=0)
 ```
 
-This is a very common approach in a comparison of k independent groups.
-Test for any deviation from the null hypothesis using an F test.
-If this is statistically significant, then use a Tukey follow-up test to see which pairs of means differ from one another.
-
-This hypothesis specifies equality by specifying that every pair of means is equal.
+The null hypothesis specifies that every pair of means is equal.
 For four groups, this implies six equalities: 1=2, 1=3, 1=4, 2=3, 2=4, and 3=4.
@@
 
-It gets messy very fast. That's okay. If you have a lot of groups that you are comparing, you have to make it has to involve a lot of comparisons. Or do you?
+It gets messy very fast. That's okay. If you have a lot of groups that you are comparing, you have to make a lot of comparisons. Or do you?
 
@@
 
-If one of the groups is a control or placebo, then you might consider an alternative formulation. Here's what the hypothesis looks like, assuming that the control is group #1.
+If one of your groups is a control or placebo, AND you want only to compare each treatment to the control, this is  what the hypothesis looks like, assuming that the control is group #1.
 
-```{r, fig.width=5, fig.height=1, echo=FALSE}
+```{r, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[1] ~ " for all i=2,...,k")
-h1 <- bquote(H[1] ~ ": " ~ mu[i] != mu[1] ~ " for at least one i")
 par(mar=rep(0.1, 4))
 plot(0:1, 0:1, axes=FALSE, type="n")
-text(0.1, 0.7, h0, adj=0)
-text(0.1, 0.3, h1, adj=0)
+text(0.1, 0.5, h0, adj=0)
 ```
@@
 It involves only three comparisons if you have four groups total, and only nine comparisons if you have ten groups total.
-You can quickly identify who is better than the control.
+You can quickly identify which treatment or treatments is different than the control.
 This also gives you simplicity and a bit of extra power and precision.
@@
 It won't hurt your overall alpha level, as long as you only look at each treatment versus the control.
-Simply compare each treatment mean minus the control mean to a cutoff value
-
-Reject $H_0$ if $|\bar X_i-\bar X_1| \geq d(\alpha, g, n) \sqrt{2MSE/n}$
-
-where d is obtained from a table such as [this one][bobb1] on the [Statology blog][bobb0].
+Simply compare each treatment mean minus the control mean to a cutoff value using percentiles from a special table for Dunnett's test.
 The value of d depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group).
@@
 
+### Example-seed viability
+
+Let's look at an example in an experiment examining the effect of different disinfection protocols on a piece of dental equipment, orthodontic elastormeric ligatures (think braces).
+In this experiment, 120 ligatures were divided into 6 groups of 20.
+Each group received a different disinfection protocol, with the exception of the first group, which received no disinfection.
+This group served as a control group.
+
+![](http://www.pmean.com/new-images/22/dunnetts-test-01.jpg)
+
+Osorio LB et al. Disinfection of Orthodontic Elastomers and Its Effects on Tensile Strength. Turkish Journal of Orthodontics, 2021, 35(1): 22-26.
+
+Disinfection is important, but does it reduce the tensile strength of the ligatures. Here are the summary statistics for the maximum strength measurement.
+
+```{r, echo=FALSE, comment=""}
+mn1 <- c(21.39, 22.58, 16.94, 21.93, 19.93, 22.00)
+sd1 <- c( 3.35,  2.88,  3.50,  3.05,  1.86, 3.71)
+data.frame(mean=mn1, standard_deviation=sd1)
+```
+
+```{r, echo=FALSE, comment=""}
+suppressMessages(suppressWarnings(library(tidyverse)))
+mse <- mean(sd1^2)
+d <- 2.73
+n <- 20
+lo <- mn1[1] - d*sqrt(2*mse/20)
+hi <- mn1[1] + d*sqrt(2*mse/20)
+```
+
+The paper does not include the raw data, but you can still calculate all the key statistics needed to run Dunnett's test.
+You can create a simple graph that identifies if any of the disinfection protocols shows significantly less strength than the control.
+
+```{r, echo=FALSE, comment=""}
+data.frame(g=2:6, mn=mn1[-1]) %>%
+  ggplot(aes(g, mn)) +
+    geom_point() +
+    geom_segment(
+      x=2:6, 
+      xend=2:6, 
+      y=mn1[-1], 
+      yend=mn1[1]) +
+    geom_segment(x=1.9, y=lo, xend=6.1, yend=lo) +
+    geom_segment(x=1.9, y=hi, xend=6.1, yend=hi) +
+    geom_segment(x=1.9, y=mn1[1], xend=6.1, yend=mn1[1], lty="dotted") +
+    expand_limits(y=c(16, 26)) +
+    expand_limits(x=c(0.9, 6.1)) +
+    xlab("Disinfection protocol") +
+    ylab("Tensile strength") +
+    scale_x_continuous(breaks=1:6, minor_breaks=NULL) +
+    scale_y_continuous(breaks=2*(8:13))
+```
+
+Only the third disinfection protocol has significantly less strength compared to the control group with no disinfection.
+This graph shows a two-sided comparison, but you may prefer a one-sided comparison.
+This is a very easy modification.
+
 ### Caveats
 
-Modifications of Dunnett's test are tricky when the group sizes are unequal.
+Things get a bit dicey when the group sizes are unequal. The formulas get a bit messier and the intervals are approximate. This is generally true for most approaches to handling multi-group data.
 
-Now you lose something when you simplify the hypothesis.
+Keep in mind that you lose something when you simplify the hypothesis.
 Suppose you have six groups, a control and five different treatments.
@@
 
-Let's consider as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.
+Let's consider this as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.

### 2022-08-29 text/22/analysis-of-means.Rmd
Good first draft of analysis-of-means.Rmd

@@
 
-### Comparison to an overall mean
+### The comparison to the overall mean hypothesis
 
-Suppose you are in a setting like the one described above, where there is an expectation of similarity from one r goal is to establish whether any groups differ from the overall mean, Then you would write your hypothesis as
-
+Suppose you are in a setting like the one described above, where there is an expectation of similarity between each group mean and the overall mean. Then you would write your hypothesis as 
 ```{r, fig.width=5, fig.height=0.33, echo=FALSE}
@@
 This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results.
-If any do not, then you want to identify the group or groups that deviate from the norm and study them further.
+If any do not, then you want to identify the group or groups that deviate from the norm and study factors that may account for the deviation.
+
+The analysis of means approach that compares each group mean to the overall mean is easy to implement and it lends itself to a simple graphical display.
+
+You don't need to compute the traditional F statistic for Analysis of Variance first, because the Analysis of Means approach controls the overall Type I error rate.
+This protects you from the accusation of p-hacking, even if the number of groups is very large.
 
@@
 
+Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant.
+Four plants were assessed for each strain.
+Here are the descriptive statistics.
+The research question is whether all strains have a comparable amount of CryA1c in their leaves.
+
+![](http://www.pmean.com/new-images/22/analysis-of-means-00.png)
+
+(Image taken from Wikipedia)
+
 Desiree M.
@@
 
-Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant.
-Four plants were assessed for each strain.
-Here are the descriptive statistics.
-The research question is whether all strains have a comparable amount of CryA1c in their leaves.
+https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498
 
-The raw data was not available, but you can calculate the analysis of means limits using the group means and standard deviations.
+Here are the summary statistics.
 
-```{r}
+```{r, echo=FALSE}
 mn <- c(
-  24.87,
-  24.54,
-  23.61,
-  18.32,
-  22.72,
-  22.81)
+  21.83,
+  20.40,
+  21.62,
+  21.06,
+  20.57)
 
 sd <- c(
-  0.56,
-  1.14,
-  1.33,
-  2.45,
-  0.19,
-  1.18)
+  1.17,
+  1.09,
+  1.32,
+  1.11,
+  0.99)
+
 mse <- mean(sd^2)
-mse
 ybarbar <- mean(mn)
@@
 lo <- ybarbar-h*sqrt((t-1)*mse/N)
-lo
-hi
 ```
 
-https://www.frontiersin.org/articles/10.3389/fchem.2022.894547/full
+![](http://www.pmean.com/new-images/22/analysis-of-means-01.png)
+
+The raw data was not available, but you can calculate the analysis of means limits using the group means and standard deviations.
+You can plot the individual means versus the limits provided by analysis of means calculations.
+
+![](http://www.pmean.com/new-images/22/analysis-of-means-02.png)
+
+Notice that all five means lie inside the limits. None of the five strains shows a  statisticially significant difference from the overall mean.
+
+### Caveats
+
+The formulas get a bit more complex and the graph gets a bit messier if the sample size differs from one group to another.
+
+More importantly, the only comparisons that you can make with the analysis of means approach is a comparison to the overall mean.
+If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control over the overall Type I error rate.
+
+The Analysis of Means, just like the Dunnett's test described in an earlier blog post, provides a simple approach to testing a different type of hypothesis.
+Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.

### 2022-08-30 text/22/dunnetts-test.Rmd
Good first draft of analysis-of-means.Rmd
Minor fix to dunnetts-test

@@
 Simply compare each treatment mean minus the control mean to a cutoff value using percentiles from a special table for Dunnett's test.
-The value of d depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group).
+The critical value for Dunnett's test depends on alpha (the desired Type I error rate for a two-sided test), g (the number of groups including the control group) and n (the number of data points in each group).
 Note that some tables define g as the number of groups excluding the control group.

### 2022-09-06 text/22/analysis-of-means.Rmd
Updated some of the 2017 files

@@
 output:
-  word_document: default
   html_document: default
@@
 
+### Comparison to an overall mean
+
 You are in a setting where you are expecting to see a degree of uniformity among several groups.
 For example, you have a group of clinics who all treat the same type of patients.
-The patients themselves may differ markedly from one to another, but they are assigned in a more or less random way to each clinic.
-So, although one patient may differ markedly from another, the average result should be the same for across all the clinics.
+The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient.
+But because of how the patients are assigned to a particular clinic (more or less randomly), these should balance out on average.
 
-The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient.
-But because of how the patients are assigned to a particular clinic, these should balance out on average.
 If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment.
@@
 Suppose you are in a setting like the one described above, where there is an expectation of similarity between each group mean and the overall mean. Then you would write your hypothesis as
+
 ```{r, fig.width=5, fig.height=0.33, echo=FALSE}
@@
 
-The analysis of means approach that compares each group mean to the overall mean is easy to implement and it lends itself to a simple graphical display.
+The analysis of means approach compares each group mean to the overall mean.
+It is easy to implement and it lends itself to a simple graphical display.
+You need a table of critical values, which depend on alpha (the overall Type I error rate), t (the number of groups), and n (the number of observations within each group).
+Some tables use the degrees of freedom for error in place of n.
 
@@
 Your goal is not to use statistics to hunt out different groups to reward or punish them.
-You are using statistics to help in understanding why deviations from the norm occur.
+You are using statistics to help in understanding if deviations from the norm occur and then study those deviating groups to understand why they deviate.
 
@@
 Here are the descriptive statistics.
-The research question is whether all strains have a comparable amount of CryA1c in their leaves.
+The research question is whether any strain differs from the overall mean level of CryA1c.
 
@@
 
+You may prefer a one-sided test in this setting, such as testing whether any strain is deficient in the CryA1c levels. You can make a very easy modification to get one-sided tests.
+
 ### Caveats
@@
 
+You also need to specify the Analysis of Means hypothesis prior to looking at your data. Peeking at the data and then choosing your hypothesis is cheating.
+
 The Analysis of Means, just like the Dunnett's test described in an earlier blog post, provides a simple approach to testing a different type of hypothesis.
 Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.
+
+### Details about the calculations
+
+To get the limits for the graph shown above, you need the Analysis of Means table of critical values plus an estimate of Mean Squared Error. The general formula (as shown in [this article][jay1]) is
+
+$\ $ $\bar Y_{..} \pm h(\alpha, t, N) \sqrt{(t-1)MSE/N}$
+
+where $\bar Y_{..}$ is the overall mean and $h(\alpha, t, N)$ comes from a table such as [this one][lev1] (see page 3) or [this one][sim3].
+
+In our example alpha=0.05, t=5, and N=20. The original article does not provide MSE, but you can calculate it as the pooled variance using the descriptive statistics shown above. The limits are
+
+```{r, echo=FALSE}
+# Make the numbers look nice
+ybarbar <- round(ybarbar, 2)
+mse <- round(mse, 2)
+lo <- round(lo, 2)
+hi <- round(hi, 2)
+```
+
+$\ $ $`r ybarbar` \pm 2.88 \sqrt{(`r t`-1)`r mse`/`r N`}$
+
+or
+
+$\ $ $`r lo`$ and $`r hi`$
+
+[jay1]: https://digitalcommons.pvamu.edu/cgi/viewcontent.cgi?article=1122&context=aam
+
+[lev1]: https://wps.prenhall.com/wps/media/objects/9431/9657451/Ch_11/levine-smume6_topic_ANOM.pdf
+
+[sim3]: http://www.pmean.com/07/AnomTable05Part1.html

### 2022-09-06 text/22/dunnetts-test.Rmd
Updated some of the 2017 files

@@
 Calculate the traditional measures in analysis of variance, including mean squared error (MSE).
-You can skip the traditional F Test that you see in most ANOVA tables.
+You can skip the F Test that you see in most ANOVA tables.
 It won't hurt your overall alpha level, as long as you only look at each treatment versus the control.
@@
 
-Most statistical software packages will include Dunnett's test as an option for ANOVA. Just remember that you don't need to use the initial F test as a screen before jumping into the Dunnett's test.
+Most statistical software packages will include Dunnett's test as an option for ANOVA. Just remember that you don't need to use the initial F test as a screen before jumping into Dunnett's test.
 
@@
 
+(Image taken from G&H Orthodontics)
+
 Osorio LB et al. Disinfection of Orthodontic Elastomers and Its Effects on Tensile Strength. Turkish Journal of Orthodontics, 2021, 35(1): 22-26.
 
-Disinfection is important, but does it reduce the tensile strength of the ligatures. Here are the summary statistics for the maximum strength measurement.
+Disinfection is important, but does it reduce the tensile strength of the ligatures? Here are the summary statistics for the maximum strength measurement.
 
@@
 The paper does not include the raw data, but you can still calculate all the key statistics needed to run Dunnett's test.
-You can create a simple graph that identifies if any of the disinfection protocols shows significantly less strength than the control.
+You can then create a simple graph that identifies if any of the disinfection protocols shows significantly less strength than the control.
 
@@
 
+The dotted line is the mean for the control group.
+Any mean falling outside the solid lines is statistically significantly different from the control mean.
+Notice that there is no comparison at 1 because you don't compare the control group to itself.
+
 Only the third disinfection protocol has significantly less strength compared to the control group with no disinfection.
@@
 
-Let's consider this as a reminder that there is no such thing as a free lunch. There are always trade-offs. No approach is superior in all settings.
+Let's consider this as a reminder that there is no such thing as a free lunch.
+There are always trade-offs.
+No approach is superior in all settings.
+If, however, you are truly only interested in comparing each treatment to the control and you specify this prior to looking at the data, then Dunnett's test is a very attractive option.
+
+### Details about the calculations
+
+The limits require a critical value from a table for Dunnett's test, the mean of the control group and Mean Squared Error.
+
+$\ $ $\bar X_1 \pm d(\alpha, t, n) \sqrt{2 MSE/n}$
+
+where $\alpha$ is the Type I error rate, t in the number of groups, and n is the number of observations in each group.
+The value of d can be found in various tables, such as [this one][bob1] or [this one][win1].
+While the paper that I drew the example from does not include the raw data, you can calculate Mean Squared Error from the summary statistics.
+The limits in this example are
+
+```{r echo=FALSE}
+# Make the numbers look nice.
+mse <- round(mse, 2)
+lo <- round(lo, 2)
+hi <- round(hi, 2)
+```
+
+$\ $ $`r mn1[1]` \pm 2.73 \sqrt{2 (`r mse`)/`r n`}$
+
+$\ $ $`r lo`$ and $`r hi`$
+
+[bob1]: https://www.statology.org/dunnetts-table/
+[win1]: https://users.stat.ufl.edu/~winner/tables/dunnett-2side.pdf

### 2022-10-18 src/clean_files.R
Minor edits to clean_files.R

@@ -11,7 +11,7 @@ suppressMessages(suppressWarnings(library(rmarkdown)))
 suppressMessages(suppressWarnings(library(stringr  )))
 suppressMessages(suppressWarnings(library(tidyverse)))
 
-source(file="src/standard_functions.R")
+# source(file="src/standard_functions.R")
 
 clean_files <- function(
     old, 
@@ -20,24 +20,26 @@ clean_files <- function(
     subdir_list=c("99", zpad(0:22)),
     file_pattern="*\\.md$") {
   if (!exists("ok_to_replace")) ok_to_replace <- FALSE 
-  k <- 0
+  k1 <- 0
+  k2 <- 0
   for (subdir in subdir_list) {
     print(subdir)
     glue("{dir_root}/{subdir}") %>%
       list.files %>%
       str_subset(file_pattern) -> file_list
-    glue{
+    glue(
       "\nSearching through ",
       "{length(file_list)} ",
       "files in {dir_root}/",
-      "{subdir}\n\n" -> message
+      "{subdir}\n\n") -> message
     if (verbose) cat(message)
     for (i_file in file_list) {
+      k1 <- k1+1
       fn <- glue("{dir_root}/{subdir}/{i_file}")
       tx <- read_lines(fn)
       tf <- str_subset(tx, old)
-      if (length(found_lines)==0) next
-      k <- k+1
+      if (length(tf)==0) next
+      k2 <- k2+1
       cat(glue("\n\n{fn}\n\n"))
       if (new != "Not yet") {
         ty <- str_replace_all(tf, old, new)
@@ -49,13 +51,13 @@ clean_files <- function(
           str_replace_all(old, new) %>%
           write_lines(fn)
       }
-      }
     }
   }
   glue(
     "\n\nTotal files ",
-    "{ifelse(ok_to_replace, "", "to be ")}",
-    "replaced = {k}"}) %>% cat
+    "{ifelse(ok_to_replace, '', 'to be ')}",
+    "replaced = {k2} out of ",
+    "{k1} files searched.") %>% cat
 }
 
 verbose <- TRUE
@@ -63,7 +65,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'xcvgit',
-  'dissertation defense',
+  '\\[sim1\\]:',
   dir_root="text", 
+  subdir_list="05",
   file_pattern="md$")

### 2022-10-20 text/17/first-jags.Rmd
Updated 17/consulting-trap.md
Standardized tags

@@
 - Bayesian statistics
-- Incomplete pages
+- Incomplete page
 output: html_document
@@
 
-Make sure that you have [jags installed][jags1] on your computer as well as the [rjags package][jags2]
+Make sure that you have [jags installed][jags1] on your computer as well as the [rjags package][jags2].
 

### 2022-10-20 src/clean_files.R

@@ -65,7 +65,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '\\[sim1\\]:',
+  '- Incomplete pages',
+  '- Incomplete page',
   dir_root="text", 
-  subdir_list="05",
   file_pattern="md$")

### 2022-10-20 src/clean_files.R
Updated 05/BaselineAdjustment

@@ -65,7 +65,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  '- Incomplete pages',
-  '- Incomplete page',
+  '- Quotations',
+  '- Quotation',
   dir_root="text", 
   file_pattern="md$")

### 2022-10-25 text/05/MergingInR.Rmd
Removed backslashes
Added "Being updated" tags to 2005 files

@@
 tags:
+- Being updated
 - R software
@@
 allow me to backtrack and find information in the original files if
-needed. Here\'s a simple example of how this would work.
+needed. Here's a simple example of how this would work.
 

### 2022-10-25 src/archive/step0-global-change.R

@@ -1,20 +1,65 @@
-# step0-skim-files.R
+# step0-global-change.R
 # written by Steve Simon
-# created 2020-01-15
+# created 2020-02-05
 
-# This program finds all the bibtex files
-# (extension .bib) in r1 and converts the
-# names.
-# Step 0-1. Preliminaries
+# This program finds files and
+# makes changes from one string
+# to another.
 
 source(file="src/prelims.R")
-if (!exists("verbose")) verbose <- TRUE
-if (!exists("update_all")) update_all <- TRUE
+if (!exists("v")) v <- TRUE
 
-skim_bib_files("name")
-skim_bib_files("name", "../source/r1")
-skim_yaml_files("category")
+file_list <- build_yr_list("text", "*.md", v=FALSE)
 
-# Save everything.
+if (v) "\nProcessing" %b% length(file_list) %b% "files.\n\n" %>% cat
 
-save.image("data/step0.RData")
+process_changes <- function(i_file, s0, s1, dry_run=TRUE) {
+  tx <- readLines(i_file)
+  tx %>% 
+    str_c(collapse="\n") %>%
+    str_replace_all(s0, s1) %>%
+    str_split("\n") %>%
+    unlist -> new_tx
+  n_changed <- which(tx != new_tx)
+  if (length(n_changed)==0) return(tx)
+  i_file    %>% br    %>% cat
+  for (i in n_changed) {
+    tx[i]     %>% br    %>% cat
+    new_tx[i] %>% br(2) %>% cat
+  }
+  if (!dry_run) writeLines(new_tx, i_file)
+  return(new_tx)  
+}
+
+for (i_file in file_list) {
+  d <- FALSE
+  new_tx <- process_changes(i_file, fixed('\n> '), "\n", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('[\\['), "[", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed(']\\]'), "]", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\['), "[", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\]'), "]", dry_run=d)
+  new_tx <- process_changes(i_file, fixed('\\*'), "*", dry_run=d) 
+  new_tx <- process_changes(i_file, fixed('\\"'), '"', dry_run=d) 
[…]

### 2022-10-25 src/step0-global-change.R

@@ -1,65 +1,0 @@
-# step0-global-change.R
-# written by Steve Simon
-# created 2020-02-05
-
-# This program finds files and
-# makes changes from one string
-# to another.
-
-source(file="src/prelims.R")
-if (!exists("v")) v <- TRUE
-
-file_list <- build_yr_list("text", "*.md", v=FALSE)
-
-if (v) "\nProcessing" %b% length(file_list) %b% "files.\n\n" %>% cat
-
-process_changes <- function(i_file, s0, s1, dry_run=TRUE) {
-  tx <- readLines(i_file)
-  tx %>% 
-    str_c(collapse="\n") %>%
-    str_replace_all(s0, s1) %>%
-    str_split("\n") %>%
-    unlist -> new_tx
-  n_changed <- which(tx != new_tx)
-  if (length(n_changed)==0) return(tx)
-  i_file    %>% br    %>% cat
-  for (i in n_changed) {
-    tx[i]     %>% br    %>% cat
-    new_tx[i] %>% br(2) %>% cat
-  }
-  if (!dry_run) writeLines(new_tx, i_file)
-  return(new_tx)  
-}
-
-for (i_file in file_list) {
-  d <- FALSE
-  new_tx <- process_changes(i_file, fixed('\n> '), "\n", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('[\\['), "[", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed(']\\]'), "]", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('\\['), "[", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('\\]'), "]", dry_run=d)
-  new_tx <- process_changes(i_file, fixed('\\*'), "*", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('\\"'), '"', dry_run=d) 
-  new_tx <- process_changes(i_file, fixed("\\'"), "'", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed(' '), ' ', dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('Â'), ' ', dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('\\$'), "$", dry_run=d) 
-  new_tx <- process_changes(i_file, fixed('\\@'), "@", dry_run=d) 
[…]

### 2022-11-03 text/21/quasi-likelihood.Rmd
Fixed titles in 2022 webpages

@@
 But you also have the option of working outside of these distributions.
-You can fit models that do not specify the distribtution, but which fit a particular transformation of the linear effect and particular mean-variance relationship.
+You can fit models that do not specify the distribution, but which fit a particular transformation of the linear effect and particular mean-variance relationship.
 

### 2022-11-05 src/clean_files.R
Added sas-regular-expressions

@@ -11,7 +11,16 @@ suppressMessages(suppressWarnings(library(rmarkdown)))
 suppressMessages(suppressWarnings(library(stringr  )))
 suppressMessages(suppressWarnings(library(tidyverse)))
 
-# source(file="src/standard_functions.R")
+source(file="src/standard_functions.R")
+
+# I had some trouble with write_lines until
+# I used the option "lazy=FALSE" in the
+# previous read_lines function. See
+# https://community.rstudio.com/t/error-cannot-
+#   open-file-for-writing-write-csv/111125
+# Also note readr 2.1.0 comments at
+# https://cran.r-project.org/web/packages/
+#    readr/news/news.html
 
 clean_files <- function(
     old, 
@@ -23,7 +32,7 @@ clean_files <- function(
   k1 <- 0
   k2 <- 0
   for (subdir in subdir_list) {
-    print(subdir)
+    cat(glue("\n\n{subdir}"))
     glue("{dir_root}/{subdir}") %>%
       list.files %>%
       str_subset(file_pattern) -> file_list
@@ -36,7 +45,7 @@ clean_files <- function(
     for (i_file in file_list) {
       k1 <- k1+1
       fn <- glue("{dir_root}/{subdir}/{i_file}")
-      tx <- read_lines(fn)
+      tx <- read_lines(fn, lazy=FALSE)
       tf <- str_subset(tx, old)
       if (length(tf)==0) next
       k2 <- k2+1
@@ -47,7 +56,6 @@ clean_files <- function(
       }
       cat(paste0(tf, collapse="\n\n"))
       if (ok_to_replace) {
-        print(fn)
         tx %>%
           str_replace_all(old, new) %>%
           write_lines(fn)
@@ -66,8 +74,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'tags:',
-  'tags:\n- Being updated',
+  '- Research designs',
+  '- Research design',
   dir_root="text", 
-  subdir_list="01",
   file_pattern="md$")

### 2022-11-13 text/22/posix-time.Rmd
Added posix-time

---
title: "The POSIX time standard"
author: "Steve Simon"
date: "2022-11-12"
output:
  html_document: default
category: Blog post
tags: Data science
source: new
---

I thought I understood everything about how computers represent time, but apparently not. There is an important time standard, POSIX time, that I was not all that aware of. Here's what I found out after a few Google searches.

### POSIX time, Unix time, and UTC

Wikipedia has a nice article about [Unix time][wik1], and there are apparently some subtle differences between the POSIX standard for time and Unix time, the standard approach for time used in Unix systems. As I understand it, the differences are fairly minor. Both differ slightly from another standard, [Coordinated Universal Time (UTC)][wik2].

[wik1]: https://en.wikipedia.org/wiki/Unix_time
[wik2]: https://en.wikipedia.org/wiki/Coordinated_Universal_Time

There are some issues about how these standards handle [leap seconds][wik3]. Leap seconds have been added to various years since 1972 to account for irregularities and the gradual slowing in the earth's rotation. This may be important in certain sciences like Astronomy, but most statisticians do not track time values to that level of precision.

[wik3]: https://en.wikipedia.org/wiki/Leap_second

### Internal storage of POSIX time values

The POSIX standard stores date time values internally as the number of seconds since midnight on January 1, 1970 (1970-01-01 00:00:00). Midnight is considered the start of a day rather than the end of it, so the date-time values transition from

+ -1 = 1969-12-31 11:59:59 to 
+ 0 = 1970-01-01 00:00:00

For Unix systems, the POSIX time value is stored as either a 32 bit signed integer or a 64 bit signed integer. If you do the math, the 32 bit signed integer will reach its upper limit sometime in 2038, which could lead to a miniature revival of the Y2K crisis.

The POSIX standard allows you to specify a time zone, which helps if you are coordinating data collection across multiple time zones and if synchronizing these values at levels of time less than a day are important to you.

### Functions in R for handling POSIX times

R uses its own standard for storing date/time values. It counts the number of days rather than the number of seconds since 1970-01-01 and this means that intervals of time less than one day are stored behind the decimal place. So noon on January 1, 1970 (1970-01-01 12:00:00) would be stored as 0.5, exactly half a day since midnight.

There are several functions that allow you to convert among these standards.

The function as.Date (part of base R) will convert either a string or a POSIX formatted date. You can also convert a numeric value to a date, but you need to specify the origin or index date.

The functions as.POSIXct and as.POSIXlt (also part of base R) will convert strings or an R formatted date into the POSIX standard. There are important differences in these functions, but only from the perspective of internal storage.

### Some examples in R

When you print dates in R, the way the R displays them hides the internal format. So dates stored as strings, dates stored in the R internal format and dates stored in the POSIX standard all look the same from the outside.

```{r}
date_string <- "1970-01-01"
date_internal_r_format <- as.Date(date_string)
date_posix_ct <- as.POSIXct(date_string)
date_posix_lt <- as.POSIXlt(date_string)

date_string
date_internal_r_format
date_posix_ct
date_posix_lt
```

The only apparent difference is that the POSIX values include a time zone. On my system and on the date this post was first written, the time zone is CST, but it might be different when you run this code on your system.

Use the print.default function to show how these files are stored internally.

Strings have no special features.

```{r}
print.default(date_string)
```

The R internal format is a number with a class attribute of Date.

```{r}
print.default(date_internal_r_format)
```

The POSIX ct format is a number with a class attribute of POSIXt and POSIXct and an additional attribute of time zone.

```{r}
print.default(date_posix_ct)
```

Note that the numeric value here is not zero, even I specified only the date and not the time. On the date this post was first run and in the time zone that my computer lives in, the value is 21600. There are 86,400 seconds in a day, so 21,600 is one quarter of this. So when I only specify the date, it adjusts for the time zone (CST) which is 6 hours behind Greenwich Mean Time (GMT).

The POSIX lt format is list with class attributes of POSIXt and POSIXlt. It is a list and not a number. The list has the individual components of a date/time starting at the second unit and continuing through to the year.

```{r}
print.default(date_posix_lt)
```

Notice that the as.POSIXlt function does not adjust for the difference in time zones.

If you care about this level of detail, both the as.POSIXct and as.POSIXlt functions include a time zone argument (tz).

If you want to convert numbers to date formats, you need to specify an origin argument in your code. Note that 1 day equals 86,400 seconds.

```{r}
date_internal_r_format <- as.Date(1, origin="1970-01-01")
date_posix_ct <- as.POSIXct(86400, origin="1970-01-01")
date_posix_lt <- as.POSIXlt(86400, origin="1970-01-01")

date_internal_r_format
date_posix_ct
date_posix_lt
```

Notice the adjustment (for me 6 hours) made by as.POSIXct and as.POSIXlt for the differential in time zone between CST and GMT. Also notice that this time the adjustment is in the opposite direction.

If you think about these adjustments, they are logical, but you should probably specify the timezone argument if you want to control these adjustments.

```{r}
as.POSIXct(86400, origin="1970-01-01", tz="GMT")
as.POSIXlt(86400, origin="1970-01-01", tz="America/Chicago")
```

### Choosing between POSIX and the R internal standard

So how should you store time values? Should you use the internal R standard or the POSIX standard?

Most of the time, I would recommend using the R internal standard. It is simple and is less likely to cause problems with libraries that might be unaware of the POSIX standard and end up misinterpreting dates stored in that format.

You should use the POSIX standard for certain situations

+ If your data arrives in POSIX format and it comes in several waves,
+ If you need to export data to a system that uses the POSIX format, or
+ If you need to track different time zones.

### But wait, there's more!

There are many nice functions in the lubridate package that does as good a job if not better than the functions available in base R. That is a topic for another blog post. The lubridate [vignette][lub1] or [cheat sheet][lub2] or the [tidyverse overview][lub3] are nice introductions to many of the helpful features of this package.

There is another package in R that is commonly cited, [chron][chr1]. I have not used chron and cannot comment intelligently on its features.

[chr1]: https://cran.r-project.org/package=chron

[lub1]: https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html
[lub2]: https://github.com/rstudio/cheatsheets/blob/main/lubridate.pdf
[lub3]: https://lubridate.tidyverse.org/

### 2022-11-27 src/check_links.Rmd
Replaced link_files with check_links

---
title: "Compute last modified dates"
author: "Steve Simon"
date: "Created 2022-05-26"
output: html_document
---

### Setup

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo=TRUE)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(tidyverse)))
```

### Read html files

```{r read_old}
old_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
old_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
old_pattern <- "html$"
new_path <- "../text/"
new_pattern <- "md$"
subdir_list=c("99", paste0("0", 0:9), 10:22)
md_files <- NULL
html_files <- NULL
for (yr in rev(subdir_list)) {
  print(yr)

  file_path <- paste0(new_path, yr)
  file_path %>%
    list.files(pattern="md$") %>%
    data.frame(yr=yr, fn=.) %>%
    bind_rows(md_files) -> md_files

  file_path <- paste0(old_path, yr)
  list.files(file_path, pattern="html$") -> yr_files
  if (length(yr_files)==0) next
  html_files <- bind_rows(
    data.frame(yr=yr, fn=yr_files),
    html_files)
}
```

### Read html files

```{r read_html}
html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
subdir_list=c("99", paste0("0", 0:9), 10:22)
html_files <- NULL
for (yr in rev(subdir_list)) {
  file_path <- paste0(html_path, yr)
  list.files(file_path, pattern="html$") -> yr_files
  if (length(yr_files)==0) next
  html_files <- bind_rows(
    data.frame(yr=yr, fn=yr_files),
    html_files)
}
count(html_files, yr)
```

### Read md files

```{r read_md}
md_path <- "../text/"
subdir_list=c("99", paste0("0", 0:9), 10:22)
md_files <- NULL
for (yr in rev(subdir_list)) {
  file_path <- paste0(md_path, yr)
  file_path %>%
    list.files(pattern="md$") %>%
    data.frame(yr=yr, fn=.) %>%
    bind_rows(md_files) -> md_files
}
count(md_files, yr)
```

### List duplicates

```{r list-duplicates}
md_files %>%
  group_by(fn) %>%
  mutate(n=n()) %>%
  filter(n>1) %>%
  select(-n) %>%
  arrange(fn, yr) %>%
  data.frame
html_files %>% count(yr)
```

Next step, search for 
+ [sim1]: http://www.pmean.com/??/...
+ <p>This page has moved to <a href="http://new.pmean.com/.../">a new website</a>.</p>

### Links found in md files

```{r md_links}
sim1 <- "[sim1]: http://www.pmean.com/"
full_file_name <- paste0(new_path, md_files$yr, "/", md_files$fn)
md_links <- NULL
for (i_fn in full_file_name) {
  read_lines(i_fn) %>%
    str_subset(fixed(sim1)) %>%
    str_replace(fixed(sim1), "") -> md_link
  md_link <- ifelse(length(md_link>0), md_link[1], "No link")
  md_file <- str_replace(i_fn, new_path, "")
  md_year <- str_sub(md_file, 1, 2)
  md_file <- str_sub(md_file, 4)
  md_links <- bind_rows(md_links, data.frame(md_year, md_file, md_link))
}
kable(md_links[1:10, ])
md_links %>%
  mutate(link=(md_link!="No link")) %>%
  count(md_year, link) %>%
  data.frame
```

### Links found in html files

```{r html_links}
moved1 <- '<p>This page has moved to <a href="http://new.pmean.com/'
moved3 <- '<p>This page is moving to a <a href="http://new.pmean.com/'
moved4 <- '/">new website</a>.</p>'
moved2 <- '/">a new website</a>.</p>'
full_file_name <- paste0(old_path, html_files$yr, "/", html_files$fn)
html_links <- NULL
for (i_fn in full_file_name) {
  read_lines(i_fn) %>%
    str_subset(glue("({moved1}|{moved3})")) %>%
    str_replace(fixed(moved1), "") %>%
    str_replace(fixed(moved2), "") %>%
    str_replace(fixed(moved3), "") %>%
    str_replace(fixed(moved4), "") -> html_link
  html_link <- ifelse(length(html_link>0), html_link[1], "No link")
  html_file <- str_replace(i_fn, old_path, "")
  html_year <- str_sub(html_file, 1, 2)
  html_file <- str_sub(html_file, 4)
  html_links <- bind_rows(html_links, data.frame(html_year, html_file, html_link))
}
kable(html_links[1:10, ])
html_links %>%
  mutate(link=(html_link!="No link")) %>%
  count(html_year, link) %>%
  data.frame
```

```{r save_everything}
save(md_links, html_links, file="../data/linkages.RData")
```

### 2022-11-29 src/check_links.Rmd
Fixed duplicate file names

@@
 ---
-title: "Compute last modified dates"
+title: "Check links between old and new pages"
 author: "Steve Simon"
@@
 ```{r read_old}
-old_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
-old_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
-old_pattern <- "html$"
-new_path <- "../text/"
-new_pattern <- "md$"
+html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
+html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
+html_pattern <- "html$"
+md_path <- "../text/"
+md_pattern <- "md$"
 subdir_list=c("99", paste0("0", 0:9), 10:22)
@@
 
-  file_path <- paste0(new_path, yr)
+  file_path <- paste0(md_path, yr)
   file_path %>%
@@
 
-  file_path <- paste0(old_path, yr)
+  file_path <- paste0(html_path, yr)
   list.files(file_path, pattern="html$") -> yr_files
@@
 ```{r read_html}
+html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
 html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
-html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
 subdir_list=c("99", paste0("0", 0:9), 10:22)
 html_files <- NULL
-for (yr in rev(subdir_list)) {
+for (yr in subdir_list) {
   file_path <- paste0(html_path, yr)
@@
 sim1 <- "[sim1]: http://www.pmean.com/"
-full_file_name <- paste0(new_path, md_files$yr, "/", md_files$fn)
+full_file_name <- paste0(md_path, md_files$yr, "/", md_files$fn)
 md_links <- NULL
@@
   md_link <- ifelse(length(md_link>0), md_link[1], "No link")
-  md_file <- str_replace(i_fn, new_path, "")
+  md_file <- str_replace(i_fn, md_path, "")
   md_year <- str_sub(md_file, 1, 2)
@@
 moved2 <- '/">a new website</a>.</p>'
-full_file_name <- paste0(old_path, html_files$yr, "/", html_files$fn)
+full_file_name <- paste0(html_path, html_files$yr, "/", html_files$fn)
 html_links <- NULL
@@
   html_link <- ifelse(length(html_link>0), html_link[1], "No link")
-  html_file <- str_replace(i_fn, old_path, "")
+  html_file <- str_replace(i_fn, html_path, "")
   html_year <- str_sub(html_file, 1, 2)

### 2022-11-30 src/check_links.Rmd
Improved display of counts in check_links.Rmd

--- 2022-11-27 src/check_links.Rmd
@@
 kable(md_links[1:10, ])
+```
+
+```{r md_year_summary}
 md_links %>%
@@
   count(md_year, link) %>%
-  data.frame
+  pivot_wider(
+    id_cols=md_year, 
+    names_from=link, 
+    names_prefix="linked_", 
+    values_from=n,
+    values_fill=0) %>%
+  mutate(total=linked_FALSE+linked_TRUE) %>%
+  data.frame -> yr_count
+yr_count %>%
+  summarize(
+    linked_FALSE=sum(linked_FALSE),
+    linked_TRUE=sum(linked_TRUE),
+    total=sum(total)) %>%
+  mutate(md_year="All") %>%
+  bind_rows(yr_count, .)
 ```
@@
   count(html_year, link) %>%
-  data.frame
+  pivot_wider(
+    id_cols=html_year, 
+    names_from=link, 
+    names_prefix="linked_", 
+    values_from=n,
+    values_fill=0) %>%
+  mutate(total=linked_FALSE+linked_TRUE) %>%
+  data.frame -> yr_count
+yr_count %>%
+  summarize(
+    linked_FALSE=sum(linked_FALSE),
+    linked_TRUE=sum(linked_TRUE),
+    total=sum(total)) %>%
+  mutate(html_year="All") %>%
+  bind_rows(yr_count, .)
 ```

### 2022-11-30 src/check_links.Rmd
Added d3-overview

--- 2022-11-29 src/check_links.Rmd
@@
 }
-kable(md_links[1:10, ])
 md_links %>%
@@
 }
-kable(html_links[1:10, ])
 html_links %>%
@@
 
+### List all linkages
+
+```{r md-linkages}
+kable(md_links)
+```
+
+```{r html-linkages}
+kable(html_links)
+```
+
 ```{r save_everything}

### 2022-12-06 src/check_links.Rmd

@@
 }
+
 md_links %>%

### 2022-12-06 text/22/definition-schoenfeld-residuals.Rmd

---
title: "Schoenfeld residuals"
author: "Steve Simon"
date: '2022-11-30'
output:
  html_document: default
category: Blog post
tags: Survival analysis
source: new
---

There are several different types of residuals in a Cox proportional hazards model. One of them is the Schoenfeld residuals. They are used to test the proportional hazards assumption.

<!---more--->

To understand the Schoenfeld residuals, you need to first recall the partial likelihood that is used to estimate hazard ratios in Cox regression model.

$L_i(\beta)  
 =\frac{\lambda(Y_i\mid X_i)}{\sum_{j:Y_j\ge Y_i}\lambda(Y_i\mid X_j)} 
 =\frac{\lambda_0(Y_i)\theta_i}{\sum_{j:Y_j\ge Y_i}\lambda_0(Y_i)\theta_j}
 =\frac{\theta_i}{\sum_{j:Y_j\ge Y_i}\theta_j},$

where

$\theta_j = exp(X_j \beta)$

### 2022-12-06 text/22/definition-schoenfeld-residuals.Rmd
Added comments about likelihood in definition-schoenfeld-residuals.

@@
 
+The likelihood function is the foundation for most statistical methods.
+You are trying to estimate some unknown parameters $\theta$ and each subject has a density function that is related to data measured on that subject and the unknown parameters.
+
+Here's a simple example.
+Assume that there is no censoring and that the event time is exponential with a scale parameter $\lambda_0$ for the control group and a mean parameter that is $\lambda_0+\lambda_1$ for the exposed or treatment group.
+The data in this case is the actual event time Y_i and an indicator variable X_i that is equal to zero for the control group and 1 for the exposed group.
+
+To simplify things, I will let
+
+$\lambda=(\lambda_0, \lambda_1)$
+
+$Y=(Y_1, Y_2, ..., Y_n)$
+
+and define X similarly.
+
+The density function,evaluated at the observed survival time is
+
+$f_i(Y_i, X_i, \lambda)=(\lambda_0+\lambda_1X_i)exp(-Y_i(\lambda_0+\lambda_1X_i))$
+
+We want to find "good values" of $\lambda$, ones that are associated with large (and therefore more likely) values of the density.
+The maximum likelihood estimate is the value of $lambda$ that maximizes the products of all these densities.
+For the exponential case, this would be
+
+$L(\lambda, Y, X)=\prod_i (\lambda_0+\lambda_1X_i) exp(-Y_i/(\lambda_0+\lambda_1X_i))$
+
+Let's make up some simple data for this example. Let Y=(1, 5, 2) and X=(0, 0, 1). The likelihood function in this simple case is
+
+$L(\lambda, Y, X)=\lambda_0 exp(-1 \lambda_0) \lambda_0 exp(-5 \lambda_0) (\lambda_0+\lambda_1) exp(-5(\lambda_0+\lambda_1))$
+
+It is often simpler to maximize the log of the likelihood.
+There are several reasons for this.
+First, a product of densities turns into a sum of log densities and sums are easier to work with than products.
+Second, many densities have an exponential function built into them and that function "disappears" after taking logs.
+I will use lower case for the log likelihood.
+
+$l(\lambda, Y, X)=log\lambda_0 - \lambda_0 + log\lambda_0 - 5 \lambda_0 + log(\lambda_0 + \lambda_1) -2 (\lambda_0+\lambda_1))$
+
+or
+
+$l(\lambda, Y, X) = 2 log\lambda_0 -9 \lambda_0 + log(\lambda_0+\lambda_1) -2 \lambda_1$
+
+Let's graph this function.
+
+```{r}
+n0 <- 20
+n1 <- 21
+
+u0 <- seq(1/n0, 0.7, length=n0)
+u1 <- seq(0, 1.4, length=n1)
+
+l <- function(u0, u1) {
+  2*log(u0)-9*u0+log(u0+u1)-2*u1
+}
+
+l_matrix <- matrix(nrow=n0, ncol=n1)
+
+for (i0 in 1:n0) {
+  for (i1 in 1:n1) {
+    l_matrix[i0, i1] <- l(u0[i0], u1[i1])
+  }
+}
+
+plot(u0, l_matrix[1:n0, 1])
+plot(u1, l_matrix[1, 1:n1])
+persp(u0, u1, l_matrix, theta=15, phi=30)
+```
+
 $L_i(\beta)  

### 2022-12-09 text/22/definition-schoenfeld-residuals.Rmd
Updated 3D plots in definition-schoenfeld-residuals

@@
 We want to find "good values" of $\lambda$, ones that are associated with large (and therefore more likely) values of the density.
-The maximum likelihood estimate is the value of $lambda$ that maximizes the products of all these densities.
-For the exponential case, this would be
+The maximum likelihood estimate is the value of $\lambda$ that maximizes the products of all these densities.
 
-$L(\lambda, Y, X)=\prod_i (\lambda_0+\lambda_1X_i) exp(-Y_i/(\lambda_0+\lambda_1X_i))$
+Let's make up some simple data for this example. Let Y=(3, 5, 1) and X=(0, 0, 1). The likelihood function in this simple case is
 
-Let's make up some simple data for this example. Let Y=(1, 5, 2) and X=(0, 0, 1). The likelihood function in this simple case is
-
-$L(\lambda, Y, X)=\lambda_0 exp(-1 \lambda_0) \lambda_0 exp(-5 \lambda_0) (\lambda_0+\lambda_1) exp(-5(\lambda_0+\lambda_1))$
+$L(\lambda, Y, X)=\lambda_0 exp(-3 \lambda_0) \lambda_0 exp(-5 \lambda_0) (\lambda_0+\lambda_1) exp(-(\lambda_0+\lambda_1))$
 
@@
 
-$l(\lambda, Y, X)=log\lambda_0 - \lambda_0 + log\lambda_0 - 5 \lambda_0 + log(\lambda_0 + \lambda_1) -2 (\lambda_0+\lambda_1))$
+$l(\lambda, Y, X)=log\lambda_0 - 3 \lambda_0 + log\lambda_0 - 5 \lambda_0 + log(\lambda_0 + \lambda_1) - (\lambda_0+\lambda_1)$
 
@@
 
-$l(\lambda, Y, X) = 2 log\lambda_0 -9 \lambda_0 + log(\lambda_0+\lambda_1) -2 \lambda_1$
+$l(\lambda, Y, X) = 2 log\lambda_0 - 9 \lambda_0 + log(\lambda_0+\lambda_1) -\lambda_1$
 
@@
 ```{r}
-n0 <- 20
-n1 <- 21
+library(plot3D)
 
-u0 <- seq(1/n0, 0.7, length=n0)
-u1 <- seq(0, 1.4, length=n1)
-
-l <- function(u0, u1) {
-  2*log(u0)-9*u0+log(u0+u1)-2*u1
+l <- function(lambda0, lambda1) {
+  2*log(lambda0)-9*lambda0+log(lambda0+lambda1)-lambda1
 }
 
-l_matrix <- matrix(nrow=n0, ncol=n1)
+n0 <- 200
+n1 <- 200
+
+lambda0 <- seq(1/n0, 1, length=n0)
+lambda1 <- seq(0, 1, length=n1)
+
+fine_mesh <- matrix(nrow=n0, ncol=n1)
 
@@
   for (i1 in 1:n1) {
-    l_matrix[i0, i1] <- l(u0[i0], u1[i1])
+    fine_mesh[i0, i1] <- l(lambda0[i0], lambda1[i1])
   }
 }
+```
 
-plot(u0, l_matrix[1:n0, 1])
-plot(u1, l_matrix[1, 1:n1])
-persp(u0, u1, l_matrix, theta=15, phi=30)
+persp3D(
+  x=lambda0, 
+  y=lambda1, 
+  z=fine_mesh, 
+  theta=15,
+  phi=30, 
+  xlab="lambda0",
+  ylab="lambda1",
+  zlab="log likelihood")
+segments3D(
+  x0=0.25,
+  y0=0.75, 
+  z0=min(fine_mesh),
+  x1=0.25,
+  y1=0.75,
+  z1=max(fine_mesh),
+  add=TRUE)
+points3D(
+  x=0.25,
+  y=0.75, 
+  z=min(fine_mesh),
+  add=TRUE)
+points3D(
+  x=0.25,
+  y=0, 
+  z=min(fine_mesh),
+  add=TRUE)
 ```
+
+```
+
+To find out where the maximum occurs, take the first two partial derivatives and set them equal to zero.
+
+$\frac{\partial /l}{\partial \lambda0} = 2/\lambda_0 -9 + 1/(\lambda_0+\lambda_1) = 0$
+
+$\frac{\partial /l}{\partial \lambda1} = 1/(\lambda_0+\lambda_1) - 1 = 0$
+
+The second equation tells you that
+
+$\lambda_0+\lambda_1=1$
+
+Substituting that into the second equation gives you
+
+$\lambda_0=0.25$
+
+and then you can show that
+
+$\lambda_1=0.75$
 

### 2022-12-09 text/22/definition-schoenfeld-residuals.Rmd
Minor edits to definition-schoenfeld-residuals

--- 2022-12-06 text/22/definition-schoenfeld-residuals.Rmd
@@
 To understand the Schoenfeld residuals, you need to first recall the partial likelihood that is used to estimate hazard ratios in Cox regression model.
+
+### Likelihood
 
@@
 We want to find "good values" of $\lambda$, ones that are associated with large (and therefore more likely) values of the density.
-The maximum likelihood estimate is the value of $lambda$ that maximizes the products of all these densities.
+The maximum likelihood estimate is the value of $\lambda$ that maximizes the products of all these densities.
 For the exponential case, this would be
@@
 Let's graph this function.
+
+```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo=FALSE)
+suppressMessages(suppressWarnings(library(plot3D)))
+```
 
@@
 persp(u0, u1, l_matrix, theta=15, phi=30)
+
+cycle_fast <- function(g=1001, cmin=0, cmax=1, rep=g) {
+  cyc <- seq(cmin, cmax, length=g)
+  return(rep(cyc, times=rep))
+}
+
+cycle_slow <- function(g=1001, cmin=0, cmax=1, rep=g) {
+  cyc <- seq(cmin, cmax, length=g)
+  return(rep(cyc, each=rep))
+}
+cycle_fast(3)
+cycle_slow(3)
+
+#  persp3D(z=d, xlab="lambda1", ylab="lambda2",zlab="Density", zlim=c(0, zmax),
+#          theta=th, phi=ph, colvar=NULL, col="white", border="black",
+#          bty="u", col.axis="gray90")
+
 ```
+
+This is a bit lopsided, but not too far off from a three dimensional equivalent of a parabola. There are several interesting things you can do with the log likelihood.
+
+First, you can find values of $\lambda_1$ and $\lambda_2$ that maximize this function. This is the maximum likelihood estimate.
+
+In this example, the maximum value occurs at $\lambda_1=0.33$ and $\lambda_2=0.5$. (Note: double check this!!!)
+
+Second, you can compare the log likelihood at the maximum to an alternative hypothesized values for $\lambda_1$ and $\lambda_2$.
+If the log likelihood is a lot smaller, then you should reject your hypothesized values.
+If the log likelihood at the hypothesized values is only slightly smaller, then you should accept these hypothesized values are reasonable.
+This is the maximum likelihood test.
+
+For example, you can hypothesize that $\lambda_2$ is zero. You let the value of $\lambda_1$ be unconstrained. This is effectively calling the two groups identical.
+
+The maximum value for the log likelihood is `r max(l_matrix)`.
+If you set $\lambda_2=0$, the log_likelihood is `r max(l_matrix[ , 1])`.
+Since these values are close, you can consider the hypothesis to be reasonable.
+
+You can also look at the partial derivative of the log likelihood at the hypothesized values.
+If the partial derivative is close to zero, that means that you are pretty close in log likelihood terms to the maximum likelihood, since the partial derivatives at the maximum have to be zero.
+In such a case, you can accept your hypothesized value as reasonable.
+If the partial derivative is large, you have a steep uphill climb until you reach the maximum likelihood.
+You are nowhere near the maximum and should reject the hypothesized value as unreasonable.
+
+In this example, the slope is approximately 0.05. Since this value is small, you can accept the hypothesized values as reasonable.
+
+Finally, you can compute the matrix of second partial derivatives at the maximum likelihood estimate.
+If this matrix is large, that means that the log likelihoods fall off rapidly once you move away from the peak.
+You can conclude that your maximum likelihood estimate has a lot of precision.
+Conversely, if the matrix of second derivatives is small, that means that the likelihood goes away from the maximum quite slowly for a large distance.
+In this case, your maximum likelihood estimate has poor precision.
 

### 2022-12-09 text/22/definition-schoenfeld-residuals.Rmd

@@
 The maximum likelihood estimate is the value of $\lambda$ that maximizes the products of all these densities.
-For the exponential case, this would be
 
-$L(\lambda, Y, X)=\prod_i (\lambda_0+\lambda_1X_i) exp(-Y_i/(\lambda_0+\lambda_1X_i))$
+Let's make up some simple data for this example. Let Y=(3, 5, 1) and X=(0, 0, 1). The likelihood function in this simple case is
 
-Let's make up some simple data for this example. Let Y=(1, 5, 2) and X=(0, 0, 1). The likelihood function in this simple case is
-
-$L(\lambda, Y, X)=\lambda_0 exp(-1 \lambda_0) \lambda_0 exp(-5 \lambda_0) (\lambda_0+\lambda_1) exp(-5(\lambda_0+\lambda_1))$
+$L(\lambda, Y, X)=\lambda_0 exp(-3 \lambda_0) \lambda_0 exp(-5 \lambda_0) (\lambda_0+\lambda_1) exp(-(\lambda_0+\lambda_1))$
 
@@
 ```{r}
-n0 <- 20
-n1 <- 21
+library(plot3D)
 
-u0 <- seq(1/n0, 0.7, length=n0)
-u1 <- seq(0, 1.4, length=n1)
-
-l <- function(u0, u1) {
-  2*log(u0)-9*u0+log(u0+u1)-2*u1
+l <- function(lambda0, lambda1) {
+  2*log(lambda0)-9*lambda0+log(lambda0+lambda1)-lambda1
 }
 
-l_matrix <- matrix(nrow=n0, ncol=n1)
+n0 <- 50
+n1 <- 50
+
+lambda0 <- seq(1/n0, 1, length=n0)
+lambda1 <- seq(0, 1, length=n1)
+
+fine_mesh <- matrix(nrow=n0, ncol=n1)
 
@@
   for (i1 in 1:n1) {
-    l_matrix[i0, i1] <- l(u0[i0], u1[i1])
+    fine_mesh[i0, i1] <- l(lambda0[i0], lambda1[i1])
   }
@@
 
-plot(u0, l_matrix[1:n0, 1])
-plot(u1, l_matrix[1, 1:n1])
-persp(u0, u1, l_matrix, theta=15, phi=30)
-
-cycle_fast <- function(g=1001, cmin=0, cmax=1, rep=g) {
-  cyc <- seq(cmin, cmax, length=g)
-  return(rep(cyc, times=rep))
-}
-
-cycle_slow <- function(g=1001, cmin=0, cmax=1, rep=g) {
-  cyc <- seq(cmin, cmax, length=g)
-  return(rep(cyc, each=rep))
-}
-cycle_fast(3)
-cycle_slow(3)
-
-#  persp3D(z=d, xlab="lambda1", ylab="lambda2",zlab="Density", zlim=c(0, zmax),
-#          theta=th, phi=ph, colvar=NULL, col="white", border="black",
-#          bty="u", col.axis="gray90")
-
+persp3D(
+  x=lambda0, 
+  y=lambda1, 
+  z=fine_mesh, 
+  zlim=c(min(fine_mesh), -5),
+  contour=TRUE,
+  theta=15,
+  phi=30, 
+  xlab="lambda0",
+  ylab="lambda1",
+  zlab="log likelihood")
+segments3D(
+  x0=0.25,
+  y0=0.75, 
+  z0=min(fine_mesh),
+  x1=0.25,
+  y1=0.75,
+  z1=max(fine_mesh),
+  add=TRUE)
+text3D(
+  x=0.25,
+  y=0.75, 
+  z=min(fine_mesh),
+  labels=round(max(fine_mesh), 1),
+  col="red",
+  add=TRUE)
+segments3D(
+  x0=0.3333,
+  y0=0, 
+  z0=min(fine_mesh),
+  x1=0.3333,
+  y1=0,
+  z1=max(fine_mesh[ , 1]),
+  add=TRUE)
+text3D(
+  x=0.3333,
+  y=0., 
+  z=min(fine_mesh),
+  labels=round(max(fine_mesh[ , 1]), 1),
+  col="red",
+  add=TRUE)
 ```
@@
 
-The maximum value for the log likelihood is `r max(l_matrix)`.
-If you set $\lambda_2=0$, the log_likelihood is `r max(l_matrix[ , 1])`.
+The maximum value for the log likelihood is `r max(fine_mesh)`.
+If you set $\lambda_2=0$, the log_likelihood is `r max(fine_mesh[ , 1])`.
 Since these values are close, you can consider the hypothesis to be reasonable.
@@
 
+To find out where the maximum occurs, take the first two partial derivatives and set them equal to zero.
+
+$\frac{\partial /l}{\partial \lambda0} = 2/\lambda_0 -9 + 1/(\lambda_0+\lambda_1) = 0$
+
+$\frac{\partial /l}{\partial \lambda1} = 1/(\lambda_0+\lambda_1) - 1 = 0$
+
+The second equation tells you that
+
+$\lambda_0+\lambda_1=1$
+
+Substituting that into the second equation gives you
+
+$\lambda_0=0.25$
+
+and then you can show that
+
+$\lambda_1=0.75$
+
+>>>>>>> 17772f400cc3acd5b2c61623fc3cf2c98df50124
+
 $L_i(\beta)  

### 2022-12-15 text/22/box-with-no-whiskers.Rmd
Started box-with-no-whiskers

---
title: "My boxplot shows a box but no whiskers"
author: "Steve Simon"
date: '2022-12-15'
output:
  html_document: default
category: Blog post
tags: Graphical display
source: new
---

One of my favorite graphs is the box plot. It is a visualization of a five number summary. Those five numbers are minimum, 25th percentile, median (aka the 50th percentile), 75th percentile, and maximum. The box in a box plot extends from the 25th to the 75th percentile and a line inside the box represents the median.

It's great for comparing two groups. Here's a boxplot showing two groups. Notice that the first group tends to have larger values.

```{r, fig.height=1.5, echo=FALSE}
suppressMessages(suppressWarnings(library(tidyverse)))
n1 <- 10
n2 <- 10
pseudo_boxplot <- function(n1, n2, a1, b1, a2, b2) {
  x1 <- qbeta((1:n1)/(n1+1), 6, 2)
  x2 <- qbeta((1:n2)/(n2+1), 5, 3)
  data.frame(x=c(x1, x2), g=paste0("g", rep(1:2, c(n1, n2)))) -> df
  print(df)
  df %>%
    ggplot(aes(g, x)) +
      geom_boxplot() +
      coord_flip()
}

pseudo_boxplot(10, 10, 6, 2, 5, 3)
```

What happens if you try to compute the five number summary with a dataset that only has only two values?

```{r, fig.height=1.5, echo=FALSE}
pseudo_boxplot(10, 2, 6, 2, 5, 3)
```

### 2023-01-06 src/yearly-archive.R
Added 2022 archive

@@ -51,4 +51,4 @@ archive_old <- function(yr) {
 yrs <- c("2003", "2007", "2011", "2015", "2019", "2021")
 paste0(brack(yrs), paren("http://new.pmean.com/" %0% yrs %0% "/"), collapse=" |\n") %>% cat
 cat("\n\n")
-archive_new("21")
+archive_new("22")

### 2023-01-06 text/22/2022.Rmd
Added 2023.md

---
title: "Pages written in 2022"
source: "New"
date: "2022-12-31"
categories:
- Blog post
tags:
- Website details
output: html_document
---

```{r setup, echo=FALSE}
library(glue)
library(magrittr)
library(stringr)
yr <- 2023:1999
glue("[{yr}](http://new.pmean.com/{yr}/)") %>%
  paste(collapse=" | ") -> yr_list
list.files() %>%
  str_subset("md$") %>%
  str_replace(fixed(".Rmd"), "") %>%
  str_replace(fixed(".md"), "") -> f
n <- length(f):1
header <- glue("New website ({n[1]} files)")
glue("+ [{f}](http://new.pmean.com/{f}/)  ") %>%
  paste(collapse="  \n") -> file_list
```

`r yr_list`

`r header`

`r file_list`

### 2023-01-07 text/22/2022.Rmd
Added date checks to 2022.Rmd

@@
 ```{r setup, echo=FALSE}
+knitr::opts_chunk$set(echo=FALSE)
 library(glue)
 library(magrittr)
+library(readr)
 library(stringr)
@@
 list.files() %>%
-  str_subset("md$") %>%
+  str_subset("md$") -> f
+```
+
+```{r}
+problem_files <- NULL
+for (fn in f) {
+  file_text <- read_lines(fn)
+  file_text %>%
+    str_subset("^source\\: ") %>%
+    str_remove(fixed("source: ")) %>%
+    str_remove_all(fixed('"')) %>%
+    tolower -> file_source
+  if (length(file_source)>1) {
+    problem_files %<>% append(glue("{fn} has multiple sources\n"))
+  }
+  if (length(file_source)>0 & file_source[1]!="new") {
+    problem_files %<>% append(glue("{fn} has source={file_source}\n"))
+  }
+  file_text %>%
+    str_subset(fixed("date: ")) %>%
+    str_remove(fixed("date: ")) -> file_date
+  if (length(file_date)==0) {
+    problem_files %<>% append(glue("{fn} has no date"))
+  }
+}
+```
+
+```{r}
+f %>%
   str_replace(fixed(".Rmd"), "") %>%
@@
 `r file_list`
+
+```{r}
+if (length(problem_files)==0) {
+  problem_files <- "All files passed date and source quality checks"
+}
+print(problem_files)
+```

### 2023-01-10 src/pages.Rmd
Started work on pages.Rmd

---
title: "Pages written in xxxx template"
source: new
date: "2023-01-10"
categories:
- Blog post
tags:
- Website details
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(stringr)))
```

```{r}

check_source <- function(file_text) {
  file_text %>%
    str_subset("^source\\: ") %>%
    str_remove(fixed("source: ")) %>%
    str_remove_all('"')  %>%
    str_remove_all("'") -> file_source

  if (length(file_source)==0) return(
    glue("{fn} has no source\n"))

  if (length(file_source)>1) return(
    glue("{fn} has multiple sources\n"))

  if (file_source!="new") return(
    glue("{fn} has source={file_source}\n"))
  return(NULL)
}

check_dates <- function(file_text) {
  file_text %>%
    str_subset(fixed("date: ")) %>%
    str_remove(fixed("date: ")) %>%
    str_remove_all('"') %>%
    str_remove_all("'")  -> file_date
    if (length(file_date)==0) file_date <-"No date"
    return(file_date)
}

list_pages <- function(page_year) {
  # get yr_list
  yr <- 2023:1999
  glue("[{yr}](http://new.pmean.com/{yr}/)") %>%
    paste(collapse=" | ") -> yr_list

  # open files and find information
  problem_files <- NULL
  date_list <- NULL
  for (fn in f) {
    file_text <- read_lines(glue("../text/21/{fn}"))
    problem_files <- append(
      problem_files,
      check_source(file_text))
    date_list <- append(
      date_list,
      check_dates(file_text))
  }
  f %>%
    str_replace(fixed(".Rmd"), "") %>%
    str_replace(fixed(".md"), "") -> f
  n <- length(f)
  o <- order(date_list, f)
  header <- glue("New website ({n} files)")
  glue(
    "+ {date_list[o]} ",
    "[{f[o]}]",
    "(http://new.pmean.com/{f[o]}/)  ") %>%
    paste(collapse="  \n") -> file_list
  return(
    paste0(
      yr_list,
      header,
      file_list, 
      problem_files))
}
```

```{r}

```

`r yr_list`

`r header`

`r file_list`

```{r}
if (length(problem_files)==0) {
  problem_files <- "All files passed date and source quality checks"
}
```

`r paste(problem_files, collapse="\n")`

### 2023-01-10 text/20/non-breaking-spaces.Rmd
Started work on pages.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-01-07"

### 2023-01-10 text/21/blogdown-first-steps.Rmd
Started work on pages.Rmd

--- 2021-10-16 text/Rmd/first-steps-with-blogdown/index.Rmd
@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-10-09"

### 2023-01-10 text/21/direct-standardization.Rmd
Started work on pages.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-09-20"

### 2023-01-10 text/22/2022.Rmd
Started work on pages.Rmd

--- 2023-01-06 text/22/2022.Rmd
@@
 title: "Pages written in 2022"
-source: "New"
+source: new
 date: "2022-12-31"
@@
 ```{r setup, echo=FALSE}
-library(glue)
-library(magrittr)
-library(stringr)
+knitr::opts_chunk$set(echo=FALSE)
+suppressMessages(suppressWarnings(library(glue)))
+suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(readr)))
+suppressMessages(suppressWarnings(library(stringr)))
 yr <- 2023:1999
@@
 list.files() %>%
-  str_subset("md$") %>%
+  str_subset("md$") -> f
+```
+
+```{r}
+problem_files <- NULL
+date_list <- NULL
+for (fn in f) {
+  file_text <- read_lines(fn)
+  file_text %>%
+    str_subset("^source\\: ") %>%
+    str_remove(fixed("source: ")) %>%
+    str_remove_all('"')  %>%
+    str_remove_all("'") -> file_source
+  if (length(file_source)>1) {
+    problem_files %<>% append(glue("{fn} has multiple sources\n"))
+  }
+  if (length(file_source)>0 & file_source[1]!="new") {
+    problem_files %<>% 
+      append(glue("{fn} has source={file_source}\n"))
+  }
+  file_text %>%
+    str_subset(fixed("date: ")) %>%
+    str_remove(fixed("date: ")) %>%
+    str_remove_all('"') %>%
+    str_remove_all("'")  -> file_date
+  if (length(file_date)==0) {
+    problem_files %<>% append(glue("{fn} has no date"))
+  }
+  date_list <- append(date_list, file_date[1])
+}
+```
+
+```{r}
+f %>%
   str_replace(fixed(".Rmd"), "") %>%
   str_replace(fixed(".md"), "") -> f
-n <- length(f):1
-header <- glue("New website ({n[1]} files)")
-glue("+ [{f}](http://new.pmean.com/{f}/)  ") %>%
+n <- length(f)
+o <- order(date_list, f)
+header <- glue("New website ({n} files)")
+glue("+ {date_list[o]} [{f[o]}](http://new.pmean.com/{f[o]}/)  ") %>%
   paste(collapse="  \n") -> file_list
@@
 `r file_list`
+
+```{r}
+if (length(problem_files)==0) {
+  problem_files <- "All files passed date and source quality checks"
+}
+```
+
+`r paste(problem_files, collapse="\n")`

### 2023-01-11 text/20/ihme-model.Rmd

--- 2020-07-23 text/20/ihme-covid-model.Rmd
@@
 author: "Steve Simon"
-source: "New"
+source: new
+date: 2020-04-23
 categories:

### 2023-01-11 text/20/map-projections.Rmd

--- 2020-03-23 text/20/map-projections.Rmd
@@
 ---
-title: "Blog post: Different map projections"
+title: "Different map projections"
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-03-15"
-category: Blog post
-tags: Graphical display
+categories:
+- Blog post
+tags:
+- Graphical display
 output: html_document
 ---
+install.packages("rnaturalearthhires", repos="http://packages.ropensci.org", type="source")
 

### 2023-01-11 text/20/poor-color-chioce.Rmd

--- 2021-09-30 text/Rmd/misuse-of-gradient/index.Rmd
@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-03-11"

### 2023-01-11 text/20/reading-text-files.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-04-30"

### 2023-01-11 text/20/rmarkdown-papers.Rmd

--- 2020-12-21 text/20/markdown-papers.Rmd
@@
 title: "How I use R Markdown to write research papers"
-source: "New"
+source: new
+date: "2020-11-21"
 categories:

### 2023-01-11 text/20/smart-quotes.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-03-02"

### 2023-01-11 text/21/anonymizing-ids.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-06-10"

### 2023-01-11 text/21/dark-side.Rmd

@@
 title: "The dark side of big data"
-source: "New"
+source: new
 date: "2021-10-28"

### 2023-01-11 text/21/glm-diagnostics.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-10-18"

### 2023-01-11 text/21/hurricane-frequencies.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-12-11"

### 2023-01-11 text/21/negative-binomial-regression.Rmd

@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-08-30"

### 2023-01-11 text/21/read-acs-data.Rmd

@@
 title: "Reading information from the American Community Survey into R"
-source: "New"
+source: new
 date: "2021-09-29"

### 2023-01-11 text/22/testing-d3.Rmd

@@
 date: "2022-06-01"
-source: "New"
+source: new
 categories: 

### 2023-01-11 src/pages.Rmd
Continued work on pages.Rmd

@@
 ```{r}
+write_yaml <- function(page_year) {
+  date_line <- paste0("date", ": 20", page_year, "-12-31")
+  output_text <- glue('
+  ---
+  title: "Pages written in 20{page_year}"
+  source: new
+  {date_line}
+  categories:
+  - Blog post
+  tags:
+  - Website details
+  output: html_document
+  ---')
+  return(output_text)
+}
+```
 
+```{r}
+write_links <- function(yr_list) {
+  glue("[{yr}](http://new.pmean.com/{yr}/)") %>%
+    paste(collapse=" | ")
+}
+```
+
+```{r}
 check_source <- function(file_text) {
@@
 
-  if (length(file_source)==0) return(
-    glue("{fn} has no source\n"))
+  if (length(file_source)==0) return("Error: no source")
 
-  if (length(file_source)>1) return(
-    glue("{fn} has multiple sources\n"))
+  if (length(file_source)>1) return("Error: multiple sources")
 
-  if (file_source!="new") return(
-    glue("{fn} has source={file_source}\n"))
-  return(NULL)
+  return(file_source)
 }
+```
 
+```{r}
 check_dates <- function(file_text) {
@@
     str_remove_all("'")  -> file_date
-    if (length(file_date)==0) file_date <-"No date"
+  if (length(file_date)==0) return("Error: no date")
+  if (length(file_date)==0) return("Error: multiple dates")
   return(file_date)
 }
+```
 
-list_pages <- function(page_year) {
-  # get yr_list
-  yr <- 2023:1999
-  glue("[{yr}](http://new.pmean.com/{yr}/)") %>%
-    paste(collapse=" | ") -> yr_list
-
+```{r}
+write_list <- function(page_year) {
+  # get file list
+  list.files(path=glue("../text/{page_year}", pattern="*.md")) %>%
+    str_subset("md$") -> f
   # open files and find information
-  problem_files <- NULL
+  file_source <- NULL
   date_list <- NULL
   for (fn in f) {
-    file_text <- read_lines(glue("../text/21/{fn}"))
-    problem_files <- append(
-      problem_files,
+    file_text <- read_lines(glue("../text/{page_year}/{fn}"))
+    # get source
+    file_source <- append(
+      file_source,
       check_source(file_text))
+    #get dates
     date_list <- append(
@@
   }
+  # strip extensions
   f %>%
@@
   o <- order(date_list, f)
+  # create RMarkdown text
   header <- glue("New website ({n} files)")
-  glue(
+  file_list <- glue(
     "+ {date_list[o]} ",
@@
     "(http://new.pmean.com/{f[o]}/)  ") %>%
-    paste(collapse="  \n") -> file_list
-  return(
-    paste0(
-      yr_list,
-      header,
-      file_list, 
-      problem_files))
+    paste(collapse="  \n")
+  # search for problem files
+
+  return(paste0(header, file_list))
 }
@@
 ```{r}
-
+write_yaml(21)
 ```
 
-`r yr_list`
-
-`r header`
-
-`r file_list`
-
-```{r}
-if (length(problem_files)==0) {
-  problem_files <- "All files passed date and source quality checks"
-}
-```
-
-`r paste(problem_files, collapse="\n")`
+`r write_list(21)`

### 2023-01-12 src/pages.Rmd
Added source link to 2020 definition pages

@@
 ---
-title: "Pages written in xxxx template"
+title: "Pages written in a given year"
 source: new
@@
   source: new
-  {date_line}
+  date: 20{page_year}-12-31
   categories:
@@
 write_links <- function(yr_list) {
-  glue("[{yr}](http://new.pmean.com/{yr}/)") %>%
+  glue("[{yr_list}](http://new.pmean.com/{yr_list}/)") %>%
     paste(collapse=" | ")
@@
   file_text %>%
-    str_subset(fixed("date: ")) %>%
+    str_subset("^date\\: ") %>%
     str_remove(fixed("date: ")) %>%
@@
   if (length(file_date)==0) return("Error: no date")
-  if (length(file_date)==0) return("Error: multiple dates")
+  if (length(file_date)>1) return("Error: multiple dates")
   return(file_date)
@@
   # open files and find information
-  file_source <- NULL
+  source_list <- NULL
   date_list <- NULL
@@
     # get source
-    file_source <- append(
-      file_source,
+    source_list <- append(
+      source_list,
       check_source(file_text))
@@
   n <- length(f)
+  # search for problem files
+  bad_date <- str_detect(date_list, "Error")
+  bad_source <- str_detect(source_list, "Error")
+  # create RMarkdown text
+  links <- glue(
+    "+ {date_list} ",
+    "[{f}]",
+    "(http://new.pmean.com/{f}/)  ")
+  # store in a sorted data frame
   o <- order(date_list, f)
-  # create RMarkdown text
-  header <- glue("New website ({n} files)")
-  file_list <- glue(
-    "+ {date_list[o]} ",
-    "[{f[o]}]",
-    "(http://new.pmean.com/{f[o]}/)  ") %>%
-    paste(collapse="  \n")
-  # search for problem files
-
-  return(paste0(header, file_list))
+  return(
+    data.frame(
+      f, 
+      links,
+      date_list,
+      source_list,
+      bad_date,
+      bad_source)[o, ])
 }
@@
 ```{r}
-write_yaml(21)
+for (four_digit_year in c("2022")) {
+  two_digit_year <- str_sub(four_digit_year, 3, 4)
+  yaml_text <- write_yaml(two_digit_year)
+  l <- write_list(two_digit_year)
+  n <- dim(l)[1]
+  fn <- glue("../text/{two_digit_year}/{four_digit_year}.md")
+  write_lines(
+    yaml_text, 
+    file=fn,
+    append=FALSE)
+  write_lines(
+    glue("New website ({n} files)\n\n"),
+    file=fn,
+    append=TRUE)
+  write_lines(
+    l$links, 
+    file=fn,
+    append=TRUE)
+  problem_files <- c(
+    paste(l$f, l$date_list)[l$bad_date],
+    paste(l$f, l$source_list)[l$bad_source])
+  all_good <- "All files passed data and source quality checks"
+  if (length(problem_files)==0) problem_files <- all_good
+  write_lines(
+    c("\n", problem_files),
+    file=fn,
+    append=TRUE)
+}
 ```
 
-`r write_list(21)`
+Links created for `r dim(l)[1]` pages in `r four_digit_year`.
+
+`r paste(problem_files, collapse="\n")`
+
+`r l$source_list`

### 2023-01-12 text/21/cubic-splines.Rmd
Added source link to 2020 definition pages

@@
 title: "Basics of cubic spline models"
+source: new
 author: "Steve Simon"

### 2023-01-12 text/21/venn-diagram.Rmd
Added source link to 2020 definition pages

--- 2021-10-30 text/21/test.Rmd
@@
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2021-10-30"

### 2023-01-12 text/21/very-simple-splines.Rmd
Added source link to 2020 definition pages

@@
 title: "Very simple spline models"
+source: new
 author: "Steve Simon"

### 2023-01-12 src/pages.Rmd
Cleaned up 2018 files

@@
   bad_source <- str_detect(source_list, "Error")
+  old_source <- (!str_detect(source_list, "new")) & (!bad_source)
+  old_link <- glue("Link to earlier page [here]({source_list}).")
+  old_link[!old_source] <- ""
   # create RMarkdown text
@@
     "[{f}]",
-    "(http://new.pmean.com/{f}/)  ")
+    "(http://new.pmean.com/{f}/). {old_link}  ")
   # store in a sorted data frame
@@
 ```{r}
-for (four_digit_year in c("2022")) {
+for (four_digit_year in c("2018")) {
   two_digit_year <- str_sub(four_digit_year, 3, 4)

### 2023-01-12 text/21/poisson-inferences.Rmd
Cleaned up 2018 files

@@
 author: "Steve Simon"
-source: "New"
-date: "2020-08-30"
+source: new
+date: "2021-08-30"
 categories:

### 2023-01-12 src/pages.Rmd
Clean up 2014 pages

@@
 ```{r}
-for (four_digit_year in c("2018")) {
+for (four_digit_year in c("2014")) {
   two_digit_year <- str_sub(four_digit_year, 3, 4)
@@
     paste(l$f, l$source_list)[l$bad_source])
-  all_good <- "All files passed data and source quality checks"
+  all_good <- "All files passed date and source quality checks"
   if (length(problem_files)==0) problem_files <- all_good

### 2023-01-12 src/pages.Rmd
Added source to 2014 files

@@
 ```{r}
-for (four_digit_year in c("2014")) {
+for (four_digit_year in c("2012")) {
   two_digit_year <- str_sub(four_digit_year, 3, 4)
@@
 
-`r paste(problem_files, collapse="\n")`
+`r paste(paste("+", problem_files), collapse="\n")`
 

### 2023-01-12 src/pages.Rmd
Created master page for 1999

--- 2023-01-12 src/pages.Rmd
@@
 ```{r}
-for (four_digit_year in c("2014")) {
+for (four_digit_year in c("1999")) {
   two_digit_year <- str_sub(four_digit_year, 3, 4)
@@
   write_lines(
-    c("\n", problem_files),
+    "\n\n### Quality check\n",
+    file=fn,
+    append=TRUE)
+  write_lines(
+    paste0("+ ", problem_files),
     file=fn,
@@
 
-`r paste(problem_files, collapse="\n")`
+`r paste(paste("1.", problem_files), collapse="\n")`
 

### 2023-01-12 src/pages.Rmd
Added source to 2000 files

@@
 write_yaml <- function(page_year) {
-  date_line <- paste0("date", ": 20", page_year, "-12-31")
+  century <- ifelse(page_year==99, 19, 20)
   output_text <- glue('
   ---
-  title: "Pages written in 20{page_year}"
+  title: "Pages written in {century}{page_year}"
   source: new
-  date: 20{page_year}-12-31
+  date: {century}{page_year}-12-31
   categories:

### 2023-01-13 src/pages.Rmd
Updated 2000 master list

@@
 ```{r}
-for (four_digit_year in c("1999")) {
+for (four_digit_year in c("2000")) {
   two_digit_year <- str_sub(four_digit_year, 3, 4)

### 2023-02-19 text/21/glm-deviance.Rmd
Added year-end-summary.R

@@
 ```{r}
-fn <- "http://www.stata-press.com/data/lf2/couart2.dta"
+fn <- "http://www.pmean.com/00files/couart2.dta"
 articles <- read.dta(fn)
@@
 [rod2]: https://data.princeton.edu/wws509/r/overdispersion
-[rod3]: http://www.stata-press.com/data/lf2/couart2.dta
+[rod3]: http://www.pmean.com/00files/couart2.dta
 [sim3]: http://www.pmean.com/00files/couart2.dta

### 2023-02-19 src/clean_files.R

@@ -74,7 +74,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'source: New',
-  'source: new',
+  fixed('http://www.stata-press.com/data/lf2/couart2.dta'),
+  'http://www.pmean.com/00files/couart2.dta',
   dir_root="text", 
   file_pattern="md$")

### 2023-02-19 src/year-end-summary.R

@@ -1,0 +1,128 @@
+# This program produces Markdown files summarizing
+# every webpage written in a given year
+
+# Preliminaries
+suppressMessages(suppressWarnings(library(glue)))
+suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(readr)))
+suppressMessages(suppressWarnings(library(stringr)))
+
+# pad single digits with leading zero
+zpad <- function(x) {
+  ifelse(nchar(x)==1, paste0("0", x), x)
+}
+
+# Create yaml header
+create_yaml <- function(yyyy) {
+  glue(
+    '---
+     title: "Pages written in 2022"
+     source: new
+     date: {yyyy}-12-31
+     categories:
+     - Blog post
+     tags:
+     - Website details
+     output: html_document
+     ---')
+}
+
+# Create links to other years
+create_links <- function(yyyy_range) {
+  yy_range <- str_sub(yyyy_range, 3)
+  glue("[{yyyy_range}](http://new.pmean.com/{yyyy_range}/)") %>%
+    paste(collapse=" | ")
+}
+
+# Check for problems
+
+# Check source
+check_source <- function(file_text) {
+  message <- NULL
+  file_text %>%
+    str_subset("^source\\: ") %>%
+    str_remove(fixed("source: ")) %>%
+    str_remove_all('"')  %>%
+    str_remove_all("'") -> file_source
+  if (length(file_source)==0) {
+    message <- glue("{fn} has no sources\n")
+  }
+  if (length(file_source)>1) {
+    message <- glue("{fn} has multiple sources\n")
+  }
+  return(message)
+}
+
+# Check date
+check_date <- function(file_text) {
+  message <- NULL
+  file_text %>%
+    str_subset("^date\\: ") %>%
+    str_remove(fixed("date: ")) %>%
+    str_remove_all('"') %>%
+    str_remove_all("'")  -> message
+  if (length(file_date)==0) {message <- glue("No date for {fn}")}
+  if (length(file_date)>1) {message <- glue("Multiple dates for {fn}")}
+  return(message)
+}
+
+# Check title
+
+# Update if no problems found
+date_list %>% 
+  str_subset(" for ") %>%
+  append(problem_files) -> problem_files
+  
+if (length(problem_files)==0) {
+  problem_files <- "All files passed date and source quality checks"
+}
+
[…]

### 2023-02-19 src/year-end-summary.R
Updated links from 2006 webpages to original sources

@@ -61,21 +61,12 @@ check_date <- function(file_text) {
     str_remove(fixed("date: ")) %>%
     str_remove_all('"') %>%
     str_remove_all("'")  -> message
-  if (length(file_date)==0) {message <- glue("No date for {fn}")}
-  if (length(file_date)>1) {message <- glue("Multiple dates for {fn}")}
+  if (length(message)==0) {message <- glue("No date for {fn}")}
+  if (length(message)>1) {message <- glue("Multiple dates for {fn}")}
   return(message)
 }
 
 # Check title
-
-# Update if no problems found
-date_list %>% 
-  str_subset(" for ") %>%
-  append(problem_files) -> problem_files
-  
-if (length(problem_files)==0) {
-  problem_files <- "All files passed date and source quality checks"
-}
 
 # Create links for each file
 create_file_list <- function(f, date_list) {
@@ -89,7 +80,9 @@ create_file_list <- function(f, date_list) {
 
 # Put it all together
 yyyy_list <- c(1999:2023)
-for (yyyy in yyyy_list[-(6:7)]) {
+n_problems <- 0
+n_files <- 0
+for (yyyy in yyyy_list) {
   yy <- str_sub(yyyy, 3, 4)
   output_file <- glue("text/{yy}/{yyyy}.md")
   glue("text/{yy}") %>%
@@ -103,11 +96,19 @@ for (yyyy in yyyy_list[-(6:7)]) {
     date_list %<>% append(check_date(file_text))
   }
   
-  if (length(problem_files)==0) {
+  # Update if no problems found
+  date_list %>% 
+    str_subset(" for ") %>%
+    append(problem_files) -> problem_files
+
+  n_year <- length(problem_files)  
+  n_problems <- n_problems + n_year
+  n_files <- n_files + length(f)
+  if (n_year==0) {
     problem_files <- "All files passed date and source quality checks"
   }
 
-  print(problem_files)
+  if (n_year > 0) {print(problem_files)}
 
   create_yaml(yyyy) %>% write_lines(output_file)
   write_lines(" ", output_file, append=TRUE)
@@ -124,5 +125,7 @@ for (yyyy in yyyy_list[-(6:7)]) {
   problem_files %>% write_lines(output_file, append=TRUE)
   # write_lines(" ", output_file, append=TRUE)
 
-  glue("{yyyy}.md created, ({length(f)} files)") %>% print
[…]

### 2023-02-19 src/year-end-summary.R
Fixed typo in year-end-summary.R

@@ -45,7 +45,7 @@ check_source <- function(file_text) {
     str_remove_all('"')  %>%
     str_remove_all("'") -> file_source
   if (length(file_source)==0) {
-    message <- glue("{fn} has no sources\n")
+    message <- glue("{fn} has no source\n")
   }
   if (length(file_source)>1) {
     message <- glue("{fn} has multiple sources\n")

### 2023-02-19 src/check-old-pages.R
Added first draft of check-old-pages.R

@@ -1,0 +1,47 @@
+# This program looks for key links in the old webpages
+
+# Preliminaries
+suppressMessages(suppressWarnings(library(glue)))
+suppressMessages(suppressWarnings(library(magrittr)))
+suppressMessages(suppressWarnings(library(readr)))
+suppressMessages(suppressWarnings(library(stringr)))
+
+# pad single digits with leading zero
+zpad <- function(x) {
+  ifelse(nchar(x)==1, paste0("0", x), x)
+}
+
+sub_dir_list <- c("99", zpad(0:12), "12a", zpad(13:18))
+
+path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
+
+key_text <- "<p>This page is moving to a <a href="
+alt_text <- "<p>This page is being phased out."
+n_files <- 0
+n_unlinked <- 0
+for (sub_dir in sub_dir_list) {
+  # List all html files in a specific folder
+  full_path <- glue("{path}/{sub_dir}/") 
+  full_path %>%
+    list.files %>%
+    str_subset("html$") -> html_list
+  i_files <- length(html_list)
+  i_unlinked <- 0
+  for (fn in html_list) {
+    # Read text from html files
+    file_text <- read_lines(glue("{full_path}/{fn}"))
+    # Search for key phrase
+    alt_line <- str_subset(file_text, alt_text)
+    if (length(alt_line) > 0) {next}
+    key_line <- str_subset(file_text, key_text)
+    if (length(key_line) > 0) {next}
+    # List files without key phrase
+    i_unlinked <- i_unlinked + 1
+    print(glue("{sub_dir}/{fn} has no link to new page"))
+  }
+  print(glue("{sub_dir} has {i_files} files. {i_unlinked} are unlinked."))
+  n_files <- n_files + i_files
+  n_unlinked <- n_unlinked + i_unlinked
+}
+print(glue("All directories have a total of  {n_files} files. {n_unlinked} are unlinked."))
+      

### 2023-02-20 src/copy_files.R
Updated copy_files.R to include 2023 files

@@ -15,7 +15,7 @@ source(file="src/standard_functions.R")
 copy_files <- function(
   old_path="text", 
   new_path="c:/Users/simons/Documents/b04/content",
-  old_dir_list=c("99", zpad(0:22)),
+  old_dir_list=c("99", zpad(0:23)),
   file_pattern="md$",
   qc=FALSE) {
 

### 2023-02-21 src/year-end-summary.R
Update year-end-summary.R

@@ -110,7 +110,12 @@ for (yyyy in yyyy_list) {
 
   if (n_year > 0) {print(problem_files)}
 
+  print("Check 1")
+  unlink(output_file)
+  file.remove(output_file)
+  print("Check 2")
   create_yaml(yyyy) %>% write_lines(output_file)
+  print("Check 3")
   write_lines(" ", output_file, append=TRUE)
   
   create_links(yyyy_list) %>% write_lines(output_file, append=TRUE)

### 2023-02-21 src/year-end-summary.R
Bug fix for year-end-summary.R

@@ -112,7 +112,7 @@ for (yyyy in yyyy_list) {
 
   print("Check 1")
   unlink(output_file)
-  file.remove(output_file)
+  # file.remove(output_file)
   print("Check 2")
   create_yaml(yyyy) %>% write_lines(output_file)
   print("Check 3")

### 2023-02-21 src/year-end-summary.Rmd
Convert year-end-summary to Rmd

---
title: "Create year end summaries"
author: "Steve Simon"
date: "Created 2022-05-26"
output: html_document
---

This program produces Markdown files summarizing every webpage written in a given year

```{r Preliminaries}
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(stringr)))

# pad single digits with leading zero
zpad <- function(x) {
  ifelse(nchar(x)==1, paste0("0", x), x)
}
```

```{r}
# Create yaml header
create_yaml <- function(yyyy) {
  glue(
    '---
     title: "Pages written in 2022"
     source: new
     date: {yyyy}-12-31
     categories:
     - Blog post
     tags:
     - Website details
     output: html_document
     ---')
}
```

```{r}
# Create links to other years
create_links <- function(yyyy_range) {
  yy_range <- str_sub(yyyy_range, 3)
  glue("[{yyyy_range}](http://new.pmean.com/{yyyy_range}/)") %>%
    paste(collapse=" | ")
}
```

```{r}
# Check for problems

# Check source
check_source <- function(file_text) {
  message <- NULL
  file_text %>%
    str_subset("^source\\: ") %>%
    str_remove(fixed("source: ")) %>%
    str_remove_all('"')  %>%
    str_remove_all("'") -> file_source
  if (length(file_source)==0) {
    message <- glue("{fn} has no source\n")
  }
  if (length(file_source)>1) {
    message <- glue("{fn} has multiple sources\n")
  }
  return(message)
}

# Check date
check_date <- function(file_text) {
  message <- NULL
  file_text %>%
    str_subset("^date\\: ") %>%
    str_remove(fixed("date: ")) %>%
    str_remove_all('"') %>%
    str_remove_all("'")  -> message
  if (length(message)==0) {message <- glue("No date for {fn}")}
  if (length(message)>1) {message <- glue("Multiple dates for {fn}")}
  return(message)
}

# Check title

# Create links for each file
create_file_list <- function(f, date_list) {
  f %>%
    str_replace(fixed(".Rmd"), "") %>%
    str_replace(fixed(".md"), "") -> short_name
  o <- order(date_list, f)
  glue("+ {date_list[o]} [{short_name[o]}](http://new.pmean.com/{short_name[o]}/)  ") %>%
    paste(collapse="  \n")
}
```

```{r}
# Put it all together
yyyy_list <- c(1999:2023)
n_problems <- 0
n_files <- 0
for (yyyy in yyyy_list) {
  yy <- str_sub(yyyy, 3, 4)
  output_file <- glue("text/{yy}/{yyyy}.md")
  glue("text/{yy}") %>%
    list.files %>%
    str_subset("md$") -> f
  problem_files <- NULL
  date_list <- NULL
  for (fn in f) {
    file_text <- read_lines(glue("text/{yy}/{fn}"))
    problem_files %<>% append(check_source(file_text))
    date_list %<>% append(check_date(file_text))
  }

  # Update if no problems found
  date_list %>% 
    str_subset(" for ") %>%
    append(problem_files) -> problem_files

  n_year <- length(problem_files)  
  n_problems <- n_problems + n_year
  n_files <- n_files + length(f)
  if (n_year==0) {
    problem_files <- "All files passed date and source quality checks"
  }

  if (n_year > 0) {print(problem_files)}

  print("Check 1")
  unlink(output_file)
  # file.remove(output_file)
  print("Check 2")
  create_yaml(yyyy) %>% write_lines(output_file)
  print("Check 3")
  write_lines(" ", output_file, append=TRUE)

  create_links(yyyy_list) %>% write_lines(output_file, append=TRUE)
  write_lines(" ", output_file, append=TRUE)

  glue("New website ({length(f)} files)") %>% write_lines(output_file, append=TRUE)
  write_lines(" ", output_file, append=TRUE)

  create_file_list(f, date_list) %>% write_lines(output_file, append=TRUE)
  write_lines(" ", output_file, append=TRUE)

  problem_files %>% write_lines(output_file, append=TRUE)
  # write_lines(" ", output_file, append=TRUE)

  glue("{yyyy}.md created, {length(f)} files, {n_year} problems detected") %>% print
}

print(glue("All years: {n_problems} problems identified in {n_files} files"))
```

### 2023-02-21 src/check_links.Rmd
Reformatted tags in yaml headers

@@
 ```{r read_old}
+html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
 html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
-html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
 html_pattern <- "html$"

### 2023-02-21 src/year-end-summary.Rmd
Reformatted tags in yaml headers

@@
 }
+
+update <- FALSE
+print("Check 1")
 ```
@@
 for (yyyy in yyyy_list) {
+  cat("\n")
+  cat(yyyy)
+  cat("\n\n")
   yy <- str_sub(yyyy, 3, 4)
-  output_file <- glue("text/{yy}/{yyyy}.md")
-  glue("text/{yy}") %>%
+  output_file <- glue("../text/{yy}/{yyyy}.md")
+  glue("../text/{yy}") %>%
     list.files %>%
@@
   for (fn in f) {
-    file_text <- read_lines(glue("text/{yy}/{fn}"))
+    file_text <- read_lines(glue("../text/{yy}/{fn}"))
     problem_files %<>% append(check_source(file_text))
@@
   if (n_year > 0) {print(problem_files)}
+  print(problem_files)
 
-  print("Check 1")
-  unlink(output_file)
-  # file.remove(output_file)
-  print("Check 2")
+  if (update==FALSE) {next}
+
   create_yaml(yyyy) %>% write_lines(output_file)
-  print("Check 3")
   write_lines(" ", output_file, append=TRUE)
@@
   problem_files %>% write_lines(output_file, append=TRUE)
-  # write_lines(" ", output_file, append=TRUE)
 

### 2023-02-21 text/04/early-stopping-animal.Rmd
Reformatted tags in yaml headers

@@
 ---
+
 I was helping someone with power calculations for an animal experiment.
-There were several ways to estimate power and they all seemed to point
-to the use of either 8 animals per group or 12 animals per group. I
-explained that if she was more risk averse, she should use 12 animals
-and if she were more work averse, she should use 8 animals. Her answer
-surprised me a bit. She said that she liked the idea of using 12 animals
-per group. That way, if the results achieved significance after only 8
-animals, she could go back to the Animal Care and Use Committee and tell
-that she saved a few animals.
+There were several ways to estimate power and they all seemed to point to the use of either 8 animals per group or 12 animals per group.
+I explained that if she was more risk averse, she should use 12 animals and if she were more work averse, she should use 8 animals.
+Her answer surprised me a bit.
+She said that she liked the idea of using 12 animals per group.
+That way, if the results achieved significance after only 8 animals, she could go back to the Animal Care and Use Committee and tell that she saved a few animals.
 
-This is an example of early stopping of a clinical trial (also described
-as an interim analysis). Generally, if you think you might want to stop
-a trial early, you should specify the conditions in your protocol prior
-to the start of data collection. Without such a rule, any examination of
-the data prior to the end of the study is a protocol deviation. As such,
-it should be reported to the IRB (if it is human research) or the ACUC
-(if it is animal research).
+This is an example of early stopping of a clinical trial (also described as an interim analysis).
+Generally, if you think you might want to stop a trial early, you should specify the conditions in your protocol prior to the start of data collection.
+Without such a rule, any examination of the data prior to the end of the study is a protocol deviation.
+As such, it should be reported to the IRB (if it is human research) or the ACUC (if it is animal research).
 
-There are reasonable examples of why you might want to stop a study
-early even without planning ahead for such an event. For example, if a
-serious unexpected side effect appears, you might not want to offer the
-new drug to any more patients.
+There are reasonable examples of why you might want to stop a study early even without planning ahead for such an event.
+For example, if a serious unexpected side effect appears, you might not want to offer the new drug to any more patients.
 
-But stopping the study early (like any other protocol deviation) does
-reduce the credibility of the research findings, sometimes by a little
-bit, sometimes by a lot.
+But stopping the study early (like any other protocol deviation) does reduce the credibility of the research findings, sometimes by a little bit, sometimes by a lot.
 
@@
 
-In the graph shown above the test never quite reaches statistical
-significance.
+In the graph shown above the test never quite reaches statistical significance.
 
@@
 
-Here's one where it just reaches statistical significance at the last
-sample.
+Here's one where it just reaches statistical significance at the last sample.
 
@@
 
-Here's one where it reaches statistical significance rather early. You
-would think that it would be nice to stop the experiment early here, but
-there is no guarantee that just because the p-value is below the
-significant level early on, that it will stay below that level the rest
-of the way.
+Here's one where it reaches statistical significance rather early.
+You would think that it would be nice to stop the experiment early here, but there is no guarantee that just because the p-value is below the significant level early on, that it will stay below that level the rest of the way.
 
@@
 
-Notice that the above graph shows statistical significance early, but
-then that significance disappears by the end of the experiment. Now this
-doesn't happen very often. Out of 100 graphs only 4 or 5 showed this
-type of pattern. Most of the time if the p-value crosses below the 0.05
-threshold early, it stays below that threshold. But these exceptions
-serve as a warning. If you stop the study early and make no adjustments
-to your p-value, your results may be invalid.
+Notice that the above graph shows statistical significance early, but then that significance disappears by the end of the experiment.
+Now this doesn't happen very often.
+Out of 100 graphs only 4 or 5 showed this type of pattern.
+Most of the time if the p-value crosses below the 0.05 threshold early, it stays below that threshold.
+But these exceptions serve as a warning.
+If you stop the study early and make no adjustments to your p-value, your results may be invalid.
 
-The adjustments to the p-value (and confidence interval) are rather
-complex. Here's some more S-plus (R) code that shows some of the
-complexities of the situation. This program simulates an experiment with
-12 animals per group, but looks at the p-value with after 4, 8, and 12
-animals have been tested.
+The adjustments to the p-value (and confidence interval) are rather complex.
+Here's some more S-plus (R) code that shows some of the complexities of the situation.
+This program simulates an experiment with 12 animals per group, but looks at the p-value with after 4, 8, and 12 animals have been tested.
 
@@
 
-Notice that each p-value considered in isolation maintains an alpha
-level of approximately 0.05. But when you combine the p-values and allow
-for statistical significance when either the n=4, n=8, or n=12 p-value
-achieves significance, the alpha level more than doubles.
+Notice that each p-value considered in isolation maintains an alpha level of approximately 0.05.
+But when you combine the p-values and allow for statistical significance when either the n=4, n=8, or n=12 p-value achieves significance, the alpha level more than doubles.
 
@@
 
-Ludbrook J. Interim analyses of data as they accumulate in laboratory
-experimentation. BMC Med Res Methodol 2003: 3(1); 15.
-[Medline](http://www.ncbi.nlm.nih.gov/entrez/query.fcgi?cmd=Retrieve&db=PubMed&list_uids=12930561&dopt=Abstract)
-[Abstract](http://www.biomedcentral.com/1471-2288/3/15/abstract)
-Avialable in [html format](http://www.biomedcentral.com/1471-2288/3/15)
-or [pdf format](http://www.biomedcentral.com/content/pdf/1471-2288-3-15.pdf)
+Ludbrook J.
+Interim analyses of data as they accumulate in laboratory experimentation.
+BMC Med Res Methodol 2003: 3(1); 15.
+Available in [html format](http://www.biomedcentral.com/1471-2288/3/15) or [pdf format](http://www.biomedcentral.com/content/pdf/1471-2288-3-15.pdf)
 
-+ The Ludbrook reference points out that laboratory
-experiments often fail to account properly for early stopping
-(interim analysis) and suggests an approach for handling this
-properly.
++ The Ludbrook reference points out that laboratory experiments often fail to account properly for early stopping (interim analysis) and suggests an approach for handling this properly.
 
-You can find an [earlier version](http://www.pmean.com/04/EarlyStoppingAnimal.html) of this page on my [original website](http://www.pmean.com/original_site.html).
+You can find an [earlier version][sim1] of this page on my [original website][sim2].
+
+[sim1]: http://www.pmean.com/04/EarlyStoppingAnimal.html
+[sim2]: http://www.pmean.com/original_site.html

### 2023-02-21 text/14/axis.Rmd
Reformatted tags in yaml headers

@@
 category: Blog post
-tags: Sample size
+tags:
+- Sample size justification
 output: html_document

### 2023-02-21 text/15/difference-of-two-exponentials.Rmd
Reformatted tags in yaml headers

@@
 ---
-title: "PMean: Nonlinear regression for the difference of two exponentials"
+title: "Nonlinear regression for the difference of two exponentials"
 author: "Steve Simon"

### 2023-02-21 text/15/missing-values-in-r.Rmd
Reformatted tags in yaml headers

--- 2021-11-12 text/15/missing-values-in-r.Rmd
@@
 ---
-title: "PMean: Missing values in R talk"
+title: "Missing values in R talk"
 author: "Steve Simon"

### 2023-02-21 text/15/relationships-in-r.Rmd
Reformatted tags in yaml headers

@@
 ---
-title: "PMean: Examining relationships in R"
+title: "Examining relationships in R"
 author: "Steve Simon"

### 2023-02-21 text/22/analysis-of-means.Rmd
Reformatted tags in yaml headers

@@
 category: Blog post
-tags: Analysis of means
+tags:
+- Analysis of means
 source: new

### 2023-02-21 text/22/box-with-no-whiskers.Rmd
Reformatted tags in yaml headers

@@
 category: Blog post
-tags: Graphical display
+tags:
+- Graphical display
 source: new

### 2023-02-21 text/22/definition-schoenfeld-residuals.Rmd
Reformatted tags in yaml headers

@@
 category: Blog post
-tags: Survival analysis
+tags:
+- Survival analysis
 source: new

### 2023-02-21 text/22/posix-time.Rmd
Reformatted tags in yaml headers

@@
 category: Blog post
-tags: Data science
+tags:
+- Data science
 source: new

### 2023-02-21 src/clean_files.R

@@ -74,7 +74,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  fixed('http://www.stata-press.com/data/lf2/couart2.dta'),
-  'http://www.pmean.com/00files/couart2.dta',
+  'tags: ',
+  'tags:\n- ',
   dir_root="text", 
   file_pattern="md$")

### 2023-02-21 src/clean_files.R
Added early-micro-array-pages

@@ -74,7 +74,6 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  'tags: ',
-  'tags:\n- ',
+  fixed('(Being updated)'),
   dir_root="text", 
   file_pattern="md$")

### 2023-02-22 src/year-end-summary.Rmd
Fixed duplicate listing in year-end-summary.Rmd

@@
     '---
-     title: "Pages written in 2022"
+     title: "Pages written in {yyyy}"
      source: new
@@
   if (n_year > 0) {print(problem_files)}
-  print(problem_files)
 

### 2023-02-24 src/check-old-pages.Rmd
Converted check-old-pages to RMarkdown

---
title: "Check links on old pages"
author: "Steve Simon"
date: "Created 2022-02-20"
output: html_document
---

This program looks for key links in the old webpages

```{r setup}
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(stringr)))

# pad single digits with leading zero
zpad <- function(x) {
  ifelse(nchar(x)==1, paste0("0", x), x)
}
```

```{r find-files}
sub_dir_list <- c("99", zpad(0:12), "12a", zpad(13:18))

path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website"
```

```{r search-files}
key_text <- "<p>This page"
n_files <- 0
n_unlinked <- 0
for (sub_dir in sub_dir_list) {
  # List all html files in a specific folder
  full_path <- glue("{path}/{sub_dir}/") 
  full_path %>%
    list.files %>%
    str_subset("html$") -> html_list
  i_files <- length(html_list)
  i_unlinked <- 0
  for (fn in html_list) {
    # Read text from html files
    file_text <- read_lines(glue("{full_path}/{fn}"))
    # Search for key phrase
    key_line <- str_subset(file_text, key_text)
    if (length(key_line) > 0) {next}
    # List files without key phrase
    i_unlinked <- i_unlinked + 1
    print(glue("{sub_dir}/{fn} has no link to new page"))
  }
  cat("\n")
  print(glue("{sub_dir} has {i_files} files. {i_unlinked} are unlinked."))
  cat("\n")
  n_files <- n_files + i_files
  n_unlinked <- n_unlinked + i_unlinked
}
print(glue("All directories have a total of  {n_files} files. {n_unlinked} are unlinked."))
```

### 2023-02-24 src/year-end-summary.Rmd

@@
   if (length(file_source)==0) {
-    message <- glue("{fn} has no source\n")
+    message <- glue("{yy}/{fn} has no source\n")
   }
   if (length(file_source)>1) {
-    message <- glue("{fn} has multiple sources\n")
+    message <- glue("{yy}/{fn} has multiple sources\n")
   }
@@
     str_remove_all("'")  -> message
-  if (length(message)==0) {message <- glue("No date for {fn}")}
-  if (length(message)>1) {message <- glue("Multiple dates for {fn}")}
+  if (length(message)==0) {message <- glue("{yy}/{fn} has no date")}
+  if (length(message)>1) {message <- glue("{yy}/{fn} has multiple dates")}
   return(message)
@@
 for (yyyy in yyyy_list) {
-  cat("\n")
-  cat(yyyy)
-  cat("\n\n")
   yy <- str_sub(yyyy, 3, 4)
@@
 
+  cat("\n")
+  cat(glue("{yyyy} has a total of {length(f)} files."))
+  cat("\n\n")
+
   if (update==FALSE) {next}

### 2023-02-24 src/check-old-pages.Rmd
Moved copyright to 00 folder

@@
 author: "Steve Simon"
-date: "Created 2022-02-20"
+date: "Created 2023-02-20"
 output: html_document
@@
 
-path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
+# path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
 path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website"

### 2023-02-26 src/check-old-pages.Rmd
Added 11/harrell-quotations

@@
 
-# path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
-path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website"
+path1 <- "c:/Users"
+path2 <- "dropbox/professional/fun stuff/update-website"
+
+list.files(path1) %>%
+  str_subset("(Simons)|(steve)") -> me
+
+path3 <- glue("{path1}/{me}/{path2}")
+
+print(glue("Searching in {path3}"))
+
+list.files(path3)
 ```
@@
   # List all html files in a specific folder
-  full_path <- glue("{path}/{sub_dir}/") 
+  full_path <- glue("{path3}/{sub_dir}/") 
   full_path %>%

### 2023-02-28 text/06/bayesian-accrual-model.Rmd
Started work on gamma hedging prior

---
title: A simple Bayesian model for accrual
author: Steve Simon
source: http://www.pmean.com/06/BayesianAccrualModel.html
date: 2006-11-17
categories:
- Blog post
tags:
- Accrual problems in clinical trials
- Bayesian statistics
output: html_document
---

Suppose you are a researcher in charge of a long term study. You plan to collect data on 120 patients. The goal is to finish your study in ten years, which means getting 12 patients per year or one every thirty days on average. Recruiting patients though appears to be harder than you had expected. You recruited your first patient on day 56, 26 days behind schedule. The second patient is not recruited until day 93. About two years into the study (day 768), you have just recruited your 10th patient. It looks like recruitment might be behind schedule. Is it time to take action?

A Bayesian model of accrual times can help you to discern whether recruitment is behind schedule and project an estimated completion date allowing for uncertainty. The Bayesian model starts with an elicitation of prior beliefs about accrual rates. Then you combine the prior belief with the actual accrual data and compute a posterior distribution of accrual times. You can then use the posterior distribution to make meaningful probability statements about the current accrual rate and the projected completion time.

Elicitation of prior beliefs is difficult in most problems, but should be somewhat easier for accrual. You would not embark on a research project if you did not have at least some idea of how many patients would meet your inclusion criteria and how many would agree to participate.

You can model the accrual process as a series of waiting times between patients. A simple distribution for waiting times is the exponential distribution. The probability density function for the exponential distribution is quite simple and easy to work with.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel01.gif)

The value lambda ($\lambda$) of this distribution is of great interest, because it will tell you how quickly the trial will end. If lambda is equal to 30, then the average waiting time will be 30 days, exactly what you were hoping for.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel02.gif)

The 5th and 95th percentiles would be 1.5 days and 90 days respectively (see the gray shading in the picture above). This wide range is quite natural for the exponential distribution, and in many applications, the exponential distribution is a reasonably accurate approximation. If you expected to see a narrower range in waiting times, then a different distribution, such as the gamma distribution, might be a better choice.

An exponential distribution with a larger lambda value would indicate an accrual process that is much slower.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel03.gif)

Shown above is a graph of the exponential distribution with lambda=60. For this distribution, the mean is 60 days and the 5th and 95th percentiles are 3.0 and 180, respectively.

Determining the value of lambda, therefore, is critical in being able to recognize if the accrual rate is satisfactory or not. With a Bayesian model for accrual, you develop a prior distribution for lambda, and combine this with data from this clinical trial. Each time you get a new patient, you add a waiting time to the data and update the prior distribution to get a better feel for the accrual process.

There are a wide range of attractive prior distributions for accrual, but the inverse gamma distribution is attractive because of its close relationship to the exponential distribution. The probability density function for the inverse gamma distribution is

![](http://www.pmean.com/new-images/06/BayesianAccrualModel04.gif)

The mean and variance of the inverse gamma distribution are

![](http://www.pmean.com/new-images/06/BayesianAccrualModel05.gif)

The second factor in the formula for the standard deviation represents the coefficient of variation (CV) also known as the relative standard deviation. Notice that as alpha increases, the CV decreases.

If you accumulate n waiting times

- x~1~
- x~2~
- ...
- x~n~

then the likelihood function is

![](http://www.pmean.com/new-images/06/BayesianAccrualModel06.gif)

The posterior distribution is

![](http://www.pmean.com/new-images/06/BayesianAccrualModel07.gif)

which is proportional to an inverse gamma distribution. The parameters of the inverse gamma distribution are

![](http://www.pmean.com/new-images/06/BayesianAccrualModel08.gif)

The posterior mean is

![](http://www.pmean.com/new-images/06/BayesianAccrualModel09.gif)

which can be written as a weighted average of the prior mean and the mean of the data.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel10.gif)

with greater weight being given to the mean of the data as the sample size increases. Let's examine some possible prior distributions that you might use.

The value of alpha indicates the degree of certainty that you have, prior to collecting the data, about the accrual rates. Deciding how much certainty you have is a bit of a challenge, but the folks who work regularly with Bayesian models have developed a variety of approaches to eliciting this information.

Once you have specified the value for alpha, you can set the value of beta so that the expected value of the distribution is equal to your expected accrual rate (30 in this case).

If you set set alpha to a small value, such as 2, this reflects the fact that you are unable or unwilling to pin down the accrual rate. A value of beta = 30 will yield the distribution shown below.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel11.gif)

The gray area represents the range of the 5th percentile (6.3 days) to the 95th percentile (84.4 days). These rates represent a broad range of accrual times, from a very fast accrual time that would allow the study to finish in an average of 2.1 years to an extremely slow accrual time that would require an astounding 28.1 years average time to completion.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel12.gif)

If you set alpha to 10 and beta to 270, you get the distribution shown above. It has the same expected value (30 days), but the 5th and 95th percentiles are closer (17.2 days and 49.8 days, respectively). These values correspond to average completion times of 5.7 years to 16.6 years.

Finally, someone who was very confident about accrual rates (perhaps because this trial is a minor variation on an existing trial) might let alpha = 50 and beta = 1470. The 5th and 95th percentiles are 23.6 days and 37.7 days, respectively. This leads to a clinical trial that would finish up, on average, in somewhere between 7.9 and 12.6 years.

Let's choose the first prior and apply it to the existing accrual times. On day 768, you have just gotten your 10th patient. With a prior distribution with alpha = 2 and beta = 30, the posterior distribution has alpha = 2+ 10 = 12 and beta = 30 +768 = 798.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel13.gif)

With only ten patients, you do not have a very precise estimate of the accrual rate. The estimated mean accrual rate is 72.5 days. That is slightly lower than the rate based on the data alone, because the prior distribution does contain some information about the accrual rate that is incorporated into the posterior mean.

The 5th and 95th percentiles are 43.8 and 115.2 days, respectively. This is extremely wide and reminds you that we have limited data with only ten patients and limited information from the prior distribution. You do, though, have more than enough information to conclude that this trial is seriously in trouble, since even the 5th percentile is well above your target of 30 days.

Let's choose the last prior distribution and apply it to the existing accrual times. On day 768, you have just gotten your 10th patient. Since you specified alpha = 50 and beta = 1470, you have a posterior distribution with alpha = 50+10 = 60 and beta=1470 + 768 = 2238.

Note that the average of this posterior distribution is 37.9 days and the 5th and 95th percentiles are 30.5 and 46.8 respectively. The strong prior belief outweighs the evidence from just 10 observations, but even here you can recognize problems. The 5th percentile for accrual rates is slightly above your target of 30 days in spite of your strong optimistic prior belief.

With the Bayesian models, you can also make predictions about future unobserved variables. You know that the waiting time is exponential, but you are not 100% certain about the parameter lambda in this distribution. With the first prior, it could easily be as small as 43.8 or as large as 115.2.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel14.gif)

The graph above shows what an exponential distribution looks like with lambda = 43.8, lambda = 115.2, and everything in between.

With the last prior distribution, the value of lambda could easily be as small as 30.5 or as large as 46.8.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel15.gif)

The graph shown above indicates what an exponential distribution looks like with lambda = 30.5, lambda = 46.8, and everything in between.

To calculate a predictive distribution, you need to compute a weighted average of exponential distributions with weights determined by the posterior distribution. This is accomplished through a simple integration. The calculations are more tedious than difficult.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel16.gif)

The density function is closely related to a Pareto distribution. For large values of alpha, it is not too far from an exponential distribution. The expected value and standard deviation of this distribution are

![](http://www.pmean.com/new-images/06/BayesianAccrualModel17.gif)

Notice the square root term in the standard deviation formula. This again represents the coefficient of variation. It is always larger than 1, but approaches 1 as alpha increases. The CV for an exponential distribution is exactly 1. So this formula reminds you that an exponential distribution with uncertainty in the parameter lambda leads to an increase in uncertainty.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel18.gif)

The graph above shows what this predictive distribution looks like for the first prior distribution (black line). There is a small deviation from this distribution and the exponential distribution (red line).

![](http://www.pmean.com/new-images/06/BayesianAccrualModel19.gif)

The graph above shows what the predictive distribution looks like for the last prior distribution. There is almost no visible difference between the predictive distribution and the exponential distribution.

If you assume that the remaining 110 patients accrue at an exponential rate, then the sum of these accrual times will be distributed as a gamma distribution. The density function of a gamma distribution is

![](http://www.pmean.com/new-images/06/BayesianAccrualModel20.gif)

which has mean and standard deviation equal to

![](http://www.pmean.com/new-images/06/BayesianAccrualModel21.gif)

Set theta to the posterior mean and k to the number of patients remaining. With the first prior distribution, set theta = 72.5 and k = 110. The mean of the remaining time in the study is 7,975 days (21.8 years) and the standard deviation is 760.4 days or 2.1 years.   The 5th and 95th percentiles are 18.5 years and 25.4 years respectively.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel22.gif)

For the last prior distribution, the predictive distribution for the remainder of time would have theta = 39.7 and k = 110.

![](http://www.pmean.com/new-images/06/BayesianAccrualModel23.gif)

The 5th and 95th percentiles are 9.7 and 13.3 years. Since we have already spent 2.1 years in the study, you can state that there is a fairly good chance that the full study will take at least 11.8 years rather than the planned 10 years.

If the number of patients remaining in the study is large, you could also rely on the central limit theorem. With the first prior distribution, the mean and standard deviation of the predictive distribution for a future waiting time are 72.5 and 79.4. The sum of 110 such variables would be approximately normal with mean = $72.5 \times 110 = 7,975$ days or 21.8 years and the standard deviation would be $79.4 \times \sqrt{110} = 832.8$ days or 2.3 years. The 5th and 95th percentiles of a normal distribution with this mean and variance would be 18.0 and 25.6 years, respectively.

You can find an [earlier version][sim1] of this page on my [website][sim2].

[sim1]: http://www.pmean.com/06/BayesianAccrualModel.html
[sim2]: http://www.pmean.com

### 2023-02-28 text/23/gamma-hedging-hyperprior.Rmd
Started work on gamma hedging prior

---
title: "The hedging hyperprior for the exponential distribution"
source: "New"
date: "2023-02-28"
categories:
- Blog post
tags:
- Bayesian statistics
output: html_document
---

I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the gamma distribution.

<!---more--->

Consider a Bayesian analysis of an outcome variable that has an exponential distribution. The density of the gamma distribution is

$f(x) = \theta^{-1} e^{-x / \theta}$

You have an informative prior on $\theta$ that has an inverse gamma distribution

$g(\theta; \alpha, \beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}$

and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of

$\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$

and a posterior distribution of

$h(\theta | \mathbf{X}) \propto \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}\theta^{-n} e^{-n \bar X / \theta}$

which simplifies to

$h(\theta | \mathbf{X}) \propto \theta^{-\alpha-1-n}e^{-(\beta + n\bar X) / \theta}$

which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean

$\frac{\beta + n\bar X}{\alpha-1+n}$

can be written as a weighted average of the prior mean and the mean of the observed data.

$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$

Let's illustrate this with an informative prior with $\alpha$ = 10 and $\beta$ = 27.

This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", 1/qgamma(0.95, 10, 27))` and `r sprintf("%.2f", 1/qgamma(0.05, 10, 27))` with 95% probability. If you observe 30 values of x with a mean of 7.7.

### 2023-03-04 text/23/gamma-hedging-hyperprior.Rmd
Started work on modified- power-prior

@@
 
+```{r setup}
+suppressMessages(suppressWarnings(library(tidyverse)))
+knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=2)
+```
+
 I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the gamma distribution.
@@
 <!---more--->
+
+### A simple Baysian analysis with an informative prior.
 
@@
 
-$\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$
+$L(\mathbf{X})=\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$
 
@@
 
-Let's illustrate this with an informative prior with $\alpha$ = 10 and $\beta$ = 27.
+```{r}
 
-This corresponds to a fairly informative prior distribution.
-This distribution places $\theta$ within `r sprintf("%.2f", 1/qgamma(0.95, 10, 27))` and `r sprintf("%.2f", 1/qgamma(0.05, 10, 27))` with 95% probability.
-If you observe 30 values of x with a mean of 7.7.
+a0 <- 10
+b0 <- 30
+
+n <- 30
+xbar <- 0.7
+
+a1 <- a0 + n
+b1 <- b0 + n*xbar
+
+posterior_alpha <- a0+n
+posterior_beta <- b0+n*xbar
+
+x_grid <- seq(0.01, 8, by=0.01)
+
+f <- function(x, theta) {
+  theta^(-1) * 
+    exp(-x/theta)
+}
+
+g <- function(theta, alpha, beta) {
+  beta^alpha/gamma(alpha) *
+    theta^(-alpha-1) * 
+    exp(-beta/theta)
+}
+
+prior05 <- 1/qgamma(c(0.95), a0, b0) 
+prior95 <- 1/qgamma(c(0.05), a0, b0)
+
+post05 <- 1/qgamma(c(0.95), a1, b1) 
+post95 <- 1/qgamma(c(0.05), a1, b1)
+
+```
+
+Let's illustrate this with an informative prior with $\alpha$ = `r a0` and $\beta$ = `r b0`.
+
+```{r}
+g_grid <- g(x_grid, a0, b0)
+sum(g_grid)*0.01 # Check the g is a density.
+outside <- x_grid < prior05 | x_grid > prior95
+x_color <- 
+  ifelse(outside, "white", "gray")
+data.frame(x=x_grid, y=g_grid) %>%
+  ggplot(aes(x, y)) +
+    geom_segment(aes(xend=x, yend=0), color=x_color) +
+    geom_line() + xlab(" ") + ylab(" ")
+```
+
+This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.
+
+Let's see what happens with this informative prior if you observe `r n` values of x with a mean of `r xbar`. The likelihood is
+
+```{r}
+f_grid <- f(xbar, x_grid)^10
+x_color <- "white"
+data.frame(x=x_grid, y=f_grid) %>%
+  ggplot(aes(x, y)) +
+    geom_line() + xlab(" ") + ylab(" ")
+```
+
+Multiply these two together to get the posterior distribution.
+
+```{r}
+h_grid <- g(x_grid, a1, b1)
+outside <- x_grid < post05 | x_grid > post95
+x_color <- 
+  ifelse(outside, "white", "gray")
+data.frame(x=x_grid, y=h_grid) %>%
+  ggplot(aes(x, y)) +
+    geom_segment(aes(xend=x_grid, yend=0), color=x_color) +
+    geom_line() + xlab(" ") + ylab(" ")
+```
+
+The posterior distribution is not close to either the prior distribution or the likelihood.
+It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
+It excludes the prior mean of `r sprintf("%.2f", b0/(a0-1))` and it excludes `r xbar`, the mean of the data.
+
+### The power prior
+
+The power prior was developed to combat the excessive influence of an informative prior when the data conflicts with that prior.
+
+Instead of the standard Bayesian approach where you multiply the Likelihood (L) by the prior (g), you use
+
+$h(\theta|\mathbf{X}) \propto L(\mathbf{X}|\theta)g(\theta)^\tau g_0(theta)$
+
+where $g_0$ is a non-informative prior and $\tau$ is between 0 and 1.
+
+For $\tau$=1, the prior is used at full strength.
+When $\tau$=0, the informative prior effectively disappears, leaving you with just the non-informative prior.
+For values of $\tau$ between 0 and 1, the informative prior is weakened, but still exerts some pull on the posterior distribution.
+
+For intermediate values of $\tau$, the formulation is not consistent with Bayes rule and you need to include a "fudge factor" of
+
+$\int g(\theta)^{\tau}g_0(\theta)d\theta$
+
+A simpler approach is to include $\tau$ in the prior distribution and then place a hyperprior on $\tau$.
+
+The parameter $\alpha$ in an inverse gamma distribution can be thought of as a prior sample size.
+Large values of $\alpha$ produce informative priors that are equivalent to adding $\alpha$ pseudo-data values.
+If you multiply $\alpha$ by $\tau$, you can reduce the amount of pseudo data being added.
+
+### The hedging hyperprior

### 2023-03-04 text/23/modified-power-prior.Rmd
Started work on modified- power-prior

--- 2023-02-28 text/23/gamma-hedging-hyperprior.Rmd
@@
 ---
-title: "The hedging hyperprior for the exponential distribution"
+title: "The modified power prior"
 source: "New"
-date: "2023-02-28"
+date: "2023-03-04"
 categories:
@@
 
-I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the gamma distribution.
+```{r setup}
+suppressMessages(suppressWarnings(library(tidyverse)))
+knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=2)
+```
+
+I want to write a paper about the hedging hyperprior. Here is a brief description of an approach it is comparable to, the modified power prior.
 
@@
 
-Consider a Bayesian analysis of an outcome variable that has an exponential distribution. The density of the gamma distribution is
+### Informative prior distributions
 
-$f(x) = \theta^{-1} e^{-x / \theta}$
+*Justify the use of informative priors.*
 
-You have an informative prior on $\theta$ that has an inverse gamma distribution
+### The power prior
 
-$g(\theta; \alpha, \beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}$
+The power prior was developed to combat the excessive influence of an informative prior when the data conflicts with that prior.
 
-and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of
+Instead of the standard Bayesian approach where you multiply the Likelihood (L) by the prior (g), you use
 
-$\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$
+$h(\theta|\mathbf{X}) \propto L(\mathbf{X}|\theta)g(\theta)^\tau g_0(theta)$
 
-and a posterior distribution of
+where $g_0$ is a non-informative prior and $\tau$ is between 0 and 1.
 
-$h(\theta | \mathbf{X}) \propto \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}\theta^{-n} e^{-n \bar X / \theta}$
+For $\tau$=1, the prior is used at full strength.
+When $\tau$=0, the informative prior effectively disappears, leaving you with just the non-informative prior.
+For values of $\tau$ between 0 and 1, the informative prior is weakened, but still exerts some pull on the posterior distribution.
 
-which simplifies to
+For intermediate values of $\tau$, the formulation is not consistent with Bayes rule and you need to include a "fudge factor" of
 
-$h(\theta | \mathbf{X}) \propto \theta^{-\alpha-1-n}e^{-(\beta + n\bar X) / \theta}$
+$\int g(\theta)^{\tau}g_0(\theta)d\theta$
 
-which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean
+A simpler approach is to include $\tau$ in the prior distribution and then place a hyperprior on $\tau$.
 
-$\frac{\beta + n\bar X}{\alpha-1+n}$
+The parameter $\alpha$ in an inverse gamma distribution can be thought of as a prior sample size.
+Large values of $\alpha$ produce informative priors that are equivalent to adding $\alpha$ pseudo-data values.
+If you multiply $\alpha$ by $\tau$, you can reduce the amount of pseudo data being added.
 
-can be written as a weighted average of the prior mean and the mean of the observed data.
-
-$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$
-
-Let's illustrate this with an informative prior with $\alpha$ = 10 and $\beta$ = 27.
-
-This corresponds to a fairly informative prior distribution.
-This distribution places $\theta$ within `r sprintf("%.2f", 1/qgamma(0.95, 10, 27))` and `r sprintf("%.2f", 1/qgamma(0.05, 10, 27))` with 95% probability.
-If you observe 30 values of x with a mean of 7.7.
+### The hedging hyperprior

### 2023-03-04 src/check_links.Rmd
Update 2013 pages

@@
 ```{r read_html}
-html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
-html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
+path_prefix <- "c:/Users"
+path_suffix <- "dropbox/professional/fun stuff/update-website"
+
+list.files(path_prefix) %>%
+  str_subset("(Simons)|(steve)") -> me
+
+full_path <- glue("{path_prefix}/{me}/{path_suffix}")
+
+print(glue("Searching in {full_path}"))
+
+list.files(full_path)
+
 subdir_list=c("99", paste0("0", 0:9), 10:22)
@@
 for (yr in subdir_list) {
-  file_path <- paste0(html_path, yr)
+  file_path <- glue("{full_path}/{yr}")
   list.files(file_path, pattern="html$") -> yr_files
@@
 sim1 <- "[sim1]: http://www.pmean.com/"
-full_file_name <- paste0(md_path, md_files$yr, "/", md_files$fn)
+full_file_name <- glue("{full_path}/{md_files$yr}/{md_files$fn}")
 md_links <- NULL

### 2023-03-06 text/23/gamma-hedging-hyperprior.Rmd
Added contour plot to gamma-hedging-hyperprior

@@
 ```{r setup}
+suppressMessages(suppressWarnings(library(MASS)))
 suppressMessages(suppressWarnings(library(tidyverse)))
@@
 ```{r}
+x_max <- 3
 
-a0 <- 10
-b0 <- 30
+a0 <- 41
+b0 <- 44
 
-n <- 30
-xbar <- 0.7
+n <- 60
+xbar <- 2.1
 
@@
 
-x_grid <- seq(0.01, 8, by=0.01)
+x_grid <- seq(0.01, x_max, length=1000)
 
@@
 
-prior05 <- 1/qgamma(c(0.95), a0, b0) 
-prior95 <- 1/qgamma(c(0.05), a0, b0)
+prior05 <- 1/qgamma(c(0.975), a0, b0) 
+prior95 <- 1/qgamma(c(0.025), a0, b0)
 
-post05 <- 1/qgamma(c(0.95), a1, b1) 
-post95 <- 1/qgamma(c(0.05), a1, b1)
+post05 <- 1/qgamma(c(0.975), a1, b1) 
+post95 <- 1/qgamma(c(0.025), a1, b1)
 
@@
 ```{r}
-f_grid <- f(xbar, x_grid)^10
+f_grid <- f(xbar, x_grid)^n
 x_color <- "white"
@@
 
-### The power prior
+### The hedging hyperprior
 
-The power prior was developed to combat the excessive influence of an informative prior when the data conflicts with that prior.
+When the likelihood of your data and your informative prior are in conflict, you may wish to downweight the informative prior. You can do this by including a hedging hyperprior.
 
-Instead of the standard Bayesian approach where you multiply the Likelihood (L) by the prior (g), you use
+Define your prior distribution as an inverse gamma distributions with parameters $1+\lambda (\alpha-1)$ and $\lambda \beta$ where $\lambda$ ranges between 0 and 1.
+Then place a hyperprior on the parameter $\lambda$.
+The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.
 
-$h(\theta|\mathbf{X}) \propto L(\mathbf{X}|\theta)g(\theta)^\tau g_0(theta)$
+```{r}
+knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
+```
 
-where $g_0$ is a non-informative prior and $\tau$ is between 0 and 1.
+```{r}
+tau_grid <- seq(0, 1, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-For $\tau$=1, the prior is used at full strength.
-When $\tau$=0, the informative prior effectively disappears, leaving you with just the non-informative prior.
-For values of $\tau$ between 0 and 1, the informative prior is weakened, but still exerts some pull on the posterior distribution.
+gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
+density <- outer(theta_grid, tau_grid, gstar)
+op <- par(bg = "white")
+z_max <- max(density, na.rm=TRUE)
+persp(
+  theta_grid, 
+  tau_grid, 
+  density, 
+  zlim=c(0, 1.1*z_max),
+  xlab="theta",
+  ylab="tau",
+  theta = 135, 
+  phi = 30, 
+  expand=0.5,
+  ticktype="detailed")
+```
 
-For intermediate values of $\tau$, the formulation is not consistent with Bayes rule and you need to include a "fudge factor" of
+It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small.
+This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
 
-$\int g(\theta)^{\tau}g_0(\theta)d\theta$
+```{r}
+density <- outer(theta_grid, tau_grid, gstar)
+density[11:40, ] <- NA
+op <- par(bg = "white")
+z_max <- max(density, na.rm=TRUE)
+persp(
+  theta_grid, 
+  tau_grid, 
+  density, 
+  zlim=c(0, 1.1*z_max),
+  xlab="theta",
+  ylab="tau",
+  theta = 135, 
+  phi = 30, 
+  expand=0.5,
+  ticktype="detailed")
+```
 
-A simpler approach is to include $\tau$ in the prior distribution and then place a hyperprior on $\tau$.
+The likelihood function is constant across all values of $\tau$, so its surface looks like the following.
 
-The parameter $\alpha$ in an inverse gamma distribution can be thought of as a prior sample size.
-Large values of $\alpha$ produce informative priors that are equivalent to adding $\alpha$ pseudo-data values.
-If you multiply $\alpha$ by $\tau$, you can reduce the amount of pseudo data being added.
+```{r}
+tau_grid <- seq(0, 1, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-### The hedging hyperprior
+fstar <- function(theta, tau) {f(xbar, theta)^n}
+likelihood <- outer(theta_grid, tau_grid, fstar)
+z_max <- max(likelihood, na.rm=TRUE)
+persp(
+  theta_grid, 
+  tau_grid, 
+  likelihood, 
+  zlim=c(0, 1.1*z_max),
+  xlab="theta",
+  ylab="tau",
+  theta = 135, 
+  phi = 30, 
+  expand=0.5,
+  ticktype="detailed")
+```
+
+The product of the prior and the likelihood is proportional to the posterior distribution.
+
+```{r}
+tau_grid <- seq(0, 1, length=60)
+theta_grid <- seq(0, x_max, length=60)
+
+hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
+density <- outer(theta_grid, tau_grid, hstar)
+z_max <- max(density, na.rm=TRUE)
+persp(
+  theta_grid, 
+  tau_grid, 
+  density, 
+  zlim=c(0, 1.1*z_max),
+  xlab="theta",
+  ylab="tau",
+  theta = 135, 
+  phi = 30, 
+  expand=0.5,
+  ticktype="detailed")
+```
+
+A contour plot shows where the probability is concentrated. The labels on the contour lines represent how much probabilty is found inside the contour line.
+
+```{r}
+density <- density / sum(density, na.rm=TRUE)
+density %>%
+  as.vector %>%
+  sort %>%
+  rev -> sorted_density
+cumulative_p <- 0
+for (p in sorted_density) {
+  cumulative_p <- p + cumulative_p
+  density[density==p] <- cumulative_p
+}
+contour(
+  theta_grid, 
+  tau_grid, 
+  density,
+  labcex=1,
+  levels=c(10, 25, 50, 75, 90)/100,
+  xlab="theta",
+  ylab="tau")
+```
+
+You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).

### 2023-03-07 text/23/exponential-hedging-hyperprior.Rmd
Converted to correct name for exponential-hedging-hyperprior

--- 2023-03-06 text/23/gamma-hedging-hyperprior.Rmd
@@
 
-I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the gamma distribution.
+I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the exponential distribution.
 
@@
 
-Consider a Bayesian analysis of an outcome variable that has an exponential distribution. The density of the gamma distribution is
+Consider a Bayesian analysis of an outcome variable that has an exponential distribution. The density of the exponential distribution is
 
@@
 g_grid <- g(x_grid, a0, b0)
-sum(g_grid)*0.01 # Check the g is a density.
 outside <- x_grid < prior05 | x_grid > prior95

### 2023-03-07 text/23/normal-hedging-hyperprior.Rmd
Started work on normal-hedging-hyperprior

--- 2023-03-07 text/23/exponential-hedging-hyperprior.Rmd
@@
 ---
-title: "The hedging hyperprior for the exponential distribution"
+title: "The hedging hyperprior for the normal distribution"
 source: "New"
@@
 
-I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the exponential distribution.
+To do:
+
++ Rewrite normals using a precision parameter
++ Change the hyperparameter from lambda to delta, kappa, tau, psi, or omega.
+
+I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the normal distribution.
 
@@
 
-Consider a Bayesian analysis of an outcome variable that has an exponential distribution. The density of the exponential distribution is
+There are lots of guides on the Internet on using a normal prior for data that is normally distributed with the population standard deviation $\sigma$ known or unknown.
+I used papers written by [Kevin P.
+Murphy][mur1] and [Michael I.
+Jordan][jor1] to get these formulas.
 
-$f(x) = \theta^{-1} e^{-x / \theta}$
+[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
+[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf
 
-You have an informative prior on $\theta$ that has an inverse gamma distribution
+For simplicity, all sums and products are assumed to go from i=1 to n.
 
-$g(\theta; \alpha, \beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}$
+Consider a Bayesian analysis of an outcome variable that has a normal distribution. The density of the normal distribution is
+
+$f(X) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+
+Assume for simplicity that $\sigma$ is known. A more general case, with $\sigma$ unknown is described on a separate web page.
+
+You have an informative prior on $\mu$ that also has a normal distribution
+
+$g(\mu; \mu_0, \sigma_0)=\frac{1}{\sigma_0 \sqrt{2 \pi}}e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$
+
+Simplify this by removing any constants not involving $\mu$.
+
+$g(\mu; \mu_0, \sigma_0) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$
 
@@
 
-$L(\mathbf{X})=\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$
+$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\mu)^2 / 2\sigma^2}$
 
-and a posterior distribution of
+Rewrite this by incorporating $\bar{X}$ into the squared term
 
-$h(\theta | \mathbf{X}) \propto \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}\theta^{-n} e^{-n \bar X / \theta}$
+$L(\mathbf{X} | \mu)= \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\bar{X}+\bar{X}-\mu)^2 / 2\sigma^2}$
+
+$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-((n-1)S^2 / 2\sigma^2}e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$
+
+Pull out anything that is constant with respect to $X_i$ or $\mu$.
+
+$L(\mathbf{X}| \mu) \propto e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$
+
+The posterior distribution is proportional to the product of the prior and the likelihood.
+
+$h(\mu | \mathbf{X}) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}e^{-n(\bar{X}-\mu)^2) / 2\sigma^2}$
 
@@
 
-$h(\theta | \mathbf{X}) \propto \theta^{-\alpha-1-n}e^{-(\beta + n\bar X) / \theta}$
+$h(\mu | \mathbf{X}) \propto e^{-((\frac{\sigma}{\sigma_0}(\mu_0-\mu))^2 + n (\bar{X}-\mu)^2) / 2\sigma^2}$
 

### 2023-03-09 text/23/exponential-hedging-hyperprior.Rmd
Converted normal-hedging-hyperprior to use precisions

@@
 ```{r}
-tau_grid <- seq(0, 1, length=60)
+lambda_grid <- seq(0, 1, length=60)
 theta_grid <- seq(0, x_max, length=60)
 
-gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
-density <- outer(theta_grid, tau_grid, gstar)
+gstar <- function(theta, lambda) {g(theta, lambda*(a0-1)+1, lambda*(b0-1)+1)}
+density <- outer(theta_grid, lambda_grid, gstar)
 op <- par(bg = "white")
@@
   theta_grid, 
-  tau_grid, 
+  lambda_grid, 
   density, 
@@
   xlab="theta",
-  ylab="tau",
+  ylab="lambda",
   theta = 135, 
@@
 
-It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small.
+It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large.
 This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
+There is a similar pattern on the far side of the graph.
+The prior trends towards a higher prior for small values of $\lambda$ when $\theta$ is very small.
+
+This illustrates how the hedging hyperprior does its magic.
+It puts more probability on large values of $\lambda$ (gives closer to full weight for the informative prior) when you observe data values close to the prior mean.
+It puts more probability on small values of $\lambda$ (gives much less weight to the informative prior) when the data values are much larger or much smaller than the prior mean.
 
 ```{r}
-density <- outer(theta_grid, tau_grid, gstar)
-density[11:40, ] <- NA
+density <- outer(theta_grid, lambda_grid, gstar)
+# density[12:32, ] <- NA
+# density[11:40, ] <- NA
+density[density>0.3] <- 0.3
 op <- par(bg = "white")
@@
 
-The likelihood function is constant across all values of $\tau$, so its surface looks like the following.
+The likelihood function is constant across all values of $\lambda$, so its surface looks like the following.
 
 ```{r}
-tau_grid <- seq(0, 1, length=60)
+lambda_grid <- seq(0, 1, length=60)
 theta_grid <- seq(0, x_max, length=60)
 
-fstar <- function(theta, tau) {f(xbar, theta)^n}
-likelihood <- outer(theta_grid, tau_grid, fstar)
+fstar <- function(theta, lambda) {f(xbar, theta)^n}
+likelihood <- outer(theta_grid, lambda_grid, fstar)
 z_max <- max(likelihood, na.rm=TRUE)
@@
   theta_grid, 
-  tau_grid, 
+  lambda_grid, 
   likelihood, 
@@
 ```{r}
-tau_grid <- seq(0, 1, length=60)
+lambda_grid <- seq(0, 1, length=60)
 theta_grid <- seq(0, x_max, length=60)
 
-hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
-density <- outer(theta_grid, tau_grid, hstar)
+hstar <- function(theta, lambda) {g(theta, lambda*(a0-1)+1, lambda*(b0-1)+1)*f(xbar, theta)^n}
+density <- outer(theta_grid, lambda_grid, hstar)
 z_max <- max(density, na.rm=TRUE)
@@
   theta_grid, 
-  tau_grid, 
+  lambda_grid, 
   density,
@@
   xlab="theta",
-  ylab="tau")
+  ylab="lambda")
 ```
 
-You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).
+You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).

### 2023-03-09 text/23/normal-hedging-hyperprior.Rmd
Converted normal-hedging-hyperprior to use precisions

@@
 
+The mathematical details are a bit easier to follow if you let
+
+$\tau = \frac{1}{\sigma}$
+
+The parameter $\tau$ is called the precision. The normal density function in terms of $\tau$ is
+
+$f(X) = \frac{\tau}{\sqrt{2 \pi}}e^{- \tau^2 (X-\mu)^2}$
+
+Simplify this by removing any constants not involving $X$, $\mu$, or $\tau$.
+
+$f(X) \propto \tau e^{- \tau^2 (X-\mu)^2}$
+
 You have an informative prior on $\mu$ that also has a normal distribution
 
-$g(\mu; \mu_0, \sigma_0)=\frac{1}{\sigma_0 \sqrt{2 \pi}}e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$
+$g(\mu; \mu_0, \tau_0) \propto \tau_0 e^{-\tau_0^2 (\mu-\mu_0)^2}$
 
-Simplify this by removing any constants not involving $\mu$.
+You have collected a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of
 
-$g(\mu; \mu_0, \sigma_0) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$
-
-and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of
-
-$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\mu)^2 / 2\sigma^2}$
+$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2\sum(X_i-\mu)^2}$
 
@@
 
-$L(\mathbf{X} | \mu)= \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\bar{X}+\bar{X}-\mu)^2 / 2\sigma^2}$
+$L(\mathbf{X} | \mu) \propto \tau^n e^{-\sum \tau (X_i-\bar{X}+\bar{X}-\mu)^2}$
 
-$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-((n-1)S^2 / 2\sigma^2}e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$
+When you square the term inside the sum, the cross-product cancels out, leaving you with
 
-Pull out anything that is constant with respect to $X_i$ or $\mu$.
+$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2(\sum(X_i-\bar{X})^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$
 
-$L(\mathbf{X}| \mu) \propto e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$
+$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2((n-1)S^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}e^{-n(\bar{X}-\mu)^2) / 2\sigma^2}$
+$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu-\mu_0)^2} \tau^n e^{-\tau((n-1)S^2} e^{-\tau^2\sum(\bar{X}-\mu)^2)}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{-((\frac{\sigma}{\sigma_0}(\mu_0-\mu))^2 + n (\bar{X}-\mu)^2) / 2\sigma^2}$
+$h(\mu | \mathbf{X}) \propto $
 
-which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean
+which, thanks to the miracle of conjugancy, is a normal distribution. The posterior mean
 
-$\frac{\beta + n\bar X}{\alpha-1+n}$
+$\frac{\tau n\bar{X}\beta + \tau_0 \mu_0}{n\tau+\tau0}$
 
@@
 
-$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$
+The posterior precision is
+
+$\tau0+n\tau$
 

### 2023-03-12 text/23/normal-hedging-hyperprior.Rmd
Initial graphs for normal-hedging-hyperprior

@@
 
-$f(X) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
 
-Assume for simplicity that $\sigma$ is known. A more general case, with $\sigma$ unknown is described on a separate web page.
+Assume for simplicity that $\sigma$ is known and equal to $\sigma_F$. A more general case, with $\sigma$ unknown is described on a separate web page.
 
@@
 
-$\tau = \frac{1}{\sigma}$
+$\tau_F = \frac{1}{\sigma_F^2}$
 
-The parameter $\tau$ is called the precision. The normal density function in terms of $\tau$ is
+The parameter $\tau$ is called the precision.
+Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance, The normal density function in terms of $\tau_F$ is
 
-$f(X) = \frac{\tau}{\sqrt{2 \pi}}e^{- \tau^2 (X-\mu)^2}$
+$f(X; \mu, \tau_F) = \sqrt{\frac{\tau_F}{2 \pi}}e^{- \frac{1}{2}\tau_F (X-\mu)^2}$
 
-Simplify this by removing any constants not involving $X$, $\mu$, or $\tau$.
+Simplify this by removing any constants not involving $X$, $\mu$, or $\tau_F$.
 
-$f(X) \propto \tau e^{- \tau^2 (X-\mu)^2}$
+$f(X) \propto e^{- \frac{1}{2} \tau_F (X-\mu)^2}$
 
-You have an informative prior on $\mu$ that also has a normal distribution
+You have an informative prior on $\mu$ that also has a normal distribution with a prior mean $\mu_0$ and a prior precision $\tau_0$.
 
-$g(\mu; \mu_0, \tau_0) \propto \tau_0 e^{-\tau_0^2 (\mu-\mu_0)^2}$
+$g(\mu; \mu_0, \tau_0) \propto  e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2}$
+
+For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$.
+The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\mu_0$ to the observed data.
 
@@
 
-$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2\sum(X_i-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto \tau^n e^{-\sum \tau (X_i-\bar{X}+\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2(\sum(X_i-\bar{X})^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$
 
-$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2((n-1)S^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu-\mu_0)^2} \tau^n e^{-\tau((n-1)S^2} e^{-\tau^2\sum(\bar{X}-\mu)^2)}$
+$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$
+
+[Note that the following formulas need a lot of work.]
+
+Drop the middle exponential term, which is constant with respect to $\mu$. Set
+
+$\mu_W=\frac{\tau_0 \mu_0 + n \tau_F \bar{X}}{\tau_0 + n \tau_F}$
+
+and rewrite the other two exponentials as
+
+$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu - \mu_W + \mu_W - mu_0)^2} \tau_F^n e^{-\tau_F(n-1)S^2} e^{-n \tau_F^2(\mu - \mu_W + \mu_W - \bar{X})^2)}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto $
+$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)^2(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
 
@@
 
-$\frac{\tau n\bar{X}\beta + \tau_0 \mu_0}{n\tau+\tau0}$
+$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$
 
-can be written as a weighted average of the prior mean and the mean of the observed data.
+and the posterior precision is
 
-The posterior precision is
+$n \tau_F + \tau_0$
 
-$\tau0+n\tau$
+The posterior mean can be written as a weighted average of the prior mean and the mean of the observed data.
 
 ```{r}
-x_max <- 3
+x_max <- 10
 
-a0 <- 41
-b0 <- 44
+mu0 <- 8
+tau0 <- 4
 
-n <- 60
-xbar <- 2.1
+tauF <- 10
 
-a1 <- a0 + n
-b1 <- b0 + n*xbar
+n <- 10
+xbar <- 5
 
-posterior_alpha <- a0+n
-posterior_beta <- b0+n*xbar
+mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
+tau1 <- tau0+n*tauF
 
@@
 
-f <- function(x, theta) {
-  theta^(-1) * 
-    exp(-x/theta)
+f <- function(x, mu, tau) {
+  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
 }
 
-g <- function(theta, alpha, beta) {
-  beta^alpha/gamma(alpha) *
-    theta^(-alpha-1) * 
-    exp(-beta/theta)
-}
+g <- f
 
-prior05 <- 1/qgamma(c(0.975), a0, b0) 
-prior95 <- 1/qgamma(c(0.025), a0, b0)
+prior05 <- qnorm(0.025, mu0, tau0) 
+prior95 <- qnorm(0.975, mu0, tau0)
 
-post05 <- 1/qgamma(c(0.975), a1, b1) 
-post95 <- 1/qgamma(c(0.025), a1, b1)
+post05 <- qnorm(0.025, mu1, tau1) 
+post95 <- qnorm(0.025, mu1, tau1)
 
@@
 
-Let's illustrate this with an informative prior with $\alpha$ = `r a0` and $\beta$ = `r b0`.
+Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.
 
 ```{r}
-g_grid <- g(x_grid, a0, b0)
+g_grid <- g(x_grid, mu0, tau0)
 outside <- x_grid < prior05 | x_grid > prior95
@@
 ```{r}
-f_grid <- f(xbar, x_grid)^n
+f_grid <- f(xbar, x_grid, tauF)^n
 x_color <- "white"
@@
 ```{r}
-h_grid <- g(x_grid, a1, b1)
+h_grid <- g(x_grid, mu1, tau1)
 outside <- x_grid < post05 | x_grid > post95
@@
 It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
-It excludes the prior mean of `r sprintf("%.2f", b0/(a0-1))` and it excludes `r xbar`, the mean of the data.
+It excludes the prior mean of `r sprintf("%.2f", mu0)` and it excludes `r xbar`, the mean of the data.
 
@@
 ```{r}
-tau_grid <- seq(0, 1, length=60)
-theta_grid <- seq(0, x_max, length=60)
+lambda_grid <- seq(0, 1, length=60)
+mu_grid <- seq(0, x_max, length=60)
 
-gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
-density <- outer(theta_grid, tau_grid, gstar)
+gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
+density <- outer(mu_grid, lambda_grid, gstar)
 op <- par(bg = "white") 
@@
 persp(
-  theta_grid, 
-  tau_grid, 
+  mu_grid, 
+  lambda_grid, 
   density, 
   zlim=c(0, 1.1*z_max),
-  xlab="theta",
-  ylab="tau",
+  xlab="mu",
+  ylab="lambda",
   theta = 135, 
@@
 
-It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small.
+It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small.
 This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
@@
 ```{r}
-density <- outer(theta_grid, tau_grid, gstar)
+density <- outer(mu_grid, lambda_grid, gstar)
 density[11:40, ] <- NA
@@
 ```{r}
-tau_grid <- seq(0, 1, length=60)
-theta_grid <- seq(0, x_max, length=60)
+lambda_grid <- seq(0, 1, length=60)
+mu_grid <- seq(0, x_max, length=60)
 
-fstar <- function(theta, tau) {f(xbar, theta)^n}
-likelihood <- outer(theta_grid, tau_grid, fstar)
+fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
+likelihood <- outer(mu_grid, lambda_grid, fstar)
 z_max <- max(likelihood, na.rm=TRUE)
 persp(
-  theta_grid, 
-  tau_grid, 
+  mu_grid, 
+  lambda_grid, 
   likelihood, 
   zlim=c(0, 1.1*z_max),
-  xlab="theta",
-  ylab="tau",
+  xlab="mu",
+  ylab="lambda",
   theta = 135, 
@@
 ```{r}
-tau_grid <- seq(0, 1, length=60)
-theta_grid <- seq(0, x_max, length=60)
+lambda_grid <- seq(0, 1, length=60)
+mu_grid <- seq(0, x_max, length=60)
 
-hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
-density <- outer(theta_grid, tau_grid, hstar)
+hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
+density <- outer(mu_grid, lambda_grid, hstar)
 z_max <- max(density, na.rm=TRUE)
 persp(
-  theta_grid, 
-  tau_grid, 
+  mu_grid, 
+  lambda_grid, 
   density, 
   zlim=c(0, 1.1*z_max),
-  xlab="theta",
-  ylab="tau",
+  xlab="mu",
+  ylab="lambda",
   theta = 135, 
@@
 contour(
-  theta_grid, 
-  tau_grid, 
+  mu_grid, 
+  lambda_grid, 
   density,
@@
   levels=c(10, 25, 50, 75, 90)/100,
-  xlab="theta",
-  ylab="tau")
+  xlab="mu",
+  ylab="lambda")
 ```
 
-You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).
+You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).

### 2023-03-14 text/23/normal-hedging-hyperprior-part-1.Rmd
Barely started work on normal-hedging-hyperprior-part-2

--- 2023-03-12 text/23/normal-hedging-hyperprior.Rmd
@@
 ---
-title: "The hedging hyperprior for the normal distribution"
+title: "The hedging hyperprior for the normal distribution: unknown mean, known precision"
 source: "New"
@@
 
-To do:
-
-+ Rewrite normals using a precision parameter
-+ Change the hyperparameter from lambda to delta, kappa, tau, psi, or omega.
-
 I want to write a paper about the hedging hyperprior. I need to work out some simple examples first. Here is an example involving the normal distribution.
@@
 
-[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
-[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf
-
 For simplicity, all sums and products are assumed to go from i=1 to n.
@@
 
-$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+For this derivation, it is easier to write the density in terms of
 
-Assume for simplicity that $\sigma$ is known and equal to $\sigma_F$. A more general case, with $\sigma$ unknown is described on a separate web page.
+$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
 
-The mathematical details are a bit easier to follow if you let
+where
 
-$\tau_F = \frac{1}{\sigma_F^2}$
+$\tau=\frac{1}{\sigma*2}$
 
-The parameter $\tau$ is called the precision.
-Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance, The normal density function in terms of $\tau_F$ is
+The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,
 
-$f(X; \mu, \tau_F) = \sqrt{\frac{\tau_F}{2 \pi}}e^{- \frac{1}{2}\tau_F (X-\mu)^2}$
+Assume that $\tau$ is known and equal to $\tau_F$.
+A more general case, with $\tau$ unknown is described on a separate web page.
+For simplicity, we will drop any multiplicative constants not involving X or $\mu$
 
-Simplify this by removing any constants not involving $X$, $\mu$, or $\tau_F$.
+$f(X; \mu) \propto e^{-\frac{1}{2}\tau_F(X-\mu)^2}$
 
-$f(X) \propto e^{- \frac{1}{2} \tau_F (X-\mu)^2}$
+The conjugate prior for a normal distribution is also a normal distribution.
 
-You have an informative prior on $\mu$ that also has a normal distribution with a prior mean $\mu_0$ and a prior precision $\tau_0$.
+$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$
 
-$g(\mu; \mu_0, \tau_0) \propto  e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2}$
+where the prior mean is $\mu_0$ and the prior precision is $\tau_0$.
 
@@
 
-[Note that the following formulas need a lot of work.]
+Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get
 
-Drop the middle exponential term, which is constant with respect to $\mu$. Set
+$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
 
-$\mu_W=\frac{\tau_0 \mu_0 + n \tau_F \bar{X}}{\tau_0 + n \tau_F}$
-
-and rewrite the other two exponentials as
-
-$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu - \mu_W + \mu_W - mu_0)^2} \tau_F^n e^{-\tau_F(n-1)S^2} e^{-n \tau_F^2(\mu - \mu_W + \mu_W - \bar{X})^2)}$
-
-which simplifies to
-
-$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)^2(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
-
-which, thanks to the miracle of conjugancy, is a normal distribution. The posterior mean
+which is a normal distribution. The posterior mean
 
@@
 
+$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$
+
 ```{r}
-x_max <- 10
+x_max <- 8
 
-mu0 <- 8
-tau0 <- 4
+mu0 <- 4
+tau0 <- 8
 
-tauF <- 10
-
+tauF <- 0.6
 n <- 10
-xbar <- 5
+xbar <- 2
 
@@
 
-prior05 <- qnorm(0.025, mu0, tau0) 
-prior95 <- qnorm(0.975, mu0, tau0)
+prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
+prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))
 
-post05 <- qnorm(0.025, mu1, tau1) 
-post95 <- qnorm(0.025, mu1, tau1)
+post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
+post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))
 
@@
 It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
-It excludes the prior mean of `r sprintf("%.2f", mu0)` and it excludes `r xbar`, the mean of the data.
+It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.
 
@@
 
-Define your prior distribution as an inverse gamma distributions with parameters $1+\lambda (\alpha-1)$ and $\lambda \beta$ where $\lambda$ ranges between 0 and 1.
-Then place a hyperprior on the parameter $\lambda$.
-The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.
+Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision
+
+$\tau_F+\lambda (\tau_0-\tau_F)$
+
+where $\lambda$ ranges between 0 and 1 and $\tau_F$ is less than $\tau_0$.
+When $\lambda=0$, the prior precision reduces to $\tau_F$ which is equivalent to a prior sample size of $n_0=1$.
+When $lambda=1$, the prior precision is $\tau_0$ which represents the full strength of the informative priori distribution.
+
+Place a hyperprior on the parameter $\lambda$. The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.
 
@@
 
+This is the joint prior distribution of $\mu_0$ and $\lambda$.
+
 It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small.
@@
 density <- outer(mu_grid, lambda_grid, gstar)
-density[11:40, ] <- NA
+density[16:44, ] <- NA
 op <- par(bg = "white")
@@
   labcex=1,
-  levels=c(10, 25, 50, 75, 90)/100,
+  levels=c(50, 75, 90)/100,
   xlab="mu",
@@
 You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).
+
+[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
+[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf
+[wik1]: https://en.wikipedia.org/wiki/Completing_the_square

### 2023-03-14 text/23/normal-hedging-hyperprior-part-2.Rmd
Barely started work on normal-hedging-hyperprior-part-2

--- 2023-03-14 text/23/normal-hedging-hyperprior-part-1.Rmd
@@
 ---
-title: "The hedging hyperprior for the normal distribution: unknown mean, known precision"
+title: "The hedging hyperprior for the normal distribution: known mean, unknown precision"
 source: "New"

### 2023-03-16 text/23/normal-hedging-hyperprior-part-2.Rmd
Added gamma distribution to normal-heding-prior-part-2

@@
 source: "New"
-date: "2023-02-28"
+date: "2023-03-14"
 categories:
@@
 
+$f(X; \mu, \sigma) = \frac{\tau}{\sigma sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+
 For this derivation, it is easier to write the density in terms of
 
-$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\mu)^2}$
 
@@
 
-Assume that $\tau$ is known and equal to $\tau_F$.
-A more general case, with $\tau$ unknown is described on a separate web page.
-For simplicity, we will drop any multiplicative constants not involving X or $\mu$
+Assume that $\mu$ is known and equal to $\mu_F$.
+An earlier case with $\tau$ known and $\mu$ unknown appears on a different web page.
+A more general case, with both $\mu$ and $\tau$ unknown is also described on a separate web page.
+For simplicity, we will drop any multiplicative constants not involving X or $\tau$
 
-$f(X; \mu) \propto e^{-\frac{1}{2}\tau_F(X-\mu)^2}$
+$f(X; \tau) \propto \sqrt{\tau}e^{-\frac{1}{2}\tau(X-\mu_F)^2}$
 
-The conjugate prior for a normal distribution is also a normal distribution.
+The conjugate prior for a normal distribution is a gamma distribution.
 
-$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$
+$g(\tau; \alpha_0, \beta_0) \propto \frac{\beta_0^\alpha_0}{\Gamma(\alpha)} \tau^{\alpha-1} e^{-\beta \tau}$
 
-where the prior mean is $\mu_0$ and the prior precision is $\tau_0$.
+where
 
-For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$.
-The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\mu_0$ to the observed data.
+$\beta_0 = 1 / \tau_0$
+
+For an informative prior distribution, $\alpha$ is large. This prior is is equivalent to adding $n_0$ pseudo data values of $\tau_0$ to the observed data.
 

### 2023-03-18 text/23/normal-hedging-hyperprior-part-1.Rmd
Added gamma distribution to normal-heding-prior-part-2
Converted formulas in normal-hedging-hyperprior to use theta

@@
 
-### A simple Baysian analysis with an informative prior.
+### A simple Bayesian analysis with an informative prior.
 
@@
 
+$f(X; \theta, \sigma) = \frac{\tau}{\sigma \sqrt{2 \pi}}e^{-(X-\theta)^2 / 2\sigma^2}$
+
 For this derivation, it is easier to write the density in terms of
 
-$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+$f(X; \theta, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\theta)^2}$
 
@@
 
-$\tau=\frac{1}{\sigma*2}$
+$\tau=\frac{1}{\sigma^2}$
 
@@
 A more general case, with $\tau$ unknown is described on a separate web page.
-For simplicity, we will drop any multiplicative constants not involving X or $\mu$
+For simplicity, we will drop any multiplicative constants not involving X or $\theta$
 
-$f(X; \mu) \propto e^{-\frac{1}{2}\tau_F(X-\mu)^2}$
+$f(X; \theta) \propto e^{-\frac{1}{2}\tau_F(X-\theta)^2}$
 
@@
 
-$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$
+$g(\theta; \theta_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\theta-\theta_0)^2}$
 
-where the prior mean is $\mu_0$ and the prior precision is $\tau_0$.
+where the prior mean is $\theta_0$ and the prior precision is $\tau_0$.
 
 For an informative prior distribution, $\tau_0$ is usually much smaller than $\tau_F$.
-The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\mu_0$ to the observed data.
+The ratio $n_0=\tau_F/\tau_0$ is the effective prior sample size and is equivalent to adding $n_0$ pseudo data values of $\theta_0$ to the observed data.
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$
+$L(\mathbf{X} | \theta) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\theta)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \theta) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\theta)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \theta) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\theta)^2}$
 
-$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$
+$L(\mathbf{X} | \theta) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\theta-\bar{X})^2}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$
+$h(\theta | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\theta-\theta_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\theta-\bar{X})^2)}$
 
-Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get
+Drop the middle exponential term, which is constant with respect to $\theta$ and use high school algebra ([completing the square][wik1]) to get
 
-$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
+$h(\theta | \mathbf{X}) \propto e^{-(n\tau_F+\tau_0)(\theta-\frac{n \tau_F \bar{X} + \tau_0 \theta_0}{n\tau_F+\tau_0})^2}$
 
@@
 
-$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$
+$\frac{n \tau_F \bar{X} + \tau_0 \theta_0}{n\tau_F+\tau_0}$
 
@@
 
-$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$
+$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\theta_0$
 
@@
 
-mu0 <- 4
+theta0 <- 4
 tau0 <- 8
@@
 
-mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
+theta1 <- (tau0*theta0+n*tauF*xbar)/(tau0+n*tauF)
 tau1 <- tau0+n*tauF
@@
 
-f <- function(x, mu, tau) {
-  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
+f <- function(x, theta, tau) {
+  sqrt(tau/2*pi)*exp(-0.5*tau*(x-theta)^2)
 }
@@
 
-prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
-prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))
+prior05 <- qnorm(0.025, theta0, 1/sqrt(tau0)) 
+prior95 <- qnorm(0.975, theta0, 1/sqrt(tau0))
 
-post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
-post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))
+post05 <- qnorm(0.025, theta1, 1/sqrt(tau1)) 
+post95 <- qnorm(0.975, theta1, 1/sqrt(tau1))
 
@@
 
-Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.
+Let's illustrate this with an informative prior with $\theta_0$ = `r theta0` and $\tau_0$ = `r tau0`.
 
 ```{r}
-g_grid <- g(x_grid, mu0, tau0)
+g_grid <- g(x_grid, theta0, tau0)
 outside <- x_grid < prior05 | x_grid > prior95
@@
 ```{r}
-h_grid <- g(x_grid, mu1, tau1)
+h_grid <- g(x_grid, theta1, tau1)
 outside <- x_grid < post05 | x_grid > post95
@@
 It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
-It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.
+It excludes the prior mean of `r theta0` and it excludes `r xbar`, the mean of the data.
 
@@
 
-Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision
+Define your prior distribution as a normal distributions with prior mean $\theta0$ and prior precision
 
@@
 lambda_grid <- seq(0, 1, length=60)
-mu_grid <- seq(0, x_max, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
-density <- outer(mu_grid, lambda_grid, gstar)
+gstar <- function(theta, lambda) {g(theta, theta0, lambda*(tau0-tauF)+tauF)}
+density <- outer(theta_grid, lambda_grid, gstar)
 op <- par(bg = "white") 
@@
 persp(
-  mu_grid, 
+  theta_grid, 
   lambda_grid, 
@@
   zlim=c(0, 1.1*z_max),
-  xlab="mu",
+  xlab="theta",
   ylab="lambda",
@@
 
-This is the joint prior distribution of $\mu_0$ and $\lambda$.
+This is the joint prior distribution of $\theta_0$ and $\lambda$.
 
-It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small.
+It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small.
 This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
@@
 ```{r}
-density <- outer(mu_grid, lambda_grid, gstar)
+density <- outer(theta_grid, lambda_grid, gstar)
 density[16:44, ] <- NA
@@
 lambda_grid <- seq(0, 1, length=60)
-mu_grid <- seq(0, x_max, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
-likelihood <- outer(mu_grid, lambda_grid, fstar)
+fstar <- function(theta, lambda) {f(xbar, theta, tau0)^n}
+likelihood <- outer(theta_grid, lambda_grid, fstar)
 z_max <- max(likelihood, na.rm=TRUE)
 persp(
-  mu_grid, 
+  theta_grid, 
   lambda_grid, 
@@
 lambda_grid <- seq(0, 1, length=60)
-mu_grid <- seq(0, x_max, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
-density <- outer(mu_grid, lambda_grid, hstar)
+hstar <- function(theta, lambda) {g(theta, theta0, lambda*(tau0-tauF) + tauF)*f(xbar, theta, tauF)^n}
+density <- outer(theta_grid, lambda_grid, hstar)
 z_max <- max(density, na.rm=TRUE)
 persp(
-  mu_grid, 
+  theta_grid, 
   lambda_grid, 
@@
 contour(
-  mu_grid, 
+  theta_grid, 
   lambda_grid, 
@@
   levels=c(50, 75, 90)/100,
-  xlab="mu",
+  xlab="theta",
   ylab="lambda")
@@
 
-You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).
+You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r xbar`).
 

### 2023-03-18 text/23/normal-hedging-hyperprior-part-2.Rmd
Converted formulas in normal-hedging-hyperprior to use theta

@@
 
-Consider a Bayesian analysis of an outcome variable that has a normal distribution. The density of the normal distribution is
+Consider a Bayesian analysis of an outcome variable that has a normal distribution.
 
-$f(X; \mu, \sigma) = \frac{\tau}{\sigma sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+For this derivation, it is easier to write the density in terms of the precision
 
-For this derivation, it is easier to write the density in terms of
-
-$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\mu)^2}$
+$f(X; \mu, \theta) = \sqrt{\frac{\theta}{2 \pi}}e^{-\frac{1}{2}\theta (X-\mu)^2}$
 
@@
 
-$\tau=\frac{1}{\sigma*2}$
+$\theta=\frac{1}{\sigma*2}$
 
-The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,
+Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$.
+An earlier case with precision known and the mean  unknown appears on a different web page.
+A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page.
+For simplicity, we will drop any multiplicative constants not involving X or $\theta$
 
-Assume that $\mu$ is known and equal to $\mu_F$.
-An earlier case with $\tau$ known and $\mu$ unknown appears on a different web page.
-A more general case, with both $\mu$ and $\tau$ unknown is also described on a separate web page.
-For simplicity, we will drop any multiplicative constants not involving X or $\tau$
-
-$f(X; \tau) \propto \sqrt{\tau}e^{-\frac{1}{2}\tau(X-\mu_F)^2}$
+$f(X; \theta) \propto \sqrt{\theta}e^{-\frac{1}{2}\theta(X-\mu_F)^2}$
 
@@
 
-$g(\tau; \alpha_0, \beta_0) \propto \frac{\beta_0^\alpha_0}{\Gamma(\alpha)} \tau^{\alpha-1} e^{-\beta \tau}$
+$g(\theta; \alpha_0, \beta_0) = \frac{\beta_0^{\alpha_0}}{\Gamma(\alpha_0)} \theta^{\alpha_0-1} e^{-\beta_0 \theta}$
 
-where
+Set the prior parameters to
 
-$\beta_0 = 1 / \tau_0$
+$\alpha_0=n_0$
 
-For an informative prior distribution, $\alpha$ is large. This prior is is equivalent to adding $n_0$ pseudo data values of $\tau_0$ to the observed data.
+$\beta_0 = n_0 \theta_0$
+
+and remove any multiplicative constants with respect to $\theta$ to get
+
+$g(\theta; n_0, \theta_0) \propto \theta^{n_0-1} e^{-n_0 \theta_0 \theta}$
+
+In this parameterization, $n_0$ is the prior sample size.
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$
-
-Rewrite this by incorporating $\bar{X}$ into the squared term
-
-$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$
-
-When you square the term inside the sum, the cross-product cancels out, leaving you with
-
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$
-
-$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$
+$L(\mathbf{X} | \mu) \propto \theta^{n/2}e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$
+$h(\theta | \mathbf{X}) \propto \theta^{n_0-1} e^{-n_0 \theta / \theta0} \theta^{n/2} e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$
 
-Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get
+which can be rewritten as
 
-$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
+$h(\theta | \mathbf{X}) \propto \theta^{n/2 + n_0-1} e^{-(n_0 \theta0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$
 
-which is a normal distribution. The posterior mean
+which is a gamma distribution with posterior parameters
 
-$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$
+$\alpha_1 = n/2+n_0$ and
 
-and the posterior precision is
+$\beta_1 = n_0 theta_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$
 
-$n \tau_F + \tau_0$
+This can also be shown to relate to a t-distribution
 
-The posterior mean can be written as a weighted average of the prior mean and the mean of the observed data.
-
-$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$
+$\frac{\Gamma \left(\frac{\nu+1}{2} \right)} {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)} \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$
 
@@
 
-mu0 <- 4
-tau0 <- 8
+n0 <- 4
+theta0 <- 0.5
 
-tauF <- 0.6
+alpha0 <- n0
+beta0 <- n0*theta0
+
+muF <- 2
 n <- 10
-xbar <- 2
 
-mu1 <- (tau0*mu0+n*tauF*xbar)/(tau0+n*tauF)
-tau1 <- tau0+n*tauF
+x <- muF+ round(rnorm(n), 1)/theta0
+
+ss <- sum((x-muF)^2)
+
+alpha1 <- alpha0 + n
+beta1 <- beta0 + 0.5*ss
 
@@
 
-f <- function(x, mu, tau) {
-  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
+f <- function(x, mu, theta) {
+  sqrt(theta/2*pi)*exp(-0.5*theta*(x-mu)^2)
 }
 
-g <- f
+g <- function(x, alpha, beta) {
+  beta^alpha/gamma(alpha)*x^(alpha-1)*exp(-beta*x)
+}
 
-prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
-prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))
+prior05 <- qgamma(0.025, alpha0, beta0) 
+prior95 <- qgamma(0.975, alpha0, beta0)
 
-post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
-post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))
+post05 <- qgamma(0.025, alpha1, beta1)  
+post95 <- qgamma(0.975, alpha1, beta1)
 
@@
 
-Let's illustrate this with an informative prior with $\mu_0$ = `r mu0` and $\tau_0$ = `r tau0`.
+Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\theta_0$ = `r theta0`.
 
 ```{r}
-g_grid <- g(x_grid, mu0, tau0)
+g_grid <- g(x_grid, alpha0, beta0)
 outside <- x_grid < prior05 | x_grid > prior95
@@
 
-Let's see what happens with this informative prior if you observe `r n` values of x with a mean of `r xbar`. The likelihood is
+Let's see what happens with this informative prior if you observe `r n` values of x: `r paste0(x, collapse=", ")`. The likelihood is
 
 ```{r}
-f_grid <- f(xbar, x_grid, tauF)^n
+f_grid <- rep(NA, length(x_grid))
+for (i in 1:length(x_grid)) {
+  f_grid[i] <- prod(f(x, muF, x_grid[i]))
+}
 x_color <- "white"
@@
 ```{r}
-h_grid <- g(x_grid, mu1, tau1)
+h_grid <- g(x_grid, alpha1, beta1)
 outside <- x_grid < post05 | x_grid > post95
@@
 It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
-It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.
+It excludes the prior mean of `r theta0` and it excludes `r ss/(n-1)`, the mean of the data.
 
@@
 
-Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision
+Define your prior distribution as a gamma distributions with prior
 
-$\tau_F+\lambda (\tau_0-\tau_F)$
+mean $\theta_0$ and prior precision
 
-where $\lambda$ ranges between 0 and 1 and $\tau_F$ is less than $\tau_0$.
-When $\lambda=0$, the prior precision reduces to $\tau_F$ which is equivalent to a prior sample size of $n_0=1$.
-When $lambda=1$, the prior precision is $\tau_0$ which represents the full strength of the informative priori distribution.
+$\alpha = 1 + \lambda(n0-1)$
+
+$\beta = \alpha \theta_0 $
+
+where $\lambda$ ranges between 0 and 1.
+When $\lambda=0$, the prior precision reduces to $alpha=1$ which is equivalent to a very diffuse prior.
+When $lambda=1$, the prior precision is $\alpha=\alpha_0$ or the full strength prior.
 
@@
 lambda_grid <- seq(0, 1, length=60)
-mu_grid <- seq(0, x_max, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
-density <- outer(mu_grid, lambda_grid, gstar)
+gstar <- function(theta, lambda) {g(theta, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
+density <- outer(theta_grid, lambda_grid, gstar)
 op <- par(bg = "white") 
@@
 
-This is the joint prior distribution of $\mu_0$ and $\lambda$.
+This is the joint prior distribution of $\theta$ and $\lambda$.
 
-It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\mu$ is very large or very small.
+It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small.
 This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
 
-```{r}
-density <- outer(mu_grid, lambda_grid, gstar)
+```{r, eval=FALSE}
+density <- outer(theta_grid, lambda_grid, gstar)
 density[16:44, ] <- NA
@@
 
-The likelihood function is constant across all values of $\tau$, so its surface looks like the following.
+The likelihood function is constant across all values of $\theta$, so its surface looks like the following.
 
-```{r}
+```{r, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
-mu_grid <- seq(0, x_max, length=60)
+theta_grid <- seq(0, x_max, length=60)
 
-fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
+fstar <- function(mu, lambda) {f(xbar, mu, theta0)^n}
 likelihood <- outer(mu_grid, lambda_grid, fstar)
@@
 
-```{r}
+```{r, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
+hstar <- function(mu, lambda) {g(mu, mu0, lambda*(theta0-thetaF) + thetaF)*f(xbar, mu, thetaF)^n}
 density <- outer(mu_grid, lambda_grid, hstar)
@@
 
-```{r}
+```{r, eval=FALSE}
 density <- density / sum(density, na.rm=TRUE)
@@
 
-You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r mu0`) is largely ignored in favor of the mean of the data (`r xbar`).
+You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r "??"`).
 

### 2023-03-21 text/23/normal-hedging-hyperprior-part-1.Rmd
Started work on normal-hedging-hyperprior-part3

@@
 
-The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,
+The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance.
 
-Assume that $\tau$ is known and equal to $\tau_F$.
-A more general case, with $\tau$ unknown is described on a separate web page.
+Assume that $\tau$ is known and equal to $\tau_F$. Other cases with [mean known and precision unknown][sim2] and with [both mean and precision unknown][sim3] appear on a different web pages.
+
+[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
+[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3
+
 For simplicity, we will drop any multiplicative constants not involving X or $\theta$

### 2023-03-21 text/23/normal-hedging-hyperprior-part-3.Rmd
Started work on normal-hedging-hyperprior-part3

--- 2023-03-18 text/23/normal-hedging-hyperprior-part-2.Rmd
@@
 ---
-title: "The hedging hyperprior for the normal distribution: known mean, unknown precision"
+title: "The hedging hyperprior for the normal distribution: unknown mean, unknown precision"
 source: "New"
-date: "2023-03-14"
+date: "2023-03-21"
 categories:
@@
 
-$f(X; \mu, \theta) = \sqrt{\frac{\theta}{2 \pi}}e^{-\frac{1}{2}\theta (X-\mu)^2}$
+$f(X; \mu, \tau) = 
+  \sqrt{\frac{\tau}{2 \pi}}
+  e^{-\frac{1}{2}\tau (X-\mu)^2}$
 
-where
+Assume that both the mean and the precision of the normal distribution are unknown.
+Earlier cases with [mean unknown and precision known][sim1] and with [mean known and precision unknown][sim2] appear on a different web pages.
 
-$\theta=\frac{1}{\sigma*2}$
+[sim1]: http://new.pmean.com/normal-hedging-hyperprior-part1
+[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
 
-Assume that the mean of the normal distribution, $\mu$, is known and equal to $\mu_F$.
-An earlier case with precision known and the mean  unknown appears on a different web page.
-A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page.
-For simplicity, we will drop any multiplicative constants not involving X or $\theta$
+For simplicity, we will drop any multiplicative constants not involving X, $\mu$, or $\tau$.
 
-$f(X; \theta) \propto \sqrt{\theta}e^{-\frac{1}{2}\theta(X-\mu_F)^2}$
+$f(X; \mu, \tau) \propto 
+  \sqrt{\tau}
+  e^{-\frac{1}{2}\tau(X-\mu)^2}$
 
-The conjugate prior for a normal distribution is a gamma distribution.
+With two unknown parameters, you need two prior distributions. The prior distribution for $\tau$ is
 
-$g(\theta; \alpha_0, \beta_0) = \frac{\beta_0^{\alpha_0}}{\Gamma(\alpha_0)} \theta^{\alpha_0-1} e^{-\beta_0 \theta}$
+$k(\tau) = 
+  \frac{(n_0 \tau_0)^{n_0}}{\Gamma(n_0)}
+  \tau^{n_0-1}
+  e^{-\beta_0 \tau}$
 
-Set the prior parameters to
+which is a gamma distribution with
 
-$\alpha_0=n_0$
+$\alpha=n_0$ and $\beta=n_0\tau_0$
 
-$\beta_0 = n_0 \theta_0$
+Use a conditional prior for $\mu$ given $\tau$.
 
-and remove any multiplicative constants with respect to $\theta$ to get
+$g(\mu | \tau) =
+  \sqrt{\frac{\tau}{2 \pi}}
+  e^{-\frac{1}{2}\tau (\mu-\mu_0)^2}$
 
-$g(\theta; n_0, \theta_0) \propto \theta^{n_0-1} e^{-n_0 \theta_0 \theta}$
+You can get an unconditional prior for $\mu$ by integrating out the prior on $\tau$
 
-In this parameterization, $n_0$ is the prior sample size.
+$g(\mu) = \int
+  \sqrt{\frac{\tau}{2 \pi}}
+  e^{-\frac{1}{2}\tau (\mu-\mu_0)^2}
+  \frac{(n_0 \tau_0)^{n_0}}{\Gamma(n_0)}
+  \tau^{n_0-1}
+  e^{-\beta_0 \tau}d\tau$
+
+Pull out any multiplicative constants not involve $\tau$.
+
+$g(\mu) = 
+  \frac{1}{\sqrt{2 \pi}}
+  \frac{(n_0 \tau_0)^{n_0}}{\Gamma(n_0)}
+  \int\tau^{n_0-\frac{1}{2}}
+  e^{-\tau(\frac{1}{2} (\mu-\mu_0)^2-\beta_0)}
+  d\tau$
 

### 2023-03-22 text/23/normal-hedging-hyperprior-part-3.Rmd
Tweaking formulas in normal-hedging-hyperprior-part-3

@@
   \tau^{n_0-1}
-  e^{-\beta_0 \tau}d\tau$
+  e^{-\frac{n_0 \tau}{\tau0}}d\tau$
 
-Pull out any multiplicative constants not involve $\tau$.
+Pull out any multiplicative constants not involving $\tau$.
 
@@
   \frac{1}{\sqrt{2 \pi}}
-  \frac{(n_0 \tau_0)^{n_0}}{\Gamma(n_0)}
+  \frac{(n_0 \tau_0)^{n_0}}{\Gamma(n_0)}$
   \int\tau^{n_0-\frac{1}{2}}
-  e^{-\tau(\frac{1}{2} (\mu-\mu_0)^2-\beta_0)}
-  d\tau$
+  e^{-\tau(
+    \frac{1}{2}(\mu-\mu_0)^2 - 
+    \frac{n_0}{\tau0}}d\tau)$
+
+What remains inside the integral could be a gamma distribution with
+
+$\alpha=n_0 + \frac{1}{2}$
+
+$\beta = 
+  \frac{1}{2}(\mu-\mu_0)^2 - 
+  \frac{n_0}{\tau0}$
+
+if there were the correct normalizing constants. Multiply inside the integral by the normalizing constant
+
+$\frac{(\frac{1}{2} (\mu-\mu_0)^2-\frac{n0}{\tau_0})^{n_0 + \frac{1}{2}}}{\Gamma(n_0+ \frac{1}{2})}$
+
+When you divide outside by the normalizing constant, you get
+
+$\frac{1}{\sqrt{2 \pi}}$
+$\frac
+  {(n_0 \tau_0)^{n_0}}
+  {\Gamma(n_0)}$
+$\frac
+  {\Gamma(n_0 + \frac{1}{2})}
+  {(\frac{1}{2} (\mu-\mu_0)^2 - 
+     \frac{n0}{\tau_0})^
+       {n_0 + \frac{1}{2}}}$
+
+This can also be shown to relate to a t-distribution
+
+$\frac
+    {\Gamma \left(\frac{\nu+1}{2} \right)}
+    {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)}
+  \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$
 

### 2023-03-28 text/23/exponential-hedging-hyperprior.Rmd
Cleaning up hedging hyperprior formulas

@@
 
-You have an informative prior on $\theta$ that has an inverse gamma distribution
+You have an informative prior on $\theta$ that has an inverse gamma distribution with prior parameters $\alpha_0$ and $\beta_0$.
 
-$g(\theta; \alpha, \beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}$
+$g(\theta; \alpha_0, \beta_0) = 
+  \frac
+    {\beta^\alpha_0}
+    {\Gamma(\alpha_0)}
+  \theta^{-\alpha_0-1}
+  e^{-\beta_0 / \theta}$
 
-and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of
+The mean of this prior distribution is
 
-$L(\mathbf{X})=\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$
+$\frac{\beta}{\alpha_0-1}$
+
+You collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of
+
+$L(\mathbf{X})
+  = \prod_{i=1}^n f(X_i)
+  = \theta^{-n} e^{-n \bar X / \theta}$
 
@@
 
-$h(\theta | \mathbf{X}) \propto \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}\theta^{-n} e^{-n \bar X / \theta}$
+$h(\theta|\mathbf{X}) \propto 
+  \frac
+    {\beta_0^{\alpha_0-1}}
+    {\Gamma(\alpha_0)}
+  \theta^{-\alpha_0-1}
+  e^{-\beta_0 / \theta}
+  \theta^{-n}
+  e^{-n \bar X / \theta}$
 
@@
 
-$h(\theta | \mathbf{X}) \propto \theta^{-\alpha-1-n}e^{-(\beta + n\bar X) / \theta}$
+$h(\theta | \mathbf{X}) \propto 
+  \theta^{-\alpha_0-1-n}
+  e^{-(\beta_0 + n\bar X) / \theta}$
 
-which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean
+which, thanks to the miracle of conjugancy, is an inverse gamma distribution. With posterior paramaters
 
-$\frac{\beta + n\bar X}{\alpha-1+n}$
+$\alpha_1 = \alpha_0 + n$
+
+$\beta_1 = \beta_0 + n\bar X$
+
+The posterior mean
+
+$\frac{\beta_0 + n\bar X}{\alpha_0-1+n}$
 
@@
 
-$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$
+$\Big(
+  \frac
+    {\alpha_0-1}
+    {\alpha_0-1+n}\Big)
+  \frac
+    {\beta_0}
+    {\alpha-1} + 
+  \Big(
+    \frac
+      {n}
+      {\alpha-1+n}\Big)
+  \bar X$
+
+### A simple numerical example
 

### 2023-03-28 text/23/normal-hedging-hyperprior-part-1.Rmd
Cleaning up hedging hyperprior formulas

--- 2023-03-14 text/23/normal-hedging-hyperprior-part-1.Rmd
@@
 
+$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+
 For this derivation, it is easier to write the density in terms of
 
-$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$
+$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-\frac{1}{2}\tau (X-\mu)^2}$
 
@@
 
-$\tau=\frac{1}{\sigma*2}$
+$\tau=\frac{1}{\sigma^2}$
 
-The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,
+The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance.
 
-Assume that $\tau$ is known and equal to $\tau_F$.
-A more general case, with $\tau$ unknown is described on a separate web page.
+Assume that $\tau$ is known and equal to $\tau_F$. Other cases with [mean known and precision unknown][sim2] and with [both mean and precision unknown][sim3] appear on different web pages.
+
+[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
+[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3
+
 For simplicity, we will drop any multiplicative constants not involving X or $\mu$
@@
 
-$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$
+$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau_0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau_0})^2}$
 
@@
 
-$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$
+$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau_0}$
 
@@
 
-This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.
+This corresponds to a fairly informative prior distribution. This distribution places $\mu$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.
 
@@
 The posterior distribution is not close to either the prior distribution or the likelihood.
-It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
+It places $\mu$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
 It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.
@@
   ylab="lambda",
-  theta = 135, 
+  mu = 135, 
   phi = 30, 

### 2023-03-28 text/23/normal-hedging-hyperprior-part-2.Rmd
More formula updates on hedging hyperprior

@@
 
-Consider a Bayesian analysis of an outcome variable that has a normal distribution.
+Consider a Bayesian analysis of an outcome variable that has a normal distribution with known mean and unknown precision.
+Assume that both the mean and the precision of the normal distribution are unknown.
+Other cases with [mean unknown and precision known][sim1] and with [both mean and precision unknown][sim3] appear on a different web pages.
 
-For this derivation, it is easier to write the density in terms of the precision
-
-$f(X; \mu, \theta) = \sqrt{\frac{\theta}{2 \pi}}e^{-\frac{1}{2}\theta (X-\mu)^2}$
-
-where
-
-$\theta=\frac{1}{\sigma*2}$
+[sim1]: http://new.pmean.com/normal-hedging-hyperprior-part1
+[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3
 
@@
 A more general case, with both the mean and the precision of the normal distribution unknown is also described on a separate web page.
-For simplicity, we will drop any multiplicative constants not involving X or $\theta$
+For simplicity, we will drop any multiplicative constants not involving X or $\tau$
 
-$f(X; \theta) \propto \sqrt{\theta}e^{-\frac{1}{2}\theta(X-\mu_F)^2}$
+$f(X; \tau) \propto
+  \sqrt{\tau}
+  e^{-\frac{1}{2}\tau(X-\mu_F)^2}$
 
@@
 
-$g(\theta; \alpha_0, \beta_0) = \frac{\beta_0^{\alpha_0}}{\Gamma(\alpha_0)} \theta^{\alpha_0-1} e^{-\beta_0 \theta}$
+$g(\tau; \alpha_0, \beta_0) =
+  \frac
+    {\beta_0^{\alpha_0}}
+    {\Gamma(\alpha_0)}
+  \tau^{\alpha_0-1}
+  e^{-\beta_0 \tau}$
 
@@
 
-$\beta_0 = n_0 \theta_0$
+$\beta_0=n_0\tau_0$
 
-and remove any multiplicative constants with respect to $\theta$ to get
+and remove any multiplicative constants with respect to $\tau$ to get
 
-$g(\theta; n_0, \theta_0) \propto \theta^{n_0-1} e^{-n_0 \theta_0 \theta}$
+$g(\tau; n_0, \tau_0) \propto \tau^{n_0-1} e^{-n_0 \tau_0 \tau}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto \theta^{n/2}e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$
+$L(\mathbf{X} | \mu) \propto
+  \tau^{n/2}
+  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$
 
@@
 
-$h(\theta | \mathbf{X}) \propto \theta^{n_0-1} e^{-n_0 \theta / \theta0} \theta^{n/2} e^{- \frac{1}{2} \theta\sum(X_i-\mu_0)^2}$
+$h(\tau | \mathbf{X}) \propto
+  \tau^{n_0-1}
+  e^{-n_0 \tau / \tau0}
+  \tau^{n/2}
+  e^{- \frac{1}{2} \tau\sum(X_i-\mu_0)^2}$
 
@@
 
-$h(\theta | \mathbf{X}) \propto \theta^{n/2 + n_0-1} e^{-(n_0 \theta0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$
+$h(\tau | \mathbf{X}) \propto
+  \tau^{n/2 + n_0-1}
+  e^{-(n_0 \tau0 + \frac{1}{2}\sum(X_i-\mu_0)^2)}$
 
@@
 
-$\beta_1 = n_0 theta_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$
+$\beta_1 = n_0 \tau_0 + \frac{1}{2}\sum(X_i-\mu_0)^2)$
 
-This can also be shown to relate to a t-distribution
-
-$\frac{\Gamma \left(\frac{\nu+1}{2} \right)} {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)} \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$
+### A simple numerical example
 
@@
 n0 <- 4
-theta0 <- 0.5
+tau0 <- 0.5
 
 alpha0 <- n0
-beta0 <- n0*theta0
+beta0 <- n0*tau0
 
@@
 
-x <- muF+ round(rnorm(n), 1)/theta0
+x <- muF+ round(rnorm(n), 1)/tau0
 
@@
 
-f <- function(x, mu, theta) {
-  sqrt(theta/2*pi)*exp(-0.5*theta*(x-mu)^2)
+f <- function(x, mu, tau) {
+  sqrt(tau/2*pi)*exp(-0.5*tau*(x-mu)^2)
 }
@@
 
-Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\theta_0$ = `r theta0`.
+Let's illustrate this with an informative prior with $n_0$ = `r n0` and $\tau_0$ = `r tau0`.
 
@@
 
-This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.
+This corresponds to a fairly informative prior distribution. This distribution places $\tau$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.
 
@@
 The posterior distribution is not close to either the prior distribution or the likelihood.
-It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
-It excludes the prior mean of `r theta0` and it excludes `r ss/(n-1)`, the mean of the data.
+It places $\tau$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability.
+It excludes the prior mean of `r tau0` and it excludes `r ss/(n-1)`, the mean of the data.
 
@@
 
-mean $\theta_0$ and prior precision
+mean $\tau_0$ and prior precision
 
@@
 
-$\beta = \alpha \theta_0 $
+$\beta = \alpha \tau_0 $
 
@@
 lambda_grid <- seq(0, 1, length=60)
-theta_grid <- seq(0, x_max, length=60)
+tau_grid <- seq(0, x_max, length=60)
 
-gstar <- function(theta, lambda) {g(theta, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
-density <- outer(theta_grid, lambda_grid, gstar)
+gstar <- function(tau, lambda) {g(tau, 1+lambda*(alpha0-1), (1+lambda*(alpha0-1))*beta0)}
+density <- outer(tau_grid, lambda_grid, gstar)
 op <- par(bg = "white") 
@@
 persp(
-  theta_grid, 
+  tau_grid, 
   lambda_grid, 
@@
   zlim=c(0, 1.1*z_max),
-  xlab="theta",
+  xlab="tau",
   ylab="lambda",
@@
 
-This is the joint prior distribution of $\theta$ and $\lambda$.
+This is the joint prior distribution of $\tau$ and $\lambda$.
 
-It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large or very small.
+It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\tau$ is very large or very small.
 This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.
@@
 ```{r, eval=FALSE}
-density <- outer(theta_grid, lambda_grid, gstar)
+density <- outer(tau_grid, lambda_grid, gstar)
 density[16:44, ] <- NA
@@
 
-The likelihood function is constant across all values of $\theta$, so its surface looks like the following.
+The likelihood function is constant across all values of $\tau$, so its surface looks like the following.
 
@@
 lambda_grid <- seq(0, 1, length=60)
-theta_grid <- seq(0, x_max, length=60)
+tau_grid <- seq(0, x_max, length=60)
 
-fstar <- function(mu, lambda) {f(xbar, mu, theta0)^n}
+fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
 likelihood <- outer(mu_grid, lambda_grid, fstar)
@@
 
-hstar <- function(mu, lambda) {g(mu, mu0, lambda*(theta0-thetaF) + thetaF)*f(xbar, mu, thetaF)^n}
+hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
 density <- outer(mu_grid, lambda_grid, hstar)
@@
   levels=c(50, 75, 90)/100,
-  xlab="theta",
+  xlab="tau",
   ylab="lambda")
@@
 
-You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r theta0`) is largely ignored in favor of the mean of the data (`r "??"`).
+You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r tau0`) is largely ignored in favor of the mean of the data (`r "??"`).
 

### 2023-04-04 text/23/non-standard-evaluation.Rmd
Added non-standard-evaluation

---
title: "Non-standard evaluation inside functions and loops"
source: "New"
date: "2023-04-04"
categories:
- Blog post
tags:
- R programming
output: html_document
---

```{r setup, echo=FALSE}
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(tidyverse)))
```

One of the big changes that the tidyverse made to the R programming language is the use of non-standard evaluation (NSE). This is sometimes described more explicitly as tidy evaluation. What NSE gives you is shorter and cleaner code by dispensing with quote marks and data frame prefixes in many situations. Unfortunately, NSE requires a bit of extra work when you are using it inside a loop or a function.

I want to briefly describe how NSE allows you to write shorter and cleaner code and then explain the adaptations that you need when using it inside functions and loops.

<!---more--->

### Dispensing with quote marks

Suppose you have a data frame with columns labelled w, x, y, and z. If you wanted to create a smaller data frame with just w and x, you could use the following code.

```{r}
old_data <- data.frame(
  w=0:3,
  x=1:4,
  y=2:5,
  z=3:6)
new_data <- old_data[ , c("w", "x")]
new_data
```

But `select`, a function in dplyr can do the same work without the quote marks.

```{r}
new_data <- select(old_data, w, x)
new_data
```

There are other nice features in select, such as the ability to select columns that meet certain criteria (such as `is.numeric`) or column names that start or end with a certain string.

### Dispensing with data frame prefixes

In a talk I attended many years ago, a really smart R programmer said that you should always keep the names of data frames very short. The reason, he explained, is that you end up typing the name of that data frame over and over and over.

Here's an example. Suppose you wanted to select just those rows in a data frame where the variable w is less than two. The r code to do this is

```{r}
new_data <- old_data[old_data$w < 2, ]
new_data
```

Notice how you had to type `old_data` twice? It is not unusual in R to maybe have to type that data frame name three or even four times in a single statement.

The `filter` function in dplyr simplifies the code.

```{r}
new_data <- filter(old_data, w < 2)
new_data
```

### What, exactly, is non-standard evaluation?

I wish I could explain NSE better. In particular, I cannot explain the mechanics behind NSE. There are some excellent references, but these are very abstract and difficult to read.

Brodie Gaslam. Standard and non-standard evaluation in R. Blog post, 2022-05-05. Available in [html format][gas1].

Chapters 17 to 21 of Hadley Wickham. Advanced R, Second edition. Available in [html format][wic1]

Hadley Wickham. Non-standard evaluation. Vignette for lazyeval, 2019-03-15. Available in [html format][wic2].

[gas1]: https://www.brodieg.com/2020/05/05/on-nse/
[wic1]: https://adv-r.hadley.nz/index.html
[wic2]: https://cran.r-project.org/web/packages/lazyeval/vignettes/lazyeval.html

### What is lazy evaluation?

Another feature of NSE in R is called lazy evaluation. Lazy evaluation can sometimes greatly improve the efficiency of a program by deferring evaluation of an argument inside a function until it is really needed. It is unclear to me why lazy evaluation is lumped in with NSE, but the reference listed below tries to explain this.

Chapter 6 in Greg Wilson. R for Data Engineers, 2021-01-11. Available in [html format][wil1].

[wil1]: https://tidynomicon.github.io/tidynomicon/nse.html

### Non-standard evaluation inside a loop.

Sometimes you want to do a repetitive task by using the `for` loop in R. Suppose, for example, that you wanted to calculate the median of each column in a data frame. The standard approach in R works easily.

```{r}
for (v in c("w", "x")) {
  print(table(old_data[ , v]))
}
```

But this does not work for `count`, the dplyr equivalent to table

```{r, error=TRUE}
for (v in c("w", "x")) {
  count(old_data, v)
}
```

The dplyr functions create a hidden data frame (`.data`) and you need to reference this hidden data.

```{r, error=TRUE}
for (v in c("w", "x")) {
  print(count(old_data, .data[[v]]))
}
```

### Using NSE inside a function

It gets tricky when you want to write a function that includes something like `select` or `filter`. Consider building a function to reduce the number of columns in a data frame.

It's pretty easy with standard R code.

```{r}
reduce <- function(old_data, variable_name) {
  old_data[ , variable_name]
}
reduce(old_data, "w")
```

But the following code won't work.

```{r, error=TRUE}
reduce <- function(old_data, variable_name) {
  select(old_data, variable_name)
}
reduce(old_data, w)
```

Putting the variable_name in quotes does work, but produces a warning.

```{r, error=TRUE}
reduce(old_data, "w")
```

If you want to use non-standard evaluation into a function, then you have to "embrace" the one or more of the function arguments. Embracing sounds like something romantic, but it actually just means putting a pair of curly braces around the variable.

```{r, error=TRUE}
reduce <- function(old_data, variable_name) {
  select(old_data, {{variable_name}})
}
reduce(old_data, w)
```

If you want to reduce to two or more columns, use the `...` function argument.

```{r, error=TRUE}
reduce <- function(old_data, ...) {
  select(old_data, ...)
}
reduce(old_data, w, x)
```

If you need to send strings to a function, use `.data[[ ]]`.

```{r, error=TRUE}
reduce <- function(old_data, variable_string) {
  select(old_data, .data[[variable_string]])
}
reduce(old_data, "w")
```

This only works for a single string. If you try to pass a vector of strings, you get an error.

```{r, error=TRUE}
reduce <- function(old_data, variable_string) {
  select(old_data, .data[[variable_string]])
}
reduce(old_data, c("w", "x"))
```

In this case, the dplyr library offer two functions, `any_of` and `all_of`.

```{r, error=TRUE}
reduce <- function(old_data, variable_string) {
  select(old_data, any_of(variable_string))
}
reduce(old_data, c("w", "x"))
```

The two functions work similarly, but differ in how they handle cases where strings in the vector do not match up with the column names.

### Some final thoughts

The complexities caused by NSE when used inside loops and functions does tend to defeat the purpose of NSE. You might be tempted to ditch the dplyr package (and other tidyverse packages like ggplot2) because they require confusing code hacks when used inside loops and functions.

This is an over-reaction, in my opinion. There are so many benefits to using dplyr in regular r code, that the added complexity when you need to incorporate it inside loops and functions is a small price to pay.

You always have the option of using the power and simplicity dplyr for most of your code and only falling back on earlier code when you need to do something inside a loop or function.

As for me, I plan to "embrace" the complexities associated with NSE. I like dplyr, ggplot2, and so many other of the tidyverse packages so much that I will find a way to remember things like `.data`, the `...` argument, and the `any_of / all_of` functions

### 2023-04-04 text/23/non-standard-evaluation.Rmd
Added an extra note to non-standard-evaluation

@@
 
+### A brief warning
+
+The example I show below are simple.
+It helps me to understand things better by working out simple examples.
+But the examples are so simple that you might ask yourself why bother with this, or why not try something even simpler.
+Bear with me, though, because the simple examples illustrate some general principles that will help you in more complex settings.
+
 ### Dispensing with quote marks
@@
 
-But `select`, a function in dplyr can do the same work without the quote marks.
+But `select`, a function in dplyr, can do the same work without the quote marks.
 

### 2023-04-04 text/23/non-standard-evaluation.Rmd
Minor touch-up of non-standard-evaluation

@@
 
-You always have the option of using the power and simplicity dplyr for most of your code and only falling back on earlier code when you need to do something inside a loop or function.
+You always have the option of using the power and simplicity of dplyr for most of your code and only falling back on earlier code when you need to do something inside a loop or function.
 
-As for me, I plan to "embrace" the complexities associated with NSE.
-I like dplyr, ggplot2, and so many other of the tidyverse packages so much that I will find a way to remember things like `.data`, the `...` argument, and the `any_of / all_of` functions
+As for me, I like dplyr, ggplot2, and so many other of the tidyverse packages so much that I will plan to "embrace" coding hacks like `.data`, the `...` argument, and the `any_of / all_of` functions

### 2023-05-19 text/22/analysis-of-means.Rmd
Added ai-fairness.

--- 2022-09-06 text/22/analysis-of-means.Rmd
@@
 ---
-title: "A second alternative to the pairwise testing approach"
+title: "A simple visual alternative to all pairwise comparisons"
 author: "Steve Simon"
@@
 
-In an earlier blog post, I mentioned Dunnett's test, a comparison of multiple treatments to a control group as an alternative to the pairwise approach to testing that is commonly used in analysis of variance.
-Here's a second interesting alternative.
-
-### Comparison to an overall mean
-
-You are in a setting where you are expecting to see a degree of uniformity among several groups.
-For example, you have a group of clinics who all treat the same type of patients.
-The time spent with each patient, the total cost of the medications prescribed, the length of the notes written in the medical record--these can differ from patient to patient.
-But because of how the patients are assigned to a particular clinic (more or less randomly), these should balance out on average.
-
-If there is one clinic that has different average results, it has to be because that the physicians at that clinic have a different approach to treatment.
-If you can identify a clinic (it could be more than one clinic, of course) that differs from the norm, then investigate.
-Perhaps that clinic is doing something "wrong" that causes the discrepancy, or maybe that clinic is the only one doing things right.
-In any case understanding why you see unexpected deviations in a process that should be uniform across clinics provides you with an opportunity to learn something.
+The classic follow-up to a significant F test in analysis of variance is examining all pairwise differences in means.
+It's great for many problems, but sometimes you may not want to focus on all pairwise differences.
+Sometimes, there are so many pairwise differences that you get lost in all the clutter.
+Other times, you are trying to find out if, like the Sesame Street song, one of these groups is not like the others.
+Perhaps, you are trying to assure that groups conform to a common standard.
+Analysis of Means (ANOM) helps in all these settings.
 
@@
 
-This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means." It fits in well with a setting where you hope that all the groups produce consistent results.
+This approach is not that common, but you can find it in most statistical software programs under the name "Analysis of Means" or its acronym, ANOM.
+It fits in well with a setting where you hope that all the groups produce consistent results.
 If any do not, then you want to identify the group or groups that deviate from the norm and study factors that may account for the deviation.
 
-The analysis of means approach compares each group mean to the overall mean.
+The ANOM approach compares each group mean to the overall mean.
 It is easy to implement and it lends itself to a simple graphical display.
@@
 
-Five genetically modified strains of eggplant were tested for presence of a natural insecticide, CryA1c, in the leaves of the plant.
-Four plants were assessed for each strain.
-Here are the descriptive statistics.
-The research question is whether any strain differs from the overall mean level of CryA1c.
+Imagine that you are a crop scientist, comparing five genetically modified strains of eggplant. You developed these strains to produce a natural insecticide, CryA1c, in the eggplant leaves.
+
+You hope that all strains produce a comparable amount of insecticide.
+If any strain produces a lot more or a lot less insecticide, you will investigate full genetic profile of that strain to see why it is different from the others.
 
@@
 (Image taken from Wikipedia)
+
+The data comes from
 
@@
 
-https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498
+<!--- https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0157498 --->
+
+and my description of the testing scenario is a bit simpler than the analysis done in that paper.
 
@@
 
-The formulas get a bit more complex and the graph gets a bit messier if the sample size differs from one group to another.
+The graph gets a bit messier if the sample size differs from one group to another.
 
 More importantly, the only comparisons that you can make with the analysis of means approach is a comparison to the overall mean.
-If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control over the overall Type I error rate.
+If two groups are both statistically significantly higher than the overall mean, you cannnot make a comparison between those two groups without losing control of the overall Type I error rate.
 
@@
 
-The Analysis of Means, just like the Dunnett's test described in an earlier blog post, provides a simple approach to testing a different type of hypothesis.
+The Analysis of Means provides a simple approach to testing a different type of hypothesis.
 Because it reduces the number of comparisons from all possible pairwise differences to a comparison of each group to the overall mean, you gain some precision and can summarize your results in a simple easily understood graph.

### 2023-05-24 src/check_links.Rmd
Added links to blog pages from 2014

@@
 html_path <- "c:/Users/steve/Dropbox/professional/fun stuff/update-website/"
-html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
+# html_path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website/"
 html_pattern <- "html$"

### 2023-05-24 src/check_web.Rmd
Added links to blog pages from 2014

---
title: "Check web files for links to new pages"
author: "Steve Simon"
date: "Created 2022-05-26"
output: html_document
---

### Setup

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo=TRUE)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(tidyverse)))
```

### Read html files

```{r read_old}
path_prefix <- "c:/Users"
path_suffix <- "dropbox/professional/fun stuff/update-website"

list.files(path_prefix) %>%
  str_subset("(Simons)|(steve)") -> me

full_path <- glue("{path_prefix}/{me}/{path_suffix}")

subdir_list=c("99", paste0("0", 0:9), 10:22)
html_files <- NULL
for (yr in rev(subdir_list)) {
  file_path <- paste0(full_path, "/", yr)
  yr_files <- list.files(file_path, pattern="html")
  if (length(yr_files)==0) next
  for (fn in yr_files) {
    read_lines(paste0(file_path, "/", fn)) %>%
      str_subset('<p>This page (is moving)|(has moved)') %>%
      str_remove('.*<a href="http://new.pmean.com/') %>%
      str_remove('/">.*') %>%
      str_remove('<p>This page is being ') %>%
      str_remove('.</p>') %>%
      str_trim -> link_file
    if (length(link_file)==0) link_file <- "NOTHING"
    cat(glue("\n\n{yr}/{fn}"))
    cat(glue(" links to {link_file}."))
  }
}
```

### 2023-07-24 src/year-end-summary.Rmd
Updated year-end-summary

@@
 
-update <- FALSE
+update <- TRUE
 print("Check 1")
@@
 }
+```
 
+```{r}
 # Check date
@@
 # Check title
+```
 
+```{r}
 # Create links for each file

### 2023-07-30 text/08/declining-count-model.Rmd
More work on 2008 update pages

@@
 tags:
+- Being updated
 - Poisson regression

### 2023-08-31 src/year-end-summary.Rmd
Updated year-end-summary

@@
 check_source <- function(file_text) {
-  message <- NULL
   file_text %>%
@@
   if (length(file_source)==0) {
-    message <- glue("{yy}/{fn} has no source\n")
+    file_source <- "not listed"
   }
@@
     message <- glue("{yy}/{fn} has multiple sources\n")
+    file_source <- paste(file_source, collapse=", ")
   }
+  message <- glue(", original source is {file_source}.")
+  if (tolower(file_source)=="new") message <- "."
   return(message)
@@
 # Create links for each file
-create_file_list <- function(f, date_list) {
+create_file_list <- function(f, date_list, n_lines, source_files) {
   f %>%
@@
   o <- order(date_list, f)
-  glue("+ {date_list[o]} [{short_name[o]}](http://new.pmean.com/{short_name[o]}/)  ") %>%
+  glue(
+      "+ {date_list[o]} ",
+      "[{short_name[o]}]",
+      "(http://new.pmean.com/",
+      "{short_name[o]}/),  ",
+      "{n_lines[o]} lines",
+      "{source_files}") %>%
     paste(collapse="  \n")
@@
 yyyy_list <- c(1999:2023)
-n_problems <- 0
 n_files <- 0
@@
     str_subset("md$") -> f
-  problem_files <- NULL
+  source_files <- NULL
   date_list <- NULL
+  n_lines <- NULL
   for (fn in f) {
     file_text <- read_lines(glue("../text/{yy}/{fn}"))
-    problem_files %<>% append(check_source(file_text))
+    source_files %<>% append(check_source(file_text))
     date_list %<>% append(check_date(file_text))
+    n_lines %<>% append(length(file_text))
   }
 
-  # Update if no problems found
-  date_list %>% 
-    str_subset(" for ") %>%
-    append(problem_files) -> problem_files
-
-  n_year <- length(problem_files)  
-  n_problems <- n_problems + n_year
   n_files <- n_files + length(f)
-  if (n_year==0) {
-    problem_files <- "All files passed date and source quality checks"
-  }
-
-  if (n_year > 0) {print(problem_files)}
-
   cat("\n")
@@
 
-  create_yaml(yyyy) %>% write_lines(output_file)
+  create_yaml(yyyy) %>%
+    write_lines(output_file)
   write_lines(" ", output_file, append=TRUE)
 
-  create_links(yyyy_list) %>% write_lines(output_file, append=TRUE)
+  create_links(yyyy_list) %>%
+    write_lines(output_file, append=TRUE)
   write_lines(" ", output_file, append=TRUE)
 
-  glue("New website ({length(f)} files)") %>% write_lines(output_file, append=TRUE)
+  glue("New website ({length(f)} files)") %>%
+    write_lines(output_file, append=TRUE)
   write_lines(" ", output_file, append=TRUE)
 
-  create_file_list(f, date_list) %>% write_lines(output_file, append=TRUE)
+  create_file_list(f, date_list, n_lines, source_files) %>% 
+    write_lines(output_file, append=TRUE)
   write_lines(" ", output_file, append=TRUE)
 
-  problem_files %>% write_lines(output_file, append=TRUE)
-
-  glue("{yyyy}.md created, {length(f)} files, {n_year} problems detected") %>% print
+  glue("{yyyy}.md created.") %>% print
 }
 
-print(glue("All years: {n_problems} problems identified in {n_files} files"))
 ```

### 2023-09-01 text/00/what-is-kaplan-meier-curve.Rmd
Converted 00/what-is-kaplan-meier-curve to Rmd format

---
title: What is the Kaplan Meier curve?
source: http://www.pmean.com/00/kaplan.html
author: Steve Simon
date: 2000-06-27
categories:
- Blog post
tags:
- Ask Professor Mean
- Survival analysis
output: html_document
---

*Dear Professor Mean, When I read my medical journals, I keep on coming across terms like Kaplan-Meier Product Limit estimate or Kaplan-Meier survival curve. What do these terms mean and when are they used?*

<!---More--->

Dear Reader,

Often we want to measure how long it takes for something to occur. The most common (and the most morbid) example is how long it takes for someone to die. For this outcome, we want to estimate the fraction of patients who survive for at least one month, at least three months, etc. This estimate is known as a survival curve.

The term survival is sometimes misleading, because we can use it for other less severe outcomes like how long until a cancer relapse, or how long until an infection occurs. Sometimes it can even be used for a positive outcome, like how long it takes for a couple to conceive. But for the rest of this example, we'll keep things simple by assuming that the outcome is time until death.

Estimating a survival curve is often complicated by the uncooperative way in which research subjects sometimes behave. For example, some subjects decide to leave a study part of the way through. Others refuse to die before the study ends. We label these uncooperative subjects as censored observations. They survived for at least three months, but then we lost touch with them. Or they survived at least three years, but then we had to terminate the study.

**Short explanation**

The Kaplan-Meier estimate is a simple way of computing the survival curve in spite of all these troublesome research subjects. It involves computing the number of people who died at a certain time point, divided by the number of people who were still in the study at that time. We multiply these probabilities by any earlier computed probabilities, which is one reason this is called a "product limit estimate."

The Kaplan-Meier survival curve is often illustrated graphically. It looks like a poorly designed staircase, with vertical steps downward at the time of death of each individual subject.

Often we will compare curves for two different groups of subjects. For example, what is the survival pattern for subjects on a standard therapy compared to a newer therapy. We can look for gaps in these curves in a horizontal or vertical direction. A vertical gap means that at a specific time point, one group had a greater fraction of subjects surviving. A horizontal gap means that it took longer for one group to experience a certain fraction of deaths.

**More details**

To compute a survival curve, you need to note the time of occurrence of events (e.g., failures, deaths). It is possible for two or more events to occur at the same time, in which case the number of distinct times is less than the number of deaths or failures. You need to place the t's in order from smallest to largest. That is,

$t_1 < t_2 <  t_3 < ...$

You also need to define the starting point of the study,

$t_0=0$

The basic computations for the Kaplan-Meier survival curve rely on the computation of conditional survival probabilities. In particular, the probability

$P[T > t_i | T > t_{i-1}]$

which can be interpreted as the probability of your surviving to a specific time, given that you survived to the previous time. This probability is easy to calculate if you know the number of deaths or failures at a specific time and if you know the number of patients at risk at that same time.

A more difficult (but more important) probability is the unconditional probability of survival,

$P[T > t_i]$

which represents the simple probability of survival to a specific time. You can use a relationship between this unconditional probability and the conditional probability:

$P[T > t_i]$

$\ =\ P[T > t_i | T > t_{i-1}]$

$\qquad \times \  P[T > t_{i-1}]$

At first glance, this does not seem to help, because the right hand
side of the equation still includes an unconditional probability. But
you can apply this approach again to get

$P[T > t_i]$

$\ =\ P[T > t_i | T > t_{i-1}]$

$\qquad \times \  P[T > t_{i-1} | T > t_{i-2}]$

$\qquad \times \  P[T > t_{i-2}]$

and you can continue along these lines to get

$P[T > t_i]$

$\ =\ P[T > t_i | T > t_{i-1}]$

$\qquad \times \  P[T > t_{i-1} | T > t_{i-2}]$

$\qquad \times \  P[T > t_{i-2} | T > t_{i-3}]$

$\qquad \times \  ... \ \times\  P[T > t_0]$

This last probability represents the probability of surviving at the start of the study. Unless we intentionally recruit dead subjects, this probability has to be 1. Therefore, the unconditional probability is equal to the cumulative product of conditional probabilities.

At each time point, you should count $d_i$, the number of deaths at time $t_i$ and $c_i$, the number of censored values at time $t_i$.

Armed with this information, you can now compute a Kaplan-Meier survival curve. First you need to calculate the number of patients at risk,

$n_i=n_{i-1}-c_i-d_i$

In other words, the number at risk at any specific time point is just the number at risk at the previous time point, minus the number of deaths/failures and the number of censored observations. For convenience, you should define $n_0$ as the number of patients in the study, $c_0$ as the number of censored observations prior to the first death, and $d_0$ as the number of deaths at the start of the study, which is always zero.

Next you extimate the conditional probability of survival:

$P[T > t_i | T > t_{i-1}]=1-\frac{d_i}{n_i}$

Finally, the unconditional probability of survival is simply the cumulative product of the conditional probabilities:

$P[T > t_i]=\prod\big(1-\frac{d_i}{n_i}\big)$

**Example**

The following example is from Chadha et al (2000). The authors studied a sample of 36 pediatric patients undergoing acute peritoneal dialysis through Cook Catheters. They wished to examine how long these catheters performed properly. They noted the date of complication (either occlusion, leakage, exit-site infection, or peritonitis).

```{}
Day Removed Failed
--- ------- ------
 1     8      2
 2     2      2
 3     1      1
 4     1      1
 5     5      3
 6     0      2
 7     0      1
10     0      2
12     0      2
13     0      1
```

If you wanted to estimate the probability that a catheter will survive its first day, that's easy. There were 34 catheters, 2 did not survive the first day, 15 failed on days 2-13. For 17 of the catheters, you did not know when they would have failed, but you do know that they all survived at least one day.

So the probability of surviving the first day is 32/34 = 94%.

But how would you estimate the probability of surviving two days? four days? ten days?

This is tricky, because the censored observations provide information up to the day of censoring, but cannot tell us anything more about surviving beyond that day. What you need to do is compute the number of catheters at risk for failure on a particular day. It would exclude any catheters that failed on previous days and it would exclude any catheters that were censored on previous days.

```{}
Day Removed Failed   At risk
--- ------- ------  ---------
 1     8      2            34
 2     2      2     34-8-2=24
 3     1      1     24-2-2=20
 4     1      1     20-1-1=18
 5     5      3     18-1-1=16
 6     0      2     16-5-3= 8
 7     0      1      8-0-2= 6
10     0      2      6-0-1= 5
12     0      2      5-0-2= 3
13     0      1      3-0-2= 1
```

You then need to compute the conditional probability of surviving at each time point given that the catheter survived the previous time point. This conditional probability would be

(number at risk - number of failures)/(number at risk)

```{}
                              Conditional
Day Removed Failed   At risk  probability
--- ------- ------  --------- -----------
 1     8      2            34  32/34=0.94
 2     2      2     34-8-2=24  22/24=0.92
 3     1      1     24-2-2=20  19/20=0.95
 4     1      1     20-1-1=18  17/18=0.94
 5     5      3     18-1-1=16  13/16=0.81
 6     0      2     16-5-3= 8   6/ 8=0.75
 7     0      1      8-0-2= 6   5/ 6=0.83
10     0      2      6-0-1= 5   3/ 5=0.60
12     0      2      5-0-2= 3   1/ 3=0.33
13     0      1      3-0-2= 1   0/ 1=0.00
```

Then we compute the cumulative product of these probabilities. This represents the Kaplan-Meier estimate of the survival probability.

```{}
                              Conditional     Cumulative
Day Removed Failed   At risk  probability      product  
--- ------- ------  --------- -----------   --------------
 1     8      2            34  32/34=0.94             0.94
 2     2      2     34-8-2=24  22/24=0.92   0.94*0.92=0.86
 3     1      1     24-2-2=20  19/20=0.95   0.86*0.95=0.82
 4     1      1     20-1-1=18  17/18=0.94   0.82*0.94=0.77
 5     5      3     18-1-1=16  13/16=0.81   0.77*0.81=0.62
 6     0      2     16-5-3= 8   6/ 8=0.75   0.62*0.75=0.46
 7     0      1      8-0-2= 6   5/ 6=0.83   0.46*0.83=0.38
10     0      2      6-0-1= 5   3/ 5=0.60   0.38*0.60=0.23
12     0      2      5-0-2= 3   1/ 3=0.33   0.23*0.33=0.08
13     0      1      3-0-2= 1   0/ 1=0.00   0.08*0.00=0.00
```

Here is a graph of these survival probabilities.

![](http://www.pmean.com/images/00/kaplan01a.gif){width="476" height="389"}

The plot has a "stair step" pattern, because we don't know the survival probability at fractional days (such as 2.5 days) and at some integer days (such as 9 days). By convention, we estimate the survival probability for these values as equaling the survival probability of the closest value that is still smaller (the 2 day survival probability for 2.5 days, and the 7 day survival probability at 9 days).

Notice that the estimated median survival time (the time at which 50% of the catheters survived) is six days.

**Reference**

Chada V, Warady BA, Blowey DL, Simckes AM, Alon US. Tenckhoff Catheters Prove Superior to Cook Catheters in Pediatric Acute Peritoneal Dialysis *American Journal of Kidney Diseases* (2000), 35(6):1111-1116.

**Further reading**

There are many beginning level books on biostatistics that discuss the Kaplan-Meier curve, such as Woolson's book. You can find a more advanced and detailed approach in Collett's book.

+ Collett D. Modelling Survival Data in Medical Research. London England: Chapman and Hall (1994). ISBN: 0-412-44890-4.
+ Woolson RF. Statistical Methods for the Analysis of Biomedical Data. New York NY: John Wiley & Sons, Inc. (1987). ISBN: 0-471-80615-3.

You can find an [earlier version][sim1] of this page on my [original website][sim2].

[sim1]: http://www.pmean.com/00/kaplan.html
[sim2]: http://www.pmean.com/original_site.html

### 2023-09-03 text/23/taylor-series-approximation.Rmd
Added taylor-series-approximation

---
title: "Using a linear Taylor series approximation to estimate rounding error"
source: "New"
author: Steve Simon
date: 2023-09-02
categories:
- Blog post
tags:
- Mathematics
output: html_document
---

When I run, I have an app that announces the distance traveled (in kilometers) at the end of every minute. But since the app rounds to one decimal place, there is a bit of uncertainty in the pace that I might compute. Here's how I would address this uncertainty using a Taylor series approximation.

<!---more--->

```{r, echo=FALSE}
d0 <- 2.5
t0 <- 20
pace0 <- t0 / d0
d1 <- d0 - 0.05
d2 <- d0 + 0.05
pace1 <- round(t0 / d1, 4)
pace2 <- round(t0 / d2, 4)
approx1 <- round(t0/d0-(d1-d0)*t0/(d0^2), 4)
approx2 <- round(t0/d0-(d2-d0)*t0/(d0^2), 4)
```

Pace is the inverse of speed. You divide the time in minutes by the distance traveled. The math is usually pretty easy. It is a simple division of two numbers both of which (for my runs) are in the double digits.

### Simple example of the linear Taylor series approximation

For example, suppose my GPS told me that I had just finished `r d0` kilometers in `r t0` minutes. I'm actually slower than that, but I used to run at that pace. The pace is

$\frac{`r t0`}{`r d0`}=`r pace0`$ minutes per kilometer.

My app rounds distance to one decimal, so that denominator of `r d0` might be as small as `r d1` or as large as `r d2`. I could compute the two ratios,

$\frac{`r t0`}{`r d1`}=`r pace1`$ and

$\frac{`r t0`}{`r d2`}=`r pace2`$

but dividing by a three digit number is tricky. I don't carry a calculator and I don't have the opportunity to write out the calculation on a pad of paper.

Taylor series approximation to the rescue! The linear Taylor series approximation is

$f(x) \approx f(a)+(x-a)f'(a)$

This approximation is pretty good in most cases. You have to have the value of a be not too far from the value of x. You also need the function f to be reasonably well behaved. This means no extreme fluctuations and no sudden transitions in shape.

This is certainly the case for the function

$f(x)=\frac{`r t0`}{x}$

The derivative is

$f'(x)=-\frac{`r t0`}{x^2}$

So the linear Taylor series approximation is

$f(`r d1`) \approx \frac{`r t0`}{`r d0`}-(`r d1`-`r d0`)\frac{`r t0`}{`r d0`^2}=`r approx1`$ and

$f(`r d2`) \approx \frac{`r t0`}{`r d0`}-(`r d2`-`r d0`)\frac{`r t0`}{`r d0`^2}=`r approx2`$ minutes per kilometer

Notice that you do divide by nice round numbers like `r d0` and $`r d0`^2$ rather than `r d1` and `r d2`.

### Graphical illustation of the linear Taylor series approximation

Here's a graphical illustration of the linear Taylor series approximation. The green line represents pace as it actually should be calculated and the red line represents the linear Taylor series approximation.

```{r, echo=FALSE}
library(ggplot2, quiet=TRUE)
d <- seq(d0-1, d0+1, length=501)
pace <- t0/d
truth <- data.frame(d, t0/d)
approx <- data.frame(d, pace=t0/d0 - (d-d0)*t0/d0^2)

ggplot(data.frame(d, pace)) +
  geom_line(aes(d, pace), col="darkgreen") +
  geom_line(data=approx, aes(d, pace), color="red")
```

The approximation is good when the distance is close to `r d0` kilometers, but does poorly further away. Fortunately, we are interested in this approximation at `r d1` and `r d2`.

Let's zoom in.

```{r, echo=FALSE}
library(ggplot2, quiet=TRUE)
d <- seq(d0-0.1, d0+0.1, length=501)
pace <- t0/d
truth <- data.frame(d, t0/d)
approx <- data.frame(d, pace=t0/d0 - (d-d0)*t0/d0^2)

ggplot(data.frame(d, pace)) +
  geom_line(aes(d, pace), col="darkgreen") +
  geom_line(data=approx, aes(d, pace), color="red") +
  geom_segment(x=d1+0.02, y=pace1+0.05, xend=d1, yend=pace1, color="darkgreen") +
  geom_segment(x=d2+0.02, y=pace2+0.05, xend=d2, yend=pace2, color="darkgreen") +
  geom_label(x=d1+0.02, y=pace1+0.05, label=pace1, color="darkgreen") +
  geom_label(x=d2+0.02, y=pace2+0.05, label=pace2, color="darkgreen") +
  geom_segment(x=d1-0.01, y=approx1-0.05, xend=d1, yend=approx1, color="red") +
  geom_segment(x=d2-0.01, y=approx2-0.05, xend=d2, yend=approx2, color="red") +
  geom_label(x=d1-0.01, y=approx1-0.05, label=approx1, color="red") +
  geom_label(x=d2-0.01, y=approx2-0.05, label=approx2, color="red")
```

### General application of the linear Taylor series approximation

Let's look at a more general case. You run a total of d kilometers in time t, but because of rounding, d could be a bit smaller (d-0.05) or a bit larger (d+0.05). Treating t as constant, define

$f(d)=\frac{t}{d}$

Then, the linear Taylor series approximation is

$f(d + 0.05) \approx \frac{t}{d} - 0.05\frac{t}{d^2}$ and

$f(d - 0.05) \approx \frac{t}{d} + 0.05\frac{t}{d^2}$.

You can factor this to

$f(d + 0.05) \approx \frac{t}{d}\Big(1 - \frac{0.05}{d}\Big)$ and

$f(d - 0.05) \approx \frac{t}{d}\Big(1 + \frac{0.05}{d}\Big)$

So after running `r d0` kilometers, the rounding error causes an inaccuracy in the pace calculation by about 0.05 / `r d0` or `r 100*0.05/d0`% in either direction. After running `r 2*d0` kilometers, the rounding error causes an inaccuracy of no more than 0.05 / `r 2*d0` or `r 100*0.05/(2*d0)`% in either direction. Notice that the signs flip. If you run slightly less than `r d0` kilometers, your pace is higher (slower). If you run slightly more than `r d0` kilometers, your pace is lower (faster).

You can state things in an even greater degree of generality. The rounding error in distance (as a percentage) translates directly to the rounding error in pace. This is an approximation, but it turns out to be a pretty good approximation as long as the percentage rounding error is small.

### 2023-09-18 text/23/too-many.rmd
Added maps to too-many

---
title: Too many colors spoil the graph
source: "New"
author: Steve Simon
date: 2023-09-12
categories:
- Blog post
tags:
- Being updated
output: word_document
---

When you draw a graph, either a bar chart, a scatterplot, or even a pie chart, you have the choice of a broad range of colors that you can use. R, for example, has 657 different colors from aliceblue to yellowgreen. SAS has 13 shades of orange, 33 shades of blue, and 47 shades of green. They even have different shades of black. You have a wealth of colors, but you can't sue all of them in the same graph. The ideal number of colors is 2.

<!---more--->

Here's a brief exercise. It is adapted from an example of Todd Iverson and Silas Bergen in a course they teach on data visualization. Set a time and count the number of 5's in this graph.

![](http://www.pmean.com/new-images/23/too-many-02.png)

It took a bit of time. I tried to help you by using a different color for each number. I hope it helped. Let's try that exercise again. Count the number of fives.one more time.

![](http://www.pmean.com/new-images/23/too-many-03.png)

Were you faster were you this time? Much faster, I'm guessing, and it's because only two colors were used.

Use two colors when you want to draw emphasis to a particular subgroup on your graph. Okay, you might claim foul here. It's easier to pick the 5's, but not so easy to pick the 4's and 7's. Fair enough, but this is something you should always consider when drawing a graph.

What elements of the graph do you want to emphasize? You can't emphasis some elements without de-emphasizing others.  If you use a different color for each group, you end up emphasizing none of the groups.

Here's another lesson that I learned from Naomi Robbins, another expert on data visualization. She points out that you don't make every word in a sentence a different color. You use a single color to draw emphasis to a particular word or phrase. The same should be true for your graphs.

Restraint is also important for gradients. Often in a graph, you want to show changes in a continuous variable across a range of colors.

```{r, echo=FALSE, warning=FALSE}
library(usmap)
library(ggplot2)
plot_usmap(
    data = statepov, 
    values = "pct_pov_2014", 
    color = "red") -> us
us +
  scale_fill_continuous(
        low = "white", 
        high = "red", 
        name = "Poverty rate (2014)", 
        label = scales::comma) + 
  theme(legend.position = "right")
```

Here's a map with a transition from a very light red to a very dark red.

```{r, echo=FALSE, warning=FALSE}
us +
  scale_fill_continuous(
        low = "green", 
        high = "blue", 
        name = "Poverty rate (2014)", 
        label = scales::comma) + 
  theme(legend.position = "right")
```

You can even show a transition between three different colors. This transition from green to yellow to red is used commonly in genomic data visualizations.

```{r, echo=FALSE, warning=FALSE}
gr <- c("green", "yellow", "red")
us +
  scale_fill_gradientn(
        colors=gr, 
        name = "Poverty rate (2014)", 
        label = scales::comma) + 
  theme(legend.position = "right")
```

But what you can't do is use all the colors in the rainbow.

![](http://www.pmean.com/new-images/23/rainbow-gradient.jpg)

This image shows the rainbow gradient. It transitions between red, yellow, green, cyan, blue, and magenta. These are the fundamental colors on your computer display. Although the rainbow gradient is used a lot in computer visualizations, it is widely derided by visualization experts.

Here's what a map of poverty rates looks like with the rainbow gradient.

```{r, echo=FALSE, warning=FALSE}
gr <- c("red", "yellow", "green", "cyan", "blue", "magenta")
us +
  scale_fill_gradientn(
        colors=gr, 
        name = "Poverty rate (2014)", 
        label = scales::comma) + 
  theme(legend.position = "right")
```

There are many problems with the cycle of rainbow colors.

First, the ends of the rainbow do not show enough contrast. In the above map, the low poverty rate of New Hampshire is a red that differs from high poverty rate of Mississippi shown as magenta.

Second, the transitions in the rainbow gradient are uneven. This leads to a banding effect which distorts the continuous scale.

Third, the brightness of the colors varies widely across the rainbow gradient.

Fourth, the colors in the rainbow gradient can sometimes come across as harsh. This is subjective, of course, but colors that are a bit lighter or a bit darker than the rainbow often are easier on the eyes.

Your computer can display over 16 million colors on the screen. The temptation is to cram a lot of different colors into a single graph. But a bit of discretion is important here. Less if often more when it comes to colors in your data visualization.

### 2023-09-19 text/23/too-many.rmd
Minor edits to too-many

@@
 
-Here's a brief exercise. It is adapted from an example of Todd Iverson and Silas Bergen in a course they teach on data visualization. Set a time and count the number of 5's in this graph.
+Here's a brief exercise. It is adapted from an example of Todd Iverson and Silas Bergen in a course they teach on data visualization. Set a timer and count the number of 5's in this graph.
 
@@
 
-It took a bit of time. I tried to help you by using a different color for each number. I hope it helped. Let's try that exercise again. Count the number of fives.one more time.
+How long did it take? I tried to help you by using a different color for each number. Let's try that exercise again. Count the number of fives in the image below.
 
@@
 
-Here's a map with a transition from a very light red to a very dark red.
+Here's a map showing poverty rates for each state in the United States. It uses a gradient from a very light red to a very dark red.
 
@@
 
-First, the ends of the rainbow do not show enough contrast. In the above map, the low poverty rate of New Hampshire is a red that differs from high poverty rate of Mississippi shown as magenta.
+First, the ends of the rainbow do not show enough contrast.
+In the above map, the low poverty rate of New Hampshire (9.2%) is a red that differs only slightly from high poverty rate of Mississippi (21.9%) shown as magenta.
+If anything, the contrast is bigger between Mississippi (magenta) and Alabama (blue) even though the poverty rates of 19.2% for Alabama is a lot closer than New Hampshire's.
 
-Second, the transitions in the rainbow gradient are uneven. This leads to a banding effect which distorts the continuous scale.
+Second, the transitions in the rainbow gradient are uneven.
+Look at how broad the transitions among the greens and how narrow the transitions among the yellows.
+This leads to a banding effect which distorts the continuous scale.
 
 Third, the brightness of the colors varies widely across the rainbow gradient.
+Yellow, in particular, is a very bright color and can get lost against a white background.
+Green, Magenta, and Cyan are also rather bright.
+In contrast, blue and red are much darker than the other colors.
+In general, you want either to keep the brightness constant or vary it systematically.
+The brightness of the rainbow gradient staggers haphazardly up and down.
 

### 2023-10-10 text/20/poor-color-choice.Rmd

--- 2023-01-11 text/20/poor-color-chioce.Rmd
@@
 
-![](http://www.pmean.com/images/20/misuse-of-gradient.png)
+![](http://www.pmean.com/new-images/20/poor-color-choice-01.png)
 
@@
 angle_seq <- (1:n)/n
-df0<- data.frame(x=delta*sin(2*pi*angle_seq), y=delta*cos(2*pi*angle_seq))
+df0<- data.frame(
+    x=delta*sin(2*pi*angle_seq),
+    y=delta*cos(2*pi*angle_seq))
 sb <- (n/k)*(1:k)
@@
 ggplot(df0, aes(x, y)) +
-  geom_point(color=hsv(angle_seq, 1, 1), size=2, shape=16) +
-  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 1), size=24, shape=16) +
-  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
+  geom_point(
+      color=hsv(angle_seq, 1, 1), 
+      size=2, 
+      shape=16) +
+  geom_point(
+      data=df1, 
+      aes(x, y), 
+      color=hsv(sb/n, 1, 1), 
+      size=24, 
+      shape=16) +
+  expand_limits(
+      x=c(-1, 1), 
+      y=c(-1, 1)) +
   theme_void() +
   theme(
-    panel.background=element_rect(fill="#808080"))
+      panel.background =
+          element_rect(fill="#808080"))
 ```
@@
 ggplot(df0, aes(x, y)) +
-  geom_point(color=hsv(angle_seq, 1, 0.5), size=2, shape=16) +
-  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 0.5), size=24, shape=16) +
-  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
+  geom_point(
+      color=hsv(angle_seq, 1, 0.5), 
+      size=2, 
+      shape=16) +
+  geom_point(
+      data=df1, 
+      aes(x, y), 
+      color=hsv(sb/n, 1, 0.5), 
+      size=24, 
+      shape=16) +
+  expand_limits(
+      x=c(-1, 1), 
+      y=c(-1, 1)) +
   theme_void() +
   theme(
-    panel.background=element_rect(fill="#808080"))
+      panel.background = 
+          element_rect(fill="#808080"))
 ```
+
+You might also consider lighter colors. These are pastel colors. Going lighter or going darker will depend on many factors, but either choice is usually better than using the pure colors.
 
@@
 ggplot(df0, aes(x, y)) +
-  geom_point(color=hsv(angle_seq, 0.5, 1), size=2, shape=16) +
-  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 0.5, 1), size=24, shape=16) +
-  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
+  geom_point(
+      color=hsv(angle_seq, 0.5, 1), 
+      size=2, 
+      shape=16) +
+  geom_point(
+      data=df1, 
+      aes(x, y), 
+      color=hsv(sb/n, 0.5, 1), 
+      size=24, 
+      shape=16) +
+  expand_limits(
+      x=c(-1, 1), 
+      y=c(-1, 1)) +
   theme_void() +
   theme(
-    panel.background=element_rect(fill="#808080"))
+      panel.background =
+          element_rect(fill="#808080"))
 ```
@@
 display.brewer.pal(5, "Set1")
-brewer.pal(5, "Set1")
+pal5 <- brewer.pal(5, "Set1")
 ```
 
-These aren't taken directly from a color wheel but they are well spaced out from a visual perspective, meaning that it is easy to distinguish among these different colors.
+These aren't taken directly from a color wheel but they are well spaced out from a visual perspective, meaning that it is easy to distinguish between any two pairs of these different colors.
+You can find similar choices in Python, SAS and Stata.
+
+SPSS does not have specific color palettes for nominal data, but it does organize its colors so that most of the rows are good choices for representing nominal data and most of the columns are good choices for representing ordinal data.
+
+![](http://www.pmean.com/new-images/20/poor-color-choice-02.png)
+
+Ordinal data is often well represented by a gradient of dark to light levels of a particular color, but this is a poor choice for nominal data.
+With nominal data, you want a good amount of separation between any two colors.
+You can get this by selecting evenly spaced values on a color wheel.

### 2023-10-10 text/23/too-many-colors.rmd

--- 2023-09-19 text/23/too-many.rmd
@@
 They even have different shades of black.
-You have a wealth of colors, but you can't sue all of them in the same graph.
+You have a wealth of colors, but you can't use all of them in the same graph.
 The ideal number of colors is 2.
+Sometimes 3 is okay, but usually 2 is better.
 
@@
 
-Here's another lesson that I learned from Naomi Robbins, another expert on data visualization.
+Here's a lesson that I learned from Naomi Robbins, another expert on data visualization.
 She points out that you don't make every word in a sentence a different color.
@@
 ```{r, echo=FALSE, warning=FALSE}
-us +
-  scale_fill_continuous(
-        low = "green", 
-        high = "blue", 
-        name = "Poverty rate (2014)", 
-        label = scales::comma) + 
-  theme(legend.position = "right")
-```
-
-You can even show a transition between three different colors. This transition from green to yellow to red is used commonly in genomic data visualizations.
-
-```{r, echo=FALSE, warning=FALSE}
 gr <- c("green", "yellow", "red")
@@
 ```
+
+You can even show a transition between three different colors. This transition from green to yellow to red is used commonly in genomic data visualizations.
 
@@
 In the above map, the low poverty rate of New Hampshire (9.2%) is a red that differs only slightly from high poverty rate of Mississippi (21.9%) shown as magenta.
-If anything, the contrast is bigger between Mississippi (magenta) and Alabama (blue) even though the poverty rates of 19.2% for Alabama is a lot closer than New Hampshire's.
+If anything, the contrast is bigger between Mississippi (magenta) and Alabama (blue) even though the poverty rate of 19.2% for Alabama is a lot closer than New Hampshire's.
 

### 2023-10-31 src/copy_files.R
Added drafts of various sample size pages

@@ -34,16 +34,16 @@ copy_files <- function(
     
     new_dir <- "post"
     for (i_file in file_list) {
-      verb <- ifelse(ok_to_copy,"", "to be")
+      verb <- ifelse(ok_to_copy,"copied", "to be copied")
       k <- k+1
       old_file <- str_glue("{old_path}/{old_dir}/{i_file}") 
       new_file <- str_glue("{new_path}/{new_dir}/{i_file}")
       if (ok_to_copy) file.copy(old_file, new_file, overwrite=TRUE)
-      str_glue("\n{old_file} {verb} copied to {new_file}\n") -> message
+      str_glue("\n{old_file}{verb} copied to {new_file}\n") -> message
       if (qc) cat(message, "\n")
     }
   }
-  str_glue("\n\nTotal files {verb} copied: {k}") %>% cat
+  str_glue("\n\nTotal files {verb}: {k}") %>% cat
 }
 ok_to_copy <- TRUE
 ok_to_copy <- FALSE

### 2023-10-31 text/23/mcnemar-test.Rmd
Added mcnemar-test

---
title: "McNemar's test"
source: "New"
author: Steve Simon
date: 2023-10-31
categories:
- Blog post
tags:
- Measuring benefit and risk
output: html_document
---

I mentioned McNemar's test in a [short article][sim3] for my [Monthly Mean newsletter][sim4]. Here are more details about this test.

<!---more--->

Everytbody who knows anything about Statistics knows about the paired t-test. You have measurements done in pairs. The most common setting with paired data is a before and after measurement on the same subject. Other pairs may arise if you take measurements on a patient treated with a new therapy along with measurements on a control patient matched on age, sex, and race. Still other pairs might result from measurements on the left arm of a patient, treated with a new topical agent, and the right arm, treated with a standard topical agent.

When you have data in pairs, you calculate the difference. Then you compute the sample average of these differences as well as the standard deviation of the differences. Then you construct a test statistic, the sample average difference divided by the standard error (the standard deviation divided by the square root of the number of pairs). If this statistic is close to zero, you accept the null hypothesis, which states that the average difference in the population is equal to zero.

The paired t-test requires several assumptions.

The first assumption is that the data from one pair of measurements is independent of the data from another pair of measurements. Be careful with this assumption. You know that withing a pair, the first measurement and the second measurement are correlated. A subject who has a high "before" measurement is likely to have a high "after" measurement. Not always, but often enough to induce a correlation.

In fact, you rely on this. A positive correlation will produce a difference that is less variable than when the two measurements are uncorrelated. Pairing is a great tool for improving the precision in a research study.

The second assumption is that the difference in measurements is normally distributed. Again, be careful with this assumption. If the first measurement is non-normal and the second measurement has a similar type of non-normality, sometimes taking the difference will cancel out the non-normality. So normality of the first measurement or the second measurement is not what you need to check. Only check the normality assumption in the difference.

So what do you do if the difference in measurements is not normal? A slight deviation from normality is not a problem, especially with a large sample size. But suppose the non-normality arises because your "before" and "after" measurements are binary. There's no way that a difference in binary measurements can come close to normal.

Quinn McNemar came to the rescue in 1947 in a publication

Quinn McNemar, "Note on the sampling error of the difference between correlated proportions or percentages". Psychometrika, 1947-06-18, 12(2), 153–157, doi: [10.1007/BF02295996][mcn1], pmid: [20254758][mcn2]. Please note that this article is behind a paywall.

[mcn1]: https://doi.org/10.1007%2FBF02295996
[mcn2]: https://pubmed.ncbi.nlm.nih.gov/20254758

that described a test that now bears his name, the McNemar test.

To compute McNemar's test, you first need to calculate the two by two crosstabulation of the "before" and "after" measurements.

$\begin{matrix} a & b \\ c & d \end{matrix}$

it is arbitrary whether the rows represent the "before" measurements and the columns represent the "after" measurements or the reverse. The order of the rows and columns is also arbitrary, but you must be consistent. If the binary variable is yes/no, then you can have the first row and the first column being "yes" or you can have the first row and column being "no" but you shouldn't have the first row be "yes" and the first column be "no".

Once you have the counts. Add in the row and column totals.

$\begin{matrix} a & b & a+b \\ c & d & c+d \\ a+c & b+d & n \end{matrix}$

where n = a + b + c + d is the total number of pairs.

McNemar's test compares the proportion of observations in the first column to the proportion of observations in the first row.

Here's a simple example from page 267 of a classic book, Alan Agresti. An Introduction to Categorical Data Analysis, Second Edition. The entire book is available for free in [pdf format][agr1].

[agr1]: https://mregresion.files.wordpress.com/2012/08/agresti-introduction-to-categorical-data.pdf

![](http://www.pmean.com/new-images/23/mcnemar-test-01.png).

```{r}
a <- 833; b <- 125; c <- 2; d <- 160
n <- a + b + c + d
r1 <- round(100*(a+b)/n)
c1 <- round(100*(a+c)/n)
T <- round((b-c)^2/(b+c), 1)
```

Do you believe in heaven? Do you believe in hell? Those two questions were asked in a survey of `r n` people. They responded either yes or no. Of the respondents, `r a` believed in both heaven and hell, `r b` believed only in heaven, `r c` believed in hell but not heaven, and `r d` believed in neither. If you add up the counts in the first row, there are `r a+b` respondents who believe in heaven, or `r r1`%. There are `r a+c` respondents who believe in hell, or `r c1`%.

it looks like more people believe in heaven than in hell. But this might be due to sampling error.

Returning to the general case,

$\begin{matrix} a & b & a+b \\ c & d & c+d \\ a+c & b+d & n \end{matrix}$

the proportion in the first row and in the first column are $\frac{a+b}{n}$ and $\frac{a+c}{n}$, respectively.

These fraction differ only if the counts in the two off diagonal cells (b and c) differ. You can write the hypothesis test as

-   $H_0:\ \pi_{before} = \pi_{after}$
-   $H_0:\ \pi_{before} \ne \pi_{after}$

though in the example cited earlier, it make make more sense to write the hypotheses as

-   $H_0:\ \pi_{heaven} = \pi_{hell}$
-   $H_0:\ \pi_{heaven} \ne \pi_{hell}$

Under the null hypothesis, you would expect the values of b and c to be the same, with both equal to (b+c)/2. There may be some sampling error, of course. Using the general chi-square format

-   $T = \Sigma \frac{(O-E)^2}{E}$

would produce

-   $T = \frac{\big(b-\frac{b+c}{2}\big)^2}{\frac{b+c}{2}} + \frac{\big(c-\frac{b+c}{2}\big)^2}{\frac{b+c}{2}}$

With a bit of algebra, you can show that this equals

-   $T = \frac{(b-c)^2}{b+c}$

Compare this to a chi-square distribution with one degree of freedom.

In the survey about heaven and hell, b=`r b` and c=`r c`. The test statistic is

-   $T = \frac{(125-2)^2}{127} = `r T`$

If you want to enter this data into a program like SPSS, you can't type it as you see it.

```{}
833,125
2,160
```

It would be nice if you could do this, but it doesn't work for a variety of reasons. You need to put each count on a separate row. Specify the belief in heaven (yes/no) as the first column and belief in hell (yes/no) as the second column. The count for each combination goes in the third column.

```{}
heaven,hell,count
1,1,833
1,2,125
2,1,2
2,2,160
```

Make sure that you designate count as a weight before starting your analysis. Here is what the output looks like in SPSS

![](http://www.pmean.com/new-images/23/mcnemar-test-02.png)

Here is what it looks like in R

```{r}
mcnemar.test(matrix(c(833, 125, 2, 160), nrow=2), correct=FALSE)
```

Thanks are due to Wikipedia for helping me identify the statistician behind the test and the original publication associated with the test.

[sim3]: http://www.pmean.com/news/201105.html#3

[sim4]: http://www.pmean.com/news/Archive.html

### 2023-12-16 text/23/what-is-a-partial-correlation.Rmd
Added what-is-a-partial-correlation

---
title: "What is a partial correlation?"
source: "New"
author: Steve Simon
date: 2023-12-05
categories:
- Blog post
tags:
- Correlation
- Definitions
output: html_document
---

The concept of controlling for a third variable is an important one to understand in Statistics. When you are looking at data without the benefit of randomization, sometimes an examination of the relationship between two variables can be distorted by a third variable. The third variable might create an false association between those two variables or it might mask a real association.

There are several ways to account for that third variable when examining interfere with your analysis of a relation. A common approach is to use a partial correlation, which effectively adjusts for the third variable or holds the third variable constant when examining the relationship.

<!---more--->

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(broom)
library(tidyverse)
fn <- "https://raw.githubusercontent.com/pmean/classes/master/biostats-1/data/data-05-fev.csv"
read_csv(fn, col_names=FALSE) %>%
  data.frame -> init
names(init) <- c(
  "age",
  "fev",
  "height",
  "sex",
  "smoker")
```

## Partial correlation

-   $r_{XY\cdot Z}=\frac{r_{XY}-r_{XZ}\times r_{YZ}} {\sqrt{1-r_{XZ}^2}\sqrt{1-r_{YZ}^2}}$

Here is the formula for partial correlation. Read $r_{XY\cdot Z}$ as the correlation between X and Y holding Z constant, or the correlation between X and Y adjusted for Z.

Look at the numerator. It takes the unadjusted correlation of X and Y and subtracts a product of the correlation of each of them with Z. This tells you something important right away. The adjustment is small if the product is small, meaning either that X is not strongly associated with the third variable, Z or Y is  not strongly associated with Z. The adjustment is large when both X and Y have a strong positive or negative association with Z.

The denominator will actually modify things even further. If $r_{XZ}$ is large negative or large positive, then the term $\sqrt{1-r_{XZ}^2$ will be small. A small value in the denominator makes the whole fraction big. A similar pattern occurs when $r_{XZ}$ is large negative or large positive.

If the third variable correlates poorly with both X and Y, then the adjusted correlation is not too much different from the unadjusted correlation.

## An example of partial correlation

```{r, echo=FALSE}
rxy <- round(cor(init$fev, init$age), 2)
rxz <- round(cor(init$fev, init$height), 2)
ryz <- round(cor(init$age, init$height), 2)
```

Let's look at an example. A pediatric study of lung function measured forced expiratory volume (FEV) in a sample of children ages 3 through 17. FEV is a measure of how much air you can blow out of your lungs. It is expected to increase with age.

-   The correlation of FEV with age is `r rxy`.

But there is a third factor, height, which might account for this relationship.

-   The correlation of FEV with height is `r rxz`
-   The correlation of age with height is `r ryz`

## Unadjusted relationship between age and FEV

If you look at all of the data, you see a strong correlation between age and FEV. This graph shows the individual heights for each child at the X, Y value corresponding to the age and FEV. Notice that the younger children have heights in the high 40s and the low 50s and the older children have heights in the high 60s and low 70s.

```{r, echo=FALSE}
ggplot(init, aes(age, fev, label=height)) +
  geom_text() +
  ggtitle(paste0("Correlation = ", rxy))
h <- 48
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

When you look at the shortest kids, those with heights between `r h-2` and `r h+1.5`, you see that the correlation is weak.

```{r, echo=FALSE}
g <- 4*floor((init$height-2)/4)+4
control_height <- function(a) {
  rxy <- round(cor(init$age[g==h], init$fev[g==h]), 2)
  co=ifelse(g==h, "black", "gray")
  al=ifelse(g==h, 1, 0)
  init %>%
    mutate(co=ifelse(g==h, "black", "gray")) %>%
    ggplot(aes(age, fev, label=height)) +
      geom_text(color=co) +
      geom_text(color=co, alpha=al) +
      ggtitle(paste0("Correlation = ", rxy))
}
h <- 48
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

You see a similar weak correlation when the height is between `r h-2` and `r h+1.5`.

And so forth.

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Relationship between age and FEV controlling for height between `r h-2` and `r h+1.5`

```{r, echo=FALSE}
control_height(h)
h <- h+4
```

## Partial correlation calculation

The formal calculation of the partial correlation is shown here.

-   $\frac{`r rxy`-`r rxz` \times `r ryz`}{\sqrt{1-`r rxz`^2}\sqrt{1-`r ryz`^2}} = `r round((rxy-rxz*ryz)/(sqrt(1-rxz^2)*sqrt(1-ryz^2)), 2)`$

The correlation between age and FEV is very weak when you adjust for the child's height.

### 2024-01-03 src/year-end-summary.Rmd
Added distribution-fee-bayesian-analysis

@@
 # Put it all together
-yyyy_list <- c(1999:2023)
+yyyy_list <- c(1999:2024)
 n_files <- 0

### 2024-03-04 src/year-end-summary.Rmd
Added finding-examples to 23

@@
 # Put it all together
-yyyy_list <- c(1999:2024)
+yyyy_list <- c(2023:2023)
 n_files <- 0

### 2024-03-04 src/year-end-summary.Rmd

@@
 ---
+
+```{r}
+yyyy_list <- c(1999:2024)
+```
 
@@
 # Put it all together
-yyyy_list <- c(2023:2023)
 n_files <- 0

### 2024-05-07 text/pending/titanic-spline.Rmd
Added titanic-spline

---
title: Illustrating splines using the Titanic dataset
source: "New"
author: Steve Simon
date: 2024-05-07
categories:
- Blog post
tags:
- Nonlinear regression
- R programming
output: html_document
---

Splines provide a useful way to model relationships that are more complex than a simple linear relationship. They work in a variety of regression models. Here is an illustration of how to use a spline in a logistic regression model with data from survival of passengers on the Titanic.

<!---more--->

```{r, echo=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "titanic-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

The Titanic was a large cruise ship, the biggest of its kind in 1912. It was thought to be unsinkable, but when it set sail from England to American in its maiden voyage, it struck an iceberg and sank, killing many of the passengers and crew. You can get fairly good data on the characteristics of passengers who died and compare them to those that survived. The data indicate a strong effect due to age and gender, representing a philosophy of "women and children first" that held during the boarding of life boats. Let's look at the effect of age on survival using a logistic regression model.

Here is a brief description of the diamond pricing dataset, taken from the [data dictionary][sim3] on my github site.

`r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/titanic-data-dictionary.yaml

```{r}
f2 <- "titanic_v00.txt"
raw_data <- read.table(
  file=paste0(f0, f2), 
  sep="\t", 
  header=TRUE)
head(raw_data)
```

### 2024-05-07 src/copy_files.R

@@ -15,7 +15,7 @@ source(file="src/standard_functions.R")
 copy_files <- function(
   old_path="text", 
   new_path="c:/Users/simons/Documents/b04/content",
-  old_dir_list=c("99", zpad(0:23)),
+  old_dir_list=c("99", zpad(0:24)),
   file_pattern="md$",
   qc=FALSE) {
 

### 2024-05-09 text/23/what-is-a-partial-correlation.Rmd
Added tricky-program

@@
 fn <- "https://raw.githubusercontent.com/pmean/classes/master/biostats-1/data/data-05-fev.csv"
-read_csv(fn, col_names=FALSE) %>%
+read_csv(
+    fn, 
+    col_names=FALSE, 
+    col_types="n") %>%
     data.frame -> init

### 2024-05-09 text/pending/diamond-spline.Rmd

---
title: Illustrating splines using the diamond pricing dataset
source: "New"
author: Steve Simon
date: 2024-05-07
categories:
- Blog post
tags:
- Nonlinear regression
- R programming
output: html_document
---

Splines provide a useful way to model relationships that are more complex than a simple linear relationship. They work in a variety of regression models. Here is an illustration of how to use a spline in a dataset looking at pricing for diamonds.

<!---more--->

```{r, echo=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "diamonds-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

Here is a brief description of the diamond pricing dataset, taken from the [data dictionary][sim3] on my github site.

`r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/diamonds-data-dictionary.yaml

You can also find the [original source of the data][jse1] and a [more detailed description][jse2], both found on the [Journal of Statistics Education datasets website][jse3].

[jse1]: http://jse.amstat.org/v9n2/4Cdata.txt
[jse2]: http://jse.amstat.org/v9n2/4C.txt
[jse3]: https://jse.amstat.org/jse_data_archive.htm

```{r}
f2 <- "diamonds.txt"
raw_data <- read.fwf(file=paste0(f0, f2), widths=c(5,2,5,4,5))
di <- raw_data
names(di) <- c(
  "carat",
  "colour",
  "clarity",
  "certification",
  "price"
)
```

Make adjustments for exchange rate and inflation.

```{r}
exchange_rate <- 1.72
inflation_rate <- 1.81
di$adjusted_price <- di$price*inflation_rate/exchange_rate
head(di)
```

Here are a few descriptive statistics

```{r}
summary(di$carat)
summary(di$price)
summary(di$adjusted_price)
count(di, colour)
count(di, clarity)
count(di, certification)
```

Suppose that you want to build a model that predicts the adjusted price using carat, colour, and clarity. You suspect, however, that the relationship between carat and adjusted price is nonlinear.

```{r}
ggplot(data=di, aes(carat, adjusted_price)) +
  geom_point(pch=1) -> scatter1
scatter1
```

First, examine a linear trend.

```{r}
scatter1+geom_smooth(method="lm")
```

It looks like you could do a bit better than a simple linear fit, and a spline is a good option.

There are many options for splines in R, and you might have additional options beyond splines that also work well.

Choosing among the splines is tricky. To be honest, most of the choices will work out well and the choices end up being between various gradations of good to great.

A type of spline that I like a lot are the restricted cubic splines developed by Frank Harrell in the rms package. You implement these splines by adding an rcs function to your model formula. You use the number of knots to control the complexity from splines that represent a gentle deviation from normality to splines that wiggle all around and are far from linear. The default option for knots in rcs usually works fairly well.

There is a problem with direct interpretation, though. The coefficients for a linear fit are fairly easy to interpret.

```{r}
linear_fit <- lm(adjusted_price ~ carat, data=di)
linear_coefficients <- tidy(linear_fit)
linear_coefficients
```

The slope, `r round(linear_coefficients[2, 2])`,  represents the estimated average change in price when carats increase by one unit.

Things don't have such a nice interpretation with your spline model.

```{r}
spline_fit <- lm(adjusted_price ~ rcs(carat), data=di)
tidy(spline_fit)
```

The coefficients from the restricted cubic spline are pretty much uninterpretable. You have to visualize the spline graphically.

```{r}
carat_sequence <- data.frame(
  carat = seq(min(di$carat), max(di$carat), length=100))
carat_sequence$prediction <- 
  predict(spline_fit, newdata=carat_sequence)
scatter1 + geom_line(
  data=carat_sequence,
  aes(carat, prediction))
```

You might be interested in seeing if the spline fits better than a simple linear function. Fit both a linear term and a spline and then compare to the fit with just a linear term.

```{r}
linear_plus_spline_fit <- lm(
  adjusted_price ~ carat + rcs(carat), 
  data=di)
anova(linear_fit, linear_plus_spline_fit)
```

The p-value is small, indicating that adding a spline to a linear fit significantly improves the prediction model.

### 2024-05-09 text/24/survival-spline.Rmd
Added survival-spline

---
title: Illustrating splines using the Worcester Heart Attack Study
source: "New"
author: Steve Simon
date: 2024-05-07
categories:
- Blog post
tags:
- Survival analysis
- R programming
output: html_document
---

Splines provide a useful way to model relationships that are more complex than a simple linear relationship. They work in a variety of regression models. Here is an illustration of how to use a spline in a Cox regression model with data from the Worcester Heart Attack Study.

<!---more--->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "whas100-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

Here is a brief description of the whas100 dataset, taken from the [data dictionary][sim3] on my github site.

`r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/whas100-data-dictionary.yaml

Here are the first few rows of data and the last few rows of data. Row 101 needs to be removed.

```{r}
f2 <- "whas100.dat"
raw_data <- read.fwf(
  file=paste0(f0, f2),
  widths=c(7, 12, 12, 5, 9, 5, 7, 4, 8))
wa <- raw_data
names(wa) <- c(
  "id",
  "admitdate",
  "foldate",
  "los",
  "lenfol",
  "fstat",
  "age",
  "gender",
  "bmi")
head(wa)
tail(wa)
wa <- wa[-101, ]
wa$lenfol <- wa$lenfol/365.25
```

Here are a few descriptive statistics

```{r}
summary(wa$id)
summary(wa$los)
summary(wa$lenfol)
count(wa, fstat)
summary(wa$age)
count(wa, gender)
summary(wa$bmi)
```

Here is an overall survival curve.

```{r}
plot(Surv(wa$lenfol, wa$fstat))
```

```{r}
linear_fit <- coxph(
  Surv(wa$lenfol, wa$fstat)~age,
  data=wa)
linear_fit
```

```{r}
spline_fit <- coxph(
  Surv(wa$lenfol, wa$fstat)~rcs(age),
  data=wa)
spline_fit
```

```{r}
age_sequence <- data.frame(
  age = seq(min(wa$age), max(wa$age), length=100))
age_sequence$prediction <- 
  predict(spline_fit, newdata=age_sequence, type="lp")
ggplot(data=age_sequence, aes(age, prediction)) +
  geom_line()
```

### 2024-05-09 text/24/titanic-spline.Rmd
Added survival-spline

---
title: Illustrating splines using the Titanic dataset
source: "New"
author: Steve Simon
date: 2024-05-07
categories:
- Blog post
tags:
- Nonlinear regression
- R programming
output: html_document
---

Splines provide a useful way to model relationships that are more complex than a simple linear relationship. They work in a variety of regression models. Here is an illustration of how to use a spline in a logistic regression model with data from survival of passengers on the Titanic.

<!---more--->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "titanic-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

The Titanic was a large cruise ship, the biggest of its kind in 1912. It was thought to be unsinkable, but when it set sail from England to American in its maiden voyage, it struck an iceberg and sank, killing many of the passengers and crew. You can get fairly good data on the characteristics of passengers who died and compare them to those that survived. The data indicate a strong effect due to age and gender, representing a philosophy of "women and children first" that held during the boarding of life boats. Let's look at the effect of age on survival using a logistic regression model.

Here is a brief description of the diamond pricing dataset, taken from the [data dictionary][sim3] on my github site.

`r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/titanic-data-dictionary.yaml

Here are the first few rows of data.

```{r}
f2 <- "titanic_v00.txt"
ti0 <- 
  data.frame(
    read_tsv(
      file=paste0(f0, f2),
      col_types="ccncn"))
names(ti0) <- tolower(names(ti0))
head(ti0)
```

Here are a few descriptive statistics

```{r}
summary(ti0$Age)
count(ti0, pclass)
count(ti0, sex)
count(ti0, survived)
```

The boxplots reveal little differences between the ages of survivors and deaths. If something is going on, it is subtle.

```{r}
ggplot(data=ti0, aes(factor(survived), age)) + 
  geom_boxplot()
```

Fit a linear model first.

```{r}
ti1 <- filter(ti0, !is.na(age))
linear_fit <- glm(
  survived ~ age,
  family=binomial,
  data=ti1)
tidy(linear_fit)
```

There may be a downward trend in the odds of survival over time, but it is not statistically significant.

Plot the predicted values.

```{r}
age_sequence <- data.frame(
  age = seq(min(ti1$age), max(ti1$age), length=100))
age_sequence$prediction <- 
  predict(linear_fit, newdata=age_sequence, type="response")
ggplot(data=ti1, aes(age, survived)) +
  geom_point(pch=1) +
  geom_line(
    data=age_sequence,
    aes(age, prediction))
```

The plot is linear, but you should really look at it on the log odds scale because the logistic regression model is linear in the log odds. You need to change the predict function from type="response" to type="link" to get predictions on a log odds scale.

```{r}
age_sequence$prediction <- 
  predict(linear_fit, newdata=age_sequence, type="link")
ggplot(data=age_sequence, aes(age, prediction)) +
  geom_line()
```

Now fit a spline function.

```{r}
spline_fit <- glm(
  survived ~ rcs(age),
  family=binomial,
  data=ti1)
tidy(spline_fit)
```

The coefficients from the restricted cubic spline are pretty much uninterpretable. You have to visualize the spline graphically. First do this on the log odds scale to see how far from linear the spline fit is.

```{r}
age_sequence$prediction <- 
  predict(spline_fit, newdata=age_sequence, type="link")
ggplot(data=age_sequence, aes(age, prediction)) +
  geom_line()
```

Now plot on the original scale to get a picture of what is going on.

```{r}
age_sequence$prediction <- 
  predict(spline_fit, newdata=age_sequence, type="response")
ggplot(data=ti1, aes(age, survived)) +
  geom_point(pch=1) +
  geom_line(
    data=age_sequence,
    aes(age, prediction))
```

It looks like "women and children" first might actually be "women, children, and old people first".

### 2024-05-09 text/pending/titanic-imputation.Rmd
Added pending folder

--- 2024-05-09 text/24/titanic-spline.Rmd
@@
 ---
-title: Illustrating splines using the Titanic dataset
+title: Illustrating imputation of missing data using the Titanic dataset
 source: "New"
@@
 tags:
-- Nonlinear regression
+- Missing data
 - R programming
@@
 
-Here is a brief description of the diamond pricing dataset, taken from the [data dictionary][sim3] on my github site.
+Here is a brief description of this dataset, taken from the [data dictionary][sim3] on my github site.
 
@@
 ```{r}
-summary(ti0$Age)
+summary(ti0$age)
 count(ti0, pclass)
@@
 ```
+
+Notice the large number of missing values for age.
 

### 2024-05-09 text/pending/survival-lasso.Rmd
Started work on survival-lasso

---
title: Illustrating the lasso model using the Worcester Heart Attack Study
source: "New"
author: Steve Simon
date: 2024-05-07
categories:
- Blog post
tags:
- Survival analysis
- R programming
output: html_document
---

The lasso model allows you to effectively identify important features in a large regression model. Here is an illustration of how to use a the lasso model with survival data from the Worcester Heart Attack Study.

<!---more--->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(broom)
library(glmnet)
library(pspline)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
f1 <- "whas500-data-dictionary.yaml"
dd <- read_yaml(paste0(f0, f1))
```

Here is a brief description of the whas100 dataset, taken from the [data dictionary][sim3] on my github site.

`r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/whas100-data-dictionary.yaml

Here are the first few rows of data and the last few rows of data. Row 101 needs to be removed.

```{r}
vnames <- names(dd$vars)
f2 <- "whas500.dat"
w0 <- read_table(
  file=paste0(f0, f2), 
  col_names=c(vnames),
  col_types="nnnnnnnnnnnnnnncccnnnn")
head(w0)
```

Here are a few descriptive statistics

```{r}
dd$vars[["id"]]
summary(w0$id)

dd$vars[["age"]]
summary(w0$age)

dd$vars[["gender"]]
count(w0, gender)

dd$vars[["sysbp"]]
summary(w0$sysbp)

dd$vars[["diasbp"]]
summary(w0$diasbp)

dd$vars[["bmi"]]
summary(w0$bmi)

dd$vars[["cvd"]]
count(w0, cvd)

dd$vars[["afb"]]
count(w0, afb)

dd$vars[["sho"]]
count(w0, sho)

dd$vars[["chf"]]
count(w0, chf)

dd$vars[["av3"]]
count(w0, av3)

dd$vars[["miord"]]
count(w0, miord)

dd$vars[["mitype"]]
count(w0, mitype)

dd$vars[["year"]]
count(w0, year)

dd$vars[["los"]]
summary(w0$los)

dd$vars[["dstat"]]
count(w0, dstat)

dd$vars[["lenfol"]]
summary(w0$lenfol)

dd$vars[["fstat"]]
count(w0, fstat)
```

Here is an overall survival curve.

```{r}
plot(Surv(w0$lenfol, w0$fstat))
```

You can find the lasso regression model in the glmnet package. One important difference with glmnet is that it does not use a formula to define your model. Instead you provide matrices for your dependent variable and your independent variables.

### 2024-05-09 text/pending/titanic-imputation.Rmd
Started work on survival-lasso

@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-library(broom)
-library(pspline)
+library(mice)
 library(readr)
@@
 
-Notice the large number of missing values for age.
-
-The boxplots reveal little differences between the ages of survivors and deaths. If something is going on, it is subtle.
+Notice the large number of missing values for age. The first three passegers with missing ages are
 
 ```{r}
-ggplot(data=ti0, aes(factor(survived), age)) + 
-  geom_boxplot()
+missing_rows <- 13:15
+ti0 %>% slice(missing_rows)
 ```
 
-Fit a linear model first.
+Create mutliply imputed values for age.
+The default is to use every variable in the dataset other than age to impute the value of age.
+You don't want to use the name of the passenger, of course, so be sure to drop it before the imputation step.
 
 ```{r}
-ti1 <- filter(ti0, !is.na(age))
-linear_fit <- glm(
-  survived ~ age,
-  family=binomial,
-  data=ti1)
-tidy(linear_fit)
+ti1 <- select(ti0, -name)
+ti_mice <- mice(ti1)
 ```
 
-There may be a downward trend in the odds of survival over time, but it is not statistically significant.
-
-Plot the predicted values.
+The mice function creates a complex object.
+Go ahead and explore the various components, but be forewarned that this is messy.
+You can extract simple parts of the imputation with various functions.
+The complete function shows the complete datasets with the imputed values.
+Here are the first three rows where age was imputed the first time.
 
 ```{r}
-age_sequence <- data.frame(
-  age = seq(min(ti1$age), max(ti1$age), length=100))
-age_sequence$prediction <- 
-  predict(linear_fit, newdata=age_sequence, type="response")
-ggplot(data=ti1, aes(age, survived)) +
-  geom_point(pch=1) +
-  geom_line(
-    data=age_sequence,
-    aes(age, prediction))
+ti_mice %>%
+  complete(1) %>%
+  slice(missing_rows)
 ```
 
-The plot is linear, but you should really look at it on the log odds scale because the logistic regression model is linear in the log odds.
-You need to change the predict function from type="response" to type="link" to get predictions on a log odds scale.
+Look at the imputations for the second, third, etc. times. Notice that the values change with each imputation.
 
 ```{r}
-age_sequence$prediction <- 
-  predict(linear_fit, newdata=age_sequence, type="link")
-ggplot(data=age_sequence, aes(age, prediction)) +
-  geom_line()
+ti_mice %>%
+  complete(2) %>%
+  slice(missing_rows)
+ti_mice %>%
+  complete(3) %>%
+  slice(missing_rows)
+ti_mice %>%
+  complete(4) %>%
+  slice(missing_rows)
+ti_mice %>%
+  complete(5) %>%
+  slice(missing_rows)
+
 ```
 
-Now fit a spline function.
+Use the with function of the mice package to fit a model to apply a particular analysis to each of the five imputed datasets.
 
 ```{r}
-spline_fit <- glm(
-  survived ~ rcs(age),
-  family=binomial,
-  data=ti1)
-tidy(spline_fit)
+ti_with <- with(
+  ti_mice, 
+  glm(
+    survived~pclass+rcs(age)+sex, 
+    family=binomial)) 
 ```
 
-The coefficients from the restricted cubic spline are pretty much uninterpretable.
-You have to visualize the spline graphically.
-First do this on the log odds scale to see how far from linear the spline fit is.
+Again, the object created here is complex. You can extract the individual analyses fairly easily. Here is the analysis of the first imputed dataset.
 
 ```{r}
-age_sequence$prediction <- 
-  predict(spline_fit, newdata=age_sequence, type="link")
-ggplot(data=age_sequence, aes(age, prediction)) +
-  geom_line()
+ti_with$analyses[[1]]
 ```
 
-Now plot on the original scale to get a picture of what is going on.
+The key variable here is sexmale, which shows how much different the log odds of survival are between men and women.
+
+Notice that the estimate for sexmale changes from one analysis to another, though not by much.
 
 ```{r}
-age_sequence$prediction <- 
-  predict(spline_fit, newdata=age_sequence, type="response")
-ggplot(data=ti1, aes(age, survived)) +
-  geom_point(pch=1) +
-  geom_line(
-    data=age_sequence,
-    aes(age, prediction))
+ti_with$analyses[[2]]
+ti_with$analyses[[3]]
+ti_with$analyses[[4]]
+ti_with$analyses[[5]]
 ```
 
-It looks like "women and children" first might actually be "women, children, and old people first".
+Combine all these analyses with the pool function.
+
+```{r}
+ti_pool <- pool(ti_with)
+summary(ti_pool)
+log_or <- summary(ti_pool)[8, 2]
+```
+
+To complete things, compute the odds ratio and confidence interval.
+
+```{r}
+se <- summary(ti_pool)[8, 3]
+or <- exp(log_or)
+lo <- exp(log_or-1.96*se)
+hi <- exp(log_or+1.96*se)
+paste0(round(or, 3), " (", round(lo, 3), ",", round(hi, 3), ")")
+```

### 2024-05-11 text/pending/survival-lasso.Rmd
First full draft of survival-lasso

@@
 
+I've followed the [glmnet vignette for Cox regression][tay1] closely. You might find that explanation to be clearer than what I present below.
+
+[tay1]: https://glmnet.stanford.edu/articles/Coxnet.html
+
 <!---more--->
@@
 library(glmnet)
+library(glue)
 library(pspline)
@@
 library(yaml)
-f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "whas500-data-dictionary.yaml"
-dd <- read_yaml(paste0(f0, f1))
+g0 <- "https://raw.githubusercontent.com/pmean/"
+r0 <- "datasets/master/"
+s1 <- "yaml-functions.R"
+f1 <- "whas500.yaml"
+
+source(paste0(g0, r0, s1))
+dd <- read_yaml(paste0(g0, r0, f1))
+vnames <- extract_names(dd)
+vlabels <- extract_labels(dd)
+vscales <- extract_scales(dd)
 ```
 
-Here is a brief description of the whas100 dataset, taken from the [data dictionary][sim3] on my github site.
+Here is a brief description of the whas500 dataset, taken from the [data dictionary][sim3] on my github site.
 
@@
 w0 <- read_table(
-  file=paste0(f0, f2), 
+  file=paste0(g0, r0, f2), 
   col_names=c(vnames),
@@
 ```{r}
-dd$vars[["id"]]
-summary(w0$id)
+categorical <- 
+  (vscales=="binary") |
+  (vscales=="nominal") |
+  (vscales=="ordinal")
+for (i_vars in which(categorical)) {
+  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
+  w0 %>%
+    count(.data[[vnames[i_vars]]]) %>%
+    data.frame %>%
+    print
+  cat("\n")
+}
+```
 
-dd$vars[["age"]]
-summary(w0$age)
-
-dd$vars[["gender"]]
-count(w0, gender)
-
-dd$vars[["sysbp"]]
-summary(w0$sysbp)
-
-dd$vars[["diasbp"]]
-summary(w0$diasbp)
-
-dd$vars[["bmi"]]
-summary(w0$bmi)
-
-dd$vars[["cvd"]]
-count(w0, cvd)
-
-dd$vars[["afb"]]
-count(w0, afb)
-
-dd$vars[["sho"]]
-count(w0, sho)
-
-dd$vars[["chf"]]
-count(w0, chf)
-
-dd$vars[["av3"]]
-count(w0, av3)
-
-dd$vars[["miord"]]
-count(w0, miord)
-
-dd$vars[["mitype"]]
-count(w0, mitype)
-
-dd$vars[["year"]]
-count(w0, year)
-
-dd$vars[["los"]]
-summary(w0$los)
-
-dd$vars[["dstat"]]
-count(w0, dstat)
-
-dd$vars[["lenfol"]]
-summary(w0$lenfol)
-
-dd$vars[["fstat"]]
-count(w0, fstat)
+```{r}
+continuous <- 
+  (vscales=="interval") |
+  (vscales=="ratio")
+for (i_vars in which(continuous)) {
+  cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
+  print(summary(w0[vnames[i_vars]]))
+  cat("\n")
+}
 ```
@@
 Instead you provide matrices for your dependent variable and your independent variables.
+In a Cox regression model, the dependent matrix has a column for time and a second column with an indicator variable (1 means the event occurred and 0 means censoring).
+
+```{r}
+ymat <- cbind(w0$lenfol, w0$fstat)
+dimnames(ymat)[2] <- list(c("time", "status"))
+xmat <- as.matrix(w0[ , 8:14])
+dimnames(xmat)[2] <- list(vnames[8:14])
+lasso_fit <- glmnet(xmat, ymat, family="cox")
+plot(lasso_fit, label=TRUE)
+plot(lasso_fit, xvar="norm", label=TRUE)
+plot(lasso_fit, xvar="lambda", label=TRUE)
+plot(lasso_fit, xvar="dev", label=TRUE)
+```
+
+```{r}
+cv_fit <- cv.glmnet(xmat, ymat, family = "cox", type.measure="mae")
+plot(cv_fit)
+plot(lasso_fit, xvar="lambda")
+abline(v=log(cv_fit$lambda.1se))
+cv_fit$lambda.1se
+coef(lasso_fit, cv_fit$lambda.1se)
+```
+
+```{r}
+cv_fit$lambda.min
+plot(lasso_fit, xvar="lambda")
+abline(v=log(cv_fit$lambda.min))
+coef(lasso_fit, cv_fit$lambda.min)
+```

### 2024-05-11 text/pending/titanic-imputation.Rmd
First full draft of survival-lasso

@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
+library(glue)
 library(mice)
@@
 library(yaml)
-f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "titanic-data-dictionary.yaml"
-dd <- read_yaml(paste0(f0, f1))
+g0 <- "https://raw.githubusercontent.com/pmean/"
+r0 <- "datasets/master/"
+f1 <- "titanic.yaml"
+s1 <- "yaml-functions.R"
+source(paste0(g0, r0, s1))
+dd <- read_yaml(paste0(g0, r0, f1))
+vnames <- extract_names(dd)
+vlabels <- extract_labels(dd)
 ```
-
-The Titanic was a large cruise ship, the biggest of its kind in 1912.
-It was thought to be unsinkable, but when it set sail from England to American in its maiden voyage, it struck an iceberg and sank, killing many of the passengers and crew.
-You can get fairly good data on the characteristics of passengers who died and compare them to those that survived.
-The data indicate a strong effect due to age and gender, representing a philosophy of "women and children first" that held during the boarding of life boats.
-Let's look at the effect of age on survival using a logistic regression model.
 
@@
 
-```{r}
+```{r, echo=FALSE}
 f2 <- "titanic_v00.txt"
-ti0 <- 
+raw_data <- 
   data.frame(
     read_tsv(
-      file=paste0(f0, f2),
+      file=paste0(g0, r0, f2),
       col_types="ccncn"))
-names(ti0) <- tolower(names(ti0))
+# Re-arranging columns just for convenient display
+o <-  c(1, 3, 2, 4, 5)
+ti0 <- select(raw_data, o)
+vnames <- vnames[o]
 head(ti0)
@@
 
-```{r}
-summary(ti0$age)
-count(ti0, pclass)
-count(ti0, sex)
-count(ti0, survived)
+```{r, echo=FALSE}
+cat(vnames[1], ": ", vlabels[1], "\n")
+summary(ti0$Age)
+cat(vnames[2], ": ", vlabels[2], "\n")
+count(ti0, PClass)
+cat(vnames[3], "\n")
+count(ti0, Sex)
+cat(vnames[4], "\n")
+cat(as.yaml(dd$vars$Survived$values), "\n")
+count(ti0, Survived)
 ```
 
-Notice the large number of missing values for age. The first three passegers with missing ages are
+I have hidden the R code up to this point, as it is mundane and not of great interest. I will show the R code and output for the rest of the analysis.
+
+Notice the large number of missing values for age. The first three passengers with missing ages are
 
 ```{r}
-missing_rows <- 13:15
+missing_rows <- which(is.na(ti0$Age))[1:3]
 ti0 %>% slice(missing_rows)
@@
 ```{r}
-ti1 <- select(ti0, -name)
+ti1 <- select(ti0, -Name)
+ti1$i_female <- as.numeric(ti1$Sex=="female")
+ti1 <- select(ti1, -Sex)
 ti_mice <- mice(ti1)
@@
   glm(
-    survived~pclass+rcs(age)+sex, 
+    Survived~PClass+rcs(Age, 3)+i_female, 
     family=binomial)) 
@@
 ti_pool <- pool(ti_with)
-summary(ti_pool)
-log_or <- summary(ti_pool)[8, 2]
+ti_summary <- summary(ti_pool)
+ti_summary
 ```
@@
 ```{r}
-se <- summary(ti_pool)[8, 3]
-or <- exp(log_or)
-lo <- exp(log_or-1.96*se)
-hi <- exp(log_or+1.96*se)
-paste0(round(or, 3), " (", round(lo, 3), ",", round(hi, 3), ")")
+female <- ti_summary$term=="i_female"
+log_or <- ti_summary[female, "estimate"]
+se <- ti_summary[female, "std.error"]
+or <- round(exp(log_or), 1)
+lo <- round(exp(log_or-1.96*se), 1)
+hi <- round(exp(log_or+1.96*se), 1)
+glue("Odds ratio for females is {or}, 95% CI ({lo},{hi})")
 ```

### 2024-05-11 text/pending/survival-lasso.Rmd
More work on survival-lasso

@@
 library(rms)
+library(survival)
 library(tidyverse)
@@
 
+Here is the traditional Cox model
+
+```{r}
+simple_fit <- coxph(
+  Surv(w0$lenfol, w0$fstat) ~
+      cvd +
+      afb + 
+      sho + 
+      chf + 
+      av3 +
+      miord +
+      mitype,
+    data=w0)
+coef(simple_fit)
+```
+
 You can find the lasso regression model in the glmnet package.
@@
 dimnames(xmat)[2] <- list(vnames[8:14])
-lasso_fit <- glmnet(xmat, ymat, family="cox")
-plot(lasso_fit, label=TRUE)
-plot(lasso_fit, xvar="norm", label=TRUE)
+lasso_fit <- glmnet(xmat, ymat, family="cox", standardize=FALSE)
 plot(lasso_fit, xvar="lambda", label=TRUE)
-plot(lasso_fit, xvar="dev", label=TRUE)
 ```
 
+Let;s spend a bit of time looking at this graph. The left hand side of the graph represents a very small penalty.
+
 ```{r}
-cv_fit <- cv.glmnet(xmat, ymat, family = "cox", type.measure="mae")
+fit_left <- as.matrix(coef(lasso_fit, min(lasso_fit$lambda)))
+fit_left
+plot(lasso_fit, xvar="lambda", label=TRUE)
+text(
+  log(min(lasso_fit$lambda)), 
+  fit_left, 
+  unlist(dimnames(fit_left)[1]))
+```
+
+Notice how the coefficients are very close to the Cox regression model. There is very little shrinkage and only one variable, av3, are zeroed out, meaning that it is totally removed from the model.
+
+```{r}
+fit_median <- as.matrix(coef(lasso_fit, median(lasso_fit$lambda)))
+fit_median
+plot(lasso_fit, xvar="lambda", label=TRUE)
+text(
+  log(median(lasso_fit$lambda)), 
+  fit_median, 
+  unlist(dimnames(fit_median)[1]))
+```
+
+At the middle (median) value of lambda, several of the variables are zeroed out. Most other variables are shrunk towards zero, relative to the fit with the smallest value of lambda.
+
+What happens at the largest value of lambda.
+
+```{r}
+fit_right <- as.matrix(coef(lasso_fit, max(lasso_fit$lambda)))
+fit_right
+plot(lasso_fit, xvar="lambda", label=TRUE)
+text(
+  log(max(lasso_fit$lambda)), 
+  fit_right, 
+  unlist(dimnames(fit_right)[1]))
+```
+
+Once lambda gets large enough, the penalty is so great that the algorithm zeroes out every single variable.
+
+Now you need to decide what lambda works best. You can do this through cross-validation.
+
+```{r}
+cv_fit <- cv.glmnet(xmat, ymat, family = "cox", standardize=FALSE)
 plot(cv_fit)
-plot(lasso_fit, xvar="lambda")
-abline(v=log(cv_fit$lambda.1se))
-cv_fit$lambda.1se
-coef(lasso_fit, cv_fit$lambda.1se)
 ```
+
+There are two choices that are commonly selected.
+
+You could set the value of lambda to the value that minimizes the criteria.
+The default criteria is the partial likelihood deviance, but you can use other criteria to pick out the "best" value of lambda.
 
@@
 abline(v=log(cv_fit$lambda.min))
-coef(lasso_fit, cv_fit$lambda.min)
+fit_min <- as.matrix(coef(lasso_fit, cv_fit$lambda.min))
+fit_min
+text(log(cv_fit$lambda.min), fit_min, unlist(dimnames(fit_min)[1]))
 ```
+
+The developers of glmnet felt that the value of lambda that minimized a fit criteria was not aggressive enough in zeroing out and shrinking coefficients.
+They suggested that instead of choosing the "best" value of lambda, choose the largest value of lambda that is still within one standard error of the criteria produced by the "best" lambda.
+
+```{r}
+plot(lasso_fit, xvar="lambda", label=TRUE)
+abline(v=log(cv_fit$lambda.1se))
+cv_fit$lambda.1se
+fit_1se <- as.matrix(coef(lasso_fit, cv_fit$lambda.1se))
+fit_1se
+text(log(cv_fit$lambda.1se), fit_1se, unlist(dimnames(fit_1se)[1]))
+```
+
+Using this criteria, there are only two variables which help in predicting risk of death, and one of them is very weak because its coefficient is very close to zero.

### 2024-05-27 src/clean_files.R
Updated clean_files

@@ -73,8 +73,6 @@ verbose <- TRUE
 verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
-clean_files(
-  fixed ('This page is being updated from a version on the original website.'),
-  fixed('This page is currently being updated from the earlier version of my website. Sorry that it is not yet fully available.'),
+clean_files("page_update: complete",
   dir_root="text", 
   file_pattern="md$")

### 2024-05-27 text/24/diamond-spline.Rmd
Moved r-datasets from pending to 24

--- 2024-05-09 text/pending/diamond-spline.Rmd
@@
 
-```{r, echo=FALSE}
+```{r initialization, echo=FALSE}
 library(broom)
@@
 f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "diamonds-data-dictionary.yaml"
+f1 <- "diamonds.yaml"
 dd <- read_yaml(paste0(f0, f1))
@@
 
-```{r}
+```{r read-diamonds}
 f2 <- "diamonds.txt"
@@
 
-```{r}
+```{r adjustments}
 exchange_rate <- 1.72
@@
 
-```{r}
+```{r descriptives}
 summary(di$carat)
@@
 
-```{r}
+```{r plot}
 ggplot(data=di, aes(carat, adjusted_price)) +
@@
 
-```{r}
+```{r linear-trend}
 scatter1+geom_smooth(method="lm")
@@
 
-```{r}
+```{r linear-fit}
 linear_fit <- lm(adjusted_price ~ carat, data=di)
@@
 
-```{r}
+```{r spline-fit}
 spline_fit <- lm(adjusted_price ~ rcs(carat), data=di)
@@
 
-```{r}
+```{r spline-plot}
 carat_sequence <- data.frame(
@@
 
-```{r}
+```{rlinear-plus-spline}
 linear_plus_spline_fit <- lm(

### 2024-05-27 text/24/survival-lasso.Rmd
Moved r-datasets from pending to 24

--- 2024-05-11 text/pending/survival-lasso.Rmd
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-library(broom)
 library(glmnet)
 library(glue)
-library(pspline)
 library(readr)
-library(rms)
 library(survival)
@@
 
+## Description of the whas500 dataset
+
 Here is a brief description of the whas500 dataset, taken from the [data dictionary][sim3] on my github site.
@@
 
-```{r}
+```{r, echo=FALSE}
 vnames <- names(dd$vars)
+# for some reason, R wants to read in 23 
+# columns rather than 22. The last column 
+# is all blanks and will be tossed out.
+#
+# Dates are always a bit messy and I do
+# not need them for this analysis, so I
+# will keep things simple and read them 
+# in as strings.
 f2 <- "whas500.dat"
@@
   file=paste0(g0, r0, f2), 
-  col_names=c(vnames),
+  col_names=vnames,
   col_types="nnnnnnnnnnnnnnncccnnnn")
-head(w0)
+w1 <- data.frame(w0)
+head(w1)
 ```
@@
 
-```{r}
+```{r, echo=FALSE}
 categorical <- 
@@
   cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
-  w0 %>%
+  w1 %>%
     count(.data[[vnames[i_vars]]]) %>%
@@
 
-```{r}
+```{r, echo=FALSE}
 continuous <- 
@@
   cat(glue("{vnames[i_vars]}: {vlabels[i_vars]}"), "\n")
-  print(summary(w0[vnames[i_vars]]))
+  print(summary(w1[vnames[i_vars]]))
   cat("\n")
@@
 
+## Simple analysis
+
+I am going to show the R code from this point onward so you can see the actual programming required to fit a lasso model.
+
 Here is an overall survival curve.
@@
 ```{r}
-plot(Surv(w0$lenfol, w0$fstat))
+plot(Surv(w1$lenfol, w1$fstat))
 ```
@@
 simple_fit <- coxph(
-  Surv(w0$lenfol, w0$fstat) ~
+  Surv(w1$lenfol, w1$fstat) ~
       cvd +
@@
       mitype,
-    data=w0)
+    data=w1)
 coef(simple_fit)
 ```
+
+## The lasso model
 
@@
 ```{r}
-ymat <- cbind(w0$lenfol, w0$fstat)
+ymat <- cbind(w1$lenfol, w1$fstat)
 dimnames(ymat)[2] <- list(c("time", "status"))
-xmat <- as.matrix(w0[ , 8:14])
+xmat <- as.matrix(w1[ , 8:14])
 dimnames(xmat)[2] <- list(vnames[8:14])
@@
 
-Let;s spend a bit of time looking at this graph. The left hand side of the graph represents a very small penalty.
+Let's spend a bit of time looking at this graph. The left hand side of the graph represents a very small penalty.
 
@@
 
-What happens at the largest value of lambda.
+What happens at the largest value of lambda?
 
@@
 Once lambda gets large enough, the penalty is so great that the algorithm zeroes out every single variable.
+
+## Cross-validation to select the optimal value of lambda
 

### 2024-05-27 text/24/titanic-imputation.Rmd
Moved r-datasets from pending to 24

--- 2024-05-11 text/pending/titanic-imputation.Rmd
@@
 
-Splines provide a useful way to model relationships that are more complex than a simple linear relationship.
-They work in a variety of regression models.
-Here is an illustration of how to use a spline in a logistic regression model with data from survival of passengers on the Titanic.
+Multiple imputation is a way to properly account for missing values without causing bias.
+Simpler forms of imputation, such as replacing missing values with the mean of the non-missing values, can produce serious problems.
+You might think that ignoring the missingness and relying just on records with complete data for all key variables would be acceptable, but this also can produce serious problems.
+I want to illustrate a simple example of multiple imputation using data on mortality from the Titanic.
 
@@
 ```{r, echo=FALSE}
-f2 <- "titanic_v00.txt"
+f2 <- "titanic.txt"
 raw_data <- 

### 2024-05-27 text/24/titanic-spline.Rmd
Moved r-datasets from pending to 24

@@
 f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "titanic-data-dictionary.yaml"
+f1 <- "titanic.yaml"
 dd <- read_yaml(paste0(f0, f1))
@@
 ```{r}
-f2 <- "titanic_v00.txt"
+f2 <- "titanic.txt"
 ti0 <- 

### 2024-06-13 text/pending/different-end-times.Rmd
Added scientific-fakery

---
title: Different end times in survival data
source: "New"
author: Steve Simon
date: 2024-05-28
categories:
- Blog post
tags:
- Survival analysis
output: html_document
---

Dear Professor Mean, I am running  a surival analysis with two groups and one group was followed for 190 days and the other for 225 days. Would that cause any bias?

<!---more--->

Dear Reader,

Your concern, clearly is that some events might have occurred in the first group between day 191 and day 225. You are smart to be careful about this. I'd have to look at the math, but I am pretty sure that it all comes out clean.

Here's a simple simulation using R.

Generate two sets of survival times.

```{r}
suppressMessages(library(survival))
suppressMessages(library(tidyverse))
set.seed(190225)
raw_time1 <- rexp(100, rate=0.004)
raw_time2 <- rexp(100, rate=0.008)
sim <- data.frame(
  g=rep(c("a", "b"), each=100), 
  raw_time=c(raw_time1, raw_time2))
head(sim)
tail(sim)
```

How many events occur from 0 to 190 days, from 191 to 225 days, and after 225 days?

```{r}
sim %>%
  group_by(g) %>%
  summarize(
    n_0_to_190=sum(raw_time <= 190),
    n_191_to_225=sum(raw_time > 190&raw_time<=225),
    n_above_225=sum(raw_time > 225))
```

Censor the first group at 190 days and the second group at 225.

```{r}
censor_time_v1 <- rep(c(190, 225), each=100)
sim %>%
  mutate(t_v1=pmin(raw_time, censor_time_v1)) %>%
  mutate(censor_v1=as.numeric(t_v1 < censor_time_v1)) -> sim_v1
sim_v1 %>%
  count(g, censor_v1)
```

Run a Cox regression model

```{r}
cox_v1 <- coxph(Surv(t_v1, censor_v1)~g, data=sim_v1)
cox_v1
```

What about those patients in the first group who might have experienced the event if they were evaluated at the full 225 days like the second group. Recreate the analysis using a common end time.

```{r}
censor_time_v2 <- rep(c(225, 225), each=100)
sim %>%
  mutate(t_v2=pmin(raw_time, censor_time_v2)) %>%
  mutate(censor_v2=as.numeric(t_v2 < censor_time_v2)) -> sim_v2
sim_v2 %>%
  count(g, censor_v2)
```

Run a Cox regression model

```{r}
cox_v2 <- coxph(Surv(t_v2, censor_v2)~g, data=sim_v2)
cox_v2
```

What if you ignored the events in the second group if they occurred in between 191 and 225 days?

```{r}
censor_time_v3 <- rep(c(190, 190), each=100)
sim %>%
  mutate(t_v3=pmin(raw_time, censor_time_v3)) %>%
  mutate(censor_v3=as.numeric(t_v3 < censor_time_v3)) -> sim_v3
sim_v3 %>%
  count(g, censor_v3)
```

Run a Cox regression model

```{r}
cox_v3 <- coxph(Surv(t_v3, censor_v3)~g, data=sim_v3)
cox_v3
```

### 2024-06-13 text/14/axis.Rmd

--- 2022-01-08 text/14/axis.Rmd
@@
 category: Blog post
-tags: Sample size
+tags:
+- Sample size justification
 output: html_document
+page_update: partial
 ---

### 2024-06-13 text/14/qqplot.Rmd

@@
 output: html_document
+page_update: partial
 ---
@@
 
-You can use the housing data set to illustrate the use of the normal probability prlot.
+You can use the housing data set to illustrate the use of the normal probability plot.
+
 ```{r read-and-display-data}
@@
 
-Figure 1. Normal probabilty plot for house prices.
+Figure 1. Normal probability plot for house prices.
 

### 2024-06-13 text/22/r-creatures.Rmd
Fixed two typos in r-creatures

@@
 tags:
-- R programming
+- R software
 output: html_document
@@
 
-You can (and probably should) use names for each element of your list. This makes it easier to pick out pieces of list.
+You can (and probably should) use names for each element of your list. This makes it easier to pick out pieces of list. Either of the first two statements will add names to a list.
 
@@
 You use the `tibble` function to create a tibble.
-But first you have to load the tidyverse packages (or just the tibble package by itself)
+But first you have to load the tidyverse packages (or just the tidyr package by itself)
 

### 2024-06-13 text/24/diamond-spline.Rmd

@@
 output: html_document
+page_update: complete
 ---

### 2024-06-13 text/24/survival-spline.Rmd

@@
 f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "whas100-data-dictionary.yaml"
+f1 <- "whas100.yaml"
 dd <- read_yaml(paste0(f0, f1))

### 2024-06-17 src/count_completed.Rmd

---
title: "Count pages completed in each year"
author: "Steve Simon"
date: "Created 2024-06-16"
output: html_document
---

### Setup

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo=TRUE)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(tidyverse)))
```

### Read md files

```{r read_md}
md_path <- "../text/"
subdir_list=c("99", paste0("0", 0:9), 10:24)
md_files <- NULL
for (yr in rev(subdir_list)) {
  file_path <- paste0(md_path, yr)
  file_path %>%
    list.files(pattern="md$") %>%
    data.frame(yr=yr, fn=.) %>%
    bind_rows(md_files) -> md_files
}
count(md_files, yr)
```

### List duplicates

```{r list-duplicates}
md_files %>%
  group_by(fn) %>%
  mutate(n=n()) %>%
  filter(n>1) %>%
  select(-n) %>%
  arrange(fn, yr) %>%
  data.frame
```

### Count page_update fields

```{r md_links}
c_text <- "page_update: "
md_files$page_update_line <- "999"
for (i_row in 1:nrow(md_files)) {
  full_file_name <- glue("{md_path}/{md_files$yr[i_row]}/{md_files$fn[i_row]}")
  read_lines(full_file_name) %>%
    str_subset(fixed(c_text)) %>%
    paste0("", collapse=", ") -> md_files$page_update_line[i_row]
}
count(md_files, page_update_line)
count(md_files, yr, page_update_line)
```

```{r save_everything}
save(md_files, file="../data/page_update_check.RData")
```

### 2024-06-20 text/22/analysis-of-means.Rmd
Update 2022 files

@@
 source: new
+page_update: complete
 ---

### 2024-06-20 text/22/definition-schoenfeld-residuals.Rmd
Update 2022 files, part 3

--- 2022-12-09 text/22/definition-schoenfeld-residuals.Rmd
@@
 category: Blog post
-tags: Survival analysis
+tags:
+- Survival analysis
 source: new
+page_update: complete
 ---

### 2024-06-20 text/22/dunnetts-test.Rmd
Update 2022 files, part 3

--- 2022-09-06 text/22/dunnetts-test.Rmd
@@
 category: Blog post
-tags: Analysis of means
+tags:
+- Analysis of means
 source: new
+page_update: complete
 ---

### 2024-06-20 text/23/normal-hedging-hyperprior-part-3.Rmd
More formula updates on hedging hyperprior
Update 2023 files, part 2

@@
 $\frac{\Gamma \left(\frac{\nu+1}{2} \right)} {\sqrt{\nu\pi}\,\Gamma \left(\frac{\nu}{2} \right)} \left(1+\frac{x^2}{\nu} \right)^{-\frac{\nu+1}{2}}$
+
+### A simple numerical example
 

### 2024-06-20 text/23/too-many-colors.rmd
Update 2023 files, part 2

@@
 output: word_document
+page_update: complete
 ---

### 2024-06-20 text/20/map-projections.Rmd
Update 2020 files

--- 2020-07-22 text/20/map-projections.Rmd
@@
 ---
-title: "Blog post: Different map projections"
+title: "Different map projections"
 author: "Steve Simon"
-source: "New"
+source: new
 date: "2020-03-15"
-categories: Blog post
-tags: Graphical display
+categories:
+- Blog post
+tags:
+- Graphical display
 output: html_document
+page_update: complete
 ---
+install.packages("rnaturalearthhires", repos="http://packages.ropensci.org", type="source")
 

### 2024-06-20 text/22/box-with-no-whiskers.Rmd

--- 2022-12-15 text/22/box-with-no-whiskers.Rmd
@@
 category: Blog post
-tags: Graphical display
+tags:
+- Graphical display
 source: new
+page_update: complete
 ---

### 2024-06-20 text/22/different-parameterizations.Rmd
Format change in differentparameterizations.Rmd

@@
 ```
+
 ### Acknowledgements

### 2024-06-20 text/22/posix-time.Rmd

--- 2022-11-13 text/22/posix-time.Rmd
@@
 category: Blog post
-tags: Data science
+tags:
+- Data science
 source: new
+page_update: complete
 ---

### 2024-06-20 text/23/normal-hedging-hyperprior-part-1.Rmd
More formula updates on hedging hyperprior

@@
 
-$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$
+$g(\mu; \mu_0, \tau_0) \propto
+  e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto 
+  e^{- \frac{1}{2} \tau_F\sum(X_i-\mu)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto
+  e^{-\frac{1}{2} \tau_F \sum (X_i-\bar{X}+\bar{X}-\mu)^2}$
 
@@
 
-$L(\mathbf{X} | \mu) \propto e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2} e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$
+$L(\mathbf{X} | \mu) \propto
+  e^{- \frac{1}{2} \tau_F (\sum(X_i-\bar{X})^2}
+  e^{- \frac{n}{2} \tau_F (\bar{X}-\mu)^2}$
 
-$L(\mathbf{X} | \mu) \propto e^{-\frac{n-1}{2}\tau_F S^2} e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$
+$L(\mathbf{X} | \mu) \propto
+  e^{-\frac{n-1}{2}\tau_F S^2}
+  e^{-\frac{n}{2} \tau_F (\mu-\bar{X})^2}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2} e^{- \frac{n-1}{2} \tau_F S^2} e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$
+$h(\mu | \mathbf{X}) \propto
+  e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2}
+  e^{- \frac{n-1}{2} \tau_F S^2}
+  e^{-\frac{n}{2} \tau_F(\mu-\bar{X})^2)}$
 
@@
 
-$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau_0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau_0})^2}$
+$h(\mu | \mathbf{X}) \propto
+  e^{
+    -(n\tau_F+\tau_0)
+    (\mu-
+      \frac
+        {n\tau_F\bar{X}+\tau_0\mu_0}
+        {n\tau_F+\tau_0})^2}$
 
@@
 $\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$
+
+### A simple numerical example
 

### 2024-06-21 src/count_completed.Rmd
Update 2006 files

@@
     str_subset(fixed(c_text)) %>%
+    str_replace(c_text, "") -> page_update
+  n <- length(page_update)
+  if (n > 1) {
+    cat(
+      glue(
+        "\n\n{n} duplicates found in {full_file_name}.\n\n"))}
+  page_update %>%
     paste0("", collapse=", ") -> md_files$page_update_line[i_row]

### 2024-06-22 src/count_completed.Rmd
Update 2008 files

--- 2024-06-17 src/count_completed.Rmd
@@
 count(md_files, page_update_line)
-count(md_files, yr, page_update_line)
+# count(md_files, yr, page_update_line)
 ```
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    count(yr, page_update_line) %>%
+    print
+  cat("\n\n")
+}
+```
+
+```{r}
+test1 <- paste0("## " 1:5)
+```
+
+`r test1`
 

### 2024-06-22 src/count_completed.Rmd
Update 2008, part 2
Update 2008 files, part 3

--- 2024-06-17 src/count_completed.Rmd
@@
 c_text <- "page_update: "
+f_text <- "[sim1]: "
 md_files$page_update_line <- "999"
+md_files$old_link <- "999"
 for (i_row in 1:nrow(md_files)) {
   full_file_name <- glue("{md_path}/{md_files$yr[i_row]}/{md_files$fn[i_row]}")
-  read_lines(full_file_name) %>%
+  read_lines(full_file_name) -> file_text
+  file_text %>%
     str_subset(fixed(c_text)) %>%
+    str_replace(fixed(c_text), "") -> page_update
+  file_text %>%
+    str_subset(fixed(f_text)) %>%
+    str_replace(f_text, "") %>%
+    paste0("") %>%
+    paste0(collapse=", ") -> old_link
+  n <- length(page_update)
+  if (n > 1) {
+    cat(
+      glue(
+        "\n\n{n} duplicates found in {full_file_name}.\n\n"))}
+  page_update %>%
     paste0("", collapse=", ") -> md_files$page_update_line[i_row]
+  md_files$old_link[i_row] <- old_link
 }
 count(md_files, page_update_line)
-count(md_files, yr, page_update_line)
+# count(md_files, yr, page_update_line)
+```
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    count(yr, page_update_line) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages with page_update = `r (update_text="n")`
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(page_update_line==update_text) %>%
+    select(-old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages with page_update = `r (update_text="partial")`
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(page_update_line==update_text) %>%
+    select(-old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages with page_update = `r (update_text="complete")`
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(page_update_line==update_text) %>%
+    select(-old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages with page_update = `r (update_text="")`
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(page_update_line==update_text) %>%
+    select(-old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages with links to previous site
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(length(old_link) > 0) %>%
+    select(yr, old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+### Pages without links to previous site
+
+```{r}
+for (y in subdir_list) {
+  md_files %>%
+    filter(yr==y) %>%
+    filter(length(old_link) == 0) %>%
+    select(yr, old_link) %>%
+    print
+  cat("\n\n")
+}
+```
+
+```{r}
+test1 <- paste0("## ", 1:5, collapse="\n\n")
+```
+
+`r test1`
+
+```{r}
+#| results: asis
+cat(r'(
+# Header
+
+Text
+)')
+```
+
+```{r}
+#| results: asis
+cat(r'(\n# Header\n\nText]\n)')
 ```

### 2024-06-22 text/08/declining-count-model.Rmd
Update 2008 files, part 3

--- 2021-11-20 text/08/declining-count-model.Rmd
@@
 tags:
+- Being updated
 - Poisson regression
 output: html_document
+page_update: partial
 ---

### 2024-06-22 src/count_completed.Rmd
Update 2018 files

@@
 c_text <- "page_update: "
-f_text <- "[sim1]: "
+f_text <- "\[sim1\]\: "
 md_files$page_update_line <- "999"
@@
   file_text %>%
-    str_subset(fixed(f_text)) %>%
+    str_subset(f_text) %>%
     str_replace(f_text, "") %>%
-    paste0("") %>%
-    paste0(collapse=", ") -> old_link
+    paste0("", collapse=", ") -> old_link
   n <- length(page_update)

### 2024-06-22 src/count_completed.Rmd
Update 2021 files

@@
 c_text <- "page_update: "
-f_text <- "\[sim1\]\: "
+f_text <- "\\[sim1\\]\\: "
 md_files$page_update_line <- "999"
@@
 
-### Pages with page_update = `r (update_text="n")`
+### Pages with page_update = `r (update_text="no")`
 

### 2024-06-22 text/21/poisson-regression.Rmd
Update 2021 files

@@
 4282
+page_update: complete
 -------

### 2024-06-22 src/count_completed.Rmd
Fixed duplicate file names

@@
     filter(yr==y) %>%
-    filter(length(old_link) > 0) %>%
+    filter(nchar(old_link) > 0) %>%
     select(yr, old_link) %>%
@@
     filter(yr==y) %>%
-    filter(length(old_link) == 0) %>%
-    select(yr, old_link) %>%
+    filter(nchar(old_link) == 0) %>%
+    select(yr, fn, old_link) %>%
     print

### 2024-06-24 src/year-end-summary.Rmd
Added titanic-spline
Updated year-end-summary

@@
      output: html_document
+     page_update: complete
      ---')

### 2024-06-24 text/01/impact-of-population-size.Rmd
Updated 00/composite-scores

---
title: The minimal impact of population size on power and precision
author: Steve Simon
date: 2001-01-19
categories:
- Blog post
tags:
- Confidence intervals
- Sample size justification
output: html_document
page_update: complete
---

### Dear Professor Mean,

Can you explain why it is okay to have similar sample sizes for populations of very different sizes. For example, why is it that a sample size for a population of 10 million doesn't have to be much larger than a sample size for a population of 10 thousand?

<!---More--->

### Dear Reader,

It is surprising, but **for the most part, you disregard the size of the population when you try to estimate an appropriate sample size**. There are some exceptions. Suppose you are trying to sample from the population of all ethical lawyers. The sampling simplifies to deciding which one of the two lawyers you want to chose.

Just kidding. I know a lot of lawyers and they are all nice.

The best analogy I have heard about sampling goes something like: "*Every cook knows that it only takes a single sip from a well-stirred soup to determine the taste*." It's a nice analogy because you can visualize what happens when the soup is poorly stirred.

With regards to why a sample size characterizes a population of 10 million and a population of 10 thousand equally--well, use the soup analogy again. A single sip is sufficient both for a small pot and a large pot.

### Finite Population Correction factor (fpc)

When the size of the sample becomes a large fraction of the size of the population, this analogy no longer holds. In this situation, we use something called a finite population correction factor (fpc). **The finite population correction factor measures how much extra precision we achieve when the sample size become close to the population size.**

The formula for fpc is

$\sqrt{\frac{N-n}{N-1}}$

where N is the size of the population and n is the size of the sample. If fpc is close to 1, then there is almost no effect. When fpc is much smaller than 1, then sampling a large fraction of the population is indeed having an effect on precision.

The following table shows what the fpc in four different situations would be.

```{r, echo=FALSE}
df <- data.frame(sample_size=c(50, 200, 4000))
df$fpc_pop_ten_thousand <- sqrt((10000-df$sample_size)/(9999))
df$fpc_pop_ten_million <- sqrt((10000000-df$sample_size)/(9999999))
df
```

When the sample size is 50, it does not matter much whether the population is 10 thousand or 10 million. When the sample size, however, is four thousand, then we have about 23% more precision with a population of ten thousand than we would for a population of ten million.

A good rule of thumb is to **use the fpc when the sample is 10% or more of the population**.

Be cautious about using the fpc. **Frequently you want to generalize to a larger population than the one you sampled from**. You may have restricted the population out of convenience, but you are interested in more than just the convenient population. This extrapolation will add to the uncertainty of your estimates, so the last thing you would want to do is to use the fpc to make your confidence intervals narrower.

For example, you might sample from a set of medical records from February to June of 2002, but you really are interested in all records: past, present, or future. Or you might sample from patients at your own hospital but you want to generalize to patients at all hospitals. **The finite population correction factor really only applies to "warehouse" type studies, where you are trying to characterize all the data in a single physical or conceptual location and there is NO interest in data outside of this location.** Warehouse studies are quite common in accounting, but they are unusual in medical research.

### Summary

Uncle Gene wants an explanation of why a sample for a population of 10 million people doesn't have to be much larger than a sample for a population of 10 thousand people. Professor Mean provides an analogy to taking a single sip from a well stirred soup. He then presents the finite population correction factor and shows that it does not make much of a difference unless your sample is a large fraction of the total population.

### Further reading

Frank Newport. Yes, Polling Works. Gallup blog, 2002-11-11. Available in [html format][new1]

"There's little question that some Americans are skeptical of polls and the process by which we use small samples to represent the views of millions of people. We pick up that skepticism when we poll people about polls (something we do from time to time!), and I certainly hear it when I am on a radio talk show or make a speech and get bombarded with questions about the believability of our polls, which are based on what seems to the questioners to be ridiculously small numbers of people."

You can find an [earlier version][sim1] of this page on my [original website][sim2].

[sim1]: http://www.pmean.com/01/population.html
[sim2]: http://www.pmean.com/original_site.html

[new1]: https://news.gallup.com/poll/7174/yes-polling-works.aspx

### 2024-06-24 src/count_completed.Rmd
Updated year-end-summary
Updated 04/fda-expert-coi

--- 2024-06-22 src/count_completed.Rmd
@@
 ```{r}
-for (y in subdir_list) {
 md_files %>%
-    filter(yr==y) %>%
   filter(page_update_line==update_text) %>%
-    select(-old_link) %>%
-    print
-  cat("\n\n")
-}
+  select(-old_link)
 ```
@@
 ```{r}
-for (y in subdir_list) {
 md_files %>%
-    filter(yr==y) %>%
   filter(page_update_line==update_text) %>%
@@
   print
-  cat("\n\n")
-}
 ```
@@
 ```{r}
-for (y in subdir_list) {
 md_files %>%
-    filter(yr==y) %>%
-    filter(length(old_link) > 0) %>%
-    select(yr, old_link) %>%
-    print
-  cat("\n\n")
-}
+  filter(nchar(old_link) > 0) %>%
+  glue_data("-   [{fn}]({old_link})\n") -> linked
 ```
+
+`r paste0(linked, collapse="\n")`
 
@@
 ```{r}
-for (y in subdir_list) {
 md_files %>%
-    filter(yr==y) %>%
-    filter(length(old_link) == 0) %>%
-    select(yr, old_link) %>%
-    print
-  cat("\n\n")
-}
+  filter(nchar(old_link) == 0) %>%
+  select(yr, fn, old_link)
 ```

### 2024-06-25 text/23/normal-hedging-hyperprior-part-1.Rmd
Updated 2022 files, part 2

@@
 
-Assume that $\tau$ is known and equal to $\tau_F$. Other cases with [mean known and precision unknown][sim2] and with [both mean and precision unknown][sim3] appear on different web pages.
+Assume that $\tau$ is known and equal to $\tau_F$. Other cases with [mean known and precision unknown][sim4] and with [both mean and precision unknown][sim5] appear on different web pages.
 
-[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
-[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3
+[sim4]: http://new.pmean.com/normal-hedging-hyperprior-part2
+[sim5]: http://new.pmean.com/normal-hedging-hyperprior-part3
 

### 2024-06-25 text/23/normal-hedging-hyperprior-part-2.Rmd
Updated 2022 files, part 2

@@
 Assume that both the mean and the precision of the normal distribution are unknown.
-Other cases with [mean unknown and precision known][sim1] and with [both mean and precision unknown][sim3] appear on a different web pages.
+Other cases with [mean unknown and precision known][sim3] and with [both mean and precision unknown][sim5] appear on a different web pages.
 
-[sim1]: http://new.pmean.com/normal-hedging-hyperprior-part1
-[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3
+[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part1
+[sim5]: http://new.pmean.com/normal-hedging-hyperprior-part3
 

### 2024-06-25 text/23/normal-hedging-hyperprior-part-3.Rmd
Updated 2022 files, part 2

@@
 Assume that both the mean and the precision of the normal distribution are unknown.
-Earlier cases with [mean unknown and precision known][sim1] and with [mean known and precision unknown][sim2] appear on a different web pages.
+Other cases with [mean unknown and precision known][sim3] and with [mean known and precision unknown][sim4] appear on a different web pages.
 
-[sim1]: http://new.pmean.com/normal-hedging-hyperprior-part1
-[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
+[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part1
+[sim4]: http://new.pmean.com/normal-hedging-hyperprior-part2
 

### 2024-06-25 src/audit-old-pages.Rmd
Started work on audit-old-pages

--- 2023-02-24 src/check-old-pages.Rmd
@@
 
-# path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
-path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website"
+path_prefix <- "c:/Users"
+path_suffix <- "dropbox/professional/fun stuff/update-website"
+
+list.files(path_prefix) %>%
+  str_subset("(Simons)|(steve)") -> me
+
+full_path <- glue("{path_prefix}/{me}/{path_suffix}")
+
+print(glue("Searching in {full_path}"))
+
+list.files(full_path)
 ```
@@
   # List all html files in a specific folder
-  full_path <- glue("{path}/{sub_dir}/") 
-  full_path %>%
+  subd_path <- glue("{full_path}/{sub_dir}/") 
+  subd_path %>% 
     list.files %>%
@@
     # Read text from html files
-    file_text <- read_lines(glue("{full_path}/{fn}"))
+    file_text <- read_lines(glue("{subd_path}/{fn}"))
     # Search for key phrase

### 2024-06-25 text/22/analysis-of-means.Rmd

@@
 
-where $\bar Y_{..}$ is the overall mean and $h(\alpha, t, N)$ comes from a table such as [this one][lev1] (see page 3) or [this one][sim3].
+where $\bar Y_{..}$ is the overall mean and $h(\alpha, t, N)$ comes from a table such as [this one][lev1] (see page 3) or [this one][sim3a].
 
@@
 
-[sim3]: http://www.pmean.com/07/AnomTable05Part1.html
+[sim3a]: http://www.pmean.com/07/AnomTable05Part1.html

### 2024-06-25 src/audit-cross-links.Rmd
Added audit-cross-links

---
title: "Check links on new pages"
author: "Steve Simon"
date: "Created 2024-06-25"
output: html_document
---

This program looks for key information in the new webpages

```{r setup}
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(readr)))
suppressMessages(suppressWarnings(library(stringr)))
suppressMessages(suppressWarnings(library(tidyverse)))

load(file="../data/check_new.RData")
load(file="../data/check_old.RData")
```

### create subsets

```{r create}
md %>% 
  filter(miss==0) %>%
  select(yr, fn, stat, base) -> md1

html %>% 
  filter(miss==0) %>%
  mutate(base=fn) %>%
  mutate(html_yr=yr) %>%
  select(base, html_yr) -> html1

match1 <- full_join(md1, html1, by="base")

md %>% 
  filter(miss==0) %>%
  mutate(base=fn) %>%
  mutate(md_yr=yr) %>%
  select(base, md_yr) -> md2

html %>% 
  filter(miss==0) %>%
  mutate(base=link) %>%
  select(fn, yr, base) -> html2

match2 <- full_join(html2, md2, by="base")

```

### mistmatch 1a, new pages that link to non-existent old pages

```{r}
match1 %>% 
  filter(is.na(html_yr))
```

### mistmatch 1b, old pages that link back without recognition

```{r}
match1 %>% 
  filter(is.na(fn))
```

### mistmatch 2a, old pages that link to non-existent new pages

```{r}
match2 %>% 
  filter(is.na(md_yr))
```

### mistmatch 2b, new pages that link back without recognition

```{r}
match2 %>% 
  filter(is.na(fn))
```

### 2024-06-25 src/audit-old-pages.Rmd
Added audit-cross-links

--- 2023-02-24 src/check-old-pages.Rmd
@@
 author: "Steve Simon"
-date: "Created 2023-02-20"
+date: "Created 2024-06-25"
 output: html_document
@@
 suppressMessages(suppressWarnings(library(stringr)))
+suppressMessages(suppressWarnings(library(tidyverse)))
 
-# pad single digits with leading zero
-zpad <- function(x) {
-  ifelse(nchar(x)==1, paste0("0", x), x)
+path1 <- "c:/Users"
+path2 <- str_subset(list.files(path1), "steve|simons")
+path3 <- "Dropbox/professional/fun stuff/update-website"
+
+"c:/Users/steve/Dropbox/professional/fun stuff/update-website"
+"c:/Users/simons/Dropbox/professional/fun stuff/update-website"
+
+html_path <- paste(path1, path2, path3, sep="/")
+subdir_list=c(
+  "00",
+  "00files",
+  "01",
+  "02",
+  "03",
+  "04",
+  "05",
+  "06",
+  "07",
+  "08",
+  "08a",
+  "09",
+  "10",
+  "11",
+  "12",
+  "12a",
+  "13",
+  "14",
+  "15",
+  "16",
+  "17",
+  "18",
+  "99",
+  "archive",
+  "category",
+  "definitions",
+  "weblog",
+  "08/accrual",
+  "08/definitions",
+  "08/diagnostic",
+  "08/faq",
+  "08/images",
+  "08/journal",
+  "08/koans",
+  "08/software",
+  "08/training",
+  "08/training/abstracts",
+  "08/training/datasets",
+  "08/training/exercises",
+  "08/training/extras",
+  "08/training/images",
+  "08/website")
+save_file <- "../data/check_old.RData"
+```
+
+### create html
+
+```{r create}
+html <- NULL
+for (yr in rev(subdir_list)) {
+  file_path <- paste(html_path, yr, sep="/")
+  file_path %>%
+    list.files(pattern="asp|htm$|html$") -> piper
+  if (length(piper) == 0) next
+  piper %>%
+    str_remove("\\..*") -> fn
+  piper %>%
+    str_remove(".*\\.") -> ext
+  data.frame(yr=yr, fn=fn, ext=ext) %>%
+    bind_rows(html) -> html
+}
+n <- nrow(html)
+```
+
+### Search for links to old files
+
+```{r links}
+search_link <- function(tx) {
+  prefix1 <- '<p>This page is moving to a <a href="'
+  prefix2 <- '<p>This page has moved to a <a href="'
+  postfix <- '">new website</a>.</p>'
+  prepost <- paste(prefix1, prefix2, postfix, sep="|")
+  website <- "http://new.pmean.com/"
+  slash   <- "/"
+  tx %>%
+    str_subset(prepost) %>%
+    str_replace(prefix1, "") %>%
+    str_replace(prefix2, "") %>%
+    str_replace(postfix, "") %>%
+    str_replace(website, "") %>%
+    str_replace(slash,   "") %>%
+    paste0("") -> piper
+  piper %>%
+    paste0(collapse=", ")
 }
@@
 
-```{r find-files}
-sub_dir_list <- c("99", zpad(0:12), "12a", zpad(13:18))
+### Search for size
 
-# path <- "c:/Users/Steve/Dropbox/professional/fun stuff/update-website"
-path <- "c:/Users/simons/Dropbox/professional/fun stuff/update-website"
+```{r size}
+search_size <- function(tx) {
+  ty <- str_remove_all(tx, "\\s")
+  length(tx)-sum(ty == "")
+}
 ```
 
-```{r search-files}
-key_text <- "<p>This page"
-n_files <- 0
-n_unlinked <- 0
-for (sub_dir in sub_dir_list) {
-  # List all html files in a specific folder
-  full_path <- glue("{path}/{sub_dir}/") 
-  full_path %>%
-    list.files %>%
-    str_subset("html$") -> html_list
-  i_files <- length(html_list)
-  i_unlinked <- 0
-  for (fn in html_list) {
-    # Read text from html files
-    file_text <- read_lines(glue("{full_path}/{fn}"))
-    # Search for key phrase
-    key_line <- str_subset(file_text, key_text)
-    if (length(key_line) > 0) {next}
-    # List files without key phrase
-    i_unlinked <- i_unlinked + 1
-    print(glue("{sub_dir}/{fn} has no link to new page"))
+### Read file text
+
+```{r read}
+full_name <-
+  paste0(
+  html_path, "/",
+  html$yr, "/",
+  html$fn, ".",
+  html$ext)
+html$size <- -1
+html$link <- "999"
+for (i in 1:n) {
+  tx <- read_lines(full_name[i])
+  html$link[i] <- search_link(tx)
+  html$size[i] <- search_size(tx)
 }
-  cat("\n")
-  print(glue("{sub_dir} has {i_files} files. {i_unlinked} are unlinked."))
-  cat("\n")
-  n_files <- n_files + i_files
-  n_unlinked <- n_unlinked + i_unlinked
-}
-print(glue("All directories have a total of  {n_files} files. {n_unlinked} are unlinked."))
+html$miss <- as.numeric(nchar(html$link) == 0)
+html$stub <- as.numeric(html$size <= 10)
+save(html, file=save_file)
 ```
+
+### Simple statistics
+
+```{r simple}
+summary(html$size)
+count(html, ext)
+count(html, miss)
+count(html, stub)
+```

### 2024-06-27 src/audit-cross-links.Rmd
Added nice-layout

@@
 load(file="../data/check_new.RData")
+n_new <- nrow(md)
 load(file="../data/check_old.RData")
+n_old <- nrow(html)
+size_limit <- 99
+
+inc <- function(x) {
+  n <- length(x)
+  m <- c(FALSE, x[-1]==x[-n])
+  x[m] <- " "
+  x
+}
+test1 <- rep(letters[1:4], 1:4)
+test2 <- inc(test1)
+data.frame(test1, test2)
 ```
 
-### create subsets
+### pages with no links from new site to old or blog site
 
-```{r create}
+```{r no-link-www}
+md %>% 
+  filter(miss==1) %>%
+  select(yr, fn, stat) -> pip1 
+print(nrow(pip1))
+glue("{nrow(pip1)} / {n_new} = {round(100*nrow(pip1)/n_new)}%")
+pip1 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
+```
+
+### pages with bad links from new site to old site
+
+```{r bad-link-www}
+html %>%
+  rename(base=fn) %>%
+  rename(subd=yr) %>%
+  select(base, subd, stub) -> html1
 md %>% 
   filter(miss==0) %>%
-  select(yr, fn, stat, base) -> md1
+  filter(site=="www") %>%
+  left_join(html1, by=c("subd", "base")) %>%
+  select(yr, fn, stat, subd, base, stub) -> pip1
+pip1 %>%
+  filter(is.na(stub)) %>%
+    select(-stub) -> pip2 
+glue("{nrow(pip2)} / {n_new} = {round(100*nrow(pip2)/n_new)}%")
+pip2 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
+```
+
+### pages with good links from new site to old site
+
+```{r good-link-www}
+pip1 %>%
+  filter(!is.na(stub))  -> pip3 
+glue("{nrow(pip3)} / {n_new} = {round(100*nrow(pip3)/n_new)}%")
+pip3 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
+```
+
+### pages with no links from old site to new site
+
+```{r no-link-new}
+html %>% 
+  filter(miss==1) %>%
+  select(yr, fn, stub) -> pip1 
+print(nrow(pip1))
+pip1 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
+```
+
+### pages with bad links from old site to new site
+
+```{r bad-link-new}
+md %>%
+  select(-link) %>%
+  rename(link=fn) %>%
+  select(link, stat) -> md1
 
@@
   filter(miss==0) %>%
-  mutate(base=fn) %>%
-  mutate(html_yr=yr) %>%
-  select(base, html_yr) -> html1
-
-match1 <- full_join(md1, html1, by="base")
-
-md %>% 
-  filter(miss==0) %>%
-  mutate(base=fn) %>%
-  mutate(md_yr=yr) %>%
-  select(base, md_yr) -> md2
-
-html %>% 
-  filter(miss==0) %>%
-  mutate(base=link) %>%
-  select(fn, yr, base) -> html2
-
-match2 <- full_join(html2, md2, by="base")
-
+  left_join(md1, by="link") %>%
+  select(yr, fn, link, stat) -> pip1
+pip1 %>%
+  filter(is.na(stat)) %>%
+  select(-stat)  -> pip4 
+print(nrow(pip4))
+pip4 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
 ```
 
-### mistmatch 1a, new pages that link to non-existent old pages
+### pages with good links from old site to new site
 
-```{r}
-match1 %>% 
-  filter(is.na(html_yr))
+```{r good-link-new}
+pip1 %>%
+  filter(!is.na(stat)) -> pip5 
+print(nrow(pip5))
+pip5 %>%
+  mutate(yr=inc(yr)) %>%
+  head(size_limit)
 ```
-
-### mistmatch 1b, old pages that link back without recognition
-
-```{r}
-match1 %>% 
-  filter(is.na(fn))
-```
-
-### mistmatch 2a, old pages that link to non-existent new pages
-
-```{r}
-match2 %>% 
-  filter(is.na(md_yr))
-```
-
-### mistmatch 2b, new pages that link back without recognition
-
-```{r}
-match2 %>% 
-  filter(is.na(fn))
-```

### 2024-06-27 src/audit-old-pages.Rmd
Added nice-layout

@@
   tx %>%
-    str_subset(prepost) %>%
+    str_subset(postfix) %>%
     str_replace(prefix1, "") %>%
@@
     str_replace(slash,   "") %>%
+    str_remove_all("\\s") %>%
+    str_trim %>%
     paste0("") -> piper
@@
 ```{r read}
+# The str_trim function is not working. There
+#  may be an encoding problem. See
+#   "about_encoding" in the stringi package
+#   https://search.r-project.org/CRAN/refmans/
+#     stringi/html/about_encoding.html
+#
+# Help may come from the "stri_enc_toutf8" function 
+#   https://search.r-project.org/CRAN/refmans/
+#     stringi/html/stri_enc_toutf8.html
+# A simpler solution may be
+#   str_remove_all("\\s")
 full_name <-
@@
 html$stub <- as.numeric(html$size <= 10)
+html <- arrange(html, yr, fn)
 save(html, file=save_file)

### 2024-06-27 text/pending/nice-layout.Rmd
Added nice-layout

---
title: Nice layout for comparison of categorical outcomes
source: "New"
author: Steve Simon
date: 2024-06-27
categories:
- Blog post
tags:
- Being updated
output: html_document
page_update: complete
---

While it is easy to get R to produce descriptive statistics for categorical outcomes in a research study, it is a bit harder to get R to produce a layout that is publication-ready. Hard, but not impossible. Here is a simple example.Here is a brief abstract.

<!---more--->

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(glue)
library(mice)
library(readr)
library(rms)
library(tidyverse)
library(yaml)
g0 <- "https://raw.githubusercontent.com/pmean/"
r0 <- "datasets/master/"
f1 <- "titanic.yaml"
s1 <- "yaml-functions.R"
source(paste0(g0, r0, s1))
dd <- read_yaml(paste0(g0, r0, f1))
vnames <- extract_names(dd)
vlabels <- extract_labels(dd)
```

Here is a brief description of this dataset, taken from the [data dictionary][sim3] on my github site.

> `r dd$description`

[sim3]: https://raw.githubusercontent.com/pmean/datasets/master/titanic-data-dictionary.yaml

Here are the first few rows of data.

```{r, echo=FALSE}
f2 <- "titanic.txt"
raw_data <- 
  data.frame(
    read_tsv(
      file=paste0(g0, r0, f2),
      col_types="ccncn"))
# Re-arranging columns just for convenient display
o <-  c(1, 3, 2, 4, 5)
ti0 <- raw_data[ , o]
vnames <- vnames[o]
head(ti0)
```

I have hidden the R code up to this point, as it is mundane and not of great interest. I will show the R code and output for the rest of the analysis.

Using the `count` command in R shows deaths and survivors among men and women.

```{r}
ti0 %>%
  group_by(Sex, Survived) %>%
  summarize(n1=n(), .groups="drop") -> tab01
tab01
```

A quick note about the code. First, it would be simpler and better to use the single function `count` instead of the two functions `group_by` and `summarize`. I prefer the two functions because you can generalize a bit more easily to other cases like using a continuous outcome.

Second, the `.groups` argument is new. When you use a `group_by` statement with two or more variables, the result is a grouped tibble. The summarize function will produce either a grouped tibble with everything except the last variable or a tibble that is not grouped. Most of the time (in my experience), you want the latter, but the default is the former. To produce a straight tibble rather than a grouped tibble, use the `.groups = "drop"` argument in `summarize`.

If you examine the results, there are a lot more male deaths than female deaths, but be careful because there were also more men than women on the boat. Calculating the proportions of deaths and survivors among men and women is fairly easy.

```{r}
tab01 %>%
  group_by(Sex) %>%
  mutate(n2=sum(n1)) %>%
  mutate(pct=round(100*n1/n2)) -> tab02
tab02
```

There will be times that the `sprintf` function will layout things a bit better than `round`. If you have numbers larger than 1000, you might consider adding a comma (e.g., 1,000) using the `format(., big.mark=",")` to make the large numbers a bit easier to read.

You can combine the numerator, denominator, and percentage into one nice package using the `glue` function.

```{r}
tab02 %>%
  mutate(out=glue("{n1}/{n2} ({pct}%)")) %>%
  select(Sex, Survived, out) -> tab03
tab03
```

Again, the `sprintf` function might work better than `glue`.

Next, arrange things so that the data on survivors is in one column and the data on deaths is in a separate column. This is done with the `pivot_wider` function.

```{r}
tab03 %>%
  pivot_wider(
    names_from=Survived, 
    values_from=out) -> tab05
tab05
```

Use the `set_names` function to produce nicer looking variable names. The `set_names` function is one of the [useful aliases][mag1] available in the magrittr package.

[mag1]: https://magrittr.tidyverse.org/reference/aliases.html

```{r}
tab05 %>%
  set_names(c(
    "Sex", 
    "Deaths", 
    "Survivors")) -> tab06
tab06
```

Next, let's use the kableExtra package to make things look really nice. There are a gazillion different options in kableExtra, and you might want to review the [kableExtra vignette][kab1]. There are [lots of other packages][mad1] out there that can create publication quality tables.

[kab1]: https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html
[mad1]: https://towardsdatascience.com/top-7-packages-for-making-beautiful-tables-in-r-7683d054e541

```{r}
library(kableExtra)
tab06 %>%
  kbl %>%
  kable_paper(
    "striped", 
    full_width=FALSE) -> tab07
tab07
```

If you have multiple independent variables, the `pack_rows` function is nice. Notice that we have to set the name of the first column to blank to avoid redundancy.

```{r}
library(kableExtra)
tab07 %>%
  pack_rows("Sex", 1, 2) -> tab08
tab08
```

Here is the code all in one place rather than step by step. It looks quite complex, but if you build things one step at a time, the complex code is not all that complex to generate.

```{r}
ti0 %>%
  group_by(Sex, Survived) %>%
  summarize(n1=n(), .groups = "drop") %>%
  group_by(Sex) %>%
  mutate(n2=sum(n1)) %>%
  mutate(pct=round(100*n1/n2)) %>%
  mutate(out=glue("{n1}/{n2} ({pct}%)")) %>%
  select(Sex, Survived, out) %>%
  pivot_wider(names_from=Survived, values_from=out) %>%
  set_names(c(" ", "Deaths", "Survivors")) %>%
  kbl %>%
  kable_paper("striped", full_width=FALSE) %>%
  pack_rows("Sex", 1, 2)
```

```{r}
ti0 %>%
  group_by(Sex, Survived) %>%
  summarize(n1=n(), .groups = "drop") %>%
  group_by(Sex) %>%
  mutate(n2=sum(n1)) %>%
  mutate(pct=100*n1/n2) %>%
  mutate(out=sprintf("%3.0f/%3.0f (%3.0f%%)", n1, n2, pct)) %>%
  select(Sex, Survived, out) %>%
  pivot_wider(names_from=Survived, values_from=out) %>%
  set_names(c(" ", "Deaths", "Survivors")) %>%
  kbl %>%
  kable_paper("striped", full_width=FALSE) %>%
  pack_rows("Sex", 1, 2)
```

### 2024-06-28 src/audit-cross-links.Rmd
Updated audit files

--- 2024-06-25 src/audit-cross-links.Rmd
@@
 ```{r setup}
+verbose <- TRUE
 suppressMessages(suppressWarnings(library(glue)))
@@
 load(file="../data/check_new.RData")
+n_new <- nrow(new)
 load(file="../data/check_old.RData")
+n_old <- nrow(old)
+size_limit <- 99
+
+inc <- function(x) {
+  n <- length(x)
+  m <- c(FALSE, x[-1]==x[-n])
+  x[m] <- " "
+  x
+}
+test1 <- rep(letters[1:4], 1:4)
+test2 <- inc(test1)
+data.frame(test1, test2)
 ```
 
-### create subsets
+## pages with no links from new site to old or blog site
 
-```{r create}
-md %>% 
-  filter(miss==0) %>%
-  select(yr, fn, stat, base) -> md1
-
-html %>% 
-  filter(miss==0) %>%
-  mutate(base=fn) %>%
-  mutate(html_yr=yr) %>%
-  select(base, html_yr) -> html1
-
-match1 <- full_join(md1, html1, by="base")
-
-md %>% 
-  filter(miss==0) %>%
-  mutate(base=fn) %>%
-  mutate(md_yr=yr) %>%
-  select(base, md_yr) -> md2
-
-html %>% 
-  filter(miss==0) %>%
-  mutate(base=link) %>%
-  select(fn, yr, base) -> html2
-
-match2 <- full_join(html2, md2, by="base")
-
+```{r no-link-www-1}
+new %>% 
+  filter(link_miss==1) %>%
+  select(dir, page) -> pip1 
+msg <- glue(
+  "{nrow(pip1)} / {n_new} = {round(100*nrow(pip1)/n_new)}%")
 ```
 
-### mistmatch 1a, new pages that link to non-existent old pages
+### `r msg`
 
-```{r}
-match1 %>% 
-  filter(is.na(html_yr))
+```{r no-lin-www-2}
+if (verbose) {
+  pip1 %>%
+    mutate(dir=inc(dir)) %>%
+    head(size_limit)
+}
 ```
 
-### mistmatch 1b, old pages that link back without recognition
+### pages with bad links from new site to old site
 
-```{r}
-match1 %>% 
-  filter(is.na(fn))
+```{r bad-link-www}
+old %>%
+  rename(back_link=link_page) %>%
+  rename(link_page=page) %>%
+  rename(link_dir=dir) %>%
+  select(link_dir, link_page, back_link, stub) -> mrg
+new %>% 
+  filter(link_miss==0) %>%
+  filter(link_site=="www") %>%
+  left_join(mrg, by=c("link_dir", "link_page")) %>%
+  select(
+    dir, 
+    page, 
+    link_dir, 
+    link_page, 
+    back_link,
+    stub) -> pip1
+pip1 %>%
+  filter(is.na(stub)) %>%
+    select(-stub) -> pip2 
+msg <- glue(
+  "{nrow(pip2)} / {n_new} = {round(100*nrow(pip2)/n_new)}%")
+pip2 %>%
+  mutate(dir=inc(dir)) %>%
+  head(size_limit)
 ```
 
-### mistmatch 2a, old pages that link to non-existent new pages
+### pages with good links from new site to old site
 
-```{r}
-match2 %>% 
-  filter(is.na(md_yr))
+```{r good-link-www}
+pip1 %>%
+  filter(!is.na(stub))  -> pip3 
+glue("{nrow(pip3)} / {n_new} = {round(100*nrow(pip3)/n_new)}%")
+pip3 %>%
+  mutate(dir=inc(dir)) %>%
+  head(size_limit)
 ```
 
-### mistmatch 2b, new pages that link back without recognition
+### pages with no links from old site to new site
 
-```{r}
-match2 %>% 
-  filter(is.na(fn))
+```{r no-link-new}
+old %>% 
+  filter(link_miss==1) %>%
+  select(dir, page) -> pip1 
+print(nrow(pip1))
+pip1 %>%
+  mutate(dir=inc(dir)) %>%
+  head(size_limit)
 ```
+
+### pages with bad links from old site to new site
+
+```{r bad-link-new}
+new %>%
+  rename(back_link=link_page) %>%
+  rename(link_page=page) %>%
+  select(link_page, back_link, update) -> mrg
+
+old %>%
+  filter(link_miss==0) %>%
+  left_join(mrg, by="link_page") %>%
+  select(dir, page, link_page, update) -> pip1
+
+pip1 %>%
+  filter(is.na(update)) %>%
+  select(-update)  -> pip4
+
+print(nrow(pip4))
+pip4 %>%
+  mutate(dir=inc(dir)) %>%
+  head(size_limit)
+```
+
+### pages with good links from old site to new site
+
+```{r good-link-new}
+pip1 %>%
+  filter(!is.na(update)) -> pip5 
+print(nrow(pip5))
+pip5 %>%
+  mutate(dir=inc(dir)) %>%
+  head(size_limit)
+```

### 2024-06-28 src/audit-old-pages.Rmd
Updated audit files

@@
 
-html_path <- paste(path1, path2, path3, sep="/")
-subdir_list=c(
+old_path <- paste(path1, path2, path3, sep="/")
+old_list=c(
   "00",
@@
 
-### create html
+### create old
 
 ```{r create}
-html <- NULL
-for (yr in rev(subdir_list)) {
-  file_path <- paste(html_path, yr, sep="/")
+old <- NULL
+for (dir in rev(old_list)) {
+  file_path <- paste(old_path, dir, sep="/")
   file_path %>%
-    list.files(pattern="asp|htm$|html$") -> piper
+    list.files(pattern="asp$|htm$|html$") -> piper
   if (length(piper) == 0) next
   piper %>%
-    str_remove("\\..*") -> fn
+    str_remove("\\..*") -> page
   piper %>%
     str_remove(".*\\.") -> ext
-  data.frame(yr=yr, fn=fn, ext=ext) %>%
-    bind_rows(html) -> html
+  data.frame(dir=dir, page=page, ext=ext) %>%
+    bind_rows(old) -> old
 }
-n <- nrow(html)
+rows <- nrow(old)
 ```
@@
 ```{r size}
-search_size <- function(tx) {
-  ty <- str_remove_all(tx, "\\s")
-  length(tx)-sum(ty == "")
-}
-```
-
-### Read file text
-
-```{r read}
 # The str_trim function is not working. There
@@
 #   str_remove_all("\\s")
+
+search_size <- function(tx) {
+  ty <- str_remove_all(tx, "\\s")
+  length(tx)-sum(ty == "")
+}
+```
+
+### Read file text
+
+```{r read}
 full_name <-
   paste0(
-  html_path, "/",
-  html$yr, "/",
-  html$fn, ".",
-  html$ext)
-html$size <- -1
-html$link <- "999"
-for (i in 1:n) {
+  old_path, "/",
+  old$dir, "/",
+  old$page, ".",
+  old$ext)
+
+old$size <- -1
+old$link_page <- "999"
+
+for (i in 1:rows) {
   tx <- read_lines(full_name[i])
-  html$link[i] <- search_link(tx)
-  html$size[i] <- search_size(tx)
+  old$size[i] <- search_size(tx)
+  old$link_page[i] <- search_link(tx)
 }
-html$miss <- as.numeric(nchar(html$link) == 0)
-html$stub <- as.numeric(html$size <= 10)
-html <- arrange(html, yr, fn)
-save(html, file=save_file)
+```
+
+```{r modify-variables}
+old$stub <- as.numeric(old$size <= 10)
+old$link_miss <- as.numeric(nchar(old$link_page) == 0)
+```
+
+```{r save}
+old %>%
+  select(
+    dir, 
+    page, 
+    ext, 
+    size,
+    stub,
+    link_miss,
+    link_page) %>%
+  arrange(dir, page) -> old
+
+save(old, file=save_file)
 ```
@@
 ```{r simple}
-summary(html$size)
-count(html, ext)
-count(html, miss)
-count(html, stub)
+summary(old$size)
+count(old, ext)
+count(old, link_miss)
+count(old, stub)
 ```

### 2024-06-28 src/archive/count_completed.Rmd
Convert audit pages to Quarto

---
title: "Count pages completed in each year"
author: "Steve Simon"
date: "Created 2024-06-16"
output: html_document
---

### Setup

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo=TRUE)
suppressMessages(suppressWarnings(library(glue)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(lubridate)))
suppressMessages(suppressWarnings(library(tidyverse)))
md_path <- "../text/"
subdir_list=c("99", paste0("0", 0:9), 10:24)
save_file <- "../data/page_update_check.RData"
```

### create md_files

```{r create, eval=FALSE}
md_files <- NULL
for (yr in rev(subdir_list)) {
  file_path <- paste0(md_path, yr)
  file_path %>%
    list.files(pattern="md$") %>%
    data.frame(yr=yr, fn=.) %>%
    bind_rows(md_files) -> md_files
}
c_text <- "page_update: "
f_text <- "\\[sim1\\]\\: "
md_files$page_update_line <- "999"
md_files$old_link <- "999"
for (i_row in 1:nrow(md_files)) {
  full_file_name <- glue("{md_path}/{md_files$yr[i_row]}/{md_files$fn[i_row]}")
  read_lines(full_file_name) -> file_text
  file_text %>%
    str_subset(fixed(c_text)) %>%
    str_replace(fixed(c_text), "") -> page_update
  file_text %>%
    str_subset(f_text) %>%
    str_replace(f_text, "") %>%
    paste0("", collapse=", ") -> old_link
  n <- length(page_update)
  if (n > 1) {
    cat(
      glue(
        "\n\n{n} duplicates found in {full_file_name}.\n\n"))}
  page_update %>%
    paste0("", collapse=", ") -> md_files$page_update_line[i_row]
  md_files$old_link[i_row] <- old_link
}
save(md_files, file=save_file)
```

```{r}
load(file=save_file)
```

### List duplicates

```{r list-duplicates}
md_files %>%
  group_by(fn) %>%
  mutate(n=n()) %>%
  filter(n>1) %>%
  select(-n) %>%
  arrange(fn, yr) %>%
  data.frame -> piper
if (nrow(piper) > 0) piper
if (nrow(piper) == 0) "None found"
```

### Count page_update fields

```{r md_links}
count(md_files, page_update_line)
```

```{r}
for (y in subdir_list) {
  md_files %>%
    filter(yr==y) %>%
    count(yr, page_update_line) %>%
    print
  cat("\n\n")
}
```

### Pages with page_update = `r (update_text="no")`

```{r}
md_files %>%
  filter(page_update_line==update_text) %>%
  select(-old_link) -> piper
if (nrow(piper) > 0) piper
if (nrow(piper) == 0) "None found"
```

### Pages with page_update = `r (update_text="partial")`

```{r}
md_files %>%
  filter(page_update_line==update_text) %>%
  select(-old_link) -> piper
if (nrow(piper) > 0) piper
if (nrow(piper) == 0) "None found"
```

### Pages with page_update = `r (update_text="complete")`

```{r}
md_files %>%
  filter(page_update_line==update_text) %>%
  select(-old_link)
```

### Pages with page_update = `r (update_text="")`

```{r}
md_files %>%
  filter(page_update_line==update_text) %>%
  select(-old_link) -> piper
if (nrow(piper) > 0) piper
if (nrow(piper) == 0) "None found"
```

### Pages with links to previous site

```{r}
md_files %>%
  filter(nchar(old_link) > 0) %>%
  glue_data("-   [{fn}]({old_link})\n") -> linked
```

`r paste0(linked, collapse="\n")`

### Pages without links to previous site

```{r}
md_files %>%
  filter(nchar(old_link) == 0) %>%
  select(yr, fn, old_link) -> piper
if (nrow(piper) > 0) piper
if (nrow(piper) == 0) "None found"
```

```{r}
test1 <- paste0("## ", 1:5, collapse="\n\n")
```

`r test1`

```{r}
#| results: asis
cat(r'(
# Header

Text
)')
```

```{r}
#| results: asis
cat(r'(\n# Header\n\nText]\n)')
```

### 2024-06-28 src/audit-old-pages.qmd
Convert audit pages to Quarto

--- 2024-06-28 src/audit-old-pages.Rmd
@@
 date: "Created 2024-06-25"
-output: html_document
+execute:
+  echo: false
+format:
+  html:
+    embed-resources: true
 ---
@@
 
-### create old
+## Create old
 
@@
 
-### Search for links to old files
+## Search for links to old files
 
@@
 
-### Search for size
+## Search for size
 
@@
 
-### Read file text
+## Read file text
 
@@
 
+## Modify variables
+
 ```{r modify-variables}
@@
 ```
+
+## Save results
 
@@
 
-### Simple statistics
+## Simple statistics
 

### 2024-06-29 text/01/impact-of-population-size.Rmd
Work on stash-files

@@
 
-```{r, echo=FALSE}
+```{r fpc, echo=FALSE}
 df <- data.frame(sample_size=c(50, 200, 4000))

### 2024-06-29 text/24/diamond-spline.Rmd
Work on stash-files

@@
 library(broom)
+library(glue)
 library(pspline)
@@
 library(yaml)
-f0 <- "https://raw.githubusercontent.com/pmean/datasets/master/"
-f1 <- "diamonds.yaml"
-dd <- read_yaml(paste0(f0, f1))
+
+f0 <- "https://raw.githubusercontent.com"
+f1 <- "pmean/datasets/master"
+f2 <- "singapore-diamond-prices.yaml"
+
+dd <- read_yaml(glue("{f0}/{f1}/{f2}"))
 ```
@@
 ```{r read-diamonds}
-f2 <- "diamonds.txt"
-raw_data <- read.fwf(file=paste0(f0, f2), widths=c(5,2,5,4,5))
+f2 <- "singapore-diamond-prices.txt"
+raw_data <- read.fwf(
+  file=glue("{f0}/{f1}/{f2}"), 
+  widths=c(5,2,5,4,5))
 di <- raw_data

### 2024-07-02 src/stash-rmd.R
Added stash-rmd

@@ -1,0 +1,52 @@
+# stash-rmd.R
+# author: Steve Simon
+# date: Created 2024-06-30
+# purpose: use knit to convert Rmd files to md
+
+library(dplyr)
+library(glue)
+library(knitr)
+library(magrittr)
+library(stringr)
+
+## Get file list
+
+yrs <- c(paste0("0", 0:9), as.character(c(10:24, 99)))
+all_files <- NULL
+
+for (y in rev(yrs)) {
+  f0 <- "text"
+  fn <- list.files(glue("{f0}/{y}"))
+  if (length(fn)==0) next
+  data.frame(y, fn) %>%
+    bind_rows(all_files) -> all_files
+}
+count(all_files, y)
+
+## Split files by extension
+
+all_files %>%
+  mutate(ext=str_remove(fn, ".*\\.")) %>%
+  mutate(fn=str_remove(fn, "\\..*")) -> all_files
+count(all_files, ext)
+
+## Define directories for copying
+
+b0 <- "C:/Users"
+b1 <- str_subset(list.files(b0), "steve|simons")
+b2 <- "Dropbox/professional"
+b3 <- "fun stuff/update-website/rmd"
+b4 <- glue("{b0}/{b1}/{b2}/{b3}")
+
+## Knit and copy Rmd files
+
+all_files %>%
+  filter(ext == "Rmd") %>%
+  mutate(a=glue("{f0}/{y}/{fn}.Rmd")) %>%
+  mutate(b=glue("{b4}/{fn}.md")) %>%
+  select(a, b) -> Rmd
+head(Rmd)
+n <- nrow(Rmd)
+for (i in 1:n) {
+  knit(input=Rmd$a[i], output=Rmd$b[i])
+}

### 2024-07-02 text/01/impact-of-population-size.Rmd
Updated 21/ihme-model

@@
 - Confidence intervals
+- Rmarkdown code
 - Sample size justification

### 2024-07-02 text/04/early-stopping-animal.Rmd
Updated 21/ihme-model

--- 2023-02-21 text/04/early-stopping-animal.Rmd
@@
 - Early stopping
+- Rmarkdown code
 output: html_document
+page_update: partial
 ---
@@
 
-```{r, echo=TRUE}
+```{r early-stopping-animal-01, echo=TRUE}
 n <- 5000

### 2024-07-02 text/05/merging-in-r.Rmd
Updated 21/ihme-model

--- 2022-10-25 text/05/MergingInR.Rmd
@@
 - R software
+- Rmarkdown code
 output: html_document
+page_update: partial
 ---
@@
 
-```{r}
+```{r merging-in-r-01}
 # randomly select 20 letters without replacement.
@@
 
-```{r}
+```{r merging-in-r-02}
 intersect(x0,x1)
@@
 
-```{r}
+```{r merging-in-r-03}
 setdiff(x0,x1)
@@
 
-```{r}
+```{r merging-in-r-04}
 x2 <- merge(list(x=x0,id0=1:20),list(x=x1,id1=1:20))
@@
 
-```{r}
+```{r merging-in-r-05}
 x3 <- merge(list(x=x0,id0=1:20),list(x=x1,id1=1:20),all=T)
@@
 
-```{r}
+```{r merging-in-r-06}
 x0 %in% x1
@@
 
-```{r}
+```{r merging-in-r-07}
 x1 %in% x0
@@
 
-```{r}
+```{r merging-in-r-08}
 x0[x0 %in% x1]
@@
 or the setdiff function,
-```{r}
+```{r merging-in-r-09}
 x0[!x0 %in% x1]
@@
 
-A less intuitive version of the %in% operator is the match function. Additional information on merging can be found at
-
--   [www.ats.ucla.edu/stat/r/notes/managing.htm](http://www.ats.ucla.edu/stat/r/notes/managing.htm)
+A less intuitive version of the %in% operator is the match function. Additional information on merging can be found at the [UCLA Statistics website][ucl].
 
@@
 [sim2]: http://www.pmean.com/original_site.html
+
+[ucl]: www.ats.ucla.edu/stat/r/notes/managing.htm

### 2024-07-02 text/08/declining-count-model.Rmd
Updated 21/ihme-model

--- 2021-11-20 text/08/declining-count-model.Rmd
@@
 - Poisson regression
+- Rmarkdown code
 output: html_document
+page_update: partial
 ---
@@
 
-```{r}
+```{r declining-count-model-01}
 ct <- c(9,4,2,3,0,0,1)
@@
 
-```{r}
+```{r declining-count-model-02}
 round(predict(pmod),4)
@@
 
-```{r}
+```{r declining-count-model-03}
 round(exp(predict(pmod)),4)
@@
 by a classic [geometric
-series](http://en.wikipedia.org/wiki/Geometric_series)
+series][wik1].
 
@@
 
-You can find an [earlier version](http://www.pmean.com/08/DecliningCount.html) of this page on my [original website](http://www.pmean.com/original_site.html).
+You can find an [earlier version][sim1] of this page on my [original website](http://www.pmean.com/original_site.html).
+
+[sim1]: http://www.pmean.com/08/DecliningCount.html
+[sim2]: http://www.pmean.com/original_site.html
+
+[wik1]: http://en.wikipedia.org/wiki/Geometric_series

### 2024-07-02 text/11/standard-deviation-ci.Rmd
Updated 21/ihme-model

--- 2021-10-29 text/11/standard-deviation-ci.Rmd
@@
 - Confidence intervals
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r standard-deviation-ci-01}
 qchisq(0.025, 79)
@@
 
-```{r}
+```{r standard-deviation-ci-02}
 qchisq(0.975, 79)

### 2024-07-02 text/14/axis.Rmd
Updated 21/ihme-model

--- 2022-01-08 text/14/axis.Rmd
@@
 category: Blog post
-tags: Sample size
+tags:
+- Data visualization
+- Rmarkdown graphs
 output: html_document
+page_update: partial
 ---
@@
 
-```{r basic-cubic}
+```{r axis-01}
 x <- seq(-2,2,length=1000)
@@
 
-```{r move-tick-marks}
+```{r axis-02}
 xticks <- c(-sqrt(2),-sqrt(2/3),0,sqrt(2/3),sqrt(2))
@@
 
-```{r nicer-labels}
+```{r axis-03}
 plot(x,y,axes=FALSE)
@@
 
-```{r math-expressions}
+```{r axis-04}
 plot(x,y,axes=FALSE)
@@
 
-```{r crossed-axes}
+```{r axis-05}
 xticks <- c(-sqrt(2),-sqrt(2/3),sqrt(2/3),sqrt(2))
@@
 
-```{r new-labels}
+```{r axis-06}
 plot(x,y,xlab="Abcissa",ylab="Ordinate")
@@
 
-```{r tick-label-orientation}
+```{r axis-07}
 # las=0 is the default
@@
 
-```{r square-plot}
+```{r axis-08}
 par(pty="s")
@@
 
-```{r equal-axis-limits}
+```{r axis-09}
 par(pty="s")

### 2024-07-02 text/14/qqplot.Rmd
Updated 21/ihme-model

@@
 - R software
+- Rmarkdown graphs
 output: html_document
@@
 ---
-
-This program was written by Steve Simon, http://www.pmean.com, and is available at http://www.pmean.com/14/qqplot.html.
-
-See a list of my other R programs at http://www.pmean.com/category/Rsoftware.html
 
@@
 
-```{r read-and-display-data}
+```{r qqplot-01}
 fn <- "http://www.pmean.com/00files/housing.txt"
@@
 
-```{r descriptive-statistics}
+```{r qqplot-02}
 al$age <- as.numeric(al$AgeYears)
@@
 
-```{r normal-probability-plot}
+```{r qqplot-03}
 qqnorm(al$Price)
@@
 
-```{r regression model}
+```{r qqplot-04}
 al.model <- lm(Price~SquareFeet+CustomBuild,data=al)

### 2024-07-02 text/15/difference-of-two-exponentials.Rmd
Updated 21/ihme-model

--- 2023-02-21 text/15/difference-of-two-exponentials.Rmd
@@
 - R software
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r difference-in-two-exponentials-01}
 head(Theoph)
@@
 
-```{r}
+```{r difference-in-two-exponentials-02}
 tail(Theoph)

### 2024-07-02 text/15/missing-values-in-r.Rmd
Updated 21/ihme-model

@@
 
-```{r}
+```{r missing-values-in-r-01}
 sl <- read.table(file="http://www.statsci.org/data/general/sleep.txt",header=TRUE)
@@
 
-```{r}
+```{r missing-values-in-r-02}
 names(sl)
@@
 
-```{r}
+```{r missing-values-in-r-03}
 head(sl)
@@
 
-```{r}
+```{r missing-values-in-r-04}
 tail(sl)
@@
 
-```{r}
+```{r missing-values-in-r-05}
 summary(sl)
@@
 
-```{r}
+```{r missing-values-in-r-06}
 table(table(sl$Species))
@@
 
-```{r}
+```{r missing-values-in-r-07}
 length(unique(sl$Species))
@@
 
-```{r}
+```{r missing-values-in-r-08}
 table(sl$Gestation)
@@
 
-```{r}
+```{r missing-values-in-r-09}
 addmargins(table(sl$Gestation))
@@
 
-```{r}
+```{r missing-values-in-r-10}
 table(sl$Gestation,useNA="always")
@@
 
-```{r}
+```{r missing-values-in-r-11}
 table(sl$Gestation,useNA="ifany")
@@
 
-```{r}
+```{r missing-values-in-r-12}
 table(sl$Danger,useNA="ifany")
@@
 
-```{r}
+```{r missing-values-in-r-13}
 table(sl$Danger,useNA="always")
@@
 
-```{r}
+```{r missing-values-in-r-14}
 mean(sl$Gestation,na.rm=FALSE)
@@
 
-```{r}
+```{r missing-values-in-r-15}
 mean(sl$Gestation,na.rm=TRUE)
@@
 
-```{r}
+```{r missing-values-in-r-16}
 mod1 <- lm(TotalSleep~Gestation,data=sl,na.action=na.omit)
@@
 
-```{r}
+```{r missing-values-in-r-17}
 which(sl$Gestation==NA)
@@
 
-```{r}
+```{r missing-values-in-r-18}
 which(is.na(sl$Gestation))
@@
 
-```{r}
+```{r missing-values-in-r-19}
 which(is.finite(sl$Gestation))
@@
 
-```{r}
+```{r missing-values-in-r-20}
 sl$Gestation==NA
@@
 
-```{r}
+```{r missing-values-in-r-21}
 sl$Gestation==645
@@
 
-```{r}
+```{r missing-values-in-r-22}
 tst.na <- NA
@@
 
-```{r}
+```{r missing-values-in-r-23}
 tst.tvl <- c(TRUE,FALSE,NA)
@@
 
-```{r}
+```{r missing-values-in-r-24}
 and.table <- outer(tst.tvl,tst.tvl,"&")
@@
 
-```{r}
+```{r missing-values-in-r-25}
 or.table <- outer(tst.tvl,tst.tvl,"|")
@@
 
-```{r}
+```{r missing-values-in-r-26}
 sl$Gestation<50
@@
 
-```{r}
+```{r missing-values-in-r-27}
 1/0
@@
 
-```{r}
+```{r missing-values-in-r-28}
 -1/0
@@
 
-```{r}
+```{r missing-values-in-r-29}
 sqrt(-1)
@@
 
-```{r}
+```{r missing-values-in-r-30}
 Inf>3
@@
 
-```{r}
+```{r missing-values-in-r-31}
 NaN>3
@@
 
-```{r}
+```{r missing-values-in-r-32}
 Inf==Inf
@@
 
-```{r}
+```{r missing-values-in-r-33}
 NaN==NaN
@@
 
-```{r}
+```{r missing-values-in-r-34}
 Inf+1
@@
 
-```{r}
+```{r missing-values-in-r-35}
 1/Inf
@@
 
-```{r}
+```{r missing-values-in-r-36}
 0*Inf
@@
 
-```{r}
+```{r missing-values-in-r-37}
 Inf-Inf
@@
 
-```{r}
+```{r missing-values-in-r-38}
 sqrt(as.complex(-1))
@@
 
-```{r}
+```{r missing-values-in-r-39}
 tst.cases <- c(3,NA,NaN,Inf,-Inf)
@@
 
-```{r}
+```{r missing-values-in-r-40}
 is.finite.table <- is.finite(tst.cases)

### 2024-07-02 text/15/relationships-in-r.Rmd
Updated 21/ihme-model

--- 2023-02-21 text/15/relationships-in-r.Rmd
@@
 
-```{r}
+```{r relationships-in-r-01}
 suppressMessages(suppressWarnings(library(tidyverse)))
@@
 
-```{r}
+```{r relationships-in-r-02}
 plot(home$sqft,home$price)
@@
 
-```{r}
+```{r relationships-in-r-03}
 sb <- is.finite(home$age)
@@
 
-```{r}
+```{r relationships-in-r-04}
 cor(home[,c("price","sqft","age")],use="pairwise.complete.obs")
@@
 
-```{r}
+```{r relationships-in-r-05}
 round(cor(home[,c("price","sqft","age")],use="pairwise.complete.obs"),1)
@@
 
-```{r}
+```{r relationships-in-r-06}
 boxplot(home$price~home$features)
@@
 
-```{r}
+```{r relationships-in-r-07}
 boxplot(home$price~home$northeast)
@@
 
-```{r}
+```{r relationships-in-r-08}
 boxplot(home$price~home$custom_build)
@@
 
-```{r}
+```{r relationships-in-r-09}
 boxplot(home$price~home$corner_lot)
@@
 
-```{r}
+```{r relationships-in-r-10}
 plot(home$price,home$northeast)
@@
 
-```{r}
+```{r relationships-in-r-11}
 plot(home$price,home$custom_build)
@@
 
-```{r}
+```{r relationships-in-r-12}
 plot(home$price,home$corner_lot)
@@
 
-```{r}
+```{r relationships-in-r-13}
 round(cor(home[,c("northeast","custom_build","corner_lot")],home$price),1)

### 2024-07-02 text/20/ihme-model.Rmd
Updated 21/ihme-model

--- 2023-01-11 text/20/ihme-model.Rmd
@@
 - Epidemiology
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
 
-```{r setup, echo=FALSE}
+```{r ihme-model-01, echo=FALSE}
 knitr::opts_chunk$set(echo=FALSE)
@@
 
-```{r g1, fig.width=4, fig.height=2}
+```{r ihme-model-02, fig.width=4, fig.height=2}
 data.frame(
@@
 
-```{r g2, fig.width=4, fig.height=2}
+```{r ihme-model-03, fig.width=4, fig.height=2}
 data.frame(
@@
 
-```{r g3, fig.width=4, fig.height=2}
+```{r ihme-model-04, fig.width=4, fig.height=2}
 data.frame(
@@
 Preprint, 2020-04-21.
-Available in [pdf format](http://www.healthdata.org/sites/default/files/files/Projects/COVID/RA_COVID-forecasting-USA-EEA_042120.pdf).
+Available in [pdf format][mur1].
 
@@
 
-Murray CJL.
-Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months.
-Preprint, 2020-03-26.
-Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf).
+Murray CJL. Forecasting COVID-19 impact on hospital bed-days, ICU-days, ventilator-days and deaths by US state in the next 4 months. Preprint, 2020-03-26. Available in [pdf format][mur2].
 
@@
 doi:10.1038/s41597-020-0448-0.
-Available in [html format](https://www.nature.com/articles/s41597-020-0448-0) or [pdf format](https://www.nature.com/articles/s41597-020-0448-0.pdf).
+Available in [html format][xub1] or [pdf format][xub2].
 
@@
 
-IHME COVID-19 model FAQs. Available in [html format](http://www.healthdata.org/covid/faqs).
+IHME COVID-19 model FAQs. Available in [html format][ihm1].
 
@@
 
-Aizenman N.
-5 Key Facts Not Explained In White House COVID-19 Projections.
-National Public Radio 2020-04-01.
-Available in [html format](https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections).
+Aizenman N. 5 Key Facts Not Explained In White House COVID-19 Projections. National Public Radio 2020-04-01. Available in [html format][aiz1].
 
@@
 
-Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M.
-What 5 Coronavirus Models Say the Next Month Will Look Like.
-The New York Times, 2020-04-22.
-Available in [html format](https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html).
+Bui Q, Katz J, Parlapiano A, and Sanger-Katz, M. What 5 Coronavirus Models Say the Next Month Will Look Like. The New York Times, 2020-04-22. Available in [html format][bui1].
 
@@
 
-Chow D.
-What we know about the coronavirus model the White House unveiled.
-NBC News, 2020-04-01.
-Available in [html format](https://www.nbcnews.com/science/science-news/what-we-know-about-coronavirus-model-white-house-unveiled-n1173601).
+Chow D. What we know about the coronavirus model the White House unveiled. NBC News, 2020-04-01. Available in [html format][cho1].
 
@@
 
-Druke N, Silver N.
-How One Modeler Is Trying To Forecast The Toll Of COVID-19.
-FiveThirtyEight 2020-04-22.
-Available in [video format](https://fivethirtyeight.com/features/politics-podcast-how-one-modeler-is-trying-to-forecast-the-toll-of-covid-19/) and as a [podcast](https://podcasts.apple.com/us/podcast/model-talk-forecasting-the-toll-of-covid-19/id1077418457).
+Druke N, Silver N. How One Modeler Is Trying To Forecast The Toll Of COVID-19. FiveThirtyEight 2020-04-22. Available in [video format][dru1] and as a [podcast][dru2].
 
@@
 
-Begley S.
-Influential Covid-19 model uses flawed methods and shouldn?t guide U.S.
-policies, critics say.
-Stat 2020-04-17.
-Available in [html format](https://www.statnews.com/2020/04/17/influential-covid-19-model-uses-flawed-methods-shouldnt-guide-policies-critics-say/).
+Begley S. Influential Covid-19 model uses flawed methods and shouldn?t guide U.S. policies, critics say. Stat 2020-04-17. Available in [html format][beg1].
 
@@
 
-Etzioni R, COVID-19 Models: What Makes Them Tick? Timmerman Report, 2020-04-07. Available in [html format](https://timmermanreport.com/2020/04/covid-19-models-what-makes-them-tick/).
+Etzioni R, COVID-19 Models: What Makes Them Tick? Timmerman Report, 2020-04-07. Available in [html format][etz1].
 
@@
 
-Mortality Modeling FAQ. Available in [html format](https://covid-19.tacc.utexas.edu/mortality-projection-faq/).
+Mortality Modeling FAQ. Available in [html format][mor1].
 
@@
 
-Social Distancing: Data Models for a Model Response to an Outbreak. Available in [html format](https://sph.umich.edu/news/2020posts/data-models-for-a-model-response-to-an-outbreak.html).
+Social Distancing: Data Models for a Model Response to an Outbreak. Available in [html format][soc1].
 
@@
 
-Robinson D. David's COVID Numbers. Available in [html format](https://davidscovidnumbers.wordpress.com/).
+Robinson D. David's COVID Numbers. Available in [html format][rob1].
 
@@
 2020.
-Available in [html format](https://doi.org/10.7326/M20-1565).
+Available in [html format][jew1].
 
@@
 doi:10.1016/j.plrev.2016.07.005.
-Available in [html format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/) or [pdf format](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/pdf/nihms-802148.pdf).
+Available in [html format][cho2] or [pdf format][cho3].
 
@@
 Science 2020-04-14.
-DOI: 10.1126/science.abb5793, Available in [html format](https://science.sciencemag.org/content/early/2020/04/24/science.abb5793).
+DOI: 10.1126/science.abb5793, Available in [html format][kis1].
 
@@
 
-FRED Modeling System. Available in [html format](https://fred.publichealth.pitt.edu/).
+FRED Modeling System. Available in [html format][fre1].
 
@@
 Preprint published in Medrxiv, 2020-04-11.
-Available in [pdf format](https://www.medrxiv.org/content/10.1101/2020.04.08.20058214v1.full.pdf).
+Available in [pdf format][aya1].
 
@@
 
-[US Predictions](https://mrc-ide.github.io/covid19usa/#/)
+[US Predictions][usp1]
 
@@
 Preprint published in Arxiv, 2020-04-30.
-Available in [pdf format](https://arxiv.org/pdf/2004.14103.pdf).
+Available in [pdf format][wan1].
 
-+ This article fits an SEIR model to individual counties in the United States with a large number of county specific covariates.
-The model produces a [nice dashboard](https://covid19.stat.iastate.edu/).
++ This article fits an SEIR model to individual counties in the United States with a large number of county specific covariates. The model produces a [nice dashboard][ias1].
 
@@
 Thanks to Chris Barker and Ruth Etzioni for helpful suggestions.
+
+[aiz1]: https://www.npr.org/sections/health-shots/2020/04/01/824744490/5-key-facts-the-white-house-isnt-saying-about-their-covid-19-projections
+[aya1]: https://www.medrxiv.org/content/10.1101/2020.04.08.20058214v1.full.pdf
+[beg1]: https://www.statnews.com/2020/04/17/influential-covid-19-model-uses-flawed-methods-shouldnt-guide-policies-critics-say/
+[bui1]: https://www.nytimes.com/interactive/2020/04/22/upshot/coronavirus-models.html
+[cho1]: https://www.nbcnews.com/science/science-news/what-we-know-about-coronavirus-model-white-house-unveiled-n1173601
+[cho2]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/
+[cho3]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5348083/pdf/nihms-802148.pdf
+[dru1]: https://fivethirtyeight.com/features/politics-podcast-how-one-modeler-is-trying-to-forecast-the-toll-of-covid-19/
+[dru2]: https://podcasts.apple.com/us/podcast/model-talk-forecasting-the-toll-of-covid-19/id1077418457
+[etz1]: https://timmermanreport.com/2020/04/covid-19-models-what-makes-them-tick/
+[fre1]: https://fred.publichealth.pitt.edu/
+[ias1]: https://covid19.stat.iastate.edu/
+[ihm1]: http://www.healthdata.org/covid/faqs
+[jew1]: https://doi.org/10.7326/M20-1565
+[kis1]: https://science.sciencemag.org/content/early/2020/04/24/science.abb5793
+[mor1]: https://covid-19.tacc.utexas.edu/mortality-projection-faq/
+[mur1]: http://www.healthdata.org/sites/default/files/files/Projects/COVID/RA_COVID-forecasting-USA-EEA_042120.pdf
+[mur2]: https://www.medrxiv.org/content/10.1101/2020.03.27.20043752v1.full.pdf
+[rob1]: https://davidscovidnumbers.wordpress.com/
+[soc1]: https://sph.umich.edu/news/2020posts/data-models-for-a-model-response-to-an-outbreak.html
+[usp1]: https://mrc-ide.github.io/covid19usa/#/
+[wan1]: https://arxiv.org/pdf/2004.14103.pdf
+[xub1]: https://www.nature.com/articles/s41597-020-0448-0
+[xub2]: https://www.nature.com/articles/s41597-020-0448-0.pdf

### 2024-07-03 text/24/different-end-times.Rmd
Moved pending files

---
title: Different end times in survival data
source: "New"
author: Steve Simon
date: 2024-05-28
categories:
- Blog post
tags:
- Rmarkdown code
- Survival analysis
output: html_document
---

Dear Professor Mean, I am running  a survival analysis with two groups and one group was followed for 190 days and the other for 225 days. Would that cause any bias?

<!---more--->

Dear Reader,

Your concern, clearly is that some events might have occurred in the first group between day 191 and day 225. You are smart to be careful about this. It's easiest to look at a simulation, and if you want to delve into the math, everything comes out clean. There is no bias caused by the differential follow-up times.

## Some simple simulated survival times

Here's a simple simulation using R.

Generate two sets of survival times. The rate is twice as large in the second group, so you should expect to see events occur earlier in the second group and events below a certain threshold should occur more often in the second group.

```{r}
suppressMessages(library(survival))
suppressMessages(library(tidyverse))
set.seed(190225)
raw_time1 <- rexp(100, rate=0.004)
raw_time2 <- rexp(100, rate=0.008)
sim <- data.frame(
  g=rep(c("a", "b"), each=100), 
  raw_time=c(raw_time1, raw_time2))
head(sim)
tail(sim)
```

How many events occur from 0 to 190 days, from 191 to 225 days, and after 225 days?

```{r}
sim %>%
  group_by(g) %>%
  summarize(
    n_0_to_190=sum(raw_time <= 190),
    n_191_to_225=sum(raw_time > 190&raw_time<=225),
    n_above_225=sum(raw_time > 225)) -> sim_counts
sim_counts
```

There are indeed more events at a threshold of 190 or below in group B (`r sim_counts$n_0_to_190[2]`) than there are in group A (`r sim_counts$n_0_to_190[1]`). There are, as well, more events at a threshold of 225 and below in group B (`r 100-sim_counts$n_above_225[2]`) than there are in group A  (`r 100-sim_counts$n_above_225[1]`).

There are `r sim_counts$n_191_to_225[1]+sim_counts$n_191_to_225[1]` events that are in the "Twilight Zone". This means that `r sim_counts$n_191_to_225[1]` of the Twilight Zone events in group A are censored, while `r sim_counts$n_191_to_225[2]` of the Twilight Zone events in group B are still counted as events.

Does this bias things? The way to find out is to run the data with different censoring thresholds (190 for group A and 225 for group B) then run it with a common threshold of 190 for both groups.

## Analysis with censoring thresholds of 190 (A) and 225 (B)

```{r}
censor_time_v1 <- rep(c(190, 225), each=100)
sim %>%
  mutate(t_v1=pmin(raw_time, censor_time_v1)) %>%
  mutate(censor_v1=as.numeric(t_v1 < censor_time_v1)) -> sim_v1
sim_v1 %>%
  count(g, censor_v1)
```

Run a Cox regression model

```{r}
cox_v1 <- coxph(Surv(t_v1, censor_v1)~g, data=sim_v1)
cox_v1
```

## Analysis with censoring thresholds of 190 for both A and B

What if you set the censoring times to 190 days for both groups? That effectively converts `r sim_counts$n_191_to_225[2]` patients in the second group from events to observations censored at 190 days.

```{r}
censor_time_v2 <- rep(c(190, 190), each=100)
sim %>%
  mutate(t_v2=pmin(raw_time, censor_time_v2)) %>%
  mutate(censor_v2=as.numeric(t_v2 < censor_time_v2)) -> sim_v2
sim_v2 %>%
  count(g, censor_v2)
```

Run a Cox regression model

```{r}
cox_v2 <- coxph(Surv(t_v2, censor_v2)~g, data=sim_v2)
cox_v2
```

Wow! The results are identical. What this says is that there is no information in any events in group B that occur after the last event in group A.

The moral of the story is that if you have a different censoring time for two groups and observe some events in the second group that are larger than the largest event time in the first group, it doesn't bias things because converting these from events to censored observations does not affect the hazard ratio.

## Looking at the partial likelihood

The Cox regression model uses a partial likelihood which looks rather ugly.

$l_p(\beta)=\Pi\frac{e^{X_i \beta}}{\Sigma\ e^{X_j \beta}}$

where the product is evaluated only at event times (not censored times). The sum in the denominator is the sum over the patients still at risk of the event. For events in group B that occur after the maximum event time of group A, the only patients at risk are those in the second group. So then the $e^{X_j \beta}$ terms in the denominator are all the same and all equal to the $e^{X_i \beta}$ term in the numerator. Cancel out the betas and you are left with a multiplicative constant for any Twilight Zone events in group B.

So the estimate you get by maximizing the partial likelihood over all events is the same as the estimate you get by maximizing the partial likelihood ignoring the events in group B that occur after the maximum event time in group A.

You can make a similar argument about the log rank test. This test can be recast as a function of observed deaths in a group minus the expected deaths in that group. After the last event in group A, the observed deaths in group B have to equal the expected deaths. So those Twilight Zone events in group B contribute exactly zero to the observed minus expected calculations.

### 2024-07-03 text/24/nice-layout.Rmd
Moved pending files

--- 2024-06-27 text/pending/nice-layout.Rmd
@@
 Hard, but not impossible.
-Here is a simple example.Here is a brief abstract.
+Here is a simple example.
 
@@
 library(kableExtra)
-tab07 %>%
+tab06 %>%
+  set_names(c(" ", "Deaths", "Survivors")) %>%
+  kbl %>%
+  kable_paper(
+    "striped", 
+    full_width=FALSE) %>%
   pack_rows("Sex", 1, 2) -> tab08
@@
 
+Here is what the code would look like if you used sprintf instead of glue.
+
 ```{r}

### 2024-07-06 text/20/non-breaking-spaces.Rmd
Updated rmd files

--- 2023-01-10 text/20/non-breaking-spaces.Rmd
@@
 - Data management
+- Rmarkdown code
 output: html_document
+page_update: partial
 ---
@@
 
-```{r insert}
+```{r  non-breaking-spaces-01}
 tst1 <- "?"
@@
 
-```{r chartoraw}
+```{r  non-breaking-spaces-02}
 charToRaw(tst1)
@@
 
-```{r search}
+```{r  non-breaking-spaces-03}
 tst1=="\xa0"

### 2024-07-06 text/20/poor-color-choice.Rmd
Updated rmd files

--- 2023-10-10 text/20/poor-color-choice.Rmd
@@
 - Graphical display
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r setup, echo=FALSE}
+```{r poor-color-choice-01, echo=FALSE}
 knitr::opts_chunk$set(echo=FALSE, fig.width=5, fig.height=5)
@@
 
-```{r color-circle1, echo=FALSE, fig.width=5, fig.height=5}
+```{r poor-color-choice-02, echo=FALSE, fig.width=5, fig.height=5}
 suppressMessages(suppressWarnings(library(ggplot2)))
@@
 
-```{r color-circle2, echo=FALSE, fig.width=5, fig.height=5}
+```{r poor-color-choice-03, echo=FALSE, fig.width=5, fig.height=5}
 ggplot(df0, aes(x, y)) +
@@
 
-```{r color-circle3, echo=FALSE, fig.width=5, fig.height=5}
+```{r poor-color-choice-04, echo=FALSE, fig.width=5, fig.height=5}
 ggplot(df0, aes(x, y)) +
@@
 
-```{r brew1, fig.height=1}
+```{r poor-color-choice-05, fig.height=1}
 par(mar=rep(0, 4))

### 2024-07-06 text/20/reading-text-files.Rmd
Updated rmd files

--- 2023-01-11 text/20/reading-text-files.Rmd
@@
 
-```{r ascii, echo=FALSE}
+```{r reading-text-files-01, echo=FALSE}
 symbol_codes <- c(33:126)

### 2024-07-06 text/20/rmarkdown-papers.Rmd
Updated rmd files

--- 2023-01-11 text/20/rmarkdown-papers.Rmd
@@
 tags:
+- R programming
+- Rmarkdown graphics
 - Writing research papers
 output: html_document
+page_update: complete
 ---
@@
 
-```{r, echo=FALSE, fig.width=6, fig.height=3}
+```{r rmarkdown-papers-01, echo=FALSE, fig.width=6, fig.height=3}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, echo=FALSE}
+```{r rmarkdown-papers-02, echo=FALSE}
 sz <- 1.2
@@
 
-```{r}
+```{r rmarkdown-papers-03}
 lm(breaks ~ wool*tension, data = warpbreaks) %>%

### 2024-07-06 text/20/smart-quotes.Rmd
Updated rmd files

--- 2023-01-11 text/20/smart-quotes.Rmd
@@
 - Data management
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---
@@
 
-```{r setup, echo=FALSE}
+```{r smart-quotes-01, echo=FALSE}
 knitr::opts_chunk$set(echo = FALSE, fig.width=4.5, fig.height=0.5)
@@
 
-```{r straight-double-quotes}
+```{r smart-quotes-02}
 cplot('"')
@@
 
-```{r straight-single-quotes}
+```{r smart-quotes-03}
 cplot("'")
@@
 
-```{r backtick}
+```{r smart-quotes-04}
 cplot("`")
@@
 
-```{r assign-double-quotes-1}
+```{r smart-quotes-05}
 cplot("x  < -  ' \" '")
@@
 
-```{r assign-double-quotes-2}
+```{r smart-quotes-06}
 cplot('x  < -  \" \\ "\ "')
@@
 
-```{r charToRaw-1}
+```{r smart-quotes-07}
 cplot("charToRaw(' \" ')")
@@
 
-```{r charToRaw-2}
+```{r smart-quotes-08}
 cplot(charToRaw('"'))
@@
 
-```{r charToRaw-3}
+```{r smart-quotes-09}
 cplot("charToRaw(\" ' \")")
@@
 
-```{r charToRaw-4}
+```{r smart-quotes-10}
 cplot(charToRaw("'"))
@@
 
-```{r charToRaw-5}
+```{r smart-quotes-11}
 cplot("charToRaw(\" ` \")")
@@
 
-```{r charToRaw-6}
+```{r smart-quotes-12}
 cplot(charToRaw("`"))
@@
 
-```{r convert-hex-1}
+```{r smart-quotes-13}
 cplot('"\\x22"')
@@
 
-```{r convert-hex-2}
+```{r smart-quotes-14}
 cplot("\x22")
@@
 
-```{r left-double-quote}
+```{r smart-quotes-15}
 cplot("\U201C")
@@
 
-```{r right-double-quote}
+```{r smart-quotes-16}
 cplot("\U201D")
@@
 
-```{r left-single-quote}
+```{r smart-quotes-17}
 cplot("\U2018")
@@
 
-```{r right-single-quote}
+```{r smart-quotes-18}
 cplot("\U2019")
@@
 
-```{r rawToChar-8}
+```{r smart-quotes-19}
 cplot('charToRaw( \" \U201C \" )')
@@
 
-```{r rawToChar-9}
+```{r smart-quotes-20}
 x <- str_c(as.character(charToRaw("\U201C")), collapse=" ")
@@
 
-```{r accented}
+```{r smart-quotes-21}
 cplot("\U00E9")
@@
 
-```{r cedilla}
+```{r smart-quotes-22}
 cplot("\U00C7")
@@
 
-```{r tilde}
+```{r smart-quotes-23}
 cplot("\U00F1")
@@
 
-```{r sharp-S}
+```{r smart-quotes-24}
 cplot("\U00DF")
@@
 
-```{r thorn}
+```{r smart-quotes-25}
 cplot("\U00FE")
@@
 
-```{r greek}
+```{r smart-quotes-26}
 cplot("\U03B1 \U03B2 \U03B3")
@@
 
-```{r arabic}
+```{r smart-quotes-27}
 cplot("\U062A\U0633\U062C\U0651\U0644")
@@
 
-```{r chinese}
+```{r smart-quotes-28}
 cplot("\U4E2D\U6587")
@@
 
-```{r emojis}
+```{r smart-quotes-29}
 cplot("\U1F642\U1F622")
@@
 
-```{r unicode-1}
+```{r smart-quotes-30}
 cplot("\"\\U201C\"")
@@
 
-```{r unicode-2}
+```{r smart-quotes-31}
 cplot("\U201C")
@@
 
-```{r unicode-3}
+```{r smart-quotes-32}
 cplot("\"\\U201D\"")
@@
 
-```{r unicode-4}
+```{r smart-quotes-33}
 cplot("\U201D")
@@
 
-```{r unicode-5}
+```{r smart-quotes-34}
 cplot("\"\\U2018\"")
@@
 
-```{r unicode-6}
+```{r smart-quotes-35}
 cplot("\U2018")
@@
 
-```{r unicode-7}
+```{r smart-quotes-36}
 cplot("\"\\U2019\"")
@@
 
-```{r unicode-8}
+```{r smart-quotes-37}
 cplot("\U2019")
@@
 
-```{r minus-1}
+```{r smart-quotes-38}
 cplot("-")
@@
 
-```{r em-dash-1}
+```{r smart-quotes-39}
 cplot("\"\\U2014\"")
@@
 
-```{r em-dash-2}
+```{r smart-quotes-40}
 cplot("\U2014")
@@
 
-```{r en-dash-1}
+```{r smart-quotes-41}
 cplot("\"\\U2013\"")
@@
 
-```{r en-dash-2}
+```{r smart-quotes-42}
 cplot("\U2013")
@@
 
-There's a [nice web page](https://www.cl.cam.ac.uk/~mgk25/ucs/quotes.html) about the historical developments of computer codes for quote marks and dashes and [another page](https://cran.r-project.org/web/packages/utf8/vignettes/utf8.html) that talks about computer codes in general from the perspective of an R programmer.
+There's a [nice web page][kuh1] about the historical developments of computer codes for quote marks and dashes and [another page][utf8] that talks about computer codes in general from the perspective of an R programmer.
 
-There are some other variants, such as the prime symbols, described in [this Wikipedia page](https://en.wikipedia.org/wiki/Prime_(symbol)).
+There are some other variants, such as the prime symbols, described in [this Wikipedia page][wik1].
+
+[kuh1]: https://www.cl.cam.ac.uk/~mgk25/ucs/quotes.html
+[utf8]: https://cran.r-project.org/web/packages/utf8/vignettes/utf8.html
+[wik1]: https://en.wikipedia.org/wiki/Prime_(symbol)

### 2024-07-06 text/21/anonymizing-ids.Rmd
Updated rmd files

--- 2023-01-11 text/21/anonymizing-ids.Rmd
@@
 - Privacy in research
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/blogdown-first-steps.Rmd
Updated rmd files

--- 2023-01-10 text/21/blogdown-first-steps.Rmd
@@
 - Website details
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/cubic-splines.Rmd
Updated rmd files

--- 2021-11-27 text/21/cubic-splines.Rmd
@@
 - Linear regression
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/dark-side.Rmd
Updated rmd files

--- 2023-01-11 text/21/dark-side.Rmd
@@
 - Incomplete page
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/direct-standardization.Rmd
Updated rmd files

--- 2023-01-10 text/21/direct-standardization.Rmd
@@
 - Epidemiology
+- Incomplete pages
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/glm-deviance.Rmd
Updated rmd files

--- 2023-02-19 text/21/glm-deviance.Rmd
@@
 - Generalized linear model
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/hurricane-frequencies.Rmd
Updated rmd files

--- 2023-01-11 text/21/hurricane-frequencies.Rmd
@@
 
-```{r setup, echo=FALSE}
+```{r hurricane-frequencies-01, echo=FALSE}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r read1}
+```{r hurricane-frequencies-02}
 fn <- "https://dasl.datadescription.com/download/data/3279"
@@
 
-```{r read2}
+```{r hurricane-frequencies-03}
 raw_data[9:16, ]
@@
 
-```{r plot}
+```{r hurricane-frequencies-04}
 df <- data.frame(x=seq(1855, 1995, by=10), y=raw_data$Count[1:15])
@@
 
-```{r poly1}
+```{r hurricane-frequencies-05}
 add_trend <- function(point_plot, df, degree) {
@@
 
-```{r poly2}
+```{r hurricane-frequencies-06}
 add_trend(point_plot, df, 2)
@@
 
-```{r poly3}
+```{r hurricane-frequencies-07}
 add_trend(point_plot, df, 3)
@@
 
-```{r poly4}
+```{r hurricane-frequencies-08}
 add_trend(point_plot, df, 4)
@@
 
-```{r poly5}
+```{r hurricane-frequencies-09}
 add_trend(point_plot, df, 5)
@@
 
-```{r poly6}
+```{r hurricane-frequencies-10}
 add_trend(point_plot, df, 6)
@@
 
-```{r poly7}
+```{r hurricane-frequencies-11}
 add_trend(point_plot, df, 7)
@@
 
-```{r poly8}
+```{r hurricane-frequencies-12}
 add_trend(point_plot, df, 8)
@@
 
-```{r poly9}
+```{r hurricane-frequencies-13}
 add_trend(point_plot, df, 9)
@@
 
-```{r poly10}
+```{r hurricane-frequencies-14}
 add_trend(point_plot, df, 10)
@@
 
-```{r poly11}
+```{r hurricane-frequencies-15}
 add_trend(point_plot, df, 11)
@@
 
-```{r poly12}
+```{r hurricane-frequencies-16}
 add_trend(point_plot, df, 12)
@@
 
-```{r poly13}
+```{r hurricane-frequencies-17}
 add_trend(point_plot, df, 13)
@@
 
-```{r poly14}
+```{r hurricane-frequencies-18}
 add_trend(point_plot, df, 14)
@@
 
-```{r reset}
+```{r hurricane-frequencies-19}
 knitr::opts_chunk$set(echo=TRUE)
@@
 
-```{r setup}
+```{r hurricane-frequencies-01}
 ```
@@
 
-```{r read1}
+```{r hurricane-frequencies-02}
 ```
 
-```{r read2}
+```{r hurricane-frequencies-03}
 ```
@@
 
-```{r plot}
+```{r hurricane-frequencies-04}
 ```
@@
 
-```{r poly1}
+```{r hurricane-frequencies-05}
 ```
@@
 
-```{r poly2}
+```{r hurricane-frequencies-06}
 ```
 
-```{r poly3}
+```{r hurricane-frequencies-07}
 ```
 
-```{r poly4}
+```{r hurricane-frequencies-08}
 ```
 
-```{r poly5}
+```{r hurricane-frequencies-09}
 ```
 
-```{r poly6}
+```{r hurricane-frequencies-10}
 ```
 
-```{r poly7}
+```{r hurricane-frequencies-11}
 ```
 
-```{r poly8}
+```{r hurricane-frequencies-12}
 ```
 
-```{r poly9}
+```{r hurricane-frequencies-13}
 ```
 
-```{r poly10}
+```{r hurricane-frequencies-14}
 ```
 
-```{r poly11}
+```{r hurricane-frequencies-15}
 ```
 
-```{r poly12}
+```{r hurricane-frequencies-16}
 ```
 
-```{r poly13}
+```{r hurricane-frequencies-17}
 ```
 
-```{r poly14}
+```{r hurricane-frequencies-18}
 ```

### 2024-07-06 text/21/negative-binomial-regression.Rmd
Updated rmd files

--- 2023-02-19 text/21/negative-binomial-regression.Rmd
@@
 
-```{r setup}
+```{r negative-binomial-regression-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r bar-charts-n}
+```{r negative-binomial-regression-02}
 data.frame(x=0:25) %>%
@@
 
-```{r bar-charts-p}
+```{r negative-binomial-regression-03}
 data.frame(x=0:25) %>%
@@
 
-```{r}
+```{r negative-binomial-regression-01}
 library(MASS)

### 2024-07-06 text/21/poisson-inferences.Rmd
Updated rmd files

--- 2023-01-12 text/21/poisson-inferences.Rmd
@@
 - Probability distributions
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/21/read-acs-data.Rmd
Updated rmd files

--- 2023-01-11 text/21/read-acs-data.Rmd
@@
 tags:
+- Rmarkdown graphs
 - Website details
 output: html_document
+page_update: complete
 ---

### 2024-07-06 text/22/testing-d3.Rmd
Updated rmd files

--- 2023-01-11 text/22/testing-d3.Rmd
@@
 
-```{r setup}
+```{r testing-d3-01}
 library(r2d3)
@@
 
-```{d3, data=runif(30), options=list(color='steelblue')}
+```{d3 testing-d3-02, data=runif(30), options=list(color='steelblue')}
 svg.selectAll('rect')

### 2024-07-06 text/21/anonymizing-ids.Rmd
Updated 2022 Rmd files

@@
 
-```{r, echo=FALSE}
+```{r anonymizing-ids-01, echo=FALSE}
 suppressMessages(suppressWarnings(library(dplyr)))
@@
 
-```{r, echo=FALSE}
+```{r anonymizing-ids-02, echo=FALSE}
 e <- data.frame(private_id=as.character(private_id), random_id=as.character(private_id+480035489))
@@
 
-```{r, echo=FALSE}
+```{r anonymizing-ids-03, echo=FALSE}
 d <- data.frame(
@@
 
-```{r, echo=FALSE}
+```{r anonymizing-ids-04, echo=FALSE}
 e %>%
@@
 
-```{r, echo=FALSE}
+```{r anonymizing-ids-05, echo=FALSE}
 h %>%

### 2024-07-06 text/21/cubic-splines.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup}
+```{r cubic-splines-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r}
+```{r cubic-splines-02}
 library(ggplot2)
@@
 
-```{r}
+```{r cubic-splines-03}
 x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
@@
 
-```{r}
+```{r cubic-splines-04}
 xm <- data.frame(
@@
 
-```{r}
+```{r cubic-splines-05}
 kable(xm[1:21, 1:16])
@@
 
-```{r}
+```{r cubic-splines-06}
 display_curve <- function(dat, lb) {
@@
 
-```{r}
+```{r cubic-splines-07}
 
@@
 
-```{r}
+```{r cubic-splines-08}
 lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
@@
 
-```{r}
+```{r cubic-splines-09}
 lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
@@
 
-```{r}
+```{r cubic-splines-10}
 lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
@@
 
-```{r}
+```{r cubic-splines-11}
 lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
@@
 
-```{r}
+```{r cubic-splines-12}
 round(cor(xm[c(2:4, 8, 12, 16)]), 2)
@@
 
-```{r}
+```{r cubic-splines-13}
 cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
@@
 
-```{r}
+```{r cubic-splines-14}
 for (j in 1:dim(cubic_spline)[2]) {
@@
 
-```{r}
+```{r cubic-splines-15}
 round(cor(cubic_spline), 1)
@@
 
-```{r}
+```{r cubic-splines-16}
 lm(xm$y~cubic_spline) %>%
@@
 
-```{r}
+```{r cubic-splines-17}
 linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
@@
 
-```{r}
+```{r cubic-splines-18}
 for (j in 1:dim(linear_spline)[2]) {
@@
 
-```{r}
+```{r cubic-splines-19}
 lm(xm$y~linear_spline) %>%
@@
 
-```{r}
+```{r cubic-splines-20}
 quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
@@
 
-```{r}
+```{r cubic-splines-21}
 for (j in 1:dim(quadratic_spline)[2]) {
@@
 
-```{r}
+```{r cubic-splines-22}
 lm(xm$y~quadratic_spline) %>%
@@
 
-```{r}
+```{r cubic-splines-23}
 natural_spline <- ns(xm$c2, knots=c(5, 10, 15), intercept=TRUE)
@@
 
-```{r}
+```{r cubic-splines-24}
 extended_x <- seq(-5, 25, by=1/64)
@@
 
-```{r}
+```{r cubic-splines-25}
 extended_y <- rep(NA, length(extended_x))
@@
 
-```{r}
+```{r cubic-splines-26}
 for (k in 3:7) {
@@
 
-```{r}
+```{r cubic-splines-27}
 for (k in 3:7) {

### 2024-07-06 text/21/dark-side.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup, echo=FALSE}
+```{r dark-side-01, echo=FALSE}
 suppressMessages(suppressWarnings(library(dplyr)))
@@
 
-```{r, fig.width=5, fig.height=5, echo=FALSE}
+```{r dark-side-02, fig.width=5, fig.height=5, echo=FALSE}
 o <- 0.6
@@
 
-```{r, fig.width=5, fig.height=5, echo=FALSE}
+```{r dark-side-02, fig.width=5, fig.height=5, echo=FALSE}
 o <- 0.88

### 2024-07-06 text/21/direct-standardization.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup}
+```{r direct-standardization-01}
 suppressMessages(suppressWarnings(library(broom)))

### 2024-07-06 text/21/glm-deviance.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup}
+```{r glm-deviance-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, eval=FALSE}
+```{r glm-deviance-02, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r glm-deviance-03}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r glm-deviance-04}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r glm-deviance-05}
 pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
@@
 
-```{r}
+```{r glm-deviance-06}
 ggplot(rd, aes(r)) +
@@
 
-```{r}
+```{r glm-deviance-07}
 ggplot(rd, aes(p, r)) +
@@
 
-```{r}
+```{r glm-deviance-08}
 ggplot(rd, aes(fem, r)) +
@@
 
-```{r}
+```{r glm-deviance-09}
 ggplot(rd, aes(ment, r)) +
@@
 
-```{r}
+```{r glm-deviance-10}
 ggplot(rd, aes(phd, r)) +
@@
 
-```{r}
+```{r glm-deviance-11}
 ggplot(rd, aes(mar, r)) +
@@
 
-```{r}
+```{r glm-deviance-12}
 ggplot(rd, aes(kid5, r)) +

### 2024-07-06 text/21/glm-diagnostics.Rmd
Updated 2022 Rmd files

--- 2023-02-19 text/21/glm-diagnostics.Rmd
@@
 - Generalized linear model
+- Incomplete page
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r setup}
+```{r glm-diagnostics-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, eval=FALSE}
+```{r glm-diagnostics-02, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r glm-diagnostics-03}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r glm-diagnostics-04}
 ggplot(articles, aes(art)) + 
@@
 
-```{r}
+```{r glm-diagnostics-05}
 data.frame(x=0:max(articles$art)) %>%
@@
 
-```{r}
+```{r glm-diagnostics-06}
 articles %>%
@@
 
-```{r}
+```{r glm-diagnostics-07}
 quantile(articles$ment)
@@
 
-```{r}
+```{r glm-diagnostics-08}
 quantile(articles$phd)
@@
 
-```{r}
+```{r glm-diagnostics-09}
 articles %>%
@@
 
-```{r}
+```{rglm-diagnostics-10}
 articles %>%
@@
 
-```{r}
+```{r glm-diagnostics-11}
 articles %>%
@@
 
-```{r}
+```{r glm-diagnostics-12}
 ggplot(articles, aes(factor(fem), art)) +
@@
 
-```{r}
+```{r glm-diagnostics-13}
 ggplot(articles, aes(ment, art)) +
@@
 
-```{r}
+```{r glm-diagnostics-14}
 ggplot(articles, aes(phd, art)) +
@@
 
-```{r}
+```{r glm-diagnostics-15}
 ggplot(articles, aes(factor(mar), art)) +
@@
 
-```{r}
+```{r glm-diagnostics-16}
 ggplot(articles, aes(factor(kid5), art)) +
@@
 
-```{r}
+```{r glm-diagnostics-17}
 pois_fem <- glm(art ~ fem, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-18}
 pois_ment <- glm(art ~ ment, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-19}
 pois_phd <- glm(art ~ phd, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-20}
 pois_mar <- glm(art ~ mar, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-21}
 pois_kid5 <- glm(art ~ kid5, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-22}
 pois_all_ivs <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles_scaled)
@@
 
-```{r}
+```{r glm-diagnostics-23}
 tidy(pois_fem) %>%
@@
 
-```{r}
+```{r glm-diagnostics-24}
 r_pois <- augment(pois_all_ivs, articles_scaled, type.predict="response", type.residuals="deviance")
@@
 
-```{r}
+```{r glm-diagnostics-26}
 r_pois %>%
@@
 
-```{r}
+```{r glm-diagnostics-27}
 ggplot(r_pois, aes(factor(fem), .resid)) + 
@@
 
-```{r}
+```{r glm-diagnostics-28}
 ggplot(r_pois, aes(ment, .resid)) + 
@@
 
-```{r}
+```{r glm-diagnostics-29}
 ggplot(r_pois, aes(phd, .resid)) + 
@@
 
-```{r}
+```{r glm-diagnostics-30}
 ggplot(r_pois, aes(factor(mar), .resid)) + 
@@
 
-```{r}
+```{r glm-diagnostics-31}
 ggplot(r_pois, aes(factor(kid5), .resid)) + 
@@
 
-```{r}
+```{r glm-diagnostics-32}
 articles %>%
@@
 
-```{r}
+```{r glm-diagnostics-33}
 library(MASS)
@@
 
-```{r}
+```{r glm-diagnostics-34}
 zip1 <- 
@@
 
-```{r}
+```{r glm-diagnostics-35}
 zip2 <- 
@@
 
-```{r}
+```{r glm-diagnostics-36}
 zip3 <- 
@@
 
-```{r}
+```{r glm-diagnostics-37}
 zinb <- 
@@
 
-```{r}
+```{r glm-diagnostics-99}
 save.image(file="glm_diagnostics.RData")

### 2024-07-06 text/21/poisson-inferences.Rmd
Updated 2022 Rmd files

--- 2023-01-12 text/21/poisson-inferences.Rmd
@@
 tags:
+- Incomplete page
 - Probability distributions
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r setup}
+```{r poisson-inferences-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r bar-charts}
+```{r  poisson-inferences-02}
 data.frame(x=0:25) %>%
@@
 
-```{r}
+```{r poisson-inferences-03}
 rate_per_thousand <- rate*1000
@@
 
-```{r}
+```{r poisson-inferences-04}
 count_ci <- c(x-1.96*sqrt(x), x+1.96*sqrt(x))
@@
 
-```{r}
+```{r poisson-inferences-05}
 rate_ci <- count_ci * 1000 / n
@@
 
-```{r}
+```{r poisson-inferences-06}
 round(rate_ci, 1)
@@
 
-```{r}
+```{r poisson-inferences-07}
 log_count <- log(x)
@@
 
-```{r}
+```{r poisson-inferences-08}
 back_transformed_ci <- exp(log_ci)
@@
 
-```{r}
+```{r poisson-inferences-09}
 rate_ci <- back_transformed_ci * 1000 / n
@@
 
-```{r}
+```{r poisson-inferences-10}
 round(rate_ci, 1)
@@
 
-```{r}
+```{r poisson-inferences-11}
 ln_ratio <- log(r1/r2)
@@
 
-```{r}
+```{r poisson-inferences-12}
 ci2 <- exp(ci1)
@@
 
-```{r}
+```{r poisson-inferences-13}
 injuries <- data.frame(x=c(x1, x2), T=c(T1, T2), g=1:0)

### 2024-07-06 text/21/poisson-regression.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup}
+```{r poisson-regression-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, eval=FALSE}
+```{r poisson-regression-02, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r poisson-regression-03}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r poisson-regression-04}
 pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
@@
 
-```{r}
+```{r poisson-regression-05}
 pois1 %>%
@@
 
-```{r}
+```{r poisson-regression-06}
 log_coefficients %>%
@@
 
-```{r}
+```{r poisson-regression-07}
 fn <- "http://www.statsci.org/data/general/motorins.txt"
@@
 
-```{r}
+```{r poisson-regression-08}
 pois2 <- 
@@
 
-```{r}
+```{r poisson-regression-09}
 pois2 %>%
@@
 
-```{r}
+```{r poisson-regression-10}
 mean(articles$art)
@@
 
-```{r}
+```{r poisson-regression-11}
 glance(pois1)

### 2024-07-06 text/21/quasi-likelihood.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup}
+```{r quasi-likelihood-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, eval=FALSE}
+```{r quasi-likelihood-02, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r quasi-likelihood-03}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r quasi-likelihood-04}
 pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
@@
 
-```{r}
+```{r quasi-likelihood-05}
 pois1 %>%
@@
 
-```{r}
+```{r quasi-likelihood-06}
 quasi1 <- glm(art ~ fem + ment + phd + mar + kid5, family=quasipoisson, data = articles)
@@
 
-```{r}
+```{r quasi-likelihood-07}
 quasi1 %>%

### 2024-07-06 text/21/read-acs-data.Rmd
Updated 2022 Rmd files

--- 2023-01-11 text/21/read-acs-data.Rmd
@@
 tags:
-- Website details
+- Incomplete page
+- Rmarkdown graphs
+- Secondary data
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r read-acs-data-01}
 library(tidycensus)

### 2024-07-06 text/21/venn-diagram.Rmd
Updated 2022 Rmd files

@@
 
-```{r setup, echo=FALSE}
+```{r venn-diagrams-01, echo=FALSE}
 suppressMessages(suppressWarnings(library(dplyr)))
@@
 
-```{r, fig.width=5, fig.height=5, echo=FALSE}
+```{r venn-diagrams-02, fig.width=5, fig.height=5, echo=FALSE}
 o <- 0.6
@@
 
-```{r, fig.width=5, fig.height=5, echo=FALSE}
+```{r venn-diagrams-03, fig.width=5, fig.height=5, echo=FALSE}
 o <- 0.88

### 2024-07-06 text/21/very-simple-splines.Rmd
Updated 2022 Rmd files

@@
 
-```{r}
+```{r very-simple-splines-01}
 library(ggplot2)
@@
 
-```{r}
+```{r very-simple-splines-02}
 linear_spline_ivs <- data.frame(x0=rep(1, 20), x1=c(rep(0, 15), 1:5))
@@
 
-```{r}
+```{r very-simple-splines-03}
 simple_spline_data <- 
@@
 
-```{r}
+```{r very-simple-splines-04}
 x_extra <- seq(0, 20, length=1000)
@@
 
-```{r}
+```{r very-simple-splines-05}
 linear_spline_predictions <- augment(linear_spline_model, newdata=enhanced_data)
@@
 
-```{r}
+```{r very-simple-splines-06}
 simple_spline_data %>%
@@
 
-```{r}
+```{r very-simple-splines-07}
 x <- seq(0, 1, length=100)
@@
 
-```{r}
+```{r very-simple-splines-08}
 x <- seq(0, 1, length=100)
@@
 
-```{r}
+```{r very-simple-splines-09}
 par(mar=rep(0.1,4))

### 2024-07-06 text/21/zero-inflated-poisson.Rmd
Updated 2022 Rmd files

--- 2023-02-19 text/21/zero-inflated-poisson.Rmd
@@
 
-```{r setup}
+```{r zero-inflated-poisson-01}
 suppressMessages(suppressWarnings(library(broom)))
@@
 
-```{r, eval=FALSE}
+```{r zero-inflated-poisson-02, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r zero-inflated-poisson-03}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r zero-inflated-poisson-04}
 pois1 %>%
@@
 
-```{r}
+```{r zero-inflated-poisson-05}
 zip1 <- zeroinfl(art ~ fem + ment + phd + mar + kid5 | 1, family=poisson, data = articles)

### 2024-07-06 text/22/analysis-of-means.Rmd
Updated 2022 Rmd files

@@
 category: Blog post
-tags: Analysis of means
+tags: 
+- Analysis of means
+- Rmarkdown graphs
 source: new
@@
 
-```{r, fig.width=5, fig.height=0.33, echo=FALSE}
+```{r analysis-of-means-01, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
@@
 
-```{r, fig.width=5, fig.height=0.33, echo=FALSE}
+```{r analysis-of-means-02, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu ~ " for all i=1,...,k")
@@
 
-```{r, echo=FALSE}
+```{r analysis-of-means-03, echo=FALSE}
 mn <- c(
@@
 
-```{r, echo=FALSE}
+```{r analysis-of-means-04, echo=FALSE}
 # Make the numbers look nice

### 2024-07-06 text/22/box-with-no-whiskers.Rmd
Updated 2022 Rmd files

@@
 
-```{r, fig.height=1.5, echo=FALSE}
+```{r box-with-no-whiskers-01, fig.height=1.5, echo=FALSE}
 suppressMessages(suppressWarnings(library(tidyverse)))
@@
 
-```{r, fig.height=1.5, echo=FALSE}
+```{r box-with-no-whiskers-02, fig.height=1.5, echo=FALSE}
 pseudo_boxplot(10, 2, 6, 2, 5, 3)

### 2024-07-06 text/22/definition-schoenfeld-residuals.Rmd
Updated 2022 Rmd files

--- 2022-12-09 text/22/definition-schoenfeld-residuals.Rmd
@@
 category: Blog post
-tags: Survival analysis
+tags:
+- Rmarkdown code
+- Survival analysis
 source: new
+page_update: complete
 ---
@@
 
-```{r setup, echo=FALSE}
+```{r definition-schoenfeld-residuals-01, echo=FALSE}
 knitr::opts_chunk$set(echo=FALSE)
@@
 
-```{r}
+```{r definition-schoenfeld-residuals-02}
 library(plot3D)
@@
 
->>>>>>> 17772f400cc3acd5b2c61623fc3cf2c98df50124
-
 $L_i(\beta)  

### 2024-07-06 text/22/different-parameterizations.Rmd
Updated 2022 Rmd files

@@
 - Analysis of variance
+- Rmarkdown code
 output: html_document
@@
 
-```{r setup}
+```{r different-parameterizations-01}
 library(MASS)
@@
 
-```{r}
+```{r different-parameterizations-02}
 means <- c(52, 54, 62, 72)
@@
 
-```{r}
+```{r different-parameterizations-03}
 data.frame(g, y) %>%
@@
 
-```{r}
+```{r different-parameterizations-04}
 x1 <- as.numeric(g==1)
@@
 
-```{r}
+```{r different-parameterizations-05}
 A <- cbind(x1, x2, x3, x4) 
@@
 
-```{r}
+```{r different-parameterizations-06}
 beta_A <- coef(lm(y~x1+x2+x3+x4-1))
@@
 
-```{r}
+```{r different-parameterizations-07}
 inv(t(A)%*%A)%*%t(A) %>% fractions
@@
 
-```{r}
+```{r different-parameterizations-08}
 x0 <- rep(1, 12)
@@
 
-```{r}
+```{r different-parameterizations-09}
 B <- cbind(x0, x1, x2, x3) 
@@
 
-```{r}
+```{r different-parameterizations-10}
 beta_B <- coef(lm(y~x0+x1+x2+x3-1))
@@
 
-```{r}
+```{r different-parameterizations-11}
 inv(t(B)%*%B)%*%t(B) %>% fractions
@@
 
-```{r}
+```{r different-parameterizations-12}
 B[1:3, ]
@@
 
-```{r}
+```{r different-parameterizations-13}
 B[4:6, ]
@@
 
-```{r}
+```{r different-parameterizations-14}
 x0 <- rep(1, 12)
@@
 
-```{r}
+```{r different-parameterizations-15}
 D <- cbind(x0, x1, x2, x3) 
@@
 
-```{r}
+```{r different-parameterizations-16}
 beta_D <- coef(lm(y~x0+x1+x2+x3-1))
@@
 
-```{r}
+```{r different-parameterizations-17}
 inv(t(D)%*%D)%*%t(D) %>% fractions
@@
 
-```{r}
+```{r different-parameterizations-18}
 B
@@
 
-```{r, echo=FALSE}
+```{r different-parameterizations-19, echo=FALSE}
 V <- matrix(c(
@@
 
-```{r}
+```{r different-parameterizations-20}
 A%*%V
@@
 
-```{r}
+```{r different-parameterizations-21}
 U=inv(V)
@@
 
-```{r}
+```{r different-parameterizations-22}
 B%*%U
@@
 
-```{r}
+```{r different-parameterizations-23}
 beta_A
@@
 
-```{r}
+```{r different-parameterizations-24}
 beta_B
@@
 
-```{r}
+```{r different-parameterizations-25}
 inv(V)%*%beta_A
@@
 
-```{r}
+```{r different-parameterizations-26}
 V%*%beta_B
@@
 
-```{r}
+```{r different-parameterizations-27}
 A
@@
 
-```{r}
+```{r different-parameterizations-28}
 D
@@
 
-```{r, echo=FALSE}
+```{r different-parameterizations-29, echo=FALSE}
 V <- matrix(c(
@@
 
-```{r}
+```{r different-parameterizations-30}
 A%*%V
@@
 
-```{r}
+```{r different-parameterizations-31}
 U=inv(V)
@@
 
-```{r}
+```{r different-parameterizations-32}
 D%*%U
@@
 
-```{r}
+```{r different-parameterizations-33}
 beta_A
@@
 
-```{r}
+```{r different-parameterizations-34}
 beta_D
@@
 
-```{r}
+```{r different-parameterizations-35}
 inv(V)%*%beta_A
@@
 
-```{r}
+```{r different-parameterizations-36}
 V%*%beta_D
@@
 
-```{r, echo=FALSE}
+```{r different-parameterizations-37, echo=FALSE}
 U <- matrix(c(
@@
 
-```{r}
+```{r different-parameterizations-38}
 A
@@
 
-```{r}
+```{r different-parameterizations-39}
 V <- inv(U)
@@
 
-```{r}
+```{r different-parameterizations-40}
 E <- A%*%inv(U)
@@
 
-```{r}
+```{r different-parameterizations-41}
 x0 <- E[ , 1]

### 2024-07-06 text/22/dunnetts-test.Rmd
Updated 2022 Rmd files

--- 2022-09-06 text/22/dunnetts-test.Rmd
@@
 category: Blog post
-tags: Analysis of means
+tags:
+- Analysis of means
+- Rmarkdown graphs
 source: new
+page_update: complete
 ---
@@
 
-```{r, fig.width=5, fig.height=0.33, echo=FALSE}
+```{r dunnetts-test-01, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[j] ~ " for all i,j")
@@
 
-```{r, fig.width=5, fig.height=0.33, echo=FALSE}
+```{r dunnetts-test-02, fig.width=5, fig.height=0.33, echo=FALSE}
 h0 <- bquote(H[0] ~ ": " ~ mu[i] == mu[1] ~ " for all i=2,...,k")
@@
 
-```{r, echo=FALSE, comment=""}
+```{r dunnetts-test-03, echo=FALSE, comment=""}
 mn1 <- c(21.39, 22.58, 16.94, 21.93, 19.93, 22.00)
@@
 
-```{r, echo=FALSE, comment=""}
+```{r dunnetts-test-04, echo=FALSE, comment=""}
 suppressMessages(suppressWarnings(library(tidyverse)))
@@
 
-```{r, echo=FALSE, comment=""}
+```{r dunnetts-test-05, echo=FALSE, comment=""}
 data.frame(g=2:6, mn=mn1[-1]) %>%
@@
 
-```{r echo=FALSE}
+```{r dunnetts-test-06, echo=FALSE}
 # Make the numbers look nice.

### 2024-07-06 text/22/mapping-longitude-latitude.Rmd
Updated 2022 Rmd files

@@
 
-```{r usa-map}
+```{r mapping-longitude-latitude-01}
 library(maps); library(ggplot2); library(mapproj)
@@
 
-```{r peek}
+```{r mapping-longitude-latitude-02}
 head(states)
@@
 
-```{r plot3d}
+```{r mapping-longitude-latitude-03}
 library(dplyr)
@@
 
-```{r square}
+```{r mapping-longitude-latitude-04}
 scatter3D(

### 2024-07-06 text/22/posix-time.Rmd
Updated 2022 Rmd files

--- 2022-11-13 text/22/posix-time.Rmd
@@
 category: Blog post
-tags: Data science
+tags:
+- Data science
+- Rmarkdown code
 source: new
+page_update: complete
 ---
@@
 
-```{r}
+```{r posix-time-01}
 date_string <- "1970-01-01"
@@
 
-```{r}
+```{r posix-time-02}
 print.default(date_string)
@@
 
-```{r}
+```{r posix-time-03}
 print.default(date_internal_r_format)
@@
 
-```{r}
+```{r posix-time-04}
 print.default(date_posix_ct)
@@
 
-```{r}
+```{r posix-time-05}
 print.default(date_posix_lt)
@@
 
-```{r}
+```{r posix-time-06}
 date_internal_r_format <- as.Date(1, origin="1970-01-01")
@@
 
-```{r}
+```{r posix-time-07}
 as.POSIXct(86400, origin="1970-01-01", tz="GMT")

### 2024-07-06 text/22/r-creatures.Rmd
Updated 2022 Rmd files

--- 2024-06-13 text/22/r-creatures.Rmd
@@
 - R software
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r r-creatures-01}
 secret_of_the_universe <- 42
@@
 
-```{r}
+```{r r-creatures-02}
 buckle_my_shoe <- c(1, 2)
@@
 
-```{r}
+```{r r-creatures-03}
 fantastic_four <- 1:4
@@
 
-```{r}
+```{r r-creatures-04}
 length(vowels)
@@
 
-```{r}
+```{r r-creatures-05}
 vowels[1]
@@
 
-```{r}
+```{r r-creatures-06}
 vowels[1:2]
@@
 
-```{r}
+```{r r-creatures-07}
 c(buckle_my_shoe, fantastic_four)
@@
 
-```{r}
+```{r r-creatures-08}
 primes <- c(
@@
 
-```{r}
+```{r r-creatures-09}
 primes["first"]
@@
 
-```{r}
+```{r r-creatures-10}
 names(primes)
@@
 
-```{r}
+```{r r-creatures-11}
 mixed_vector <- c(1, "two")
@@
 
-```{r}
+```{r r-creatures-12}
 mixed_list <- list(1, "two")
@@
 
-```{r}
+```{r r-creatures-13}
 mixed_list[[2]]
@@
 
-```{r}
+```{r r-creatures-14}
 groceries <- list("bread", "fruit")
@@
 
-```{r}
+```{r r-creatures-15}
 unlist(shopping)
@@
 
-```{r}
+```{r r-creatures-16}
 names(shopping) <- c("grocery_list", "clothing_list")
@@
 
-```{r}
+```{r r-creatures-17}
 shopping$clothing_list
@@
 
-```{r}
+```{r r-creatures-18}
 diagonal_matrix <- matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), nrow=3)
@@
 
-```{r}
+```{r r-creatures-19}
 dim(diagonal_matrix)
@@
 
-```{r}
+```{r r-creatures-20}
 diagonal_matrix[1, 3]
@@
 
-```{r}
+```{r r-creatures-21}
 diagonal_matrix[1, ]
@@
 
-```{r}
+```{r r-creatures-22}
 dimnames(diagonal_matrix) <- list(c("A", "B", "C"), c("D", "E", "F"))
@@
 
-```{r}
+```{r r-creatures-23}
 cbind(fantastic_four, not_so_fantastic_four)
@@
 
-```{r}
+```{r r-creatures-24}
 chain <- c(
@@
 
-```{r}
+```{r r-creatures-25}
 fast_food_alt <- data.frame(fat, sodium, calories)
@@
 
-```{r}
+```{r r-creatures-26}
 fast_food[, 2]
@@
 
-```{r}
+```{r r-creatures-27}
 library(tidyr)
@@
 
-```{r}
+```{r r-creatures-28}
 library(survival)
@@
 
-```{r}
+```{r r-creatures-29}
 print(tibble(colon), n=25)
@@
 
-```{r}
+```{r r-creatures-30}
 fast_food[5, ]
@@
 
-```{r}
+```{r r-creatures-31}
 real_fast_food <- fast_food[-4, ]
@@
 
-```{r}
+```{r r-creatures-32}
 fast_food_regression <- lm(calories~fat+sodium, data=real_fast_food)
@@
 
-```{r}
+```{r r-creatures-33}
 names(fast_food_regression)
@@
 
-```{r}
+```{r r-creatures-34}
 fast_food_regression$coefficients
@@
 
-```{r}
+```{r r-creatures-35}
 fast_food_regression$residuals
@@
 
-```{r}
+```{r r-creatures-36}
 fast_food_regression$df.residual
@@
 
-```{r}
+```{r r-creatures-37}
 fast_food_regression$qr
@@
 
-```{r}
+```{r r-creatures-38}
 fast_food_summary <- summary(fast_food_regression)
@@
 
-```{r}
+```{r r-creatures-39}
 fast_food_summary$coefficients
@@
 
-```{r}
+```{r r-creatures-40}
 fast_food_summary$coefficients[c("fat", "sodium"), "Pr(>|t|)"]
@@
 
-```{r}
+```{r r-creatures-41}
 fast_food_summary$r.squared
@@
 
-```{r}
+```{r r-creatures-42}
 fast_food_summary$fstatistic

### 2024-07-07 text/23/exponential-hedging-hyperprior.Rmd
Updated 2024 Rmd files

--- 2023-03-28 text/23/exponential-hedging-hyperprior.Rmd
@@
 - Bayesian statistics
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-01}
 x_max <- 3
@@
 post95 <- 1/qgamma(c(0.025), a1, b1) 
-
 ```
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-02}
 g_grid <- g(x_grid, a0, b0)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-03}
 f_grid <- f(xbar, x_grid)^n
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-04}
 h_grid <- g(x_grid, a1, b1)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-05}
 knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-06}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-07}
 density <- outer(theta_grid, lambda_grid, gstar)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-08}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-09}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r exponential-hedging-hyperprior-10}
 density <- density / sum(density, na.rm=TRUE)

### 2024-07-07 text/23/mcnemar-test.Rmd
Updated 2024 Rmd files

--- 2023-10-31 text/23/mcnemar-test.Rmd
@@
 - Measuring benefit and risk
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---
@@
 
-```{r}
+```{r mcnemar-test-01}
 a <- 833; b <- 125; c <- 2; d <- 160
@@
 
-```{r}
+```{r mcnemar-test-02}
 mcnemar.test(matrix(c(833, 125, 2, 160), nrow=2), correct=FALSE)

### 2024-07-07 text/23/modified-power-prior.Rmd
Updated 2024 Rmd files

--- 2023-03-04 text/23/modified-power-prior.Rmd
@@
 - Bayesian statistics
+- Incomplete pages
 output: html_document
+page_update: complete
 ---
 
-```{r setup}
+```{r modified-power-prior-01}
 suppressMessages(suppressWarnings(library(tidyverse)))

### 2024-07-07 text/23/non-standard-evaluation.Rmd
Updated 2024 Rmd files

--- 2024-06-13 text/23/non-standard-evaluation.Rmd
@@
 
-```{r}
+```{r non-standard-evaluation-01}
 old_data <- data.frame(
@@
 
-```{r}
+```{r non-standard-evaluation-02}
 new_data <- select(old_data, w, x)
@@
 
-```{r}
+```{r non-standard-evaluation-03}
 new_data <- old_data[old_data$w < 2, ]
@@
 
-```{r}
+```{r non-standard-evaluation-04}
 new_data <- filter(old_data, w < 2)
@@
 
-```{r}
+```{r non-standard-evaluation-05}
 for (v in c("w", "x")) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-06, error=TRUE}
 for (v in c("w", "x")) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-07, error=TRUE}
 for (v in c("w", "x")) {
@@
 
-```{r}
+```{r non-standard-evaluation-08}
 reduce <- function(old_data, variable_name) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-09, error=TRUE}
 reduce <- function(old_data, variable_name) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-10, error=TRUE}
 reduce(old_data, "w")
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-11, error=TRUE}
 reduce <- function(old_data, variable_name) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-12, error=TRUE}
 reduce <- function(old_data, ...) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-13, error=TRUE}
 reduce <- function(old_data, variable_string) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-14, error=TRUE}
 reduce <- function(old_data, variable_string) {
@@
 
-```{r, error=TRUE}
+```{r non-standard-evaluation-15, error=TRUE}
 reduce <- function(old_data, variable_string) {

### 2024-07-07 text/23/normal-hedging-hyperprior-part-1.Rmd
Updated 2024 Rmd files

@@
 - Bayesian statistics
+- Rmarkdown graphs
 output: html_document
@@
 
-```{r setup}
+```{r normal-hedging-prior-01}
 suppressMessages(suppressWarnings(library(MASS)))
@@
 
-```{r}
+```{r normal-hedging-prior-02}
 x_max <- 8
@@
 
-```{r}
+```{r normal-hedging-prior-03}
 g_grid <- g(x_grid, mu0, tau0)
@@
 
-```{r}
+```{r normal-hedging-prior-04}
 f_grid <- f(xbar, x_grid, tauF)^n
@@
 
-```{r}
+```{r normal-hedging-prior-05}
 h_grid <- g(x_grid, mu1, tau1)
@@
 
-```{r}
+```{r normal-hedging-prior-06}
 knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
@@
 
-```{r}
+```{r normal-hedging-prior-07}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r normal-hedging-prior-08}
 density <- outer(mu_grid, lambda_grid, gstar)
@@
 
-```{r}
+```{r normal-hedging-prior-09}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r normal-hedging-prior-10}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r}
+```{r normal-hedging-prior-11}
 density <- density / sum(density, na.rm=TRUE)

### 2024-07-07 text/23/normal-hedging-hyperprior-part-2.Rmd
Updated 2024 Rmd files

@@
 
-```{r setup}
+```{r normal-hedging-prior-20}
 suppressMessages(suppressWarnings(library(MASS)))
@@
 
-```{r}
+```{r normal-hedging-prior-21}
 x_max <- 8
@@
 
-```{r}
+```{r normal-hedging-prior-22}
 g_grid <- g(x_grid, alpha0, beta0)
@@
 
-```{r}
+```{r normal-hedging-prior-23}
 f_grid <- rep(NA, length(x_grid))
@@
 
-```{r}
+```{r normal-hedging-prior-24}
 h_grid <- g(x_grid, alpha1, beta1)
@@
 
-```{r}
+```{r normal-hedging-prior-25}
 knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
@@
 
-```{r}
+```{r normal-hedging-prior-26}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-27, eval=FALSE}
 density <- outer(tau_grid, lambda_grid, gstar)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-28, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-29, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-30, eval=FALSE}
 density <- density / sum(density, na.rm=TRUE)

### 2024-07-07 text/23/normal-hedging-hyperprior-part-3.Rmd
Updated 2024 Rmd files

@@
 
-```{r setup}
+```{r normal-hedging-prior-40}
 suppressMessages(suppressWarnings(library(MASS)))
@@
 
-```{r}
+```{r normal-hedging-prior-41}
 x_max <- 8
@@
 
-```{r}
+```{r normal-hedging-prior-42}
 g_grid <- g(x_grid, alpha0, beta0)
@@
 
-```{r}
+```{r normal-hedging-prior-43}
 f_grid <- rep(NA, length(x_grid))
@@
 
-```{r}
+```{r normal-hedging-prior-44}
 h_grid <- g(x_grid, alpha1, beta1)
@@
 
-```{r}
+```{r normal-hedging-prior-45}
 knitr::opts_chunk$set(echo=TRUE, fig.width=8, fig.height=8)
@@
 
-```{r}
+```{r normal-hedging-prior-46}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-47, eval=FALSE}
 density <- outer(theta_grid, lambda_grid, gstar)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-48, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-49, eval=FALSE}
 lambda_grid <- seq(0, 1, length=60)
@@
 
-```{r, eval=FALSE}
+```{r normal-hedging-prior-50, eval=FALSE}
 density <- density / sum(density, na.rm=TRUE)

### 2024-07-07 text/23/taylor-series-approximation.Rmd
Updated 2024 Rmd files

--- 2023-09-03 text/23/taylor-series-approximation.Rmd
@@
 - Mathematics
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r, echo=FALSE}
+```{r taylor-series-approximation-01, echo=FALSE}
 d0 <- 2.5
@@
 
-```{r, echo=FALSE}
+```{r taylor-series-approximation-02, echo=FALSE}
 library(ggplot2, quiet=TRUE)
@@
 
-```{r, echo=FALSE}
+```{r taylor-series-approximation-03, echo=FALSE}
 library(ggplot2, quiet=TRUE)

### 2024-07-07 text/23/too-many-colors.rmd
Updated 2024 Rmd files

@@
 
-```{r, echo=FALSE, warning=FALSE}
+```{r too-many-colors-01, echo=FALSE, warning=FALSE}
 library(usmap)
@@
 
-```{r, echo=FALSE, warning=FALSE}
+```{r too-many-colors-02, echo=FALSE, warning=FALSE}
 gr <- c("green", "yellow", "red")
@@
 
-```{r, echo=FALSE, warning=FALSE}
+```{r too-many-colors-03, echo=FALSE, warning=FALSE}
 gr <- c("red", "yellow", "green", "cyan", "blue", "magenta")

### 2024-07-07 text/23/what-is-a-partial-correlation.Rmd
Updated 2024 Rmd files

--- 2023-12-16 text/23/what-is-a-partial-correlation.Rmd
@@
 - Definitions
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r, echo=FALSE, warning=FALSE, message=FALSE}
+```{r partial-correlation-01, echo=FALSE, warning=FALSE, message=FALSE}
 library(broom)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-02, echo=FALSE}
 rxy <- round(cor(init$fev, init$age), 2)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-03, echo=FALSE}
 ggplot(init, aes(age, fev, label=height)) +
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-04, echo=FALSE}
 g <- 4*floor((init$height-2)/4)+4
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-05, echo=FALSE}
 control_height(h)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-06, echo=FALSE}
 control_height(h)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-07, echo=FALSE}
 control_height(h)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-08, echo=FALSE}
 control_height(h)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-09, echo=FALSE}
 control_height(h)
@@
 
-```{r, echo=FALSE}
+```{r partial-correlation-10, echo=FALSE}
 control_height(h)

### 2024-07-07 text/24/diamond-spline.Rmd
Updated 2024 Rmd files

@@
 
-```{r initialization, echo=FALSE}
+```{r diamond-spline-01, echo=FALSE}
 library(broom)
@@
 
-```{r read-diamonds}
+```{r diamond-spline-02}
 f2 <- "singapore-diamond-prices.txt"
@@
 
-```{r adjustments}
+```{r diamond-spline-03}
 exchange_rate <- 1.72
@@
 
-```{r descriptives}
+```{r diamond-spline-04}
 summary(di$carat)
@@
 
-```{r plot}
+```{r diamond-spline-05}
 ggplot(data=di, aes(carat, adjusted_price)) +
@@
 
-```{r linear-trend}
+```{r diamond-spline-06}
 scatter1+geom_smooth(method="lm")
@@
 
-```{r linear-fit}
+```{r diamond-spline-07}
 linear_fit <- lm(adjusted_price ~ carat, data=di)
@@
 
-```{r spline-fit}
+```{r diamond-spline-08}
 spline_fit <- lm(adjusted_price ~ rcs(carat), data=di)
@@
 
-```{r spline-plot}
+```{r diamond-spline-09}
 carat_sequence <- data.frame(
@@
 
-```{rlinear-plus-spline}
+```{r diamond-spline-10}
 linear_plus_spline_fit <- lm(

### 2024-07-07 text/24/different-end-times.Rmd
Updated 2024 Rmd files

@@
 
-```{r}
+```{r different-end-times-01}
 suppressMessages(library(survival))
@@
 
-```{r}
+```{r different-end-times-02}
 sim %>%
@@
 
-```{r}
+```{r different-end-times-04}
 censor_time_v1 <- rep(c(190, 225), each=100)
@@
 
-```{r}
+```{r different-end-times-05}
 cox_v1 <- coxph(Surv(t_v1, censor_v1)~g, data=sim_v1)
@@
 
-```{r}
+```{r different-end-times-06}
 censor_time_v2 <- rep(c(190, 190), each=100)
@@
 
-```{r}
+```{r different-end-times-07}
 cox_v2 <- coxph(Surv(t_v2, censor_v2)~g, data=sim_v2)

### 2024-07-07 text/24/nice-layout.Rmd
Updated 2024 Rmd files

@@
 tags:
-- Being updated
+- Incomplete pages
+- R programming
+- Rmarkdown code
 output: html_document
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r nice-layout-01, echo=FALSE, message=FALSE, warning=FALSE}
 library(glue)
@@
 
-```{r, echo=FALSE}
+```{r nice-layout-02, echo=FALSE}
 f2 <- "titanic.txt"
@@
 
-```{r}
+```{r nice-layout-03}
 ti0 %>%
@@
 
-```{r}
+```{r nice-layout-04}
 tab01 %>%
@@
 
-```{r}
+```{r nice-layout-05}
 tab02 %>%
@@
 
-```{r}
+```{r nice-layout-06}
 tab03 %>%
@@
 
-```{r}
+```{r nice-layout-07}
 tab05 %>%
@@
 
-```{r}
+```{r nice-layout-08}
 library(kableExtra)
@@
 
-```{r}
+```{r nice-layout-09}
 library(kableExtra)
@@
 
-```{r}
+```{r nice-layout-10}
 ti0 %>%
@@
 
-```{r}
+```{r nice-layout-11}
 ti0 %>%

### 2024-07-07 text/24/survival-lasso.Rmd
Updated 2024 Rmd files

--- 2024-05-27 text/24/survival-lasso.Rmd
@@
 - R programming
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---

### 2024-07-07 text/24/survival-spline.Rmd
Updated 2024 Rmd files

--- 2024-06-13 text/24/survival-spline.Rmd
@@
 tags:
+- Incomplete pages
 - Survival analysis
 - R programming
+- Rmarkdown graphs
 output: html_document
+page_update: complete
 ---
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r survival-spline-01, echo=FALSE, message=FALSE, warning=FALSE}
 library(broom)
@@
 
-```{r}
+```{r survival-spline-02}
 f2 <- "whas100.dat"
@@
 
-```{r}
+```{r survival-spline-03}
 summary(wa$id)
@@
 
-```{r}
+```{r survival-spline-04}
 plot(Surv(wa$lenfol, wa$fstat))
@@
 
-```{r}
+```{r survival-spline-05}
 linear_fit <- coxph(
@@
 
-```{r}
+```{r survival-spline-06}
 spline_fit <- coxph(
@@
 
-```{r}
+```{r survival-spline-07}
 age_sequence <- data.frame(

### 2024-07-07 text/24/titanic-imputation.Rmd
Updated 2024 Rmd files

--- 2024-05-27 text/24/titanic-imputation.Rmd
@@
 - R programming
+- Rmarkdown code
 output: html_document
+page_update: complete
 ---
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r titanic-imputation-01, echo=FALSE, message=FALSE, warning=FALSE}
 library(glue)
@@
 
-```{r, echo=FALSE}
+```{r titanic-imputation-02, echo=FALSE}
 f2 <- "titanic.txt"
@@
 
-```{r, echo=FALSE}
+```{r titanic-imputation-03, echo=FALSE}
 cat(vnames[1], ": ", vlabels[1], "\n")
@@
 
-```{r}
+```{r titanic-imputation-04}
 missing_rows <- which(is.na(ti0$Age))[1:3]
@@
 
-```{r}
+```{r titanic-imputation-05}
 ti1 <- select(ti0, -Name)
@@
 
-```{r}
+```{r titanic-imputation-06}
 ti_mice %>%
@@
 
-```{r}
+```{r titanic-imputation-07}
 ti_mice %>%
@@
 
-```{r}
+```{r titanic-imputation-08}
 ti_with <- with(
@@
 
-```{r}
+```{r titanic-imputation-09}
 ti_with$analyses[[1]]
@@
 
-```{r}
+```{r titanic-imputation-10}
 ti_with$analyses[[2]]
@@
 
-```{r}
+```{r titanic-imputation-11}
 ti_pool <- pool(ti_with)
@@
 
-```{r}
+```{r titanic-imputation-12}
 female <- ti_summary$term=="i_female"

### 2024-07-07 text/24/titanic-spline.Rmd
Updated 2024 Rmd files

--- 2024-05-27 text/24/titanic-spline.Rmd
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r titanic-spline-01, echo=FALSE, message=FALSE, warning=FALSE}
 library(broom)
@@
 
-```{r}
+```{r titanic-spline-02}
 f2 <- "titanic.txt"
@@
 
-```{r}
+```{r titanic-spline-03}
 summary(ti0$Age)
@@
 
-```{r}
+```{r titanic-spline-04}
 ggplot(data=ti0, aes(factor(survived), age)) + 
@@
 
-```{r}
+```{r titanic-spline-05}
 ti1 <- filter(ti0, !is.na(age))
@@
 
-```{r}
+```{r titanic-spline-06}
 age_sequence <- data.frame(
@@
 
-```{r}
+```{r titanic-spline-07}
 age_sequence$prediction <- 
@@
 
-```{r}
+```{r titanic-spline-08}
 spline_fit <- glm(
@@
 
-```{r}
+```{r titanic-spline-09}
 age_sequence$prediction <- 
@@
 
-```{r}
+```{r titanic-spline-10}
 age_sequence$prediction <- 

### 2024-07-07 text/24/wavelets.Rmd
Updated 2024 Rmd files

--- 2022-02-06 text/14/wavelets.Rmd
@@
 - R software
+- Rmarkdown graphs
 output: html_document
+page_update: partial
 ---
@@
 
-```{r setup}
+```{r wavelets-01}
 knitr::opts_chunk$set(fig.height=1)
@@
 
-```{r read-ecg-data}
+```{r wavelets-02}
+library("ggplot2")
+library("magrittr")
 library("rwt")
@@
 
-```{r graphing-functions}
+```{r wavelets-03}
 graph_with_axis <- function(x, lab, limits=range(x)) {
@@
 
-```{r graph-ecg, fig.height=2}
+```{r wavelets-04, fig.height=2}
 graph_with_axis(ecg, "ecg\nsignal")
@@
 
-```{r apply-fft}
+```{r wavelets-05}
 ft.ecg <- fft(ecg)/2048
@@
 
-```{r complex-plot, fig.height=5}
+```{r wavelets-06, fig.height=5}
 par(mar=c(0,0,0,0))
@@
 
-```{r inverse-fft}
+```{r wavelets-07}
 ift.ecg <- fft(ft.ecg,inverse=TRUE)
-head(ift.ecg)
+ead(ift.ecg)
 tail(ift.ecg)
@@
 
-```{r compute-mean}
+```{r wavelets-08}
 mean(ecg)
@@
 
-```{r graph-cosine-1}
+```{r wavelets-09}
 t.unit <- (1:2048)/2048
@@
 
-```{r compare-to-original-curve}
+```{r wavelets-10}
 {
@@
 
-```{r second-component}
+```{r wavelets-11}
 ft.comp2 <- ft.ecg
@@
 
-```{r graph-third-component}
+```{r wavelets-12}
 ft.comp3 <- ft.ecg
@@
 
-```{r graph-fourth-component}
+```{r wavelets-13}
 ft.comp4 <- ft.ecg
@@
 
-```{r thirty-second}
+```{r wavelets-14}
 ft.comp32 <- ft.ecg
@@
 
-```{r count}
+```{r wavelets-15}
 m <- max(comp32)
@@
 
-```{r rwt-package}
+```{r wavelets-16}
 d4.rwt <- daubcqf(4)
@@
 
-```{r wavelets-package}
+```{r wavelets-17}
 d4.wavelets <- wt.filter("d4")
@@
 
-```{r graph-d4}
+```{r wavelets-18}
 wt.comp <- function(x, y) {
@@
 
-```{r calculate-functions}
+```{r wavelets-19}
 calculate_mother <- function(i, L) {
@@
 
-```{r mother-and-father}
+```{r wavelets-20}
 {
@@
 
-```{r shifted}
+```{r wavelets-21}
 {
@@
 
-```{r wrapped}
+```{r wavelets-22}
 {
@@
 
-```{r stretch-and-squeeze}
+```{r wavelets-23}
 {  
@@
 
-```{r most-extreme}
+```{r wavelets-24}
 {
@@
 
-```{r high-pass}
+```{r wavelets-25}
 g <- daubcqf(4)$h.0
@@
 
-These coefficients represent the "high pass" filter. There is a complementary "low pass" filter that you get by reversing the coefficients and changing the sign of every other value.
-
-Note: To get the math to work on my end, tThe low pass filter provided by the rwt package needs to be negated.
-
-```{r low-pass}
+These coefficients represent the "high pass" filter.
+There is a complementary "low pass" filter that you get by reversing the coefficients and changing the sign of every other value.
+The term "high pass" means that it passes high frequencies above a certain cutoff and attenuates low frequencies.
+The term "low pass" means the opposite.
+
+Note that not all wavelets develop the low pass filter in the way described above. The construction is known as a quadrature mirror filter. There is an attractive simplicity to this approach.
+
+```{r wavelets-26}
 h <- -daubcqf(4)$h.1
@@
 
-These two filter pairs are known as a quadrature mirror filter because of the reflection. These two filters are orthonormal.
-
-```{r orthonormal}
+These two filters are orthonormal.
+
+```{r wavelets-27}
 # The inline function %*% calculates a dot product of two vectors.
@@
 
-```{r overlap}
+```{r wavelets-28}
 g[1:2] %*% g[3:4]
@@
 
-```{r orthogonal-to-linear}
+```{r wavelets-29}
 h %*% (1:4)
@@
 
-```{r first-level-wavelet}
+```{r wavelets-30}
 wt.1 <- mdwt(ecg, daubcqf(4)$h.0, L=1)
@@
 
-```{r first-level-dot-products}
+```{r wavelets-31}
 c(V1[   1], g %*% ecg[1:4])
@@
 
-```{r second-level-wavelet}
+```{r wavelets-32}
 wt.2 <- mdwt(ecg, daubcqf(4)$h.0, L=2)
@@
 
-```{r second-level-dot-products}
+```{r wavelets-33}
 c(V2[  1], g %*% V1[1:4])
@@
 
-```{r transform-V1}
+```{r wavelets-34}
 quality_check <- mdwt(V1, daubcqf(4)$h.0, L=1)
@@
 
-```{r third-level-wavelet}
+```{r wavelets-35}
 wt.3 <- mdwt(ecg, daubcqf(4)$h.0, L=3)
@@
 
-```{r third-level-dot-products}
+```{r wavelets-36}
 c(V3[  1], g %*% V2[1:4])
@@
 
-```{r ninth-level-wavelet}
+```{r wavelets-37}
 wt.9 <- mdwt(ecg, daubcqf(4)$h.0, L=9)
@@
 
-```{r reconstuct-V8}
+```{r wavelets-38}
 V9_diluted <- rep(0, 8)
@@
 
-```{r reconstuct-V7}
+```{r wavelets-39}
 V8_diluted <- rep(0, 16)
@@
 
-```{r reconstuct-V6}
+```{r wavelets-40}
 V7_diluted <- rep(0, 32)
@@
 
-```{r graph-v}
+```{r wavelets-41}
 
@@
 
-```{r graph-w}
+```{r wavelets-42}
 
@@
 
-```{r read-dog-image}
+```{r wavelets-43}
 knitr::opts_chunk$set(fig.width=5, fig.height=5)
@@
 
-```{r display-dog-image}
+```{r wavelets-44}
 img.plot <- function(imat,n=256) {
@@
 
-```{r d8-wavelet-g}
+```{r wavelets-45}
 g <- daubcqf(8)$h.0
@@
 
-```{r d8-wavelet-h}
+```{r wavelets-46}
 h <- -daubcqf(8)$h.1
@@
 
-```{r d8-in-1d, fig.height=1}
+```{r wavelets-47, fig.height=1}
 d8_father_coefficients <- rep(0, 2048)
@@
 
-```{r decompose-and-restructure}
+```{r wavelets-48}
 wt.img.4 <- mdwt(img, daubcqf(8)$h.0, L=4)
@@
 
-```{r check-2d-calculations}
+```{r wavelets-49}
 wt.img.1 <- mdwt(img, daubcqf(8)$h.0, L=1)
@@
 
-```{r level-1-image-g-g}
+```{r wavelets-50}
+x <- c(700, 300, 700)
+y <- c(700, 300, 300)
+m <- c(
+  "W1\nh %*% g", 
+  "W1\ng %*% h", 
+  "W1\nh %*% h"
+)
+data.frame(x, y, m) %>%
+  ggplot(aes(x, y, label=m)) +
+  expand_limits(x=c(100, 900), y=c(100, 900)) + 
+  geom_segment(x=100, y=100, xend=900, yend=100) +
+  geom_segment(x=900, y=100, xend=900, yend=900) +
+  geom_segment(x=900, y=900, xend=100, yend=900) +
+  geom_segment(x=100, y=900, xend=100, yend=100) +
+  geom_segment(x=500, y=100, xend=500, yend=900) +
+  geom_segment(x=100, y=500, xend=900, yend=500) +
+  theme_void() + 
+  geom_text() -> layout_1
+layout_1 +
+    geom_text(x=300, y=700, label="V1\ng %*% g")
+layout_1+
+  geom_segment(x=300, y=500, xend=300, yend=900) +
+  geom_segment(x=100, y=700, xend=500, yend=700) +
+  geom_text(x=400, y=800, label="W2\nh %*% g") + 
+  geom_text(x=200, y=600, label="W2\ng %*% h") + 
+  geom_text(x=400, y=600, label="W2\nh %*% h") -> layout_2
+layout_2 +
+  geom_text(x=200, y=800, label="V2\ng %*% g")
+layout_2 +
+  geom_segment(x=200, y=700, xend=200, yend=900) +
+  geom_segment(x=100, y=800, xend=300, yend=800) +
+  geom_text(x=250, y=850, label="W3") + 
+  geom_text(x=150, y=750, label="W3") + 
+  geom_text(x=250, y=750, label="W3") + 
+  geom_text(x=150, y=850, label="etc.")
+
+```
+
+```{r wavelets-51}
 gg1 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-1-image-g-h}
+```{r wavelets-52}
 gh1 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-1-image-h-g}
+```{r wavelets-53}
 hg1 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-1-image-h-h}
+```{r wavelets-54}
 hh1 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-4-image-g-g}
+```{r wavelets-55}
 gg4 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-4-image-g-h}
+```{r wavelets-56}
 gh4 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-4-image-h-g}
+```{r wavelets-57}
 hg4 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-4-image-h-h}
+```{r wavelets-58}
 hh4 <- matrix(0, nrow=256, ncol=256)
@@
 
-```{r level-4-detail}
+```{r wavelets-59}
 detail4 <- gh4 + hg4 + hh4
@@
 
-```{r compress}
+```{r wavelets-60}
 pctl_99 <- quantile(abs(wt.img.4$y), probs=0.99)

### 2024-07-19 text/24/different-end-times.Rmd
Added missing-semicolon

@@
 sim %>%
+  mutate(
+    n_0_to_190 =
+      as.numeric(raw_time <= 190)) %>%
+  mutate(
+    n_191_to_225 =
+      as.numeric((raw_time > 190) & (raw_time<=225))) %>%
+  mutate(
+    n_above_225 =
+      as.numeric(raw_time > 225)) %>%
   group_by(g) %>%
   summarize(
-    n_0_to_190=sum(raw_time <= 190),
-    n_191_to_225=sum(raw_time > 190&raw_time<=225),
-    n_above_225=sum(raw_time > 225)) -> sim_counts
+    n_0_to_190=sum(n_0_to_190),
+    n_191_to_225=sum(n_191_to_225),
+    n_above_225=sum(n_above_225)) %>%
+  data.frame -> sim_counts
 sim_counts
+
 ```
 
-There are indeed more events at a threshold of 190 or below in group B (`r sim_counts$n_0_to_190[2]`) than there are in group A (`r sim_counts$n_0_to_190[1]`).
-There are, as well, more events at a threshold of 225 and below in group B (`r 100-sim_counts$n_above_225[2]`) than there are in group A  (`r 100-sim_counts$n_above_225[1]`).
-
-There are `r sim_counts$n_191_to_225[1]+sim_counts$n_191_to_225[1]` events that are in the "Twilight Zone".
-This means that `r sim_counts$n_191_to_225[1]` of the Twilight Zone events in group A are censored, while `r sim_counts$n_191_to_225[2]` of the Twilight Zone events in group B are still counted as events.
+There are several events that are in the "Twilight Zone". The Twilight Zone events in group A are censored, while the Twilight Zone events in group B are still counted as events.
 
@@
 
-What if you set the censoring times to 190 days for both groups? That effectively converts `r sim_counts$n_191_to_225[2]` patients in the second group from events to observations censored at 190 days.
+What if you set the censoring times to 190 days for both groups? That effectively converts patients in the Twilight Zone in the second group from events to observations censored at 190 days.
 

### 2024-07-21 text/21/negative-binomial-regression.Rmd

@@
 
-```{r, eval=FALSE}
+```{r negative-binomial-04, eval=FALSE}
 fn <- "http://faculty.missouri.edu/huangf/data/poisson/articles.csv"
@@
 
-```{r}
+```{r negative-binomial-05}
 fn <- "http://www.pmean.com/00files/couart2.dta"
@@
 
-```{r}
+```{r negative-binomial-06}
 pois1 <- glm(art ~ fem + ment + phd + mar + kid5, family=poisson, data = articles)
@@
 
-```{r negative-binomial-regression-01}
+```{r negative-binomial-regression-07}
 library(MASS)

### 2024-07-21 text/24/titanic-imputation.Rmd

@@
 o <-  c(1, 3, 2, 4, 5)
-ti0 <- select(raw_data, o)
-vnames <- vnames[o]
+ti0 <- raw_data
+# ti0 <- raw_data[ , o]
+# vnames <- vnames[o]
+# vlabels <- vlabels[o]
 head(ti0)
-```
-
-Here are a few descriptive statistics
-
-```{r titanic-imputation-03, echo=FALSE}
-cat(vnames[1], ": ", vlabels[1], "\n")
-summary(ti0$Age)
-cat(vnames[2], ": ", vlabels[2], "\n")
-count(ti0, PClass)
-cat(vnames[3], "\n")
-count(ti0, Sex)
-cat(vnames[4], "\n")
-cat(as.yaml(dd$vars$Survived$values), "\n")
-count(ti0, Survived)
 ```
@@
 
-Create mutliply imputed values for age.
+Create multiply imputed values for age.
 The default is to use every variable in the dataset other than age to impute the value of age.
@@
 ```{r titanic-imputation-05}
-ti1 <- select(ti0, -Name)
+ti1 <- ti0
 ti1$i_female <- as.numeric(ti1$Sex=="female")
-ti1 <- select(ti1, -Sex)
+ti1 <- ti1[ , c("PClass", "Age", "Survived", "i_female")]
 ti_mice <- mice(ti1)
@@
 
-The key variable here is sexmale, which shows how much different the log odds of survival are between men and women.
+The key variable here is i_female, which shows how much different the log odds of survival are between men and women.
 

### 2024-08-01 src/clean_files.R
Added tags-template.qmd

@@ -51,6 +51,7 @@ clean_files <- function(
       k2 <- k2+1
       cat("\n\n")
       cat(glue("===> {fn}"))
+      cat("\n")
       if (new != "Not yet") {
         ty <- str_replace_all(tf, old, new)
         tf <- paste(tf, ty, sep="\n")
@@ -75,6 +76,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  fixed("[sim"),
+  "Accrual problems in clinical trials",
+  new="Accrual problems",
   dir_root="text", 
   file_pattern="md$")

### 2024-08-02 src/clean_files.R
Added _quarto.yml

@@ -76,7 +76,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "Accrual problems in clinical trials",
-  new="Accrual problems",
+  "category:",
+  new="categories:",
   dir_root="text", 
   file_pattern="md$")

### 2024-08-02 src/copy_files.R

@@ -14,8 +14,8 @@ source(file="src/standard_functions.R")
 
 copy_files <- function(
   old_path="text", 
-  new_path="c:/Users/simons/Documents/b04/content",
-  old_dir_list=c("99", zpad(0:24)),
+  new_path="../qblog",
+  old_dir_list=c("99", zpad(0:24), "tags"),
   file_pattern="md$",
   qc=FALSE) {
 
@@ -32,7 +32,7 @@ copy_files <- function(
     cat(message)
     cat("\n")
     
-    new_dir <- "post"
+    new_dir <- "posts"
     for (i_file in file_list) {
       verb <- ifelse(ok_to_copy,"copied", "to be copied")
       k <- k+1

### 2024-08-04 text/14/axis.Rmd
Fixed tags

@@
 date: "2014-01-01"
-category: Blog post
+categories: Blog post
 tags:

### 2024-08-04 text/15/difference-of-two-exponentials.Rmd
Fixed tags

@@
 date: "2015-10-05"
-category: 
+categories: 
 - Blog post

### 2024-08-04 text/22/definition-schoenfeld-residuals.Rmd
Fixed tags

@@
   html_document: default
-category: Blog post
+categories: Blog post
 tags:

### 2024-08-04 text/22/r-creatures.Rmd
Fixed tags

@@
 date: "2022-04-14"
-category: 
+categories: 
 - Blog post

### 2024-08-04 text/16/circle.Rmd
Changed circle to Rmd

---
title: "Drawing the perfect circle"
author: "Steve Simon"
source: "http://blog.pmean.com/circle/"
date: "2016-10-31"
categories: Blog post
tags:
- Cheat sheet
- R software
output: html_document
page_update: complete
---

I have an R cheat sheet, [How Big Is Your Graph][sim3], that explains how to measure the size of various features of your graph in R. This blog post illustrates how you can use some of the commands described in that cheat sheet to draw a perfect circle.

<!---More--->

When you try to draw a circle on a graph, it often ends up looking like an ellipse. Here's an example.

```{r}
draw.circle <- function(x0, y0, r, ...) {
  # draw a circle of radius r centered at x0, y0
  pi.seq <- seq(0, 2*pi, length=100)
  x.circle <- x0 + r * cos(pi.seq)
  y.circle <- y0 + r * sin(pi.seq)
  polygon(x.circle, y.circle, ...)
}
plot(0, 0, xlim=c(0, 20), ylim=c(0, 50), type="n", xlab=" ", ylab=" ")
draw.circle(5, 25, 5, lty="dotted")
text(5, 80, "r=5 (usr)")
```

![Circle in user coordinates](http://www.pmean.com/new-images/16/circle01.jpg){width="448" height="320"}

``` {.r}
dv <- dev.size()
ma <- par("mai")
us <- par("usr")
```

There are three things that are contributing to this problem.

1.  The default graph in R Markdown is rectangular. You can measure the size of a graph in inches with the **dev.size()** function: 7, 5

2.  The default margins are uneven with more margins on the top and bottom compared to the left and right sides. You can measure the size of the margins in inches with **par("mai")**: 1.02, 0.82, 0.82, 0.42. The size of the bottom margin is listed first, followed by the left, top, and right margins.

3.  The number of units on this graph are different for the x-axis and the y-axis. You can measure the minimum and maximum values on the x and y axes using **par("usr")**: -0.8, 20.8, -2, 52. Note that, by default, R adds an extra 4% to each end of the axis to reduce problems with clipping.

It's not necessarily bad to see an ellipse where you were expecting a circle. But if you do want a perfect circle, you need to calculate the size of the plotting region and adjust appropriately.

```{}
draw.circle <- function(x0, y0, r, adjust.units=function() {list(x=1, y=1)}, ...) {
  # draw a circle of radius r centered at x0, y0
  # with an adjustment in the horizontal and vertical dimensions
  # The default adjustment is no adjustment
  pi.seq <- seq(0, 2*pi, length=100)
  adj <- adjust.units()
  x.circle <- x0 + adj$x * r * cos(pi.seq)
  y.circle <- y0 + adj$y * r * sin(pi.seq)
  polygon(x.circle, y.circle, ...)
}

convert_to_inches <- function() {
  # This funciton adjusts from user coordinates to inches.
  # There is a short-cut: ratio <- par("cxy") / par("cin")

  # Range of the plotting region in user coordinates
  plot.range.usr <- c(par("usr")[2] - par("usr")[1], par("usr")[4] - par("usr")[3])
  if (verbose) cat("\nThe range of the plotting region is", plot.range.usr, "in user coordinates")

  # Range of the plotting region in inches
  plot.range.in <- par("pin")
  if (verbose) cat("\nThe range of the plotting region is", plot.range.in, "in inches")

  # Number of inches per user coordinate
  ratio <- plot.range.usr / plot.range.in
  if (verbose) cat("\nThere are", ratio, "user coordinates per inch")
  return(list(x=ratio[1], y=ratio[2]))
}

plot(0, 0, xlim=c(0, 20), ylim=c(0, 50), type="n", xlab=" ", ylab=" ")
draw.circle( 5, 25, 5, lty="dotted")
text( 5, 25, "r=5 (usr)")
verbose <- TRUE
draw.circle(15, 25, 0.5, convert_to_inches, col="red")
```

```{}
## The range of the plotting region is 21.6 54 in user coordinates
## The range of the plotting region is 5.759999 3.159999 in inches
## There are 3.75 17.08861 user coordinates per inch
```

```{}
text(15, 25, "r=0.5 (in)")
```

![Circle in user coordinates](http://www.pmean.com/new-images/16/circle02.jpg){width="448" height="320"}

The concept of inches in a graph is a loose one at best. The half inch radius shown above might appears slightly bigger or slightly smaller than a half inch and it will probably change when you move from one computer monitor to another. The size will probably change when you print this graph.

If you're lucky, the relative size of an inch will mean the same thing in the horizontal and vertical direction, and this will give you a perfect circle, though of an uncertain size. But there is no guarantee even of this. See, for example, [R FAQ 5.3 Circles appear as ovals on screen][cran1]

You may wish to specify the size of your circle in pixels instead. This may or may not be a more stable measure. To do this, you need a measure of pixels per inch.

You can also draw a circle using **pch=21** or **pch="o"** or **pch="O"**. The latter two choices are not quite perfect circles, but come pretty close. You can control the size of these circles using the pointsize and/or cex arguments. With **pch=21**, you can control the fill color with the bg argument.

```{}
plot(0, 0, xlim=c(0, 20), ylim=c(0, 50), type="n", xlab=" ", ylab=" ")
co <- c("black", "blue", "red", "green")
points(rep( 5, 4), 10*(1:4), pch=21,  cex=1:4, col=co, bg=rev(co))
points(rep(10, 4), 10*(1:4), pch="o", cex=1:4, col=co)
points(rep(15, 4), 10*(1:4), pch="O", cex=1:4, col=co)
```

![Circle in user coordinates](http://www.pmean.com/new-images/16/circle03.jpg){width="448" height="320"}

Notice the subtle differences in shape and thickness.

If you try this code on your system, your results may not match the results you see here, for a wide variety of reasons that are impossible to list. Graphics are often the result of trial and error, and if you don't get the results you want right away, keep plugging away.

You can find an [earlier version][sim1] of this page on my [blog][sim2].

[sim1]: http://blog.pmean.com/circle/
[sim2]: http://blog.pmean.com

[sim3]: http://blog.pmean.com/cheatsheets

[cran1]: https://cran.r-project.org/bin/windows/base/rw-FAQ.html#Circles-appear-as-ovals-on-screen

### 2024-08-04 text/20/reading-text-files.Rmd
Removed brackets from generic code chunks

@@
 
-```{}
+```
 ![A non-text file displayed in notepad](../../../web/images/reading-text-files01.png)
@@
 
-```{}
+```
 ![Data set on movies](../../../web/images/read-text-files03.png)
@@
 
-```{}
+```
 ![Data dictionary](../../../web/images/read-text-files06.png)
@@
 
-```{}
+```
 ![Example of a space delimited file](../../../web/images/read-text-files02.png)
@@
 
-```{}
+```
 ![Dataset showing COVID-19 deaths over time](../../../web/images/read-text-files04.png)
@@
 
-```{}
+```
 ![Dataset of revenues for selected movies](../../../web/images/read-text-files05.png)

### 2024-08-04 src/clean_files.R

@@ -76,7 +76,7 @@ verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 clean_files(
-  "category:",
-  new="categories:",
+  fixed("```{}"),
+  new="```",
   dir_root="text", 
   file_pattern="md$")

### 2024-08-04 text/18/teaching-and-research.Rmd
Fixed tricky-program

---
title: "Pmean: My teaching and research statement"
author: "Steve Simon"
source: "http://blog.pmean.com/teaching-and-research/"
date: "2018-04-02"
categories:
- Blog post
tags:
- Professional details
output: html_document
page_update: partial
---

I am applying to a variety of jobs and some of them ask for a statement
about teaching or research. Here's something I wrote where the employer
asked for a combination of the two.

<!---More--->

## Teaching

I have taught in a wide range of non-traditional formats: short courses
at regional, national, and international conferences, and webinars for a
geographically diverse audience. Students in these classes have a choice
and will not tolerate a poorly taught class. I take it as validation of
my teaching quality that I have been re-invited over and over again to
give talks by the same organizations (American Society of Andrology, The
Analysis Factor, International Research Conference on Complementary
Medicine, Medical Library Association, Midwest Society of Pediatric
Research).

I don't have (and don't believe in) an overarching philosophy of
teaching. Good teaching comes from the little things, and I attribute my
teaching success to three little things: finding compelling examples,
using humor to make a point, and spending a large portion of my limited
teaching time on small group exercises. I also want to mention some of
the special challenges that I have faced with teaching in an online
(webinar) format.

## Compelling teaching examples

It takes a lot of time to find compelling teaching examples. They need
to be vivid and memorable. They need to reflect the practice of
statistics in the real world, but at the same time they need to avoid
obscure details that only specialists in an area can follow.

In a book chapter that I wrote for Big Data Analysis for Bioinformatics
and Biomedical Discoveries, I was faced with a challenge to find a
compelling data set to illustrate the R programming language in action.
Most data sets used in Bioinformatics, however, are narrowly focused and
require a fairly specialized knowledge to fully comprehend. I chose a
data set that had broader appeal and which was easily followed by anyone
with a reasonable medical background, Son et al. Database of mRNA gene
expression profiles of multiple human organs, Genome Res 2005.
15:443-450. This study took tissue samples from 19 different organs of
30 people (158 samples total) and fed them to a DNA microarray. This
provided expression levels of almost 19 thousand genes. This made the
data set large enough to qualify for a big data analysis. More
importantly, the comparison was concrete and easy to discuss. How does
gene expression differ in the heart versus the liver, or in the pancreas
versus the spleen? Also important to me was that the data used in Son et
al was freely available for download at the journal's website.

Another compelling data set is mortality on the Titanic, a large ocean
liner that in 1912 struck an iceberg and sunk. One of the first things
you might do with this data set is to calculate a two by two table of
gender versus survival. The Titanic sunk during an era where society
really did believe in the mantra "women and children first" though
perhaps more so among first and second class passengers. The
crosstabulation of gender and survival would look something like this.

```{r titanic, echo=FALSE}
d <- list(c("Female", "Male"), c("No", "Yes"))
ti <- matrix(c(154, 709, 308, 142), 2, 2, dimnames=d)
print(ti)
```

You can bring these numbers to life by pointing out that Kate Winslet
was in the upper right corner of the 308 women who survived and that
Leonardo DiCaprio, sadly, was in the lower left corner of the 709 men
who died. The data set has an interesting feature: the relative risk of
death is 2.5 times higher for males than females, but the odds ratio is
much higher, almost 10. It also allows you to study a rather complex
interaction of gender and passenger class. I've used the Titanic data
set in a tutorial article about measures of risk (Simon SD.
Understanding the odds ratio and relative risk. J Androl. 2001.
22(4):533-6), in a short course on logistic regression, and for some of
the homework assignments in the Introduction to R, Introduction to SAS,
and Introduction to SPSS classes. I take as validation of this approach
the fact that other groups (most notably Kaggle) have also adopted the
Titanic as an instructional data set.

Compelling examples are not limited to data sets. In my book,
Statistical Evidence in Medical Trials, I talk about blinding with
specific attention to blinding in surgical trials. Surgery is easily
understood: you cut someone open, take something out, and sew them back
up again. But it requires special challenges to ensure blinding, such as
the use of extra large bandages to cover up the size of the incision. I
make special reference to a study of Parkinson's Disease where fetal
cells were injected directly into the brain. The study had a control
group, and in order to maintain the blind, the control patients also had
their heads shaved, were put under an anesthetic, and the surgeons
actually drilled into their skulls. The only difference was that nothing
was injected after drilling if you were in the control group. Needless
to say, this was a highly controversial study with one ethical expert
writing a critical article with the title "I Need a Placebo Like I Need
a Hole in the Head." The gruesome thought of skull drilling makes this
example memorable.

There is no more compelling example relating to research methodology
than the efforts to find a treatment for AIDS. Controversies surrounding
AIDS research has changed the way we approach research today. When I
lecture about surrogate outcomes, it is easy enough to come up with
examples where reliance on surrogate outcomes has led us astray. The
classic example is the CAST trial which showed that the use of
anti-arrythmic drugs in patients with heart rhythm problems actually led
to an increase in mortality versus placebo. It is harder to find
examples where reliance on surrogate outcomes has helped us. AIDS
trials, however, do clearly illustrate the value of surrogate outcomes.
In the first studies of anti-retroviral treatments for AIDS patients,
scientists originally advocated the use of "hard" endpoints like
mortality and opportunistic infections. But AIDS advocates pushed for
surrogate outcomes like changes in CD4 cell counts because hard
endpoints took too long and required waiting until you had accumulated a
sufficient number of bad events in the untreated group. It is clear that
reliance on surrogate outcomes reduced the number of deaths in the
clinical trial itself and allowed much earlier FDA approval of these
treatments.

## Using humor to make a point

My audiences are often apprehensive. Will this guy speak a lot of Greek
and formulas to me? A bit of humor very early in the talk will often
allay some of those fears. I try hard, however, to use humor as well to
make a point. Humor comes from an exaggeration of a basic idea to levels
of absurdity, which you can use profitably to emphasize your basic idea
later in your talk.

In a lecture on sample size justification for a short course on grant
writing, I started with a (fictional) story of a researcher who gets a
six year, ten million dollar NIH grant and writes up a report
summarizing the research saying "This is a new and innovative surgical
approach and we are 95% confident that the cure rate is somewhere
between 3% and 98%." This becomes a repeated touchstone for the rest of
the talk. The agency that you are seeking to get grant money from wants
some assurance that the results will be informative when you finish your
work, and not a confidence interval that goes from 3% to 98%.

The previous joke is an original of mine, but I'm not above stealing
other people's jokes. There is a classic about two statisticians
travelling on an airplane and one engine after another explodes. After
each explosion, the pilot announces that everything is okay, but the
flight is delayed further and further after each explosion. After the
third explosion, one statistician turns to the other and say "Boy I hope
that last engine doesn't explode" [dramatic pause] "or we'll be up
here forever!". I use this joke when I teach linear regression, because
it illustrates a dangerous extrapolation beyond the range of observed
data. Later, I introduce the intercept term and offer an interpretation
(the estimated average value of Y when X equals zero). Then I point out
that such an interpretation is a dangerous extrapolation when you have
no data near the value X=0. In fact,the estimated travel time of a jet
with X=0 engines is also an intercept term requiring the exact type of
dangerous extrapolation.

I also use humor to assess my audience as well. At the very start, I'll
tell a corny joke about degrees of freedom (a joke so bad that I won't
repeat it here). I look at who laughs and who doesn't. When most of my
audience laughs, I compliment them by telling them that if you
understood that joke, you won't have any problems with the rest of my
lecture. When most of my audience is quiet, I tell them that they are
going to learn a lot today (and I teach things a bit more slowly).

Humor can be overdone, and students are not listening to you for
entertainment. I try to keep my jokes short and get them out of the way
early. But a little bit of humor does seem to help a lot.

## Spending time on small group exercises

I hate to do small group exercises. I really hate them. When you are
giving a short course at a research conference, you have a limited time
frame, usually three to four hours. Small group exercises can easily eat
up a third to a half of that time. I'd much rather be talking because
there's so much to teach and so little time. Even so, I include small
group exercises in almost all my short courses. The feedback I have
gotten from students has been that the small group exercises are the
best part of the class.

Constructing a small group exercise is not easy. You have to find a
problem that students can tackle, so it can't be too hard, but it has to
be challenging enough to force them to use some of the things they have
just learned. It has to be something that everyone in the group feels
that they can contribute. And it has to be something that each small
group can summarize to the entire class in five minutes or less.

The reason that small group exercises work is that students have a
variety of different experiences and can learn from each other. The more
experienced students benefit because they have to understand the
concepts at a higher level in order to articulate these concepts to the
rest of the group. The less experienced students benefit because they
get a chance to hear about the same concepts a second time and from a
different perspective.

In the short course on grant writing, we gave each group a different
research paper (actually, just the abstract from the paper) and asked
them to suggest a new study with a different research design that builds
on the paper's findings and that an agency might want to fund. Showing
just the abstract is a big no-no in Evidence Based Medicine, but we
apologized and explained that if we handed them a full paper they would
run out of time before they even finished the research paper. Each group
then appointed a spokesperson to summarize the original paper (just the
PICO: Patient group, Intervention, Comparison group, and Outcome
measure) and then propose their new research design. Once the
spokesperson was done, we asked if anyone else in the small group wanted
to add anything and opened it up for comments and questions by the other
students. Getting our students to visualize a new research study was one
of the big successes of our short course.

In a short course on setting up an independent statistical consulting
practice, I was actually able to get two small group exercises squeezed
into a four hour format. The first exercise was pairing up students and
asking them to share with their partner where they hoped to be five
years from now. This served as an ice breaker and got students to think
about what a career as a consultant might mean for them. The second
small group exercise took some fictional accounts (loosely based on
reality) of bad client interactions. The groups had to decide on the
best course of action, and this quite honestly might involve just
walking away from the project completely, an option that is always
available to an independent consultant. The short time frame did not
allow each group to summarize their findings to the rest of the class,
but the students still found the small group exercises to be valuable.

In a short course, Statistics for Medical Librarians, that I have taught
in an online format and at multiple regional and national meetings of
the Medical Library Association, I used small group exercises to dissect
statistics found in actual peer-reviewed publications. I teach concepts
like confidence intervals, provide a non-technical interpretation, and
then ask students to do the same in small groups. I show them actual
papers (actually abstracts again to save time) that show how a
confidence interval might look "in the wild." Then each group has to
summarize the abstract using a PICO format and then interpret the
highlighted confidence interval in that abstract. All of the abstracts
that I choose are from Open Source journals so the librarians can track
down the full articles after the class ends, if they are so inclined.

## Online teaching

I wanted to mention a bit about the special challenges associated with
online teaching, because I see a transition to online teaching almost
everywhere I look. I have done webinars, a form of online training, for
several several organizations. I've learned a lot from these webinars,
though they are not the same as teaching a semester long class in an
online format.

One lesson I have learned is that you have to work much much harder at
establishing a rapport with your students. In every webinar I have
taught, I do not have the option of requiring the students to make
themselves visible on a webcam, and only in a few did I have the option
of having students ask questions orally rather than through a chat box.
This makes the webinars less fun, quite frankly, than a short course
taught to a live audience. But I have embraced the webinar format
because it allows the participation of a large number of students who
are otherwise geographically isolated.

To make up for not seeing and often not hearing the students, I try to
make myself seen--using my own webcam when possible. I also take extra
time to more directly encourage student participation. It starts with a
joke. I introduce the joke with an exhortation. It is difficult, I point
out, to say something funny and then hear total silence on the other
end, so I after I tell this joke, I want you to type something in the
chat box. A "haha" if you liked the joke or a "groan" if you hated it.
Then I read the chorus of responses after the joke. It reinforces the
connection between me and my audience and it lowers the barrier for
students to use the chat box later for more serious comments.

When you ask for questions during a webinar, you need to wait long
enough for people to type their questions. And you need to follow up
your answer with another exhortation "Did that make sense" because you
don't get the feedback from the knowing nods or the glazed eye
confusion.

A webinar format demands a good Powerpoint presentation. I do not always
use Powerpoint for live talks and short courses and subscribe to the
admonition of many that "Powerpoint is Evil." For a webinar, however, if
you don't give students a Powerpoint slide to look at, it makes it that
much harder to keep a student's attention.

I've taught two traditional courses (Introduction to R and Introduction
to SPSS) in an online format. These are in an asynchronous format.
Students listen to mp4 files where you lecture and show Powerpoint
slides and live screen shots of R and SPSS as those programs grind
through their computations. The one issue I've noticed with the
asynchronous format is that if one student asks a question, the other
students would not normally get to benefit from hearing that question
and hearing my response. They also lose out on the opportunity to add
something to my answer if they can. So it is very important to share
questions and answers that otherwise would only come to you privately
via email.

Neither Introduction to R nor Introduction to SPSS lends itself well to
small group exercises, but I do recognize by watching others at my
current job that small group exercises take a much different form in an
online format. In particular, they require much more written
communication than oral communication, and the effort, especially for an
asynchronous format, is spread out over a longer time frame. This can
actually be an advantage because you need to explain yourself more
clearly in a written format and the slower pace encourages more thought
as well.

No doubt that I will encounter other issues as I teach more classes in
an online format. I learned a lot about online teaching at various talks
at the Joint Statistical Meetings, and plan to keep my eyes and ears
open for other people's experiences.

## Research

My best research efforts have been collaborative. I feel like these
efforts represent something that neither I nor my co-authors could have
developed by ourselves. I've had many wonderful research collaborations
over the years. I want to highlight two of them: patient accrual in
clinical trials and with mining information from the electronic health
record. I also need to describe my efforts to help others be successful
in their research endeavors.

## Patient accrual

In 2006, I gave a journal club presentation on how to use control charts
to monitor the process of patient accrual in a clinical trial. Too many
studies, I hate to say, fail to meet their sample size requirements
because researchers grossly underestimated the amount of time it would
take to recruit patients. I discussed this control chart in the context
of a clinical trial, but it really applies to any prospective research
study that collects data from human volunteers.

Byron Gajewski, one of the other faculty members at the journal club,
suggested that this problem might be better solved with a Bayesian
approach. That turned my research around 180 degrees but it was worth
it. His suggestion led to a very profitable avenue of research for the
two of us.

The beauty of a Bayesian approach is that it requires the specification
of a prior distribution. The prior distribution represents what you know
and what you don't know about how quickly volunteers will show up on
your doorstep asking to join your study. You can think of the prior
distribution as a way to quantify your ignorance. You choose a very
broad and variable prior distribution if you know very little about
accrual patterns. This might be because you're new to the area, you're
using a novel recruiting approach, and/or there's very little experience
of others that you can draw upon. You choose a very narrow and precise
prior distribution if you've worked on this type of study many times
before, your recruiting techniques are largely unchanged, and/or there's
lots of experience of others that you can draw upon. What you don't do,
even if you are very unsure about accrual, is to use a flat or
non-informative prior. A flat prior would be like admitting "I don't
know: the study might take ten weeks or it might take ten years and I
think both possibilities are equally likely." Someone with that level of
ignorance would be unqualified to conduct the research.

The very act of asking someone to produce a prior distribution will
force them to think about accrual, and that in and of itself is a good
thing. But the advantage of specifying a prior shows during the trial
itself. As the trial progresses you get actual data on the accrual rate,
and you can combine that with your prior distribution, as any good
Bayesian would, to get an updated estimate of how long the trial will
take. Here is where the precision of the prior distribution kicks in. If
you have a broad prior with high variance, then even a little bit of bad
news about accrual during the trial itself will lead to a drastic
revision in your estimated time to complete the trial. You will act
quickly, either by adding extra centers to a multi-center trial, hiring
an extra research co-ordinator to beat the bushes for volunteers, or (if
the news is bad enough) cutting your losses by ending the trial early
for futility. If you have a narrow prior with low variance, then you've
done this trial often enough that you don't panic over a bit of bad
news. If the data keeps coming in and it shows a much slower accrual
rate than you expected, then you will eventually reach a point where you
need to take action. But there's a cost associated with a premature
overreaction that a precise prior will protect you against.

Dr. Gajewski put one of his graduate student, Joyce Jiang, on the trail,
and she contributed several additional publications after completing a
successful dissertation defense of her extensions in this area. I worked
very closely with Drs. Gajewski and Jiang, and found an interesting
theoretical contribution to Bayesian data analysis that was hidden in
their work.

One of the problems with getting researchers to produce a prior
distribution is that they sometimes are wrong--spectacularly wrong. If
you have a strong prior attached to a prior that is sharply at odds with
the actual accrual data, you'd like to find a way to discount that prior
distribution, but you'd like to keep that strong prior for the precision
it gives you when the prior and the actual accrual data agreed with one
another. They came up with a very clever solution. Attach a hyperprior
to the precision of the prior distribution. If the accrual data and the
prior are in sync, the precision stays high. But if there is a serious
discrepancy between the accrual data and the prior, the hyperprior
shifts and leads to a much weaker prior distribution.

I dubbed the method they proposed the hedging hyperprior, and suggested
that it might work in other Bayesian settings as well. It turns out to
be equivalent to the modified Power prior proposed by Yuyan Duan in
2006, but the formulation of the hedging hyperprior is both simpler and
more intuitive. I have presented a simple example applying the hedging
hyperprior to the beta-binomial model and am preparing a manuscript for
publication.

The work that Dr. Jiang did on her dissertation was not just limited to
the accrual problem but included an additional Bayesian application to a
validation model using expert opinion. The strength of her work in these
two areas led to her appointment to a post-doctoral fellowship at Yale
University. Dr. Gajewski and I continue to collaborate with her on these
models. We have four peer-reviewed publications and an R package so far,
and plan to collaborate with other researchers in this area.

Closely related to my research on patient accrual is an effort to audit
the records of Institutional Review Boards (IRBs). Too often,
researchers fail to obtain the sample size that they promised in the
original research protocol, mainly because subject recruitment takes
longer than expected. It is very easy to compare the protocol submitted
to the IRB to the final report. In the study of 135 submissions to one
IRB, more than half of the researchers failed to reach their enrollment
targets and the average shortfall was more than 50%. I have approached
many other IRBs to ask to replicate this work, but none have shown any
interest. But I plan to continue to ask anyone who works on an IRB to
help me.

## Mining the electronic health record

In January 2016, I was offered the opportunity to work on a research
grant funded by the Patient Centered Outcomes Research Institute.<U+00A0> The
grant supported the Greater Plains Collaborative, a consortium of ten
academic health care centers in the midwest. It was run out of
Enterprise Analytics, located at Kansas University Medical Center. I
jumped at the chance and dropped much of my other work to focus on this
grant.

My assignment, derived through discussions with the head of Enterprise
Analytics, Russ Waitman, was to develop a develop a phenotype of breast
cancer from information in the electronic health record (EHR) and
validate it against information in the breast cancer registry.

Such a phenotype would have great value in identifying patients for
prospective clinical trials. The advances in high throughput genome
sequencing and the linkage of that information with the EHR allows for
exploration of novel precision medicine options. Developing a phenotype
from the EHR is fraught with peril because information in the EHR on
basic issues like diagnoses and treatments is often coded
inconsistently. Having a link between the EHR and a tumor registry
provides an external validation of the accuracy the EHR phenotype.

The EHR at KUMC (and the other sites in the Greater Health
Collaborative) is stored in an Oracle database and is accessible through
an i2b2 system that is very easy to use. My work, however, required
access to the full database as well as some of the metadata. With the
help of Dr. Waitman and the SQL expert sitting across from me, Dan
Connolly, I was able to pull information directly from Oracle.

I used a big data model, LASSO regression, to predict whether a patient
was in the breast cancer tumor registry and set up sparse matrices as
input to better manage the size of the data sets. The breast cancer
cases were compared against three separate control groups and in spite
of the massive size of the independent variable matrix (more than 45,000
columns), this model ran in under ten minutes. The resulting sensitivity
and specificity were very high, putting to rest concerns that the EHR
data might be too incomplete or inconsistent to produce an accurate
phenotype. The LASSO regression model could easily be run for other
tumor types, and just as quickly validated. I have presented these
results at a local research conference and plan to submit a
peer-reviewed publication soon.

An interesting side effect of this research is also worth mentioning. I
had a passing knowledge of SQL prior to my work with Dr. Waitman and Mr.
Connally, but I had to quickly learn and apply a broad range of SQL
programming to get the data into R and the LASSO regression model. SQL
is a fairly easy language to learn, but many of the students in the
Bioinformatics program at UMKC do not have even a cursory knowledge of
SQL. So I am partnering with a database expert at UMKC to develop a new
class, Introduction to SQL, that will cover some of the basic skills
that an researcher would need to query data stored in a relational
database. This class is not intended to teach someone to become a
database administrator, but rather a competent user of other people's
databases. It will parallel to a large extent classes that I have
already helped develop and teach: Introduction to R, Introduction to
SAS, and Introduction to SPSS.

My work on the PCORI grant has been transferred a different grant and I
plan to work with partners at Truman Health Center and Children's Mercy
Hospitals to develop more research utilization of EHR at their
locations. I also have been asked to help develop an analytics platform
simplifies data mining of the EHR through a standardized library of
functions interfacing SQL databases and R. This library would pull
appropriate meta data descriptors as well, expanding the types of
analyses available to the end user.

## Helping others with research

I am a great believer in the Harry Truman quote "Anything is possible if
you don't care who gets the credit." A major portion of my career has
been helping others become successful researchers. This work is quiet,
behind the scenes, and often leads to very little recognition for me.
But I enjoy watching someone developed from a scared and timid person
starting out with their very first research study to someone who has
learned enough that now he/she is mentoring others.

A large part of my work is helping people who are struggling in their
graduate studies. It might be some extra tutoring for that seemingly
impossible statistics class. More often, though, it is guiding students
through the difficult process of writing and defending their
dissertation. I did this for free for doctors, nurses, and other health
care professionals that I worked with at Children's Mercy Hospitals and
Clinics. After I left that job, I started my own consulting business,
and I got lots of referrals to graduate students. They typically are
struggling with their disserations and with a dissertation committee
that was not giving adequate direction on the data analysis. For a
dissertation, you can't do the data analysis for them because they have
to be able to explain their work during the dissertation defense. You
have to teach them those things that they didn't pick up in their
earlier statistics class and teach them so thoroughly that they can
offer a clear and convincing presentation of the analysis that they ran.
You have to help them anticipate the types of questions that they might
get and how best to answer those questions. Perhaps the most important
thing is to get them a sense of self-confidence that they are working on
an important problem and that they have a solid and defensible plan for
solving that problem.

Another big portion of my work is helping people write their first
grant. The first grant is almost always for an amount too small to
support a statistician as a co-investigator. But these are the grants
that need to most help and support. I help with the research design, the
sample size justification, and the data analysis plan. But my work is
not just limited to that. If the specific aims are vague, if the
literature review rambles incoherently, or if the research budget is
unrealistic, I offer gentle suggestions to fix these problems. I try
very hard not to overstep my bounds; I don't know the science as well as
they do. But I can often provide valuable feedback by looking at things
from the perspective of an outsider who has seen hundreds of grants in
dozens of different scientific fields.

I've taken many classes in grant writing to better understand the whole
process and to improve my ability to work with new researchers. But
recently I have taught the classes myself. In 2012, I co-developed a
class in grant writing for researchers in Complementary and Alternative
Medicine with a prominent statistician working for what was then called
the National Center for Complementary and Alternative Medicine and with
another statistician working at a major college of Chiropractic
Medicine. We repeated this class in 2014 for the same conference. I
provided lectures on designing a pilot study, justifying your sample
size, and what a scientist should look for in a collaborating
statisticians. I gained the most benefit, however, by listening to the
lectures of the other statisticians. One of the best was a discussion
about insuring consistency between the specific aims, the research
design, the data analysis plan, and the budget. Another excellent talk
was on hiring a data manager and preparing a solid data management plan
(including budgets!).

When you include the small internal grants that provide seed money for
new researchers, I have helped write hundreds of grants, too many to
track. I do list those grants where I am on the budget, but these
represent just the tip of the iceberg.

Finally, I help new researchers navigate the daunting process of getting
Institutional Review Board (IRB) approval for their studies. At
Children's Mercy Hospital and Clinics, I was housed very close to the
people who ran the IRB and got to know them very well. It would have
been a conflict of interest for me to serve on the IRB, because I would
be reviewing protocols that I helped write. But I did work extensively
with the IRB, answering technical questions, accepting referrals from
researchers whose protocols had glaring problems with scientific rigor,
and providing training courses on ethical issues associated with
research. Statisticians, I believe, are key players in assuring the
ethical conduct of research. Often researchers are barred from the
optimal research design by ethical constraints and our job is to help
find an alternative design that meets the needs of the researcher while
still protecting the rights of the research volunteers.

### 2024-08-05 src/clean_files.R
Fixed names that duplicate tags

@@ -75,8 +75,9 @@ verbose <- TRUE
 verbose <- FALSE
 ok_to_replace <- TRUE
 ok_to_replace <- FALSE
+
 clean_files(
-  fixed("```{}"),
-  new="```",
+  fixed("- Big data"),
+  new="- Data science",
   dir_root="text", 
   file_pattern="md$")

### 2024-08-08 text/pending/wavelets.Rmd
Updated references-from-jsm-2024

--- 2024-07-07 text/24/wavelets.Rmd
@@
 date: "2014-01-01"
-category: 
+categories: 
 - Blog post

### 2024-08-15 text/04/early-stopping-animal.Rmd

@@
 
-```{}
+```
 par(mar=c(4,5,0,0)+0.1,cex=2,pch="+")

### 2024-08-15 text/15/missing-values-in-r.Rmd

@@
 date: "2015-02-07"
-category:
+categories:
 - Blog post

### 2024-08-15 text/16/circle.Rmd

@@
 
-```{}
+```
 draw.circle <- function(x0, y0, r, adjust.units=function() {list(x=1, y=1)}, ...) {
@@
 
-```{}
+```
 ## The range of the plotting region is 21.6 54 in user coordinates
@@
 
-```{}
+```
 text(15, 25, "r=0.5 (in)")
@@
 
-```{}
+```
 plot(0, 0, xlim=c(0, 20), ylim=c(0, 50), type="n", xlab=" ", ylab=" ")

### 2024-08-15 text/21/read-acs-data.Rmd

@@
 key <- Sys.getenv("CENSUS_API_KEY")
-census_api_key("11bafd82737f7d75b3305ada1e772501d0eab002")
+census_api_key(key)
 tarr <- get_acs(geography = "tract", variables = "B19013_001",

### 2024-08-15 text/22/analysis-of-means-example.Rmd

--- 2024-07-06 text/22/analysis-of-means.Rmd
@@
   html_document: default
-category: Blog post
+categories: Blog post
 tags: 

### 2024-08-15 text/22/different-parameterizations.Rmd

@@
 date: "2022-07-06"
-category: 
+categories: 
 - Blog post

### 2024-08-15 text/22/mapping-longitude-latitude.Rmd

@@
 date: "2022-01-06"
-category: 
+categories: 
 - Blog post

### 2024-12-13 text/23/too-many-colors.rmd
Updated 23/too-many-colors

@@
     data = statepov, 
-    values = "pct_pov_2014", 
+    values = "pct_pov_2021", 
     color = "red") -> us
@@
         high = "red", 
-        name = "Poverty rate (2014)", 
+        name = "Poverty rate (2021)", 
         label = scales::comma) + 

### 2024-12-13 src/clean_files.R

@@ -77,7 +77,6 @@ ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 
 clean_files(
-  fixed("- Big data"),
-  new="- Data science",
+  fixed("avelet"),
   dir_root="text", 
   file_pattern="md$")

### 2024-12-13 src/clean_files.R
Added corgis-datasets

@@ -77,6 +77,6 @@ ok_to_replace <- TRUE
 ok_to_replace <- FALSE
 
 clean_files(
-  fixed("avelet"),
+  "- [Dd]atasets$",
   dir_root="text", 
   file_pattern="md$")

### 2024-12-14 src/copy-files.qmd
Added sql-versus-r

---
title: "Copy files from build-website to qblog"
author: "Steve Simon"
format: html
---

This program was written by Steve Simon on 2024-11-09 and is placed in the public domain.

## Preliminaries

```{r}
#| label: setup
#| message: false
#| warning: false

library(glue)
library(tidyverse)
```

## Set old and new locations

```{r}
#| label: locations
old_path <- "../text"
old_directories <- c("99", paste0("0", (0:9)), 10:24)
new_path <- "../../qblog/posts"
file_pattern <- "md$"
```

## Find files function

```{r}
#| label: find-files
files <- NULL
subdirectories <- NULL
for (dir in old_directories) {
  glue("{old_path}/{dir}") |>
    list.files() |>
    str_subset(file_pattern) -> to_be_copied

  n_to_be_copied <- length(to_be_copied)
  subdirectories <- append(subdirectories, rep(dir, n_to_be_copied))
  files <- append(files, to_be_copied)

  message <- glue(
    "Searching through {n_to_be_copied} files in {old_path}/{dir}")
  cat(message)
  cat("\n")
}

files |>
  str_remove("\\..*") -> roots

files |> 
  str_remove("^.*\\.") -> extensions
```

```{r}
#| label: create-file-names
new_paths <- glue("{new_path}/{roots}")
new_files <- glue("{new_paths}/index.{extensions}")
old_files <- glue("{old_path}/{subdirectories}/{roots}.{extensions}")
```

```{r}
#| label: move
n_files <- length(files)
for (i in 1:n_files) {
  if (!dir.exists(new_paths[i])) {dir.create(new_paths[i])}
  test1 <- file.copy(from=old_files[i], to=new_files[i], overwrite=TRUE)
  cat(glue("{test1}: {old_files[i]} copied to {new_files[i]}"))
  cat("\n")
  cat(list.files(new_paths[i]))
  cat("\n")
}
cat(str_glue("Total files copied: {n_files}"))
```

### 2024-12-14 text/23/what-is-a-partial-correlation.Rmd

@@
 library(tidyverse)
-fn <- "https://raw.githubusercontent.com/pmean/classes/master/biostats-1/data/data-05-fev.csv"
+fn <- "https://raw.githubusercontent.com/pmean/data/refs/heads/main/files/fev.csv"
 read_csv(

### 2025-02-09 text/pending/pretty-functions.Rmd
Added new-quarto-program to pending

---
title: "New RMarkdown program"
author: "Steve Simon"
date: "`r Sys.Date()`"
output: html_document
---

Most of the statistical summary functions in R produce plain vanilla output. That's actually good because it allows you to customize things the way you like. I have been finding that I use pretty close to the same customizations over and over again, so I thought I should standardize them in a few simple R functions.

<!---more--->

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(broom)
library(glue)
library(palmerpenguins)
# data(package = 'palmerpenguins')
library(tidyverse)
```

## pretty_p

The pretty_p function represents my efforts to avoid displaying small p-values using scientific notation. Some researchers get confused when they see something like 7.93e- 2. Even for those researchers who are used to scientific notation, it just makes things hard to read. This function replaces very small p-values with "p < 0.001" and rounds larger p-values to two significant figures.

```{r}
#| label: function-1

pretty_p <- function(p_table) {
  # p_table is the output from the tidy function in the broom package
  p_table |>
    mutate(p.value =
      case_when(
        p.value <  0.001 ~ "p < 0.001",
        p.value >= 0.001 ~ glue("p = {signif(p.value, 2)}")))
}
```

#### Comment on the code

The [mutate][ref01] function modifies an existing variable or creates a new variable. The [case_when][ref02] function uses logic statements to assign values. Both are part of the [dplyr][ref03]/[tidyverse][ref04] libraries. The [signif][ref05] function, part of the base R package rounds a value to a specified number of significant digits. The [glue][ref06] function, part of the [glue][ref07] package, will insert variables inside of a string by surrounding the variable by curly brackets.

[ref01]: https://dplyr.tidyverse.org/reference/mutate.html
[ref02]: https://dplyr.tidyverse.org/reference/case_when.html
[ref03]: https://dplyr.tidyverse.org/
[ref04]: https://www.tidyverse.org/
[ref05]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/Round.html
[ref06]: https://glue.tidyverse.org/
[ref07]: https://glue.tidyverse.org/articles/glue.html

#### Example

Here is an example of how you might use the pretty_p function. I am using the [Palmer Penguins][ref08] dataset. Here is the output from tidy without pretty_p.

[ref08]: https://allisonhorst.github.io/palmerpenguins/

```{r}
lm(flipper_length_mm ~ species + island, data=penguins) |>
  tidy()
```

Here is the output from tidy with pretty_p.

```{r}
lm(flipper_length_mm ~ species + island, data=penguins) |>
  tidy() |>
  pretty_p()
```

## pretty_n

I also like to replace counts with percentages, but show the numerator and denominator in parentheses afterwards. This function will take results from the count function (or from any dataset with a variable named "n"), compute a total and percentage, and then display the results in a nice format.

```{r}
pretty_n <- function(n_table) {
  # n_table is output from the count function
  n_table |>
  	mutate(total=sum(n)) |>
  	mutate(pct=round(100*n/total)) |>
	  mutate(pct=glue("{pct}% ({n}/{total})")) |>
	  select(-n, -total)
}
```

#### Comment on the code

The mutate and glue functions are described above. The round function is described in the same link as the signif function. The [select][ref09] function will keep or drop columns of data.

[ref09]: https://dplyr.tidyverse.org/reference/select.html

#### Example

Here is an example of how you might use the pretty_n function. Here is the output from count without pretty_n.

```{r}
penguins |>
  count(species)
```

Here is the output from count with pretty_n.

```{r}
penguins |>
  count(species) |>
  pretty_n()
```

## pretty_mean

For continuous variables, I usually like to compute a few descriptive statistics: the mean, standard deviation, minimum, maximum, and the number of missing values. This is easy enough to do but the code is long and tedious. The pretty_mean function is just as long and tedious (if not more so), but once you enter it, you can save a lot of time starting with the second variable that you need to summarize. I actually used to like the [summary][ref10] function, but it does not include a standard deviation and it does not work well with the [group_by][ref11] function in dplyr.

```{r}
pretty_mean <- function(d, v) {
  d |>
    summarize(n_missing=sum(is.na({{v}}))) -> d1
  d |>
    filter(!is.na({{v}})) |>
    summarize(
      across({{v}}, 
        list(
          mean=mean, 
          sd=sd,
          min=min,
          max=max))) |>
    bind_cols(d1)
}
```

### 2025-02-09 text/25/pretty-functions.Rmd
Finalized pretty-functions

--- 2025-02-09 text/pending/pretty-functions.Rmd
@@
 ---
-title: "New RMarkdown program"
+title: "Functions to make plain R output look nicer"
 author: "Steve Simon"
-date: "`r Sys.Date()`"
+source: new
+date: "2025-02-09"
+categories:
+- Blog post
+tags:
+- R programming
 output: html_document
+page_update: complete
 ---
@@
 
+[ref10]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/summary.html
+[ref11]: https://dplyr.tidyverse.org/reference/group_by.html
+
 ```{r}
 pretty_mean <- function(d, v) {
+  # d is a data frame or tibble, v is a variable in d.
   d |>
@@
 ```
+
+#### Comment on the code
+
+Using tidyverse functions [inside a loop or function][ref12] is tricky.
+With an argument that represents a variable name, you either enclose the argument in double curly braces ("embrace" is the term they use) or you specify the argument inside of .data[[]].
+
+The [across][ref13] function, part of the dplyr/tidyverse libraries, allows you to select specific variable(s) for further operations within dplyr functions like summarize.
+I have had some difficulty with this function, but it seems to work well here.
+
+The [bind_cols][ref14] function, part of the dplyr library, combines two data frames or tibbles side by side.
+
+[ref12]: https://dplyr.tidyverse.org/articles/programming.html
+[ref13]: https://dplyr.tidyverse.org/reference/across.html
+[ref14]: https://dplyr.tidyverse.org/reference/bind_cols.html
+
+#### Example
+
+Here is the code you would write without the pretty_mean function.
+
+```{r}
+penguins |>
+  summarize(
+    body_mass_g_mean=mean(body_mass_g, na.rm=TRUE),
+    body_mass_g_sd=sd(body_mass_g, na.rm=TRUE),
+    body_mass_g_min=min(body_mass_g, na.rm=TRUE),
+    body_mass_g_max=max(body_mass_g, na.rm=TRUE),
+    n_missing=sum(is.na(body_mass_g)))
+```
+
+Here is the code with the pretty_mean function.
+
+```{r}
+penguins |>
+  pretty_mean(body_mass_g)
+```
+
+Writing functions like these only make sense if you find yourself doing the same sort of thing three or more times in one program.
+There is a nice side effect of writing the function.
+If you decide that you want a different appearance or you want to round some of your results or you want to feed the data into a [nicely formatted table][ref15], it only takes a small modification in one spot.
+
+[ref15]: https://cran.r-project.org/web/packages/insight/vignettes/display.html

### 2025-02-11 text/25/group-by-choices.Rmd
Added group-by-choices

---
title: "New RMarkdown program"
author: "Steve Simon"
source: new
date: "2025-02-11"
categories:
- Blog post
tags:
- R programming
output: html_document
page_update: complete
---

You have an interesting choice in an R program When you use the group_by function with two or more arguments and follow that with the summarize function. Here is a practical illustration of how your choices.

<!---more--->

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)
library(tidyverse)
```

The palmer penguins library provides an interesting dataset on body measurements of several penguin species conducted by the researchers at Palmer Station, Antarctica. The group_by and summarize functions are part of the dplyr library which is one of the ones loaded when you ask for the tidyverse library.

## Illustration of group_by with two arguments followed by summarize

```{r}
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarize(n=n())
```

To simplify the output, I am removing some missing values. In a real analysis, you should not be so cavalier in just tossing out entire rows of data when one of the important variables is missing.

Notice the warning message produced by summarize. You can suppress this warning by adding warning=FALSE as a chunk option, or by specifying a .groups argument in the summarize function.

Two of the .groups options are "drop_last" and "drop". The options don't matter if you stop with summarize. If you continue onward with additional calculations, the options sometimes (but not always) become important. Here is an example where the .groups options are important.

## Calculating percentages using .groups="drop"

```{r}
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarize(n=n(), .groups="drop") |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

With the .groups="drop" option, all grouping is removed after the summarize function. This means that mutate(total=sum(n)) calculates the sum across all six of rows produced by summarize. Effectievly that produces percentages in the second mutate function will that add up to 100% across all six rows. Actually, they only add up to 99% but this is because of rounding.

## Calculating percentages using .groups="drop_last"

```{r}
penguins |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarize(n=n(), .groups="drop_last") |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

The default option for the .groups argument in summarize is .groups="drop_last".

With the .groups="drop_last" option, grouping is removed for sex, the last argument in group_by. This means that mutate(total=sum(n)) calculates the sum across each species. Effectively, this produce percentages in the second mutate function that add up to 100% across each species.

## Calculating a third set of percentages

```{r}
penguins |>
  filter(!is.na(sex)) |>
  group_by(sex, species) |>
  summarize(n=n(), .groups="drop_last") |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

If you reverse the order of the two arguments in the group_by function, you get a third option. In this case, grouping is removed for species instead of sex. This means that mutate(total=sum(n)) calculates the sum across each species. Effectively, this produce percentages in the second mutate function that add up to 100% across each sex.

## The count function works differently

The count function Notice that while the documentation on count says that

"df %>% count(a, b) is roughly equivalent to df %>% group_by(a, b) %>% summarise(n = n())"

it is not exactly equivalent. It becomes obvious first when you see that replacing group_by and summarize with count does not produce a warning message.

```{r}
penguins |>
  filter(!is.na(sex)) |>
  count(species, sex)
```

When you follow up with summarize functions to compute total and pct, you get the following.

```{r}
penguins |>
  filter(!is.na(sex)) |>
  count(species, sex) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

So count implicitly uses the .groups="drop" argument, which is not the default when you combine group_by with summarize.

The count function does not have a .groups argument. If you wanted to do something effectively the same as .groups="drop_last" you just need to insert group_by statement after count.

```{r}
penguins |>
  filter(!is.na(sex)) |>
  count(species, sex) |>
  group_by(species) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total))
```

Notice that the percentages here add up to 100% within each species.

### 2025-02-11 text/25/group-by-choices.Rmd
Added links to group-by-choices

@@
 ---
-title: "New RMarkdown program"
+title: "Choices with the group_by and summarize functions"
 author: "Steve Simon"
@@
 You have an interesting choice in an R program When you use the group_by function with two or more arguments and follow that with the summarize function.
-Here is a practical illustration of how your choices.
+Here is a practical illustration of how your choices can make a difference.
 
@@
 Notice that the percentages here add up to 100% within each species.
+
+## More information
+
+Like most of the tidyverse documentation, the information on summarize is very complete.
+Sometimes so complete as to be overwhelming.
+Look first at the summarise section in the [vignette on grouping][ref01].
+The [documentation page on count][ref02] does not go into detail about the inconsistency between the count function and the group_by followed by summary functions, but is still worth reading.
+
+[ref01]: https://dplyr.tidyverse.org/articles/grouping.html#summarise
+[ref02]: https://dplyr.tidyverse.org/reference/count.html

### 2025-03-24 src/copy-files.qmd
Added wikipedia-white-space
Updated copy-files to also read 2025 pages

@@
 author: "Steve Simon"
-format: html
+format:
+  html:
+    embed-resources: true
 ---
@@
 old_path <- "../text"
-old_directories <- c("99", paste0("0", (0:9)), 10:24)
+old_directories <- c("99", paste0("0", (0:9)), 10:25)
 new_path <- "../../qblog/posts"

### 2025-03-24 src/rmd-check.qmd
Updated copy-files to also read 2025 pages

---
title: "Check that all Rmd files render properly"
author: "Steve Simon"
format:
  html:
    embed-resources: true
---

This program was written by Steve Simon on 2024-11-09 and is placed in the public domain.

## Preliminaries

```{r}
#| label: setup
#| message: false
#| warning: false

library(glue)
library(quarto)
library(tidyverse)
```

## Set old and new locations

```{r}
#| label: locations

old_path <- "../text"
old_directories <- c(25:10, paste0("0", (9:0)), "99")
new_path <- "../../qblog/posts"
file_pattern <- "Rmd$"
```

## Find files

```{r}
#| label: find-files

rmd_files <- NULL
for (dir in old_directories) {
  glue("{old_path}/{dir}") |>
    list.files() |>
    str_subset(file_pattern) -> to_be_tested

  n_to_be_tested <- length(to_be_tested)
  message <- glue(
    "Examining {n_to_be_tested} files in {old_path}/{dir}")
  cat(message)
  cat("\n")

  rmd_files <- append(rmd_files, glue("{old_path}/{dir}/{to_be_tested}"))
}
```

## List libraries

```{r}
#| label: list-libraries

lib <- NULL
for (i in 1:length(rmd_files)) {
  read_lines(rmd_files[i]) |>
    str_subset("library\\(") |>
    append(lib) -> lib
}
lib |>
  unique() |>
  sort() -> lib
cat(lib, sep="\n")
```

## Render files

```{r}
#| message: false
#| warning: false
#| label: render-files

for (i in 1:length(rmd_files)) {
  print(glue("{i}: {rmd_files[i]}"))
  tmp <- quarto_render(input=rmd_files[i], quiet=TRUE)
}
```

### 2025-04-12 text/pending/invalid-names.qmd
Started work on nexus-bioinformatics-2025

---
title: "Working with invalid names in R"
author: "Steve Simon"
source: new
date: "2025-04-09"
categories:
- Blog post
tags:
- R programming
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: true
  echo: true
page_update: complete
---

The R programming language has a few rules about names that you can assign for various objects. This can cause some issues with data frames and tibbles that you might be importing or exporting. There is a fairly standard, but somewhat awkward way to handle this.

<!---more--->

The rules for names in R are fairly restrictive. A name cannot include blanks. It cannot include most special characters, such as the dash (-) or the slash (/). The two exceptions are the dot (.) and the underscore (_). Any combination of those two symbols plus any number (0-9) or any letter (A-Z, a-z) is fine, with one exception. You can't start a name with a number. So a1 is okay, but 1a is not.

## Invalid names during data import

You may encounter invalid names when you import data from a text file, from a spreadsheet, or from other statistical software. You can live with the invalid names with the careful use of backticks and quote marks. You can also change the names to names that are valid. I'll show both approaches, though I strongly prefer changing to valid names.

When you import a text file or a spreadsheet, the file will often include the names of each variable as the first line. These names might not respect the limitations that R imposes.

Less common, but still a problem is the names from files created by other statistical software. The rules that those programs have for variable names may not work with R.

## Use backticks or quote marks

You can still work with data where one or more variables have invalid names by surrounding the name with backticks or quote marks.

Let's work with a small file with two invalid names. First load the tidyverse library. I'll also need the gt package for the last example.

```{r}
#| label: setup
#| message: false
#| warning: false

library(gt)
library(tidyverse)
```

Read in the file using the read_csv function in the readr library (included when you loaded tidyverse) and display the results.

```{r}
#| label: read

beatles_1 <- read_csv(
  file = "invalid-names.csv",
  col_names = TRUE,
  col_types = "cc")
beatles_1
```

The results display just fine, but if you try to work with individual variables, it all falls apart.

```{r}
#| label: last-name-1

beatles_1$last-name
```

You can work with this variable by surrounding it with backticks

```{r}
#| label: last-name-2

beatles_1$`last-name`
```

or with quotes.

```{r}
#| label: last-name-3

beatles_1[ ,"last-name"]
```

Sometimes the backticks and the quote marks are interchangeable

```{r}
#| label: last-name-4

beatles_1$"last-name"
```

and sometimes they are not. With a bit of trial and error, you can figure out what works and what doesn't.

```{r}
#| label: last-name-5

beatles_1[ ,`last-name`]
```

## The rename function

Always having to surround your names with backticks or quotes get very tedious very quickly. It's a good idea to change the names right away. The rename function in dplyr is an easy way to do this.

```{r}
#| label: rename

beatles_1 |>
  rename(
    last_name = `last-name`,
    first_name = `1st-name`) -> beatles_2

beatles_2
```

You could also use the names function, part of base R.

```{r}
#| label: names

names(beatles_1) <- c("last_name", "first_name")
beatles_1
```

Once you have converted to valid names, you don't have to use backticks and quote marks.

## The name_repair argument

Most (but not all) of the functions in the readr library have a name_repair argument. The default value for the name_repair argument is "unique". This will change the name of a variable to avoid the unpleasant prospect of having two variables with the same name, but it won't fix names that fail to follow the rules of R.

You can, however, use the "universal" value for the name_repair. This will modify any variables to insure unique names, but will also slightly modify the names to avoid any conflicts with

```{r}
#| label: repair

beatles_3 <- read_csv(
  file = "invalid-names.csv",
  col_names = TRUE,
  col_types = "cc",
  name_repair = "universal")
beatles_3
```

The universal repair places dots strategically to make the variable name valid. The results may look a bit ugly at times, but that's a small price to pay.

## Replace with generic names

You can tell R to ignore the names given in the first line. If you do this, R will create generic names. Remember to skip to the second line of the file, or R will use the variable names as the first row of the data frame or tibble.

```{r}
#| label: generic

beatles_4 <- read_csv(
  file = "invalid-names.csv",
  col_names = FALSE,
  skip = 1,
  col_types = "cc")

beatles_4
```

I generally discourage the use of generic names like X1 and X2. But it does work.

## Introducing invalid names during output

There are a few times when you might want to deliberately create invalid names. One reason is that you are outputtng a dataset using a fancy format and want the headings to look nice.

You should look at the gt library, which does an awesome job of creating nice looking tables. Here's some sample code.

```{r}
#| label: gt-1

beatles_1 |>
  gt() |> 
  tab_header(
    title = "The Fab Four",
    subtitle = "In alphabetical order")
```

The table looks nice, except for the names in each column. This table would be greatly improved using mixed capitalization and spaces. You can do this with the rename function and backticks.

```{r}
#| label: gt-2

beatles_1 |>
  rename(
    `Last Name` = last_name,
    `First Name` = first_name) |>
  gt() |> 
  tab_header(
    title = "The Fab Four",
    subtitle = "In alphabetical order")
```

Now doesn't that look so much nicer!

## Summary

You might encounter invalid names for variables in a data frame or tibble when you import data. You can accommodate those invalid names using backticks and/or quote marks. It is usually better to convert to valid names. You can do this several ways.

You may sometimes want to create invalid names intentionally to make nicer looking tables. Use the rename function and backticks to accomplish this.

### 2025-05-09 text/23/mcnemar-test.Rmd
Removed brackets from generic code chunks

--- 2024-06-20 text/23/mcnemar-test.Rmd
@@
 
-```{r}
+```{r, echo=FALSE}
 a <- 833; b <- 125; c <- 2; d <- 160
@@
 
-In the survey about heaven and hell, b=`r b` and c=`r c`. The test statistic is
+In the survey about heaven and hell, b=125 and c=2. The test statistic is
 
--   $T = \frac{(125-2)^2}{127} = `r T`$
+-   $T = \frac{(125-2)^2}{127} = 119.1$
 
-If you want to enter this data into a program like SPSS, you can't type it as you see it.
+If you want to enter this data into a program like SPSS or R, you can't type it as you see it.
 
@@
 Specify the belief in heaven (yes/no) as the first column and belief in hell (yes/no) as the second column.
+Notice how the order is staggered to allow all four combinations to appear.
 The count for each combination goes in the third column.
@@
 heaven,hell,count
-1,1,833
-1,2,125
-2,1,2
-2,2,160
+"yes","yes",833
+"yes","no",125
+"no","yes",2
+"no","no",160
 ```
 
-Make sure that you designate count as a weight before starting your analysis. Here is what the output looks like in SPSS
+Make sure that you designate count as a weight before starting your analysis. In SPSS, select
 
-![](http://www.pmean.com/new-images/23/mcnemar-test-02.png)
+-   Analyze | Nonparametric Tests | Legacy Dialogs | 2 Related Samples
 
-Here is what it looks like in R
+from the menu.
 
-```{r}
-mcnemar.test(matrix(c(833, 125, 2, 160), nrow=2), correct=FALSE)
-```
+In R, use the
+
+-   mcnemar.test
+
+function.
 

### 2025-05-16 text/24/survival-sample-size-example.Rmd

---
title: Survival analysis sample size calculation
source: "New"
author: Steve Simon
date: 2024-11-26
categories: Blog post
tags:
- Sample size justification
- Survival analysis
output: html_document
page_update: no
---

Someone asked about getting a sample size for a survival analysis project. Here is the scenario. This is a two group comparison. The control group is expected to see 75% survival at one year. The treatment group is expected to do better, with a hazard ratio of 0.8. Patients are going to be recruited for a full calendar year and then accrual will stop. Patients will continue to be followed for another two years. This means that the patients recruited on the first day of the study will be followed for three full years and patients recruited on the last day of accrual will be followed for two full years.

<!---more--->

Let's start with a simple model. Maybe too simple, but you have to start somewhere. Assume that events occur with an exponential distribution. The cumulative density function of the exponential distribution is

-   $F(x)=1-e^{-\lambda x}$

The value of lambda for the control group has to satisfy the equation

-   $1-e^{-\lambda 1} = 0.25$

With a bit of algebra, you get

-   $\lambda$ = - ln(0.75) = 0.2876

I always get nervous with this, because it is really easy to make a minor error and all of a sudden, you have something that is off by a full order of magnitude. So check your work. In R, the function

-   pexp(1, rate=0.2876) = `r pexp(1, rate=0.2876)`

There are similar functions, in SAS, Stata, etc.

Now, you expect the treated group to have a hazard ratio of 0.8. For the exponential distribution, this means that they have a smaller parameter,

-   $\lambda$ = -ln(0.75)\*0.8 = `r -log(0.75)*0.8`

Now, if you follow a patient for an average of 2.5 years, what is the probability that they will experience the event?

For the control group, it is

-   $1-e^{-0.2876 \times 2.5}$ = `r 1-exp(-0.2876*2.5)`

and for the treatment group, it is

-   $1-e^{-0.2301 \times 2.5}$ = `r 1-exp(-0.2301*2.5)`

Now power is driven by the number of events, not the number of patients. How many events would you need to see? The formula found in Schoenfeld (1983)

Schoenfeld, D. A. (1983). Sample-size formula for the proportional-hazards regression model. Biometrics, 499-503. Available in [html format][sch1] is

-   $\frac{(z_\beta + z_{1-alpha/2})^2}{P_A P_B (ln(\Delta))^2}$

where $P_A$ and $P_B$ are the proportions in the control and treatment groups and $\Delta$ is the hazard ratio. Assuming equal sample sizes in each group, you would get

-   $\frac{(0.84 + 1.96)^2}{0.25*(ln(0.8))^2}$ = `r (1.96+0.84)^2/(0.25*log(0.8)^2)`

Now how many patients (not events) would you need? Solve the equation

-   0.5128 n + 0.4374 n = 630

to get

-   n = `r 630/(0.5128+0.4374)`

which is the number of patients in each group.

[sch1]: https://www.biostat.wisc.edu/~chappell/641/papers/paper31.pdf
